### 2021-09-14 ReadingGroupPres.tex
Add files

\documentclass[10pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{multicol}
\usepackage{booktabs}
\usepackage[scale=2]{ccicons}
\usepackage[style=authoryear]{biblatex}
\bibliography{demo.bib}
\usepackage{pgfplots}

\usepgfplotslibrary{dateplot}

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

\title{Hospital Information Technology and Physician Response}

\subtitle{Hanna Glenn}
 \date{\today}
% \date{}
%\author{Presented by: Hanna Glenn}
% \institute{Center for modern beamer themes}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

\begin{document}

\maketitle

\setbeamercolor{background canvas}{bg=white}

\begin{frame}{Table of contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents%[hideallsubsections]
\end{frame}

\section[Motivation]{Motivation}

\begin{frame}{The Beginning of HIT}
\centering
    \includegraphics[scale=.6]{graphics/News Clip4.PNG}
\end{frame}

\begin{frame}[fragile]{HIT has had Great (Expected) Potential in Healthcare}
\begin{alertblock}{Cost Saving}
\begin{itemize}
    \item A 2005 estimate finds possible cost reduction of hundreds of billions of dollars (\cite{hillestad2005})
\end{itemize}
\end{alertblock}

\begin{alertblock}{Quality Improvement}
\begin{itemize}
    \item Patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
    \item Significant policy push for EHR implementation with this goal in mind: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
\end{itemize}
\end{alertblock}

\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} (\cite{stats})

\end{frame}

\begin{frame}[fragile]{What does this mean for physicians?}
\begin{itemize}
    \item Day-to-day, this means physicians spend more time entering information into a computer
    \item Loss of autonomy as EHRs have progressed
    \item Physicians have experienced significant burn-out from EHRs
\end{itemize}

\end{frame}

\begin{frame}{Physician Burnout}
\begin{center}
    \includegraphics[scale=.4]{graphics/News Clip1.PNG}
\end{center}
\end{frame}

\begin{frame}{Physician Burnout}
\begin{center}
    \includegraphics[scale=.5]{graphics/News Clip2.PNG}
\end{center}
\end{frame}

\begin{frame}{Physician Burnout}
\begin{center}
    \includegraphics[scale=.3]{graphics/News Clip3.PNG}
\end{center}
\end{frame}

\begin{frame}{This Paper}

Does EHR implementation affect labor market decisions of physicians, specifically where, how much, and whether or not they work?

\end{frame}

\section{Contribution}

\begin{frame}{Branches of Literature}
\centering
\includegraphics[scale=.45]{graphics/Contribution_litgraphic.PNG}

\end{frame}

\begin{frame}{Physician Labor Literature}
\centering
    \includegraphics[scale=.45]{graphics/labor_litgraphic.PNG}
\end{frame}

\begin{frame}{EHR Literature}
    \centering
    \includegraphics[scale=.45]{graphics/EHR_litgraphic.PNG}
\end{frame}

\section{Model}

\begin{frame}{Physician's Incentives}
\begin{itemize}
    \item Physician decides to work, possibly at a handful of hospitals
    \item Some proportion of these implement a new technology (an EHR)
    \item Two ways to think about the effect:
    \begin{enumerate}
        \item The EHR could impose a burden on the physician, leading them to shift work to other hospitals or lower workload
        \item The EHR could increase the physician's attachment to the implementing hospital, leading them to shift away from other hospitals and towards the EHR hospital
    \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}{Physician's Incentives: reduced form}
\begin{equation*}
    share_h=f()
\end{equation*}

\end{frame}

\begin{frame}{Variation in EHR Adoption}
\centering
\includegraphics[scale=.55]{Objects/TYP_plot_hospEHR_year.pdf}
\end{frame}

\begin{frame}{Frame Title}

\end{frame}

\section{Plans}

\end{document}

### 2021-09-14 Path.R

@@ -1,0 +1,2 @@
+created_data_path <- "/Users/hannaglenn/Documents/EHR_physicianlabor/CreatedData/"
+raw_data_path <- "/Users/hannaglenn/Documents/EHR_physicianlabor/RawData/"

### 2021-09-14 Scripts/AHANPI_cw.R

@@ -1,0 +1,328 @@
+library(dplyr)
+library(readr)
+
+# -------------------------------   Create NPI-AHA Crosswalk -----------------------------
+#                                   Hanna Glenn
+#                                   8/27/2021
+
+# This script reads in data from the AHA survey with connected NPI numbers. Many hospitals
+# will skip this question in all or some years. I will fill in any NPI from NPI lookup
+# that are missing and match in name and location. I will also observe when there are 
+# multiple NPIs connected to one AHAID and address this. The resulting dataset is 
+# "AHANPI_cw.csv". 
+
+AHA.NPI <- read_csv(paste0(path.currentdata,"AHA-NPI.csv"))
+  # 44077 obs of 10 variables
+
+AHA.NPI <- AHA.NPI %>%
+  filter(MSTATE %in% c("AL","AK","AZ","AR","CA","CO", "CT","DE","DC","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD",
+                       "MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY")) %>%
+  rename(AHAID=ID, Hospname=MNAME, Street=MLOCADDR, City=MLOCCITY, Beds=BDTOT, State=MSTATE, Zip=MLOCZIP, NPI=NPINUM) %>%
+  select(YEAR, AHAID, NPI, Hospname, Street, City, State, Zip, Beds)
+  # 43615 obs of 9 variables
+
+# Create a dataset to see how many AHA IDs are missing an NPI (number of obs gives number of unique AHA IDs)
+num_missing_npi <- AHA.NPI %>%
+  group_by(AHAID) %>%
+  filter(is.na(NPI)) %>%
+  count() 
+  # 3162 obs 
+
[…]

### 2021-09-14 Scripts/MedPhys_PUF.R

@@ -1,0 +1,52 @@
+library(dplyr)
+library(readr)
+
+# -----------------------------   Read in and Clean Provider Utilization and Payment Files ------------------------
+#                                 Hanna Glenn, Emory University
+#                                 8/31/2021
+
+
+# This script reads in the "Medicare Physician and Other Supplier PUF" data from 2012-2015.
+# The data will provide the dependent variable for my third year paper. 
+
+# Read in Provider Utilization and Payment Files ----------------------------------------------------------------------
+for (t in 2012:2013){
+  PartByear <- read_tsv(paste0(path.currentdata,"Medicare_Provider_Util_Payment_PUF_CY",t,".txt"))
+  PartByear <- PartByear %>%
+    select(NPI,HCPCS_CODE,BENE_UNIQUE_CNT,BENE_DAY_SRVC_CNT) %>%
+    rename(num_services=BENE_UNIQUE_CNT,num_services_day=BENE_DAY_SRVC_CNT) %>%
+    filter(NPI>1) %>%
+    mutate(year=t)
+  
+  assign(paste0("PartB",t),PartByear)
+}
+
+for (t in 2014:2015){
+  PartByear <- read_tsv(paste0(path.currentdata,"Medicare_Provider_Util_Payment_PUF_CY",t,".txt"))
+  PartByear <- PartByear %>%
+    select(npi,hcpcs_code,bene_unique_cnt,bene_day_srvc_cnt) %>%
+    rename(num_services=bene_unique_cnt,num_services_day=bene_day_srvc_cnt, NPI=npi, HCPCS_CODE=hcpcs_code) %>%
+    filter(NPI>1) %>%
+    mutate(year=t)
+  
+  assign(paste0("PartB",t),PartByear)
+}
+
+MedPhys_PUF <- rbind(PartB2012, PartB2013, PartB2014, PartB2015)
+
+MedPhys_PUF <- MedPhys_PUF %>%
+  distinct()
+
+# I need to create an aggregated version so that each NPI has one utilization measure
+MedPhys_PUF <- MedPhys_PUF %>%
+  group_by(NPI,year) %>%
+  mutate(total_services=sum(num_services),NPI=as.numeric(NPI)) %>%
+  mutate(total_services=ifelse(total_services==0,NA,total_services)) %>%
+  select(year,NPI,total_services) %>%
+  distinct()
+
[…]

### 2021-09-14 Scripts/TYP_sumstats.R

@@ -1,0 +1,164 @@
+library(tidyverse)
+library(reshape)
+library(knitr)
+library(kableExtra)
+
+options(knitr.kable.NA=" ")
+cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+# ------------------  Summary Stats and Figures for Third Year Paper -----------------------
+#                     Hanna Glenn, Emory University
+#                     8/27/2021
+
+# This script creates preliminary summary stats and figures to explore relationships
+# between EHR and physician decisions. Some of these will be used in my third year paper.
+# The data used is created in "ThirdYearPaper.R"
+
+# Read in ThirdYearPaper.rds
+ThirdYearPaper <- read_rds(paste0(created_data_path,"ThirdYearPaper.rds"))
+
+# General Summary Stats Tables: Separate for Physician, Hospital, Pair level ----------------------------------------------------------------
+
+# Physician Level
+TYP_sumstats_physician <- ThirdYearPaper %>% ungroup() %>%
+  summarise_at(c("Total Services (Part B)"="total_services","Number of Hospitals Worked With"="num_hospitals"), 
+               list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))), na.rm=TRUE) %>%
+  mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
+  gather(key=var,value=value) %>%
+  extract(col="var",into=c("variable", "statistic"), regex=("(.*)_(.*)$")) %>%
+  spread(key=statistic, value=value) %>%
+  relocate(variable,n,m,sd,min,max) 
+
+knitr::kable(TYP_sumstats_physician, "latex",
+             col.names=c("Variable", "N", "Mean", "Std. Dev.", "Min", "Max"),
+             digits=2,
+             caption="Physician Level Variables",
+             booktabs=T,
+             escape=F,
+             align=c("l", "c","c","c","c","c"),
+             position="h",
+             format.args = list(big.mark = ",")) %>%
+  kable_styling( full_width=F, latex_options=c("scale_down") ) %>%
+  save_kable("objects/sumstats_physician_table.pdf", density=300)
+
+# Hospital Level
[…]

### 2021-09-14 Scripts/ThirdYearPaper.R

@@ -1,0 +1,261 @@
+library(tidyverse)
+library(readr)
+library(plm)
+
+# --------------------    Build Variables Onto Main Hospital-Physician Pair Data --------------------------
+#                         Hanna Glenn, Emory University
+#                         9/3/2021
+
+# This script develops the main dataset to be used in my third year paper.
+# First, I bring in Physician Compare for information about medical school graduation and gender. Next, I bring in 
+# MedPhys_PUF, which has information on total Part B claims for a physician from 2012-2015. 
+# For EHR information, I use the main and supplement AHA surveys and CMS data on hospitals which received meaningful use subsidies.
+# The final dataset is called "ThirdYearPaper.rds"
+
+# Read in physician hospital pairs master file
+ThirdYearPaper <- read_rds(paste0(created_data_path, "phys_hosp_pairs.rds"))
+  # This data does onto have any repeats of year, HospNPI, DocNPI
+
+# Create Experience and Gender Variables ------------------------------------------------------------
+# Read in Physician Compare
+PhysCompare <- read_csv(paste0(raw_data_path,"DAC_NationalDownloadableFile.csv"))
+
+PhysCompare <- PhysCompare %>%
+  rename(gender=gndr,grad_year=Grd_yr) %>%
+  select(NPI,gender,grad_year)
+
+# Merge
+ThirdYearPaper <- ThirdYearPaper %>%
+  left_join(PhysCompare, by=c("DocNPI"="NPI"))
+
+
+ThirdYearPaper <- ThirdYearPaper %>%
+  distinct()
+
+# Create total_services Variable --------------------------------------------------------------------
+# Read in MedPhys_PUF.R (created from Physician Compare in "MedPhys_PUF.R)
+MedPhys_PUF <- read_rds(paste0(created_data_path,"MedPhys_PUF.rds"))
+
+ThirdYearPaper <- ThirdYearPaper %>%
[…]

### 2021-09-14 Scripts/npidata.R

@@ -1,0 +1,63 @@
+library(readr)
+library(dplyr)
+
+# ------------------  Building NPI Taxonomy Dataset ------------------------------------- 
+#                     Hanna Glenn, Emory University
+#                     8/26/2021
+
+
+# This document uses data from National Uniform Claim Committee (NUCC) on taxonomy code definitions
+# as well as the NPPES dataset on all NPI agents. This file will merge the two to result in a 
+# dataset called "npidata.csv", which contains all NPI agents matched to their entity type and other
+# characteristics that come from their tax code. 
+
+
+# First, we read in the NUCC taxonomy code file. 
+tax.define <- read_csv(paste0(path.currentdata,"nucc_taxonomy_210.csv"))
+tax.define <- tax.define %>%
+  select(Code, Grouping) %>%
+  rename(t_code=Code,group=Grouping)
+  # This should have 864 obs of 2 variables (as of 8/26/2021)
+
+# Now we read in the NPI data which contains taxonomy code of each NPI
+# For the current project I don't need it, but the file also contins enumeration date of the NPI number
+npi.taxcodes <- read_csv(paste0(path.currentdata,"npidata_pfile.csv"),
+                         col_types = cols(NPI=col_character()))
+npi.taxcodes <- npi.taxcodes %>%
+  rename(npi=NPI,enumdate='Provider Enumeration Date',entitytype='Entity Type Code',t_code='Healthcare Provider Taxonomy Code_1',zip='Provider Business Practice Location Address Postal Code', state='Provider Business Practice Location Address State Name') %>%
+  select(npi,t_code,zip,state)
+  # Should be 6843788 obs of 4 variables
+
+# Merge these two datasets 
+npi.data <- merge(tax.define,npi.taxcodes,by="t_code")
+
+# Filter this data to only the types of NPI I need for my question (How does EHR implementation affect
+# working decicisions of physicians?). I only need hospitals and physicians. These are defined in the
+# group names. 
+# I may use the indicators for specific entities later on
+npi.data <- npi.data %>%
+  filter(str_detect(t_group,fixed("hospital", ignore_case=TRUE)) | str_detect(t_group,fixed("physician", ignore_case=TRUE))) %>%
+  mutate(GeneralAcuteHosp=ifelse(str_detect(t_code, "^282N"),1,0)) %>%
[…]

### 2021-09-14 Scripts/phys_hosp_pairs.R

@@ -1,0 +1,107 @@
+library(tidyr)
+library(readr)
+library(stringr)
+library(dplyr)
+library(plm)
+library(janitor)
+
+
+# ------------------------------------- Building Balanced Physician Hospital Pairs ------------------------------------
+#                                       Hanna Glenn, Emory University
+#                                       9/2/2021
+
+# This script reads in (1) the CMS Physician Shared Patient Data from 2009-2015, (2) phys_npidata and hosp_npidata that were
+# created in "npidata.R". These data are all merged to end up with physician hospital pairs and their shared count. 
+# The resulting dataset, called "phys_hosp_pars.rds" consists of a balanced dataset of all physician, hospital pairs 
+# from 2009-2015.
+
+# Read in npi datasets and combine into one------------------
+phys_npidata <- read_rds(paste0(path.currentdata,"phys_npidata.rds"))
+hosp_npidata <- read_rds(paste0(path.currentdata,"hosp_npidata.rds"))
+
+temp_npidata <- bind_rows(phys_npidata,hosp_npidata) 
+temp_npidata <- temp_npidata %>%
+  select(npi,PCP,hospital)
+
+# Read in Shared Patient Data, merge to temp_npi along the way ----------------------
+for (t in 2009:2015){
+  SP.year <- read_csv(paste0(path.currentdata,"pspp",t,"_90.csv"),
+                      col_types=cols(npi1=col_character(),
+                                     npi2=col_character()
+                      ))
+  SP.year <- SP.year %>%
+    select(npi1,npi2,year,samedaycount) %>%
+    distinct()
+  
+  SP.year<- SP.year %>%
+    left_join(temp_npidata, by=c("npi1"="npi")) %>%
+    rename(PCP1=PCP,hospital1=hospital) 
+  
+  #filter for npi1 being either hospital or PCP
+  SP.year <- SP.year %>%
+    filter(PCP1==1 | hospital1==1)
+  
+  SP.year <- SP.year %>%
+    left_join(temp_npidata, by=c("npi2"="npi")) %>%
+    rename(PCP2=PCP,hospital2=hospital) %>%
[…]

### 2021-09-15 ReadingGroupPres.tex
Updates from Overleaf

@@
 
-\begin{frame}{Table of contents}
+\begin{frame}{Table of Contents}
   \setbeamertemplate{section in toc}[sections numbered]
@@
 \begin{itemize}
-    \item Patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
-    \item Significant policy push for EHR implementation with this goal in mind: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
+    \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
+    \item Significant policy push for EHR implementation with these goal in mind: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
 \end{itemize}
@@
 
-\begin{frame}{Physician Burnout}
+\begin{frame}[noframenumbering]{Physician Burnout}
 \begin{center}
@@
 
-Does EHR implementation affect labor market decisions of physicians, specifically where, how much, and whether or not they work?
+Does EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work??
 
@@
 
-\section{Model}
+\section{Model and Data}
 
@@
     \begin{enumerate}
-        \item The EHR could impose a burden on the physician, leading them to shift work to other hospitals or lower workload
-        \item The EHR could increase the physician's attachment to the implementing hospital, leading them to shift away from other hospitals and towards the EHR hospital
+        \item The EHR could impose a burden (cost) on the physician, leading them to shift work to other hospitals or lessen their own workload
+        \item The EHR could increase the physician's attachment to the implementing hospital (through the learning of a complicated new skill), leading them to lessen work at other hospitals and increase work at EHR hospital
     \end{enumerate}
@@
 
-\begin{frame}{Physician's Incentives: reduced form}
+\begin{frame}{Physician's Incentives: Reduced Form}
 \begin{equation*}
-    share_h=f()
+    p_{iht}=\textbf{p}()
 \end{equation*}

### 2021-09-16 ReadingGroupPres.tex
Updates from Overleaf

@@
 
-Does EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work??
+Does EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work?
 
@@
     \item Some proportion of these implement a new technology (an EHR)
-    \item Two ways to think about the effect:
+    \item Three ways to think about the effect:
     \begin{enumerate}
@@
         \item The EHR could increase the physician's attachment to the implementing hospital (through the learning of a complicated new skill), leading them to lessen work at other hospitals and increase work at EHR hospital
+        \item Other factors impact physicians more than the hospital's technology, so they do not change their behavior when EHRs are implemented
     \end{enumerate}
@@
 
-\begin{frame}{Physician's Incentives: Reduced Form}
+\begin{frame}{Relevant Data}
     \begin{equation*}
-    p_{iht}=\textbf{p}()
+    p_{iht}=\textbf{p}(EHR_{ht},EHR_{-h,t},X_i,H_j,...)
+\end{equation*}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Relevant Data}
+\begin{equation*}
+    \textcolor{blue}{p_{iht}}=\textbf{p}(EHR_{ht},EHR_{-h,t},X_i,H_j,...)
 \end{equation*}
 
+$\textcolor{blue}{p_{iht}}:$
+\begin{itemize}
+    \item measures of physician's labor
+    \item number of shared patients with a hospital: CMS PSPD (2009-2015)
+    \item total services: Medicare Provider Utilization and Payment Data (starts in 2012)
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Relevant Data}
+\begin{equation*}
+    p_{iht}=\textbf{p}(\textcolor{red}{EHR_{ht},EHR_{-h,t}},X_i,H_j,...)
+\end{equation*}
+
+$\textcolor{red}{EHR}:$
+\begin{itemize}
+    \item Main variable (binary for whether a hospital uses an EHR) comes from AHA survey
+    \item AHA IT Supplement also has information on EHRs that can help with documentation or decision making
+    \item Meaningful use data gives information on which hospitals received a subsidy for achieving "meaningful use"
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Relevant Data}
+\begin{equation*}
+    p_{iht}=\textbf{p}(EHR_{ht},EHR_{-h,t},\textcolor{purple}{X_i},H_j,...)
+\end{equation*}
+
+$\textcolor{purple}{X_i}:$
+\begin{itemize}
+    \item Physician level characteristics
+    \item Medical school graduation date, gender: Physician Compare
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Relevant Data}
+\begin{equation*}
+    p_{iht}=\textbf{p}(EHR_{ht},EHR_{-h,t},X_i,\textcolor{orange}{H_{jt}},...)
+\end{equation*}
+
+$\textcolor{orange}{H_{jt}}:$
+\begin{itemize}
+    \item Hospital Level Characteristics
+    \item Number of beds: AHA Survey
+    \item Number of days out of the year operating: AHA Survey
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Summary Statistics}
+\centering
+\includegraphics[scale=.45]{Objects/sumstats_physician_table.pdf}\\
+\vspace{3mm}
+\includegraphics[scale=.45]{Objects/sumstats_pair_table.pdf}\\
+\vspace{3mm}
+\includegraphics[scale=.45]{Objects/sumstats_hospital_table.pdf}
 \end{frame}
@@
 
-\begin{frame}{Frame Title}
+\begin{frame}{Variation in Working Variables}
+    \centering
+    \includegraphics[scale=.5]{Objects/mainhosp_share_treatvscontrol.pdf}
+\end{frame}
 
+\begin{frame}{Variation in Meaningful Use}
+    \centering
+    \includegraphics[scale=.5]{Objects/TYP_plot_hospmeanuse_year.pdf}
+\end{frame}
+
+\begin{frame}{Variation in Working Variables}
+\centering
+\includegraphics[scale=.5]{Objects/hhi_treatvscontrol.pdf}
+\end{frame}
+
+\begin{frame}{Variation in Working Variables}
+    \centering
+    \includegraphics[scale=.4]{Objects/rel_mainhosp_share_treatvscontrol.pdf}
 \end{frame}
@@
 
+\begin{frame}{Estimation}
+First, DD
+\begin{itemize}
+    \item Parallel trends assumption
+    \item EHR implementation exogenous to physician?
+    \item Implement all the current DD techniques
+\end{itemize}
+
+Later, IV
+\begin{itemize}
+    \item State level HIPPA privacy laws
+    \item Bandwidth capacity of a given area
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Other Plans}
+\begin{itemize}
+    \item Dig deeper into meaningful use measures
+    \item Try to get location measurements
+\end{itemize}
+
+\end{frame}
+
 \end{document}

### 2021-09-16 ReadingGroupPres.tex
update slides
slide changes

--- 2021-09-15 ReadingGroupPres.tex
@@
 \begin{itemize}
-    \item Day-to-day, this means physicians spend more time entering information into a computer 
+    \item Day-to-day, this means physicians spend more time entering information into a computer and learning a new software
     \item Loss of autonomy as EHRs have progressed
@@
     \begin{enumerate}
-        \item The EHR could impose a burden (cost) on the physician, leading them to shift work to other hospitals or lessen their own workload
-        \item The EHR could increase the physician's attachment to the implementing hospital (through the learning of a complicated new skill), leading them to lessen work at other hospitals and increase work at EHR hospital
+        \item The cost of switching hospitals is higher (attachment to EHR hospital)
+        \item The cost of switching hospitals is lower (increased burden at EHR hospital)
     \end{enumerate}
@@
 
-\begin{frame}{Physician's Incentives: Reduced Form}
+\begin{frame}{Data}
 \begin{equation*}
-    p_{iht}=\textbf{p}()
+    \textcolor{blue}{p_{iht}}=f(EHR_{ht},EHR_{-h,t},X_i,H_{jt},...)
 \end{equation*}
 
+\vspace{3mm}
+
+Measures of physician labor market behavior
+\begin{itemize}
+    \item Number of shared patients with a hospital: CMS PSPD (2009-2015)
+    \item Total services: Medicare Provider Utilization and Payment Data (starts in 2012)
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Data}
+\begin{equation*}
+    p_{iht}=f(\textcolor{blue}{EHR_{ht},EHR_{-h,t}},X_i,H_{jt},...)
+\end{equation*}
+
+\vspace{5mm}
+
+\begin{itemize}
+    \item Main variable (binary for whether a hospital uses an EHR) comes from AHA survey
+    \item AHA IT Supplement also has information on EHRs that can help with documentation or decision making
+    \item Meaningful use data gives information on which hospitals received a subsidy for achieving "meaningful use"
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Data}
+\begin{equation*}
+    p_{iht}=f(EHR_{ht},EHR_{-h,t},\textcolor{blue}{X_i},H_{jt},...)
+\end{equation*}
+
+\vspace{5mm}
+
+Physician level characteristics
+\begin{itemize}
+    \item Medical school graduation date, gender: Physician Compare
+\end{itemize}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Data}
+\begin{equation*}
+    p_{iht}=f(EHR_{ht},EHR_{-h,t},X_i,\textcolor{blue}{H_{jt}},...)
+\end{equation*}
+
+\vspace{5mm}
+
+Hospital Level Characteristics
+\begin{itemize}
+    \item Number of beds: AHA Survey
+    \item Number of days out of the year operating: AHA Survey
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Summary Statistics}
+\centering
+\includegraphics[scale=.45]{Objects/sumstats_physician_table.pdf}\\
+\vspace{5mm}
+\includegraphics[scale=.45]{Objects/sumstats_pair_table.pdf}\\
+\vspace{3mm}
+\includegraphics[scale=.45]{Objects/sumstats_hospital_table.pdf}
 \end{frame}
@@
 
-\begin{frame}{Frame Title}
+\section{Plans}
+
+\begin{frame}{Original Plan: Difference in Difference}
+\begin{itemize}
+    \item Treatment: physicians who are exposed to an EHR at one of their hospitals in some year
+    \item Control: physicians who are not exposed to a hospital at \textbf{any} hospital in any year
+    \begin{itemize}
+        \item  Issues with control group because of spillovers
+    \end{itemize}
+    \item Could still try to attempt this by imposing a control group where the physician has not \textbf{yet} been exposed to an EHR at any hospital
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Other Ideas}
+\begin{itemize}
+    \item Aggregate to physician level dataset and do analysis more along the lines of the graph I showed you earlier
+    \item Have a continuous measure of EHR use that captures share of exposure to EHRs
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Miscellaneous Points}
+\begin{itemize}
+    \item Want to dig deeper into meaningful use measures
+    \item Try to get location measurements to think about where physicians are working
+    \item Bring in other measures of work from shared patient data
+\end{itemize}
 
@@
 
-\section{Plans}
+\begin{frame}{Variation in Meaningful Use}
+    \centering
+    \includegraphics[scale=.5]{Objects/TYP_plot_hospmeanuse_year.pdf}
+\end{frame}
 

### 2021-09-17 ReadingGroupPres.tex
Updates from Overleaf

@@
 
-\begin{frame}{The Beginning of HIT}
+\begin{frame}{The Potential of HIT}
 \centering
@@
 
-\begin{frame}[fragile]{HIT has had Great (Expected) Potential in Healthcare}
+\begin{frame}[fragile]{HIT had Great (Expected) Potential in Healthcare}
 \begin{alertblock}{Cost Saving}
@@
 \begin{itemize}
-    \item Day-to-day, this means physicians spend more time entering information into a computer and learning a new software
+    \item Initially, this means physicians spend more time entering information into a computer and learning a new software
     \item Loss of autonomy as EHRs have progressed
@@
 \begin{center}
-    \includegraphics[scale=.3]{graphics/News Clip3.PNG}
+    \includegraphics[scale=.45]{graphics/News Clip3.PNG}
 \end{center}
@@
 \centering
-\includegraphics[scale=.45]{graphics/Contribution_litgraphic.PNG}
+\includegraphics[scale=.5]{graphics/Contribution_litgraphic.PNG}
 
@@
 \begin{itemize}
-    \item Number of shared patients with a hospital: CMS PSPD (2009-2015)
-    \item Total services: Medicare Provider Utilization and Payment Data (starts in 2012)
+    \item Number/percent of shared patients physician $i$ has with hospital $h$: \underline{CMS Shared Patient Data} (2009-2015)
+    \item Total Part B claims for physician $i$: \underline{Medicare Provider Utilization and Payment Data} (starts in 2012)
 \end{itemize}
@@
     \item AHA IT Supplement also has information on EHRs that can help with documentation or decision making
-    \item Meaningful use data gives information on which hospitals received a subsidy for achieving "meaningful use"
+    \item (\textit{Considering Later}) Meaningful use data gives information on which hospitals received a subsidy for achieving "meaningful use"
 \end{itemize}
@@
 \begin{itemize}
-    \item Medical school graduation date, gender: Physician Compare
+    \item Medical school graduation date, gender: \underline{Physician Compare}
 \end{itemize}
@@
 \centering
-\includegraphics[scale=.45]{Objects/sumstats_physician_table.pdf}\\
-\vspace{5mm}
-\includegraphics[scale=.45]{Objects/sumstats_pair_table.pdf}\\
+
+\includegraphics[scale=.8]{Objects/sumstats_pair_table.pdf}
+
 \vspace{3mm}
-\includegraphics[scale=.45]{Objects/sumstats_hospital_table.pdf}
+
+\includegraphics[scale=.8]{Objects/sumstats_physician_table.pdf}
+
+\end{frame}
+
+\begin{frame}{Summary Statistics}
+\centering
+\includegraphics[scale=.8]{Objects/sumstats_hospital_table.pdf}
+
 \end{frame}
@@
 \centering
-\includegraphics[scale=.55]{Objects/TYP_plot_hospEHR_year.pdf}
+\includegraphics[scale=.53]{Objects/TYP_plot_hospEHR_year.pdf}
+\end{frame}
+
+\begin{frame}{Looking at Shares when EHRs are Adopted}
+    \centering
+    \includegraphics[scale=.5]{Objects/relyear_treatment.pdf}
 \end{frame}
@@
 
-\begin{frame}{Original Plan: Difference in Difference}
+\begin{frame}{Initial Plan: Difference in Difference}
 \begin{itemize}
-    \item Treatment: physicians who are exposed to an EHR at one of their hospitals in some year
-    \item Control: physicians who are not exposed to a hospital at \textbf{any} hospital in any year
-    \begin{itemize}
-        \item  Issues with control group because of spillovers
-    \end{itemize}
-    \item Could still try to attempt this by imposing a control group where the physician has not \textbf{yet} been exposed to an EHR at any hospital
+    \item Interested in two treatment effects
+    \begin{enumerate}
+        \item The effect of EHR at hospital $h$ on share at hospital $h$
+        \item The effect of EHR at hospital $h$ on share at hospital $-h$
+    \end{enumerate}
+
+    \vspace{2mm}
+
+    \item My idea is to estimate these two separately, but there are issues because of spillover effects (if hospital $-h$ implements an EHR)
+
+    \vspace{2mm}
+
+    \item Control group will be physicians who have not yet been exposed to an EHR at \textbf{any} hospital at time $t$
 \end{itemize}
@@
 \begin{itemize}
-    \item Aggregate to physician level dataset and do analysis more along the lines of the graph I showed you earlier
+    \item Aggregate to physician level dataset if there is insufficient data for the control group
     \item Have a continuous measure of EHR use that captures share of exposure to EHRs
+    \item IV: bandwidth, state privacy laws
 \end{itemize}
@@
 
-\begin{frame}{Miscellaneous Points}
+\begin{frame}{Other Points}
 \begin{itemize}
-    \item Want to dig deeper into meaningful use measures
-    \item Try to get location measurements to think about where physicians are working
+    \item Want to dig deeper into how a hospital qualified for meaningful use subsidies
+    \item Physician location
     \item Bring in other measures of work from shared patient data
@@
 
-\begin{frame}{Variation in Meaningful Use}
-    \centering
-    \includegraphics[scale=.5]{Objects/TYP_plot_hospmeanuse_year.pdf}
-\end{frame}
-
 \end{document}

### 2021-10-13 Scripts/TYP_Analysis.R
add event study objects

@@ -18,7 +18,7 @@ Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 # Event Studies for Entire Sample of Doctors ------------------------------------------------------
 # Event Study for Continuous Working Variable
 event_reg <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
-                  rel_p4 + rel_p5 + rel_p6 + avg_beds + avg_oper_days |year + DocNPI,
+                  rel_p4 + rel_p5 + rel_p6 + female + experience + avg_beds + avg_oper_days |year,
                   data=Physician_Data)
 
 event_reg_coef <- as_tibble(event_reg$coefficients, rownames="term") %>%
@@ -26,7 +26,7 @@ event_reg_coef <- as_tibble(event_reg$coefficients, rownames="term") %>%
                      "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
   dplyr::rename(estimate=phys_working)
 
-event_reg_ci <- as_tibble(confint(event_reg), rownames="term") %>%
+event_reg_ci <- as_tibble(confint.default(event_reg), rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
                      "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
   dplyr::rename(conf.low='2.5 %', conf.high='97.5 %')
@@ -40,11 +40,10 @@ event_plot_table <- event_reg_coef %>%
   arrange(rel_year)
 
 event_plot <- dwplot(event_plot_table, vline=geom_vline(xintercept=0, linetype=2),
-                     order_vars=c("rel_p6", "rel_p5", "rel_p4", "rel_p3", "rel_p2", "rel_p1", "rel_0",
-                                  "rel_m1", "rel_m2", "rel_m3", "rel_m4", "rel_m5", "rel_m6"),
                      whisker_args=list(color="black", size=1.1),
+                     vars_order =c("rel_p6", "rel_p5", "rel_p4","rel_p3","rel_p2","rel_p1","rel_0","rel_m1","rel_m2","rel_m3","rel_m4","rel_m5","rel_m6"),
                      dot_args=list(color="black")) +
[…]

### 2021-10-13 Paper/Paper.tex

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,hyperref,array,xcolor,multicol,verbatim,mathpazo}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{graphicx}
\usepackage{fullpage}

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\onehalfspacing
\addtolength{\skip\footins}{2pc plus 5pt}

\title{Physician Response to EHR Implementation in Hospitals}
\author{Hanna Glenn}
%\date{\today}

\DeclareLabeldate[online]{%
  \field{date}
  \field{year}
  \field{eventdate}
  \field{origdate}
  \field{urldate}
}

\begin{document}

\maketitle

\vspace{1.5cm}

Electronic Health Records (EHRs) have become increasingly relevant in the US since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on cost and quality of care. In particular, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). The improvement in quality of care was expected due to physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance that could help physicians avoid harmful mistakes. Such expectations on cost and quality led policymakers to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}). This legislation subsidized hospitals which used EHRs “meaningfully”\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The push towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).

Specifically, an EHR is digital version of a patient’s medical records. These include detailed accounts and notes of medical history and can become advanced enough to provide the physician suggestions for care. Physicians are the primary users of this technology and have a significant role in determining whether potential benefits of EHRs are realized. However, the day-to-day life of physicians changed drastically with the implementation of EHRs. What was expected to be a helpful tool in general has turned out to be burdensome and frustrating for the persons using it. In individual interviews, multiple physicians reported that when using EHRs, they are less satisfied with their job and have both higher burnout and stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). A physician with the capability to stop working or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high. This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market behaviors, such as where to work, how much to work, and whether to work at all.

Using CMS Shared Patient Data from 2009-2015, I construct a panel of physician-hospital pairs that captures a specific working relationship between the pair, most likely the physician working in the hospital. I link the hospitals to AHA survey data for information on EHR use and then aggregate the data to the physician level. To measure the working decisions of physicians, I utilize (1) physician level information, also from the shared patient data, that captures total working life (referrals to labs, pharmacies, other physicians, etc), and (2) an office-based physician level dataset to observe whether physicians are moving from hospitals to private practice. For my analysis, I use a difference-in-difference estimation where treatment is defined as physicians' exposure to an EHR at any of the hospitals they work in. A key identifying assumption in this analysis is that hospitals choose to implement EHRs not based on any factors that are also correlated with physician working decisions.

\newpage
\printbibliography
\end{document}

### 2021-10-14 Scripts/TYP_Analysis.R
Clean up comments

@@ -15,8 +15,8 @@ library(stargazer)
 # Read in data
 Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 
-# Event Studies for Entire Sample of Doctors ------------------------------------------------------
-# Event Study for Continuous Working Variable
+# Event Studies for Entire Sample of Doctors ----------------------------------------------------------------
+# Continuous Number of Patients Variable
 event_reg <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
                   rel_p4 + rel_p5 + rel_p6 + female + experience + avg_beds + avg_oper_days |year,
                   data=Physician_Data)
@@ -104,7 +104,7 @@ event_plot_ind <- dwplot(event_plot_ind_table, vline=geom_vline(xintercept=0, li
 event_plot_ind
 ggsave("Objects/ind_ES_fullsample.pdf")
 
-# Repeat Event Studies for Old Doctors ----------------------------------------------------------------
+# Repeat Event Studies for Senior Doctors ----------------------------------------------------------------
 # Event Study for Continuous Working Variable (Old Doctors)
 event_reg_old <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
                     rel_p4 + rel_p5 + rel_p6 + female + avg_beds + avg_oper_days | year,

### 2021-10-14 Presentations/ReadingGroupPres.tex

--- 2021-09-17 ReadingGroupPres.tex
@@
 
-\section{Model and Data}
+\section{Data}
 
@@
     \item Number/percent of shared patients physician $i$ has with hospital $h$: \underline{CMS Shared Patient Data} (2009-2015)
+    \begin{itemize}
+        \item Main dataset: physician hospital level panel
+        \item Physicians and hospitals in this data work closely together, likely the physician works \textit{in} the hospital
+    \end{itemize}
     \item Total Part B claims for physician $i$: \underline{Medicare Provider Utilization and Payment Data} (starts in 2012)
@@
 
-\begin{frame}[noframenumbering]{Data}
+\begin{frame}{Data}
 \begin{equation*}
@@
 
-    \item My idea is to estimate these two separately, but there are issues because of spillover effects (if hospital $-h$ implements an EHR)
+    \item My idea is to estimate these two separately, but there are issues because of spillover effects (hospital $-h$ implements an EHR)
 

### 2021-10-15 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

--- 2021-09-17 ReadingGroupPres.tex
@@
 \usepackage{pgfplots}
+\usepackage{bbm}
 
@@
 
-\begin{frame}{The Potential of HIT}
-\centering
-    \includegraphics[scale=.6]{graphics/News Clip4.PNG}
+\begin{frame}{What is an EHR?}
+
+\begin{itemize}
+    \item Computer-program technology that hospitals purchase
+    \item Performance of EHR depends on level purchased:
+    \begin{itemize}
+        \item All include electronic patient records
+        \item More complex systems allow for decision making assistance
+        \item More recently, patient interfaces are being developed
+    \end{itemize}
+\end{itemize}
+
 \end{frame}
@@
 \begin{itemize}
-    \item Physician decides to work, possibly at a handful of hospitals
-    \item Some proportion of these implement a new technology (an EHR)
-    \item Two ways to think about the effect:
+    \item Physician has a choice set of where to work, including multiple hospitals and office-based locations
+    \item All physicians I consider work closely with at least one hospital at some point in time
+    \item Some proportion of hospitals implement a new technology (an EHR)
+    \item How to think about the effect on labor:
     \begin{enumerate}
-        \item The cost of switching hospitals is higher (attachment to EHR hospital)
-        \item The cost of switching hospitals is lower (increased burden at EHR hospital)
+        \item EHR makes seeing patients easier (or more difficult), thus the number of patients seen increases (or decreases)
+        \item EHR imposes a burden with a high cost, so the physician switches to working at an office-based location where they have more control
+        \item A senior physician could be tipped to retire early instead of taking on the cost of learning a new technology
     \end{enumerate}
@@
 
-\begin{frame}{Data}
-\begin{equation*}
-    \textcolor{blue}{p_{iht}}=f(EHR_{ht},EHR_{-h,t},X_i,H_{jt},...)
-\end{equation*}
+\begin{frame}{Measures of Physician Labor Market Decisions}
 
-\vspace{3mm}
+     Total Patients: \underline{CMS Shared Patient Data} (2009-2015)
+    \begin{itemize}
+        \item Sum of shared patients with all entities that have an NPI
+    \end{itemize}
 
-Measures of physician labor market behavior
+     Indicator for whether they are working
     \begin{itemize}
-    \item Number/percent of shared patients physician $i$ has with hospital $h$: \underline{CMS Shared Patient Data} (2009-2015)
-    \item Total Part B claims for physician $i$: \underline{Medicare Provider Utilization and Payment Data} (starts in 2012)
+        \item =1 if Total Patients $>0$
     \end{itemize}
-\end{frame}
 
-\begin{frame}[noframenumbering]{Data}
-\begin{equation*}
-    p_{iht}=f(\textcolor{blue}{EHR_{ht},EHR_{-h,t}},X_i,H_{jt},...)
-\end{equation*}
-
-\vspace{5mm}
-
+     Indicator for working in office-based location
     \begin{itemize}
-    \item Main variable (binary for whether a hospital uses an EHR) comes from AHA survey
-    \item AHA IT Supplement also has information on EHRs that can help with documentation or decision making
-    \item (\textit{Considering Later}) Meaningful use data gives information on which hospitals received a subsidy for achieving "meaningful use"
-\end{itemize}
-\end{frame}
-
-\begin{frame}[noframenumbering]{Data}
-\begin{equation*}
-    p_{iht}=f(EHR_{ht},EHR_{-h,t},\textcolor{blue}{X_i},H_{jt},...)
-\end{equation*}
-
-\vspace{5mm}
-
-Physician level characteristics
-\begin{itemize}
-    \item Medical school graduation date, gender: \underline{Physician Compare}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[noframenumbering]{Data}
-\begin{equation*}
-    p_{iht}=f(EHR_{ht},EHR_{-h,t},X_i,\textcolor{blue}{H_{jt}},...)
-\end{equation*}
-
-\vspace{5mm}
-
-Hospital Level Characteristics
-\begin{itemize}
-    \item Number of beds: AHA Survey
-    \item Number of days out of the year operating: AHA Survey
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Summary Statistics}
-\centering
-
-\includegraphics[scale=.8]{Objects/sumstats_pair_table.pdf}
-
-\vspace{3mm}
-
-\includegraphics[scale=.8]{Objects/sumstats_physician_table.pdf}
-
-\end{frame}
-
-\begin{frame}{Summary Statistics}
-\centering
-\includegraphics[scale=.8]{Objects/sumstats_hospital_table.pdf}
-
-\end{frame}
-
-\begin{frame}{Variation in EHR Adoption}
-\centering
-\includegraphics[scale=.53]{Objects/TYP_plot_hospEHR_year.pdf}
-\end{frame}
-
-\begin{frame}{Looking at Shares when EHRs are Adopted}
-    \centering
-    \includegraphics[scale=.5]{Objects/relyear_treatment.pdf}
-\end{frame}
-
-\section{Plans}
-
-\begin{frame}{Initial Plan: Difference in Difference}
-\begin{itemize}
-    \item Interested in two treatment effects
-    \begin{enumerate}
-        \item The effect of EHR at hospital $h$ on share at hospital $h$
-        \item The effect of EHR at hospital $h$ on share at hospital $-h$
-    \end{enumerate}
-
-    \vspace{2mm}
-
-    \item My idea is to estimate these two separately, but there are issues because of spillover effects (if hospital $-h$ implements an EHR)
-
-    \vspace{2mm}
-
-    \item Control group will be physicians who have not yet been exposed to an EHR at \textbf{any} hospital at time $t$
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Other Ideas}
-\begin{itemize}
-    \item Aggregate to physician level dataset if there is insufficient data for the control group
-    \item Have a continuous measure of EHR use that captures share of exposure to EHRs
-    \item IV: bandwidth, state privacy laws
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Other Points}
-\begin{itemize}
-    \item Want to dig deeper into how a hospital qualified for meaningful use subsidies
-    \item Physician location
-    \item Bring in other measures of work from shared patient data
+        \item Comes from SK$\&$A database of office-based physicians
     \end{itemize}
@@
 
+\begin{frame}{Measures of EHR Use}
+
+Physician-level Indicator that captures exposure to \textit{any} EHR
+\begin{itemize}
+    \item With information on shared patients between physicians and hospitals, I limit to physicians working in hospitals
+    \item Fraction of hospitals worked with that use an EHR comes from the AHA survey
+    \item Similarly, I can capture hospitals that use EHRs for decision making vs. documentation using the AHA Survey IT Supplement 
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Other Characteristics}
+
+Medical school graduation date, gender: \underline{Physician Compare}
+
+\vspace{5mm}
+
+Average size of hospital worked with (measured in beds), average hospital days operating: \underline{AHA Survey}
+\end{frame}
+
+\section{Analysis}
+
+\begin{frame}{Event Studies}
+\begin{equation*}
+    Y_{it}= \sum_{t=-6}^6 \mathbbm{1}_{\{\text{relative treatment year}_{i}=t\}} + \alpha_t + \delta_{i} + \mathbb{X}_{it}
+\end{equation*}
+
+\end{frame}
+
 \end{document}

### 2021-10-19 Scripts/Final_Pairs.R
update script and add tex object

@@ -77,6 +77,9 @@ is.pbalanced(balance_check)
 # I need to think about adding more hospital characteristics to this dataset
 AHAmainsurvey <- read_csv(paste0(raw_data_path,"AHA_mainsurvey.csv"))
 
+AHAmainsurvey <- AHAmainsurvey %>%
+  mutate(ID=as.character(ID))
+
 Final_Pairs <- Final_Pairs %>% ungroup() %>%
   mutate(AHAID=as.character(AHAID)) %>%
   left_join(AHAmainsurvey,by=c("AHAID"="ID","year"="YEAR"),na_matches="never") %>%

### 2021-10-19 Scripts/Final_Pairs.R
update regression tables

@@ -390,6 +390,27 @@ Physician_Data <- Physician_Data %>%
          rel_p6=1*(rel_exposedyr==6)) %>%
   mutate(rel_m1=ifelse(minyr_EHR==0,1,rel_m1))
 
+# Create indicators for exposure and "post_year" to use in year-specific diff in diff 
+Physician_Data <- Physician_Data %>%
+  mutate(exposed_2010=1*(minyr_EHR==2010),
+         exposed_2011=1*(minyr_EHR==2011),
+         exposed_2012=1*(minyr_EHR==2012),
+         exposed_2013=1*(minyr_EHR==2013),
+         exposed_2014=1*(minyr_EHR==2014),
+         post_2010=1*(year>=2010),
+         post_2011=1*(year>=2011),
+         post_2012=1*(year>=2012),
+         post_2013=1*(year>=2013),
+         post_2014=1*(year>=2014))
+
+# Create an indicator for whether the physician stays working in all years
+Physician_Data <- Physician_Data %>%
+  group_by(DocNPI) %>%
+  mutate(n=sum(working_ind)) %>%
+  ungroup() %>%
+  mutate(working_allyears=1*(n==7)) %>%
+  select(-n)
+
 
 # Save the Data for Analysis -----------------------------------------------------------------
 saveRDS(Physician_Data,file=paste0(created_data_path,"Physician_Data.rds"))

### 2021-10-19 Scripts/TYP_Analysis.R

@@ -284,13 +284,192 @@ ggsave("Objects/ind_ES_subset2009.pdf")
 
 
 
+# Look more closely at what is happening each year -------------------------------------------
+
+# Treated in 2010 ----------------------------------------------------------------------------
+# Look at some graphs (comparing treated in 2010 to never treated or treated after 2010)
+ggplot(filter(Physician_Data, working_allyears==1 & (minyr_EHR==0 | minyr_EHR>2009)), aes(year,phys_working,color=factor(exposed_2010))) +
+  stat_summary(geom='line') + geom_vline(xintercept=2010) 
+ggplot(filter(Physician_Data, minyr_EHR==0 | minyr_EHR>2009), aes(year,working_ind,color=factor(exposed_2010))) +
+  stat_summary(geom='line') + geom_vline(xintercept=2010)
+  # Neither of these graphs point to anything substantial happening in 2010. For the working variable, the lines stay
+  # parallel for all of the years. For the working indicator, the lines dont break until 2015. 
+
+# Look at some graphs(comparing treated in 2010 to never treated only)
+ggplot(filter(Physician_Data, working_allyears==1 & (minyr_EHR==0 | minyr_EHR==2010)), aes(year,phys_working,color=factor(exposed_2010))) +
+  stat_summary(geom='line') + geom_vline(xintercept=2010) 
+ggplot(filter(Physician_Data, minyr_EHR==0 | minyr_EHR==2010), aes(year,working_ind,color=factor(exposed_2010))) +
+  stat_summary(geom='line') + geom_vline(xintercept=2010)
+# Neither of these graphs point to anything substantial happening in 2010. For the working variable, the lines stay
+# parallel for all of the years. The working indicator seems to break a little bit after 2010
+
+# DiD with comparison as never treated or treated after 2010
+cont_dd_2010 <- lm(phys_working ~ exposed_2010 + post_2010 + exposed_2010*post_2010 + experience + female + 
+                     avg_beds + avg_oper_days, data=filter(Physician_Data, working_allyears==1 & (minyr_EHR==0 | minyr_EHR>2009)))
+summary(cont_dd_2010)
+
+ind_dd_2010 <- lm(working_ind ~ exposed_2010 + post_2010 + exposed_2010*post_2010 + experience + female + 
[…]

### 2021-10-20 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{fullpage}
+\usepackage{import}
+\usepackage{adjustbox}
 
@@
 
-\title{Physician Response to EHR Implementation in Hospitals}
+\title{Labor Markets and Technological Change: Evidence from Electronic Health Records}
 \author{Hanna Glenn}
@@
 I link the hospitals to AHA survey data for information on EHR use and then aggregate the data to the physician level.
-To measure the working decisions of physicians, I utilize (1) physician level information, also from the shared patient data, that captures total working life (referrals to labs, pharmacies, other physicians, etc), and (2) an office-based physician level dataset to observe whether physicians are moving from hospitals to private practice.
-For my analysis, I use a difference-in-difference estimation where treatment is defined as physicians' exposure to an EHR at any of the hospitals they work in.
+This physician-level data gives information on an individuals' level of exposure to electronic health records.
+To measure the working decisions of physicians, I utilize (1) physician level information, also from the shared patient data, that captures total working life (referrals to labs, pharmacies, other physicians, etc), and (2) an office-based physician level data set to observe whether physicians are moving from hospitals to private practice.
+For my analysis, I use a difference-in-difference estimation where treatment is physicians' exposure to an EHR at any of the hospitals they work in.
 A key identifying assumption in this analysis is that hospitals choose to implement EHRs not based on any factors that are also correlated with physician working decisions.
 
+Paragraph about findings
+
+This paper contributes to two main portions of literature: the effects of health information technology and the factors of physician labor market decisions.
+The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
+Despite many case studies and hospital-level analysis that indicate large positive effects on health (\cite{Buntin2011TheResults}), economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements on average (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
+Further, there have been no significant cost decreases due to EHRs in the short or long run (CITE COST STUDIES).
+However, the long run is not fully realized, so there may still be potential benefits realized in the future.
+
+The relationship between this technology and physicians, which are the primary users of the technology, is ...
+
+\section{Background}
+
+\section{Data}
+
+\section{Empirical Analysis}
+
+\section{Results}
+
 \newpage
-\printbibliography
+
+\import{Objects/}{continuous.tex}
+
+\newpage
+
+\import{Objects/}{indicator.tex}
+
+\newpage
+
+\import{Objects/}{continuous_old.tex}
+
+\newpage
+
+\import{Objects/}{indicator_old.tex}
+
+\newpage
+
 \end{document}

### 2021-10-20 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 
-\title{Hospital Information Technology and Physician Response}
+\title{Labor Markets and Technological Change: Evidence from Electronic Health Records}
 
@@
 \begin{frame}{What is an EHR?}
-
 \begin{itemize}
     \item Computer-program technology that hospitals purchase
-    \item Performance of EHR depends on level purchased:
+    \vspace{3mm}
+    \item What EHR is used for depends on level purchased:
+    \vspace{2mm}
     \begin{itemize}
         \item All include electronic patient records
+        \vspace{2mm}
         \item More complex systems allow for decision making assistance
+        \vspace{2mm}
         \item More recently, patient interfaces are being developed
@@
 \end{itemize}
-
 \end{frame}
 
-\begin{frame}[fragile]{HIT had Great (Expected) Potential in Healthcare}
+\begin{frame}[fragile]{HIT: Great (Expected) Potential in Healthcare}
 \begin{alertblock}{Cost Saving}
 \begin{itemize}
-    \item A 2005 estimate finds possible cost reduction of hundreds of billions of dollars (\cite{hillestad2005})
+    \item Possible cost reduction of hundreds of billions of dollars \\ (\cite{hillestad2005})
 \end{itemize}
@@
     \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
-    \item Significant policy push for EHR implementation with these goal in mind: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
+    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
 \end{itemize}
@@
 \begin{frame}[fragile]{What does this mean for physicians?}
+In the short run:
 \begin{itemize}
-    \item Initially, this means physicians spend more time entering information into a computer and learning a new software
-    \item Loss of autonomy as EHRs have progressed
-    \item Physicians have experienced significant burn-out from EHRs
+    \item Physicians spend more time entering information into a computer and learning a new software
+    \item Physicians have experienced significant burn-out from using and learning EHRs
+\end{itemize}
+
+In the long run:
+\begin{itemize}
+    \item Could make daily tasks easier or harder
+    \item Continued loss of autonomy as technology progresses
 \end{itemize}
@@
 
-\begin{frame}[noframenumbering]{Physician Burnout}
+\begin{frame}[noframenumbering]{Physician Burnout in the Media}
 \begin{center}
@@
 
-Does EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work?
+Did EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work?
 

### 2021-10-21 Scripts/Final_Pairs.R
Fix exposed variable and graphs that use it

@@ -2,6 +2,7 @@ library(tidyverse)
 library(readr)
 library(plm)
 
+
 # --------------------    Build Variables Onto Main Hospital-Physician Pair Data --------------------------
 #                         Hanna Glenn, Emory University
 #                         9/3/2021
@@ -337,7 +338,7 @@ Final_Pairs <- Final_Pairs %>% ungroup() %>%
 
 # Create "exposed" indicator based on year and min year of EHR
 Final_Pairs <- Final_Pairs %>%
-  mutate(exposed=ifelse(year>=minyr_EHR,1,0))
+  mutate(exposed=ifelse(minyr_EHR>0 & year>=minyr_EHR,1,0))
 
 # Potential Treatment Variables: Documentation and Decision Making Indicators
 Final_Pairs <- Final_Pairs %>%

### 2021-10-21 Scripts/TYP_Analysis.R

@@ -5,6 +5,7 @@ library(lfe)
 library(dotwhisker)
 library(stargazer)
 library(estimatr)
+library(stats)
 
 # --------------------    EHR-Physician Analysis  --------------------------
 #                         Hanna Glenn, Emory University
@@ -20,7 +21,7 @@ Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 # Continuous Number of Patients Variable
 event_reg <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
                   rel_p4 + rel_p5 + rel_p6 + female + experience + avg_beds + avg_oper_days |year,
-                  data=Physician_Data)
+                  data=filter(Physician_Data,working_allyears==1))
 
 event_reg_coef <- as_tibble(event_reg$coefficients, rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
@@ -109,7 +110,7 @@ ggsave("Objects/ind_ES_fullsample.pdf")
 # Event Study for Continuous Working Variable (Old Doctors)
 event_reg_old <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
                     rel_p4 + rel_p5 + rel_p6 + female + avg_beds + avg_oper_days | year,
-                  data=filter(Physician_Data, experience>=40))
+                  data=filter(Physician_Data, working_allyears==1 & experience>=37))
 
 event_reg_coef_old <- as_tibble(event_reg_old$coefficients, rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
@@ -154,7 +155,7 @@ ggsave("Objects/cont_ES_oldsample.pdf")
 # Event Study for Indicator Working Variable (Old Doctors)
 event_reg_ind_old <- felm(working_ind ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
                         rel_p4 + rel_p5 + rel_p6 + female + avg_beds + avg_oper_days | year,
[…]

### 2021-10-21 Scripts/TYP_sumstats.R

@@ -20,7 +20,7 @@ cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"
 # Read in Final_Pairs_Variables.rds
 Physician_Data <- read_rds(paste0(created_data_path,"Physician_data.rds"))
 
-# General Summary Stats Tables: Separate for Physician, Hospital, Pair level ----------------------------------------------------------------
+# General Summary Stats Tables:  ----------------------------------------------------------------
 
 # Physician Level
 sum_stats <- Physician_Data %>% ungroup() %>%
@@ -39,7 +39,7 @@ sum_stats <- Physician_Data %>% ungroup() %>%
   relocate(variable,n,m,sd,min,max)
 
 knitr::kable(sum_stats, "latex",
-             col.names=c("Variable", "Physicians", "Mean", "Std. Dev.", "Min", "Max"),
+             col.names=c("Variable", "Physician-Year", "Mean", "Std. Dev.", "Min", "Max"),
              digits=2,
              caption="Physician Level Variables",
              booktabs=T,

### 2021-10-21 Presentations/ThirdYearPaperpres.tex

@@
 \usepackage{bbm}
+\usepackage{import}
+\usepackage{graphics}
 
@@
     \item Physicians have experienced significant burn-out from using and learning EHRs
+    \item Loss of autonomy
 \end{itemize}
@@
     \item Could make daily tasks easier or harder
-    \item Continued loss of autonomy as technology progresses
+    \item Progressing loss of autonomy as technology progresses
 \end{itemize}
@@
 
-\begin{frame}{Branches of Literature}
-\centering
-\includegraphics[scale=.5]{graphics/Contribution_litgraphic.PNG}
+\begin{frame}{Effect of Health Information Technology}
 
@@
 
-\begin{frame}{Physician Labor Literature}
-\centering
-    \includegraphics[scale=.45]{graphics/labor_litgraphic.PNG}
-\end{frame}
+\begin{frame}{Physician Labor Decisions}
 
-\begin{frame}{EHR Literature}
-    \centering
-    \includegraphics[scale=.45]{graphics/EHR_litgraphic.PNG}
 \end{frame}
@@
     \begin{enumerate}
-        \item EHR makes seeing patients easier (or more difficult), thus the number of patients seen increases (or decreases)
-        \item EHR imposes a burden with a high cost, so the physician switches to working at an office-based location where they have more control
-        \item A senior physician could be tipped to retire early instead of taking on the cost of learning a new technology
+        \item EHR makes seeing patients easier (or more difficult), thus physician can see more (or less) patients
+        \item If the cost of sing an EHR is high enough, the physician can switch to working at an office-based location where they have more control
+        \item A senior physician may have incentive to retire early instead of taking on the cost of learning a new technology
     \end{enumerate}
@@
 \begin{itemize}
-    \item With information on shared patients between physicians and hospitals, I limit to physicians working in hospitals
-    \item Fraction of hospitals worked with that use an EHR comes from the AHA survey
+    \item With information on shared patients between physicians and hospitals, I limit to physicians who work in at least one hospital at some point in the time frame
+    \item Hospital EHR use comes from AHA survey
     \item Similarly, I can capture hospitals that use EHRs for decision making vs. documentation using the AHA Survey IT Supplement 
@@
 
+\begin{frame}{Summary Stats}
+\centering
+    \includegraphics[scale=.7]{Objects/sumstats.pdf}
+\end{frame}
+
+\begin{frame}{Physician EHR Use}
+    \centering
+    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
+\end{frame}
+
 \section{Analysis}
 
-\begin{frame}{Event Studies}
-\begin{equation*}
-    Y_{it}= \sum_{t=-6}^6 \mathbbm{1}_{\{\text{relative treatment year}_{i}=t\}} + \alpha_t + \delta_{i} + \mathbb{X}_{it}
-\end{equation*}
+\begin{frame}{Analysis}
 
@@
 
+\section{Results}
+
+\begin{frame}[shrink=30]{Total Patients}
+\centering
+\import{Objects/}{continuous.tex}
+\end{frame}
+
+\begin{frame}[shrink=30]{Total Patients}
+\centering
+\import{Objects/}{continuous_old.tex}
+\end{frame}
+
+\begin{frame}[shrink=30]{Working}
+\centering
+\import{Objects/}{indicator.tex}
+\end{frame}
+
+\begin{frame}[shrink=30]{Retirement?}
+\centering
+\import{Objects/}{indicator_old.tex}
+\end{frame}
+
 \end{document}

### 2021-10-25 Scripts/TYP_Analysis.R
add event studies for the full sample

@@ -20,14 +20,14 @@ Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 
 # Event Studies for Entire Sample of Doctors ----------------------------------------------------------------
 # Continuous Number of Patients Variable
-event_reg <- felm(phys_working ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
-                  rel_p4 + rel_p5 + rel_p6 + female + experience + avg_beds + avg_oper_days |year,
-                  data=filter(Physician_Data,working_allyears==1))
+event_reg <- felm(phys_working_hosp ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
+                  rel_p4 + rel_p5 + rel_p6 + avg_beds + avg_oper_days |DocNPI + year,
+                  data=filter(Physician_Data,working_allyears_hosp==1))
 
 event_reg_coef <- as_tibble(event_reg$coefficients, rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
                      "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
-  dplyr::rename(estimate=phys_working)
+  dplyr::rename(estimate=phys_working_hosp)
 
 event_reg_ci <- as_tibble(confint.default(event_reg), rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
@@ -198,14 +198,14 @@ ggsave("Objects/ind_ES_oldsample.pdf")
 
 
 # Event Studies for Sample of Doctors excluding 2009 treatment ------------------------------------------------------
-event_reg_subset <- felm(phys_working ~ rel_m5 + rel_m4 + rel_m3 + rel_m2 + rel_0 + rel_p1 + rel_p2 + rel_p3 + 
-                    rel_p4 + rel_p5 + female + experience + avg_beds + avg_oper_days |year,
-                  data=filter(Physician_Data,working_allyears==1 & minyr_EHR>2009))
[…]

### 2021-10-25 Scripts/TYP_intensive.R

@@ -17,6 +17,98 @@ library(did)
 
 Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 
+# Event Study, Full Sample ------------------------------------------------------------
+event_fullsample <- felm(phys_working_hosp ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + 
+                                rel_0 + rel_p1 + rel_p2 + rel_p3 + 
+                                rel_p4 + rel_p5 + rel_p6 + avg_beds + avg_oper_days |DocNPI + year|0|DocNPI,
+                              data=filter(Physician_Data,working_allyears_hosp==1))
+
+event_fullsample_coef <- as_tibble(event_fullsample$coefficients, rownames="term") %>%
+  filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
+                     "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
+  dplyr::rename(estimate=phys_working_hosp)
+
+event_fullsample_ci <- as_tibble(confint.default(event_fullsample), rownames="term") %>%
+  filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
+                     "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
+  dplyr::rename(conf.low='2.5 %', conf.high='97.5 %')
+
+extra_row <- tibble(term="rel_m1", estimate=0, conf.low=0, conf.high=0, rel_year=-1)
+
+event_fullsample_table <- event_fullsample_coef %>%
+  left_join(event_fullsample_ci, by="term") %>%
+  mutate(rel_year=c(-6,-5,-4,-3,-2,0,1,2,3,4,5,6)) %>%
+  bind_rows(extra_row) %>%
+  arrange(rel_year)
+
+event_fullsample_plot <- dwplot(event_fullsample_table, vline=geom_vline(xintercept=0, linetype=2),
+                                     whisker_args=list(color="black", size=1.1),
+                                     vars_order =c("rel_p5", "rel_p4","rel_p3","rel_p2","rel_p1",
+                                                   "rel_0","rel_m1","rel_m2","rel_m3","rel_m4","rel_m5","rel_m6"),
+                                     dot_args=list(color="black")) + #xlim(-150,150) +
[…]

### 2021-10-25 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 
-Did EHR implementation in hospitals affect labor market decisions of physicians, specifically where, and how much they work?
+Did EHR implementation in hospitals affect physician labor outcomes?
+\begin{itemize}
+    \item Intensive Margin
+    \item Extensive Margin
+    \item Hospital $\rightarrow$ Office?
+\end{itemize}
 
@@
 
-\begin{frame}{Effect of Health Information Technology}
+\begin{frame}{Health Information Technology}
 
@@
 
-\begin{frame}{Physician Labor Decisions}
+\begin{frame}{Physician Labor}
+
+\end{frame}
+
+\begin{frame}{Technology and Labor Relationship}
 
@@
 
-\begin{frame}[shrink=30]{Total Patients}
+\begin{frame}{Full Sample Event Study}
 \centering
-\import{Objects/}{continuous.tex}
-\end{frame}
+\includegraphics[scale=.7]{Objects/event_hosp_fullsample.pdf}
 
-\begin{frame}[shrink=30]{Total Patients}
-\centering
-\import{Objects/}{continuous_old.tex}
-\end{frame}
-
-\begin{frame}[shrink=30]{Working}
-\centering
-\import{Objects/}{indicator.tex}
-\end{frame}
-
-\begin{frame}[shrink=30]{Retirement?}
-\centering
-\import{Objects/}{indicator_old.tex}
 \end{frame}

### 2021-10-25 Scripts/TYP_hosp_to_office.R
add event studies for probability of working, but not in hospital

@@ -18,15 +18,15 @@ library(did)
 Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 
 # Event Study, Full Sample ------------------------------------------------------------
-event_fullsample <- felm(phys_working_hosp ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + 
-                                rel_0 + rel_p1 + rel_p2 + rel_p3 + 
-                                rel_p4 + rel_p5 + rel_p6 + avg_beds + avg_oper_days |DocNPI + year|0|DocNPI,
-                              data=filter(Physician_Data,working_allyears_hosp==1))
+event_fullsample <- felm(nonhosp_ind ~ rel_m6 + rel_m5 + rel_m4 + rel_m3 + rel_m2 + 
+                           rel_0 + rel_p1 + rel_p2 + rel_p3 + 
+                           rel_p4 + rel_p5 + rel_p6 + avg_beds + avg_oper_days |DocNPI + year|0|DocNPI,
+                         data=Physician_Data)
 
 event_fullsample_coef <- as_tibble(event_fullsample$coefficients, rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
                      "rel_p2", "rel_p3", "rel_p4", "rel_p5", "rel_p6")) %>%
-  dplyr::rename(estimate=phys_working_hosp)
+  dplyr::rename(estimate=nonhosp_ind)
 
 event_fullsample_ci <- as_tibble(confint.default(event_fullsample), rownames="term") %>%
   filter(term %in% c("rel_m6", "rel_m5", "rel_m4", "rel_m3", "rel_m2", "rel_0", "rel_p1",
@@ -43,10 +43,10 @@ event_fullsample_table <- event_fullsample_coef %>%
   mutate(model="All")
 
 event_fullsample_plot <- dwplot(event_fullsample_table, vline=geom_vline(xintercept=0, linetype=2),
-                                     whisker_args=list(color="black", size=1.1),
-                                     vars_order =c("rel_p5", "rel_p4","rel_p3","rel_p2","rel_p1",
-                                                   "rel_0","rel_m1","rel_m2","rel_m3","rel_m4","rel_m5","rel_m6"),
-                                     dot_args=list(color="black")) + #xlim(-150,150) +
+                                whisker_args=list(color="black", size=1.1),
[…]

### 2021-10-26 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 \begin{itemize}
-    \item Computer-program technology that hospitals purchase
+    \item Technology purchased and implemented by hospitals
+\end{itemize}
+
+\begin{itemize}
+    \item Used for:
     \vspace{3mm}
-    \item What EHR is used for depends on level purchased:
-    \vspace{2mm}
     \begin{itemize}
-        \item All include electronic patient records
-        \vspace{2mm}
-        \item More complex systems allow for decision making assistance
-        \vspace{2mm}
-        \item More recently, patient interfaces are being developed
+        \item Electronic patient records
+        \vspace{3mm}
+        \item More complex (and expensive) EHRs allow for decision making assistance
+        \vspace{3mm}
+        \item Continuing to evolve over time
     \end{itemize}
@@
 \begin{frame}[fragile]{What does this mean for physicians?}
-In the short run:
+Immediate Effects:
 \begin{itemize}
-    \item Physicians spend more time entering information into a computer and learning a new software
-    \item Physicians have experienced significant burn-out from using and learning EHRs
-    \item Loss of autonomy
+    \item Large startup costs (time, productivity)
+    \item Possible loss of autonomy
 \end{itemize}
 
-In the long run:
+Long Run Effects:
 \begin{itemize}
-    \item Could make daily tasks easier or harder
+    \item Productivity could increase or decrease
     \item Progressing loss of autonomy as technology progresses
@@
 \begin{itemize}
-    \item Intensive Margin
+    \item Intensive Margin / Productivity
     \item Extensive Margin
@@
 \end{itemize}
-
 \end{frame}
@@
 
-\begin{frame}{Health Information Technology}
-
-\end{frame}
-
-\begin{frame}{Physician Labor}
-
-\end{frame}
-
-\begin{frame}{Technology and Labor Relationship}
+\begin{frame}{Literature}
 
@@
 
-\begin{frame}{Physician's Incentives}
+\begin{frame}{What I Want to Capture}
 \begin{itemize}
-    \item Physician has a choice set of where to work, including multiple hospitals and office-based locations
-    \item All physicians I consider work closely with at least one hospital at some point in time
-    \item Some proportion of hospitals implement a new technology (an EHR)
-    \item How to think about the effect on labor:
-    \begin{enumerate}
-        \item EHR makes seeing patients easier (or more difficult), thus physician can see more (or less) patients
-        \item If the cost of sing an EHR is high enough, the physician can switch to working at an office-based location where they have more control
-        \item A senior physician may have incentive to retire early instead of taking on the cost of learning a new technology
-    \end{enumerate}
+    \item Initial Implementation
+    \begin{itemize}
+        \item Productivity gains/losses
+        \item Immediate changes to labor market decisions: retirement or place of work
+    \end{itemize}
+    \item Loss of autonomy and cost of learning/using new technology
+    \begin{itemize}
+        \item Could cause physicians to move to settings where they have more control
+    \end{itemize}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{About the Data}
+    Main measurements come from \underline{CMS Shared Patient Data} (2009-2015)
+    \begin{itemize}
+        \item Lists 2 NPIs and the number of billings those two entities shared in the same day
+    \end{itemize}
+
+    How I use it:
+    \begin{itemize}
+        \item Limit the NPIs to physician-hospital pairs, where physicians are PCPs, internists, hospitalists
+        \item Limit the billings to those which occur in the same day
+        \item Only keep pairs with sufficient number of billings together to suggest a close relationship between physician and hospital
+        \item Aggregate to the physician level to avoid spillover effects
     \end{itemize}
@@
 
-     Total Patients: \underline{CMS Shared Patient Data} (2009-2015)
+     Total hospital patients: 
     \begin{itemize}
-        \item Sum of shared patients with all entities that have an NPI
+        \item Sum of shared patients with all hospitals
     \end{itemize}
 
-     Indicator for whether they are working
+     \vspace{4mm}
+
+     Indicator for switching work setting:
     \begin{itemize}
-        \item =1 if Total Patients $>0$
+        \item Positive billings shared with all entities but shared entities with hospitals becomes 0
+        \item \color{gray}{SK$\&$A database of office-based physicians}
     \end{itemize}
 
-     Indicator for working in office-based location
-    \begin{itemize}
-        \item Comes from SK$\&$A database of office-based physicians
-    \end{itemize}
+    \vspace{4mm}
+
+    \textcolor{gray}
+     {Indicator for retirement/ leaving the labor force}
 
@@
 
-\begin{frame}{Measures of EHR Use}
+\begin{frame}{Measure of EHR Use}
 
-Physician-level Indicator that captures exposure to \textit{any} EHR
+Physician-level treatment variable that captures exposure to \textit{any} EHR
 \begin{itemize}
-    \item With information on shared patients between physicians and hospitals, I limit to physicians who work in at least one hospital at some point in the time frame
-    \item Hospital EHR use comes from AHA survey
-    \item Similarly, I can capture hospitals that use EHRs for decision making vs. documentation using the AHA Survey IT Supplement 
+    \item Individual hospital EHR use comes from AHA survey
+    \item Treatment turns on the first time any of the shared hospitals implements an EHR
 \end{itemize}
-
 \end{frame}
@@
 
-\begin{frame}{Summary Stats}
+\begin{frame}{Summary Statistics}
 \centering
-    \includegraphics[scale=.7]{Objects/sumstats.pdf}
+    \includegraphics[scale=.8]{Objects/sumstats.pdf}
 \end{frame}
@@
 
-\begin{frame}{Analysis}
+\begin{frame}{Event Studies}
+\begin{equation*}
+    Y_{it}=\alpha_t+\delta_i
+\end{equation*}
 
@@
 
-\begin{frame}{Full Sample Event Study}
+\begin{frame}{Intensive Margin}
 \centering
-\includegraphics[scale=.7]{Objects/event_hosp_fullsample.pdf}
+\includegraphics[scale=.45]{Objects/event_plot.pdf}
+\end{frame}
 
+\begin{frame}{Intensive Margin}
+\centering
+\includegraphics[scale=.45]{Objects/event_hosp_plot.pdf}
+\end{frame}
+
+\begin{frame}{Hospital to Office?}
+    \centering
+    \includegraphics[scale=.5]{Objects/event_plot_office_to_hosp.pdf}
+\end{frame}
+
+\begin{frame}{Hospital to Office?}
+    \centering
+    \includegraphics[scale=.5]{Objects/event_hosp_plot_hosp_to_office.pdf}
 \end{frame}

### 2021-10-26 Scripts/Final_Pairs.R
update scripts and analysis for SKA data additions

@@ -212,6 +212,50 @@ Final_Pairs <- Final_Pairs %>%
 
 
 # Bring in SK&A -----------------------------------------------------------------------------------------
+ska2009 <- read_csv(paste0(raw_data_path,"SK&A/SKA2009.csv"))
+ska2009 <- ska2009 %>%
+  mutate(year=2009) %>%
+  select(npi,year)
+
+ska2010 <- read_csv(paste0(raw_data_path,"SK&A/SKA2010.csv"))
+ska2010 <- ska2010 %>%
+  mutate(year=2010) %>%
+  select(npi,year)
+
+ska2011 <- read_csv(paste0(raw_data_path,"SK&A/SKA2011.csv"))
+ska2011 <- ska2011 %>%
+  mutate(year=2011) %>%
+  select(npi,year)
+
+ska2012 <- read_csv(paste0(raw_data_path,"SK&A/SKA2012.csv"))
+ska2012 <- ska2012 %>%
+  mutate(year=2012) %>%
+  select(npi,year)
+
+ska2013 <- read_csv(paste0(raw_data_path,"SK&A/SKA2013.csv"))
+ska2013 <- ska2013 %>%
+  mutate(year=2013) %>%
+  select(npi,year)
+
+ska2014 <- read_csv(paste0(raw_data_path,"SK&A/SKA2014.csv"))
+ska2014 <- ska2014 %>%
+  mutate(year=2014) %>%
+  select(npi,year)
+
+ska2015 <- read_csv(paste0(raw_data_path,"SK&A/SKA2015.csv"))
+ska2015 <- ska2015 %>%
+  mutate(year=2015) %>%
+  select(npi,year)
+
+ska <- rbind(ska2009, ska2010, ska2011, ska2012, ska2013, ska2014, ska2015)
+
+ska <- ska %>%
+  mutate(ska=1, npi=as.character(npi))
+
+# Merge Final_Pairs and SKA
+Final_Pairs <- Final_Pairs %>%
+  left_join(ska, by=c("DocNPI"="npi","year")) %>%
+  mutate(ska=if_else(is.na(ska),0,ska))
 
   
 
@@ -364,12 +408,31 @@ Final_Pairs <- Final_Pairs %>%
   ungroup() %>%
   mutate(frac_EHR_patients=ifelse(num_patients>0,num_patients_EHR/num_patients,0))
 
+# Treatment Variable: start office work
[…]

### 2021-10-26 Scripts/TYP_intensive.R

@@ -42,6 +42,15 @@ event_fullsample_table <- event_fullsample_coef %>%
   arrange(rel_year) %>%
   mutate(model="All")
 
+# Calculate the sample mean of y one period before adoption
+mean_fullsample <- Physician_Data %>%
+  filter(rel_m1==1) %>%
+  ungroup() %>%
+  mutate(avg=mean(phys_working_hosp)) %>%
+  distinct(avg)
+
+avg_fullsample <- formatC(mean_fullsample$avg, digits = 0, format = "f")
+
 event_fullsample_plot <- dwplot(event_fullsample_table, vline=geom_vline(xintercept=0, linetype=2),
                                      whisker_args=list(color="black", size=1.1),
                                      vars_order =c("rel_p5", "rel_p4","rel_p3","rel_p2","rel_p1",
@@ -59,7 +68,9 @@ event_fullsample_plot <- dwplot(event_fullsample_table, vline=geom_vline(xinterc
                             "rel_m3"="t-3",
                             "rel_m4"="t-4",
                             "rel_m5"="t-5",
-                            "rel_m6"="t-6")) + ggtitle("\nEffect of EHR Implementation on Patients Billed in Hospital\n")
+                            "rel_m6"="t-6")) +
+  #scale_x_discrete(labels=c("0"=paste0("0 (",avg_fullsample," , ",avg_oldsample,")"))) +
+  ggtitle("\nEffect of EHR Implementation on Patients Billed in Hospital\n")
 
 event_fullsample_plot
 ggsave("Objects/event_fullsample.pdf")
@@ -89,6 +100,15 @@ event_oldsample_table <- event_oldsample_coef %>%
   arrange(rel_year) %>%
   mutate(model="Experience>35")
 
+# Calculate the sample mean of y one period before adoption
+mean_oldsample <- Physician_Data %>%
+  filter(rel_m1==1 & experience>35) %>%
+  ungroup() %>%
+  mutate(avg=mean(phys_working_hosp)) %>%
+  distinct(avg)
+
+avg_oldsample <- formatC(mean_oldsample$avg, digits = 0, format = "f")
+
 event_oldsample_plot <- dwplot(event_oldsample_table, vline=geom_vline(xintercept=0, linetype=2),
                                 whisker_args=list(color="black", size=1.1),
                                 vars_order =c("rel_p5", "rel_p4","rel_p3","rel_p2","rel_p1",
[…]

### 2021-10-26 Scripts/TYP_sumstats.R

@@ -67,9 +67,10 @@ knitr::kable(sum_stats, "latex",
              label=NA,
              align=c("l", "c","c","c","c","c","c","c","c","c","c"),
              position="h",
+             linesep = "\\addlinespace",
              format.args = list(big.mark = ",")) %>%
   kable_styling( full_width=F, latex_options=c("scale_down") ) %>%
-  add_header_above(c(" "=1, "Full Sample"=5, "Hospitalists" =5)) %>%
+  add_header_above(c(" "=1, "Full Sample"=5, "Works with Only 1 Hospital" =5)) %>%
   save_kable("objects/sumstats.pdf", density=300)
 
 
@@ -85,14 +86,14 @@ sum_stats_year <- Physician_Data %>% group_by(year) %>%
 
 # Create a dataframe out of the summary stats to put in a ggplot
 sum_stats_year <- as.data.frame(sum_stats_year)
-sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Fraction of Physicians Working","Fraction of Hospitals using EHR",
-                                                                                    "Fraction of Physicians Exposed to an EHR"))
+sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Fraction of Hospitals \nusing EHR",
+                                                                          "Fraction of Physicians \nExposed to an EHR"))
 sum_stats_year <- sum_stats_year %>%
   dplyr::rename("Variable"="variable")
 
 ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) + 
   geom_line() +geom_point() + labs(x="\nYear", y="Percent\n", 
-                                   title="\nPhysician Variables by Year\n") + 
+                                   title="\nEHR use Over Time\n") + 
   scale_colour_manual(values=cbbPalette) + ylim(.2,1)  + theme(legend.key.size=unit(.3,'cm'),legend.key.height = unit(.4, 'cm'),legend.key.width = unit(.3, 'cm'))
 
 ggsave("objects/sum_stats_year.pdf", width=8, height=5, units="in")

### 2021-10-26 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 \begin{frame}{Event Studies}
-\begin{equation*}
-    Y_{it}=\alpha_t+\delta_i
-\end{equation*}
 
@@
 
-\begin{frame}{Intensive Margin}
+\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
             \centering
-\includegraphics[scale=.45]{Objects/event_plot.pdf}
+            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample.pdf}
+            \caption{\small All Physicians\\}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample.pdf}
+            \caption{\small Phys. $>$ 35 yrs Experience}
+        \end{minipage}
+    \end{figure}
 \end{frame}
 
-\begin{frame}{Intensive Margin}
+\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
             \centering
-\includegraphics[scale=.45]{Objects/event_hosp_plot.pdf}
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample.pdf}
+            \caption{\small All Physicians \\(Only 1 Hospital)}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample.pdf}
+            \caption{\small Phys. $>$ 35yrs\\ (Only 1 Hospital)}
+        \end{minipage}
+    \end{figure}
 \end{frame}
 
-\begin{frame}{Hospital to Office?}
+\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
             \centering
-    \includegraphics[scale=.5]{Objects/event_plot_office_to_hosp.pdf}
+            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_ind.pdf}
+            \caption{\small All Physicians}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_ind.pdf}
+            \caption{\small Phys. $>$ 35yrs Experience}
+        \end{minipage}
+    \end{figure}
 \end{frame}
 
-\begin{frame}{Hospital to Office?}
+\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
             \centering
-    \includegraphics[scale=.5]{Objects/event_hosp_plot_hosp_to_office.pdf}
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_ind.pdf}
+            \caption{\small All Physicians \\(Only 1 Hospital)}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_ind.pdf}
+            \caption{\small Phys. $>$ 35yrs Experience\\ (Only 1 Hospital)}
+        \end{minipage}
+    \end{figure}
 \end{frame}

### 2021-10-27 Scripts/TYP_sumstats.R
fix n value in summary stats

@@ -80,20 +80,20 @@ sum_stats_year <- Physician_Data %>% group_by(year) %>%
   summarise_at(c("Fraction of Physicians Working"="working_ind", "Fraction of Hospitals using EHR"="frac_EHR", 
                  "Fraction of Physicians Exposed to an EHR"="exposed"), list(m=mean), na.rm=T) %>%
   dplyr::rename("Fraction of Physicians Working"="Fraction of Physicians Working_m", 
-                "Fraction of Hospitals using EHR"="Fraction of Hospitals using EHR_m",
-                "Fraction of Physicians Exposed to an EHR"="Fraction of Physicians Exposed to an EHR_m")
+                "Hospitals using EHR"="Fraction of Hospitals using EHR_m",
+                "Physicians Exposed to an EHR"="Fraction of Physicians Exposed to an EHR_m")
 
 
 # Create a dataframe out of the summary stats to put in a ggplot
 sum_stats_year <- as.data.frame(sum_stats_year)
-sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Fraction of Hospitals \nusing EHR",
-                                                                          "Fraction of Physicians \nExposed to an EHR"))
+sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Hospitals using EHR",
+                                                                          "Physicians Exposed to an EHR"))
 sum_stats_year <- sum_stats_year %>%
   dplyr::rename("Variable"="variable")
 
 ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) + 
   geom_line() +geom_point() + labs(x="\nYear", y="Percent\n", 
-                                   title="\nEHR use Over Time\n") + 
+                                   title="\nEHR Use Over Time\n") + 
   scale_colour_manual(values=cbbPalette) + ylim(.2,1)  + theme(legend.key.size=unit(.3,'cm'),legend.key.height = unit(.4, 'cm'),legend.key.width = unit(.3, 'cm'))
 
 ggsave("objects/sum_stats_year.pdf", width=8, height=5, units="in")

### 2021-10-27 Presentations/ThirdYearPaperpres.tex

@@
 
-\begin{frame}{Event Studies}
+\begin{frame}{Event Study}
+\begin{equation*}
+    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-5}^5 \beta_kz_{i,t-k} + \varepsilon_{it}
+\end{equation*}
 
@@
 
+\begin{frame}{Dep. Variable: Appears in Office-Based Data}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_office.pdf}
+            \caption{\small All Physicians}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_office.pdf}
+            \caption{\small Phys.$>$35yrs Experience}
+        \end{minipage}
+    \end{figure}
+\end{frame}
+
+\begin{frame}{Dep. Variable: Appears in Office-Based Data}
+\begin{figure}[ht]
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_office.pdf}
+            \caption{\small All Physicians \\(Only 1 Hospital)}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_office.pdf}
+            \caption{\small Phys.$>$35yrs Experience\\ (Only 1 Hospital)}
+        \end{minipage}
+    \end{figure}
+\end{frame}
+
 \end{document}

### 2021-10-27 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 \usepackage{appendixnumberbeamer}
-\usepackage{multicol}
+\usepackage{pgfpages}
 \usepackage{booktabs}
-\usepackage[scale=2]{ccicons}
-\usepackage[style=authoryear]{biblatex}
-\bibliography{demo.bib}
+\usepackage{xcolor}
+
 \usepackage{pgfplots}
-\usepackage{bbm}
-\usepackage{import}
-\usepackage{graphics}
 
@@
 
+\setbeameroption{show notes on second screen=right}
+\setbeamertemplate{note page}{\pagecolor{yellow!5}\insertnote}\usepackage{palatino}
 \usepackage{xspace}
@@
  \date{\today}
-% \date{}
-%\author{Presented by: Hanna Glenn}
-% \institute{Center for modern beamer themes}
-% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}
+
+\setbeamertemplate{itemize item}{\scalebox{.6}{$\blacktriangleright$ }}     
+\setbeamertemplate{itemize subitem}{\scalebox{.8}{$\centerdot$ }}
 
@@
   \tableofcontents%[hideallsubsections]
+
+  \note[item]{Roadmap of the presentation}
 \end{frame}
@@
 
-\begin{frame}{What is an EHR?}
+\begin{frame}[fragile]{What is an EHR?}
 \begin{itemize}
@@
 \end{itemize}
+
+\note[item]{Get everyone on the same page about how to think about EHRs}
+\note[item]{A product in the form of technology that is sold to hospitals (and others, but we focus on hospital setting)}
+\note[item]{The most standard element of an EHR is that it converts paper records to electronic records}
+\note[item]{New elements of EHRs: interropability, patient interface, decision-making assistance}
+
 \end{frame}
@@
 \begin{itemize}
-    \item Possible cost reduction of hundreds of billions of dollars \\ (\cite{hillestad2005})
+    \item Possible cost reduction of hundreds of billions of dollars \\ \scriptsize (Hillestad et al 2005, HA)
 \end{itemize}
@@
     \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
-    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs \nocite{hitech}
+    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs 
 \end{itemize}
@@
 
-\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} (\cite{stats})
+\vspace{4mm}
+
+\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
+
+\note[item]{Why do we care about EHRs or health information technology in the first place?}
 
@@
 
-\begin{frame}[fragile]{What does this mean for physicians?}
+\begin{frame}[fragile]{What could this mean for physicians?}
+Physicians are the ones who put this technology to use, how they respond will partially determine the technology's impacts
+
+\vspace{3mm}
+
 Immediate Effects:
@@
 
+\vspace{4mm}
+
 Long Run Effects:
 \begin{itemize}
-    \item Productivity could increase or decrease
+    \item Productivity could increase
     \item Progressing loss of autonomy as technology progresses
@@
 
-\begin{frame}[noframenumbering]{Physician Burnout in the Media}
+\begin{frame}[noframenumbering]{Physician Response to EHRs}
 \begin{center}
@@
 
-\begin{frame}[noframenumbering]{Physician Burnout in the Media}
+\begin{frame}[noframenumbering]{Physician Response to EHRs}
 \begin{center}
-    \includegraphics[scale=.45]{graphics/News Clip3.PNG}
+    \includegraphics[scale=.35]{graphics/News Clip3.PNG}
 \end{center}
@@
 \begin{frame}{Literature}
+\footnotesize 
+\begin{columns}
+    \column{0.34\textwidth}
+        \centering
+        \underline{ EHR $\rightarrow$ Healthcare }
+        \begin{itemize}
+            \item Small improvements for severe health conditions \\ \vspace{1mm}
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
 
+            \footnotesize
+
+            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
+            \tiny (Appari et al 2013)
+
+            \footnotesize
+
+            \item No decrease in cost \\ \vspace{1mm}
+            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
+
+            \footnotesize
+
+            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
+            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
+
+        \end{itemize}
+
+    \column{0.33\textwidth}
+    \color{gray}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \begin{itemize}
+        \color{gray}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011)
+
+            \footnotesize
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2020)
+        \end{itemize}
+
+    \column{0.33\textwidth}
+        \color{gray}
+        \centering
+        Physician Labor \underline{ Determinants }
+        \begin{itemize}
+        \color{gray}
+            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
+            \tiny (Baltagi et al 2005, Sloan 2018)
+
+            \footnotesize
+
+            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002)
+
+            \footnotesize
+
+            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
+            \tiny (Dewa et al 2014)
+        \end{itemize}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Literature}
+\footnotesize 
+\begin{columns}
+    \column{0.34\textwidth}
+    \color{gray}
+        \centering
+        \underline{ EHR $\rightarrow$ Healthcare }
+        \begin{itemize}
+        \color{gray}
+            \item Small improvements for severe health conditions \\ \vspace{1mm}
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
+
+            \footnotesize
+
+            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
+            \tiny (Appari et al 2013)
+
+            \footnotesize
+
+            \item No decrease in cost \\ \vspace{1mm}
+            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
+
+            \footnotesize
+
+            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
+            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
+
+        \end{itemize}
+
+    \column{0.33\textwidth}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \begin{itemize}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011)
+
+            \footnotesize
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2020)
+        \end{itemize}
+
+    \column{0.33\textwidth}
+    \color{gray}
+        \centering
+        Physician Labor \underline{ Determinants }
+        \begin{itemize}
+        \color{gray}
+            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
+            \tiny (Baltagi et al 2005, Sloan 2018)
+
+            \footnotesize
+
+            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002)
+
+            \footnotesize
+
+            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
+            \tiny (Dewa et al 2014)
+        \end{itemize}
+\end{columns}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Literature}
+\footnotesize 
+\begin{columns}
+    \column{0.34\textwidth}
+    \color{gray}
+        \centering
+        \underline{ EHR $\rightarrow$ Healthcare }
+        \begin{itemize}
+        \color{gray}
+            \item Small improvements for severe health conditions \\ \vspace{1mm}
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
+
+            \footnotesize
+
+            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
+            \tiny (Appari et al 2013)
+
+            \footnotesize
+
+            \item No decrease in cost \\ \vspace{1mm}
+            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
+
+            \footnotesize
+
+            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
+            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
+
+        \end{itemize}
+
+    \column{0.33\textwidth}
+    \color{gray}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \begin{itemize}
+        \color{gray}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011)
+
+            \footnotesize
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2020)
+        \end{itemize}
+
+    \column{0.33\textwidth}
+        \centering
+        Physician Labor \underline{ Determinants }
+        \begin{itemize}
+            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
+            \tiny (Baltagi et al 2005, Sloan 2018)
+
+            \footnotesize
+
+            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002)
+
+            \footnotesize
+
+            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
+            \tiny (Dewa et al 2014)
+        \end{itemize}
+\end{columns}
 \end{frame}
@@
     \item Initial Implementation
+    \vspace{3mm}
     \begin{itemize}
-        \item Productivity gains/losses
-        \item Immediate changes to labor market decisions: retirement or place of work
+        \item Productivity gains/losses?
+        \vspace{3mm}
+        \item Retirement or change in place of work?
+        \vspace{3mm}
     \end{itemize}
-    \item Loss of autonomy and cost of learning/using new technology
-    \begin{itemize}
-        \item Could cause physicians to move to settings where they have more control
-    \end{itemize}
+    \item Do the effects change over time after the technology is implemented?
+
 \end{itemize}
@@
     \begin{itemize}
-        \item Lists 2 NPIs and the number of billings those two entities shared in the same day
+        \item Lists 2 NPIs and the number of billings those two entities shared
     \end{itemize}
+    \vspace{3mm}
 
@@
 
-     Total hospital patients: 
+     Hospital patients: 
     \begin{itemize}
@@
         \item Positive billings shared with all entities but shared entities with hospitals becomes 0
-        \item \color{gray}{SK$\&$A database of office-based physicians}
+        \item Database of office-based physicians
     \end{itemize}
@@
 Physician-level treatment variable that captures exposure to \textit{any} EHR
+\vspace{3mm}
 \begin{itemize}
     \item Individual hospital EHR use comes from AHA survey
+    \vspace{3mm}
     \item Treatment turns on the first time any of the shared hospitals implements an EHR
@@
 
-Medical school graduation date, gender: \underline{Physician Compare}
+Medical school graduation date: \underline{Physician Compare}
 
@@
     \includegraphics[scale=.8]{Objects/sumstats.pdf}
+
+    \note[item]{Need to divide n by 7 to get actual number of physicians}
 \end{frame}
@@
 
+\section{Continuing Work}
+
+\begin{frame}{Plans}
+    \begin{itemize}
+        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicaid data
+    \end{itemize}
+\end{frame}
+
 \end{document}

### 2021-10-28 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 
-\setbeameroption{show notes on second screen=right}
-\setbeamertemplate{note page}{\pagecolor{yellow!5}\insertnote}\usepackage{palatino}
+\setbeameroption{slides only}
+%\setbeameroption{show notes on second screen=right}
 \usepackage{xspace}
@@
 
-\begin{frame}[fragile]{What is an EHR?}
+\begin{frame}{Technology and Displacement}
+
+\end{frame}
+
+\begin{frame}[fragile]{What is an Electronic Health Record?}
 \begin{itemize}
@@
         \vspace{3mm}
-        \item More complex (and expensive) EHRs allow for decision making assistance
-        \vspace{3mm}
-        \item Continuing to evolve over time
+        \item Decision making assistance
     \end{itemize}
@@
 \begin{itemize}
-    \item Possible cost reduction of hundreds of billions of dollars \\ \scriptsize (Hillestad et al 2005, HA)
+    \item Possible cost reduction of hundreds of billions of dollars \\ \scriptsize (Hillestad et al 2005)
 \end{itemize}
@@
 \note[item]{Why do we care about EHRs or health information technology in the first place?}
-
-\end{frame}
-
-\begin{frame}[fragile]{What could this mean for physicians?}
-Physicians are the ones who put this technology to use, how they respond will partially determine the technology's impacts
-
-\vspace{3mm}
-
-Immediate Effects:
-\begin{itemize}
-    \item Large startup costs (time, productivity)
-    \item Possible loss of autonomy
-\end{itemize}
-
-\vspace{4mm}
-
-Long Run Effects:
-\begin{itemize}
-    \item Productivity could increase
-    \item Progressing loss of autonomy as technology progresses
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[noframenumbering]{Physician Response to EHRs}
-\begin{center}
-    \includegraphics[scale=.4]{graphics/News Clip1.PNG}
-\end{center}
+\note[item]{Verbally explain the bridge to physicians here (the slide I took out)}
 \end{frame}
@@
 
-Did EHR implementation in hospitals affect physician labor outcomes?
+Did EHR implementation in hospitals affect physician labor market decisions?
 \begin{itemize}
@@
     \item Extensive Margin
-    \item Hospital $\rightarrow$ Office?
+    \item Hospital $\rightarrow$ Office
 \end{itemize}
@@
             \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
+            \tiny(Agha 2014; McCullough et al 2016)
 
@@
             \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
+            \tiny (Agha 2014; Dranove et al 2019)
 
@@
         \end{itemize}
+\end{columns}
 
-\end{columns}
+\note[item]{\footnotesize Agha 2014 (JHE) uses claims data and difference in difference methodology.
+Find 1.3 percent increase in billed costs and no improvements in health coutcomes, data is 1998-2005.}
+
+\note[item]{McCullough, Parente, Town 2016, RAND. claims data 2002-2007. Find no reduction in moretality for the median patient, only those who require complex care with coordination.}
+
+\note[item]{Appari, Johnson, Anthony 2013. Health services review. fixed effects linear panel model. quality measure comes from hospital compare, not at the patient level.}
+
+\note[item]{Dranove, Forman, Goldfarb, Greenstein (2019) American Economic Journal: Economic Policy.
+data from 1996 to 2009.
+Specific findings: hospitals have a rise in cost after 3 years, some hospitals in certain locations have a decrease in costs after 5 years but others still have a higher cost after 6 years.}
+
 \end{frame}
@@
 \end{columns}
+
+\note[item]{Zeira and Johnson 2011. }
 \end{frame}
@@
 
-\begin{frame}{What I Want to Capture}
-\begin{itemize}
-    \item Initial Implementation
-    \vspace{3mm}
-    \begin{itemize}
-        \item Productivity gains/losses?
-        \vspace{3mm}
-        \item Retirement or change in place of work?
-        \vspace{3mm}
-    \end{itemize}
-    \item Do the effects change over time after the technology is implemented?
-
-\end{itemize}
-
-\end{frame}
-
 \begin{frame}{About the Data}
-    Main measurements come from \underline{CMS Shared Patient Data} (2009-2015)
+    Main data: \underline{CMS Shared Patient Data} (2009-2015)
     \begin{itemize}
@@
     \begin{itemize}
-        \item Limit the NPIs to physician-hospital pairs, where physicians are PCPs, internists, hospitalists
-        \item Limit the billings to those which occur in the same day
-        \item Only keep pairs with sufficient number of billings together to suggest a close relationship between physician and hospital
-        \item Aggregate to the physician level to avoid spillover effects
+        \item Physician-hospital pairs (PCPs, internists, hospitalists)
+        \item Billings that occur on same day
+        \item Close relationship between hospital and physician
+        \item Aggregate to the physician level
     \end{itemize}
@@
 
-     Hospital patients: 
-    \begin{itemize}
-        \item Sum of shared patients with all hospitals
-    \end{itemize}
+     Patients visited in hospitals
 
@@
     \begin{itemize}
-        \item Positive billings shared with all entities but shared entities with hospitals becomes 0
-        \item Database of office-based physicians
+        \item 0 patients visited but maintaining billable activity
     \end{itemize}
@@
     \vspace{3mm}
-    \item Treatment turns on the first time any of the shared hospitals implements an EHR
+    \item Treatment turns on the first time any of the physicians' hospitals implements an EHR
 \end{itemize}
@@
 
-\vspace{5mm}
-
-Average size of hospital worked with (measured in beds), average hospital days operating: \underline{AHA Survey}
 \end{frame}
 
-\begin{frame}{Summary Statistics}
+\begin{frame}{Data}
 \centering
@@
 
-    \note[item]{Need to divide n by 7 to get actual number of physicians}
 \end{frame}
@@
     \begin{itemize}
-        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicaid data
+        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
+        \item Investigate other estimation strategies
     \end{itemize}

### 2021-11-01 Presentations/ThirdYearPaperpres.tex

@@
 \usepackage{xcolor}
+\usepackage{mathtools}
 
@@
 \setbeameroption{slides only}
+%\setbeameroption{show only notes}
 %\setbeameroption{show notes on second screen=right}
@@
 
+\setbeamertemplate{caption}{\raggedright\insertcaption\par}
 \setbeamertemplate{itemize item}{\scalebox{.6}{$\blacktriangleright$ }}     
@@
 
-\begin{frame}{Table of Contents}
-  \setbeamertemplate{section in toc}[sections numbered]
-  \tableofcontents%[hideallsubsections]
-
-  \note[item]{Roadmap of the presentation}
-\end{frame}
-
 \section[Motivation]{Motivation}
 
-\begin{frame}{Technology and Displacement}
+\begin{frame}{The Broad Question}
+\begin{itemize}
+    \item Longstanding topic in economics: how does technology affect employment?
+    \vspace{3mm}
+    \item Been investigated in many industries
+    \vspace{3mm}
+    \item Specific healthcare setting
+\end{itemize}
 
+\note{An interesting and prominent question in economics is how technology affects employment and other labor market factors.
+This question has been studied for a long time and has entire chapters of books dedicated to it.
+People are generally very interested in how their jobs and the jobs of their children will be affected by technology.
+Empirically, this has been studied across sectors and industries and countries.
+In this paper, I am thinking about a very specific type of technology: electronic health records.
+}
 \end{frame}
@@
 \begin{frame}[fragile]{What is an Electronic Health Record?}
+At the least:
 \begin{itemize}
-    \item Technology purchased and implemented by hospitals
+    \item Switching from transcribing to computerization
 \end{itemize}
 
+\vspace{3mm}
+
+More:
 \begin{itemize}
-    \item Used for:
-    \vspace{3mm}
-    \begin{itemize}
-        \item Electronic patient records
-        \vspace{3mm}
+    \item Communication
     \item Decision making assistance
 \end{itemize}
-\end{itemize}
 
-\note[item]{Get everyone on the same page about how to think about EHRs}
-\note[item]{A product in the form of technology that is sold to hospitals (and others, but we focus on hospital setting)}
-\note[item]{The most standard element of an EHR is that it converts paper records to electronic records}
-\note[item]{New elements of EHRs: interropability, patient interface, decision-making assistance}
+\note[item]{So what actually are electronic health records?
+At their most basic, they computerize patient records that used to be kept on paper.
+For example, before EHRs a doctor would look at an ultrasound and they would talk into a microphone about what they saw.
+Then the recording would go to transcribers who record what the doctor said.
+So, when basic EHRs are implemented they record things directly into a computer instead of talking into a microphone.
+More complex systems have similar transitions.
+Systems can have messaging components that have transformed communication from letter writing to sitting down and hitting a few buttons.
+Another aspect you can imagine is a physician sitting down with a patient and receiving a popup on the screen that tells them the patient should have had a colonoscopy by now.
+So these systems change a lot about how physicians practice.}
 
@@
 
-\note[item]{Why do we care about EHRs or health information technology in the first place?}
-\note[item]{Verbally explain the bridge to physicians here (the slide I took out)}
+\note[item]{\tiny Technology improvements in health to improve efficiency have been a huge deal. The reason for this is two-fold: the expected cost saving and the expected quality improvements.}
+\note[item]{Before EHRs were standard, people expected the cost saving of these things to be massive.
+One study in particular estimated cost savings of hundreds of billions of dollars if HIT was fully implemented.}
+\note[item]{In addition to the cost saving, the assumption was that new HIT would indirectly affect patient care/outcomes as well.
+Doctors could see more patients and faster, they have assistance to prevent unnecessary mistakes, etc.}
+\note[item]{These expectations are what led to the incentivizing of using EHRs "meaningfully".
+All of these factors led to hospitals switching over to EHRs: 9 percent of hospitals with EHRs grew to 84 percent in 2015.}
+
+\note[item]{Physicians are the primary users of the new technology.
+There is this stigma in healthcare about how difficult it has been for physicians to adjust to the new technologies.
+A main reason for this is the loss of autonomy and control a physician feels in their job.
+They go through all of this school just to have a computer tell them what to do?
+But we don't know if physicians have changed their behavior in terms of labor market decisions.
+Since they are the bridge to the potential benefits of HIT and the realized benefits of HIT.}
 \end{frame}
@@
 \end{center}
+
+\note[item]{I want to show you a couple examples of what is typical to see in the media about how physicians respond to EHRs.
+This headline is about how the very thing that is supposed to improve doctors' interactions with patients is the very thing causing compications.}
 \end{frame}
@@
 \end{center}
+
+\note[item]{Another headline here talks about EHRs ruining medicine for physicians.
+The subheading here says that this issue ranks 2nd in the list of issues ruining medicine for physicians.
+So do these technologies actually cause physicians to change their behavior?
+That is what I want to investigate.}
 \end{frame}
@@
 \end{itemize}
+
+\note[item]{This paper will specifically look at the effect of EHRs on physicians in three dimensions.
+First, on the intensive margin.
+Conditional on staying in the job, do EHRs lead to more or less patients being taken on?
+Second, on the extensive margin.
+Do EHRs lead physicians to actually retire early or leave their jobs?
+Third, I want to look at the work setting of the physician.
+Do EHRs in hospitals lead physicians to switch from hospital settings to office-based settings?
+Or transfer some of their workload to facilities in which they have more autonomy?}
+
 \end{frame}
@@
 \begin{frame}{Literature}
+\small
+\begin{columns}
+\setlength{\tabcolsep}{-5pt}
+    \column{0.33\textwidth}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \vspace{-1mm}
+        \begin{itemize}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011, EJ)
+
             \footnotesize
-\begin{columns}
-    \column{0.34\textwidth}
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2021, Labour)
+        \end{itemize}
+
+    \column{0.33\textwidth}
         \centering
+        \color{gray}
+        Physician Labor \underline{ Determinants }
+        \vspace{-1mm}
+        \begin{itemize}
+        \color{gray}
+            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
+            \tiny (Baltagi et al 2005, HE; Sloan 2018, ILRR)
+
+            \footnotesize
+
+            \item Personal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002, JHHSA)
+
+            \footnotesize
+
+            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
+            \tiny (Dewa et al 2014, BMC HSR)
+        \end{itemize}
+
+        \column{0.37\textwidth}
+        \centering
+        \color{gray}
         \underline{ EHR $\rightarrow$ Healthcare }
+        \vspace{-1mm}
         \begin{itemize}
+        \color{gray}
             \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014; McCullough et al 2016)
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
 
@@
             \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013)
+            \tiny (Appari et al 2013, HSR)
 
@@
             \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014; Dranove et al 2019)
+            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)
 
@@
             \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
+            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
 
@@
 
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
-        \color{gray}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
-        \end{itemize}
-
-    \column{0.33\textwidth}
-        \color{gray}
-        \centering
-        Physician Labor \underline{ Determinants }
-        \begin{itemize}
-        \color{gray}
-            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
-
-            \footnotesize
-
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
-
-            \footnotesize
-
-            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
-        \end{itemize}
 \end{columns}
@@
 \footnotesize 
+\small
 \begin{columns}
-    \column{0.34\textwidth}
+\setlength{\tabcolsep}{-5pt}
+    \column{0.33\textwidth}
+        \centering
         \color{gray}
-        \centering
-        \underline{ EHR $\rightarrow$ Healthcare }
+        \underline{ Technology $\rightarrow$ Labor }
+        \vspace{-1mm}
         \begin{itemize}
         \color{gray}
-            \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
-
-            \footnotesize
-
-            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013)
-
-            \footnotesize
-
-            \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
-
-            \footnotesize
-
-            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
-
-        \end{itemize}
-
-    \column{0.33\textwidth}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
             \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
+            \tiny (Zeira $\&$ Joseph 2011, EJ)
 
@@
             \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
+            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
 
@@
             \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
-        \end{itemize}
-
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        Physician Labor \underline{ Determinants }
-        \begin{itemize}
-        \color{gray}
-            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
-
-            \footnotesize
-
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
-
-            \footnotesize
-
-            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
-        \end{itemize}
-\end{columns}
-
-\note[item]{Zeira and Johnson 2011. }
-\end{frame}
-
-\begin{frame}[noframenumbering]{Literature}
-\footnotesize 
-\begin{columns}
-    \column{0.34\textwidth}
-    \color{gray}
-        \centering
-        \underline{ EHR $\rightarrow$ Healthcare }
-        \begin{itemize}
-        \color{gray}
-            \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
-
-            \footnotesize
-
-            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013)
-
-            \footnotesize
-
-            \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
-
-            \footnotesize
-
-            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
-
-        \end{itemize}
-
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
-        \color{gray}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
+            \tiny (Willis and Hudomiet 2021, Labour)
         \end{itemize}
@@
         Physician Labor \underline{ Determinants }
+        \vspace{-1mm}
         \begin{itemize}
             \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
+            \tiny (Baltagi et al 2005, HE; Sloan 2018, ILRR)
 
@@
 
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
+            \item Personal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002, JHHSA)
 
@@
             \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
+            \tiny (Dewa et al 2014, BMC HSR)
         \end{itemize}
+
+        \column{0.37\textwidth}
+        \centering
+        \color{gray}
+        \underline{ EHR $\rightarrow$ Healthcare }
+        \vspace{-1mm}
+        \begin{itemize}
+        \color{gray}
+            \item Small improvements for severe health conditions \\ \vspace{1mm}
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
+
+            \footnotesize
+
+            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
+            \tiny (Appari et al 2013, HSR)
+
+            \footnotesize
+
+            \item No decrease in cost \\ \vspace{1mm}
+            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)
+
+            \footnotesize
+
+            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
+            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
+
+        \end{itemize}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}[noframenumbering]{Literature}
+\small
+\begin{columns}
+\setlength{\tabcolsep}{-5pt}
+    \column{0.33\textwidth}
+    \color{gray}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \vspace{-1mm}
+        \begin{itemize}
+        \color{gray}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011, EJ)
+
+            \footnotesize
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2021, Labour)
+        \end{itemize}
+
+    \column{0.33\textwidth}
+        \centering
+        \color{gray}
+        Physician Labor \underline{ Determinants }
+        \vspace{-1mm}
+        \begin{itemize}
+        \color{gray}
+            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
+            \tiny (Baltagi et al 2005, HE; Sloan 2018, ILRR)
+
+            \footnotesize
+
+            \item Personal life factors matter more for retirement than work conditions \\ \vspace{1mm}
+            \tiny (Bahrami 2002, JHHSA)
+
+            \footnotesize
+
+            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
+            \tiny (Dewa et al 2014, BMC HSR)
+        \end{itemize}
+
+        \column{0.37\textwidth}
+        \centering
+        \underline{ EHR $\rightarrow$ Healthcare }
+        \vspace{-1mm}
+        \begin{itemize}
+            \item Small improvements for severe health conditions \\ \vspace{1mm}
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
+
+            \footnotesize
+
+            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
+            \tiny (Appari et al 2013, HSR)
+
+            \footnotesize
+
+            \item No decrease in cost \\ \vspace{1mm}
+            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)
+
+            \footnotesize
+
+            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
+            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
+
+        \end{itemize}
+
 \end{columns}
@@
 
-\begin{frame}{About the Data}
-    Main data: \underline{CMS Shared Patient Data} (2009-2015)
+\begin{frame}{Data}
+    Main data: \underline{CMS Shared Patient Data (2009-2015)}
     \begin{itemize}
-        \item Lists 2 NPIs and the number of billings those two entities shared
+        \item Hospital-physician pairs: bill for same patient
+        \item Physician works in the hospital
     \end{itemize}
+    EHR Use:
+    \begin{itemize}
+        \item Individual hospital EHR: AHA Survey
+    \end{itemize}
+    Other:
+    \begin{itemize}
+        \item Physician Compare
+        \item Hospital Characteristics
+    \end{itemize}
+
     \vspace{3mm}
 
-    How I use it:
-    \begin{itemize}
-        \item Physician-hospital pairs (PCPs, internists, hospitalists)
-        \item Billings that occur on same day
-        \item Close relationship between hospital and physician
-        \item Aggregate to the physician level
-    \end{itemize}
+    Aggregated to the physician level
 \end{frame}
 
-\begin{frame}{Measures of Physician Labor Market Decisions}
-
-     Patients visited in hospitals
-
-     \vspace{4mm}
-
-     Indicator for switching work setting:
+\begin{frame}{Variables}
+Labor Market Outcomes:
+\vspace{-2mm}
 \begin{itemize}
-        \item 0 patients visited but maintaining billable activity
+    \item Patients visited in hospitals
+    \item Maintaining billable activity only outside of hospitals
+    \item \color{gray} Level of billable activity outside of hospitals
+    \item \color{gray} Retirement/leaving the labor force
 \end{itemize}
 
-    \vspace{4mm}
+EHR Use:
+\vspace{-2mm}
+\begin{itemize}
+    \item Indicator for any exposure to EHR
+    \item \color{gray} Fraction of hospitals exposed to EHR
+\end{itemize}
 
-    \textcolor{gray}
-     {Indicator for retirement/ leaving the labor force}
+Other:
+\vspace{-2mm}
+\begin{itemize}
+    \item Years of experience
+    \item Hospital characteristics
+\end{itemize}
 
+\note[item]{So now we can talk }
 \end{frame}
 
-\begin{frame}{Measure of EHR Use}
-
-Physician-level treatment variable that captures exposure to \textit{any} EHR
-\vspace{3mm}
-\begin{itemize}
-    \item Individual hospital EHR use comes from AHA survey
-    \vspace{3mm}
-    \item Treatment turns on the first time any of the physicians' hospitals implements an EHR
-\end{itemize}
+\begin{frame}{Data}
+    \centering
+    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
 \end{frame}
 
-\begin{frame}{Other Characteristics}
-
-Medical school graduation date: \underline{Physician Compare}
+\begin{frame}{Data}
+\centering
+    \includegraphics[scale=.5]{Objects/cont_treatment_graph.pdf}
 
@@
     \centering
-    \includegraphics[scale=.8]{Objects/sumstats.pdf}
-
-\end{frame}
-
-\begin{frame}{Physician EHR Use}
-    \centering
-    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.5]{Objects/ind_treatment_graph.pdf}
 \end{frame}
@@
 \end{equation*}
+
+\vspace{4mm}
+
+\begin{align*}
+    y_{it} &: \text{Outcomes} \hspace{30mm}\\
+    \alpha_i, \delta_t &: \text{Fixed Effects}\hspace{30mm}\\
+    q_{it} &: \text{physician/hospital characteristics}\hspace{30mm}\\
+    z_{i,t-k} &: \text{exposure to EHR} \hspace{30mm}
+\end{align*}
 
@@
 
-\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample.pdf}
-            \caption{\small Phys. $>$ 35yrs\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
 \begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
@@
 
-\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_ind.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_ind.pdf}
-            \caption{\small Phys. $>$ 35yrs Experience\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Appears in Office-Based Data}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_office.pdf}
-            \caption{\small All Physicians}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_office.pdf}
-            \caption{\small Phys.$>$35yrs Experience}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Appears in Office-Based Data}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_office.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_office.pdf}
-            \caption{\small Phys.$>$35yrs Experience\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
 \section{Continuing Work}
@@
 
+\begin{frame}{}
+\centering
+    Thank you!
+\end{frame}
+
 \end{document}

### 2021-11-01 Presentations/ThirdYearPaperpres.tex

--- 2021-10-28 Presentations/ThirdYearPaperpres.tex
@@
 
-\begin{frame}{Table of Contents}
-  \setbeamertemplate{section in toc}[sections numbered]
-  \tableofcontents%[hideallsubsections]
-
-  \note[item]{Roadmap of the presentation}
-\end{frame}
-
 \section[Motivation]{Motivation}
 
-\begin{frame}{Technology and Displacement}
+\begin{frame}{The Broad Question}
+\begin{itemize}
+    \item Longstanding topic in economics: relationship between technology and employment
+    \vspace{2mm}
+    \begin{itemize}
+        \item Displacement?
+        \item Productivity?
+        \item Wages?
+    \end{itemize}
+    \vspace{3mm}
+    \item Investigated in many settings
+    \vspace{3mm}
+    \item This paper: specific healthcare setting
+\end{itemize}
 
@@
 \begin{frame}[fragile]{What is an Electronic Health Record?}
+``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"
+
+\vspace{3mm}
+
+What it can do:
 \begin{itemize}
-    \item Technology purchased and implemented by hospitals
-\end{itemize}
-
-\begin{itemize}
-    \item Used for:
-    \vspace{3mm}
-    \begin{itemize}
-        \item Electronic patient records
-        \vspace{3mm}
+    \item Paper $\rightarrow$ computerization
+    \item Communication
     \item Decision making assistance
 \end{itemize}
-\end{itemize}
-
-\note[item]{Get everyone on the same page about how to think about EHRs}
-\note[item]{A product in the form of technology that is sold to hospitals (and others, but we focus on hospital setting)}
-\note[item]{The most standard element of an EHR is that it converts paper records to electronic records}
-\note[item]{New elements of EHRs: interropability, patient interface, decision-making assistance}
 
@@
 
-\note[item]{Why do we care about EHRs or health information technology in the first place?}
-\note[item]{Verbally explain the bridge to physicians here (the slide I took out)}
 \end{frame}
 
-\begin{frame}[noframenumbering]{Physician Response to EHRs}
+\begin{frame}{Physician Response to EHRs}
 \begin{center}
-    \includegraphics[scale=.5]{graphics/News Clip2.PNG}
+    \includegraphics[scale=.3]{graphics/News Clip3.PNG}
+
+    \vspace{3mm}
+
+    \includegraphics[scale=.3]{graphics/News Clip2.PNG}
 \end{center}
-\end{frame}
 
-\begin{frame}[noframenumbering]{Physician Response to EHRs}
-\begin{center}
-    \includegraphics[scale=.35]{graphics/News Clip3.PNG}
-\end{center}
 \end{frame}
@@
 \end{itemize}
+
 \end{frame}
@@
 \begin{frame}{Literature}
+\small
+\begin{columns}
+\setlength{\tabcolsep}{-5pt}
+    \column{0.4\textwidth}
+        \centering
+        \underline{ Technology $\rightarrow$ Labor }
+        \vspace{-1mm}
+        \begin{itemize}
+            \item Erosion effect vs. wage effect \\ \vspace{1mm}
+            \tiny (Zeira $\&$ Joseph 2011, EJ)
+
             \footnotesize
-\begin{columns}
-    \column{0.34\textwidth}
+
+            \item Cause retirement? \\ \vspace{1mm}
+            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
+
+            \footnotesize
+
+            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
+            \tiny (Willis and Hudomiet 2021, Labour)
+        \end{itemize}
+
+        \column{0.4\textwidth}
         \centering
+        \color{gray}
         \underline{ EHR $\rightarrow$ Healthcare }
+        \vspace{-1mm}
         \begin{itemize}
+        \color{gray}
             \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014; McCullough et al 2016)
+            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
 
@@
 
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
-        \color{gray}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
-        \end{itemize}
-
-    \column{0.33\textwidth}
-        \color{gray}
-        \centering
-        Physician Labor \underline{ Determinants }
-        \begin{itemize}
-        \color{gray}
-            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
-
-            \footnotesize
-
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
-
-            \footnotesize
-
-            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
-        \end{itemize}
 \end{columns}
 
-\note[item]{\footnotesize Agha 2014 (JHE) uses claims data and difference in difference methodology.
-Find 1.3 percent increase in billed costs and no improvements in health coutcomes, data is 1998-2005.}
-
-\note[item]{McCullough, Parente, Town 2016, RAND. claims data 2002-2007. Find no reduction in moretality for the median patient, only those who require complex care with coordination.}
-
-\note[item]{Appari, Johnson, Anthony 2013. Health services review. fixed effects linear panel model. quality measure comes from hospital compare, not at the patient level.}
-
-\note[item]{Dranove, Forman, Goldfarb, Greenstein (2019) American Economic Journal: Economic Policy.
-data from 1996 to 2009.
-Specific findings: hospitals have a rise in cost after 3 years, some hospitals in certain locations have a decrease in costs after 5 years but others still have a higher cost after 6 years.}
-
-\end{frame}
-
-\begin{frame}[noframenumbering]{Literature}
-\footnotesize 
-\begin{columns}
-    \column{0.34\textwidth}
-    \color{gray}
-        \centering
-        \underline{ EHR $\rightarrow$ Healthcare }
-        \begin{itemize}
-        \color{gray}
-            \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
-
-            \footnotesize
-
-            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013)
-
-            \footnotesize
-
-            \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
-
-            \footnotesize
-
-            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
-
-        \end{itemize}
-
-    \column{0.33\textwidth}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
-        \end{itemize}
-
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        Physician Labor \underline{ Determinants }
-        \begin{itemize}
-        \color{gray}
-            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
-
-            \footnotesize
-
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
-
-            \footnotesize
-
-            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
-        \end{itemize}
-\end{columns}
-
-\note[item]{Zeira and Johnson 2011. }
-\end{frame}
-
-\begin{frame}[noframenumbering]{Literature}
-\footnotesize 
-\begin{columns}
-    \column{0.34\textwidth}
-    \color{gray}
-        \centering
-        \underline{ EHR $\rightarrow$ Healthcare }
-        \begin{itemize}
-        \color{gray}
-            \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)
-
-            \footnotesize
-
-            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013)
-
-            \footnotesize
-
-            \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)
-
-            \footnotesize
-
-            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)
-
-        \end{itemize}
-
-    \column{0.33\textwidth}
-    \color{gray}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \begin{itemize}
-        \color{gray}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2020)
-        \end{itemize}
-
-    \column{0.33\textwidth}
-        \centering
-        Physician Labor \underline{ Determinants }
-        \begin{itemize}
-            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
-            \tiny (Baltagi et al 2005, Sloan 2018)
-
-            \footnotesize
-
-            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
-            \tiny (Bahrami 2002)
-
-            \footnotesize
-
-            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
-            \tiny (Dewa et al 2014)
-        \end{itemize}
-\end{columns}
 \end{frame}
@@
 
-\begin{frame}{About the Data}
-    Main data: \underline{CMS Shared Patient Data} (2009-2015)
+\begin{frame}{Data}
+    \underline{Main: CMS Shared Patient Data (2009-2015)}
     \begin{itemize}
-        \item Lists 2 NPIs and the number of billings those two entities shared
+        \item Hospital-physician pairs: bill for same patient under Medicare
+        \item Physician (PCP, hospitalist, internist) works in the hospital
     \end{itemize}
+
     \vspace{3mm}
 
-    How I use it:
+    \underline{EHR Use:}
     \begin{itemize}
-        \item Physician-hospital pairs (PCPs, internists, hospitalists)
-        \item Billings that occur on same day
-        \item Close relationship between hospital and physician
-        \item Aggregate to the physician level
+        \item AHA Survey
     \end{itemize}
+
+    \vspace{3mm}
+
+    \underline{Other:}
+    \begin{itemize}
+        \item Physician Compare
+    \end{itemize}
+
+    \vspace{3mm}
+
+    Aggregated to physician level
 \end{frame}
 
-\begin{frame}{Measures of Physician Labor Market Decisions}
-
-     Patients visited in hospitals
-
-     \vspace{4mm}
-
-     Indicator for switching work setting:
+\begin{frame}{Variables}
+Labor Market Outcomes:
+\vspace{-2mm}
 \begin{itemize}
-        \item 0 patients visited but maintaining billable activity
+    \item Patients visited in hospitals
+    \item \color{gray} Maintaining billable activity only outside of hospitals
+    \item \color{gray} Level of billable activity outside of hospitals
+    \item \color{gray} Retirement/leaving the labor force
 \end{itemize}
 
-    \vspace{4mm}
+EHR Use:
+\vspace{-2mm}
+\begin{itemize}
+    \item Indicator for any exposure to EHR
+    \item \color{gray} Fraction of hospitals exposed to EHR
+\end{itemize}
 
-    \textcolor{gray}
-     {Indicator for retirement/ leaving the labor force}
-
-\end{frame}
-
-\begin{frame}{Measure of EHR Use}
-
-Physician-level treatment variable that captures exposure to \textit{any} EHR
-\vspace{3mm}
+Other:
+\vspace{-2mm}
 \begin{itemize}
-    \item Individual hospital EHR use comes from AHA survey
-    \vspace{3mm}
-    \item Treatment turns on the first time any of the physicians' hospitals implements an EHR
+    \item Years of experience
+    \item Hospital characteristics
 \end{itemize}
-\end{frame}
-
-\begin{frame}{Other Characteristics}
-
-Medical school graduation date: \underline{Physician Compare}
 
@@
     \centering
-    \includegraphics[scale=.8]{Objects/sumstats.pdf}
-
+    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
 \end{frame}
 
-\begin{frame}{Physician EHR Use}
+\begin{frame}{Data}
 \centering
-    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.5]{Objects/cont_treatment_graph.pdf}
+
 \end{frame}
@@
 \end{equation*}
+
+\vspace{4mm}
+
+\begin{align*}
+    y_{it} &: \text{Outcomes} \hspace{30mm}\\
+    \alpha_i, \delta_t &: \text{Fixed Effects}\hspace{30mm}\\
+    q_{it} &: \text{Physician/hospital characteristics}\hspace{30mm}\\
+    z_{i,t-k} &: \text{Exposure to EHR} \hspace{30mm}
+\end{align*}
 
@@
 
-\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample.pdf}
-            \caption{\small Phys. $>$ 35yrs\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_ind.pdf}
-            \caption{\small All Physicians}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_ind.pdf}
-            \caption{\small Phys. $>$ 35yrs Experience}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_ind.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_ind.pdf}
-            \caption{\small Phys. $>$ 35yrs Experience\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Appears in Office-Based Data}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_office.pdf}
-            \caption{\small All Physicians}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_office.pdf}
-            \caption{\small Phys.$>$35yrs Experience}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Dep. Variable: Appears in Office-Based Data}
-\begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_office.pdf}
-            \caption{\small All Physicians \\(Only 1 Hospital)}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_office.pdf}
-            \caption{\small Phys.$>$35yrs Experience\\ (Only 1 Hospital)}
-        \end{minipage}
-    \end{figure}
-\end{frame}
-
 \section{Continuing Work}

### 2021-11-01 Scripts/TYP_sumstats.R
fix EHR graph label

@@ -56,7 +56,7 @@ sum_stats_year <- sum_stats_year %>%
   dplyr::rename("Variable"="variable")
 
 ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) + 
-  geom_line() +geom_point() + labs(x="\nYear", y="Percent\n", 
+  geom_line() +geom_point() + labs(x="\nYear", y="Proportion\n", 
                                    title="\nEHR Use Over Time\n") + 
   scale_colour_manual(values=cbbPalette) + ylim(.2,1)  + theme(legend.key.size=unit(.3,'cm'),legend.key.height = unit(.4, 'cm'),legend.key.width = unit(.3, 'cm'))
 

### 2021-11-01 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
         \centering
-        \color{gray}
+
+        \pause
+
+        \color{blue}
+        \underline{ Contribution }
+        \vspace{1mm}
+
+        \color{blue}
+        Speaks to this question in a particular setting 
+            \begin{itemize}
+            \color{blue}
+                \item Important because health care spending is extremely high
+                \item Important to think about physician labor markets because of shortages and access to care issues
+            \end{itemize}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Literature}
+\small
+\begin{columns}
+\setlength{\tabcolsep}{-5pt}
+        \column{0.4\textwidth}
+        \centering
         \underline{ EHR $\rightarrow$ Healthcare }
@@
         \begin{itemize}
-        \color{gray}
             \item Small improvements for severe health conditions \\ \vspace{1mm}
@@
             \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
+        \end{itemize}
 
+        \pause
+
+        \column{0.4\textwidth}
+        \centering
+        \color{blue}
+        \underline{ Contribution }
+        \vspace{1mm}
+        \begin{itemize}
+        \color{blue}
+            \item No empirical work on EHR/physician effects
+            \item Speaks to the puzzling results
+            \item Time period 
         \end{itemize}
@@
             \includegraphics[width=\textwidth]{Objects/xtevent_fullsample.pdf}
-            \caption{\small All Physicians\\}
         \end{minipage}
@@
             \includegraphics[width=\textwidth]{Objects/xtevent_oldsample.pdf}
-            \caption{\small Phys. $>$ 35 yrs Experience}
         \end{minipage}
@@
         \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
-        \item Investigate other estimation strategies
+        \item Endogenous treatment? (Bandwidth, state privacy laws as IV)
+        \item Is there substitutability with other types of workers?
     \end{itemize}

### 2021-11-01 Presentations/ThirdYearPaperpres.tex

@@
     \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
-    \item Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs 
 \end{itemize}
@@
 
-\textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
+Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs
+\begin{itemize}
+    \item \textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
+\end{itemize}
 
@@
         \underline{ Contribution }
+
         \vspace{1mm}
 
-        \color{blue}
-        Speaks to this question in a particular setting 
             \begin{itemize}
@@
 Labor Market Outcomes:
-\vspace{-2mm}
+\begin{itemize}
+\item Productivity:
 \begin{itemize}
     \item Patients visited in hospitals
+\end{itemize}
+\end{itemize}
+\begin{itemize}
+\item Retirement
+\end{itemize}
+\begin{itemize}
+\item Work Setting:
+\begin{itemize}
     \item \color{gray} Maintaining billable activity only outside of hospitals
     \item \color{gray} Level of billable activity outside of hospitals
-    \item \color{gray} Retirement/leaving the labor force
+\end{itemize}
 \end{itemize}

### 2022-01-10 Paper/Paper.tex
Updates from Overleaf

@@
 
-Electronic Health Records (EHRs) have become increasingly relevant in the US since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
+The relationship between technology and labor markets is a longstanding topic of interest in economics and in policy making.
+This relationship is relevant in many aspects: technology causing displacement, whether low-skilled vs.
+high-skilled workers are differentially impacted, or the productivity gains that can be realized due to technology.
+In many capacities, across differing countries and industries, people tend to care about how technology affects their jobs and their children's future jobs.
+This paper seeks to understand this relationship under a major technology change in the U.S.
+healthcare system, the implementation of electronic health care record systems.
+
+Electronic Health Records (EHRs) have become increasingly relevant in the U.S.
+since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
 The movement towards digitization in health care was expected to have immediate and substantial impacts on cost and quality of care.
 In particular, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-The improvement in quality of care was expected due to physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance that could help physicians avoid harmful mistakes.
+Improvement in quality of care was expected due to physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance to avoid harmful mistakes.
 Such expectations on cost and quality led policymakers to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}).
@@
 Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
-The push towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
+The movement towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
 
@@
 What was expected to be a helpful tool in general has turned out to be burdensome and frustrating for the persons using it.
-In individual interviews, multiple physicians reported that when using EHRs, they are less satisfied with their job and have both higher burnout and stress levels.
+In individual interviews, multiple physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels.
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-A physician with the capability to stop working or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high.
-This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market behaviors, such as where to work, how much to work, and whether to work at all.
+A physician with the choice to stop working or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high.
+This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market behaviors: where to work, productivity levels, and retirement.
 
@@
 To measure the working decisions of physicians, I utilize (1) physician level information, also from the shared patient data, that captures total working life (referrals to labs, pharmacies, other physicians, etc), and (2) an office-based physician level data set to observe whether physicians are moving from hospitals to private practice.
-For my analysis, I use a difference-in-difference estimation where treatment is physicians' exposure to an EHR at any of the hospitals they work in.
+For my main analysis, I use an event study estimation where treatment is defined as a physicians' exposure to an EHR at any of the hospitals they work in.
 A key identifying assumption in this analysis is that hospitals choose to implement EHRs not based on any factors that are also correlated with physician working decisions.
 
-Paragraph about findings
+\textcolor{red}{Paragraph about findings}
 
-This paper contributes to two main portions of literature: the effects of health information technology and the factors of physician labor market decisions.
+This paper contributes to our understanding of two main ideas: the broad effects of technology on labor markets, and the more specific effects of EHR implementation in various healthcare dimensions.
+
+This setting is an important application to the broad understanding of how technology impacts labor for various reasons.
+First, health is a major expense in the United States, accounting for approximately 17 percent of GDP.
+The potential cost savings of EHRs are extremely important to understand given their potential to also improve patient satisfaction, as policymakers are continuously interested in achieving these two things simultaneously.
+Further, there has been discussion in both economics and medical literature about whether we have a physician shortage and, similarly, access to care issues in the U.S.
+(\cite{cooper2002economic}).
+Thus it is important to understand whether electronic health records are driving physicians out of the market or decreasing the amount of patients that can be seen by one physician, both of which could potentially worsen these issues.
+\textcolor{red}{This paragraph needs significant editing}
+
 The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
-Despite many case studies and hospital-level analysis that indicate large positive effects on health (\cite{Buntin2011TheResults}), economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements on average (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
-Further, there have been no significant cost decreases due to EHRs in the short or long run (CITE COST STUDIES).
-However, the long run is not fully realized, so there may still be potential benefits realized in the future.
-
-The relationship between this technology and physicians, which are the primary users of the technology, is ...
+Despite many case studies and hospital-level analysis that indicate large positive effects on health (\cite{Buntin2011TheResults}), economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
+Further, there have been no significant cost decreases due to EHRs in the short or long run.
+If any cost reductions are realized at all, it is not until at least 6 years after implementation (\cite{dranove2014trillion}).
+However, it may be that it takes more time to see the potential cost reductions realized, so there may still be potential benefits realized in the future.
+Finally, there have been preliminary case studies and interviews seeking to understand physicians' opinions on EHRs.
+These have drastically different findings depending on the setting studied, where there is evidence of older physicians having more frustration regarding the technology than younger physicians.
+This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) since these studies are all considering time periods before 2014, they do not capture the full period of EHR implementation.
 
@@
 
-\section{Empirical Analysis}
+Using various linked datasets, I construct a physician-level panel spanning from 2009-2015 which measures physicians' exposure to EHRs over time and other relevant characteristics.
+The data used to construct this panel are described in detail below.
 
-\section{Results}
+The main data I utilize is the CMS Shared Patients Data.
+For a pair of National Provider Identifiers (NPIs), this data records the number of patients who bill both of the entities under Medicare in the same day.
+For example, if a Medicare patient is referred to a specialist by a primary care physician, then those two entities have a shared patient in common.
+I limit the entities to only include the number of shared patients of physician-hospital pairs.
+I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
+To achieve this, I limit the qualifications for a physician-hospital pair in my data.
+The data only includes doctors who have a primary care type of role, excluding any specialists.
+The reason for this is that a primary care physician who does rounds in a hospital likely takes on a hospitalist role and does not do much office work.
+I further limit to those pairs which have a substantial number of patients billed together in the same day.
+This is to avoid including the potential physician who works in an office but happens to send a small number of patients to a nearby hospital.
+Substantial in this case is defined as having at least 30 patients bill both the physician and the hospital in the same day in at least one year, \textcolor{red}{where the exact number chosen is explored in the appendix}.
+With these limitations, the physicians left in the data are likely doing a number of rounds physically inside hospitals, since it is unlikely that many patients will see a primary care doctor in a private office and then immediately go to a hospital in the same day.
 
-\newpage
+Using hospital NPI, I link the physician-hospital pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
+I then aggregate to the physician level, where physician EHR use is measured as the percentage of their hospitals which use an EHR in a given year.
+Additional physician-level characteristics, such as years of experience, come from Physician Compare.
 
-\import{Objects/}{continuous.tex}
+Summary statistics from the data constructed are shown below. \textcolor{red}{Summary stats table} \textcolor{red}{Talk about what the summary stats show here}
 
-\newpage
+Figure 1 shows the adoption of EHRs by hospitals and the exposure to EHRs by physicians over time. \textcolor{red}{talk about graph here} \textcolor{red}{put graph here}
 
-\import{Objects/}{indicator.tex}
+\section{Retirement}
 
-\newpage
+\section{Work Setting}
 
-\import{Objects/}{continuous_old.tex}
-
-\newpage
-
-\import{Objects/}{indicator_old.tex}
-
-\newpage
+\section{Physician Productivity}
 

### 2022-01-10 Presentations/ThirdYearPaperpres.tex
Updates from Overleaf

@@
 
-\subtitle{Hanna Glenn}
+\subtitle{Hanna Glenn\\ \small Advisor: Dr. Ian McCarthy}
  \date{\today}
@@
         \item Productivity?
-        \item Wages?
     \end{itemize}
@@
     \item Intensive Margin / Productivity
-    \item Extensive Margin
+    \item Extensive Margin / Retirement
     \item Hospital $\rightarrow$ Office
@@
             \color{blue}
-                \item Important because health care spending is extremely high
-                \item Important to think about physician labor markets because of shortages and access to care issues
+                \item Healthcare spending is extremely high
+                \item Population of physicians - possible shortages / access to care issues
             \end{itemize}
@@
             \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
-
-            \footnotesize
-
-            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
-            \tiny (Appari et al 2013, HSR)
 
@@
 \begin{itemize}
-\item Retirement
+\item \color{gray} Retirement
 \end{itemize}
 \begin{itemize}
-\item Work Setting:
+\item \color{gray} Work Setting:
 \begin{itemize}
@@
 \begin{equation*}
-    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-5}^5 \beta_kz_{i,t-k} + \varepsilon_{it}
+    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-4}^4 \beta_kz_{i,t-k} + \varepsilon_{it}
 \end{equation*}
@@
 
-\section{Continuing Work}
+\begin{frame}{Summary}
+\begin{itemize}
+    \item Despite physician response to EHRs, those who stick around do become more productive
+    \vspace{3mm}
 
-\begin{frame}{Plans}
-    \begin{itemize}
-        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
-        \item Endogenous treatment? (Bandwidth, state privacy laws as IV)
-        \item Is there substitutability with other types of workers?
+    \item Complementary relationship between physicians and EHRs
+    \vspace{3mm}
+
+    \item Older physicians see (slightly) greater productivity increases 
+    \vspace{3mm}
+
 \end{itemize}
@@
 
-\begin{frame}{}
+\section{Continuing Work}
+
+\begin{frame}{Plans for this paper}
+    \begin{itemize}
+        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
+        \item See if there are any specific years driving these results
+        \item Endogenous treatment? (Bandwidth, state privacy laws as IV)
+    \end{itemize}
+\end{frame}
+
+\begin{frame}[plain]{}
 \centering

### 2022-01-13 Paper/Paper.tex

--- 2021-10-13 Paper/Paper.tex
@@
 \usepackage{fullpage}
+\usepackage{import}
+\usepackage{adjustbox}
+\usepackage{booktabs}
+\usepackage[font=footnotesize,labelfont=bf]{caption}
+\captionsetup{justification=raggedright,singlelinecheck=false}
 
@@
 
-Electronic Health Records (EHRs) have become increasingly relevant in the US since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
+The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making.
+Their interaction is relevant in several aspects: technology causing displacement, whether low-skilled vs.
+high-skilled workers are differentially impacted, or the productivity gains that can be realized due to technology.
+In many capacities, across differing countries and industries, people tend to care about how technology affects current and future labor markets.
+This paper seeks to better understand this relationship under a major technology change in the U.S.
+healthcare system, the implementation of electronic health care record systems in hospitals.
+
+Electronic Health Records (EHRs) have become increasingly relevant in the U.S.
+since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
 The movement towards digitization in health care was expected to have immediate and substantial impacts on cost and quality of care.
-In particular, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-The improvement in quality of care was expected due to physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance that could help physicians avoid harmful mistakes.
-Such expectations on cost and quality led policymakers to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}).
+For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
+Improvement in quality of care was expected as a result of physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance to avoid harmful mistakes.
+Such expectations about cost and quality led the government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}).
 This legislation subsidized hospitals which used EHRs “meaningfully”\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
@@
 Physicians are the primary users of this technology and have a significant role in determining whether potential benefits of EHRs are realized.
-However, the day-to-day life of physicians changed drastically with the implementation of EHRs.
-What was expected to be a helpful tool in general has turned out to be burdensome and frustrating for the persons using it.
-In individual interviews, multiple physicians reported that when using EHRs, they are less satisfied with their job and have both higher burnout and stress levels.
+However, the day-to-day life of physicians changed drastically with the rapid implementation of EHRs.
+What was expected to be a helpful tool turned out to be burdensome and frustrating for the persons using it.
+In individual interviews, multiple physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels.
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-A physician with the capability to stop working or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high.
-This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market behaviors, such as where to work, how much to work, and whether to work at all.
+A physician with the choice to retire or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high.
+This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market outcomes: where to work, productivity levels, and retirement.
 
@@
 I link the hospitals to AHA survey data for information on EHR use and then aggregate the data to the physician level.
-To measure the working decisions of physicians, I utilize (1) physician level information, also from the shared patient data, that captures total working life (referrals to labs, pharmacies, other physicians, etc), and (2) an office-based physician level dataset to observe whether physicians are moving from hospitals to private practice.
-For my analysis, I use a difference-in-difference estimation where treatment is defined as physicians' exposure to an EHR at any of the hospitals they work in.
-A key identifying assumption in this analysis is that hospitals choose to implement EHRs not based on any factors that are also correlated with physician working decisions.
+The key independent variable for my main analysis is an individuals physicians' exposure to electronic health records.
+There are three outcomes of interest in this paper.
+First, I consider the decision for a physician to retire.
+\textcolor{red}{This is measured using blank}.
+Second, I consider whether those physicians who stay in the labor force choose to substitute work towards office-based practices.
+\textcolor{red}{This information comes from blank.} Finally, for physicians who continue working in hospitals, I analyze the productivity impacts of EHRs.
+The measure of productivity used is the number of Medicaid patients that bill the both the physician and a hospital in the same day, which comes from the CMS Shared Patient Data.
 
-\newpage
-\printbibliography
+The estimation is an event-study approach, where treatment is whether a physician is exposed to an EHR.
+I assume the decision of a hospital to implement an EHR is exogenous to the physician, but explore this assumption in more detail in Section \textcolor{red}{blank}.
+\textcolor{red}{To extend the analysis I also utilize difference in difference with continuous treatment, using fraction of hospitals using an EHR as the treatment variable.}
+
+\textcolor{red}{Paragraph about findings}
+
+This paper contributes to the general understanding of two main questions: the broad effects of technology on labor markets, and the more specific effects of EHR implementation in various healthcare dimensions.
+
+How technology affects labor market outcomes depends drastically on the industry and setting being considered.
+Health is a major component of expense in the United States, accounting for approximately 17 percent of GDP.
+Within the health sector, health information technology has been a major change implemented widely.
+There is a continual discussion amongst policymakers and researchers about how to lower healthcare spending without sacrificing quality of care.
+Thus, the potential cost savings of EHRs are extremely important to understand given their potential to also improve patient satisfaction, since there is a continual interest in achieving these two things simultaneously.
+Further, there is the question in both economics and medical literature of whether there exists a physician shortage and, similarly, access to care issues in the U.S.
+(\cite{cooper2002economic}).
+A relevent factor of this discussion is whether electronic health records are driving physicians out of the market or decreasing the amount of patients that can be seen by one physician, both of which have the potential to worsen these issues.
+
+The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
+Despite many case studies and hospital-level analysis that indicate large improvements in health outcomes such as mortality (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
+Further, there have been no significant cost decreases due to EHRs in the short or long run.
+If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}).
+However, the incentive programs put in place could impact the cost effects of EHRs, and 6 years may not be a long enough time frame to see the full picture of cost effects.
+While this paper does not specifically investigate hospital costs, it does speak to potential drivers of cost effects.
+Finally, there have been preliminary case studies and interviews seeking to understand physicians' outlook on EHRs.
+These differ depending on the setting studied, where there is evidence of older physicians having more frustration regarding the technology than younger physicians.
+\textcolor{red}{I need a citation here and probably to expand more on the idea of physician opinions.}
+
+This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010, capturing more so the effect of computerization.
+
+\section{Background}
+
+\textcolor{red}{Need to figure out what goes in this section. History of EHRs? Specifics of the subsidies and meaningful use?}
+
+\section{Data}
+
+Using various linked datasets, I construct a physician-level panel spanning from 2009-2015 which measures physicians' exposure to EHRs over time and other relevant characteristics.
+The different data sets used to construct this panel are described below.
+
+The main data I utilize is the CMS Shared Patients Data.
+For a pair of National Provider Identifiers (NPIs), it records the number of patients who bill both of the entities under Medicaid in the same day.
+For example, if a Medicaid patient is referred to a specialist by a primary care physician, then those two physicians have a shared patient in common.
+I limit the entities to only include the number of shared patients for physician-hospital pairs.
+I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
+To achieve this, I limit the qualifications for a sufficient physician-hospital pair.
+First, I only consider doctors who have a primary care role, excluding any specialists.
+The reason for this is that a primary care physician who does rounds in a hospital is likely to be primarily employed by that hospital (not brought in from a practice).
+Another reason for this is that primary care doctors interact directly with EHRs during checkups, unlike a surgeon.
+Second, I limit pairs to those which have a substantial number of patients billed together in the same day.
+This is to avoid including the potential physician who works in an office but happens to send a small number of patients to a nearby hospital.
+I consider a substantial number of patients to be at least 30 patients bill both the physician and the hospital in the same day in at least one year, \textcolor{red}{where the exact number chosen is explored in the appendix}.
+The physicians remaining in the data are likely doing a number of rounds physically inside hospitals, since it is unlikely that many patients will see a primary care doctor in a private office and then immediately go to a hospital in the same day.
+
+Using hospital NPI, I link the physician-hospital pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
+I then aggregate to the physician level, where physician EHR use is measured as the percentage of their hospitals which use an EHR in a given year.
+Additional physician-level characteristics, such as years of experience, come from Physician Compare.
+
+Overall summary statistics of the variables are shown in \hyperref[fig:summarystatistics]{Table 1}.
+Average hospital operating days is a physician-level variable that averages the number of operating days for that physicians hospital network.
+While there are some low outliers with a low value, on average physicians are working in hospitals open year round.
+Similarly, the average size of the hospital worked with averages the number of beds of each hospital in a physician's network.
+While these two average variables can technically vary by year, they tend to stay constant across the data, which is why they are included in the static category.
+Approximately 30\% of the sample of physicians are female and the average physician graduated medical school in 1993.
+Note that there are no physicians in the data who graduated after 2009, since they would be entering the labor force during the span of the data.
+On average, a physician works with between 1 and 2 hospitals as well as hospital systems.
+While the number of hospitals that a physician cares for patients in varies by year, the number of hospitals in their network is constant.
+The physician level variables that vary over time deal are the treatment variables regarding EHR exposure and the number of Medicaid patients seen.
+For all physician-years, physicians were exposed to an EHR in 56\% of the hospitals they work in and 70\% of the sample is exposed to any EHR.
+Once a physician becomes exposed to an EHR, they cannot go back to being unexposed.
+Finally, physicians are seeing an average of 620 Medicaid patients in hospitals per year.
+\textcolor{red}{In Appendix A1, I include summary statistics by subsample of physician age and year.}
+
+\begin{table}[ht]
+    \import{Table Code}{overall stats.tex}
+    \label{fig:summarystatistics}
+    \vspace{1mm}
+    \caption*{\footnotesize Obs: 1,044,001}
+\end{table}
+
+For a clear look at the variation in treatment variables over time, \hyperref[fig:figure1]{Figure 1} shows the average of these variables over time.
+In 2009, 42\% of the sample of physicians were exposed to an EHR.
+That is, over half of the sample had no affiliation with EHRs at the beginning of the sample period.
+By 2015, 92\% of physicians had exposure to an EHR in at least one hospital in their network.
+Similarly, the fraction of hospitals in a physician's network which were using an EHR was 30\% in 2009 and grew to 82\% in 2015.
+These statistics are consistent with the surge in incentives to implement EHRs in hospitals.
+
+\begin{figure}
+    \caption{Treatment Variables Over Time}
+    \vspace{-2mm}
+    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
+    \label{fig:figure1}
+\end{figure}
+
+\section{Retirement}
+
+\section{Work Setting}
+
+\section{Physician Productivity}
+
+An important aspect in the implementation of EHRs is whether or not they increase productivity in users.
+Productivity, in this case, has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
+Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
+That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
+To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have patients in hospitals even after EHR exposure.
+
+It is reasonable to believe that physicians are fully utilizing the EHRs in the hospital they work in, since by 2015 electronic health records were widely implemented and considered unavoidable.
+An objection to this assumption may be that older physicians stay in a hospital, but the hospital hires a data assistant to do technology work for the physician.
+In this cae, the productivity gain could be from having the data assistant instead of EHR use.
+I investigate this below.
+
+\subsection{Data Assistants}
+
+\textcolor{red}{get data from AHA on data assistance. It may be that I just need to show a graph here of the number of data assistant jobs at hospitals. Need to see from there. }
+
+\subsection{Event Study and Results}
+
+The analysis in this section focuses on the effects of hospital EHR implementation on physician productivity within hospitals, measured by the total number of Medicaid patients billed in hospitals in a given year.
+The estimating equation is given as follows:
+
+\begin{equation*}
+    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-5}^5 \beta_kz_{i,t-k} + \varepsilon_{it},
+\end{equation*}
+
+where $y_{it}$ is the number of patients, $\alpha_i$ and $\delta_i$ are physician and time fixed effects, $q_{it}$ are hospital characteristics such as number of beds and days open during the year, $z_{i,t-k}$ is an indicator for whether the physician was exposed to an EHR in any hospital in time $t-k$, and $\epsilon_{it}$ is a random error term.
+
+The identifying assumption is that physicians are exposed endogenously to electronic health records.
+That is, hospital management decides and implements EHRs, and this decision is not correlated with physician productivity or labor market decisions.
+This is not unreasonable for large hospitals with ample resources and many physicians.
+However, a small hospital may give experienced physicians some level of voice in the decision to implement new technologies.
+Results under this assumption are shown below, and  \textcolor{red}{the strength of this assumption is considered in the next section.
+}
+
+An event study plot containing estimates and confidence intervals of  $\{\beta_k\}_{k=-5}^5$  is given in in Figure \textcolor{red}{number}.
+The left hand side is the full sample of physicians at any experience level.
+As is standard in event studies, the identification of the coefficients depends on the assumption that there are no pre-trends.
+The p-value of a Wald test for pre-trends in both event studies indicate no observable pre-trends in the data.
+
+For physicians at any experience level, EHR exposure yields no immediate affect on physician productivity, but leads to a clear increase in the number of patients seen beginning 2 years after exposure.
+After 5 years of exposure, physicians are seeing, on average, 93.67 more Medicaid patients per year.
+Relative to the mean in the year before exposure, 796.95, this is an 11.75\% increase in the number of patients seen.
+When the sample of physicians is limited to senior physicians (at least 35 years of experience), this effect disappears.
+In fact, the results show a slight decrease of \textcolor{red}{put the exact number here} in the number of patients seen in the year after exposure, and no effect following that year.
+This graph limits event time from -4 to 4 due to restricted sample size in year 2010 and 2014.
+
+\begin{figure}[ht]
+\caption{}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/prod_eventstudy_fullsample.pdf}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/prod_eventstudy_oldsample.pdf}
+        \end{minipage}
+\end{figure}
+
+\subsection{Relaxing Exogeneity of Implementation}
+
+\textcolor{red}{I need to put somewhere that i dropped 2015 for this analysis since it cut off at a different point than the other years}
+
 \end{document}

### 2022-01-13 Table Code/overall stats.tex

\footnotesize
\caption{\footnotesize Summary Statistics}
\vspace{-2mm}
\begin{tabular}[t]{lcccc}
\toprule
Variable & Mean & Std. Dev. & Min & Max\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Static}}\\
\hspace{1em}Average Hospital Operating Days  & 364.89 & 5.11 & 41 & 366\\
\hspace{1em}Average Size of Hospitals Worked With (Beds)  & 373.12 & 252.40 & 10 & 2574\\
\hspace{1em}Female  & 0.31 & 0.46 & 0 & 1\\
\hspace{1em}Medical School Grad. Year  & 1993 & 10.04 & 1949 & 2008\\
\hspace{1em}Number of Hospitals Worked With  & 1.87 & 1.30 & 1 & 40\\
\hspace{1em}Number of Systems Worked With  & 1.48 & 0.77 & 1 & 13\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Dynamic}}\\
\hspace{1em}Exposure to an EHR  & 0.70 & 0.46 & 0 & 1\\
\hspace{1em}Fraction of Hospitals with EHR  & 0.56 & 0.44 & 0 & 1\\
\hspace{1em}Number of Patients & 620.54 & 988.74 & 0 & 74639\\
\bottomrule
\end{tabular}

### 2022-01-13 Paper/Paper.tex

@@
 An objection to this assumption may be that older physicians stay in a hospital, but the hospital hires a data assistant to do technology work for the physician.
-In this cae, the productivity gain could be from having the data assistant instead of EHR use.
+In this case, the productivity gain could be from having the data assistant instead of EHR use.
 I investigate this below.
@@
 
-\textcolor{red}{get data from AHA on data assistance. It may be that I just need to show a graph here of the number of data assistant jobs at hospitals. Need to see from there. }
+One concern with this measure of productivity is that an increase in the number of patients seen by a physician could be caused by an outside factor that is also correlated with EHR adoption in hospitals.
+A logical outside factor could be the existence of employees with the purpose of utilizing electronic health records on a physician's behalf.
+Employees of this type are assigned a medical tax ID when working in hospitals; they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+All of these employees are the equivalent of transcribers for health information technology.
+
+Using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
+Figure \textcolor{red}{number} is a frequency plot of the year of activation for every tax code that falls in the categories listed above.
+The total number of these registered employees is 875.
 
@@
 
+\renewcommand*{\bibfont}{\footnotesize}
+
+\printbibliography
+
+\newpage
+
+\appendix
+
+\section{Tables and Figures}
+
+\begin{table}[ht]
+\import{Table Code}{split stats.tex}
+\end{table}
+
 \end{document}

### 2022-01-21 Table Code/split stats.tex

\scriptsize
\begin{tabular}[t]{lcccccccccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Senior Physicians} & \multicolumn{6}{c}{Young Physicians} \\
\cmidrule(l{3pt}r{3pt}){2-7} \cmidrule(l{3pt}r{3pt}){8-13}
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{\# of Patients} & \multicolumn{2}{c}{Frac. Hosp. w/ EHR} & \multicolumn{2}{c}{Exposure EHR} & \multicolumn{2}{c}{\# of Patients} & \multicolumn{2}{c}{Frac. Hosp. w/ EHR} & \multicolumn{2}{c}{Exposure EHR} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9} \cmidrule(l{3pt}r{3pt}){10-11} \cmidrule(l{3pt}r{3pt}){12-13}
Year & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev.\\
\midrule
2009 & 928 & 1304. & 0.27 & 0.39 & 0.38 & 0.49 & 611 & 976 & 0.30 & 0.40 & 0.42 & 0.49\\
2010 & 734 & 1172 & 0.32 & 0.41 & 0.47 & 0.50 & 512 & 912 & 0.34 & 0.42 & 0.49 & 0.50\\
2011 & 910 & 1268 & 0.42 & 0.44 & 0.60 & 0.49 & 666 & 1008 & 0.44 & 0.44 & 0.62 & 0.48\\
2012 & 888 & 1317 & 0.54 & 0.44 & 0.71 & 0.45 & 684 & 1047 & 0.57 & 0.43 & 0.74 & 0.44\\
2013 & 829 & 1298 & 0.64 & 0.42 & 0.81 & 0.39 & 667 & 1045 & 0.67 & 0.41 & 0.83 & 0.38\\
\addlinespace
2014 & 836 & 1272 & 0.73 & 0.39 & 0.87 & 0.34 & 705 & 1062 & 0.75 & 0.37 & 0.89 & 0.32\\
2015 & 482 & 752 & 0.81 & 0.34 & 0.92 & 0.28 & 418 & 636 & 0.83 & 0.33 & 0.93 & 0.26\\
\bottomrule
\end{tabular}

### 2022-02-18 Scripts/AHANPI_cw.R
remove old event study graphs from objects folder and update scripts after recording all judgement calls made

@@ -1,328 +1,0 @@
-library(dplyr)
-library(readr)
-
-# -------------------------------   Create NPI-AHA Crosswalk -----------------------------
-#                                   Hanna Glenn
-#                                   8/27/2021
-
-# This script reads in data from the AHA survey with connected NPI numbers. Many hospitals
-# will skip this question in all or some years. I will fill in any NPI from NPI lookup
-# that are missing and match in name and location. I will also observe when there are 
-# multiple NPIs connected to one AHAID and address this. The resulting dataset is 
-# "AHANPI_cw.csv". 
-
-AHA.NPI <- read_csv(paste0(path.currentdata,"AHA-NPI.csv"))
-  # 44077 obs of 10 variables
-
-AHA.NPI <- AHA.NPI %>%
-  filter(MSTATE %in% c("AL","AK","AZ","AR","CA","CO", "CT","DE","DC","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD",
-                       "MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY")) %>%
-  rename(AHAID=ID, Hospname=MNAME, Street=MLOCADDR, City=MLOCCITY, Beds=BDTOT, State=MSTATE, Zip=MLOCZIP, NPI=NPINUM) %>%
-  select(YEAR, AHAID, NPI, Hospname, Street, City, State, Zip, Beds)
-  # 43615 obs of 9 variables
-
-# Create a dataset to see how many AHA IDs are missing an NPI (number of obs gives number of unique AHA IDs)
-num_missing_npi <- AHA.NPI %>%
-  group_by(AHAID) %>%
-  filter(is.na(NPI)) %>%
-  count() 
-  # 3162 obs 
-
[…]

### 2022-02-18 Scripts/analysis.R

@@ -1,6 +1,6 @@
 library(did)
 library(dplyr)
-library(fixest)
+library(ggplot2)
 
 # ------------------------------------- ANALYSIS  ------------------------------------
 #                                       Hanna Glenn, Emory University
@@ -12,81 +12,115 @@ library(fixest)
 
 Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
+
 # Retirement ---------------------------------------------------------------------------
 
-retire_es_all <- feols(retire~ rel_m4+rel_m3+ rel_m2+ rel_0+ rel_p1+
-                                rel_p2+ rel_p3 + rel_p4 + age| DocNPI+year, 
-                              data=filter(Physician_Data,year<2017))
-retire_es_young <- feols(retire~ rel_m4+rel_m3+ rel_m2+ rel_0+ rel_p1+
-                         rel_p2+ rel_p3 + rel_p4 + age| DocNPI+year, 
-                       cluster="DocNPI",
-                       data=filter(Physician_Data,age<50 & year<2017))
-retire_es_old <- feols(retire~ rel_m4+rel_m3+ rel_m2+ rel_0+ rel_p1+
-                         rel_p2+ rel_p3 + rel_p4 + age| DocNPI+year, 
-                       cluster="DocNPI",
-                       data=filter(Physician_Data,age>=50 & year<2017))
-reference=4
-names(reference)<-"-1"
-
-coefplot(list(retire_es_all,retire_es_old), keep=c("-4","-3","-2","-1","0",
-                                                    "1", "2","3","4"),
-         ref=reference, dict=c("rel_m4"="-4", "rel_m3"="-3", "rel_m2"="-2", "rel_0"="0",
-                               "rel_p1"="1","rel_p2"="2","rel_p3"="3","rel_p4"="4"),
-         main="",
-         zero=TRUE,
-         col=1:2)
-
-wald_young<-wald(any_chronic_es_young, keep=c("rel_m4","rel_m3","rel_m2"), cluster="st_fips")
-p_young <- wald_young$p
-
-wald_old<-wald(any_chronic_es_old, keep=c("rel_m4","rel_m3","rel_m2"), cluster="st_fips")
-p_old <- wald_old$p
-
-legend("topright", col = 1:2, pch = 20, lwd = 1, lty = 1:2,
-       legend = c("< 35 Yrs.", "35-55 Yrs."))
-
[…]

### 2022-02-18 Scripts/data1_npi_list.R

@@ -36,10 +36,29 @@ MDPPAS <- MDPPAS %>%
 MDPPAS <- MDPPAS %>%
   dplyr::filter(sum>0) 
 
-# Create a list of npi numbers to use when creating the shared patient data
+# Figure out what to do with doctors who switch specialties
+switchers <- MDPPAS %>%
+  dplyr::filter(sum==1) %>%
+  dplyr::group_by(npi) %>%
+  dplyr::mutate(n=1) %>%
+  dplyr::mutate(yrs=sum(n)) %>%
+  dplyr::ungroup() %>%
+  dplyr::filter(yrs>1) %>%
+  dplyr::mutate(drop=1)
+
+switchers <- switchers %>%
+  dplyr::distinct(npi,drop)
+
 npi_list <- MDPPAS %>%
+  dplyr::left_join(switchers,by="npi") %>%
+  dplyr::filter(is.na(drop))
+
+# Create a list of npi numbers to use when creating the shared patient data
+npi_list <- npi_list %>%
   dplyr::distinct(npi)
-  # 245k obs
+  # 243k obs
+
+
 
 saveRDS(npi_list,file=paste0(created_data_path,"npi_list.rds"))
 

### 2022-02-18 Scripts/data2_npidata.R

@@ -1,5 +1,6 @@
 library(readr)
 library(dplyr)
+library(stringr)
 
 # ------------------  Building NPI Taxonomy Dataset ------------------------------------- 
 #                     Hanna Glenn, Emory University
@@ -42,7 +43,8 @@ npidata_type <- npidata %>%
   mutate(PCP=ifelse(str_detect(t_class,"Internal Medicine") |
                       str_detect(t_class,"Hospitalist") | 
                       str_detect(t_class,"Family Medicine") |
-                      str_detect(t_class,"General Practice"),1,0),
+                      str_detect(t_class,"General Practice") |
+                      str_detect(t_class,"Pediatrics"),1,0),
          hospital=ifelse(str_detect(t_group,fixed("hospital", ignore_case=TRUE)),1,0))
 
 # Create specialist indicator

### 2022-02-18 Scripts/data3_pairs.R

@@ -15,9 +15,9 @@ library(janitor)
 # This script reads in (1) the CMS Physician Shared Patient Data from 2009-2015, (2) phys_npidata and hosp_npidata that were
 # created in "data2_npidata.R" and (3) the npi list created in "data1_MDPPAS.R".
 # These data are all merged to end up with qualifying physician hospital pairs and their shared count. 
-# The resulting dataset, called "data2_pairs.rds" consists of a balanced dataset of physician hospital pairs 
+# The resulting dataset, called "data2_pairs.rds" consists of a dataset of physician hospital pairs 
 # from 2009-2015.
-# Note: preliminary filtering is done on these pairs to only save those with more than 30 shared patients total. 
+# Note: preliminary filtering is done on these pairs to only save those with sufficient number of patients. 
 
 # Read in npi datasets and combine into one------------------
 phys_npidata <- read_rds(paste0(created_data_path,"phys_npidata.rds"))
@@ -58,7 +58,7 @@ for (t in 2009:2015){
 }
 
 phys_hosp_pairs <- rbind(PSPD2009, PSPD2010, PSPD2011, PSPD2012, PSPD2013, PSPD2014, PSPD2015)
-  # 12.5 mill observations
+  # 12.6 mill observations
 
 # Change to physician and hospital npis instead of npi1 and npi2
 phys_hosp_pairs <- phys_hosp_pairs %>%
@@ -84,7 +84,7 @@ phys_hosp_pairs <- phys_hosp_pairs %>%
   select(year,HospNPI,DocNPI,samedaycount_combine) %>%
   distinct() %>%
   rename(samedaycount=samedaycount_combine)
-  # 7 million obs. 
+  # 7.1 million obs. 
 
 
 # Merge to NPI List of Physicians----------------
@@ -97,52 +97,26 @@ phys_hosp_pairs <- phys_hosp_pairs %>%
   select(-keep)
   # 3.7 mill obs.
 
-# Filter out any hospital-physician pairs that have sum less than 100(?) shared patients in all years
+# Filter out any hospital-physician pairs that have number of patients less 
+# than 30*number of years in the data
 phys_hosp_pairs <- phys_hosp_pairs %>% ungroup() %>%
[…]

### 2022-02-18 Scripts/data4_physician_level.R

@@ -19,11 +19,10 @@ library(kableExtra)
 # Read in physician hospital pairs created in "phys_hosp_pairs.r"
 data <- read_rds(paste0(created_data_path, "phys_hosp_pairs.rds"))
   # This data does not have any repeats of year, HospNPI, DocNPI
-  # 1.75 mill obs, 6 variables
+  # 1.3 mill obs, 5 variables
 
 data <- data %>%
-  rename(pairID=ID) %>%
-  select(-keep)
+  rename(pairID=ID) 
 
 # Create Experience Variable ------------------------------------------------------------
 # Read in Physician Compare
@@ -38,7 +37,7 @@ data <- data %>%
   left_join(PhysCompare, by=c("DocNPI"="NPI")) %>%
   distinct() %>%
   filter(grad_year<2009)
-  # Now have 1.27 mill obs
+  # Now have 1 mill obs
 
 
 #### Create EHR Variables ---------------------------------------------------------------------- ####
@@ -47,18 +46,19 @@ AHANPI_cw <- read_rds(paste0(created_data_path,"AHANPI_cw.rds"))
 
 # Merge crosswalk to main data
 data <- data %>%
+  mutate(HospNPI=as.double(HospNPI)) %>%
   left_join(AHANPI_cw,by=c("HospNPI"="NPI"))
 
 # See how many unique hospitals in the dataset have an AHAID
 num_hosp <- data %>% ungroup() %>%
   filter(!is.na(AHAID)) %>%
   distinct(HospNPI)
-  # There are 4225 unique hospitals in the data that are not missing AHAID (good)
+  # There are 4271 unique hospitals in the data that are not missing AHAID (good)
 
 # Only keep pairs with AHA hospitals since that is where EHR information comes from
 data <- data %>%
   filter(!is.na(AHAID))
-  #1 mill
+  #857k obs
 
 # Read in main AHA survey data
 # I need to think about adding more hospital characteristics to this dataset
@@ -72,7 +72,8 @@ AHAmainsurvey <- AHAmainsurvey %>%
   rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
          joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
[…]

### 2022-02-18 Scripts/data5_MDPPAS.R

@@ -1,6 +1,8 @@
 library(readr)
 library(tidyr)
 library(stringr)
+library(dplyr)
+library(stringi)
 
 # ------------------------------------- INITIALIZING PHYSICIAN DATA WITH MDPPAS ------------------------------------
 #                                       Hanna Glenn, Emory University
@@ -18,7 +20,8 @@ for (i in 2009:2017) {
     dplyr::select(-name_middle, -name_first, -spec_broad, -spec_prim_1, -spec_prim_2,
                   -phy_zip7, -claim_count7, -phy_zip8, -claim_count8, -phy_zip9, -claim_count9,
                   -phy_zip10, -claim_count10, -phy_zip11, -claim_count11, -phy_zip12, 
-                  -claim_count12)
+                  -claim_count12, -spec_prim_2_name, -tin1_legal_name, -tin2_legal_name,
+                  -group1, -group2)
   assign(paste0("MDPPAS",i),year)
 }
 
@@ -28,20 +31,6 @@ MDPPAS <- rbind(MDPPAS2009, MDPPAS2010, MDPPAS2011, MDPPAS2012, MDPPAS2013, MDPP
 # Read in Aggregated Pairs data from "data4_physician_level.R"
 Aggregated_Pairs <- read_rds(paste0(created_data_path,"Aggregated_Pairs.rds"))
 
-Aggregated_Pairs <- complete(Aggregated_Pairs,DocNPI,year=2009:2017)
-
-# fill time invariant variables 
-Aggregated_Pairs <- Aggregated_Pairs %>%
-  dplyr::group_by(DocNPI) %>%
-  tidyr::fill(grad_year,.direction="downup") %>%
-  tidyr::fill(minyr_EHR,.direction="downup") %>%
-  tidyr::fill(anyEHR_exposed,.direction="down") %>%
-  tidyr::fill(num_hospitals_constant,.direction="downup") %>%
-  tidyr::fill(num_systems,.direction="downup") %>%
-  tidyr::fill(exposed_ever,.direction="downup") %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(experience=ifelse(is.na(experience),year-grad_year,experience))
-  
 
 # Merge MDPPAS to main dataset and fill time invariant variables
 Physician_Data <- Aggregated_Pairs %>%
@@ -59,11 +48,12 @@ Physician_Data <- Physician_Data %>%
[…]

### 2022-02-18 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{setspace}
-\onehalfspacing
+\doublespacing
 \addtolength{\skip\footins}{2pc plus 5pt}
@@
 The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making.
-Their interaction is relevant in several aspects: technology causing displacement, whether low-skilled vs.
-high-skilled workers are differentially impacted, or the productivity gains that can be realized due to technology.
-In many capacities, across differing countries and industries, people tend to care about how technology affects current and future labor markets.
-This paper seeks to better understand this relationship under a major technology change in the U.S.
-healthcare system, the implementation of electronic health care record systems in hospitals.
+Their interaction has direct impacts across industries; job displacement, creation, and satisfaction are all prevalent reality for many workers (\textcolor{red}{PUT SOME CITATIONS HERE}).
+Additionally, in some cases, labor market changes that take place for one job can have rippling effects to populations of people.
+One such job is that of a physician, where physician decisions have direct and indirect effects on patients receiving care.
+For example, a physician experiencing burnout from work pressures may treat patients differently and lead to patients feeling uncomfortable seeking care (\textcolor{red}{FIND A CITE FOR THIS)}, or increased administrative burdens could lead to less time with patients and exacerbated access to care issues for those in rural settings (\textcolor{red}{FIND CITE FOR THIS}).
+There may also be positive effects on patients from physician labor market changes, such as increased access to care that comes from more physicians entering the market, or less harmful mistakes that arise from standardizing care given.
+For these reasons, I investigate physician labor markets under a major technology change in the U.S.
+healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce labor market changes in the occupation.
+\textcolor{red}{Rethink how I frame the examples, but MAKE SURE I HAVE CITATIONS FOR THESE THINGS}
 
-Electronic Health Records (EHRs) have become increasingly relevant in the U.S.
+Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care.
+EHRs have have become increasingly relevant in the U.S.
 since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
-The movement towards digitization in health care was expected to have immediate and substantial impacts on cost and quality of care.
-For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-Improvement in quality of care was expected as a result of physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance to avoid harmful mistakes.
-Such expectations about cost and quality led the government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}).
-This legislation subsidized hospitals which used EHRs “meaningfully”\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency of care.
+A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
+Such expectations led the U.S.
+government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
-The movement towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
+The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.
 
-Specifically, an EHR is digital version of a patient’s medical records.
-These include detailed accounts and notes of medical history and can become advanced enough to provide the physician suggestions for care.
-Physicians are the primary users of this technology and have a significant role in determining whether potential benefits of EHRs are realized.
-However, the day-to-day life of physicians changed drastically with the rapid implementation of EHRs.
-What was expected to be a helpful tool turned out to be burdensome and frustrating for the persons using it.
-In individual interviews, multiple physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels.
+Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized.
+However, with the rapid implementation of EHRs, the day to day life of physicians changed drastically.
+In interviews, groups of physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels.
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-A physician with the choice to retire or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high.
-This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market outcomes: where to work, productivity levels, and retirement.
+The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices.
+Further, the short and long run impacts of EHRs on physician productivity could vary greatly.
+I use modern event study approaches, including continuous treatment, to understand the impact of EHR implementation in hospitals on physician labor market outcomes: retirement, where to work, and productivity in hospitals.
 
-Using CMS Shared Patient Data from 2009-2015, I construct a panel of physician-hospital pairs that captures a specific working relationship between the pair, most likely the physician working in the hospital.
-I link the hospitals to AHA survey data for information on EHR use and then aggregate the data to the physician level.
-The key independent variable for my main analysis is an individuals physicians' exposure to electronic health records.
-There are three outcomes of interest in this paper.
-First, I consider the decision for a physician to retire.
-\textcolor{red}{This is measured using blank}.
-Second, I consider whether those physicians who stay in the labor force choose to substitute work towards office-based practices.
-\textcolor{red}{This information comes from blank.} Finally, for physicians who continue working in hospitals, I analyze the productivity impacts of EHRs.
-The measure of productivity used is the number of Medicaid patients that bill the both the physician and a hospital in the same day, which comes from the CMS Shared Patient Data.
+Using CMS shared patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct physician-level data on hospitalists and other general practice physicians connected to hospitals from 2009-2017.
+The data captures the extent to which the physician in exposed to EHRs in hospitals and various labor market outcomes, along with demographic information such as zip code, age, and gender.
+With this data I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician.
+Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals.
+For retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it.
+Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here (if no effect, take out the first part of this sentence)}.
 
-The estimation is an event-study approach, where treatment is whether a physician is exposed to an EHR.
-I assume the decision of a hospital to implement an EHR is exogenous to the physician, but explore this assumption in more detail in Section \textcolor{red}{blank}.
-\textcolor{red}{To extend the analysis I also utilize difference in difference with continuous treatment, using fraction of hospitals using an EHR as the treatment variable.}
+Given that there is a population of both younger and older physicians who stay in the labor force (\textcolor{red}{it may be that all of them stay in the market}), the next outcome to consider is whether physicians switch work settings as a result of EHR implementation in hospitals.
+That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different?
+Here, I limit the sample to any physicians who stay in the labor force after EHR exposure.
+I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.
 
-\textcolor{red}{Paragraph about findings}
+These findings imply that there are still physicians who remain employed by the same hospital after EHR implementation.
+Thus, I study the effect of the technology on productivity within hospitals, where productivity is measured by the number of patients seen.
+The population of physicians left in this analysis are those who do not retire or switch to an office based setting.
+I find that productivity does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains.
+After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs.
+This result speaks to the disconnect between expected effects and realized effects of EHRs which is discussed in the health information technology literature.
+Since more patients are being seen by physicians, efficiency and access to care have both increased while costs (Medicaid patients) and quality remain unchanged.
+\textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}
 
-This paper contributes to the general understanding of two main questions: the broad effects of technology on labor markets, and the more specific effects of EHR implementation in various healthcare dimensions.
+\textcolor{red}{Add a paragraph discussing sensitivity analysis or mechanisms.}
 
-How technology affects labor market outcomes depends drastically on the industry and setting being considered.
-Health is a major component of expense in the United States, accounting for approximately 17 percent of GDP.
-Within the health sector, health information technology has been a major change implemented widely.
-There is a continual discussion amongst policymakers and researchers about how to lower healthcare spending without sacrificing quality of care.
-Thus, the potential cost savings of EHRs are extremely important to understand given their potential to also improve patient satisfaction, since there is a continual interest in achieving these two things simultaneously.
-Further, there is the question in both economics and medical literature of whether there exists a physician shortage and, similarly, access to care issues in the U.S.
-(\cite{cooper2002economic}).
-A relevent factor of this discussion is whether electronic health records are driving physicians out of the market or decreasing the amount of patients that can be seen by one physician, both of which have the potential to worsen these issues.
+A key assumption underlying this analysis is that EHR exposure is exogenous to individual physicians.
+That is, a hospital makes the decision to implement an EHR independent of a physicians' labor market response to that implementation.
+This assumption is not unrealistic in large hospitals who employ many physicians, but may be violated in small hospitals where a physician has a voice in management.
+To address this, I first limit the analysis to large hospitals.
+However, small hospitals are especially important in studying patient access to care.
+To include small hospitals in the analysis, I instrument EHR adoption with local bandwidth capacity.
+\textcolor{red}{Relate these findings to previous findings.
+I need to make this paragraph more clear as I improve what I'm doing}
+
+This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010.
+\textcolor{red}{This needs to be expounded upon a lot.
+Perhaps one or two paragraphs for each area of contribution.}
 
 The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
-Despite many case studies and hospital-level analysis that indicate large improvements in health outcomes such as mortality (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
+Despite many case studies and hospital-level analysis that indicate s large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
 Further, there have been no significant cost decreases due to EHRs in the short or long run.
 If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}).
-However, the incentive programs put in place could impact the cost effects of EHRs, and 6 years may not be a long enough time frame to see the full picture of cost effects.
-While this paper does not specifically investigate hospital costs, it does speak to potential drivers of cost effects.
-Finally, there have been preliminary case studies and interviews seeking to understand physicians' outlook on EHRs.
-These differ depending on the setting studied, where there is evidence of older physicians having more frustration regarding the technology than younger physicians.
-\textcolor{red}{I need a citation here and probably to expand more on the idea of physician opinions.}
+As discussed earlier, these findings are puzzling due to the expected benefits of EHRs.
+This paper contributes to our knowledge of this area in re-framing past finding into a patient access to care problem instead of an efficiency problem.
 
-This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010, capturing more so the effect of computerization.
+In 2009, the
+
+\newpage
 
@@
 
-\textcolor{red}{Need to figure out what goes in this section. History of EHRs? Specifics of the subsidies and meaningful use?}
+EHRs have been an important feature of healthcare since the 1980s and 1990s.
+Early on, this technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling.
+Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected.
+The cost of utilizing an EHR was extremely large and increasing in the amount of capabilities, which led to the widespread view that they were only complements to paper records and not feasible as a substitute.
+The feasibility of using only EHRs increased as computers became more portable, creating what is known as the "physician workspace", a computer station for the physician to interface directly with EHRs.
+However, physicians kept the view that the technology was purely complementary to paper due to burdensome data entry.
+Automation in data entry was non-existent, making it extremely time consuming for the user.
+Even when automation was developed for certain data, policy required users to double check all of the data in the EHR for errors.
+
+The HITECH Act was passed in 2009, setting aside government funds to subsidies hospitals who used EHRs according to certain guidelines.
+These requirements include having at least 80\% of patients in the system, asking certain questions to patients, and protecting the security of the system.
+Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies.
+Hospitals began receiving payments for the first stage of meaningful use in 2011.
+This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
+\textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, some physicians still continue to report frustration.
+"Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
+This shows the continued hesitancy to utilize EHRs that physicians exhibit.
 
@@
 Finally, physicians are seeing an average of 620 Medicaid patients in hospitals per year.
-\textcolor{red}{In Appendix A1, I include summary statistics by subsample of physician age and year.}
 
+\vspace{5mm}
 \begin{table}[ht]
@@
 \end{table}
+\vspace{5mm}
+
+I also include summary statistics by year and physician experience for comparison in Table \textcolor{red}{number}.
+In all years, senior physicians see more patients on average than younger physicians.
+Note that in 2015 the number of Medicaid patients drops drastically.
+This is due to the data collection in 2015 cutting off earlier than previous years since it was the last year collected for the data set.
+Because of this limitation, I drop 2015 in the productivity analysis, where number of patients is the dependent variable.
+
+\textcolor{red}{I need to think about why senior physicians are more exposed than younger physicians}
+
+\vspace{5mm}
+\begin{table}[ht]
+\caption{Summary Statistics by Year and Experience Level}
+\import{Table Code}{split stats.tex}
+\end{table}
+\vspace{5mm}
 
@@
 
-\begin{figure}
+\vspace{5mm}
+\begin{figure}[ht]
     \caption{Treatment Variables Over Time}
@@
 \end{figure}
+\vspace{5mm}
 
@@
 Employees of this type are assigned a medical tax ID when working in hospitals; they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
-All of these employees are the equivalent of transcribers for health information technology.
+I will refer to any employee in these categories as a data assistant.
 
@@
 The total number of these registered employees is 875.
+The first data assistants ever registered were in 2005, when electronic medical records were being implemented.
+From 2005 to 2013, more 15-60 more employees were registered each year.
+The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occured in 2014.
+If hiring data assistants were directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians were first exposed to EHRs.
+
+\vspace{5mm}
+\begin{figure}[ht]
+\caption{Frequency of Data Assistant Enumeration by Year}
+    \includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+    \label{fig:dataassistant_histogram}
+\end{figure}
+
+The total number of registered data assistants may not equal the true number of people doing a data assistant job in hospitals.
+However, under the assumption that the number of registered data assistants is proportional to the number of actual data assistants, the claim of no correlation still holds.
 
@@
 
-\section{Tables and Figures}
-
-\begin{table}[ht]
-\import{Table Code}{split stats.tex}
-\end{table}
+\section{}
 

### 2022-02-18 Paper/Alternate Intro.tex

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,hyperref,array,xcolor,multicol,verbatim,mathpazo}
\usepackage[normalem]{ulem}
\usepackage[pdftex]{graphicx}
\usepackage{fullpage}
\usepackage{import}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage[font=footnotesize,labelfont=bf]{caption}
\captionsetup{justification=raggedright,singlelinecheck=false}

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\onehalfspacing
\addtolength{\skip\footins}{2pc plus 5pt}

\title{Electronic Health Records and Patient Access to Care: Evidence from Physician Labor Market Decisions}
\author{Hanna Glenn}
%\date{\today}

\DeclareLabeldate[online]{%
  \field{date}
  \field{year}
  \field{eventdate}
  \field{origdate}
  \field{urldate}
}

\begin{document}

\maketitle

Computerized medical records and the systems that encompass them came with expectations of revolutionizing the U.S. healthcare system. Former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on the both quality and cost of care. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of electronic health care records (EHRs) in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}). The movement towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).

A large area of research has been devoted to whether EHRs accomplished this simultaneous cost reduction and quality improvement. Economic research has concluded that neither were realized to the extent that was expected. The median patient received the same quality of care and some hospitals were still seeing cost increases from EHRs even six years after implementation. This raises the question- did any benefit come from the push towards EHR implementation? Why were the main goals not realized? An important facet to the answer may be an aspect not yet considered: patient access to care. This paper seeks to understand the effect of EHR implementation on patient access to care through the channel of physician labor market decisions.

\end{document}

### 2022-02-22 Scripts/analysis.R
update definition of retirement (only =1 in one year max) and update event study graphs with this new definition

@@ -1,6 +1,7 @@
 library(did)
 library(dplyr)
 library(ggplot2)
+library(readr)
 
 # ------------------------------------- ANALYSIS  ------------------------------------
 #                                       Hanna Glenn, Emory University
@@ -15,13 +16,17 @@ Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
 # Retirement ---------------------------------------------------------------------------
 
-# Full sample (exclude 2017 because I cannot observe retirement)
+meanage_retire <- Physician_Data %>% ungroup() %>%
+  dplyr::filter(ever_retire==1) %>%
+  dplyr::summarise_at(c("max_age"), list(mean),na.rm=TRUE)
+
+# Full sample 
 retire_es <- att_gt(yname = "retire",
               gname = "minyr_EHR",
               idname = "DocNPI",
               tname = "year",
-              xformla = ~grad_year,
-              data = filter(Physician_Data, minyr_EHR>0),
+              xformla = ~max_age,
+              data = filter(Physician_Data, minyr_EHR>0 & minyr_retire!=2016),
               est_method = "dr",
               control_group = "notyettreated",
               anticipation=0,
@@ -49,7 +54,7 @@ old_retire_es <- att_gt(yname = "retire",
                     idname = "DocNPI",
                     tname = "year",
                     xformla = ~grad_year,
-                    data = dplyr::filter(Physician_Data,age>50 & minyr_EHR>0),
+                    data = dplyr::filter(Physician_Data,max_age>45 & minyr_EHR>0),
                     est_method = "dr",
                     control_group = "notyettreated",
                     anticipation=0
@@ -105,7 +110,7 @@ old_retire_es_li <- att_gt(yname = "retire",
                         idname = "DocNPI",
                         tname = "year",
                         xformla = ~grad_year,
-                        data = dplyr::filter(Physician_Data,age>50 & minyr_EHR>0),
+                        data = dplyr::filter(Physician_Data,age>50 & minyr_EHR_int>0),
                         est_method = "dr",
                         control_group = "notyettreated",
                         anticipation=0
@@ -126,7 +131,7 @@ ggdid(old_retire_es_dyn_li, xlab="\n Relative Year", theming=FALSE, legend=FALSE
 ggsave(file="ggdid_retire_allEHR_old_li.pdf",path="Objects")
 
 
-# Office- based outcomes -------------------------------------------------------------
[…]

### 2022-02-22 Scripts/data5_MDPPAS.R

@@ -46,7 +46,10 @@ Physician_Data <- Aggregated_Pairs %>%
 Physician_Data <- Physician_Data %>%
   dplyr::mutate(birth_year=substr(birth_dt,start=6, stop=9)) %>%
   dplyr::mutate(birth_year=as.numeric(birth_year)) %>%
-  dplyr::mutate(age=year-birth_year)
+  dplyr::mutate(age=year-birth_year) %>%
+  dplyr::group_by(DocNPI) %>%
+  dplyr::mutate(max_age=max(age)) %>%
+  dplyr::ungroup()
 
 # Drop any physicians with less than 20% of patients in hospitals
 Physician_Data <- Physician_Data %>%
@@ -90,57 +93,37 @@ for (i in 2010:2017){
   
 } 
 
-# I only consider it retirement if all future claims are zero.
-# Create one retirement variable that only consider MDPPAS future claims and one that considers both hosp and MDPPAS future claims
-Physician_Data <- Physician_Data %>% dplyr::ungroup() %>%
-  dplyr::mutate(retire=ifelse(future_claims==0,1,0),
-                retire_both=ifelse(future_claims==0 & future_claims_hosp==0,1,0)) %>%
-  dplyr::mutate(retire_2016=ifelse(year==2016 & retire==1,1,NA),
-                retire_2016_both=ifelse(year==2016 & retire_both==1,1,NA)) %>%
-  dplyr::group_by(DocNPI) %>%
-  tidyr::fill(retire_2016,.direction="downup") %>%
-  tidyr::fill(retire_2016_both,.direction="downup") %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(retire=ifelse(year==2017 & retire_2016==1,1,retire),
-                retire_both=ifelse(year==2017 & retire_2016_both==1,1,retire_both)) %>%
-  dplyr::mutate(retire=ifelse(year==2017 & is.na(retire_2016),0,retire),
-                retire_both=ifelse(year==2017 & is.na(retire_2016_both),0,retire_both))
-
-# This variable creates the retirement indicator one year too early. I want the first year of retirement to be the first year showing zeros.
-# Fix this
+# Create retirement variable
 minyr_retire <- Physician_Data %>%
-  dplyr::filter(retire==1) %>%
[…]

### 2022-02-22 Paper/Paper.tex

@@
 \usepackage{setspace}
-\doublespacing
+\singlespacing
 \addtolength{\skip\footins}{2pc plus 5pt}
@@
 The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making.
-Their interaction has direct impacts across industries; job displacement, creation, and satisfaction are all prevalent reality for many workers (\textcolor{red}{PUT SOME CITATIONS HERE}).
-Additionally, in some cases, labor market changes that take place for one job can have rippling effects to populations of people.
-One such job is that of a physician, where physician decisions have direct and indirect effects on patients receiving care.
-For example, a physician experiencing burnout from work pressures may treat patients differently and lead to patients feeling uncomfortable seeking care (\textcolor{red}{FIND A CITE FOR THIS)}, or increased administrative burdens could lead to less time with patients and exacerbated access to care issues for those in rural settings (\textcolor{red}{FIND CITE FOR THIS}).
-There may also be positive effects on patients from physician labor market changes, such as increased access to care that comes from more physicians entering the market, or less harmful mistakes that arise from standardizing care given.
-For these reasons, I investigate physician labor markets under a major technology change in the U.S.
-healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce labor market changes in the occupation.
-\textcolor{red}{Rethink how I frame the examples, but MAKE SURE I HAVE CITATIONS FOR THESE THINGS}
+Their interaction has direct impacts across industries; job changes such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
+Labor market changes for one job type may also have rippling effects on populations serviced by that industry.
+One such job is that of a physician, where physician choices and labor markets directly affect patients on margins of accessibility and quality of care received.
+For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
+Thus, shocks to physician labor markets or work settings may have drastic implications for patients.
+I investigate physician labor markets under a major technology change in the U.S.
+healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.
 
@@
 since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
-The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency of care.
+The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency.
 A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
@@
 Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized.
-However, with the rapid implementation of EHRs, the day to day life of physicians changed drastically.
-In interviews, groups of physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels.
+However, with the rapid implementation of EHRs, the day to day life of physicians also changed rapidly.
+In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices.
-Further, the short and long run impacts of EHRs on physician productivity could vary greatly.
-I use modern event study approaches, including continuous treatment, to understand the impact of EHR implementation in hospitals on physician labor market outcomes: retirement, where to work, and productivity in hospitals.
+The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor force altogether.
+Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.
+I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and patients seen in the short vs.
+long run.
 
-Using CMS shared patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct physician-level data on hospitalists and other general practice physicians connected to hospitals from 2009-2017.
-The data captures the extent to which the physician in exposed to EHRs in hospitals and various labor market outcomes, along with demographic information such as zip code, age, and gender.
-With this data I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician.
+Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel on hospitalists and other general practice physicians connected to hospitals from 2009-2017.
+The data captures the extent to which the physician in exposed to EHRs in hospitals, various labor market outcomes, and demographic information such as zip code, age, and gender.
+I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician.
 Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals.
-For retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it.
-Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here (if no effect, take out the first part of this sentence)}.
+I consider exposure on two dimensions: working in any hospital that implements an EHR, and working in a low-integrated hospital that implements an EHR.
+To study retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it.
+Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here.
+(The funny thing is, I find that young physicians "retire"...
+I need to figure out what this means and possibly rephrase the last couple sentences)}.
 
-Given that there is a population of both younger and older physicians who stay in the labor force (\textcolor{red}{it may be that all of them stay in the market}), the next outcome to consider is whether physicians switch work settings as a result of EHR implementation in hospitals.
+For those who stay in the labor force, there may be margin to change physicil place of work if physicians are unhappy with the specific loss of autonomy experienced at a given hospital.
 That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different?
-Here, I limit the sample to any physicians who stay in the labor force after EHR exposure.
+To answer this question, I limit the sample to any physicians who stay in the labor force after EHR exposure and use the same event study framework to estimate the effect on whether the physician switches zip codes or increases office-based patients.
 I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.
 
-These findings imply that there are still physicians who remain employed by the same hospital after EHR implementation.
-Thus, I study the effect of the technology on productivity within hospitals, where productivity is measured by the number of patients seen.
+Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen.
 The population of physicians left in this analysis are those who do not retire or switch to an office based setting.
-I find that productivity does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains.
+I find that the number of patients seen does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains.
 After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs.
@@
 \textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}
-
-\textcolor{red}{Add a paragraph discussing sensitivity analysis or mechanisms.}
 
@@
 
+\textcolor{red}{Add a paragraph discussing sensitivity analysis or mechanisms.}
+
 This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010.
@@
 The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
-Despite many case studies and hospital-level analysis that indicate s large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
+Despite many case studies and hospital-level analysis that indicate a large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
 Further, there have been no significant cost decreases due to EHRs in the short or long run.
@@
 
-In 2009, the
+\textcolor{red}{Expand both contribution and literature review and combine them to flow better.}
 
@@
 EHRs have been an important feature of healthcare since the 1980s and 1990s.
-Early on, this technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling.
+Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling.
 Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected.
-The cost of utilizing an EHR was extremely large and increasing in the amount of capabilities, which led to the widespread view that they were only complements to paper records and not feasible as a substitute.
-The feasibility of using only EHRs increased as computers became more portable, creating what is known as the "physician workspace", a computer station for the physician to interface directly with EHRs.
-However, physicians kept the view that the technology was purely complementary to paper due to burdensome data entry.
+The cost of utilizing an EHR was extremely large, and cost increased as the amount of capabilities increased, which led to the widespread view that EHRs were strictly complements to paper records and not a feasible substitute.
+The possibility of using solely EHRs increased as computers became more portable, creating what is known as the "physician workspace": a computer station for a physician to interface directly with EHRs to record patient updates.
+However, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
 Automation in data entry was non-existent, making it extremely time consuming for the user.
-Even when automation was developed for certain data, policy required users to double check all of the data in the EHR for errors.
+Even when automation was developed for particular machines that performed monitoring, the hospital still held responsible for the accuracy of information and therefore required physicians to double check each data entry.
 
-The HITECH Act was passed in 2009, setting aside government funds to subsidies hospitals who used EHRs according to certain guidelines.
-These requirements include having at least 80\% of patients in the system, asking certain questions to patients, and protecting the security of the system.
+The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines.
+These requirements include having at least 80\% of patients in the system, recording answers to specific questions, and protecting the security of the system to ensure privacy.
 Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies.
@@
 This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
-\textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, some physicians still continue to report frustration.
+\textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, physicians still continue to report frustration.
 "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
-This shows the continued hesitancy to utilize EHRs that physicians exhibit.
 
@@
 
-Using various linked datasets, I construct a physician-level panel spanning from 2009-2015 which measures physicians' exposure to EHRs over time and other relevant characteristics.
+Using various linked datasets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time and other relevant characteristics.
 The different data sets used to construct this panel are described below.
 
-The main data I utilize is the CMS Shared Patients Data.
-For a pair of National Provider Identifiers (NPIs), it records the number of patients who bill both of the entities under Medicaid in the same day.
+\subsection{MD-PPAS}
+
+Medicare Data on Provider Practice and Specialty is a database of physicians designed to record information about claim counts and specialty from 2009-2017.
+The relevant information included is physician NPI, primary and secondary specialties, patient counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
+First, I make limitations to the sample of physicians based on specialty type.
+Primary care physicians are the most likely to be exposed to the administrative burden of EHRs in hospitals.
+Thus, I only include physicians whose primary specialty is hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed for all other years).
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
+The dependent variables are constructed using MD-PPAS claim and patient count information.
+Retirement is an indicator variable that is equal to 1 in the first year that all future claims in all zip codes are equal to 0.
+I do not observe retirement in 2017, and drop those who "retire" in 2016 since one year of zero claims does not necessarily mean retirement.
+
+\subsection{Shared Patient}
+
+CMS Shared Patient data records the number of patients who bill two entities under Medicaid in the same day.
 For example, if a Medicaid patient is referred to a specialist by a primary care physician, then those two physicians have a shared patient in common.
-I limit the entities to only include the number of shared patients for physician-hospital pairs.
+I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs.
 I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
@@
 
+\newpage
+
 \section{Retirement}
+
+It is not uncommon in developed countries for workers to plan for retirement well in advance.
+Many workers have a retirement age set and spend their working life saving to get to that point.
+An exogenous shock typically will not change this retirement age because of the large sum of money necessary to formally leave the labor force.
+However, physicians typically make this decision differently than the majority of workers.
+Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
+The reason being that the type of person who selects into being a physician will, on average, have a high regard for helping people.
+When it comes time to retire, they hesitate to abandon patients they have seen over the course of their career.
+This decision to delay retirement is not financial.
+Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
+
+This implies that an exogenous shock to work setting could, if it outweighs the concern for patients, cause an earlier exit.
+Medical articles suggest that the implementation of EHRs did lead to retirement in older physicians.
+\textcolor{red}{cite articles}.
+
+\subsection{Event Study and Results}
+
+Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in a hospital.
+The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR.
+On the right panel, only physicians over the age of 50 are included.
+These results are puzzling, as they show that young physicians are the ones "retiring", or leaving the labor force.
+\textcolor{red}{However, this may just reflect that in later years taking a one or two year break looks like retiring whereas in early years it doesn't.
+What would cause a physician to drop out of the data for one or two years?
+Add to the appendix the same analysis but dropping physicians who "retired" after 2014.
+}
+
+\begin{figure}[ht]
+\caption{}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR.pdf}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old.pdf}
+        \end{minipage}
+        \label{fig:retirefirst}
+\end{figure}
+
+I also repeat this analysis with a more strict version of exposure to EHRs.
+A physician is only considered exposed if they worked in a low-integration status hospital that implemented an EHR.
+These results are shown in Figure \ref{fig:retiresecond}.
+
+\begin{figure}[ht]
+\caption{}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_li.pdf}
+        \end{minipage}
+        \hspace{0.2cm}
+        \begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old_li.pdf}
+        \end{minipage}
+        \label{fig:retiresecond}
+\end{figure}
 
@@
 \caption{Frequency of Data Assistant Enumeration by Year}
-    \includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+
     \label{fig:dataassistant_histogram}

### 2022-02-22 Scripts/analysis.R
make grad year more strict to accomodate residency years and then update graphs

@@ -17,7 +17,7 @@ Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 # Retirement ---------------------------------------------------------------------------
 
 meanage_retire <- Physician_Data %>% ungroup() %>%
-  dplyr::filter(ever_retire==1) %>%
+  dplyr::filter(ever_retire==0) %>%
   dplyr::summarise_at(c("max_age"), list(mean),na.rm=TRUE)
 
 # Full sample 
@@ -66,7 +66,7 @@ p<-old_retire_es$Wpval
 old_retire_es_dyn <- aggte(old_retire_es, type = "dynamic")
 
 # Create a plot
-ggdid(old_retire_es_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure \n In Physicians > 50") + 
+ggdid(old_retire_es_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure \n In Physicians > 45") + 
   labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
   theme(plot.caption = element_text(hjust = 0, face= "italic")) + theme_bw() +
   scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"))
@@ -110,7 +110,7 @@ old_retire_es_li <- att_gt(yname = "retire",
                         idname = "DocNPI",
                         tname = "year",
                         xformla = ~grad_year,
-                        data = dplyr::filter(Physician_Data,age>50 & minyr_EHR_int>0),
+                        data = dplyr::filter(Physician_Data,max_age>45 & minyr_EHR_int>0),
                         est_method = "dr",
                         control_group = "notyettreated",
                         anticipation=0

### 2022-02-22 Scripts/data4_physician_level.R

@@ -32,11 +32,11 @@ PhysCompare <- PhysCompare %>%
   dplyr::rename(gender=gndr,grad_year=Grd_yr) %>%
   select(NPI,grad_year) 
 
-# Merge, then drop any physicians that graduated after 2008
+# Merge, then drop any physicians that graduated after 2005
 data <- data %>%
   left_join(PhysCompare, by=c("DocNPI"="NPI")) %>%
   distinct() %>%
-  filter(grad_year<2009)
+  filter(grad_year<2005)
   # Now have 1 mill obs
 
 
@@ -391,7 +391,7 @@ Aggregated_Pairs <- data %>%
            num_hospitals,
            frac_EHR, frac_EHR_dec, frac_EHR_doc, minyr_EHR, minyr_EHR_dec, minyr_EHR_int, minyr_EHR_dec_int,
            num_systems, hosp_patient_count)
-  #588k obs
+  #528k obs
 
 # Now complete the data
 Aggregated_Pairs <- complete(Aggregated_Pairs,DocNPI,year=2009:2017)

### 2022-02-22 Scripts/data5_MDPPAS.R

@@ -56,7 +56,7 @@ Physician_Data <- Physician_Data %>%
   dplyr::group_by(DocNPI) %>%
   dplyr::mutate(max=max(pos_inpat,na.rm=T)) %>%
   dplyr::filter(max>.2)
-  # 484k obs
+  # 403k obs
 
 # CREATE DEPENDENT VARIABLES -----------------------------------------------------------------------------------------
 

### 2022-02-22 Paper/Paper.tex
Updates from Overleaf

@@
 I do not observe retirement in 2017, and drop those who "retire" in 2016 since one year of zero claims does not necessarily mean retirement.
+The average age of the sample is 51.5, indicating a large number of older physicians, where the average age of those who retire at some point in the sample is 52.5.
+This indicates a large older population in the sample.
+\textcolor{red}{maybe show the distribution of ages by those who retire and those who don't.}
+
+\textcolor{red}{paragraph about switching setting variables}
+
+\textcolor{red}{paragraph about patient count variable}
 
@@
 CMS Shared Patient data records the number of patients who bill two entities under Medicaid in the same day.
-For example, if a Medicaid patient is referred to a specialist by a primary care physician, then those two physicians have a shared patient in common.
+For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen the same day, then those two physicians have a shared patient in common.
+I utilize this data to find physicians' EHR exposure based on the hospitals they work with, which defines treatment variables for the analysis.
 I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs.
 I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
-To achieve this, I limit the qualifications for a sufficient physician-hospital pair.
-First, I only consider doctors who have a primary care role, excluding any specialists.
-The reason for this is that a primary care physician who does rounds in a hospital is likely to be primarily employed by that hospital (not brought in from a practice).
-Another reason for this is that primary care doctors interact directly with EHRs during checkups, unlike a surgeon.
-Second, I limit pairs to those which have a substantial number of patients billed together in the same day.
-This is to avoid including the potential physician who works in an office but happens to send a small number of patients to a nearby hospital.
-I consider a substantial number of patients to be at least 30 patients bill both the physician and the hospital in the same day in at least one year, \textcolor{red}{where the exact number chosen is explored in the appendix}.
-The physicians remaining in the data are likely doing a number of rounds physically inside hospitals, since it is unlikely that many patients will see a primary care doctor in a private office and then immediately go to a hospital in the same day.
+The types of physicians are already limited to primary care physicians, but there still may be some office-based primary care physicians in the shared patient data.
+Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
+A physician is dropped if their total number of same day billings with a hospital is less than 30 times the number of years the pair appears in the data.
+The reason for this is to avoid dropping physicians who drop out of the data for many years, as this may indicate retirement.
+\textcolor{red}{I explore other thresholds in the appendix.}
 
 Using hospital NPI, I link the physician-hospital pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
-I then aggregate to the physician level, where physician EHR use is measured as the percentage of their hospitals which use an EHR in a given year.
-Additional physician-level characteristics, such as years of experience, come from Physician Compare.
+Then I record the first year a physician is exposed to an EHR based on the hospitals they share patients with.
+I create two measures of EHR exposure.
+The first is exposure to any EHR in any hospital, and the second is exposure to EHR at a "low-integration" hospital (a hospital where physicians have a low measure of vertical integration, as defined in \textcolor{red}{add cite}).
+I then aggregate to the physician level.
+
+\textcolor{red}{write about the discrepancy in years}
+
+\subsection{Summary Statistics}
 

### 2022-02-23 Scripts/data5_MDPPAS.R
update summary statistics after incorporating MDPPAS data

@@ -38,6 +38,7 @@ Physician_Data <- Aggregated_Pairs %>%
   dplyr::group_by(DocNPI) %>%
   tidyr::fill(sex, .direction="downup") %>%
   tidyr::fill(birth_dt, .direction="downup") %>%
+  tidyr::fill(frac_EHR,.direction = "downup") %>%
   dplyr::ungroup() %>%
   dplyr::mutate(female=ifelse(sex=='F',1,0)) 
 
@@ -68,15 +69,6 @@ Physician_Data <- Physician_Data %>%
   dplyr::mutate(claim_count_total=ifelse(is.na(claim_count1),NA,sum(dplyr::c_across(tidyr::starts_with("claim_count")),na.rm=T))) %>%
   dplyr::ungroup() 
 
-# Remove those who only appear in the data for one year (maybe they are a visiting doctor?)
-# Maybe I shouldn't do this?
-Physician_Data <- Physician_Data %>%
-  dplyr::mutate(pos=ifelse(is.na(claim_count_total),0,1)) %>%
-  dplyr::group_by(DocNPI) %>%
-  dplyr::mutate(sum=sum(pos)) %>%
-  dplyr::ungroup() %>%
-  dplyr::filter(sum>1) %>%
-  dplyr::select(-pos,-sum)
 
 # Create a variable that sums up the claim count in all future years
 Physician_Data <- Physician_Data %>%
@@ -225,29 +217,44 @@ Physician_Data <- Physician_Data %>%
 
 Physician_Data <- Physician_Data %>%
   dplyr::rowwise() %>%
-  dplyr::mutate(change_zip=ifelse(year==2010 & !(str_detect(zip_list2009,zip_list2010) | str_detect(zip_list2010,zip_list2009)) ,1,NA)) %>%
-  dplyr::mutate(change_zip=ifelse(year==2010 & is.na(change_zip),0,change_zip)) %>%
-  dplyr::mutate(change_zip=ifelse(year==2010 & change_zip==1 & sort(unlist(strsplit(zip_list2009, "")))==sort(unlist(strsplit(zip_list2010, ""))),0,change_zip))
+  dplyr::mutate(change_zip=ifelse(year==2010 & !(str_detect(zip_list2009,zip_list2010)[1] |
+                                                   str_detect(zip_list2010,zip_list2009)[1]) ,1,NA)) %>%
+  dplyr::mutate(change_zip=ifelse(year==2010 & is.na(change_zip) & !(is.na(zip_list2009) | 
[…]

### 2022-02-23 Scripts/sumstats.R

@@ -26,43 +26,47 @@ Physician_Data <- read_rds(paste0(created_data_path,"Physician_data.rds"))
 
 # Physician Level
 sum_stats_fullsample <- Physician_Data %>% ungroup() %>%
-  summarise_at(c("Number of Hospitals Worked With"="num_hosp_constant",
+  summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
-                 "Medical School Grad. Year"="grad_year",
-                 "Number of Patients"="hosp_count","Fraction of Hospitals with EHR"="frac_EHR",
-                 "Average Size of Hospitals Worked With (Beds)"="avg_beds", 
-                 "Average Hospital Operating Days"="avg_oper_days",
-                 "Exposure to an EHR"="anyEHR_exposed"), 
+                 "Age"="age",
+                 "Number of Patients"="claim_count_total","Fraction of Hospitals with EHR"="frac_EHR",
+                 "Exposure to an EHR"="anyEHR_exposed",
+                 "Exposure to an EHR (Low Integration)"="anyEHR_LI_exposed",
+                 "Fraction Patients in Office"="pos_office",
+                 "Ever Retire"="ever_retire",
+                 "Work in an Office"="work_in_office",
+                 "Change Zip Codes"="change_zip"), 
                list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))), na.rm=TRUE) %>%
   mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
   gather(key=var,value=value) %>%
   extract(col="var",into=c("variable", "statistic"), regex=("(.*)_(.*)$")) %>%
   spread(key=statistic, value=value) %>%
-  relocate(variable,n,m,sd,min,max) %>%
-  mutate(m=ifelse(variable=='Medical School Grad. Year',round(m,digits=0),m))
+  relocate(variable,n,m,sd,min,max) 
 
 
 
 
-knitr::kable(sum_stats_fullsample[c(1,2,4,6,7,9,3,5,8),],
+knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
              format="latex",
-             table.envir="figure",
+             table.envir="table",
              col.names=c("Variable","N","Mean","Std. Dev.", "Min", "Max"),
              digits=2,
[…]

### 2022-02-23 Table Code/split stats.tex
Updates from Overleaf

@@
-\scriptsize
-\begin{tabular}[t]{lcccccccccccc}
+\begin{table}[h]
+
+\caption{Means by Age Sample}
+\centering
+\begin{tabular}[t]{lcc}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Senior Physicians} & \multicolumn{6}{c}{Young Physicians} \\
-\cmidrule(l{3pt}r{3pt}){2-7} \cmidrule(l{3pt}r{3pt}){8-13}
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{\# of Patients} & \multicolumn{2}{c}{Frac.
-Hosp.
-w/ EHR} & \multicolumn{2}{c}{Exposure EHR} & \multicolumn{2}{c}{\# of Patients} & \multicolumn{2}{c}{Frac.
-Hosp.
-w/ EHR} & \multicolumn{2}{c}{Exposure EHR} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9} \cmidrule(l{3pt}r{3pt}){10-11} \cmidrule(l{3pt}r{3pt}){12-13}
-Year & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev. & Mean & Std. Dev.\\
+Variable & Age $>$ 45 & Age $<=$ 45\\
 \midrule
-2009 & 928 & 1304. & 0.27 & 0.39 & 0.38 & 0.49 & 611 & 976 & 0.30 & 0.40 & 0.42 & 0.49\\
-2010 & 734 & 1172 & 0.32 & 0.41 & 0.47 & 0.50 & 512 & 912 & 0.34 & 0.42 & 0.49 & 0.50\\
-2011 & 910 & 1268 & 0.42 & 0.44 & 0.60 & 0.49 & 666 & 1008 & 0.44 & 0.44 & 0.62 & 0.48\\
-2012 & 888 & 1317 & 0.54 & 0.44 & 0.71 & 0.45 & 684 & 1047 & 0.57 & 0.43 & 0.74 & 0.44\\
-2013 & 829 & 1298 & 0.64 & 0.42 & 0.81 & 0.39 & 667 & 1045 & 0.67 & 0.41 & 0.83 & 0.38\\
-\addlinespace
-2014 & 836 & 1272 & 0.73 & 0.39 & 0.87 & 0.34 & 705 & 1062 & 0.75 & 0.37 & 0.89 & 0.32\\
-2015 & 482 & 752 & 0.81 & 0.34 & 0.92 & 0.28 & 418 & 636 & 0.83 & 0.33 & 0.93 & 0.26\\
+\addlinespace[0.3em]
+\multicolumn{3}{l}{\textbf{Outcomes}}\\
+\hspace{1em}Number of Hospitals Worked With & 1.54 & 1.38\\
+\hspace{1em}Female & 0.27 & 0.41\\
+\hspace{1em}Number of Systems Worked With & 1.33 & 1.23\\
+\hspace{1em}Age & 51.83 & 37.79\\
+\hspace{1em}Number of Patients & 2578.12 & 1359.17\\
+\addlinespace[0.3em]
+\multicolumn{3}{l}{\textbf{Treatment}}\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.51 & 0.55\\
+\hspace{1em}Exposure to an EHR & 0.64 & 0.62\\
+\hspace{1em}Exposure to an EHR (Low Integration) & 0.44 & 0.39\\
+\addlinespace[0.3em]
+\multicolumn{3}{l}{\textbf{Characteristics}}\\
+\hspace{1em}Fraction Patients in Office & 0.36 & 0.19\\
+\hspace{1em}Ever Retire & 0.02 & 0.02\\
+\hspace{1em}Work in an Office & 0.66 & 0.37\\
+\hspace{1em}Change Zip Codes & 0.07 & 0.11\\
 \bottomrule
 \end{tabular}
+\end{table}

### 2022-02-23 Paper/Paper.tex

@@
 \usepackage{booktabs}
-\usepackage[font=footnotesize,labelfont=bf]{caption}
-\captionsetup{justification=raggedright,singlelinecheck=false}
+%\usepackage[font=footnotesize,labelfont=bf]{caption}
+%\captionsetup{justification=raggedright,singlelinecheck=false}
 
@@
 
-Medicare Data on Provider Practice and Specialty is a database of physicians designed to record information about claim counts and specialty from 2009-2017.
+Medicare Data on Provider Practice and Specialty is a database of physicians which details claim counts and specialty from 2009-2017.
 The relevant information included is physician NPI, primary and secondary specialties, patient counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
@@
 Primary care physicians are the most likely to be exposed to the administrative burden of EHRs in hospitals.
-Thus, I only include physicians whose primary specialty is hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed for all other years).
+Thus, I only include physicians whose primary specialty is hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
 Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
-The dependent variables are constructed using MD-PPAS claim and patient count information.
+
+The dependent variables in the analysis have to do with physician labor market outcomes, and are constructed using claim and patient count variables.
 Retirement is an indicator variable that is equal to 1 in the first year that all future claims in all zip codes are equal to 0.
 I do not observe retirement in 2017, and drop those who "retire" in 2016 since one year of zero claims does not necessarily mean retirement.
-The average age of the sample is 51.5, indicating a large number of older physicians, where the average age of those who retire at some point in the sample is 52.5.
-This indicates a large older population in the sample.
 \textcolor{red}{maybe show the distribution of ages by those who retire and those who don't.}
 
-\textcolor{red}{paragraph about switching setting variables}
-
-\textcolor{red}{paragraph about patient count variable}
+For those physicians who do not retire, I analyze whether there is evidence of switching to different places of work.
+Two outcome variables capture moving towards office based settings, one is the fraction of patients seen in an office setting and the other is an indicator for whether the physician works in an office at all.
+While I do not observe different hospitals in the data, I also attempt to capture any type of switching (to office or from hospital to hospital) by measuring if a physician changes zip codes from one year to the next.
+Finally, for physicians who do not exhibit switching settings, I utilize number of patients as an outcome variable.
 
@@
 Approximately 30\% of the sample of physicians are female and the average physician graduated medical school in 1993.
-Note that there are no physicians in the data who graduated after 2009, since they would be entering the labor force during the span of the data.
+Note that there are no physicians in the data who graduated after 2005, since they would be entering the labor force after residency during the span of the data.
 On average, a physician works with between 1 and 2 hospitals as well as hospital systems.
@@
 \vspace{5mm}
-\begin{table}[ht]
 \import{Table Code}{overall stats.tex}
-    \label{fig:summarystatistics}
-    \vspace{1mm}
-    \caption*{\footnotesize Obs: 1,044,001}
-\end{table}
 \vspace{5mm}
@@
 \vspace{5mm}
-\begin{table}[ht]
-\caption{Summary Statistics by Year and Experience Level}
 \import{Table Code}{split stats.tex}
-\end{table}
 \vspace{5mm}
@@
 The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR.
-On the right panel, only physicians over the age of 50 are included.
-These results are puzzling, as they show that young physicians are the ones "retiring", or leaving the labor force.
-\textcolor{red}{However, this may just reflect that in later years taking a one or two year break looks like retiring whereas in early years it doesn't.
-What would cause a physician to drop out of the data for one or two years?
-Add to the appendix the same analysis but dropping physicians who "retired" after 2014.
-}
+On the right panel, only physicians over the age of 45 in 2017 are included.
 
@@
         \label{fig:retirefirst}
-\end{figure}
-
-I also repeat this analysis with a more strict version of exposure to EHRs.
-A physician is only considered exposed if they worked in a low-integration status hospital that implemented an EHR.
-These results are shown in Figure \ref{fig:retiresecond}.
-
-\begin{figure}[ht]
-\caption{}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_li.pdf}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old_li.pdf}
-        \end{minipage}
-        \label{fig:retiresecond}
 \end{figure}

### 2022-02-23 Scripts/sumstats.R
update size of graph

@@ -60,65 +60,77 @@ knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
   pack_rows(index = c("Outcomes" = 5, "Treatment" = 3, "Characteristics" = 4))
 
 
-# Summary Stats broken down by age of physician and year ---------------------------------------------------------
-means_old <- Physician_Data %>% 
-  filter(max_age>45) %>% 
-  group_by(year) %>%
-  summarize_at(c("hosp_count",
-                 "frac_EHR",
-                 "anyEHR_exposed"), list(mean,sd), na.rm=TRUE) 
-  filter(year >= 1986) %>%
-  select(year,hosp_count_fn1,hosp_count_fn2,frac_EHR_fn1,frac_EHR_fn2,anyEHR_exposed_fn1,anyEHR_exposed_fn2) %>%
-  mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.)) 
-
-means_young <- Physician_Data %>% filter(experience<=35) %>% ungroup() %>%
-  group_by(year) %>%
-  summarize_at(c("hosp_count",
-                 "frac_EHR",
-                 "anyEHR_exposed"), list(mean,sd), na.rm=TRUE) %>%
-  filter(year >= 1986) %>%
-  select(year,hosp_count_fn1,hosp_count_fn2,frac_EHR_fn1,frac_EHR_fn2,anyEHR_exposed_fn1,anyEHR_exposed_fn2) %>%
+# Summary Stats of all variables by odl vs. young ---------------------------------------------------------
+means_old <- Physician_Data %>% ungroup() %>%
+  filter(max_age>45) %>%
+  summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
+                 "Female"="female", "Number of Systems Worked With"="num_systems",
+                 "Age"="age",
+                 "Number of Patients"="claim_count_total","Fraction of Hospitals with EHR"="frac_EHR",
+                 "Exposure to an EHR"="anyEHR_exposed",
+                 "Exposure to an EHR (Low Integration)"="anyEHR_LI_exposed",
+                 "Fraction Patients in Office"="pos_office",
+                 "Ever Retire"="ever_retire",
+                 "Work in an Office"="work_in_office",
+                 "Change Zip Codes"="change_zip"), list(mean), na.rm=TRUE) %>%
[…]

### 2022-02-23 Table Code/overall stats.tex

@@
-\footnotesize
-\caption{\footnotesize Summary Statistics}
-\vspace{-2mm}
-\begin{tabular}[t]{lcccc}
+\begin{table}[h]
+\caption{Summary Statistics}
+\centering
+\begin{tabular}[t]{lrrrrr}
 \toprule
-Variable & Mean & Std. Dev. & Min & Max\\
+Variable & N & Mean & Std. Dev. & Min & Max\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Static}}\\
-\hspace{1em}Average Hospital Operating Days  & 364.89 & 5.11 & 41 & 366\\
-\hspace{1em}Average Size of Hospitals Worked With (Beds)  & 373.12 & 252.40 & 10 & 2574\\
-\hspace{1em}Female  & 0.31 & 0.46 & 0 & 1\\
-\hspace{1em}Medical School Grad. Year  & 1993 & 10.04 & 1949 & 2008\\
-\hspace{1em}Number of Hospitals Worked With  & 1.87 & 1.30 & 1 & 40\\
-\hspace{1em}Number of Systems Worked With  & 1.48 & 0.77 & 1 & 13\\
+\multicolumn{6}{l}{\textbf{Outcomes}}\\
+\hspace{1em}Ever Retire & 403020 & 0.02 & 0.13 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 403020 & 0.31 & 0.36 & 0 & 1\\
+\hspace{1em}Work in an Office & 403020 & 0.57 & 0.49 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 372357 & 0.08 & 0.28 & 0 & 1\\
+\hspace{1em}Number of Patients & 403020 & 2220.46 & 3207.08 & 0 & 199765\\
 \addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Dynamic}}\\
-\hspace{1em}Exposure to an EHR  & 0.70 & 0.46 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR  & 0.56 & 0.44 & 0 & 1\\
-\hspace{1em}Number of Patients & 620.54 & 988.74 & 0 & 74639\\
+\multicolumn{6}{l}{\textbf{Treatment}}\\
+\hspace{1em}Exposure to an EHR & 403020 & 0.64 & 0.48 & 0 & 1\\
+\hspace{1em}Exposure to an EHR (Low Integration) & 403020 & 0.42 & 0.49 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 403020 & 0.52 & 0.46 & 0 & 1\\
+\addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Characteristics}}\\
+\hspace{1em}Age & 403002 & 47.71 & 9.38 & 23 & 89\\
+\hspace{1em}Female & 403020 & 0.31 & 0.46 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 403020 & 1.49 & 0.83 & 1 & 20\\
+\hspace{1em}Number of Systems Worked With & 403020 & 1.30 & 0.59 & 1 & 8\\
 \bottomrule
 \end{tabular}
+\end{table}

### 2022-02-23 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{booktabs}
-%\usepackage[font=footnotesize,labelfont=bf]{caption}
+\usepackage[font=footnotesize,labelfont=bf]{caption}
 %\captionsetup{justification=raggedright,singlelinecheck=false}
@@
 
-CMS Shared Patient data records the number of patients who bill two entities under Medicaid in the same day.
+CMS Shared Patient data is from 2009 to 2015, and records the number of patients who bill two entities under Medicaid in the same day.
 For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen the same day, then those two physicians have a shared patient in common.
-I utilize this data to find physicians' EHR exposure based on the hospitals they work with, which defines treatment variables for the analysis.
-I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs.
-I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
+I utilize this data to find physicians' EHR exposure based on the hospitals they share sufficient patients with, which defines treatment variables for the analysis.
+I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
+I care specifically about primary care physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
 The types of physicians are already limited to primary care physicians, but there still may be some office-based primary care physicians in the shared patient data.
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
-A physician is dropped if their total number of same day billings with a hospital is less than 30 times the number of years the pair appears in the data.
+A physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data.
 The reason for this is to avoid dropping physicians who drop out of the data for many years, as this may indicate retirement.
@@
 
-Using hospital NPI, I link the physician-hospital pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
-Then I record the first year a physician is exposed to an EHR based on the hospitals they share patients with.
+Using hospital NPI in each pair, I link the pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
+I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
 I create two measures of EHR exposure.
@@
 I then aggregate to the physician level.
+Since this data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017.
 
-\textcolor{red}{write about the discrepancy in years}
+Finally, I include a physician's graduation year from Physician Compare.
+I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior.
 
@@
 
-Overall summary statistics of the variables are shown in \hyperref[fig:summarystatistics]{Table 1}.
-Average hospital operating days is a physician-level variable that averages the number of operating days for that physicians hospital network.
-While there are some low outliers with a low value, on average physicians are working in hospitals open year round.
-Similarly, the average size of the hospital worked with averages the number of beds of each hospital in a physician's network.
-While these two average variables can technically vary by year, they tend to stay constant across the data, which is why they are included in the static category.
-Approximately 30\% of the sample of physicians are female and the average physician graduated medical school in 1993.
-Note that there are no physicians in the data who graduated after 2005, since they would be entering the labor force after residency during the span of the data.
-On average, a physician works with between 1 and 2 hospitals as well as hospital systems.
-While the number of hospitals that a physician cares for patients in varies by year, the number of hospitals in their network is constant.
-The physician level variables that vary over time deal are the treatment variables regarding EHR exposure and the number of Medicaid patients seen.
-For all physician-years, physicians were exposed to an EHR in 56\% of the hospitals they work in and 70\% of the sample is exposed to any EHR.
-Once a physician becomes exposed to an EHR, they cannot go back to being unexposed.
-Finally, physicians are seeing an average of 620 Medicaid patients in hospitals per year.
+Overall summary statistics of the variables are shown in \hyperref[tab:sumstats1]{Table \ref{tab:sumstats1}}. \textcolor{red}{add commentary on Table 1}
 
@@
 
-I also include summary statistics by year and physician experience for comparison in Table \textcolor{red}{number}.
-In all years, senior physicians see more patients on average than younger physicians.
-Note that in 2015 the number of Medicaid patients drops drastically.
-This is due to the data collection in 2015 cutting off earlier than previous years since it was the last year collected for the data set.
-Because of this limitation, I drop 2015 in the productivity analysis, where number of patients is the dependent variable.
-
-\textcolor{red}{I need to think about why senior physicians are more exposed than younger physicians}
+I include means of these variables in physicians either younger than or older than 45 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}.
+Senior physicians see more patients and work in more hospitals on average than younger physicians.
+Physicians over 45 are slightly more likely to be exposed to EHRs in both general hospitals and low integration hospitals, but a smaller fraction of their hospitals utilize EHRs.
+These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general.
+Finally, physicians in the higher age bracket are less likely to switch zip codes.
 
@@
 
-For a clear look at the variation in treatment variables over time, \hyperref[fig:figure1]{Figure 1} shows the average of these variables over time.
-In 2009, 42\% of the sample of physicians were exposed to an EHR.
-That is, over half of the sample had no affiliation with EHRs at the beginning of the sample period.
-By 2015, 92\% of physicians had exposure to an EHR in at least one hospital in their network.
-Similarly, the fraction of hospitals in a physician's network which were using an EHR was 30\% in 2009 and grew to 82\% in 2015.
-These statistics are consistent with the surge in incentives to implement EHRs in hospitals.
+To examine variation in treatment variables, \hyperref[fig:figure1]{Figure \ref{fig:figure1}} shows the averages over time.
+In 2009, 26\% of the sample of physicians were exposed to an EHR.
+That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
+By 2015, 84\% of physicians had exposure to an EHR in at least one hospital in their network.
+Exposure to EHRs only in low integrated hospitals follows a similar trend with lower percentages in each year.
 
@@
 \begin{figure}[ht]
+\centering
     \caption{Treatment Variables Over Time}
-    \vspace{-2mm}
     \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
@@
 \begin{figure}[ht]
-\caption{}
+\caption{Event Study Graphs}
+\vspace{2mm}
         \begin{minipage}[b]{0.47\linewidth}
@@
 
+\newpage
+
 \section{Work Setting}
+
+\newpage
 

### 2022-02-24 Scripts/data5_MDPPAS.R
update graphs with new age

@@ -62,6 +62,9 @@ Physician_Data <- Physician_Data %>%
 # CREATE DEPENDENT VARIABLES -----------------------------------------------------------------------------------------
 
 # RETIREMENT ####
+# Now I have two ways to define retirement. One using all claim counts and the other
+# using unique beneficiaries. Theoretically, they should result in the same retirement variable.
+
 
 # For retirement, I can combine all claim counts into one
 Physician_Data <- Physician_Data %>%
@@ -73,16 +76,12 @@ Physician_Data <- Physician_Data %>%
 # Create a variable that sums up the claim count in all future years
 Physician_Data <- Physician_Data %>%
   dplyr::group_by(DocNPI) %>%
-  dplyr::mutate(future_claims=ifelse(year==2009,sum(claim_count_total[year>2009],na.rm=T),NA)) %>%
-  dplyr::mutate(future_claims_hosp=ifelse(year==2009,sum(hosp_patient_count[year>2009],na.rm=T),NA))
-
+  dplyr::mutate(future_claims=ifelse(year==2009,sum(claim_count_total[year>2009],na.rm=T),NA))
 
 for (i in 2010:2017){
   Physician_Data <- Physician_Data %>%
     dplyr::group_by(DocNPI) %>%
-    dplyr::mutate(future_claims=ifelse(year==i,sum(claim_count_total[year>i],na.rm=T),future_claims)) %>%
-    dplyr::mutate(future_claims_hosp=ifelse(year==i,sum(hosp_patient_count[year>i],na.rm=T),future_claims_hosp))
-  
+    dplyr::mutate(future_claims=ifelse(year==i,sum(claim_count_total[year>i],na.rm=T),future_claims))
 } 
 
 # Create retirement variable
@@ -93,6 +92,12 @@ minyr_retire <- Physician_Data %>%
   dplyr::distinct(DocNPI, minyr_retire) %>%
   dplyr::ungroup()
 
+
+
+Physician_Data <- Physician_Data %>%
+  dplyr::left_join(minyr_retire, by="DocNPI") %>%
+  dplyr::mutate(minyr_retire=ifelse(!is.na(minyr_retire),minyr_retire+1,minyr_retire)) %>%
+  dplyr::mutate(minyr_retire=ifelse(minyr_retire==2017,NA,minyr_retire))
 
 Physician_Data <- Physician_Data %>%
   dplyr::left_join(minyr_retire, by="DocNPI") %>%
[…]

### 2022-02-24 Paper/Paper.tex

@@
 I create two measures of EHR exposure.
-The first is exposure to any EHR in any hospital, and the second is exposure to EHR at a "low-integration" hospital (a hospital where physicians have a low measure of vertical integration, as defined in \textcolor{red}{add cite}).
+The first is exposure to any EHR in any hospital, and the second is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})).
 I then aggregate to the physician level.
@@
 Finally, physicians in the higher age bracket are less likely to switch zip codes.
+\textcolor{red}{why is there equal retirement between young and old?}
 
@@
 
+\section{Empirical Strategy}
+
+In this setting, treatment effects likely vary across physicians who are exposed in different years since the underlying hospitals who implement likely vary.
+Therefore, to avoid negative weighting issues that would arise in a two way fixed effects specification, I utilize \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) to estimate average group time treatment effects.
+
+\textcolor{red}{outline assumptions necessary for identification and talk about doubly robust estimator}
+
 \section{Retirement}
@@
 Many workers have a retirement age set and spend their working life saving to get to that point.
-An exogenous shock typically will not change this retirement age because of the large sum of money necessary to formally leave the labor force.
-However, physicians typically make this decision differently than the majority of workers.
+An exogenous shock to the worker is not expected to change retirement age because of the large sum of money necessary to formally leave the labor force.
+However, physicians typically make this decision differently than most other workers.
 Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
@@
 
-This implies that an exogenous shock to work setting could, if it outweighs the concern for patients, cause an earlier exit.
-Medical articles suggest that the implementation of EHRs did lead to retirement in older physicians.
+An exogenous shock to work setting could, if it outweighs the reasons for staying, cause an earlier exit than would have occured without the shock.
+Medical articles suggest through interaction with physicians that the implementation of EHRs did lead to retirement in older physicians.
 \textcolor{red}{cite articles}.
+The aim of this section is to test that hypothesis and explore its implications.
 
-\subsection{Event Study and Results}
+\subsection{Event Study Results}
 
-Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in a hospital.
+Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital.
 The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR.
@@
 
-\subsection{Event Study and Results}
-
-The analysis in this section focuses on the effects of hospital EHR implementation on physician productivity within hospitals, measured by the total number of Medicaid patients billed in hospitals in a given year.
-The estimating equation is given as follows:
-
-\begin{equation*}
-    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-5}^5 \beta_kz_{i,t-k} + \varepsilon_{it},
-\end{equation*}
-
-where $y_{it}$ is the number of patients, $\alpha_i$ and $\delta_i$ are physician and time fixed effects, $q_{it}$ are hospital characteristics such as number of beds and days open during the year, $z_{i,t-k}$ is an indicator for whether the physician was exposed to an EHR in any hospital in time $t-k$, and $\epsilon_{it}$ is a random error term.
-
-The identifying assumption is that physicians are exposed endogenously to electronic health records.
-That is, hospital management decides and implements EHRs, and this decision is not correlated with physician productivity or labor market decisions.
-This is not unreasonable for large hospitals with ample resources and many physicians.
-However, a small hospital may give experienced physicians some level of voice in the decision to implement new technologies.
-Results under this assumption are shown below, and  \textcolor{red}{the strength of this assumption is considered in the next section.
-}
-
-An event study plot containing estimates and confidence intervals of  $\{\beta_k\}_{k=-5}^5$  is given in in Figure \textcolor{red}{number}.
-The left hand side is the full sample of physicians at any experience level.
-As is standard in event studies, the identification of the coefficients depends on the assumption that there are no pre-trends.
-The p-value of a Wald test for pre-trends in both event studies indicate no observable pre-trends in the data.
-
-For physicians at any experience level, EHR exposure yields no immediate affect on physician productivity, but leads to a clear increase in the number of patients seen beginning 2 years after exposure.
-After 5 years of exposure, physicians are seeing, on average, 93.67 more Medicaid patients per year.
-Relative to the mean in the year before exposure, 796.95, this is an 11.75\% increase in the number of patients seen.
-When the sample of physicians is limited to senior physicians (at least 35 years of experience), this effect disappears.
-In fact, the results show a slight decrease of \textcolor{red}{put the exact number here} in the number of patients seen in the year after exposure, and no effect following that year.
-This graph limits event time from -4 to 4 due to restricted sample size in year 2010 and 2014.
-
-\begin{figure}[ht]
-\caption{}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/prod_eventstudy_fullsample.pdf}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/prod_eventstudy_oldsample.pdf}
-        \end{minipage}
-\end{figure}
+\subsection{Event Study Results}
 

### 2022-02-24 Scripts/analysis.R
update scripts

@@ -41,7 +41,7 @@ retire_es_dyn <- aggte(retire_es, type = "dynamic")
 ggdid(retire_es_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure") + 
   labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
   theme(plot.caption = element_text(hjust = 0, face= "italic")) + theme_bw() +
-  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"))
+  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00")) + ylim(-1,1)
 
 # Save the plot
 ggsave(file="ggdid_retire_allEHR.pdf",path="Objects")
@@ -54,7 +54,7 @@ old_retire_es <- att_gt(yname = "retire",
                     idname = "DocNPI",
                     tname = "year",
                     xformla = ~grad_year,
-                    data = dplyr::filter(Physician_Data,max_age>45 & minyr_EHR>0),
+                    data = dplyr::filter(Physician_Data,max_age>59 & minyr_EHR>0 & minyr_retire!=2016),
                     est_method = "dr",
                     control_group = "notyettreated",
                     anticipation=0
@@ -62,14 +62,16 @@ old_retire_es <- att_gt(yname = "retire",
 # Save p-value to put in footnote
 p<-old_retire_es$Wpval
 
+ggdid(old_retire_es)
+
 # Aggregate the effects
 old_retire_es_dyn <- aggte(old_retire_es, type = "dynamic")
 
 # Create a plot
-ggdid(old_retire_es_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure \n In Physicians > 45") + 
+ggdid(old_retire_es_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure \n In Physicians >= 60") + 
   labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
   theme(plot.caption = element_text(hjust = 0, face= "italic")) + theme_bw() +
-  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"))
[…]

### 2022-02-24 Scripts/sumstats.R

@@ -60,9 +60,10 @@ knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
   pack_rows(index = c("Outcomes" = 5, "Treatment" = 3, "Characteristics" = 4))
 
 
-# Summary Stats of all variables by odl vs. young ---------------------------------------------------------
+# Summary Stats of all variables by old vs. young ---------------------------------------------------------
 means_old <- Physician_Data %>% ungroup() %>%
-  filter(max_age>45) %>%
+  filter(minyr_EHR>0 & (minyr_retire<2015 | is.na(minyr_retire))) %>%
+  filter(max_age>59) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
@@ -78,7 +79,8 @@ means_old <- Physician_Data %>% ungroup() %>%
   dplyr::rename(value_old=value)
 
 means_young <- Physician_Data %>% ungroup() %>%
-  filter(max_age<=45) %>%
+  filter(minyr_EHR>0 & (minyr_retire<2015 | is.na(minyr_retire))) %>%
+  filter(max_age<=59) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",

### 2022-03-18 Scripts/analysis.R
fix age cutoff to 60 instead of 55

@@ -217,7 +217,7 @@ retireold_cs <- att_gt(
   # xformla = NULL                 # No covariates
   xformla = ~grad_year,            # Time-invariant controls
   data = dplyr::filter(
-    Physician_Data,minyr_EHR>0 & max_age>=55),   # Remove never-treated units and young physicians
+    Physician_Data,minyr_EHR>0 & max_age>=60),   # Remove never-treated units and young physicians
   # data = Physician_Data
   est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
   control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
@@ -248,7 +248,7 @@ retireyoung_cs <- att_gt(
   # xformla = NULL                 # No covariates
   xformla = ~grad_year,            # Time-invariant controls
   data = dplyr::filter(
-  Physician_Data,minyr_EHR>0 & max_age<55),   # Remove never-treated units and young physicians
+  Physician_Data,minyr_EHR>0 & max_age<60),   # Remove never-treated units and young physicians
   # data = Physician_Data
   est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
   control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
@@ -310,7 +310,7 @@ office_fracold_cs <- att_gt(yname = "pos_office",
                          idname = "DocNPI",
                          tname = "year",
                          xformla = ~grad_year,
-                         data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & max_age>=55),
+                         data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & max_age>=60),
                          est_method = "dr",
                          control_group = "notyettreated"
 )
@@ -336,7 +336,7 @@ office_fracyoung_cs <- att_gt(yname = "pos_office",
                             idname = "DocNPI",
                             tname = "year",
                             xformla = ~grad_year,
-                            data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & max_age<55),
+                            data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & max_age<60),
                             est_method = "dr",
                             control_group = "notyettreated"
 )
@@ -386,7 +386,7 @@ office_indold_cs <- att_gt(yname = "work_in_office",
                         idname = "DocNPI",
[…]

### 2022-03-18 Table Code/split stats.tex
Updates from Overleaf

@@
-\begin{table}[h]
+\begin{table}[ht]
 
@@
 \centering
-\begin{tabular}[t]{lcc}
+\begin{tabular}[t]{lccc}
 \toprule
-Variable & Age $>$ 45 & Age $<=$ 45\\
+Variable & Age $>$ 60 & Age $<=$ 60 & Any Who Retire\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Number of Hospitals Worked With & 1.54 & 1.38\\
-\hspace{1em}Female & 0.27 & 0.41\\
-\hspace{1em}Number of Systems Worked With & 1.33 & 1.23\\
-\hspace{1em}Age & 51.83 & 37.79\\
-\hspace{1em}Number of Patients & 2578.12 & 1359.17\\
+\multicolumn{4}{l}{\textbf{Outcomes}}\\
+\hspace{1em}Ever Retire & 0.01 & 0.01 & 1.00\\
+\hspace{1em}Fraction Patients in Office & 0.44 & 0.27 & 0.15\\
+\hspace{1em}Work in an Office & 0.79 & 0.52 & 0.33\\
+\hspace{1em}Change Zip Codes & 0.06 & 0.09 & 0.08\\
+\hspace{1em}Number of Patients & 3238.30 & 2028.23 & 554.53\\
 \addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Treatment}}\\
-\hspace{1em}Fraction of Hospitals with EHR & 0.51 & 0.55\\
-\hspace{1em}Exposure to an EHR & 0.64 & 0.62\\
-\hspace{1em}Exposure to an EHR (Low Integration) & 0.44 & 0.39\\
+\multicolumn{4}{l}{\textbf{Treatment}}\\
+\hspace{1em}Exposure to an EHR & 0.77 & 0.76 & 0.77\\
+\hspace{1em}Exposure to an EHR (Low Integration) & 0.54 & 0.50 & 0.44\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.64 & 0.64\\
 \addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Fraction Patients in Office & 0.36 & 0.19\\
-\hspace{1em}Ever Retire & 0.02 & 0.02\\
-\hspace{1em}Work in an Office & 0.66 & 0.37\\
-\hspace{1em}Change Zip Codes & 0.07 & 0.11\\
+\multicolumn{4}{l}{\textbf{Characteristics}}\\
+\hspace{1em}Age & 61.06 & 44.06 & 48.29\\
+\hspace{1em}Female & 0.15 & 0.35 & 0.39\\
+\hspace{1em}Number of Hospitals Worked With & 1.67 & 1.53 & 1.35\\
+\hspace{1em}Number of Systems Worked With & 1.41 & 1.33 & 1.19\\
 \bottomrule

### 2022-03-18 Paper/Paper.tex

@@
 The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making.
-Their interaction has direct impacts across industries; job changes such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
-Labor market changes for one job type may also have rippling effects on populations serviced by that industry.
-One such job is that of a physician, where physician choices and labor markets directly affect patients on margins of accessibility and quality of care received.
-For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
-Thus, shocks to physician labor markets or work settings may have drastic implications for patients.
-I investigate physician labor markets under a major technology change in the U.S.
+Their interaction has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
+Labor market changes for one job type may have rippling effects on populations serviced by that industry.
+One such job is that of a physician, where physician choices and labor markets directly affect patient accessibility and quality of care received.
+For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
+Thus, shocks to physician labor markets or work settings potentially have drastic implications for patients.
+I investigate physician labor market changes as a result of a major technology change in the U.S.
 healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.
 
-Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care.
+Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care.
 EHRs have have become increasingly relevant in the U.S.
@@
 Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
-The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.
+The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.
 
-Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized.
-However, with the rapid implementation of EHRs, the day to day life of physicians also changed rapidly.
+Physicians, as primary users of this technology, play a meaningful role in whether potential benefits of EHRs are realized.
+However, with the rapid implementation of EHRs, the day to day life for physicians changed rapidly.
 In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor force altogether.
+The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether.
 Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.
-I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and patients seen in the short vs.
-long run.
+Treating EHR implementation as an exogenous treatment vairble, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and amount of patients seen.
 
@@
 
-\section{Background}
+\section{Institutional Details}
 
@@
 "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
+\textcolor{red}{add details about how this technology impacts physician day to day life}
 
@@
 
-I include means of these variables in physicians either younger than or older than 45 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}.
+I include means of these variables in physicians either younger than or older than 60 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}.
 Senior physicians see more patients and work in more hospitals on average than younger physicians.
-Physicians over 45 are slightly more likely to be exposed to EHRs in both general hospitals and low integration hospitals, but a smaller fraction of their hospitals utilize EHRs.
 These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general.
 Finally, physicians in the higher age bracket are less likely to switch zip codes.
+The age groups are exposed to EHRs in similar ways.
 \textcolor{red}{why is there equal retirement between young and old?}
@@
 Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital.
-The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR.
-On the right panel, only physicians over the age of 45 in 2017 are included.
+The left hand panel includes all physicians and shows no effect of EHR implementation on retirement.
+When limiting the sample to only physicians 60 and above (retirement age), the year of EHR implementation leads to a .37 percentage point increase in the probability of retiring.
 
+\vspace{5mm}
 \begin{figure}[ht]

### 2022-03-22 Scripts/analysis.R
add graph for te effect of retirement on patient count (for retirement subsection)

@@ -204,8 +204,6 @@ ggdid(retire_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title
 ggsave(file="CS_retire_allEHR.pdf",path="Objects")
 
 
-
-
 # OTHER RETIREMENT OUTCOMES --------------------------------------------------------------------------------
 
 # Senior Physicians
@@ -269,6 +267,38 @@ ggdid(retireyoung_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, 
 
 # Save the plot
 ggsave(file="CS_retireyoung_allEHR.pdf",path="Objects")
+
+
+# DO PHYSICIANS WHO RETIRE REUCE PATIENT COUNT FIRST?
+retirepatientcount_cs <- att_gt(
+  yname = "npi_unq_benes",                # LHS Variable
+  gname = "minyr_retire",             # First year a unit is treated. (set to 0 if never treated)
+  idname = "DocNPI",               # ID
+  tname = "year",                  # Time Variable
+  # xformla = NULL                 # No covariates
+  xformla = ~grad_year,            # Time-invariant controls
+  data = dplyr::filter(
+    Physician_Data,minyr_EHR>0 & ever_retire==1 & minyr_retire<2016 & minyr_retire>2010),   # Remove never-treated units and young physicians
+  # data = Physician_Data
+  est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
+  control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
+  clustervars = "DocNPI",          # Cluster Variables          
+  anticipation=0                   # can set a number of years to account for anticipation effects
+)
+# Save p-value for pre-trends to put in footnote of table
+p<-retirepatientcount_cs$Wpval
+
+# Aggregate the effects
+retirepatientcount_cs_dyn <- aggte(retirepatientcount_cs, type = "dynamic", na.rm=T)
+
+# Create a plot
+ggdid(retirepatientcount_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of Retirement on Patient Count by Length of Exposure \nSenior Physicians") + 
+  labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
[…]

### 2022-03-22 Scripts/sumstats.R

@@ -25,7 +25,7 @@ Physician_Data <- read_rds(paste0(created_data_path,"Physician_data.rds"))
 # General Summary Stats Tables:  ----------------------------------------------------------------
 
 # Physician Level
-sum_stats_fullsample <- Physician_Data %>% ungroup() %>%
+sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
@@ -54,7 +54,7 @@ knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
              caption="Summary Statistics",
              booktabs=TRUE,
              escape=F,
-             align=c("l","r","r","r","r","r"),
+             align=c("l","c","c","c","c","c"),
              position="h") %>%
   kable_styling(full_width=F) %>%
   pack_rows(index = c("Outcomes" = 5, "Treatment" = 3, "Characteristics" = 4))
@@ -62,7 +62,7 @@ knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
 
 # Summary Stats of all variables by old vs. young vs. those who retire ---------------------------------------------------------
 means_old <- Physician_Data %>% ungroup() %>%
-  filter(minyr_EHR>0 & (minyr_retire<2015 | is.na(minyr_retire))) %>%
+  filter(minyr_EHR>0) %>%
   filter(max_age>59) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
@@ -79,7 +79,7 @@ means_old <- Physician_Data %>% ungroup() %>%
   dplyr::rename(value_old=value)
 
 means_young <- Physician_Data %>% ungroup() %>%
-  filter(minyr_EHR>0 & (minyr_retire<2015 | is.na(minyr_retire))) %>%
+  filter(minyr_EHR>0) %>%
   filter(max_age<=59) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
@@ -117,7 +117,7 @@ means_bind <- means_old %>%
   left_join(means_retire,by="var")
 
[…]

### 2022-03-22 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{setspace}
-\singlespacing
+%\singlespacing
+\doublespacing
 \addtolength{\skip\footins}{2pc plus 5pt}
@@
 
-The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making.
-Their interaction has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
+The interaction between technological innovation and labor markets has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
 Labor market changes for one job type may have rippling effects on populations serviced by that industry.
-One such job is that of a physician, where physician choices and labor markets directly affect patient accessibility and quality of care received.
+It is straightforward to see this for a physician since choices and labor markets directly affect patient accessibility and quality of care received.
 For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
-Thus, shocks to physician labor markets or work settings potentially have drastic implications for patients.
+Physician labor markets are an important topic of study because of their implications on care.
 I investigate physician labor market changes as a result of a major technology change in the U.S.
@@
 Physicians, as primary users of this technology, play a meaningful role in whether potential benefits of EHRs are realized.
-However, with the rapid implementation of EHRs, the day to day life for physicians changed rapidly.
+With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly.
 In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
@@
 Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.
-Treating EHR implementation as an exogenous treatment vairble, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and amount of patients seen.
+Treating EHR implementation in hospitals as exogenous to an individual physician, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and amount of patients seen.
 
-Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel on hospitalists and other general practice physicians connected to hospitals from 2009-2017.
-The data captures the extent to which the physician in exposed to EHRs in hospitals, various labor market outcomes, and demographic information such as zip code, age, and gender.
+Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians connected to hospitals.
+The data captures the extent to which the physician in exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
 I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician.
-Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals.
-I consider exposure on two dimensions: working in any hospital that implements an EHR, and working in a low-integrated hospital that implements an EHR.
-To study retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it.
-Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here.
-(The funny thing is, I find that young physicians "retire"...
-I need to figure out what this means and possibly rephrase the last couple sentences)}.
+Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals on two dimensions: working in any hospital that implements an EHR, and working in a low-integrated hospital that implements an EHR.
+A key assumption that underlies the analysis is that EHR exposure is exogenous to physicians.
+While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decisions.
+This is the reason I include a second definition of EHR exposure that only considers hospitals where vertical integration between physicians and the hospital are low, as this indicates a much lower probability that EHR implementation is endogenous.
+In the Appendix I consider different thresholds of exposure as treatment variables for robustness.
 
-For those who stay in the labor force, there may be margin to change physicil place of work if physicians are unhappy with the specific loss of autonomy experienced at a given hospital.
-That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different?
-To answer this question, I limit the sample to any physicians who stay in the labor force after EHR exposure and use the same event study framework to estimate the effect on whether the physician switches zip codes or increases office-based patients.
+Retirement typically takes place late in life, which is no different for physicians.
+The average physician retirement age is 65 (\cite{collier2017challenges}) However, the data reveals that even physicians as young as 35 are "retiring".
+Young physicians are likely not retiring in the formal sense, but leaving a clinical setting for a more administrative setting.
+Alternative careers include teaching, consulting, or medical malpractice expert.
+My data does not allow granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether.
+The implications of physicians retiring are the same whether they are young or old, that access to care issues may have been exemplified.
+Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .1 ppt increase in the likelihood of a senior physician retiring.
+While there were occurrences of young physicians leaving clinical work, there is no evidence that it was because of EHR implementation.
+
+Not everyone chooses to forego clinical work altogether.
+However, there may be margin to change physical place of work if physicians are not satisfied with their experience at a given hospital, whether it is because of burnout, loss of autonomy, or factors unrelated to EHRs.
+I investigate whether physicians switch from primarily hospital based work to smaller office based settings, or switch hospital locations?
+To answer this question, I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR implementation on whether the physician switches zip codes or increases office-based patients.
 I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.
@@
 Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen.
-The population of physicians left in this analysis are those who do not retire or switch to an office based setting.
-I find that the number of patients seen does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains.
-After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs.
-This result speaks to the disconnect between expected effects and realized effects of EHRs which is discussed in the health information technology literature.
-Since more patients are being seen by physicians, efficiency and access to care have both increased while costs (Medicaid patients) and quality remain unchanged.
-\textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}
-
-A key assumption underlying this analysis is that EHR exposure is exogenous to individual physicians.
-That is, a hospital makes the decision to implement an EHR independent of a physicians' labor market response to that implementation.
-This assumption is not unrealistic in large hospitals who employ many physicians, but may be violated in small hospitals where a physician has a voice in management.
-To address this, I first limit the analysis to large hospitals.
-However, small hospitals are especially important in studying patient access to care.
-To include small hospitals in the analysis, I instrument EHR adoption with local bandwidth capacity.
-\textcolor{red}{Relate these findings to previous findings.
-I need to make this paragraph more clear as I improve what I'm doing}
 
@@
 
-\section{Institutional Details}
+\section{Institution}
 
 EHRs have been an important feature of healthcare since the 1980s and 1990s.
-Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling.
-Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected.
-The cost of utilizing an EHR was extremely large, and cost increased as the amount of capabilities increased, which led to the widespread view that EHRs were strictly complements to paper records and not a feasible substitute.
-The possibility of using solely EHRs increased as computers became more portable, creating what is known as the "physician workspace": a computer station for a physician to interface directly with EHRs to record patient updates.
-However, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
+Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and/or scheduling.
+Prior to computerization, physicians were not able to interface directly with EHRs, and thus were not drastically affected by their implementation.
+The cost of utilizing an EHR was large and increasing in the amount of capabilities of the system, which led to the widespread view that EHRs were strictly complements to paper records, not a feasible substitute.
+As computers were made more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
+Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
 Automation in data entry was non-existent, making it extremely time consuming for the user.
-Even when automation was developed for particular machines that performed monitoring, the hospital still held responsible for the accuracy of information and therefore required physicians to double check each data entry.
+Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry.
 
 The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines.
-These requirements include having at least 80\% of patients in the system, recording answers to specific questions, and protecting the security of the system to ensure privacy.
+The guidelines include having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy.
 Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies.
 Hospitals began receiving payments for the first stage of meaningful use in 2011.
-This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
-\textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, physicians still continue to report frustration.
-"Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
-\textcolor{red}{add details about how this technology impacts physician day to day life}
+This resulted in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
+Figure \ref{fig:meanuse} shows a comparison of U.S.
+hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
+2013.
+
+\vspace{3mm}
+\begin{figure}[ht]
+    \centering
+    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
+    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
+    \caption*{Source: HealthIT.gov}
+    \label{fig:meanuse}
+\end{figure}
+\vspace{3mm}
+
+While EHRs were widely utilized in hospitals by 2015, physicians still continue to report frustration.
+For every patient seen, a primary care physician spends an average of 16 minutes actively using an EHR, an amount of time that often exceeds face to face time with the patient (\cite{overhage2020physician}).
+The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication.
+Appendix (\textcolor{red}{section}) includes a computer screen image of what a physician may see when looking at EHRs.
+Because of the added administrative burden, "some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
 
@@
 Using various linked datasets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time and other relevant characteristics.
-The different data sets used to construct this panel are described below.
+The different data sets used to construct the panel are described below.
 
@@
 
-The dependent variables in the analysis have to do with physician labor market outcomes, and are constructed using claim and patient count variables.
+The dependent variables in the analysis have to do with physician labor market outcomes, and are constructed using claim and patient counts.
 Retirement is an indicator variable that is equal to 1 in the first year that all future claims in all zip codes are equal to 0.
@@
 
-For those physicians who do not retire, I analyze whether there is evidence of switching to different places of work.
+For physicians who do not retire, I analyze whether there is evidence of switching to different places of work.
 Two outcome variables capture moving towards office based settings, one is the fraction of patients seen in an office setting and the other is an indicator for whether the physician works in an office at all.
@@
 
-CMS Shared Patient data is from 2009 to 2015, and records the number of patients who bill two entities under Medicaid in the same day.
+CMS Shared Patient data contains annual information from 2009 to 2015, and records the number of patients who bill two entities under Medicaid in the same day.
 For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen the same day, then those two physicians have a shared patient in common.
-I utilize this data to find physicians' EHR exposure based on the hospitals they share sufficient patients with, which defines treatment variables for the analysis.
+I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with, which defines treatment variables for the analysis.
 I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
-I care specifically about primary care physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital.
+I care specifically about primary care physicians who have a close working relationship with hospitals who do rounds within at least one hospital.
 The types of physicians are already limited to primary care physicians, but there still may be some office-based primary care physicians in the shared patient data.
@@
 A physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data.
-The reason for this is to avoid dropping physicians who drop out of the data for many years, as this may indicate retirement.
+It is important to avoid dropping physicians who drop out of the data for many years, as this may indicate retirement and skew the results.
 \textcolor{red}{I explore other thresholds in the appendix.}
 
-Using hospital NPI in each pair, I link the pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
+Using hospital NPI, I link the pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
 I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
@@
 Since this data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017.
+That is, the data contains no units which are "never-treated".
 
@@
 
-Overall summary statistics of the variables are shown in \hyperref[tab:sumstats1]{Table \ref{tab:sumstats1}}. \textcolor{red}{add commentary on Table 1}
+Summary statistics for the entire sample are shown in Table \ref{tab:sumstats1}.
+Only 2\% of physicians retire over the course of the sample, \textcolor{red}{compare this to national average?
+seems low}.
+A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting.
+All physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed.
+Physicians are typically exposed in any capacity about one fourth of the way through the sample and in low-integrated hospitals about halfway through the sample.
+The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.
 
-\vspace{5mm}
 \import{Table Code}{overall stats.tex}
-\vspace{5mm}
 
-I include means of these variables in physicians either younger than or older than 60 in \hyperref[tab:sumstats2]{Table \ref{tab:sumstats2}}.
+I also include a table of means comparing physicians younger than 60, older than 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}.
 Senior physicians see more patients and work in more hospitals on average than younger physicians.
@@
 Finally, physicians in the higher age bracket are less likely to switch zip codes.
-The age groups are exposed to EHRs in similar ways.
-\textcolor{red}{why is there equal retirement between young and old?}
+The age groups are exposed to EHRs in similar ways, but are exposed to EHRs in low integrated setting slightly later.
+Retirement occurs just as often between young and physicians.
+As discussed, this could be due to physicians leaving a clinical setting for an administrative position.
+Further, it is feasible that physicians just below the threshold are retiring in the formal sense.
 
-\vspace{5mm}
 \import{Table Code}{split stats.tex}
-\vspace{5mm}
 
-To examine variation in treatment variables, \hyperref[fig:figure1]{Figure \ref{fig:figure1}} shows the averages over time.
+Figure \ref{fig:treatmentgraph} shows variation in treatment variables over time prior to dropping the never-treated observations.
 In 2009, 26\% of the sample of physicians were exposed to an EHR.
@@
 By 2015, 84\% of physicians had exposure to an EHR in at least one hospital in their network.
-Exposure to EHRs only in low integrated hospitals follows a similar trend with lower percentages in each year.
+Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates.
 
-\vspace{5mm}
 \begin{figure}[ht]
@@
     \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
-    \label{fig:figure1}
+    \includegraphics[scale=.6]{Objects/sum_stats_year.pdf}
+    \label{fig:treatmentgraph}
 \end{figure}
-\vspace{5mm}
-
-\newpage
 
@@
 In this setting, treatment effects likely vary across physicians who are exposed in different years since the underlying hospitals who implement likely vary.
-Therefore, to avoid negative weighting issues that would arise in a two way fixed effects specification, I utilize \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) to estimate average group time treatment effects.
+Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I utilize \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) to estimate average group time treatment effects (ATT(g,t)).
+There are several assumptions I make to identify ATT(g,t).
 
-\textcolor{red}{outline assumptions necessary for identification and talk about doubly robust estimator}
+\subsection{Anticipation}
+
+ The first assumption necessary for identification is that anticipation either does not exist or is well understood.
+The estimator can takes anticipation into account as long as the duration is well understood and the same for all units.
+Anticipation plays into EHR implementation in an interesting way.
+Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until years later.
+Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
+Therefore, in my main specification I assume there is no anticipation in physicians.
+In Appendix (\textcolor{red}{section}), I show results under the assumption of a one year anticipation period.
+
+\subsection{Parallel Trends with Not-yet-treated}
+
+\subsection{No Treatment Reversal and Overlap}
 
@@
 
-It is not uncommon in developed countries for workers to plan for retirement well in advance.
-Many workers have a retirement age set and spend their working life saving to get to that point.
-An exogenous shock to the worker is not expected to change retirement age because of the large sum of money necessary to formally leave the labor force.
+It is not uncommon in developed countries for workers to plan for formal retirement well in advance.
+An exogenous shock to a worker is not expected to change retirement age because of the large sum of money necessary to formally leave the labor force.
 However, physicians typically make this decision differently than most other workers.
 Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
-The reason being that the type of person who selects into being a physician will, on average, have a high regard for helping people.
-When it comes time to retire, they hesitate to abandon patients they have seen over the course of their career.
-This decision to delay retirement is not financial.
+When it comes time to retire, many physicians hesitate to abandon patients they have seen over the course of their career.
+This decision to delay retirement is not financial, but altruistic.
 Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
+Further, formal retirement is not the only way to stop seeing patients.
+There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, hospital management, etc.
+While young physicians are probably not retiring in the traditional sense, either old or young physicians could be career switching.
+The term "retirement" for the purpose of this paper remains agnostic about what the physician does after they stop seeing patients in a clinical setting.
 
-An exogenous shock to work setting could, if it outweighs the reasons for staying, cause an earlier exit than would have occured without the shock.
-Medical articles suggest through interaction with physicians that the implementation of EHRs did lead to retirement in older physicians.
+An exogenous shock to work setting, such as EHR implementation in place of work, may cause an earlier exit than would have occurred without the shock.
+Medical articles suggest through interaction with physicians that the implementation of EHRs led to retirement in older physicians.
 \textcolor{red}{cite articles}.
-The aim of this section is to test that hypothesis and explore its implications.
+The first aim of this section is to test that hypothesis..
 
-\subsection{Event Study Results}
+\subsection{Effect of EHR Exposure on Retirement Decision: Event Study Results}
 
 Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital.
-The left hand panel includes all physicians and shows no effect of EHR implementation on retirement.
-When limiting the sample to only physicians 60 and above (retirement age), the year of EHR implementation leads to a .37 percentage point increase in the probability of retiring.
+The top graph suggests that for any physician, being exposed to an EHR leads to an almost .0025 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
+Relative to the sample mean, this is an increase of approximately 1\%.
+We can examine how this result differs when considering different age groups of physician.
+These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that old physicians are driving the result in the first year after EHR exposure.
+That is, a physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a .25\% increase relative to the mean.
+The results for physicians less than 60 reveals something interesting, that the younger physicians are driving the results in the second year after exposure.
+If young physicians are choosing to leave clinical settings, this result is intuitive since career switching may take more time to prepare for than formal retirement.
 
-\vspace{5mm}
 \begin{figure}[ht]
-\caption{Event Study Graphs}
+\caption{Event Study Graphs: Retirement}
 \vspace{2mm}
-        \begin{minipage}[b]{0.47\linewidth}
 \centering
-            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR.pdf}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
+\hspace{20mm}
+\includegraphics[scale=.4]{Objects/CS_retire_allEHR.pdf}
+\vspace{3mm}
+\newline
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
             \centering
-            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old.pdf}
-        \end{minipage}
+            \includegraphics[width=\textwidth]{Objects/CS_retireold_allEHR.pdf}
+        \end{minipage}}
+        \hspace{0.1cm}
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/CS_retireyoung_allEHR.pdf}
+        \end{minipage}}
         \label{fig:retirefirst}
 \end{figure}
-\vspace{5mm}
 
-\newpage
+\subsection{Pre-Retirement Claim Counts}
+
+If physicians are retiring because of EHRs, this may lead to the question of whether the physician is choosing to retire or the hospital is forcing older physicians out as they make major changes.
+To address this, I utilize claim counts in years prior to retirement.
+Some studies suggest that physicians may significantly reduce work or work part-time in preparation for formal retirement (\cite{hedden2017patterns}, \cite{acp_2017}).
+If there is evidence of physicians reducing workload prior to retirement, that is indicative of physicians choosing to retire as a result of EHR exposure.
+Thus, I seek to answer the question do physicians of retirement age reduce patients seen due to EHRs relative to other populations?
+Did physicians who eventually retire in the sample reduce patients seen prior to retirement?
 

### 2022-03-22 Table Code/overall stats.tex
Updates from Overleaf

@@
-\begin{table}[h]
+\begin{table}[ht]
+
 \caption{Summary Statistics}
 \centering
-\begin{tabular}[t]{lrrrrr}
+\begin{tabular}[t]{lccccc}
 \toprule
@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 403020 & 0.02 & 0.13 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 403020 & 0.31 & 0.36 & 0 & 1\\
-\hspace{1em}Work in an Office & 403020 & 0.57 & 0.49 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 372357 & 0.08 & 0.28 & 0 & 1\\
-\hspace{1em}Number of Patients & 403020 & 2220.46 & 3207.08 & 0 & 199765\\
+\hspace{1em}Ever Retire & 337176 & 0.02 & 0.13 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 337176 & 0.31 & 0.36 & 0 & 1\\
+\hspace{1em}Work in an Office & 337176 & 0.57 & 0.49 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 313663 & 0.08 & 0.28 & 0 & 1\\
+\hspace{1em}Number of Patients & 337176 & 2274 & 3293 & 0 & 199765\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 403020 & 0.64 & 0.48 & 0 & 1\\
-\hspace{1em}Exposure to an EHR (Low Integration) & 403020 & 0.42 & 0.49 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 403020 & 0.52 & 0.46 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 337176 & 0.76 & 0.43 & 0 & 1\\
+\hspace{1em}Exposure to an EHR (Low Integration) & 337176 & 0.51 & 0.50 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 337176 & 0.63 & 0.44 & 0 & 1\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 403002 & 47.71 & 9.38 & 23 & 89\\
-\hspace{1em}Female & 403020 & 0.31 & 0.46 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 403020 & 1.49 & 0.83 & 1 & 20\\
-\hspace{1em}Number of Systems Worked With & 403020 & 1.30 & 0.59 & 1 & 8\\
+\hspace{1em}Age & 337176 & 47.67 & 9.35 & 23 & 89\\
+\hspace{1em}Female & 337176 & 0.31 & 0.46 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 337176 & 1.56 & 0.87 & 1 & 20\\
+\hspace{1em}Number of Systems Worked With & 337176 & 1.34 & 0.62 & 1 & 8\\
 \bottomrule
 \end{tabular}
+\label{tab:sumstats1}
 \end{table}

### 2022-04-05 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{booktabs}
+\usepackage{bbm}
 \usepackage[font=footnotesize,labelfont=bf]{caption}
@@
 Retirement typically takes place late in life, which is no different for physicians.
-The average physician retirement age is 65 (\cite{collier2017challenges}) However, the data reveals that even physicians as young as 35 are "retiring".
-Young physicians are likely not retiring in the formal sense, but leaving a clinical setting for a more administrative setting.
+The average physician retirement age is 65 (\cite{collier2017challenges}) However, the data used in my analysis reveals that even physicians as young as 35 are ``retiring".
+Young physicians are likely not retiring in the formal sense, but leaving a clinical setting for a different type of work.
 Alternative careers include teaching, consulting, or medical malpractice expert.
-My data does not allow granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether.
-The implications of physicians retiring are the same whether they are young or old, that access to care issues may have been exemplified.
-Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .1 ppt increase in the likelihood of a senior physician retiring.
-While there were occurrences of young physicians leaving clinical work, there is no evidence that it was because of EHR implementation.
+I do not observe granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether.
+Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring.
+Similarly, younger physicians were .0025 ppts more likely to retire when exposed to EHRs.
+Whether physicians are young or old, these results suggest that access to care issues may have been exemplified as EHRs were implemented, and there was a surge of experienced physicians who left the labor force and were no longer available to mentor those learning under them.
 
 Not everyone chooses to forego clinical work altogether.
-However, there may be margin to change physical place of work if physicians are not satisfied with their experience at a given hospital, whether it is because of burnout, loss of autonomy, or factors unrelated to EHRs.
-I investigate whether physicians switch from primarily hospital based work to smaller office based settings, or switch hospital locations?
-To answer this question, I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR implementation on whether the physician switches zip codes or increases office-based patients.
-I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.
+However, there may be margin to change physical place of work if physicians are not satisfied with their experience at a given hospital, whether it is because of burnout, loss of autonomy, or other factors.
+I investigate whether physicians switch from primarily hospital based work to office based settings, or more generally switch primary zip codes.
+I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR implementation on switching work settings.
+I find that younger physicians are more likely to work in an office in the year after being exposed to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
+Further, \textcolor{red}{add result about zip code switching}.
+This implies that physicians may incur significant switching costs in order to avoid a mismatch of preferences rooted in burnout or loss of autonomy.
+This is important as policymakers attempt to regulate and homogenize healthcare.
 
-Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen.
+Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen, and overall billing activity by the physician.
 
@@
 
-This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010.
-\textcolor{red}{This needs to be expounded upon a lot.
-Perhaps one or two paragraphs for each area of contribution.}
+This paper is at the intersection of both health economics and labor economics and thus contributes to our understanding of multiple important branches of literature.
+There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH act taking effect.
+Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found less impactful results, that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{dranove2014trillion}).
 
-The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively.
-Despite many case studies and hospital-level analysis that indicate a large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}).
-Further, there have been no significant cost decreases due to EHRs in the short or long run.
-If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}).
-As discussed earlier, these findings are puzzling due to the expected benefits of EHRs.
-This paper contributes to our knowledge of this area in re-framing past finding into a patient access to care problem instead of an efficiency problem.
+I both expand and improve upon this literature.
+First, I consider a different mechanism by which EHRs could indirectly affect patients through physicians' response to them.
+This is the first study to my knowledge that connects EHR implementation to physician labor market decisions.
+Further, I improve on the literature by considering the primary time period in which EHRs were rapidly implemented.
+This provides sufficient variation in the timing of EHR implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in a typical two way fixed effects setup.
 
-\textcolor{red}{Expand both contribution and literature review and combine them to flow better.}
-
-\newpage
+The interaction between labor markets and technology implementation are of major interest in labor economics.
 
@@
 
-\vspace{3mm}
 \begin{figure}[ht]
@@
 \end{figure}
-\vspace{3mm}
 
@@
 Medicare Data on Provider Practice and Specialty is a database of physicians which details claim counts and specialty from 2009-2017.
-The relevant information included is physician NPI, primary and secondary specialties, patient counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
-First, I make limitations to the sample of physicians based on specialty type.
-Primary care physicians are the most likely to be exposed to the administrative burden of EHRs in hospitals.
-Thus, I only include physicians whose primary specialty is hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
+The relevant information included is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
+I first limit the sample of physicians based on specialty type.
+Primary care physicians are exposed to the administrative burden of EHRs in hospitals more-so than specialists.
+Thus, I only include physicians whose primary specialty is reported as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years, who likely do not have strong ties to the hospitals they have limited work in.
 
-The dependent variables in the analysis have to do with physician labor market outcomes, and are constructed using claim and patient counts.
-Retirement is an indicator variable that is equal to 1 in the first year that all future claims in all zip codes are equal to 0.
-I do not observe retirement in 2017, and drop those who "retire" in 2016 since one year of zero claims does not necessarily mean retirement.
-\textcolor{red}{maybe show the distribution of ages by those who retire and those who don't.}
-
-For physicians who do not retire, I analyze whether there is evidence of switching to different places of work.
-Two outcome variables capture moving towards office based settings, one is the fraction of patients seen in an office setting and the other is an indicator for whether the physician works in an office at all.
-While I do not observe different hospitals in the data, I also attempt to capture any type of switching (to office or from hospital to hospital) by measuring if a physician changes zip codes from one year to the next.
-Finally, for physicians who do not exhibit switching settings, I utilize number of patients as an outcome variable.
+This data includes individual labor market information on physicians which I use to construct the dependent variables for my analysis.
+Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen and overall billing activity.
+These outcome variables and their construction are described in detail below.
 
@@
 
-CMS Shared Patient data contains annual information from 2009 to 2015, and records the number of patients who bill two entities under Medicaid in the same day.
-For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen the same day, then those two physicians have a shared patient in common.
-I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with, which defines treatment variables for the analysis.
+CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicaid in the same day.
+For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen by the patient in the same day, then those two physicians have that shared patient in common.
+I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with.
+This measure of exposure defines various treatment variables for my analysis.
 I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
-I care specifically about primary care physicians who have a close working relationship with hospitals who do rounds within at least one hospital.
-The types of physicians are already limited to primary care physicians, but there still may be some office-based primary care physicians in the shared patient data.
+I care specifically about primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital.
+The types of physicians are already limited to primary care, but there still may be some office-based primary care physicians in the shared patient data.
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
-A physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data.
-It is important to avoid dropping physicians who drop out of the data for many years, as this may indicate retirement and skew the results.
+Specifically, a physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data.
+This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.
 \textcolor{red}{I explore other thresholds in the appendix.}
@@
 I create two measures of EHR exposure.
-The first is exposure to any EHR in any hospital, and the second is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})).
-I then aggregate to the physician level.
-Since this data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017.
+The first is exposure to any EHR in any hospital, which attempts to capture the effect of first-time EHR use on physician labor markets.
+One concern with this approach is that a typical physician may be involved in the decision to implement an EHR and thus the treatment would be endogenous.
+While I cannot perfectly address this concern, I define an alternative treatment variable which theoretically will have less likelihood of this endogeneity problem.
+That is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})).
+In low integrated settings, physicians do not have ownership or much control in the hospital setting.
+Results for this alternative defintion of exposure are presented in Appendix ().
+I then aggregate this data to the physician level.
+Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017.
 That is, the data contains no units which are "never-treated".
@@
 Finally, I include a physician's graduation year from Physician Compare.
-I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior.
+I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.
+
+\subsection{Outcome Variables}
+
+The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
+$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
+where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$.
+In words, retire is equal to one in the first year that a physician has no claim counts at any point in the future.
+Alternatively, I could also define retirement using future number of patients.
+I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
+
+Next, I consider a physician switching place of work, such as switching to an office based setting or even selecting into different hospitals where preferences may be more aligned.
+I measure this in the data in three ways.
+I consider the fraction of patients a physician sees in an office based setting, which is directly available in the data.
+Further, I use this variable to construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office based setting.
+There is less variation in this data but allows insight to a different type of decision making where physicians may enter or leave office based setting based on EHR exposure.
+Lastly, I utilize the zip code information to define an indicator variable equal to one only if the physician's list of zip codes changes from one year to the next.
+While these variables do not give direct insight into whether physicians specifically move from one hospital to another, they do allow for exploration of whether settings are changing as a result of EHRs in general.
+
+Finally, I consider outcome variables which explore productivity and billing activity once EHRs are implemented in hospitals.
+Productivity is measured in the number of patients a physician sees in a give year, where the billing activity is the total claim count for the physician.
+I distinguish between the two because it could be thd case that productivity does not actually increase, but EHRs could encourage excessive billing of things that would not have been billed for otherwise.
+\textcolor{red}{put distribution graphs here?}
 
@@
 Summary statistics for the entire sample are shown in Table \ref{tab:sumstats1}.
-Only 2\% of physicians retire over the course of the sample, \textcolor{red}{compare this to national average?
-seems low}.
+Only 2\% of physicians retire over the course of the sample, which equates to approximately 750 physicians in the sample who retire at some point.
 A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting.
-All physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed.
+By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed.
 Physicians are typically exposed in any capacity about one fourth of the way through the sample and in low-integrated hospitals about halfway through the sample.
@@
 Finally, physicians in the higher age bracket are less likely to switch zip codes.
-The age groups are exposed to EHRs in similar ways, but are exposed to EHRs in low integrated setting slightly later.
-Retirement occurs just as often between young and physicians.
-As discussed, this could be due to physicians leaving a clinical setting for an administrative position.
-Further, it is feasible that physicians just below the threshold are retiring in the formal sense.
+The age groups are exposed to EHRs in similar ways, but younger physicians are exposed to EHRs in low integrated setting slightly later.
+Retirement occurs just as often in the different age groups, indicating the possibility of physicians leaving a clinical setting for an administrative or other position.
 
@@
 
-Figure \ref{fig:treatmentgraph} shows variation in treatment variables over time prior to dropping the never-treated observations.
+In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time prior to dropping the never-treated observations.
 In 2009, 26\% of the sample of physicians were exposed to an EHR.
@@
 Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates.
+\textcolor{red}{i need to either fix or delete this}
 
@@
 
-In this setting, treatment effects likely vary across physicians who are exposed in different years since the underlying hospitals who implement likely vary.
-Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I utilize \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) to estimate average group time treatment effects (ATT(g,t)).
-There are several assumptions I make to identify ATT(g,t).
+In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
+Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$:
+$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
+where $G_g=1$ for those in group $g$.
+A group indicates all physicians who were treated, or first exposed to an EHR, in the same year.
+To estimate the heterogeneous treatment effects, I utilize the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
 
-\subsection{Anticipation}
+There are several assumptions necessary to identify $ATT(g,t)$.
+First, I assume that treatment is not reversed once it occurs.
+That is, once a physician is exposed to an EHR, they cannot be un-exposed.
+This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure.
+An EHR is a costly technology and requires a significant amount of work to implement.
+A hospital does not have incentive to un-implement an EHR once it is implemented.
+They may add features or switch vendors, but do not reverse EHR use.
+Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR.
+Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
+Thus, physicians remain treated in all future years after exposure.
 
- The first assumption necessary for identification is that anticipation either does not exist or is well understood.
-The estimator can takes anticipation into account as long as the duration is well understood and the same for all units.
-Anticipation plays into EHR implementation in an interesting way.
+Second, I assume physicians do not anticipate EHR exposure prior to occurrence.
 Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until years later.
 Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
-Therefore, in my main specification I assume there is no anticipation in physicians.
-In Appendix (\textcolor{red}{section}), I show results under the assumption of a one year anticipation period.
+Therefore, I proceed in the main specification assuming no anticipation.
+However, I explore and present results for a one year anticipation period in Appendix ().
 
-\subsection{Parallel Trends with Not-yet-treated}
+As is usual in an event study framework, I assume a version of parallel trends based on not-yet-treated units.
+I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods.
 
-\subsection{No Treatment Reversal and Overlap}
+\section{Effect of EHR Exposure on Labor Market Outcomes}
 
-\section{Retirement}
+\subsection{Retirement}
 
@@
 Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
-When it comes time to retire, many physicians hesitate to abandon patients they have seen over the course of their career.
+When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career.
 This decision to delay retirement is not financial, but altruistic.
 Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
-Further, formal retirement is not the only way to stop seeing patients.
-There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, hospital management, etc.
-While young physicians are probably not retiring in the traditional sense, either old or young physicians could be career switching.
-The term "retirement" for the purpose of this paper remains agnostic about what the physician does after they stop seeing patients in a clinical setting.
+Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role.
+There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management.
+While it is unlikely that physicians under 55 are formally retiring, either old or young physicians could be career switching.
+The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no claim about what the physician does afterwards.
+A physician is said to retire in a given year only if it is the first year in which the physician has no future claims.
 
-An exogenous shock to work setting, such as EHR implementation in place of work, may cause an earlier exit than would have occurred without the shock.
-Medical articles suggest through interaction with physicians that the implementation of EHRs led to retirement in older physicians.
-\textcolor{red}{cite articles}.
-The first aim of this section is to test that hypothesis..
+An exogenous shock such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock.
+Media articles suggest through interaction with physicians that the implementation of EHRs led to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
+The first aim of this section is to test that hypothesis.
 
-\subsection{Effect of EHR Exposure on Retirement Decision: Event Study Results}
-
-Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in any hospital.
-The top graph suggests that for any physician, being exposed to an EHR leads to an almost .0025 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
-Relative to the sample mean, this is an increase of approximately 1\%.
-We can examine how this result differs when considering different age groups of physician.
-These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that old physicians are driving the result in the first year after EHR exposure.
-That is, a physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a .25\% increase relative to the mean.
-The results for physicians less than 60 reveals something interesting, that the younger physicians are driving the results in the second year after exposure.
-If young physicians are choosing to leave clinical settings, this result is intuitive since career switching may take more time to prepare for than formal retirement.
+In Figure \ref{fig:retirefirst}, I present the aggregated group time treatment effects of being exposed to an EHR in any hospital for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
+The top graph suggests that for any age physician, being exposed to an EHR leads to an almost .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
+While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, which is only 2 out of every hundred.
+Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure.
+It would be alarming to see any huge effects on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
+These estimates are noisy, however, given the rarity of retirement in the sample the precision does indicate a positive effect occurring.
 
 \begin{figure}[ht]
-\caption{Event Study Graphs: Retirement}
+\caption{Effect of any EHR Exposure on Retirement}
 \vspace{2mm}
@@
 \hspace{20mm}
-\includegraphics[scale=.4]{Objects/CS_retire_allEHR.pdf}
+\includegraphics[scale=.45]{Objects/CS_retire_allEHR.pdf}
 \vspace{3mm}
@@
 
-\subsection{Pre-Retirement Claim Counts}
+Now, I examine how these result differ when considering different age groups of physician.
+These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure.
+That is, a physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean, whereas younger physicians are far less likely, if at all, to retire due to EHR exposure.
+Physicians in the young category have a slightly increased likelihood of retiring in the second year after exposure relative to the first year.
+While I do not have data to test this, if young physicians are career switching then it may take an additional year to prepare for that decision relative to retirement.
 
-If physicians are retiring because of EHRs, this may lead to the question of whether the physician is choosing to retire or the hospital is forcing older physicians out as they make major changes.
-To address this, I utilize claim counts in years prior to retirement.
-Some studies suggest that physicians may significantly reduce work or work part-time in preparation for formal retirement (\cite{hedden2017patterns}, \cite{acp_2017}).
-If there is evidence of physicians reducing workload prior to retirement, that is indicative of physicians choosing to retire as a result of EHR exposure.
-Thus, I seek to answer the question do physicians of retirement age reduce patients seen due to EHRs relative to other populations?
-Did physicians who eventually retire in the sample reduce patients seen prior to retirement?
+A general concern in event study settings is whether pre-trends are driving the results.
+With each specification, I present the p-value for a Wald test for pre-trends.
+Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed in the significantly large p-values.
 
-\section{Work Setting}
+\subsection{Work Setting}
 
-\newpage
+For most physicians, EHRs will not impose a costly enough burden to induce retirement.
+However, there are other ways physicians might change their labor market behavior under sufficient incentive to avoid the technology.
+One such response is to shift from working in hospitals to working in office based settings, or to shift towards different hospitals.
+The implication of moving in this context may indicate something subtle about physician incentives, that costs may be incurred to avoid some negative aspects of EHRs such as loss of autonomy or burdensome administrative tasks.
 
-\section{Physician Productivity}
+I measure this outcome in multiple ways.
+First, I investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure.
+I do this using variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting.
 
-An important aspect in the implementation of EHRs is whether or not they increase productivity in users.
+I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
+The top panel is for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean.
+Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
+There is no evidence that older physicians are shifting towards offices.
+This is may be due to the group of senior physicians already being established in some capacity in offices prior to EHR exposure.
+This is supported by the significantly larger portion of senior physicians working in offices relative to younger physicians.
+
+Then senior physicians may be more affected on an internal margin, the fraction of patients seen in an office relative to hospital or other settings.
+The estimates for this outcome are shown in Figure \ref{fig:officesecond}.
+There is no evidence of an effect on the fraction of patients seen in an office, neither positive nor negative.
+There only seems to be a downward trend in fraction of patients seen in an office over time.
+\textcolor{red}{make sure to try this without defining fraction of patients as 0}
+
+\begin{figure}[ht]
+\caption{Effect of any EHR Exposure on Likelihood of Working in Office}
+\vspace{2mm}
+\centering
+\hspace{20mm}
+\includegraphics[scale=.45]{Objects/CS_office_ind_allEHR.pdf}
+\vspace{3mm}
+\newline
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/CS_office_indold_allEHR.pdf}
+        \end{minipage}}
+        \hspace{0.1cm}
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/CS_office_indyoung_allEHR.pdf}
+        \end{minipage}}
+        \label{fig:officefirst}
+\end{figure}
+
+\begin{figure}[ht]
+\caption{Effect of any EHR Exposure on Fraction of Patients in Office}
+\vspace{2mm}
+\centering
+\hspace{20mm}
+\includegraphics[scale=.45]{Objects/CS_office_frac_allEHR.pdf}
+\vspace{3mm}
+\newline
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/CS_office_fracold_allEHR.pdf}
+        \end{minipage}}
+        \hspace{0.1cm}
+        \fbox{\begin{minipage}[b]{0.47\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{Objects/CS_office_fracyoung_allEHR.pdf}
+        \end{minipage}}
+        \label{fig:officesecond}
+\end{figure}
+
+Another way to examine whether physicians are switching settings in a vague sense is to see whether the zip codes in which the physicians work is changing.
+Estimates for this outcome variable are shown in
+
+\subsection{Physician Productivity}
+
+An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users.
 Productivity, in this case, has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
@@
 In this case, the productivity gain could be from having the data assistant instead of EHR use.
-I investigate this below.
+I investigate this in the Sensitivity Analyses below.
+
+\subsection{Billing Activity}
+
+\section{Sensitivity Analyses}
 
@@
 \begin{figure}[ht]
+\centering
 \caption{Frequency of Data Assistant Enumeration by Year}
-
+\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
     \label{fig:dataassistant_histogram}
@@
 
-\subsection{Event Study Results}
+\section{Conclusion}
 
-\subsection{Relaxing Exogeneity of Implementation}
+In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes.
+This technology has been controversial and debated as a burdensome load on physicians that may contribute to a lack of improvements in the healthcare sector.
 
-\textcolor{red}{I need to put somewhere that i dropped 2015 for this analysis since it cut off at a different point than the other years}
+Using an event study framework, I find that physicians of all ages are more likely to stop clinical work altogether when exposed to an EHR in a hospital, where physicians of retirement age have an even larger increase relative to younger physicians.
+Further, I find that younger physicians are more likely to work in an office based setting after EHR exposure, and physicians who remain in the labor force increase both number of patients seen and billing activity.
+
+The implications of these results are blank.
+First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure.
+In a field where younger physicians learn a lot by watching those more experienced than them, this may particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
+Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid burdens externally placed on them.
+This has implications when considering the current policy movement towards regulated care, that physicians may go to extreme measures to avoid such regulation.
+Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users.
+While the technology may be a burden, physicians are able to see more patients because of their use.
+
+This paper provides an empirical foundation in a specific occurrence of the interaction between labor markets and technology.
 

### 2022-04-05 Scripts/analysis.R
update all event study graphs

@@ -4,6 +4,7 @@ library(ggplot2)
 library(readr)
 library(fixest)
 library(did2s)
+library(ggpubr)
 
 # ------------------------------------- ANALYSIS  ------------------------------------
 #                                       Hanna Glenn, Emory University
@@ -195,13 +196,11 @@ p<-retire_cs$Wpval
 retire_cs_dyn <- aggte(retire_cs, type = "dynamic", na.rm=T)
 
 # Create a plot
-ggdid(retire_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure") + 
-  labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
+cs_retire_allEHR <- ggdid(retire_cs_dyn, ylab = "Estimate and 95% CI", xlab="Relative Year", title="All Physicians") + 
+  labs(caption=paste0("p-value for pre-test of parallel trends assumption= ",round(p,3),"\n")) +
   theme(plot.caption = element_text(hjust = 0, face= "italic")) + theme_bw() +
-  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00")) + ylim(-.01,.01)
+  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") + ylim(-.01,.01)
 
-# Save the plot
-ggsave(file="CS_retire_allEHR.pdf",path="Objects")
 
 
 # OTHER RETIREMENT OUTCOMES --------------------------------------------------------------------------------
@@ -229,13 +228,12 @@ p<-retireold_cs$Wpval
 retireold_cs_dyn <- aggte(retireold_cs, type = "dynamic", na.rm=T)
 
 # Create a plot
-ggdid(retireold_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Average Effect of EHR Exposure on Retirement by Length of Exposure \nSenior Physicians") + 
-  labs(caption=paste0("\n Note: p-value for pre-test of parallel trends assumption= ",round(p,3))) +
+cs_retireold_allEHR <- ggdid(retireold_cs_dyn, xlab="\n Relative Year", theming=FALSE, legend=FALSE, title="Physicians >= 60") + 
[…]

### 2022-04-05 Scripts/sumstats.R

@@ -3,8 +3,9 @@ library(reshape)
 library(knitr)
 library(kableExtra)
 library(ggplot2)
-library(magick)
-library(webshot)
+library(readr)
+library(hexbin)
+
 
 options(knitr.kable.NA=" ")
 cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@ -130,16 +131,16 @@ knitr::kable(means_bind[c(10,9,11,12,5,7,8,6,4,2,1,3),], "latex",
 
 # EHR Info at the Physician Level (by year) -----------------------------------------------------------------
 sum_stats_year <- Physician_Data %>% filter(minyr_EHR>0) %>% group_by(year) %>%
-  summarise_at(c("Frac. Hospitals using EHR"="frac_EHR", 
-                 "Perc. Physicians Exposed to EHR"="anyEHR_exposed",
-                 "Perc. Physicians Exposed to EHR (Low Integration)"="anyEHR_LI_exposed"), list(mean), na.rm=T) 
+  summarise_at(c("Hospitals using EHR"="frac_EHR", 
+                 "Phys. Exposed to EHR"="anyEHR_exposed",
+                 "Phys. Exposed to EHR (LI)"="anyEHR_LI_exposed"), list(mean), na.rm=T) 
 
 
 # Create a dataframe out of the summary stats to put in a ggplot
 sum_stats_year <- as.data.frame(sum_stats_year)
-sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Frac. Hospitals using EHR",
-                                                                          "Perc. Physicians Exposed to EHR",
-                                                                          "Perc. Physicians Exposed to EHR (Low Integration)"))
+sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Hospitals using EHR",
+                                                                          "Phys. Exposed to EHR",
+                                                                          "Phys. Exposed to EHR (LI)"))
 sum_stats_year <- sum_stats_year %>%
   dplyr::rename("Variable"="variable")
 
@@ -150,7 +151,7 @@ ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) +
         legend.key.height = unit(.4, 'cm'),
         legend.key.width = unit(.3, 'cm'))
 
[…]

### 2022-04-05 Paper/Paper.tex

@@
 \begin{figure}[ht]
-\caption{Effect of any EHR Exposure on Retirement}
-\vspace{2mm}
     \centering
-\hspace{20mm}
-\includegraphics[scale=.45]{Objects/CS_retire_allEHR.pdf}
-\vspace{3mm}
-\newline
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_retireold_allEHR.pdf}
-        \end{minipage}}
-        \hspace{0.1cm}
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_retireyoung_allEHR.pdf}
-        \end{minipage}}
+    \caption{Effect of EHR Exposure on Retirement}
+    \includegraphics[scale=.57]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
 
-\begin{figure}[ht]
-\caption{Effect of any EHR Exposure on Likelihood of Working in Office}
-\vspace{2mm}
+\begin{figure}
     \centering
-\hspace{20mm}
-\includegraphics[scale=.45]{Objects/CS_office_ind_allEHR.pdf}
-\vspace{3mm}
-\newline
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_office_indold_allEHR.pdf}
-        \end{minipage}}
-        \hspace{0.1cm}
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_office_indyoung_allEHR.pdf}
-        \end{minipage}}
-        \label{fig:officefirst}
+    \includegraphics{obj}
+    \caption{Caption}
+    \label{fig:my_label}
 \end{figure}
-
-\begin{figure}[ht]
-\caption{Effect of any EHR Exposure on Fraction of Patients in Office}
-\vspace{2mm}
-\centering
-\hspace{20mm}
-\includegraphics[scale=.45]{Objects/CS_office_frac_allEHR.pdf}
-\vspace{3mm}
-\newline
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_office_fracold_allEHR.pdf}
-        \end{minipage}}
-        \hspace{0.1cm}
-        \fbox{\begin{minipage}[b]{0.47\linewidth}
-            \centering
-            \includegraphics[width=\textwidth]{Objects/CS_office_fracyoung_allEHR.pdf}
-        \end{minipage}}
-        \label{fig:officesecond}
-\end{figure}
-
 Another way to examine whether physicians are switching settings in a vague sense is to see whether the zip codes in which the physicians work is changing.

### 2022-04-11 Scripts/sumstats.R
update summary stats script

@@ -30,13 +30,14 @@ sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
-                 "Number of Patients"="claim_count_total","Fraction of Hospitals with EHR"="frac_EHR",
+                 "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
                  "Exposure to an EHR"="anyEHR_exposed",
                  "Exposure to an EHR (Low Integration)"="anyEHR_LI_exposed",
                  "Fraction Patients in Office"="pos_office",
                  "Ever Retire"="ever_retire",
                  "Work in an Office"="work_in_office",
-                 "Change Zip Codes"="change_zip"), 
+                 "Change Zip Codes"="change_zip",
+                 "Claim Count"="claim_count_total"), 
                list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))), na.rm=TRUE) %>%
   mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
   gather(key=var,value=value) %>%
@@ -47,7 +48,7 @@ sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
 
 
 
-knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
+knitr::kable(sum_stats_fullsample[c(4,9,13,2,11,3,5,6,8,1,7,10,12),],
              format="latex",
              table.envir="table",
              col.names=c("Variable","N","Mean","Std. Dev.", "Min", "Max"),
@@ -58,7 +59,7 @@ knitr::kable(sum_stats_fullsample[c(3,8,12,2,10,4,5,7,1,6,9,11),],
              align=c("l","c","c","c","c","c"),
              position="h") %>%
   kable_styling(full_width=F) %>%
-  pack_rows(index = c("Outcomes" = 5, "Treatment" = 3, "Characteristics" = 4))
+  pack_rows(index = c("Outcomes" = 6, "Treatment" = 3, "Characteristics" = 4))
 
 
 # Summary Stats of all variables by old vs. young vs. those who retire ---------------------------------------------------------
@@ -68,13 +69,14 @@ means_old <- Physician_Data %>% ungroup() %>%
[…]

### 2022-04-11 Paper/Paper.tex

@@
 
-\begin{figure}
+\begin{figure}[ht]
     \centering
-    \includegraphics{obj}
-    \caption{Caption}
-    \label{fig:my_label}
+    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
+    \includegraphics[scale=.57]{Objects/officefrac_plot.pdf}
+    \label{fig:officefirst}
 \end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
+    \includegraphics[scale=.57]{Objects/officeind_plot.pdf}
+    \label{fig:officesecond}
+\end{figure}
+
 Another way to examine whether physicians are switching settings in a vague sense is to see whether the zip codes in which the physicians work is changing.
 Estimates for this outcome variable are shown in
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
+    \includegraphics[scale=.57]{Objects/zip_plot.pdf}
+    \label{fig:zip}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Patient Count}
+    \includegraphics{Objects/patient_plot.pdf}
+    \label{fig:patient}
+\end{figure}
+
 \subsection{Billing Activity}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Claim Count}
+    \includegraphics{Objects/claim_plot.pdf}
+    \label{fig:claim}
+\end{figure}
 

### 2022-04-11 Paper/Paper.tex

@@
 Labor market changes for one job type may have rippling effects on populations serviced by that industry.
-It is straightforward to see this for a physician since choices and labor markets directly affect patient accessibility and quality of care received.
-For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
-Physician labor markets are an important topic of study because of their implications on care.
-I investigate physician labor market changes as a result of a major technology change in the U.S.
-healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.
+For example, physician labor markets directly affect patient accessibility and quality of care received.
+Physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
+Thus, a shock to physician labor markets have potential implications for healthcare.
+I investigate physician labor market changes as a result of a major technology shift in the U.S.
+healthcare system, the implementation of electronic health records, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.
 
@@
 Such expectations led the U.S.
-government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2009 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether.
-Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.
-Treating EHR implementation in hospitals as exogenous to an individual physician, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and amount of patients seen.
+The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings.
+However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
+Treating EHR implementation in hospitals as exogenous to an individual physician, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, patients seen, and billing activity.
 
 Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians connected to hospitals.
-The data captures the extent to which the physician in exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
-I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician.
-Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals on two dimensions: working in any hospital that implements an EHR, and working in a low-integrated hospital that implements an EHR.
-A key assumption that underlies the analysis is that EHR exposure is exogenous to physicians.
-While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decisions.
-This is the reason I include a second definition of EHR exposure that only considers hospitals where vertical integration between physicians and the hospital are low, as this indicates a much lower probability that EHR implementation is endogenous.
-In the Appendix I consider different thresholds of exposure as treatment variables for robustness.
+The data captures whether a physician in exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
+I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), (3) patients seen, and (4) billing activity.
+Common to each specification is the use of a binary treatment variable capturing a physician's EHR exposure in hospitals.
+In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
+A key assumption that underlies the main analysis is that exposure is exogenous to individual physicians.
+That is, the physician does not choose their own EHR exposure.
+While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making.
+For this reason, I include a second definition of EHR exposure that only considers exposure in hospitals where vertical integration between physician and hospital is low, as this indicates a much lower probability that EHR implementation is endogenous.
 
-Retirement typically takes place late in life, which is no different for physicians.
-The average physician retirement age is 65 (\cite{collier2017challenges}) However, the data used in my analysis reveals that even physicians as young as 35 are ``retiring".
-Young physicians are likely not retiring in the formal sense, but leaving a clinical setting for a different type of work.
-Alternative careers include teaching, consulting, or medical malpractice expert.
 I do not observe granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether.
-Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring.
-Similarly, younger physicians were .0025 ppts more likely to retire when exposed to EHRs.
-Whether physicians are young or old, these results suggest that access to care issues may have been exemplified as EHRs were implemented, and there was a surge of experienced physicians who left the labor force and were no longer available to mentor those learning under them.
+Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
+Younger physicians were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
+Whether physicians are young or old, these results suggest that access to care issues may have been exemplified as EHRs were implemented, and there was a surge of experienced physicians who left the labor force and were no longer available to influence younger physicians learning under them.
 
-Not everyone chooses to forego clinical work altogether.
-However, there may be margin to change physical place of work if physicians are not satisfied with their experience at a given hospital, whether it is because of burnout, loss of autonomy, or other factors.
+Most physicians did not choose to forego clinical work altogether.
+However, changing physical place of work might be an option if a physician is not satisfied with their current experience at a given hospital, whether it is because of burnout, loss of autonomy, or other factors.
 I investigate whether physicians switch from primarily hospital based work to office based settings, or more generally switch primary zip codes.
-I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR implementation on switching work settings.
-I find that younger physicians are more likely to work in an office in the year after being exposed to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
-Further, \textcolor{red}{add result about zip code switching}.
-This implies that physicians may incur significant switching costs in order to avoid a mismatch of preferences rooted in burnout or loss of autonomy.
-This is important as policymakers attempt to regulate and homogenize healthcare.
+I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR exposure on switching work settings.
+I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
+Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes.
+This implies that the technology imposed a large enough costs to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy.
 
-Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen, and overall billing activity by the physician.
+Finally, I investigate the effect of the technology implementation on work activity, measured by the number of patients seen, and overall billing activity by the physician.
+The subtle difference between the two measures represents an important nuance, that an increase in billing activity in itself could actually signify an increase in wastefulness as opposed to productivity if physicians are simply recording more in the EHR system.
+I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
+The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.
 
-\textcolor{red}{Add a paragraph discussing sensitivity analysis or mechanisms.}
+There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect.
+Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings.
+One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).
 
-This paper is at the intersection of both health economics and labor economics and thus contributes to our understanding of multiple important branches of literature.
-There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH act taking effect.
-Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found less impactful results, that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{dranove2014trillion}).
-
-I both expand and improve upon this literature.
-First, I consider a different mechanism by which EHRs could indirectly affect patients through physicians' response to them.
-This is the first study to my knowledge that connects EHR implementation to physician labor market decisions.
+I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through physicians' response to them.
+This is the first study to my knowledge that connects EHR implementation to various physician labor market decisions.
 Further, I improve on the literature by considering the primary time period in which EHRs were rapidly implemented.
@@
 Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in a typical two way fixed effects setup.
-
-The interaction between labor markets and technology implementation are of major interest in labor economics.
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
-    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
-    \caption*{Source: HealthIT.gov}
-    \label{fig:meanuse}
-\end{figure}
-
 While EHRs were widely utilized in hospitals by 2015, physicians still continue to report frustration.
@@
 The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication.
-Appendix (\textcolor{red}{section}) includes a computer screen image of what a physician may see when looking at EHRs.
-Because of the added administrative burden, "some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).
+Appendix () includes a computer screen image of what a physician typically sees when looking at an EHR.
 
@@
 This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.
-\textcolor{red}{I explore other thresholds in the appendix.}
 
@@
 In words, retire is equal to one in the first year that a physician has no claim counts at any point in the future.
+I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement.
 Alternatively, I could also define retirement using future number of patients.
@@
 I distinguish between the two because it could be thd case that productivity does not actually increase, but EHRs could encourage excessive billing of things that would not have been billed for otherwise.
-\textcolor{red}{put distribution graphs here?}
 
@@
 
-\import{Table Code}{overall stats.tex}
-
 I also include a table of means comparing physicians younger than 60, older than 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}.
@@
 
-\import{Table Code}{split stats.tex}
-
-In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time prior to dropping the never-treated observations.
-In 2009, 26\% of the sample of physicians were exposed to an EHR.
+In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time..
+In 2009, 32\% of the sample of physicians were exposed to an EHR.
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
-By 2015, 84\% of physicians had exposure to an EHR in at least one hospital in their network.
+Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed from 2015-2017.
 Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates.
-\textcolor{red}{i need to either fix or delete this}
-
-\begin{figure}[ht]
-\centering
-    \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.6]{Objects/sum_stats_year.pdf}
-    \label{fig:treatmentgraph}
-\end{figure}
+When I utilize this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation.
+The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time.
+Not only are physicians gradually becoming exposed to EHRs, but their exposure is actually increasing over time.
 
@@
 
-As is usual in an event study framework, I assume a version of parallel trends based on not-yet-treated units.
+As is usual in a difference in difference framework, I assume a version of parallel trends based on not-yet-treated units.
 I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods.
+This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession.
+The time period I study, 2009-2017, is reassuringly stable.
 
@@
 It is not uncommon in developed countries for workers to plan for formal retirement well in advance.
-An exogenous shock to a worker is not expected to change retirement age because of the large sum of money necessary to formally leave the labor force.
-However, physicians typically make this decision differently than most other workers.
+An exogenous shock to a worker is not expected to change retirement age because of the amount of wealth and planning necessary to formally leave the labor force.
+However, physicians typically make this decision differently than workers in other industries.
 Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
 When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career.
-This decision to delay retirement is not financial, but altruistic.
+The decision to delay retirement is not financial, but altruistic.
 Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
@@
 An exogenous shock such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock.
-Media articles suggest through interaction with physicians that the implementation of EHRs led to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
-The first aim of this section is to test that hypothesis.
+Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
+The first aim of this section is to test that hypothesis, since this claim implies potential changes to patients' access to care and to young physicians' opportunities to learn under experienced physicians.
 
-In Figure \ref{fig:retirefirst}, I present the aggregated group time treatment effects of being exposed to an EHR in any hospital for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
-The top graph suggests that for any age physician, being exposed to an EHR leads to an almost .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
-While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, which is only 2 out of every hundred.
+In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
+The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
+While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .02.
 Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure.
-It would be alarming to see any huge effects on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
-These estimates are noisy, however, given the rarity of retirement in the sample the precision does indicate a positive effect occurring.
+It would be alarming to see numerically large effects on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
+While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that retirement did occur as a result of EHR implementation.
 
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.57]{Objects/retire_plot.pdf}
-    \label{fig:retirefirst}
-\end{figure}
-
-Now, I examine how these result differ when considering different age groups of physician.
+Now, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
 These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure.
-That is, a physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean, whereas younger physicians are far less likely, if at all, to retire due to EHR exposure.
+A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
+In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure.
 Physicians in the young category have a slightly increased likelihood of retiring in the second year after exposure relative to the first year.
-While I do not have data to test this, if young physicians are career switching then it may take an additional year to prepare for that decision relative to retirement.
+While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.
 
@@
 However, there are other ways physicians might change their labor market behavior under sufficient incentive to avoid the technology.
-One such response is to shift from working in hospitals to working in office based settings, or to shift towards different hospitals.
-The implication of moving in this context may indicate something subtle about physician incentives, that costs may be incurred to avoid some negative aspects of EHRs such as loss of autonomy or burdensome administrative tasks.
-
-I measure this outcome in multiple ways.
-First, I investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure.
-I do this using variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting.
+I consider the decision to switch work settings in multiple different outcomes.
+First, I directly investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure.
+I do this using variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
+This analysis does not include physicians who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
 
@@
 There is no evidence that older physicians are shifting towards offices.
-This is may be due to the group of senior physicians already being established in some capacity in offices prior to EHR exposure.
-This is supported by the significantly larger portion of senior physicians working in offices relative to younger physicians.
+This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more "stuck" to their particular work environment.
 
-Then senior physicians may be more affected on an internal margin, the fraction of patients seen in an office relative to hospital or other settings.
+If senior physicians are already established in offices in some capacity, they may be more affected on an internal margin, the fraction of patients seen in an office relative to hospital or other settings.
 The estimates for this outcome are shown in Figure \ref{fig:officesecond}.
-There is no evidence of an effect on the fraction of patients seen in an office, neither positive nor negative.
-There only seems to be a downward trend in fraction of patients seen in an office over time.
-\textcolor{red}{make sure to try this without defining fraction of patients as 0}
+There is no evidence that EHR exposure affects the fraction of patients seen in office settings.
+Of course, this could mean multiple things.
+If total number of patients remains constant, then this implies that the absolute number of patients seen in offices also is unchanged.
+However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in a similar direction.
+I revisit the implications of this result in the next subsection where I examine the total number of patients as an outcome variable.
 
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
-    \includegraphics[scale=.57]{Objects/officefrac_plot.pdf}
-    \label{fig:officefirst}
-\end{figure}
+A much more general way to examine  physician switching behavior is to investigate whether the zip codes in which the physicians work is changing.
+Again, I exclude physicians who ever retire.
+The outcome variable is equal to 1 if the list of primary zip codes (up to twelve of them) changes.
+This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes.
+While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation.
 
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.57]{Objects/officeind_plot.pdf}
-    \label{fig:officesecond}
-\end{figure}
+The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}.
+The graph behaves similarly to the probability of working in an office in that the probability of switching zip codes increases and the result is driven solely by the younger demographic.
+However, this result is meaningfully larger; a physician is .025 ppts more likely to switch zip codes in the same year as EHR exposure and the year after, a 31.25\% increase relative to the mean.
+This result suggests that EHRs are imposing a cost high enough for physicians to exhibit significant switching behaviors, whether that is towards offices or other hospitals.
 
-Another way to examine whether physicians are switching settings in a vague sense is to see whether the zip codes in which the physicians work is changing.
-Estimates for this outcome variable are shown in
-
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.57]{Objects/zip_plot.pdf}
-    \label{fig:zip}
-\end{figure}
-
-\subsection{Physician Productivity}
+\subsection{Patient Count}
 
 An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users.
-Productivity, in this case, has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
+I consider productivity and efficiency as interchangeable, the interest here is the number of patients a physician can see in a given amount of time.
+Efficiency does not necessarily indicate quality if the amount of time spent is not optimal to the patient a care needs.
+Nevertheless, healthcare in the U.S.
+can be wasteful relative to the resources used to produce it.
+I make no claims about whether the number of patients seen is optimal, I simply explore whether the number of patients seen changes as a result of EHR exposure since a primary intention of the technology is to improve efficiency.
+
+In this case, efficiency has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
 Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
 That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
-To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have patients in hospitals even after EHR exposure.
+To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have a positive number of patients in hospitals even after EHR exposure.
 
@@
 In this case, the productivity gain could be from having the data assistant instead of EHR use.
-I investigate this in the Sensitivity Analyses below.
+I investigate this in the sensitivity analysis below.
 
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics{Objects/patient_plot.pdf}
-    \label{fig:patient}
-\end{figure}
+The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
+For physicians of any age, patient count increases by 11.5 patients in the year of EHR exposure, 24 patients in the year after exposure, 15.3 patients in the second year after exposure, and then no effect after the second year.
+These results translate to a .5-1\% increase in the number of patients seen relative to the mean.
+As with other outcomes, this result is totally due to changes by the younger group of physicians.
+Senior physicians, who are on average likely more averse to technology, are not gaining in efficiency due to EHR use whereas younger physicians, who may pick up the technology quicker, experience efficiency gains immediately.
+
+This outcome directly relates to the outcome in the previous section, the fraction of patients seen in an office setting.
+For younger physicians, since the total number of patients increases in the first couple of years after exposure, the absolute number of patients seen in offices increases as well.
+That is, it seems that younger physicians are shifting a positive amount of work towards offices but still able to increase productivity in the time spent in the hospital, thus the null effect on fraction of office patients.
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \caption{Effect of EHR Exposure on Claim Count}
-    \includegraphics{Objects/claim_plot.pdf}
-    \label{fig:claim}
-\end{figure}
+The way that physicians bill for claims is of extreme interest to researchers because incentives to report certain claims do not always line up with optimizing both outcomes and efficiency of care.
+EHRs have the potential to change claim incentives if the administrative costs of billing decrease or if EHRs encourage billing for more items than otherwise would have been recorded.
+
+On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year.
+Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count.
+Thus, billing activity increases proportionally less than average.
+EHRs do not seem to be incentivizing physicians to bill more than they would have previously per patient.
 
@@
 Using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
-Figure \textcolor{red}{number} is a frequency plot of the year of activation for every tax code that falls in the categories listed above.
+Figure \ref{fig:dataassistant_histogram} is a frequency plot of the year of activation for every tax code that falls in the categories listed above.
 The total number of these registered employees is 875.
-The first data assistants ever registered were in 2005, when electronic medical records were being implemented.
-From 2005 to 2013, more 15-60 more employees were registered each year.
-The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occured in 2014.
+The first data assistants ever registered were in 2005.
+From 2005 to 2013, 15-60 more employees were registered each year.
+The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
 If hiring data assistants were directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians were first exposed to EHRs.
-
-\vspace{5mm}
-\begin{figure}[ht]
-\centering
-\caption{Frequency of Data Assistant Enumeration by Year}
-\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
-    \label{fig:dataassistant_histogram}
-\end{figure}
 
@@
 Using an event study framework, I find that physicians of all ages are more likely to stop clinical work altogether when exposed to an EHR in a hospital, where physicians of retirement age have an even larger increase relative to younger physicians.
-Further, I find that younger physicians are more likely to work in an office based setting after EHR exposure, and physicians who remain in the labor force increase both number of patients seen and billing activity.
+Further, I find that younger physicians are more likely to work in an office based setting after EHR exposure, and physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.
 
-The implications of these results are blank.
-First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure.
-In a field where younger physicians learn a lot by watching those more experienced than them, this may particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
+The implications of these results are threefold.
+First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care.
+In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
 Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid burdens externally placed on them.
@@
 
-This paper provides an empirical foundation in a specific occurrence of the interaction between labor markets and technology.
+This paper contributes to our understanding of health information technology by providing empirical evidence of the theoretically ambiguous physician response to electronic health records being rapidly implemented in hospitals.
+This contribution is relevant to EHR vendors, hospitals, and policymakers who are involved in further implementation of complex technology.
+
+\section{Tables and Figures}
+
+\begin{figure}[htp]
+    \centering
+    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
+    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
+    \caption*{Source: HealthIT.gov}
+    \label{fig:meanuse}
+\end{figure}
+
+\import{Table Code}{overall stats.tex}
+
+\import{Table Code}{split stats.tex}
+
+\begin{figure}[p]
+\centering
+    \caption{Treatment Variables Over Time}
+    \includegraphics[scale=.55]{Objects/sum_stats_year.pdf}
+    \label{fig:treatmentgraph}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Retirement}
+    \includegraphics[scale=.57]{Objects/retire_plot.pdf}
+    \label{fig:retirefirst}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
+    \includegraphics[scale=.57]{Objects/officeind_plot.pdf}
+    \label{fig:officefirst}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
+    \includegraphics[scale=.57]{Objects/officefrac_plot.pdf}
+    \label{fig:officesecond}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
+    \includegraphics[scale=.57]{Objects/zip_plot.pdf}
+    \label{fig:zip}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Patient Count}
+    \includegraphics[scale=.57]{Objects/patient_plot.pdf}
+    \label{fig:patient}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Claim Count}
+    \includegraphics[scale=.57]{Objects/claim_plot.pdf}
+    \label{fig:claim}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\caption{Frequency of Data Assistant Enumeration by Year}
+\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+\label{fig:dataassistant_histogram}
+\end{figure}
+
+\clearpage
 
@@
 
-\newpage
-
-\appendix
-
-\section{}
-
 \end{document}

### 2022-04-14 Scripts/analysis.R
add graph of hospital adoption over time

@@ -739,6 +739,586 @@ ggsave(claim_plot,filename="Objects/claim_plot.pdf", width=9.08, height = 11.4, 
 
 
 
+##### APPENDIX: SECONDARY TREATMENT VARIABLE #####################################################################
+##################################################################################################################
+#################################################################################################################
+##################################################################################################################
+
+
+retire_cs <- att_gt(
+  yname = "retire",                # LHS Variable
+  gname = "minyr_EHR_int",             # First year a unit is treated. (set to 0 if never treated)
+  idname = "DocNPI",               # ID
+  tname = "year",                  # Time Variable
+  # xformla = NULL                 # No covariates
+  xformla = ~grad_year,            # Time-invariant controls
+  data = dplyr::filter(
+    Physician_Data,minyr_EHR_int>0),   # Remove never-treated units
+  # data = Physician_Data
+  est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
+  control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
+  clustervars = "DocNPI",          # Cluster Variables          
+  anticipation=0                   # can set a number of years to account for anticipation effects
+)
+# Save p-value for pre-trends to put in footnote of table
+p<-retire_cs$Wpval
+
+# Aggregate the effects
+retire_cs_dyn <- aggte(retire_cs, type = "dynamic", na.rm=T)
+
+# Create a plot
+cs_retire_allEHR <- ggdid(retire_cs_dyn, ylab = "Estimate and 95% CI", xlab="Relative Year", title="All Physicians") + 
+  labs(caption=paste0("p-value for pre-test of parallel trends assumption= ",round(p,3),"\n")) +
+  theme(plot.caption = element_text(hjust = 0, face= "italic")) + theme_bw() +
+  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") + ylim(-.01,.01)
+
+
+
+# OTHER RETIREMENT OUTCOMES --------------------------------------------------------------------------------
+
+# Senior Physicians
+retireold_cs <- att_gt(
[…]

### 2022-04-14 Scripts/sumstats.R

@@ -23,6 +23,8 @@ cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"
 # Read in Final_Pairs_Variables.rds
 Physician_Data <- read_rds(paste0(created_data_path,"Physician_data.rds"))
 
+
+
 # General Summary Stats Tables:  ----------------------------------------------------------------
 
 # Physician Level
@@ -267,3 +269,123 @@ ggplot(filter(Physician_Data, num_hospitals>1), aes(x=minyr_EHR)) + geom_histogr
 
 
 
+
+
+## Create hospital level implementation graph ---------------------------------------------------
+
+AHAmainsurvey <- read_csv(paste0(raw_data_path,"AHA_mainsurvey.csv"))
+phys_hosp_pairs <- readRDS(paste0(created_data_path,"phys_hosp_pairs.rds"))
+AHANPI_cw <- read_rds(paste0(created_data_path,"AHANPI_cw.rds"))
+
+AHAmainsurvey <- AHAmainsurvey %>%
+  mutate(ID=as.character(ID)) %>%
+  mutate(low_integration=ifelse(IPAHOS==1 | OPHOHOS==1 | CPHOHOS==1,1,0),
+         high_integration=ifelse(GPWWHOS==1 | MSOHOS==1 | ISMHOS==1 | EQMODHOS==1,1,0)) %>%
+  dplyr::rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
+         joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
+  select(-IPAHOS, -OPHOHOS, -CPHOHOS, -GPWWHOS, -MSOHOS, -ISMHOS, -EQMODHOS) %>%
+  select(-IPAP, -GPWP, -OPHP, -CPHP, -ISMP, -EQMP, -FNDP, -full_year, -EHLTHI, -EHLTHS,
+         -total_physicians, -joint_phys, -phys_owned)
+
+
+hosp_keep <- phys_hosp_pairs %>% ungroup() %>%
+  distinct(HospNPI) %>%
+  mutate(keep=1) 
+
+AHANPI_cw <- AHANPI_cw %>%
+  mutate(NPI=as.character(NPI))
+
+hosp_keep <- hosp_keep %>%
+  left_join(AHANPI_cw,by=c("HospNPI"="NPI")) %>%
[…]

### 2022-04-14 Paper/Paper.tex
Updates from Overleaf

@@
 
-The interaction between technological innovation and labor markets has a direct impact across industries; job characteristics such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}).
-Labor market changes for one job type may have rippling effects on populations serviced by that industry.
-For example, physician labor markets directly affect patient accessibility and quality of care received.
-Physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations on physicians such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}).
-Thus, a shock to physician labor markets have potential implications for healthcare.
-I investigate physician labor market changes as a result of a major technology shift in the U.S.
-healthcare system, the implementation of electronic health records, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.
+\section{Introduction}
+
+How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries.
+For example, industrial robots decreased the share of hours worked in low-skilled workers from 1993-2007, but increased overall labor productivity  (\cite{graetz2018robots}), and even low-skilled workers in Norway who were not displaced by robots, but saw other workers displaced, experienced lower job satisfaction due to fear of displacement (\cite{schwabe2020automation}).
+Moreover, shocks to an existing labor market could have downstream effects on customers in that market.
+For example, technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health.
+In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S.
+health care system, the implementation of electronic health records.
 
@@
 since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
-The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency.
+Researchers and government officials expected the movement towards digitization to affect efficiency in health care substantially.
 A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
@@
 
-Physicians, as primary users of this technology, play a meaningful role in whether potential benefits of EHRs are realized.
+Physicians, as primary users of this technology, play an important role in whether potential benefits of EHRs are realized.
 With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly.
 In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
-Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
+Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
 The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Treating EHR implementation in hospitals as exogenous to an individual physician, I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, patients seen, and billing activity.
+Using a difference-in-differences research design, I estimate the effects of hospital EHR implementation on physician labor market outcomes.
 
-Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians connected to hospitals.
-The data captures whether a physician in exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
-I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), (3) patients seen, and (4) billing activity.
-Common to each specification is the use of a binary treatment variable capturing a physician's EHR exposure in hospitals.
+My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in a hospital.
+The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
+I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.
+For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals.
 In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
+
+Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
+Younger physicians (less than 60) were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
+Whether physicians are young or old, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
+In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
+Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes.
+This implies that the technology imposed a large enough cost to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy.
+Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
+The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.
+
 A key assumption that underlies the main analysis is that exposure is exogenous to individual physicians.
@@
 While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making.
-For this reason, I include a second definition of EHR exposure that only considers exposure in hospitals where vertical integration between physician and hospital is low, as this indicates a much lower probability that EHR implementation is endogenous.
+For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a much lower probability that EHR implementation is endogenous.
+The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.
 
-I do not observe granular details of what physicians do when they drop out of clinical settings, so I define retirement vaguely as the decision to cease clinical work altogether.
-Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
-Younger physicians were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
-Whether physicians are young or old, these results suggest that access to care issues may have been exemplified as EHRs were implemented, and there was a surge of experienced physicians who left the labor force and were no longer available to influence younger physicians learning under them.
-
-Most physicians did not choose to forego clinical work altogether.
-However, changing physical place of work might be an option if a physician is not satisfied with their current experience at a given hospital, whether it is because of burnout, loss of autonomy, or other factors.
-I investigate whether physicians switch from primarily hospital based work to office based settings, or more generally switch primary zip codes.
-I limit the sample to any physicians who did not retire and use the same event study framework to estimate the effect of EHR exposure on switching work settings.
-I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
-Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes.
-This implies that the technology imposed a large enough costs to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy.
-
-Finally, I investigate the effect of the technology implementation on work activity, measured by the number of patients seen, and overall billing activity by the physician.
-The subtle difference between the two measures represents an important nuance, that an increase in billing activity in itself could actually signify an increase in wastefulness as opposed to productivity if physicians are simply recording more in the EHR system.
-I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
-The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.
-
+My analysis contributes to the literature examining effects of health information technology on health care.
 There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect.
@@
 
-\section{Institution}
+\section{Background on Electronic Health Records}
 
 EHRs have been an important feature of healthcare since the 1980s and 1990s.
-Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and/or scheduling.
-Prior to computerization, physicians were not able to interface directly with EHRs, and thus were not drastically affected by their implementation.
-The cost of utilizing an EHR was large and increasing in the amount of capabilities of the system, which led to the widespread view that EHRs were strictly complements to paper records, not a feasible substitute.
+Early in its existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling.
+Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation.
 As computers were made more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
@@
 Automation in data entry was non-existent, making it extremely time consuming for the user.
-Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry.
+Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry (\cite{evans2016electronic}).
 
-The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines.
+The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines.
 The guidelines include having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy.
-Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies.
-Hospitals began receiving payments for the first stage of meaningful use in 2011.
-This resulted in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
-Figure \ref{fig:meanuse} shows a comparison of U.S.
+The U.S.
+allocated subsidies according to stages of "meaningful" use: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, and stage 3 (2014) extended to data sharing between practices.
+This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
+Figure \ref{fig:meanuse} shows a geographical comparison of U.S.
 hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
@@
 
-While EHRs were widely utilized in hospitals by 2015, physicians still continue to report frustration.
+Figure \ref{fig:EPIC} shows a computer screen image of what a physician typically sees when viewing an EHR.
+
+While most hospitals fully implemented EHRs by 2015, physicians still continue to report frustration.
 For every patient seen, a primary care physician spends an average of 16 minutes actively using an EHR, an amount of time that often exceeds face to face time with the patient (\cite{overhage2020physician}).
 The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication.
-Appendix () includes a computer screen image of what a physician typically sees when looking at an EHR.
+Figure \ref{fig:EPIC} includes a computer screen image of what a physician typically sees when looking at an EHR.
 
@@
 
-Using various linked datasets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time and other relevant characteristics.
-The different data sets used to construct the panel are described below.
+Using various linked data sets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, outcomes related to physician practice location and billable activity, and other relevant physician, hospital, and practice characteristics.
+I describe the different data sets used to construct the panel below.
 
@@
 
-Medicare Data on Provider Practice and Specialty is a database of physicians which details claim counts and specialty from 2009-2017.
-The relevant information included is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
+The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017.
+Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
 I first limit the sample of physicians based on specialty type.
-Primary care physicians are exposed to the administrative burden of EHRs in hospitals more-so than specialists.
-Thus, I only include physicians whose primary specialty is reported as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years, who likely do not have strong ties to the hospitals they have limited work in.
+My analysis relies on EHR implementation in hospitals.
+Thus, I only include physicians reported their primary specialty as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
 
-This data includes individual labor market information on physicians which I use to construct the dependent variables for my analysis.
-Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen and overall billing activity.
-These outcome variables and their construction are described in detail below.
+This data includes information on physicians which I use to construct the dependent variables for my analysis.
+Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen, and overall billing activity.
+I describe these outcome variables and their construction in detail below in Section \ref{sec:outcome}.
 
-\subsection{Shared Patient}
+\subsection{Shared Patients}
 
-CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicaid in the same day.
-For example, if a Medicaid patient is referred to a specialist by a primary care physician and the specialist is seen by the patient in the same day, then those two physicians have that shared patient in common.
+CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicare.
+For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common.
+The number of shared patients are collected in 30, 90, or 180 day intervals.
+I focus on the number of shared patient billing two entities in the same day, within a 30 day interval.
 I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with.
 This measure of exposure defines various treatment variables for my analysis.
-I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
-I care specifically about primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital.
-The types of physicians are already limited to primary care, but there still may be some office-based primary care physicians in the shared patient data.
+I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
+I am specifically interested in primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital.
+The types of physicians are already limited to primary care, but there still may be office-based primary care physicians in the shared patient data.
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
@@
 
-Using hospital NPI, I link the pairs to the American Hospital Association survey, which contains information on hospital-level EHR use and other characteristics.
+\subsection{AHA Survey Data}
+
+Using hospital NPI, I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
 I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
@@
 
+\subsection{Physician Compare}
+
 Finally, I include a physician's graduation year from Physician Compare.
@@
 
-\subsection{Outcome Variables}
+\subsection{Outcome Variables}\label{sec:outcome}
 
@@
 where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$.
-In words, retire is equal to one in the first year that a physician has no claim counts at any point in the future.
+Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future.
 I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement.
@@
 I measure this in the data in three ways.
-I consider the fraction of patients a physician sees in an office based setting, which is directly available in the data.
-Further, I use this variable to construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office based setting.
-There is less variation in this data but allows insight to a different type of decision making where physicians may enter or leave office based setting based on EHR exposure.
-Lastly, I utilize the zip code information to define an indicator variable equal to one only if the physician's list of zip codes changes from one year to the next.
-While these variables do not give direct insight into whether physicians specifically move from one hospital to another, they do allow for exploration of whether settings are changing as a result of EHRs in general.
+First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
+Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting.
+This binary measure of practice setting offers less variation but captures a more discrete changes in which physicians enter or leave the office-based setting entirely.
+Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next.
+While no single variable directly measures physician movement across hospitals, these three outcomes collectively capture different ways in which physicians may change their practice setting over time.
 
-Finally, I consider outcome variables which explore productivity and billing activity once EHRs are implemented in hospitals.
-Productivity is measured in the number of patients a physician sees in a give year, where the billing activity is the total claim count for the physician.
-I distinguish between the two because it could be thd case that productivity does not actually increase, but EHRs could encourage excessive billing of things that would not have been billed for otherwise.
+Finally, I consider outcome variables which explore productivity and billing activity.
+Productivity is measured as the number of patients a physician sees in a given year, where the billing activity is the total claim count for the physician.
+I distinguish between the two because it could be the case that productivity does not actually increase, but EHRs could encourage more comprehensive billing.
 
@@
 
-Summary statistics for the entire sample are shown in Table \ref{tab:sumstats1}.
-Only 2\% of physicians retire over the course of the sample, which equates to approximately 750 physicians in the sample who retire at some point.
+I show summary statistics for the entire sample in Table \ref{tab:sumstats1}.
+Only 2\% of physicians (approximately 750) retire over the course of the panel.
 A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting.
@@
 The age groups are exposed to EHRs in similar ways, but younger physicians are exposed to EHRs in low integrated setting slightly later.
-Retirement occurs just as often in the different age groups, indicating the possibility of physicians leaving a clinical setting for an administrative or other position.
 
-In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time..
+In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time.
 In 2009, 32\% of the sample of physicians were exposed to an EHR.
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
-Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed from 2015-2017.
+Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by 2015.
 Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates.
-When I utilize this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation.
+When I employ this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation.
 The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time.
-Not only are physicians gradually becoming exposed to EHRs, but their exposure is actually increasing over time.
+This shows that, along with becoming first exposed over the sample period, physicians are also seeing EHRs implemented in more of their own hospitals over time.
 
@@
 A group indicates all physicians who were treated, or first exposed to an EHR, in the same year.
-To estimate the heterogeneous treatment effects, I utilize the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
 
@@
 This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure.
-An EHR is a costly technology and requires a significant amount of work to implement.
+An EHR is a costly technology and requires a significant amount of collaboration to implement.
 A hospital does not have incentive to un-implement an EHR once it is implemented.
@@
 
-As is usual in a difference in difference framework, I assume a version of parallel trends based on not-yet-treated units.
+As is usual in a difference-in-differences framework, I assume a version of parallel trends based on not-yet-treated units.
 I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods.
@@
 The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no claim about what the physician does afterwards.
-A physician is said to retire in a given year only if it is the first year in which the physician has no future claims.
+A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
 
@@
 Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
-The first aim of this section is to test that hypothesis, since this claim implies potential changes to patients' access to care and to young physicians' opportunities to learn under experienced physicians.
+The aim of this section is to test that hypothesis.
 
@@
 With each specification, I present the p-value for a Wald test for pre-trends.
-Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed in the significantly large p-values.
+Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed statistically, where I fail to reject the null of parallel trends.
 
@@
 
-If senior physicians are already established in offices in some capacity, they may be more affected on an internal margin, the fraction of patients seen in an office relative to hospital or other settings.
-The estimates for this outcome are shown in Figure \ref{fig:officesecond}.
-There is no evidence that EHR exposure affects the fraction of patients seen in office settings.
+If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings.
+The estimates for this outcome are shown in Figure \ref{fig:officesecond}, where I find limited evidence that EHR exposure affects the fraction of patients seen in office settings.
 Of course, this could mean multiple things.
@@
 I consider productivity and efficiency as interchangeable, the interest here is the number of patients a physician can see in a given amount of time.
-Efficiency does not necessarily indicate quality if the amount of time spent is not optimal to the patient a care needs.
-Nevertheless, healthcare in the U.S.
+Importantly, my measure of efficiency does not necessarily indicate quality.
+Nevertheless, health care in the U.S.
 can be wasteful relative to the resources used to produce it.
@@
 Thus, billing activity increases proportionally less than average.
-EHRs do not seem to be incentivizing physicians to bill more than they would have previously per patient.
+EHRs do not seem to be facilitating more billable activity.
 
@@
 In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes.
-This technology has been controversial and debated as a burdensome load on physicians that may contribute to a lack of improvements in the healthcare sector.
+This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes.
 
@@
     \label{fig:meanuse}
+\end{figure}
+
+\begin{figure}[htp]
+    \centering
+    \caption{Screenshot of EPIC System}
+    \includegraphics[scale=.5]{Objects/epic-ehr-screenshot.jpg}
+    \label{fig:EPIC}
 \end{figure}
@@
 
+\clearpage
+
+\appendix
+
+\section{Alternate Definition of EHR Exposure}
+
+The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in.
+A concern that one may have is that physicians choose their individual EHR exposure, making this an endogenous treatment variable.
+The results do not indicate that this is happening on average, since we would expect physicians not to respond on the margin of retirement or switching zip codes if they are choosing to be exposed to EHRs.
+Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions.
+However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable.
+This variable is whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
+
+\subsection{Retirement}
+
+\section{Anticipation}
+
+\clearpage
+
+\section{Tables and Figures}
+
 \end{document}

### 2022-04-14 Paper/Paper.tex

@@
 I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement.
-Alternatively, I could also define retirement using future number of patients.
-I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
+\footnote{Alternatively, I could also define retirement using future number of patients.
+I use claim count for a more conservative measure of retirement, but using patient count yields identical results.}
 
@@
 They may add features or switch vendors, but do not reverse EHR use.
+This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a random sample of hospitals over time.
 Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR.
@@
 Second, I assume physicians do not anticipate EHR exposure prior to occurrence.
-Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until years later.
+Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until a future period.
 Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
@@
 
-It is not uncommon in developed countries for workers to plan for formal retirement well in advance.
+Workers in developed countries tend to plan for formal retirement well in advance.
 An exogenous shock to a worker is not expected to change retirement age because of the amount of wealth and planning necessary to formally leave the labor force.
@@
 There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management.
-While it is unlikely that physicians under 55 are formally retiring, either old or young physicians could be career switching.
-The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no claim about what the physician does afterwards.
+While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching.
+The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
 A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
 
-An exogenous shock such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock.
+An exogenous shock to the physician such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock.
 Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
@@
 Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure.
-It would be alarming to see numerically large effects on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
-While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that retirement did occur as a result of EHR implementation.
+In this scenario, it would be alarming to see a numerically large effect on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
+While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that a positive effect on the likelihood of retirement did occur as a result of EHR implementation.
 
@@
 In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure.
-Physicians in the young category have a slightly increased likelihood of retiring in the second year after exposure relative to the first year.
+Physicians in the young category have a larger effect on the likelihood of retiring in the second year after exposure relative to the first year.
 While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.
@@
 I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
-The top panel is for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean.
+The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean.
 Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
@@
 However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in a similar direction.
-I revisit the implications of this result in the next subsection where I examine the total number of patients as an outcome variable.
+I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.
 
-A much more general way to examine  physician switching behavior is to investigate whether the zip codes in which the physicians work is changing.
+The previous results only capture switching behavior which involves office-based settings.
+A more general way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors.
 Again, I exclude physicians who ever retire.
@@
 
-\subsection{Patient Count}
+\subsection{Patient Count}\label{sec:patientcount}
 
 An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users.
-I consider productivity and efficiency as interchangeable, the interest here is the number of patients a physician can see in a given amount of time.
-Importantly, my measure of efficiency does not necessarily indicate quality.
-Nevertheless, health care in the U.S.
-can be wasteful relative to the resources used to produce it.
-I make no claims about whether the number of patients seen is optimal, I simply explore whether the number of patients seen changes as a result of EHR exposure since a primary intention of the technology is to improve efficiency.
+I measure productivity as the number of Medicare patients seen annually.
+EHRs could increase this number if they improve efficiency through less paperwork or ease of information access, or EHRs could decrease the number of patients seen if physicians are spending more time interfacing with the system and are not able to see as many patients.
 
-In this case, efficiency has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
+In this case, productivity has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
 Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
@@
 
-The way that physicians bill for claims is of extreme interest to researchers because incentives to report certain claims do not always line up with optimizing both outcomes and efficiency of care.
 EHRs have the potential to change claim incentives if the administrative costs of billing decrease or if EHRs encourage billing for more items than otherwise would have been recorded.
@@
 If hiring data assistants were directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians were first exposed to EHRs.
-
-The total number of registered data assistants may not equal the true number of people doing a data assistant job in hospitals.
-However, under the assumption that the number of registered data assistants is proportional to the number of actual data assistants, the claim of no correlation still holds.
 
@@
     \label{fig:treatmentgraph}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption{Hospital EHR Adoption Over Time}
+    \includegraphics[scale=.6]{Objects/hosp_treat.png}
+    \label{fig:hosp_treat}
 \end{figure}

### 2022-04-15 Scripts/sumstats.R
sort hospitals in figure 4 and remove low integration from summary stats tables

@@ -4,7 +4,7 @@ library(knitr)
 library(kableExtra)
 library(ggplot2)
 library(readr)
-library(hexbin)
+
 
 
 options(knitr.kable.NA=" ")
@@ -34,7 +34,6 @@ sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
                  "Age"="age",
                  "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
                  "Exposure to an EHR"="anyEHR_exposed",
-                 "Exposure to an EHR (Low Integration)"="anyEHR_LI_exposed",
                  "Fraction Patients in Office"="pos_office",
                  "Ever Retire"="ever_retire",
                  "Work in an Office"="work_in_office",
@@ -50,7 +49,7 @@ sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
 
 
 
-knitr::kable(sum_stats_fullsample[c(4,9,13,2,11,3,5,6,8,1,7,10,12),],
+knitr::kable(sum_stats_fullsample[c(4,8,12,2,10,3,5,7,1,6,9,11),],
              format="latex",
              table.envir="table",
              col.names=c("Variable","N","Mean","Std. Dev.", "Min", "Max"),
@@ -61,7 +60,7 @@ knitr::kable(sum_stats_fullsample[c(4,9,13,2,11,3,5,6,8,1,7,10,12),],
              align=c("l","c","c","c","c","c"),
              position="h") %>%
   kable_styling(full_width=F) %>%
-  pack_rows(index = c("Outcomes" = 6, "Treatment" = 3, "Characteristics" = 4))
+  pack_rows(index = c("Outcomes" = 6, "Treatment" = 2, "Characteristics" = 4))
 
 
 # Summary Stats of all variables by old vs. young vs. those who retire ---------------------------------------------------------
@@ -73,7 +72,6 @@ means_old <- Physician_Data %>% ungroup() %>%
                  "Age"="age",
                  "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
                  "Exposure to an EHR"="anyEHR_exposed",
-                 "Exposure to an EHR (Low Integration)"="anyEHR_LI_exposed",
                  "Fraction Patients in Office"="pos_office",
                  "Ever Retire"="ever_retire",
[…]

### 2022-04-15 Paper/Paper.tex

@@
 \usepackage{bbm}
-\usepackage[font=footnotesize,labelfont=bf]{caption}
+\usepackage[font=normalsize,labelfont=bf]{caption}
 %\captionsetup{justification=raggedright,singlelinecheck=false}
+\usepackage[english,dutch]{babel}
 
@@
 EHRs have have become increasingly relevant in the U.S.
-since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
+since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+In Stage 1 (2010), MU focused on data capturing and sharing.
+In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care.
+Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} (\cite{hitech}).
+President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
 Researchers and government officials expected the movement towards digitization to affect efficiency in health care substantially.
 A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-Such expectations led the U.S.
-government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2009 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
-In Stage 1 (2010), MU focused on data capturing and sharing.
-In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care.
-Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
 The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.
@@
 Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
-Younger physicians (less than 60) were .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
-Whether physicians are young or old, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
+Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
+Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
 In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
@@
 Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation.
-As computers were made more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
+As technological innovation made computers more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
 Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
@@
 
-Figure \ref{fig:EPIC} shows a computer screen image of what a physician typically sees when viewing an EHR.
+Figure \ref{fig:EPIC} shows a two computer screenshots of what physicians might see in their EHR system.
+On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of the time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}).
+These are the most time consuming portions of EHR use, according to the same study.
+In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR.
 
 While most hospitals fully implemented EHRs by 2015, physicians still continue to report frustration.
-For every patient seen, a primary care physician spends an average of 16 minutes actively using an EHR, an amount of time that often exceeds face to face time with the patient (\cite{overhage2020physician}).
-The tasks that may take place during this time include viewing or inputting patient information, ordering services such as tests or prescriptions, navigating the system, or communication.
-Figure \ref{fig:EPIC} includes a computer screen image of what a physician typically sees when looking at an EHR.
+The same functions of EHR use that take the most time are also reported to be the most frustrating (\cite{dymek2021building}).
+A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages for other members of the care team, and searching relevant information (\cite{dymek2021building}).
+Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or even longer to load.
+Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
+Physicians have been voicing concerns over EHR usability since their beginning, indicating that they are likely responding in their behavior as well.
 
@@
 My analysis relies on EHR implementation in hospitals.
-Thus, I only include physicians reported their primary specialty as hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
+Thus, I only include physicians reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
 Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
@@
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
-Specifically, a physician-hospital pair is dropped if their total number of same day billings with the hospital is less than 30 times the number of years the pair appears in the data.
+I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year.
 This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.
@@
 Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future.
-I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement.
 \footnote{Alternatively, I could also define retirement using future number of patients.
-I use claim count for a more conservative measure of retirement, but using patient count yields identical results.}
+I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
+In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}
 
@@
 By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed.
-Physicians are typically exposed in any capacity about one fourth of the way through the sample and in low-integrated hospitals about halfway through the sample.
+Physicians are typically exposed in any capacity about one fourth of the way through the sample.
 The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.
@@
 In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time.
-In 2009, 32\% of the sample of physicians were exposed to an EHR.
+In 2009, 32\% of the sample of physicians were already exposed to an EHR.
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
@@
 where $G_g=1$ for those in group $g$.
-A group indicates all physicians who were treated, or first exposed to an EHR, in the same year.
+A group indicates all physicians treated, or first exposed to an EHR, in the same year.
 To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
@@
 Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
-Thus, physicians remain treated in all future years after exposure.
+However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
+I investigate this possibility and limit the data to address it in Appendix ().
 
@@
 As is usual in a difference-in-differences framework, I assume a version of parallel trends based on not-yet-treated units.
-I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups which were treated in later periods.
+I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
 This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession.
 The time period I study, 2009-2017, is reassuringly stable.
+In my analysis, I present p-values for a Wald test of pre-trends.
+In most cases, there is no evidence of a pre-trend, however, I investigate the assumption in detail in Appendix ().
+
+An institutional assumption that I make is that physicians working in EHR-adopted hospitals are constrained to use the technology.
+My outcomes have two key determinants: the choice of physicians to learn and utilize the technology, and the realized effect of electronic health records themselves.
+The point of this paper is to identify the second determinant, and thus I assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
+That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
+
+It is reasonable to believe that physicians are fully utilizing the EHRs in the hospital they work in, since by 2015 electronic health records were widely implemented and considered unavoidable.
+An objection to this assumption may be that physicians stay in a hospital, but the hospital hires a data assistant to do technology work for the physician.
+In this case, the effects seen could be from having the data assistant instead of EHR use.
+I investigate this in Section \ref{sec:dataass}.
 
@@
 While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.
-
-A general concern in event study settings is whether pre-trends are driving the results.
-With each specification, I present the p-value for a Wald test for pre-trends.
-Visually, there are no clear indications of an increasing trend in retirement before exposure, and this is confirmed statistically, where I fail to reject the null of parallel trends.
 
@@
 
-In this case, productivity has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves.
-Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
-That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
-To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have a positive number of patients in hospitals even after EHR exposure.
-
-It is reasonable to believe that physicians are fully utilizing the EHRs in the hospital they work in, since by 2015 electronic health records were widely implemented and considered unavoidable.
-An objection to this assumption may be that older physicians stay in a hospital, but the hospital hires a data assistant to do technology work for the physician.
-In this case, the productivity gain could be from having the data assistant instead of EHR use.
-I investigate this in the sensitivity analysis below.
-
 The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
@@
 
-\section{Sensitivity Analyses}
+\subsection{Late Adopters as Comparison Group}
 
-\subsection{Data Assistants}
+In all outcomes, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
+All specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure.
+This is likely due to the composition of the hospitals adopting at different times.
+Consider time $t+4$.
+The only hospitals to analyze the effect in this period are those who adopted in 2010 compared to hospitals who adopted in 2015.
+The group of hospitals who adopted in 2015 may be compositionally different than early adopters if there is a characteristic reason for late adoption such as demographic of patients.
+For this reason, I focus discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.
 
-One concern with this measure of productivity is that an increase in the number of patients seen by a physician could be caused by an outside factor that is also correlated with EHR adoption in hospitals.
+\section{Physician EHR Avoidance}
+
+\subsection{Hospitals Hiring Data Assistants}\label{sec:dataass}
+
+One concern with this analysis is that any effect could be caused by an outside factor that is also correlated with EHR adoption in hospitals.
 A logical outside factor could be the existence of employees with the purpose of utilizing electronic health records on a physician's behalf.
@@
 The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
-If hiring data assistants were directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians were first exposed to EHRs.
+If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
+
+\subsection{Physician Influence of EHR Decision in Hospitals}
 
@@
     \centering
-    \caption{Screenshot of EPIC System}
+    \caption{Screenshot of EHR System}
     \includegraphics[scale=.5]{Objects/epic-ehr-screenshot.jpg}
+    \includegraphics[scale=.135]{Objects/EHRimage2.jpg}
     \label{fig:EPIC}
@@
 
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Retirement}
+    \includegraphics[scale=.4]{Objects/retire_plot_LI.pdf}
+    \label{fig:retire_LI}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Likelihood of Working in Office}
+    \includegraphics[scale=.4]{Objects/officeind_plot_LI.pdf}
+    \label{fig:officeind_LI}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Fraction of Patients in Office}
+    \includegraphics[scale=.4]{Objects/officefrac_plot_LI.pdf}
+    \label{fig:officefrac_LI}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Likelihood of Switching Zip Codes}
+    \includegraphics[scale=.4]{Objects/zip_plot_LI.pdf}
+    \label{fig:zip_LI}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Number of Patients}
+    \includegraphics[scale=.4]{Objects/patient_plot_LI.pdf}
+    \label{fig:patient_LI}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of Exposure in LI Setting on Billing Activity}
+    \includegraphics[scale=.4]{Objects/claim_plot_LI.pdf}
+    \label{fig:claim_LI}
+\end{figure}
+
 \end{document}

### 2022-04-15 Table Code/split stats.tex
Updates from Overleaf

@@
-\begin{table}[ht]
+\begin{table}[h]
 
@@
 \multicolumn{4}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 0.01 & 0.01 & 1.00\\
-\hspace{1em}Fraction Patients in Office & 0.44 & 0.27 & 0.15\\
-\hspace{1em}Work in an Office & 0.79 & 0.52 & 0.33\\
-\hspace{1em}Change Zip Codes & 0.06 & 0.09 & 0.08\\
-\hspace{1em}Number of Patients & 3238.30 & 2028.23 & 554.53\\
+\hspace{1em}Ever Retire & 0.02 & 0.02\\
+\hspace{1em}Fraction Patients in Office & 0.44 & 0.27 \\
+\hspace{1em}Work in an Office & 0.78 & 0.52 \\
+\hspace{1em}Change Zip Codes & 0.06 & 0.09 \\
+\hspace{1em}Number of Patients & 431.96 & 357.46 \\
+\hspace{1em}Claim Count & 3216.46 & 2019.17\\
 \addlinespace[0.3em]
@@
 \hspace{1em}Exposure to an EHR & 0.77 & 0.76 & 0.77\\
-\hspace{1em}Exposure to an EHR (Low Integration) & 0.54 & 0.50 & 0.44\\
-\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.64 & 0.64\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.63 & 0.64\\
 \addlinespace[0.3em]
 \multicolumn{4}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 61.06 & 44.06 & 48.29\\
+\hspace{1em}Age & 61.07 & 44.06 & 48.29\\
 \hspace{1em}Female & 0.15 & 0.35 & 0.39\\
 \hspace{1em}Number of Hospitals Worked With & 1.67 & 1.53 & 1.35\\
-\hspace{1em}Number of Systems Worked With & 1.41 & 1.33 & 1.19\\
+\hspace{1em}Number of Systems Worked With & 1.40 & 1.33 & 1.19\\
 \bottomrule
 \end{tabular}
+\label{tab:splitstats}
 \end{table}

### 2022-04-18 Scripts/analysis.R
update graphs and add new graphs to analysis

@@ -592,6 +592,14 @@ patient_cs <- att_gt(yname = "npi_unq_benes",
 
 p<-patient_cs$Wpval
 
+patient_cs_group <- ggdid(patient_cs, theming=FALSE, legend=FALSE, title="Results by Group") + 
+  theme_bw() + xlab("\nYear") +
+  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") +
+  geom_hline(yintercept = 0,size=.2,linetype="dashed") + 
+  theme(legend.position = "none", text=element_text(size=17,family="lm")) 
+
+ggsave(patient_cs_group, filename="Objects/patient_group.pdf", width=19, height = 15, units="in")
+
 # Aggregate the effects
 patient_cs_dyn <- aggte(patient_cs, type = "dynamic", na.rm=T)
 
@@ -766,55 +774,39 @@ ggsave(claim_plot,filename="Objects/claim_plot.pdf", width=11.4, height = 9.08, 
 ## CLAIM COUNT ONLY IN EHR HOSPITALS FOR THOSE WHO NEVER GET NEW NPI -------------------------
 
 claim2_cs <- att_gt(yname = "claim_count_total",
-                        gname = "minyr_EHR",
-                        idname = "DocNPI",
-                        tname = "year",
-                        xformla = ~grad_year,
-                        data = dplyr::filter(Physician_Data,minyr_EHR>0 & never_newnpi==1 & ever_retire==0),
-                        est_method = "dr",
-                        control_group = "notyettreated"
+                    gname = "minyr_EHR",
+                    idname = "DocNPI",
+                    tname = "year",
+                    xformla = ~grad_year,
+                    data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & never_newnpi==1 & num_hospitals==1),
+                    est_method = "dr",
+                    control_group = "notyettreated"
 )
-
+ggdid(claim2_cs)
 p<-claim2_cs$Wpval
 
 # Aggregate the effects
 claim2_cs_dyn <- aggte(claim2_cs, type = "dynamic", na.rm=T)
 
 # Create a plot
-cs_claim2_allEHR 
-ggdid(claim2_cs_dyn, theming=FALSE, legend=FALSE, title="Physicians Who Do Not Change Hospitals") + 
[…]

### 2022-04-18 Scripts/data4_physician_level.R

@@ -393,7 +393,7 @@ data <- data %>%
   left_join(EHR_hosp, by=c("year", "DocNPI")) %>%
   mutate(hosp_patient_count_EHRhosp=ifelse(is.na(hosp_patient_count_EHRhosp),0,hosp_patient_count_EHRhosp))
 
-# Create physician level patient count with EHR hospitals and merge back to data
+# Create physician level patient count with non-EHR hospitals and merge back to data
 noEHR_hosp <- data %>%
   filter(EHR==0) %>%
   group_by(year, DocNPI) %>%

### 2022-04-18 Scripts/data5_MDPPAS.R

@@ -258,7 +258,8 @@ observe <- Physician_Data %>%
 
 Physician_Data <- Physician_Data %>%
   mutate(hosp_patient_count_EHRhosp=ifelse(is.na(hosp_patient_count_EHRhosp),0,hosp_patient_count_EHRhosp),
-         hosp_patient_count_noEHRhosp=ifelse(is.na(hosp_patient_count_noEHRhosp),0,hosp_patient_count_noEHRhosp))
+         hosp_patient_count_noEHRhosp=ifelse(is.na(hosp_patient_count_noEHRhosp),0,hosp_patient_count_noEHRhosp),
+         hosp_patient_count=ifelse(is.na(hosp_patient_count),0,hosp_patient_count))
 
 Physician_Data <- Physician_Data %>%
   mutate(npi_unq_benes=ifelse(is.na(npi_unq_benes),0,npi_unq_benes))

### 2022-04-18 Scripts/sumstats.R

@@ -4,6 +4,15 @@ library(knitr)
 library(kableExtra)
 library(ggplot2)
 library(readr)
+library(showtext)
+
+font_add_google("Cormorant Garamond", "corm")
+
+font_add("lm","C:/Users/hkagele/Downloads/Latin-Modern-Roman/lmroman10-regular.otf")
+
+## Automatically use showtext to render text
+showtext_auto()
+
 
 
 
@@ -149,7 +158,8 @@ ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) +
   scale_colour_manual(values=cbbPalette) + ylim(0,1)  + xlim(2009,2015) +
   theme(legend.key.size=unit(.3,'cm'),
         legend.key.height = unit(.4, 'cm'),
-        legend.key.width = unit(.3, 'cm'))
+        legend.key.width = unit(.3, 'cm'),
+        text=element_text(size=17, family="lm"))
 
 ggsave("Objects/sum_stats_year.pdf", width=11.4, height=6, units="in")
 
@@ -381,5 +391,6 @@ AHA_sample <- AHAmainsurvey %>%
 
 
 
-panel <- panelview(AHA_sample, Y=NULL, D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE)
-ggsave("Objects/hosp_treat.png")
+panelview(AHA_sample, Y=NULL, D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE) +
+  theme(text=element_text(size=50,family="lm"))
+ggsave("Objects/hosp_treat.png", width=11.4, height=15)

### 2022-04-18 Paper/Paper.tex

@@
 %\captionsetup{justification=raggedright,singlelinecheck=false}
-\usepackage[english,dutch]{babel}
 
@@
 
-Now, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
+Next, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
 These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure.
@@
 
-\subsection{Patient Count}\label{sec:patientcount}
+\subsection{Patient Count and Billable Activity}\label{sec:patientcount}
 
 An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users.
-I measure productivity as the number of Medicare patients seen annually.
-EHRs could increase this number if they improve efficiency through less paperwork or ease of information access, or EHRs could decrease the number of patients seen if physicians are spending more time interfacing with the system and are not able to see as many patients.
+A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+Such a decrease implies a change in patient care: either more patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
+In this section, I estimate the effect of EHR implementation on a physician's number of patients seen.
 
@@
 As with other outcomes, this result is totally due to changes by the younger group of physicians.
-Senior physicians, who are on average likely more averse to technology, are not gaining in efficiency due to EHR use whereas younger physicians, who may pick up the technology quicker, experience efficiency gains immediately.
+While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is economically trivial.
+
+Since this is a demand-driven variable as well, I investigate whether any external factors may shift the demand for health care over the sample period.
 
@@
 That is, it seems that younger physicians are shifting a positive amount of work towards offices but still able to increase productivity in the time spent in the hospital, thus the null effect on fraction of office patients.
-
-\subsection{Billing Activity}
 
@@
     \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.57]{Objects/retire_plot.pdf}
+    \includegraphics[scale=.4]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.57]{Objects/officeind_plot.pdf}
+    \includegraphics[scale=.4]{Objects/officeind_plot.pdf}
     \label{fig:officefirst}
@@
     \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
-    \includegraphics[scale=.57]{Objects/officefrac_plot.pdf}
+    \includegraphics[scale=.4]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
     \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.57]{Objects/zip_plot.pdf}
+    \includegraphics[scale=.4]{Objects/zip_plot.pdf}
     \label{fig:zip}
@@
     \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.57]{Objects/patient_plot.pdf}
+    \includegraphics[scale=.4]{Objects/patient_plot.pdf}
     \label{fig:patient}
@@
     \caption{Effect of EHR Exposure on Claim Count}
-    \includegraphics[scale=.57]{Objects/claim_plot.pdf}
+    \includegraphics[scale=.4]{Objects/claim_plot.pdf}
     \label{fig:claim}

### 2022-04-18 Paper/Paper.tex

@@
 
-An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users.
+An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive.
 A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
-Such a decrease implies a change in patient care: either more patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
-In this section, I estimate the effect of EHR implementation on a physician's number of patients seen.
+Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
+Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.
+In this section, I address EHRs' effect on productivity by estimating its effect on a physician's number of patients seen.
 

### 2022-04-18 Scripts/analysis.R
add limited sample plot

@@ -664,7 +664,7 @@ cs_patientyoung_allEHR <- ggdid(patientyoung_cs_dyn, theming=FALSE, legend=FALSE
   theme(legend.position = "none", text=element_text(size=17,family="lm")) +ylim(-50,50)
 
 
-# Create office ind plot 
+# Create plot 
 patient_plot <- ggarrange(
   cs_patient_allEHR,
   ggarrange(cs_patientyoung_allEHR, cs_patientold_allEHR,
@@ -676,6 +676,32 @@ patient_plot <- ggarrange(
 )
 
 ggsave(patient_plot,filename="Objects/patient_plot.pdf", width=11.4, height = 9.08, units="in")
+
+
+# PATIENT COUNT LIMITED SAMPLE
+patient2_cs <- att_gt(yname = "npi_unq_benes",
+                    gname = "minyr_EHR",
+                    idname = "DocNPI",
+                    tname = "year",
+                    xformla = ~grad_year,
+                    data = dplyr::filter(Physician_Data,minyr_EHR>0 & ever_retire==0 & never_newnpi==1 & num_hospitals==1),
+                    est_method = "dr",
+                    control_group = "notyettreated"
+)
+ggdid(patient2_cs)
+p<-patient2_cs$Wpval
+
+# Aggregate the effects
+patient2_cs_dyn <- aggte(patient2_cs, type = "dynamic", na.rm=T)
+
+# Create a plot
+cs_patient2_allEHR <- ggdid(patient2_cs_dyn, theming=FALSE, legend=FALSE, title="Patient Count") + 
+  labs(caption=paste0("p-value= ",round(p,3))) +
+  theme_bw() + xlab("Event Time") +
+  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") +
+  geom_hline(yintercept = 0,size=.2,linetype="dashed") + 
+  theme(legend.position = "none", text=element_text(size=17,family="lm")) 
+
 
 
 ## CLAIM COUNT ---------------------------------------------------------------------------------------
@@ -789,14 +815,27 @@ p<-claim2_cs$Wpval
 claim2_cs_dyn <- aggte(claim2_cs, type = "dynamic", na.rm=T)
 
 # Create a plot
[…]

### 2022-04-18 Scripts/sumstats.R

@@ -392,5 +392,6 @@ AHA_sample <- AHAmainsurvey %>%
 
 
 panelview(AHA_sample, Y=NULL, D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE) +
-  theme(text=element_text(size=50,family="lm"))
-ggsave("Objects/hosp_treat.png", width=11.4, height=15)
+  theme(text=element_text(size=70,family="lm"))
+
+ggsave("Objects/hosp_treat.pdf")

### 2022-04-18 Paper/Paper.tex
Updates from Overleaf

@@
 As with other outcomes, this result is totally due to changes by the younger group of physicians.
-While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is economically trivial.
+While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is not largely meaningful.
 
-Since this is a demand-driven variable as well, I investigate whether any external factors may shift the demand for health care over the sample period.
+Since this is also a demand-driven variable as well, I investigate whether any external factors may shift the demand for health care over the sample period.
 
@@
 EHRs do not seem to be facilitating more billable activity.
+
+However, these results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
+Thus, I limit the sample to only physicians who work stay with the same one hospital through the entire sample.
+These physicians do not retire or see patients in another hospital.
+Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period.
+I present the results for this limited sample in Figure \ref{fig:limi}.
 
@@
     \caption{Hospital EHR Adoption Over Time}
-    \includegraphics[scale=.6]{Objects/hosp_treat.png}
+    \includegraphics[scale=.8]{Objects/hosp_treat.pdf}
     \label{fig:hosp_treat}
@@
     \label{fig:patient}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
+    \includegraphics[scale=.4]{Objects/patient2_plot.pdf}
+    \label{fig:patient2}
 \end{figure}

### 2022-04-21 Paper/Paper.tex

@@
 Moreover, shocks to an existing labor market could have downstream effects on customers in that market.
-For example, technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health.
+A prime instance of this is in health care, where technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health.
 In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S.
@@
 
-My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in a hospital.
+My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in hospitals.
 The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
-I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.
 For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals.
 In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
+I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.
 
@@
 In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
-Further, when considering a more general outcome of whether a physician switches zip codes, I observe similar trends with even larger magnitudes.
-This implies that the technology imposed a large enough cost to induce switching and that physicians may be avoiding regulation that imposes loss of autonomy.
+I observe similar trends in whether physicians change zip codes, although pre-trends may be driving this result.
 Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
@@
 
-A key assumption that underlies the main analysis is that exposure is exogenous to individual physicians.
-That is, the physician does not choose their own EHR exposure.
+A key assumption that underlies the main analysis is that physicians are constrained to using the technology in a hospital who chooses to adopt.
+I consider two possibilities where this assumption is violated.
+First, I investigate the existence of hospital employees with the purpose of using or assisting with an EHR on behalf of a physician.
+If this is occurring, the results are at least partially driven by the hiring of new employees.
+I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the majority of EHR implementation.
+Further, my main results are not sensitive to the exclusion of these years and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
+Second, I investigate whether the decision for hospitals to implement EHRs is exogenous to individual physicians.
+That is, a physician does not choose their own EHR exposure.
 While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making.
-For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a much lower probability that EHR implementation is endogenous.
+For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a lower probability that EHR implementation is endogenous.
 The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.
@@
 
-I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through physicians' response to them.
-This is the first study to my knowledge that connects EHR implementation to various physician labor market decisions.
+I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses.
+This is the first study to my knowledge that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups.
 Further, I improve on the literature by considering the primary time period in which EHRs were rapidly implemented.
 This provides sufficient variation in the timing of EHR implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in a typical two way fixed effects setup.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects.
 
@@
 
-EHRs have been an important feature of healthcare since the 1980s and 1990s.
-Early in its existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling.
+EHRs have been an important feature of health care since the 1980s.
+Early in the technology's existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling.
 Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation.
@@
 hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
-2013.
+2013, revelaing the nationwide and rapid expansion of technology.
 
 Figure \ref{fig:EPIC} shows a two computer screenshots of what physicians might see in their EHR system.
-On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of the time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}).
+On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}).
 These are the most time consuming portions of EHR use, according to the same study.
 In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR.
-
-While most hospitals fully implemented EHRs by 2015, physicians still continue to report frustration.
-The same functions of EHR use that take the most time are also reported to be the most frustrating (\cite{dymek2021building}).
-A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages for other members of the care team, and searching relevant information (\cite{dymek2021building}).
+Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use.
+The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}).
+A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages meant for other members of a care team, and searching relevant information (\cite{dymek2021building}).
 Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or even longer to load.
@@
 I describe the different data sets used to construct the panel below.
+I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
 
@@
 My analysis relies on EHR implementation in hospitals.
-Thus, I only include physicians reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed in all years but one).
+Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics, general practice or family practice in at least one year.
+However, I exclude physicians who list themselves as a specialist in all but one year.
 Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
@@
 
-CMS Shared Patient data records annual information from 2009 to 2015 on the number of patients who bill two entities under Medicare.
+CMS Shared Patient data records annual information on the number of patients who bill two entities under Medicare from 2009 to 2015.
 For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common.
@@
 I focus on the number of shared patient billing two entities in the same day, within a 30 day interval.
-I utilize this data to explore physicians' EHR exposure based on the hospitals they share sufficient patients with.
+I employ this data to detect physician who work closely with hospitals and which of those hospitals have implemented an EHR, denoting each physician's exposure to EHRs.
 This measure of exposure defines various treatment variables for my analysis.
@@
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
-I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year.
-This is because it is important to avoid dropping physicians who leave the data for many years, as this may indicate retirement and skew the results of my analysis.
+I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year, which gives a threshold of work relationship without excluding physicians who have no claims for a number of years.
 
@@
 I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
-I create two measures of EHR exposure.
-The first is exposure to any EHR in any hospital, which attempts to capture the effect of first-time EHR use on physician labor markets.
-One concern with this approach is that a typical physician may be involved in the decision to implement an EHR and thus the treatment would be endogenous.
-While I cannot perfectly address this concern, I define an alternative treatment variable which theoretically will have less likelihood of this endogeneity problem.
-That is exposure to EHR at a "low integration" hospital (a hospital where physicians have a low measure of vertical integration with the hospital (\cite{madison2004hospital})).
-In low integrated settings, physicians do not have ownership or much control in the hospital setting.
-Results for this alternative defintion of exposure are presented in Appendix ().
 I then aggregate this data to the physician level.
@@
 where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$.
-Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future.
-\footnote{Alternatively, I could also define retirement using future number of patients.
+Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future \footnote{Alternatively, I could also define retirement using future number of patients.
 I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}
+In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.
 
-Next, I consider a physician switching place of work, such as switching to an office based setting or even selecting into different hospitals where preferences may be more aligned.
+Next, I consider whether physicians exhibit behavior that suggests they have shifted towards other work settings.
 I measure this in the data in three ways.
@@
 Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next.
-While no single variable directly measures physician movement across hospitals, these three outcomes collectively capture different ways in which physicians may change their practice setting over time.
+While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from EHRs.
 
-Finally, I consider outcome variables which explore productivity and billing activity.
-Productivity is measured as the number of patients a physician sees in a given year, where the billing activity is the total claim count for the physician.
-I distinguish between the two because it could be the case that productivity does not actually increase, but EHRs could encourage more comprehensive billing.
+Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and total billing activity.
 
@@
 Only 2\% of physicians (approximately 750) retire over the course of the panel.
-A little over half of the sample works in an office in some capacity, with about a third of the number of patients being seen in an office setting.
-By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of being exposed.
-Physicians are typically exposed in any capacity about one fourth of the way through the sample.
+Just over half of the physician-year observations have value one for working in an office in some capacity, with about a third of the number of patients being seen in an office setting.
+By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure.
+Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015.
 The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.
 
-I also include a table of means comparing physicians younger than 60, older than 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}.
+I also include a table of means comparing physicians younger than 60, at least 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}.
 Senior physicians see more patients and work in more hospitals on average than younger physicians.
@@
 Finally, physicians in the higher age bracket are less likely to switch zip codes.
-The age groups are exposed to EHRs in similar ways, but younger physicians are exposed to EHRs in low integrated setting slightly later.
+The age groups are exposed to EHRs in similar ways.
 
-In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment variables over time.
+In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time.
 In 2009, 32\% of the sample of physicians were already exposed to an EHR.
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
-Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by 2015.
-Exposure to EHRs only in low integrated hospitals follows a similar trend with lower exposure rates.
-When I employ this treatment variable, I drop all physicians who are not exposed in a low-integration setting by 2015, so the sample of physicians changes according to this limitation.
+Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by then.
 The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time.
@@
 In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
-Therefore, to avoid negative weighting issues that arise in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$:
+Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$:
 $$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
@@
 To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+Other estimators that similarly address the concerns yield similar results, presented in Appendix ().
 
@@
 That is, once a physician is exposed to an EHR, they cannot be un-exposed.
-This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure.
+At the hospital level, this assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure.
 An EHR is a costly technology and requires a significant amount of collaboration to implement.
@@
 This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a random sample of hospitals over time.
-Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR.
+Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
 Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
 However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
-I investigate this possibility and limit the data to address it in Appendix ().
+I investigate this possibility and limit the data to address it in Section \ref{sec:patientcount}.
 
@@
 That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
-
 It is reasonable to believe that physicians are fully utilizing the EHRs in the hospital they work in, since by 2015 electronic health records were widely implemented and considered unavoidable.
-An objection to this assumption may be that physicians stay in a hospital, but the hospital hires a data assistant to do technology work for the physician.
+An objection to this assumption may be that physicians stay in a hospital, but the hospital hires an employee to do technology work for the physician.
 In this case, the effects seen could be from having the data assistant instead of EHR use.
@@
 
-An exogenous shock to the physician such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock.
+As an exogenous shock to work environment, EHR implementation may lead to an earlier retirement than would have otherwise occurred.
 Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
-The aim of this section is to test that hypothesis.
+The aim of this section is to test that hypothesis and evaluate whether younger physicians exhibited similar behavior.
 
@@
 Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure.
-In this scenario, it would be alarming to see a numerically large effect on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors.
+Considering the magnitude of the decision to retire, it would be alarming to see a numerically large effect on this outcome.
+A 10\% increase in the likelihood of such a decision is a meaningful jump.
 While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that a positive effect on the likelihood of retirement did occur as a result of EHR implementation.
 
-Next, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
-These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure.
+Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
+These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above.
 A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
@@
 However, there are other ways physicians might change their labor market behavior under sufficient incentive to avoid the technology.
-I consider the decision to switch work settings in multiple different outcomes.
+I consider the decision to switch work settings using multiple outcomes.
 First, I directly investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure.
-I do this using variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
+I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
 This analysis does not include physicians who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
@@
 The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean.
+This is positive, but somewhat inconsequential effect, both economically and compared to the effect of EHRs on retirement discussed above.
 Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
@@
 If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings.
-The estimates for this outcome are shown in Figure \ref{fig:officesecond}, where I find limited evidence that EHR exposure affects the fraction of patients seen in office settings.
-Of course, this could mean multiple things.
-If total number of patients remains constant, then this implies that the absolute number of patients seen in offices also is unchanged.
-However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in a similar direction.
+I show the estimates for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
+This result has varying implications depending on how total number of patients is affected.
+If total number of patients remains constant, then the absolute number of patients seen in offices is also unchanged.
+However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in the same direction.
 I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.
@@
 
-Since this is also a demand-driven variable as well, I investigate whether any external factors may shift the demand for health care over the sample period.
+This variable is also driven by demand for health care.
+A major change in the U.S.
+health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
+If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, the ACA may be driving the results.
+The ACA could potentially affect patient counts in opposite ways: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population.
+A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}).
+Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}.
+Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.
 
-This outcome directly relates to the outcome in the previous section, the fraction of patients seen in an office setting.
+Total patient count is directly related to the fraction of patients seen in an office setting.
 For younger physicians, since the total number of patients increases in the first couple of years after exposure, the absolute number of patients seen in offices increases as well.
@@
 
-EHRs have the potential to change claim incentives if the administrative costs of billing decrease or if EHRs encourage billing for more items than otherwise would have been recorded.
-
+Further, EHRs also have the potential to change billing activity differentially than patient count changes.
+If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the amount of patients has changed or not.
 On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year.
@@
 
-However, these results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
+However, for both patient and claim count, the results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
 Thus, I limit the sample to only physicians who work stay with the same one hospital through the entire sample.
@@
 Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period.
-I present the results for this limited sample in Figure \ref{fig:limi}.
+I present the results for this limited sample in Figure \ref{fig:limitedsample}.
 
@@
 
-\subsection{Hospitals Hiring Data Assistants}\label{sec:dataass}
+\subsection{Hiring Data Assistants}\label{sec:dataass}
 
-One concern with this analysis is that any effect could be caused by an outside factor that is also correlated with EHR adoption in hospitals.
-A logical outside factor could be the existence of employees with the purpose of utilizing electronic health records on a physician's behalf.
-Employees of this type are assigned a medical tax ID when working in hospitals; they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals.
+One such occurrence would be the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not actually use the EHR.
+Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
+they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
 I will refer to any employee in these categories as a data assistant.
+In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
 
-Using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
-Figure \ref{fig:dataassistant_histogram} is a frequency plot of the year of activation for every tax code that falls in the categories listed above.
-The total number of these registered employees is 875.
-The first data assistants ever registered were in 2005.
+First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
+Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
+The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
 From 2005 to 2013, 15-60 more employees were registered each year.
@@
 If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
+My results are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
+
+Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
+I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
+I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
+I find that .1\% of physician-year observations are associated with data assistants.
+This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
+However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
+Details of this analysis are given in Appendix ().
 
@@
     \centering
-    \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
-    \includegraphics[scale=.4]{Objects/patient2_plot.pdf}
-    \label{fig:patient2}
+    \caption{Effect of EHR Exposure on Patient Count by Group}
+    \includegraphics[scale=.3]{Objects/patient_group.pdf}
+    \label{fig:patientgroup}
 \end{figure}
@@
     \label{fig:claim}
+\end{figure}
+
+\begin{figure}[p]
+    \centering
+    \caption{Effect of EHR Exposure on Productivity, Limited Sample}
+    \includegraphics[scale=.4]{Objects/limitedsample_plot.pdf}
+    \label{fig:limitedsample}
 \end{figure}
@@
 
-\section{Alternate Definition of EHR Exposure}
+\section{Data}\label{app:data}
+
+This section details the process of creating the data set used in the analysis of the main paper.
+
+\subsection{Physician Specialties by Tax Code}\label{sec:taxcode}
+
+I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI_Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}.
+I then categorize NPIs by key words in their tax code description.
+Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", ``General Practice", or ``Pediatrics" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital.
+I create two data sets, one containing only PCPs and one containing only hospitals.
+These data are saved to use in identifying the relevant NPIs in the CMS Shared Patient Data.
+
+\subsection{CMS Shared Patient Data}\label{sec:sharedpat}
+
+For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}}.
+I use data that captures shared patient activity in 30 day intervals.
+I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals.
+I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs.
+Some pairs are duplicates from the hospital being listed first or second in the shared patient data.
+I combine duplicates into one observation, summing the same day count variable.
+Once duplicated are removed, there are 7.1 million observations.
+Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital.
+I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.
+The final list of pairs consists of 1.3 million observations.
+
+\subsection{Pair-Level Variables}
+
+I combine the physician-hospital pairs created in Section \ref{sec:sharedpat} with various data sets for hospital or physician level information.
+First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation year and gender.
+This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation.
+I drop any physicians who graduated medical school after 2004, since graduating after that means leaving residency or graduating during the span of my main data, 2009-2017, and potentially exhibiting labor market changes that seem associated with EHRs but are not.
+
+For hospital level variables, I use an AHA-NPI crosswalk to merge the pairs to the Annual Hospital Administration Survey (AHA Survey) from 2009-2015.
+The variables I include from this data are a hospital's number of beds, organization type, system ID, and EHR use.
+There are 4,253 unique AHA hospitals in the shared patient data.
+I drop any hospitals that aren't in the AHA Survey due to lack of information on EHR use.
+After limiting the hospitals, there are 780,000 observations of pair-years left.
+Next, I investigate the hospitals with missing information for EHR use.
+There are 89 hospitals in the data who never answer the survey question about EHR use; I drop these hospitals.
+If a hospital does not answer the survey question in one year, but the year before and after have an identical survey answer, I fill in the missing year of information.
+Then there are 4,253 observations with missing data for the EHR survey question.
+I make a further limitation to hospitals with at least 10 beds.
+There are only 58 unique hospitals in the data with less than 10 beds, and these are likely very different from the rest of the sample of hospitals.
+Then, I create a physician-level variable which captures the average number of beds in the hospitals they work with in each year.
+I sum a physician's same day count over their hospitals in a given year to create a physician level patient count variable.
+I also create this variable only for hospitals who are using an EHR and only for hospitals who are not using an EHR.
+
+I merge the remaining pairs to an additional survey, the AHA IT Supplement, only sent to hospitals who reported using an EHR.
+There are more detailed questions in this survey including the EHR vendor, and the ranking of documentation and decision making features the hospital uses.
+There is a lot of missing data in this survey, which is why I do not use it in my main analysis.
+
+I use the general survey question about each hospital's EHR use to define my key treatment variable.
+The survey asks the extent to which the hospital utilizes an EHR and allows for three answers: not at all, partially utilize an EHR, and fully utilize an EHR.
+I create a binary variable equal to one if the answer is fully utilize, zero otherwise.
+The data will be aggregated up to the physician level due to physician level outcome variables, so I first transform the pair-level EHR variable into a treatment variable at the physician level.
+I create a variable capturing the first year that at least one of a physician's connected hospitals uses an EHR.
+For the endogeneity discussion in Section \ref{sec:endog}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR.
+The level of vertical integration between hospital and physician is defined as in Madison (\citeyear{madison2004hospital}) using the organization type of the hospital.
+Hospitals classified as IPA or PHO are low-integration, and any other type is high0integration.
+I also create a variable that captures the fraction of hospitals a physician works with that use an EHR, which varies on the dimension of EHR use and on number of hospitals worked with.
+Finally, the last variable I create before aggregating to the physician level is an indicator for whether the physician works with the same set of hospitals through the entire sample.
+I create this variable by comparing a physician's number of hospitals in each year to the maximum number of hospitals they ever work with.
+If there is any year in which the a physician's number of hospitals is less than the maximum number of hospitals they ever work with, the indicator for never having a new NPI is set to 0.
+I use this variable to limit the sample in Section \ref{sec:patientcount}.
+
+Finally, I aggregate to the physician level by only keeping variables that do not vary by hospital.
+The variables in the physician level data are year, physician NPI, graduation year, average number of beds in hospitals worked with, years of experience, number of hospitals, fraction of hospitals with EHR, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, patient count, patient count in EHR hospitals, patient count in non-EHR hospitals, and never works with a new NPI.
+I complete this data to include years 2016 and 2017, but leave time varying variables missing for those years.
+Thus, this is a balanced panel of physicians.
+I save this data to be merged with physician labor market activity in Section \ref{sec:appmdppas}.
+
+\subsection{Physician Labor Market Activity}\label{sec:appmdppas}
+
+Using physician NPI, I merge the physician treatment data to Medicare Data on Provider Practice and Specialty (MD-PPAS)\footnote{Information on data located at \hyperlink{https://resdac.org/cms-data/files/md-ppas}{https://resdac.org/cms-data/files/md-ppas}}, which psnas 2009-2017.
+This data contains variables on physician specialty, Medicare claim counts in various zip codes, unique number of patients seen, fraction of patients seen in specific settings, patient demographics, and physician date of birth.
+Once the data is merged, I make a further limitation to drop any physicians with less than 20\% of their total patients in a hospital setting.
+This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
+With this limitation, there are 403,000 observations left in the data.
+
+Next, I create the dependent variables used in the analysis.
+The first outcome is whether a physician chooses to retire or not.
+First, I sum a physicians claim counts across zip codes into one variable for total claim count in a given year.
+Then, I create a variable that sums up a physicians claims in all future years.
+In 2009, this variable sums up all claims from 2010-2017, and so on.
+Then I create a variable for the first year that a physician has a future claim count of zero.
+However, this counts retirement one year too early.
+For example, if a physician has no claims from 2014 to 2017, the minimum year variable is set to 2013, but the first year of retirement is 2014.
+Thus, I add one year to the minimum year variable.
+I then define the outcome variable, retire, as one if a physician ever retires and the year is equal to the first year of retirement.
+I utilize a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.
+
+The next outcome variables consider level of labor market activity in office settings.
+The first decision to make is how to handle years where a physician has missing data for claims.
+A year of missing data can mean the physician had no claims that year, or the observation was removed due to insufficient claims (less than 11).
+A conservative assumption is to assume a year of missing data means zero claims, thus, I set all missing claim counts to zero.
+The data already contains a variable for the fraction of patients seen in an office setting, which is used in the analysis, but I further create an indicator variable for whether the physician sees any positive amount of patients in an office setting.
+
+I also create an indicator variable for whether the physician switches zip codes from one year to the next.
+The physicians in this data work in up to 12 different zip codes, so I their zip codes into one string variable, a list of all zip codes in a year.
+I make this variable into wide format: zip list 2009, zip list 2010, ...
+, zip list 2017.
+In year $t$, I define a variable change zip that is equal to one if the zip list in $t-1$ is different from the zip list in $t$, zero otherwise.
+Finally, the variables for unique patient count and claim count are already defined.
+This is the final data used in the paper, with 403,020 observations.
+
+\section{Sensitivity Analysis}
+
+\subsection{Physician Selection into EHRs}\label{sec:endog}
 
@@
 
-\subsection{Retirement}
-
-\section{Anticipation}
+\subsection{Anticipation}
 
@@
 
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Retirement}
-    \includegraphics[scale=.4]{Objects/retire_plot_LI.pdf}
-    \label{fig:retire_LI}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Likelihood of Working in Office}
-    \includegraphics[scale=.4]{Objects/officeind_plot_LI.pdf}
-    \label{fig:officeind_LI}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Fraction of Patients in Office}
-    \includegraphics[scale=.4]{Objects/officefrac_plot_LI.pdf}
-    \label{fig:officefrac_LI}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Likelihood of Switching Zip Codes}
-    \includegraphics[scale=.4]{Objects/zip_plot_LI.pdf}
-    \label{fig:zip_LI}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Number of Patients}
-    \includegraphics[scale=.4]{Objects/patient_plot_LI.pdf}
-    \label{fig:patient_LI}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of Exposure in LI Setting on Billing Activity}
-    \includegraphics[scale=.4]{Objects/claim_plot_LI.pdf}
-    \label{fig:claim_LI}
-\end{figure}
-
 \end{document}

### 2022-04-25 Scripts/analysis.R
add other sensitivity graphs and update scripts

@@ -70,7 +70,7 @@ for (d in 2010:2015){
            clean_control=ifelse(minyr_EHR>d,1,0),
            years_included=ifelse(year>=d-1 & year<=d+1,1,0)) %>%
     mutate(inclusion=years_included*(treated_unit+clean_control)) %>%
-    filter(inclusion==1)
+    dplyr::filter(inclusion==1)
   
   assign(paste0("subgroup_",d),subgroup)
 }

### 2022-04-25 Scripts/sensitivity_analysis.R

@@ -2,6 +2,8 @@ library(showtext)
 library(ggplot2)
 library(did)
 library(ggpubr)
+library(fixest)
+library(did2s)
 
 
 
@@ -26,521 +28,206 @@ showtext_auto()
 # Read in the main data created in "data5_MDPPAS.R"
 Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
+
+
+
 ### LOW INTEGRATION DEFINITION #####################################################################################
- # Retire
+varlist <- list("retire", "pos_office", "work_in_office", "change_zip", "npi_unq_benes", "claim_count_total")
+
+# Create List of Results for each age group
+models_LI <- lapply(varlist, function(x) {
+  att_gt(yname = x,
+         gname = "minyr_EHR_int",
+         idname = "DocNPI",
+         tname = "year",
+         xformla = ~grad_year,
+         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0) else dplyr::filter(Physician_Data,minyr_EHR_int>0),
+         est_method = "dr",
+         control_group = "notyettreated")
+  
+})
+
+models_LI_young <- lapply(varlist, function(x) {
+  att_gt(yname = x,
+         gname = "minyr_EHR_int",
+         idname = "DocNPI",
+         tname = "year",
+         xformla = ~grad_year,
+         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0 & max_age<60) else dplyr::filter(Physician_Data,minyr_EHR_int>0 & max_age<60),
+         est_method = "dr",
+         control_group = "notyettreated")
+  
+})
+
+models_LI_old <- lapply(varlist, function(x) {
+  att_gt(yname = x,
+         gname = "minyr_EHR_int",
+         idname = "DocNPI",
+         tname = "year",
+         xformla = ~grad_year,
+         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0 & max_age>=60) else dplyr::filter(Physician_Data,minyr_EHR_int>0 & max_age>=60),
+         est_method = "dr",
+         control_group = "notyettreated")
+  
+})
+
+# Translate to single ATT value for each age group
+Simple_LI <- lapply(models_LI, function(x){
+  aggte(x, type = "simple")
+})
+
[…]

### 2022-04-25 Paper/Paper.tex

@@
 Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
-Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
+Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching or avoiding Medicare patients.
 Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
 In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
-I observe similar trends in whether physicians change zip codes, although pre-trends may be driving this result.
 Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
@@
 
-The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
+The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I loosely call retirement. This variable is constructed as follows: 
 $$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
@@
 
-Next, I consider whether physicians exhibit behavior that suggests they have shifted towards other work settings.
+Next, I consider whether physicians exhibit behavior that suggests they have shifted setting or location of work.
 I measure this in the data in three ways.
@@
 There is no evidence that older physicians are shifting towards offices.
-This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more "stuck" to their particular work environment.
+This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more stuck to their particular work environment.
 
@@
 The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}.
-The graph behaves similarly to the probability of working in an office in that the probability of switching zip codes increases and the result is driven solely by the younger demographic.
-However, this result is meaningfully larger; a physician is .025 ppts more likely to switch zip codes in the same year as EHR exposure and the year after, a 31.25\% increase relative to the mean.
-This result suggests that EHRs are imposing a cost high enough for physicians to exhibit significant switching behaviors, whether that is towards offices or other hospitals.
+While there is a statistically positive effect in the first 2 years of exposure, there is an upward sloping pre-trend that is probably driving this.
+Thus, I make no claim that EHR exposure is leading physicians to switch zip codes.
+
+A limitation in this analysis is that physicians may shift work behavior due to the organization of the hospital system changing.
+For example, a hospital may acquire a smaller practice and send physicians to work there for a period of time.
+If this acquisition is correlated with incentives to adopt an EHR, then the effect I find is biased upward.
+Since I do not find strong evidence for switching behavior, I leave this question to future research.
 
@@
 Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.
-In this section, I address EHRs' effect on productivity by estimating its effect on a physician's number of patients seen.
+In this section, I speak to efficiency by estimating the effect of EHR implementation on patient count.
 
@@
 
-This variable is also driven by demand for health care.
+Patient count is also driven by demand for health care.
 A major change in the U.S.
@@
 If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, the ACA may be driving the results.
-The ACA could potentially affect patient counts in opposite ways: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population.
-A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}).
+The ACA has potentially opposing effects on Medicare patients: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population.
+A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
 Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}.
@@
 
-Further, EHRs also have the potential to change billing activity differentially than patient count changes.
+EHRs may affect physician incentives in such a way that billing activity changes differently from patient count.
 If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the amount of patients has changed or not.
@@
 However, for both patient and claim count, the results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
-Thus, I limit the sample to only physicians who work stay with the same one hospital through the entire sample.
+Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
 These physicians do not retire or see patients in another hospital.
@@
 I present the results for this limited sample in Figure \ref{fig:limitedsample}.
+These results are similar to those presented for the full sample, suggesting that physicians become more productive in terms of patient count in the first two years after implementation.
 
@@
 
-In all outcomes, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
-All specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure.
-This is likely due to the composition of the hospitals adopting at different times.
-Consider time $t+4$.
-The only hospitals to analyze the effect in this period are those who adopted in 2010 compared to hospitals who adopted in 2015.
-The group of hospitals who adopted in 2015 may be compositionally different than early adopters if there is a characteristic reason for late adoption such as demographic of patients.
-For this reason, I focus discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.
+In the observed results above, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
+Almost all specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure.
+One reason for this may be the composition of the hospitals which adopt at different times.
+Consider event time $t+4$.
+The physician included in the estimation for this coefficient are those exposed in 2010 compared with those who did not adopt until 2015, implying that the underlying hospitals implemented EHRs in either 2010 or 2015.
+It is likely that the group of hospitals who adopted in 2015 are compositionally different than early adopters, especially if poorer hospitals adopted later.
+For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.
 
@@
 
-One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals.
-One such occurrence would be the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not actually use the EHR.
+One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals, such as the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
 Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
 I will refer to any employee in these categories as a data assistant.
@@
 If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
-My results are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
+The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
 
@@
 \subsection{Physician Influence of EHR Decision in Hospitals}
+
+Another way physicians may avoid EHR exposure is to exhibit decision making power to either delay or encourage EHR implementation in hospitals.
+This would make the binary treatment variable endogenous, biasing the results of the main specification, as it includes physicians who are selecting into their own exposure.
+After the sample period ends, the government began penalizing hospitals for not using health information technology meaningfully, so implementation eventually occurs across the board.
+However, during the sample, physicians may impact the timing or occurrence of implementation.
+I address this question by limiting the sample of hospitals in a meaningful way, only including those hospitals which are considered to have low vertical integration with physicians.
+That is, physicians are employees and less likely to be decision makers.
+I then redefine exposure as when a physician is exposed to an EHR in low integration settings.
+I present details of how I define integration and results for this specification in Appendix \ref{sec:endog}.
+In general, group time estimates are similar to those in the main specification, with the exception of the probability of working in an office setting, which may indicate some correlation between EHR adoption in high integration settings and acquisition of physician offices.
 
@@
     \centering
-    \caption{Effect of EHR Exposure on Claim Count}
+    \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
     \includegraphics[scale=.4]{Objects/claim_plot.pdf}
@@
     \centering
-    \caption{Effect of EHR Exposure on Productivity, Limited Sample}
+    \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
     \includegraphics[scale=.4]{Objects/limitedsample_plot.pdf}
@@
 For the endogeneity discussion in Section \ref{sec:endog}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR.
-The level of vertical integration between hospital and physician is defined as in Madison (\citeyear{madison2004hospital}) using the organization type of the hospital.
+The level of vertical integration between hospital and physician is defined as in \citeauthor{dynan1998assessing} (\citeyear{dynan1998assessing}) using the organization type of the hospital.
 Hospitals classified as IPA or PHO are low-integration, and any other type is high0integration.
@@
 The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in.
-A concern that one may have is that physicians choose their individual EHR exposure, making this an endogenous treatment variable.
-The results do not indicate that this is happening on average, since we would expect physicians not to respond on the margin of retirement or switching zip codes if they are choosing to be exposed to EHRs.
+These results are biased if physicians endogenously select EHR exposure, particularly by participating in the hospital's decision to implement.
+The results do not indicate that this is a huge issue on average, since we would expect physicians not to respond on the margin of retirement or switching work setting if they are choosing to be exposed to EHRs.
 Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions.
-However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable.
-This variable is whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
+However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable: whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
+
+I define a hospital's level of integration based on organizational structure, where independent practice associations (IPAs) and physician-hospital organizations (PHOs) are classified as low vertical integration and other organization types are classified as high vertical integration (\cite{dynan1998assessing}).
+
+I present a comparison of the average treatment effect in the main specification and the average treatment effect with low-integration exposure as the binary treatment variable in Figure \ref{fig:LI}.
+For conciseness, I present group time treatment effects to a single ATT value.
+As expected, the coefficient on retirement is nearly identical to the main specification.
+However, the coefficient for the probability of working in office based settings completely switches signs, pointing to questions about acquisition of physician practices.
+In low-integration settings, physicians are less likely to shift to an office after EHR exposure by a small measure.
+However, when including physicians with more power, the do seem to start office based work.
+If high-integration hospitals are acquiring smaller practices as EHRs are implemented, physicians from the hospital may be sent to the newly acquired practice for a period of time during transition.
+This is an important result because it points to physicians shifting according to hospital incentives instead of as a method to avoid EHR implementation.
+Estimates for other coefficients are more noisy, but similar to the main specification.
 
@@
 
+\begin{figure}
+    \centering
+    \caption{Comparison of ATT, Main Specification and Low-Integration Treatment Variable}
+    \includegraphics[scale=.5]{Objects/LI_results.pdf}
+    \label{fig:LI}
+\end{figure}
+
 \end{document}

### 2022-04-25 Paper/Paper.tex
Updates from Overleaf

@@
 Therefore, I proceed in the main specification assuming no anticipation.
-However, I explore and present results for a one year anticipation period in Appendix ().
+However, I explore and present results for a one year anticipation period in Appendix \ref{app:anticipation}.
 
@@
 
-\subsection{Anticipation}
+\subsection{Anticipation}\label{app:anticipation}
+
+While there is evidence to suggest the time it takes to implement an EHR does not exceed one year, there still may be some anticipation if physicians are expecting implementation in the hospitals they work in.
+In Figure \ref{fig:anticipation}, I present aggregated estimates in the case of one year of anticipation.
+The results are virtually identical, just noisier, compared to the estimates from the main specification.
+
+\subsection{Other Estimators}
 
@@
 
-\begin{figure}
+\begin{figure}[ht]
     \centering
-    \caption{Comparison of ATT, Main Specification and Low-Integration Treatment Variable}
+    \caption{Comparison of ATT, Low-Integration Treatment Variable}
     \includegraphics[scale=.5]{Objects/LI_results.pdf}
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{ATT With Anticipation}
+    \includegraphics[scale=.57]{Objects/anticipation_graph.pdf}
+    \label{fig:anticipation}
+\end{figure}
+
 \end{document}

### 2022-05-10 Paper/Paper.tex

@@
 \addtolength{\skip\footins}{2pc plus 5pt}
+
+\usepackage{geometry}
+ \geometry{,
+ left=20mm,
+ top=20mm,
+ right=20mm,
+ bottom=20mm
+ }
 
@@
 How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries.
-For example, industrial robots decreased the share of hours worked in low-skilled workers from 1993-2007, but increased overall labor productivity  (\cite{graetz2018robots}), and even low-skilled workers in Norway who were not displaced by robots, but saw other workers displaced, experienced lower job satisfaction due to fear of displacement (\cite{schwabe2020automation}).
 Moreover, shocks to an existing labor market could have downstream effects on customers in that market.
 A prime instance of this is in health care, where technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health.
+The most  widespread and health care-altering technological innovation in recent years took place during the implementation of electronic health records.
+This huge innovation in the health care sector, where there is extensive margin for improvement, is an excellent setting to examine the relationship between innovation and labor market, and the downstream effects this may have.
 In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S.
@@
 Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care.
-EHRs have have become increasingly relevant in the U.S.
+EHRs have become increasingly relevant in the U.S.
 since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
@@
 My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in hospitals.
-The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and demographic information such as zip code, age, and gender.
+The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and other demographic information.
 For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals.
@@
 
-Contrary to previous research stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
-Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching or avoiding Medicare patients.
+My analysis contributes to the literature examining effects of health information technology on health care.
+There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect.
+Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings.
+One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).
+
+I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined.
+This is the first empirical study that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups.
+Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
+This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects.
+
+Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
+Contrary to this work, I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
+Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
 Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
@@
 I consider two possibilities where this assumption is violated.
-First, I investigate the existence of hospital employees with the purpose of using or assisting with an EHR on behalf of a physician.
+First, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
 If this is occurring, the results are at least partially driven by the hiring of new employees.
 I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the majority of EHR implementation.
-Further, my main results are not sensitive to the exclusion of these years and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
+Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
 Second, I investigate whether the decision for hospitals to implement EHRs is exogenous to individual physicians.
@@
 The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.
-
-My analysis contributes to the literature examining effects of health information technology on health care.
-There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect.
-Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
-More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings.
-One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).
-
-I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses.
-This is the first study to my knowledge that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups.
-Further, I improve on the literature by considering the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of EHR implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects.
 
@@
 hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
-2013, revelaing the nationwide and rapid expansion of technology.
+2013, revealing the nationwide and rapid expansion of technology.
 
@@
 Physicians have been voicing concerns over EHR usability since their beginning, indicating that they are likely responding in their behavior as well.
+
+\section{Physician Incentives}
+
+The physician behaviors that I consider in this analysis are closely intertwined and should not be taken as independent.
+For example, a physician's choice of work setting depends on their previous decision not to retire, and billing activity depends on work setting.
+Thus, I discuss the behaviors in terms of stages of decision making.
+
+\subsection{Retirement}
+
+Workers in developed countries tend to plan for formal retirement well in advance.
+Generally, an exogenous shock to a worker is not expected to change retirement age simply because of the amount of wealth and planning necessary to formally leave the labor force.
+Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}).
+However, physicians typically make this decision differently than workers in other industries.
+Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
+When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career.
+The decision to delay retirement is not financial, but altruistic.
+Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
+
+It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
+First, I will walk through the incentives of a physician who is already of retirement age when the EHR is implemented.
+While payment schemes, income as a whole, and retirement benefits certainly impact the utility a physician receives from working or retiring, I can model the decision as independent of those things since I assume the physician already has the financial mean to retire.
+Then, a physician will only retire if the utility from working is less than the utility from retiring, assuming the physician keeps working under indifference.
+How does exogenous EHR implementation affect these utilities?
+The utility received in retirement will not be affected by whether the hospital is using an EHR after the physician is already gone.
+However, the utility received from working is a function of work environment.
+If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility.
+If EHRs make daily tasks easier, there is a positive relationship.
+Thus, for a physician on the margin of retirement, the decision to retire only happens as a result of EHRs if they impose a cost, or a negative impact on work environment.
+Further, the cost has to be high enough to raise utility from retirement above utility from working, counteracting the altruism that is keeping the physician from retiring.
+
+Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role.
+There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management.
+While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching.
+This decision is impacted similarly to the above discussion regarding formal retirement, but is also impacted by income that would be received as a result of career switching.
+
+The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
+A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
+
+As an exogenous shock to work environment, EHR implementation may lead to an earlier retirement than would have otherwise occurred.
+Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
+The aim of this section is to test that hypothesis and evaluate whether younger physicians exhibited similar behavior.
+
+\subsection{Work Setting}
 
@@
 My analysis relies on EHR implementation in hospitals.
-Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics, general practice or family practice in at least one year.
+Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics (although Medicare primarily enrolls the elderly, Medicare also covers disabled children), general practice or family practice in at least one year.
 However, I exclude physicians who list themselves as a specialist in all but one year.
@@
 I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \textcolor{red}{()}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.
+In Appendix \ref{app:years}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.
 
@@
 To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
-Other estimators that similarly address the concerns yield similar results, presented in Appendix ().
+Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}.
 
@@
 In my analysis, I present p-values for a Wald test of pre-trends.
-In most cases, there is no evidence of a pre-trend, however, I investigate the assumption in detail in Appendix ().
+In most cases, there is no evidence of a pre-trend.
+In cases where a pre-trend could be driving the results, I make no strong claims about the effect of EHR exposure.
 
@@
 
-Workers in developed countries tend to plan for formal retirement well in advance.
-An exogenous shock to a worker is not expected to change retirement age because of the amount of wealth and planning necessary to formally leave the labor force.
-However, physicians typically make this decision differently than workers in other industries.
-Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}).
-When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career.
-The decision to delay retirement is not financial, but altruistic.
-Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.
-Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role.
-There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management.
-While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching.
-The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
-A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
-
-As an exogenous shock to work environment, EHR implementation may lead to an earlier retirement than would have otherwise occurred.
-Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
-The aim of this section is to test that hypothesis and evaluate whether younger physicians exhibited similar behavior.
-
 In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
-The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation.
+The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after implementation.
 While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .02.
@@
 health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
-If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, the ACA may be driving the results.
-The ACA has potentially opposing effects on Medicare patients: if Medicare patients are crowded out by an increased demand in Medicaid patients, or if the small changes to Medicare led to an increase in utilization from the Medicare population.
+If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the results.
+The ACA has potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
 A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
@@
 However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
-Details of this analysis are given in Appendix ().
+Details of this analysis are given in Appendix \ref{app:DA}.
 
@@
 This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes.
+The interaction between health information technology and labor markets is vastly important as the technology is continually legislated and growing each year.
 
-Using an event study framework, I find that physicians of all ages are more likely to stop clinical work altogether when exposed to an EHR in a hospital, where physicians of retirement age have an even larger increase relative to younger physicians.
-Further, I find that younger physicians are more likely to work in an office based setting after EHR exposure, and physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.
+I utilize data on shared patients, hospital EHR use, and physician claim activity.
+Using an event study framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR in a hospital, where those of retirement age have an even larger increase relative to younger physicians.
+Further, I find limited evidence that physicians are switching work settings, but that this may be driven by hospital incentives instead of physician incentives.
+Finally, physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.
 
 The implications of these results are threefold.
-First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care.
+First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care, especially if these physicians are in rural settings.
 In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
@@
 Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users.
-While the technology may be a burden, physicians are able to see more patients because of their use.
+While the technology may be a burden, physicians are able to see more patients because of their use, and are not more incentivized to over-bill.
 
@@
     \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.4]{Objects/retire_plot.pdf}
+    \includegraphics[scale=.45]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.4]{Objects/officeind_plot.pdf}
+    \includegraphics[scale=.45]{Objects/officeind_plot.pdf}
     \label{fig:officefirst}
@@
     \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
-    \includegraphics[scale=.4]{Objects/officefrac_plot.pdf}
+    \includegraphics[scale=.45]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
     \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.4]{Objects/zip_plot.pdf}
+    \includegraphics[scale=.45]{Objects/zip_plot.pdf}
     \label{fig:zip}
@@
     \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.4]{Objects/patient_plot.pdf}
+    \includegraphics[scale=.45]{Objects/patient_plot.pdf}
     \label{fig:patient}
@@
     \caption{Effect of EHR Exposure on Patient Count by Group}
-    \includegraphics[scale=.3]{Objects/patient_group.pdf}
+    \includegraphics[scale=.35]{Objects/patient_group.pdf}
     \label{fig:patientgroup}
@@
     \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
-    \includegraphics[scale=.4]{Objects/claim_plot.pdf}
+    \includegraphics[scale=.45]{Objects/claim_plot.pdf}
     \label{fig:claim}
@@
     \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
-    \includegraphics[scale=.4]{Objects/limitedsample_plot.pdf}
+    \includegraphics[scale=.45]{Objects/limitedsample_plot.pdf}
     \label{fig:limitedsample}
@@
 
-I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI_Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}.
+I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI-Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}.
 I then categorize NPIs by key words in their tax code description.
@@
 
-\subsection{Other Estimators}
+\subsection{Alternative Estimators}\label{app:estimators}
+
+There is a robust recent literature considering the issues with estimating heterogeneous treatment effects with a two way fixed effects specification, the classic approach in staggered treatment setups.
+To avoid these problems, in my main specification I use the average group time effect estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+In this section, I discuss other potential estimators and why I ultimately included average group time treatment effects as the main specification.
+
+One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}).
+This approach is commonly known as two stage difference in differences and is very clever.
+However, the approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
+Nevertheless, I limit the years of data to 2009-2015 and present estimates of this estimator in Figure \ref{fig:esimators}.
+Generally, the results have the same sign with a slightly larger magnitude, driven by the inclusion of never-treated units.
+
+Another estimation method established first used by \citeauthor{cengiz2019effect} (\citeyear{cengiz2019effect}) is commonly known as stacked regression.
+In this method, one re-frames event study data into groups of sub-experiments that are then stacked on top of each other, defining a group of treated units and "clean" control units.
+The drawback of this method is choosing an event study window that must be the same for each treatment group, where a large window eliminates many treated units and a small window does not allow for estimation over many years.
+I present the results using this method with an event window of one year in Figure \ref{fig:esimators}.
+The results are very similar to the main specification, with a smaller magnitude in some cases.
+
+\subsection{Limiting Sample Years}\label{app:years}
+
+I create the retirement outcome variable as a direct function of future claims.
+Those who retire in the early years of data have zero future claims for multiple years.
+However, those who retire after 2016 have only one or two years of zero future claims to be flagged as retired.
+While some of these physicians are likely retiring, it may be that a fraction of them only drop out of the data temporarily due to pregnancy or other factors.
+Thus, I limit the years in the analysis to 2009-2015 to ensure that those years are not driving the observed results.
+Results for this specification are shown in Figure \ref{fig:years}.
+The observed estimators are virtually identical to those in the main specification.
+
+\subsection{Physicians not Exposed to Data Assistants}\label{app:DA}
+
+In this section, I present results for the main specification with a data limitation: only including physicians who are exposed in hospitals that do not record employing any data assistants.
+These physicians may be more likely to respond negatively to EHRs and may be driving results in the main specification.
+I present these results in Figure \ref{fig:DA}.
+These results are surprisingly similar to those for exposure in low integration hospitals discussed in Appendix \ref{sec:endog}.
+Many outcomes have similar coefficients for the effect of EHR exposure, with the exception of probability of working in office settings.
+Those who are not exposed to data assistants are less likely to begin working in an office, an opposite sign from the specification in the main analysis.
+The level of integration could be correlated with hiring data assistants, which both may be correlated with the acquisition of smaller physician practices which are considered office based settings.
+
+\section{Heterogeneity Analysis}
 
@@
     \caption{Comparison of ATT, Low-Integration Treatment Variable}
-    \includegraphics[scale=.5]{Objects/LI_results.pdf}
+    \includegraphics[scale=.57]{Objects/LI_graph.pdf}
     \label{fig:LI}
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Other Event Study Estimators}
+    \includegraphics[scale=.57]{Objects/estimators_graph.pdf}
+    \label{fig:esimators}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Limited Sample Years}
+    \includegraphics[scale=.57]{Objects/years_graph.pdf}
+    \label{fig:years}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Physicians without Data Assistants}
+    \includegraphics[scale=.57]{Objects/DA_graph.pdf}
+    \label{fig:DA}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Heterogeneity Analysis}
+    \includegraphics[scale=.57]{Objects/gr}
+    \label{fig:DA}
+\end{figure}
+
 \end{document}

### 2022-06-03 Scripts/analysis.R
Update graphs after sample change (limiting fraction of inpatients to 70% or more)

@@ -180,1453 +180,147 @@ coefplot(retire_2sdid, keep=c("-4","-3","-2","-1","0",
 
 
 
+## ANALYSIS ---------------------------------------------- #########
+## Write results for each variable in a loop and create graphs ----- ##
 
-# RETIREMENT OUTCOMES --------------------------------------------------------------------------------
+varlist <- list("retire", "pos_office", "work_in_office", "change_zip", "npi_unq_benes", "claim_count_total")
 
-# Callaway and Sant'Anna (CS) ------------------------------------------
-# This is the estimator I land on, that's why it's in a different section
-
-retire_cs <- att_gt(
-  yname = "retire",                # LHS Variable
-  gname = "minyr_EHR",             # First year a unit is treated. (set to 0 if never treated)
-  idname = "DocNPI",               # ID
-  tname = "year",                  # Time Variable
-  # xformla = NULL                 # No covariates
-  xformla = ~grad_year,            # Time-invariant controls
-  data = dplyr::filter(
-    Physician_Data,minyr_EHR>0),   # Remove never-treated units
-  # data = Physician_Data
-  est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
-  control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
-  clustervars = "DocNPI",          # Cluster Variables          
-  anticipation=0                   # can set a number of years to account for anticipation effects
-)
-
-# Save p-value for pre-trends to put in footnote of table
-p<-retire_cs$Wpval
-
-# Aggregate the effects
-retire_cs_dyn <- aggte(retire_cs, type = "dynamic", na.rm=T)
-
-# Create a plot
-cs_retire_allEHR <- ggdid(retire_cs_dyn, xlab="\nEvent Time", title="All Physicians") + 
-  labs(caption=paste0("p-value= ",round(p,3),"\n")) + theme_bw() +
-  scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") + ylim(-.0025,.007) +
-  theme(text = element_text(size = 17, family="lm"))
-
-
-
-
-
-
-# Senior Physicians
-retireold_cs <- att_gt(
-  yname = "retire",                # LHS Variable
-  gname = "minyr_EHR",             # First year a unit is treated. (set to 0 if never treated)
[…]

### 2022-06-03 Scripts/data5_MDPPAS.R

@@ -53,11 +53,11 @@ Physician_Data <- Physician_Data %>%
   dplyr::mutate(max_age=max(age)) %>%
   dplyr::ungroup()
 
-# Drop any physicians with less than 20% of patients in hospitals
+# Drop any physicians with less than 70% of patients in hospitals
 Physician_Data <- Physician_Data %>%
   dplyr::group_by(DocNPI) %>%
   dplyr::mutate(max=max(pos_inpat,na.rm=T)) %>%
-  dplyr::filter(max>.2)
+  dplyr::filter(max>.7)
   # 403k obs
 
 
@@ -73,7 +73,7 @@ opt_out <- opt_out %>%
 
 Physician_Data <- Physician_Data %>%
   left_join(opt_out,by=c("DocNPI"="NPI"))
-# I see zero doctors actually opt out of medicare... can this be right?
+# I see zero doctors actually opt out of medicare.
 
 
 # CREATE DEPENDENT VARIABLES -----------------------------------------------------------------------------------------
@@ -122,10 +122,6 @@ Physician_Data <- Physician_Data %>%
 # Create variable for whether the physician ever retires
 Physician_Data <- Physician_Data %>% dplyr::ungroup() %>%
   dplyr::mutate(ever_retire=ifelse(is.na(minyr_retire),0,1)) 
-
-observe <- Physician_Data %>%
-  dplyr::filter(ever_retire==1) %>%
-  dplyr::select(DocNPI, year,age,retire,hosp_patient_count,claim_count_total,missingbefore)
 
 
 # OFFICE ####

### 2022-06-03 Scripts/sumstats.R

@@ -161,7 +161,7 @@ ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) +
         legend.key.width = unit(.3, 'cm'),
         text=element_text(size=17, family="lm"))
 
-ggsave("Objects/sum_stats_year.pdf", width=11.4, height=6, units="in")
+ggsave("Objects/sum_stats_year.pdf", width=10, height=7, units="in")
 
 # Graph continuous labor variable with lines for each year treated ------------------------------------
 # I don't think I want to include this graph in the paper
@@ -391,7 +391,7 @@ AHA_sample <- AHAmainsurvey %>%
 
 
 
-panelview(AHA_sample, Y=NULL, D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE) +
-  theme(text=element_text(size=70,family="lm"))
+panelview(AHAmainsurvey, Y=NULL,D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE) +
+  theme(text=element_text(size=10,family="lm"))
 
 ggsave("Objects/hosp_treat.pdf")

### 2022-06-03 Paper/Paper.tex
Updates from Overleaf

@@
 \maketitle
+
+\begin{abstract}
+    I investigate the relationship between technological innovation and labor market decisions by examining changes in physician behavior as a result of major technology shift in U.S.
+health care, electronic health records.
+Physician labor market decisions such as retirement, choice of work setting, and productivity potentially have downstream effects for quality of care and access to care for patients.
+I treat EHR implementation in hospitals as exogenous to physicians employed by the hospital and estimate average group time treatment effects on these labor market outcomes.
+I find that physicians of retirement age are more likely to retire, younger physicians are more likely to switch to office settings, and physicians who do not switch work settings see more patients as a results of individual EHR exposure.
+Further, I find that data assistants do not play a meaningful role in physician response to EHR implementation.
+\end{abstract}
 
@@
 
-The physician behaviors that I consider in this analysis are closely intertwined and should not be taken as independent.
-For example, a physician's choice of work setting depends on their previous decision not to retire, and billing activity depends on work setting.
+This section serves to investigate why we might expect physicians to respond to the implementation of EHRs.
+The physician behaviors that I consider in this analysis are closely related and should not be taken as independent.
+For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting.
 Thus, I discuss the behaviors in terms of stages of decision making.
@@
 
+While the majority of the analysis focuses on senior vs.
+young physicians, there are other physician characteristics that are also interesting to investigate.
+Figure \ref{fig:het} presents estimates for different subsets of physicians.
+These include female physicians,those in rural or urban zip codes, and physicians who see majority white patients relative to the entire sample.
+Most of the estimates are similar across these characteristics, with some particularly noisy due to low sample sizes (such as physicians in rural settings).
+The implications implication of this table is that EHRs did in fact lead to behavioral changes in many different physicians, not just a specific subgroup.
+
 \clearpage
@@
     \caption{Heterogeneity Analysis}
-    \includegraphics[scale=.57]{Objects/gr}
-    \label{fig:DA}
+    \includegraphics[scale=.57]{Objects/heterogeneity _graph.pdf}
+    \label{fig:het}
 \end{figure}

### 2022-06-03 Table Code/overall stats.tex
Updates from Overleaf

@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 337176 & 0.02 & 0.13 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 337176 & 0.31 & 0.36 & 0 & 1\\
-\hspace{1em}Work in an Office & 337176 & 0.57 & 0.49 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 313663 & 0.08 & 0.28 & 0 & 1\\
-\hspace{1em}Number of Patients & 337176 & 2274 & 3293 & 0 & 199765\\
+\hspace{1em}Ever Retire & 179883 & 0.03 & 0.18 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 179883 & 0.10 & 0.24 & 0 & 1\\
+\hspace{1em}Work in an Office & 179883 & 0.30 & 0.46 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 161558 & 0.12 & 0.33 & 0 & 1\\
+\hspace{1em}Number of Patients & 179883 & 329.71 & 274.35 & 0 & 7613\\
+\hspace{1em}Claim Count & 179883 & 1382.14 & 1789.34 & 0 & 64508\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 337176 & 0.76 & 0.43 & 0 & 1\\
-\hspace{1em}Exposure to an EHR (Low Integration) & 337176 & 0.51 & 0.50 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 337176 & 0.63 & 0.44 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 179883 & 0.75 & 0.43 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 179883 & 0.63 & 0.44 & 0 & 1\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 337176 & 47.67 & 9.35 & 23 & 89\\
-\hspace{1em}Female & 337176 & 0.31 & 0.46 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 337176 & 1.56 & 0.87 & 1 & 20\\
-\hspace{1em}Number of Systems Worked With & 337176 & 1.34 & 0.62 & 1 & 8\\
+\hspace{1em}Age & 179883 & 44.87 & 8.43 & 25 & 86\\
+\hspace{1em}Female & 179883 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 179883 & 1.49 & 0.82 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 179883 & 1.31 & 0.60 & 1 & 8\\
 \bottomrule
 \end{tabular}
-\label{tab:sumstats1}
+\label{tab:sumstats}
 \end{table}

### 2022-06-03 Table Code/split stats.tex
Updates from Overleaf

@@
 \multicolumn{4}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 0.02 & 0.02\\
-\hspace{1em}Fraction Patients in Office & 0.44 & 0.27 \\
-\hspace{1em}Work in an Office & 0.78 & 0.52 \\
-\hspace{1em}Change Zip Codes & 0.06 & 0.09 \\
-\hspace{1em}Number of Patients & 431.96 & 357.46 \\
-\hspace{1em}Claim Count & 3216.46 & 2019.17\\
+\hspace{1em}Ever Retire & 0.04 & 0.03 & 1.00\\
+\hspace{1em}Fraction Patients in Office & 0.15 & 0.09 & 0.06\\
+\hspace{1em}Work in an Office & 0.48 & 0.28 & 0.18\\
+\hspace{1em}Change Zip Codes & 0.10 & 0.12 & 0.09\\
+\hspace{1em}Number of Patients & 341.34 & 328.12 & 76.16\\
+\hspace{1em}Claim Count & 1762.41 & 1330.00 & 290.68\\
 \addlinespace[0.3em]
 \multicolumn{4}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 0.77 & 0.76 & 0.77\\
-\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.63 & 0.64\\
+\hspace{1em}Exposure to an EHR & 0.76 & 0.75 & 0.75\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.59 & 0.64 & 0.65\\
 \addlinespace[0.3em]
 \multicolumn{4}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 61.07 & 44.06 & 48.29\\
-\hspace{1em}Female & 0.15 & 0.35 & 0.39\\
-\hspace{1em}Number of Hospitals Worked With & 1.67 & 1.53 & 1.35\\
-\hspace{1em}Number of Systems Worked With & 1.40 & 1.33 & 1.19\\
+\hspace{1em}Age & 60.31 & 42.75 & 45.96\\
+\hspace{1em}Female & 0.18 & 0.36 & 0.40\\
+\hspace{1em}Number of Hospitals Worked With & 1.63 & 1.48 & 1.29\\
+\hspace{1em}Number of Systems Worked With & 1.38 & 1.30 & 1.16\\
 \bottomrule

### 2022-06-13 Scripts/analysis.R
update plots for paper

@@ -25,160 +25,6 @@ showtext_auto()
 # Read in the main data created in "data5_MDPPAS.R"
 Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
-# CONSIDERING ESTIMATORS WITH RETIRE AS DEPENDENT VARIABLE ----------------------------------
-
-# TWFE ----------------------------------------------------------------------------------------
-retire_twfe <- feols(retire~ rel_m4+rel_m3+ rel_m2+ rel_0+ rel_p1+
-                         rel_p2+ rel_p3 + rel_p4 | DocNPI + year, 
-                       data=filter(Physician_Data,minyr_EHR>0))
-
-reference=4
-names(reference)<-"-1"
-
-coefplot(list(retire_twfe), keep=c("-4","-3","-2","-1","0",
-                                     "1", "2","3","4"),
-         ref=reference, dict=c("rel_m4"="-4", "rel_m3"="-3", "rel_m2"="-2", "rel_0"="0",
-                               "rel_p1"="1","rel_p2"="2","rel_p3"="3","rel_p4"="4"),
-         main="",
-         zero=TRUE,
-         col=1:2)
-
-wald<-wald(retire_twfe, keep=c("rel_m4","rel_m3","rel_m2"))
-p <- wald$p
-
-mtext(text=paste0("p-value for pre-trends: ",round(p,3)),side=1,line = 3, cex = 0.8, adj = 0)
-
-
-
-
-
-# Stacked Regression (Cengiz et al 2019) ------------------------------------------------------
-
-# This method reframes event study data into groups of sub-experiments with identical event windows that are stacked on top of each other. 
-# One first chooses an event study window that is consistent for all “groups” (different treatment timing). 
-# Then, treatment variables and clean controls are defined for each subgroup. Once these sub-experiments are filtered, they are 
-# all stacked on top of each other. Finally, regress the dependent variable on relative year variables plus relative 
-# year variables interacted with treatment, as well as ID-treatment year fixed effects. Cluster standard errors at 
-# the individual level to account for duplicates in controls. The trade-off for this estimator is choosing an event window 
-# and thus losing groups of treated observations. To compare this estimator with others, I will create a table of estimates based on the 
-# event window chosen. 
-
[…]

### 2022-06-13 Scripts/data4_physician_level.R

@@ -69,7 +69,7 @@ AHAmainsurvey <- AHAmainsurvey %>%
   mutate(ID=as.character(ID)) %>%
   mutate(low_integration=ifelse(IPAHOS==1 | OPHOHOS==1 | CPHOHOS==1,1,0),
          high_integration=ifelse(GPWWHOS==1 | MSOHOS==1 | ISMHOS==1 | EQMODHOS==1,1,0)) %>%
-  rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
+  dplyr::rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
          joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
   select(-IPAHOS, -OPHOHOS, -CPHOHOS, -GPWWHOS, -MSOHOS, -ISMHOS, -EQMODHOS) %>%
   select(-IPAP, -GPWP, -OPHP, -CPHP, -ISMP, -EQMP, -FNDP, -full_year, -EHLTHI, -EHLTHS,
@@ -179,10 +179,10 @@ data <- data %>%
 VA <- data %>%
   mutate(VA=ifelse(str_detect(Hospital_name,"Veteran"),1,0)) %>%
   filter(VA==1) %>%
-  distinct(pairID,VA)
+  distinct(ID,VA)
 
 data <- data %>%
-  left_join(VA,by="pairID") %>%
+  left_join(VA,by="ID") %>%
   filter(is.na(VA)) %>%
   select(-VA)
 

### 2022-06-13 Scripts/sumstats.R

@@ -5,6 +5,7 @@ library(kableExtra)
 library(ggplot2)
 library(readr)
 library(showtext)
+library(panelView)
 
 font_add_google("Cormorant Garamond", "corm")
 
@@ -344,8 +345,7 @@ AHAmainsurvey <- AHAmainsurvey %>% ungroup() %>%
   select(-firstyear_0, -firstyear_1, -firstyear_2, -lastyear_0, -lastyear_1, -lastyear_2)
 
 AHAmainsurvey <- AHAmainsurvey %>%
-  mutate(EHR=ifelse(EHLTH==2,1,ifelse(EHLTH==0 | EHLTH==1,0,NA))) %>%
-  select(-EHLTH)
+  mutate(EHR=ifelse(EHLTH==2,1,ifelse(EHLTH==0 | EHLTH==1,0,NA)))
 
 # In some cases, 2009 or 2015 can be filled in to not have a missing value. 
 # Create dataset to fill in 2009 or 2015 conditionally
@@ -380,18 +380,20 @@ AHAmainsurvey <- AHAmainsurvey %>%
   select(-always_missing)
 
 sample <- AHAmainsurvey %>%
-  distinct(ID) %>%
-  sample_frac(.02) %>%
+  distinct(HospNPI) %>%
+  sample_frac(.02, replace=FALSE) %>%
   mutate(sample=1)
 
+
 AHA_sample <- AHAmainsurvey %>%
-  left_join(sample,by="ID") %>%
+  left_join(sample,by="HospNPI") %>%
   filter(sample==1) %>%
-  select(-sample)
+  select(-sample) 
 
 
 
-panelview(AHAmainsurvey, Y=NULL,D="EHR", index=c("ID","year"), axis.lab = "time", by.timing=TRUE) +
+
+panelview(AHA_sample, Y=NULL,D="EHR", index=c("HospNPI","year"), axis.lab = "time", by.timing=TRUE) +
   theme(text=element_text(size=10,family="lm"))
 
-ggsave("Objects/hosp_treat.pdf")
+ggsave("Objects/test.pdf")

### 2022-06-13 Paper/Paper.tex

@@
 Moreover, shocks to an existing labor market could have downstream effects on customers in that market.
-A prime instance of this is in health care, where technological innovation in health care may contribute to physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or technological change may encourage physician retirement, possibly limiting patient access to care and ultimately affecting patient health.
-The most  widespread and health care-altering technological innovation in recent years took place during the implementation of electronic health records.
-This huge innovation in the health care sector, where there is extensive margin for improvement, is an excellent setting to examine the relationship between innovation and labor market, and the downstream effects this may have.
+A prime instance of this is in health care, where technological innovation may affect patient care indirectly if it induces physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or if it brings about changes in the composition of aggregate physician labor markets in terms of total number of physicians, number of physicians working in rural settings, or number of physicians working in hospitals, which all effect patient access to care.
+The most widespread and health care-altering technological innovation in recent years took place with the implementation of electronic health records.
+Because the technology was incentivized by policy, it was implemented rapidly and on a large scale, making it an opportune setting to examine the relationship between labor markets and technology, and the potential downstream effects that took place.
 In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S.
@@
 Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards other work settings.
+The frustration of using a new technology raises the cost of working in certain settings, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards lower cost settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Using a difference-in-differences research design, I estimate the effects of hospital EHR implementation on physician labor market outcomes.
+Using an event study research design, I estimate the effect of hospital EHR implementation on individual physician labor market outcomes.
 
-My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct a physician-level panel from 2009-2017 consisting of hospitalists and other general practice physicians who work exclusively in hospitals.
-The data captures whether a physician is exposed to EHRs in various hospitals, several labor market outcomes, and other demographic information.
-For all outcomes, the main independent variable of interest is a binary treatment variable capturing a physician's EHR exposure in hospitals.
+My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct an individual-level panel from 2009-2017 consisting of physicians who work primarily in a hospital, known as hospitalists.
+The data captures whether such physicians are exposed to EHRs in their associated hospital(s), several labor market outcomes, and other demographic information.
+For all outcomes, the main independent variable of interest is a binary treatment variable capturing exposure to an EHR.
 In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
@@
 
-My analysis contributes to the literature examining effects of health information technology on health care.
-There is a robust group of empirical papers that examine the effects of EHRs on patient outcomes and hospital costs prior to the HITECH Act taking effect.
+This paper contributes to the literature's understanding of how health information technology generally affects health care.
+There is a robust set of empirical papers examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
 Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
-More relevant to EHR effects on physicians, a number of studies consider the effect of EHRs on productivity in specific settings.
+More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
 One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).
 
-I expand and improve upon this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined.
+I expand this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined.
 This is the first empirical study that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups.
@@
 This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects, and further provides insight to short vs.
+long term effects taking place.
 
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this work, I find that EHR implementation led to a .005 ppt increase (25\% increase relative to the mean percentage) in the likelihood of a senior physician retiring in the year after implementation.
-Younger physicians (less than 60) are .002 ppts more likely to retire in the second year after exposure, possibly pointing to career switching.
-Whether physicians are retirement age or not, these results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
-In those who did not retire, I find that younger physicians are .015 ppts more likely to work in an office in the year of exposure to an EHR, but that exposure does not change the average fraction of patients seen in office settings for either age group.
-Finally, I find that both the total number of patients and billing activity increases in younger physicians in the first 2-3 years after exposure to an EHR.
-The ratio of billing activity to patient count increases less than the sample average, suggesting that billing activity did not dis-proportionally increase.
+Contrary to this work, I find that EHR implementation led to a 0.006 ppt (s.e.
+0.002) increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
+Physicians less than 60 were 0.002 ppts (s.e.
+0.0009) more likely to retire after exposure, possibly pointing to career switching since retirement is measured as a lack of Medicare patients seen.
+These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
+For those who did not retire, I find no evidence of switching towards office settings in either age group.
+However, both age groups are more likely to change zip codes (by 0.045 ppts), suggesting across-hospital switching.
+Finally, I find a modest, positive effect on patient count (an increase of 14 patients relative to a mean of 330) completely driven by younger physicians, and no effect on billing activity for either age group.
+Thus, for younger physicians, EHRs allow the claim count to patient ratio to decrease, which, in the best case scenario, points to a decrease in wasteful care being billed for.
 
 A key assumption that underlies the main analysis is that physicians are constrained to using the technology in a hospital who chooses to adopt.
-I consider two possibilities where this assumption is violated.
+I consider two scenarios where this assumption would be violated.
 First, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
-If this is occurring, the results are at least partially driven by the hiring of new employees.
-I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the majority of EHR implementation.
+If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence.
+I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast EHR implementation.
 Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
@@
 
-This section serves to investigate why we might expect physicians to respond to the implementation of EHRs.
+This section serves to outline the potential responses of physicians who work primarily in hospitals to EHRs based on the costs and incentives they face.
 The physician behaviors that I consider in this analysis are closely related and should not be taken as independent.
 For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting.
-Thus, I discuss the behaviors in terms of stages of decision making.
+Thus, I discuss the behaviors in stages of decision making.
 
@@
 Workers in developed countries tend to plan for formal retirement well in advance.
-Generally, an exogenous shock to a worker is not expected to change retirement age simply because of the amount of wealth and planning necessary to formally leave the labor force.
+Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of wealth and planning necessary to formally leave the labor force.
 Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}).
@@
 It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
-First, I will walk through the incentives of a physician who is already of retirement age when the EHR is implemented.
-While payment schemes, income as a whole, and retirement benefits certainly impact the utility a physician receives from working or retiring, I can model the decision as independent of those things since I assume the physician already has the financial mean to retire.
-Then, a physician will only retire if the utility from working is less than the utility from retiring, assuming the physician keeps working under indifference.
-How does exogenous EHR implementation affect these utilities?
-The utility received in retirement will not be affected by whether the hospital is using an EHR after the physician is already gone.
-However, the utility received from working is a function of work environment.
-If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility.
+First, I consider physicians already of retirement age when a new technology is implemented.
+While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire.
+Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life.
+Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction.
+If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work.
 If EHRs make daily tasks easier, there is a positive relationship.
-Thus, for a physician on the margin of retirement, the decision to retire only happens as a result of EHRs if they impose a cost, or a negative impact on work environment.
-Further, the cost has to be high enough to raise utility from retirement above utility from working, counteracting the altruism that is keeping the physician from retiring.
+Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
+Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), suggesting a negative impact.
 
@@
 While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching.
-This decision is impacted similarly to the above discussion regarding formal retirement, but is also impacted by income that would be received as a result of career switching.
-
-The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
+This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how EHRs impact job satisfaction of current clinical work.
+As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
+Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
 A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
 
-As an exogenous shock to work environment, EHR implementation may lead to an earlier retirement than would have otherwise occurred.
-Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
-The aim of this section is to test that hypothesis and evaluate whether younger physicians exhibited similar behavior.
+\subsection{Work Setting}
 
-\subsection{Work Setting}
+For most physicians, EHRs will not impose a costly enough burden to induce retirement.
+However, there are other ways physicians might change their behavior under sufficient incentive to avoid the technology.
+For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over the technologies used in the work environment, affecting overall job satisfaction.
+Generally, physicians may have the ability to move hospitals, move from a hospital to an office, or work in multiple facilities and shift the composition of patients towards other facilities.
+I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}).
+Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
+
+First, I consider those who are in a relatively binding contract.
+For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital.
+This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings.
+A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set.
+Since these hospitalists are established in their hospital, they work a significant amount of time there.
+When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.
+Alternatively, there are hospitalists that work in multiple hospitals or work in both hospitals and offices.
+If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.
+
+\subsection{Productivity}
+
+There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation.
+A lack of response could be due to 3 factors: EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
+For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment.
+An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive.
+A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
+Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.
 
@@
 My analysis relies on EHR implementation in hospitals.
-Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, pediatrics (although Medicare primarily enrolls the elderly, Medicare also covers disabled children), general practice or family practice in at least one year.
+Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
 However, I exclude physicians who list themselves as a specialist in all but one year.
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years.
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years and I drop physicians with insufficient claim counts to show indications of behavior.
 
@@
 
-I show summary statistics for the entire sample in Table \ref{tab:sumstats1}.
-Only 2\% of physicians (approximately 750) retire over the course of the panel.
-Just over half of the physician-year observations have value one for working in an office in some capacity, with about a third of the number of patients being seen in an office setting.
+I show summary statistics for the entire sample in Table \ref{tab:sumstats}.
+Only 3\% of physicians (approximately 770) retire over the course of the panel.
+30\% of the physician-year observations have value one for working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting.
 By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure.
 Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015.
-The average physician is 48 years old, works with 1.5 hospitals and 1.3 systems.
+The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems.
 
@@
 
-In this setting, treatment effects are likely heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
+I seek to estimate average treatment effects of EHR exposure on various physician outcomes.
+I treat EHR implementation as an exogenous treatment.
+In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
 Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$:
@@
 A group indicates all physicians treated, or first exposed to an EHR, in the same year.
-To estimate the heterogeneous treatment effects, I employ the estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
 Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}.
+I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}.
+These results yield similar patterns with slightly smaller magnitudes.
 
@@
 
-For most physicians, EHRs will not impose a costly enough burden to induce retirement.
-However, there are other ways physicians might change their labor market behavior under sufficient incentive to avoid the technology.
  I consider the decision to switch work settings using multiple outcomes.
@@
 \subsection{Patient Count and Billable Activity}\label{sec:patientcount}
-
-An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive.
-A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
-Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
-Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.
-In this section, I speak to efficiency by estimating the effect of EHR implementation on patient count.
 
@@
     \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.55]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.65]{Objects/sum_stats_year.pdf}
     \label{fig:treatmentgraph}
@@
     \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.45]{Objects/retire_plot.pdf}
+    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.45]{Objects/officeind_plot.pdf}
+    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
     \label{fig:officefirst}
@@
     \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
-    \includegraphics[scale=.45]{Objects/officefrac_plot.pdf}
+    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
     \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.45]{Objects/zip_plot.pdf}
+    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
     \label{fig:zip}
@@
     \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.45]{Objects/patient_plot.pdf}
+    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
     \label{fig:patient}
@@
     \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
-    \includegraphics[scale=.45]{Objects/claim_plot.pdf}
+    \includegraphics[scale=.65]{Objects/claim_plot.pdf}
     \label{fig:claim}
@@
 I then categorize NPIs by key words in their tax code description.
-Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", ``General Practice", or ``Pediatrics" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital.
+Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital.
 I create two data sets, one containing only PCPs and one containing only hospitals.
@@
 
+Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. These results from each outcome are presented in
+
 One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}).
@@
 
+\begin{figure}
+    \centering
+    \includegraphics{objects}
+    \caption{Caption}
+    \label{fig:my_label}
+\end{figure}
+
 \begin{figure}[ht]

### 2022-06-14 Scripts/analysis.R
update appendix

@@ -172,18 +172,66 @@ ggsave(file="Objects/patient_plot.pdf",plot=plots[[5]], width=10, height=7, unit
 ggsave(file="Objects/claim_plot.pdf",plot=plots[[6]], width=10, height=7, units="in")
 
 
-# Save specific values to reference in paper
-cat(round(Models_agg[[1]][["agg_all"]][["overall.att"]],3),file="Results.tex")
-
-
-observe <- Physician_Data %>%
-  filter(ever_retire==1) %>%
-  select(year, phy_zip1, DocNPI, age, retire, pos_office, claim_count_total)
+#Save plot of patient count dis-aggregated
+ggdid(models[[5]][["all"]]) + scale_color_manual(labels = c("Pre", "Post"), values=c("#999999", "#E69F00"),name="") +
+  theme(legend.position = "none", text = element_text(size = 12, family="lm")) + labs(title="Dis-aggregated Effects of EHR Exposure on Patient Count")
+ggsave(filename = "Objects/patient_group.pdf", width=10, height=12, units="in")
 
 
 
+# Create data frame of overall ATT values
 
+ATT_all <- lapply(Models_agg, function(x){
+  c(x[["agg_all"]][["DIDparams"]][["yname"]],
+    round(x[["agg_all"]][["overall.att"]],5), 
+    round(x[["agg_all"]][["overall.se"]], 5),
+    round(x[["agg_all"]][["overall.att"]]-(1.959*x[["agg_all"]][["overall.se"]]),5),
+    round(x[["agg_all"]][["overall.att"]]+(1.959*x[["agg_all"]][["overall.se"]]),5),
+    "Any",
+    "Main")
+})
 
+ATT_young <- lapply(Models_agg, function(x){
+  c(x[["agg_young"]][["DIDparams"]][["yname"]],
+    round(x[["agg_young"]][["overall.att"]],5), 
+    round(x[["agg_young"]][["overall.se"]], 5),
+    round(x[["agg_young"]][["overall.att"]]-(1.959*x[["agg_young"]][["overall.se"]]),5),
+    round(x[["agg_young"]][["overall.att"]]+(1.959*x[["agg_young"]][["overall.se"]]),5),
+    "< 60",
+    "Main")
+})
 
+ATT_old <- lapply(Models_agg, function(x){
+  c(x[["agg_old"]][["DIDparams"]][["yname"]],
+    round(x[["agg_old"]][["overall.att"]],5), 
[…]

### 2022-06-14 Scripts/sensitivity_analysis.R

@@ -42,7 +42,7 @@ models_LI <- lapply(varlist, function(x) {
          idname = "DocNPI",
          tname = "year",
          xformla = ~grad_year,
-         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0) else dplyr::filter(Physician_Data,minyr_EHR_int>0),
+         data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR_int>0) else (if (x=="pos_office" | x=="work_in_office" | x=="change_zip") dplyr::filter(Physician_Data, minyr_EHR_int>0 & ever_retire==0) else dplyr::filter(Physician_Data, minyr_EHR_int>0 & ever_retire==0 & never_newnpi==1)),
          est_method = "dr",
          control_group = "notyettreated")
   
@@ -54,7 +54,7 @@ models_LI_young <- lapply(varlist, function(x) {
          idname = "DocNPI",
          tname = "year",
          xformla = ~grad_year,
-         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0 & max_age<60) else dplyr::filter(Physician_Data,minyr_EHR_int>0 & max_age<60),
+         data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & max_age<60) else (if (x=="pos_office" | x=="work_in_office" | x=="change_zip") dplyr::filter(Physician_Data, minyr_EHR_int>0 & ever_retire==0 & max_age<60) else dplyr::filter(Physician_Data, minyr_EHR_int>0 & ever_retire==0 & never_newnpi==1 & max_age<60)),
          est_method = "dr",
          control_group = "notyettreated")
   
@@ -66,7 +66,7 @@ models_LI_old <- lapply(varlist, function(x) {
          idname = "DocNPI",
          tname = "year",
          xformla = ~grad_year,
-         data = if (x!= "retire") dplyr::filter(Physician_Data,minyr_EHR_int>0 & ever_retire==0 & max_age>=60) else dplyr::filter(Physician_Data,minyr_EHR_int>0 & max_age>=60),
[…]

### 2022-06-14 Paper/Paper.tex

@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
+    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
+    \caption*{Source: HealthIT.gov}
+    \label{fig:meanuse}
+\end{figure}
+
 Figure \ref{fig:EPIC} shows a two computer screenshots of what physicians might see in their EHR system.
@@
 Physicians have been voicing concerns over EHR usability since their beginning, indicating that they are likely responding in their behavior as well.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Screenshot of EHR System}
+    \includegraphics[scale=.4]{Objects/epic-ehr-screenshot.jpg}
+    \includegraphics[scale=.115]{Objects/EHRimage2.jpg}
+    \label{fig:EPIC}
+\end{figure}
 
@@
 The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems.
+\import{Table Code}{overall stats.tex}
 
@@
 The age groups are exposed to EHRs in similar ways.
+
+\import{Table Code}{split stats.tex}
 
@@
 
+\begin{figure}[t]
+\centering
+    \caption{Treatment Variables Over Time}
+    \includegraphics[scale=.45]{Objects/sum_stats_year.pdf}
+    \label{fig:treatmentgraph}
+\end{figure}
+
 \section{Empirical Strategy}
 
-I seek to estimate average treatment effects of EHR exposure on various physician outcomes.
-I treat EHR implementation as an exogenous treatment.
+I treat EHR implementation as an exogenous treatment variable, and I seek to estimate average treatment effects of EHR exposure on various physician outcomes.
 In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
-Therefore, to avoid negative weighting issues this causes in a two way fixed effects specification, I estimate average treatment effects for a specific group $g$ at time $t$:
+Therefore, to avoid negative weighting issues this causes in a classical two way fixed effects event study specification, I instead estimate average treatment effects for a specific group $g$ at time $t$:
 $$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
@@
 
-There are several assumptions necessary to identify $ATT(g,t)$.
+Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot.
+When the dis-aggregated estimates are of value, I present those as well.
+Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each table.
+Simultaneous confidence bands, calculated using bootstrapped standard errors, are presented.
+
+\subsection{Assumptions}
+
+There are several assumptions necessary to identify the parameters of interest, $ATT(g,t)$.
 First, I assume that treatment is not reversed once it occurs.
@@
 They may add features or switch vendors, but do not reverse EHR use.
-This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a random sample of hospitals over time.
+This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a 2\% random sample of hospitals over time\footnote{Note that out of the 81 hospitals included in the random sample, 4 of them show a reversal of treatment status.
+Upon further inspection, each of these hospitals show such a reversal due to a discrepancy in the coding of whether the hospital uses an EHR partially or fully.
+Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not.
+Hospitals with this discretion arre coded as implementing an EHR in the first year that they answered "fully" to the survey question.}.
 Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
@@
 I investigate this possibility and limit the data to address it in Section \ref{sec:patientcount}.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Hospital EHR Adoption Over Time}
+    \includegraphics[scale=.7]{Objects/hosp_treat.pdf}
+    \label{fig:hosp_treat}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Retirement}
+    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
+    \label{fig:retirefirst}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0008), 0.002 (0.0009), 0.005 (0.002), respectively.}}
+\end{figure}
+
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
+    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
+    \label{fig:officefirst}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0004 (0.008), -0.004 (0.008), 0.02 (0.02), respectively.}}
+\end{figure}
+
 If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings.
@@
 I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
+    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
+    \label{fig:officesecond}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0003 (0.004), -0.0007 (0.005), -0.006 (0.018), respectively.}}
+\end{figure}
 
@@
 Thus, I make no claim that EHR exposure is leading physicians to switch zip codes.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
+    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
+    \label{fig:zip}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.045 (0.006), 0.046 (0.007), 0.042 (0.018), respectively.}}
+\end{figure}
 
@@
 While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is not largely meaningful.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Patient Count}
+    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
+    \label{fig:patient}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 14.1 (4.6), 15.8 (5.4), -0.29 (13.0), respectively.}}
+\end{figure}
 
@@
 \section{Tables and Figures}
-
-\begin{figure}[htp]
-    \centering
-    \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
-    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
-    \caption*{Source: HealthIT.gov}
-    \label{fig:meanuse}
-\end{figure}
-
-\begin{figure}[htp]
-    \centering
-    \caption{Screenshot of EHR System}
-    \includegraphics[scale=.5]{Objects/epic-ehr-screenshot.jpg}
-    \includegraphics[scale=.135]{Objects/EHRimage2.jpg}
-    \label{fig:EPIC}
-\end{figure}
-
-\import{Table Code}{overall stats.tex}
-
-\import{Table Code}{split stats.tex}
-
-\begin{figure}[p]
-\centering
-    \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.65]{Objects/sum_stats_year.pdf}
-    \label{fig:treatmentgraph}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \caption{Hospital EHR Adoption Over Time}
-    \includegraphics[scale=.8]{Objects/hosp_treat.pdf}
-    \label{fig:hosp_treat}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
-    \label{fig:retirefirst}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
-    \label{fig:officefirst}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
-    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
-    \label{fig:officesecond}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
-    \label{fig:zip}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
-    \label{fig:patient}
-\end{figure}
 
@@
 
-Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. These results from each outcome are presented in
+Before I present these estimators, I present results from a two way fixed effects setup that may suffer bias from negative weighting. Specifically, I estimate the following equation:
+$$\text{outcome}_{i,t}=\sum_{j=-4}^{-2} \text{rel\_year}_{j} + \sum_{j=0}^{4} \text{rel\_year}_{j} + \delta_i + \gamma_t,$$
+where outcome$_{i,t}$ is one of the six outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to expansion, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
+These results from each outcome are presented in Figure \ref{fig:twfe}. These results show similar patterns to the main specification, with slight variations in magnitude.
 
@@
     \centering
-    \includegraphics{objects}
-    \caption{Caption}
-    \label{fig:my_label}
+    \caption{Results: Two Way Fixed Effects}
+    \includegraphics[scale=.8]{Objects/twfe_plot.pdf}
+    \label{fig:twfe}
 \end{figure}
@@
     \caption{Limited Sample Years}
-    \includegraphics[scale=.57]{Objects/years_graph.pdf}
+    \includegraphics[scale=.5]{Objects/years_graph.pdf}
     \label{fig:years}

### 2022-06-14 Scripts/sensitivity_analysis.R
update appendix

@@ -127,7 +127,7 @@ LI_values_old <- as.data.frame(do.call(rbind, ATT_LI_old)) %>%
   dplyr::rename(Variable=V1, ATT=V2, SE=V3, Lower=V4, Upper=V5, Age=V6, Specification=V7)
   
 # Merge all ages and main specification results
-LI_merged <- rbind(LI_values, LI_values_young, LI_values_old) %>%
+LI_merged <- rbind(LI_values, LI_values_young, LI_values_old, main_ATT) %>%
   dplyr::mutate(ATT=as.numeric(ATT),
                 Lower=as.numeric(Lower),
                 Upper=as.numeric(Upper)) %>%
@@ -142,25 +142,25 @@ LI_merged <- rbind(LI_values, LI_values_young, LI_values_old) %>%
 # Create Graph
 dodge <- position_dodge(width=.75)
 small_values <- ggplot(dplyr::filter(LI_merged, Variable!="Claim Count" & Variable!="Number Patients" & Age=="Any"),
-                       aes(x=Variable, y=ATT)) +
+                       aes(x=Variable, y=ATT, color=Specification)) +
   geom_point(position=dodge) +
   geom_errorbar(aes(ymax=Upper,ymin=Lower, width=.5),position = dodge) + 
   geom_hline(yintercept=0, linetype="dashed", size=.1) +
   coord_flip() + theme_bw() + 
   theme(text=element_text(size=17, family="lm")) +
   xlab(" ") + ylab(" ") +
-  scale_color_manual(values=c("#999999", "#E69F00"))
+  scale_color_manual(values=c(Main="#999999", "Low-Integration"="#E69F00"))
 
 
 large_values <- ggplot(dplyr::filter(LI_merged, (Variable=="Claim Count" | Variable=="Number Patients") & Age=="Any"),
-                       aes(x=Variable, y=ATT)) +
+                       aes(x=Variable, y=ATT, color=Specification)) +
   geom_point(position=dodge) +
   geom_errorbar(aes(ymax=Upper,ymin=Lower, width=.3),position = dodge)  +
   geom_hline(yintercept=0, linetype="dashed", size=.1) +
   coord_flip() + theme_bw() + 
   theme(text=element_text(size=17, family="lm")) +
   xlab(" ") + ylab("\nATT and 95% CI") +
-  scale_color_manual(values=c("#999999", "#E69F00"))
[…]

### 2022-06-14 Paper/Paper.tex
Updates from Overleaf

@@
 In this case, the effects seen could be from having the data assistant instead of EHR use.
-I investigate this in Section \ref{sec:dataass}.
+Thus, I investigate hteir existence further.
+
+\subsection{Hiring Data Assistants}\label{sec:dataass}
+
+One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals, such as the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
+Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
+they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+I will refer to any employee in these categories as a data assistant.
+In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
+
+First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
+Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
+The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
+From 2005 to 2013, 15-60 more employees were registered each year.
+The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
+If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
+The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
+
+\begin{figure}[ht]
+\centering
+\captionsetup{width=.5\linewidth}
+\caption{Frequency of Data Assistant Enumeration by Year}
+\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+\label{fig:dataassistant_histogram}
+\vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Histtogram uses NPPES data on the enumeration of certain types of NPIs to estimate the number of new data assistants being hired in each year of my sample.}}
+\end{figure}
+
+Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
+I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
+I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
+I find that .1\% of physician-year observations are associated with data assistants.
+This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
+However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
+Details of this analysis are given in Appendix \ref{app:DA}.
 
@@
 
+For both patient and claim count, if the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
+Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
+These physicians do not retire or see patients in another hospital.
+Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period.
+
 The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
@@
 
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Patient Count by Group}
+    \includegraphics[scale=.47]{Objects/patient_group.pdf}
+    \label{fig:patientgroup}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Plot shows average group time treatment effects for each treatment group.}}
+\end{figure}
+
 Total patient count is directly related to the fraction of patients seen in an office setting.
@@
 
-However, for both patient and claim count, the results include physicians who shift place of work, which may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
-Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
-These physicians do not retire or see patients in another hospital.
-Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period.
-I present the results for this limited sample in Figure \ref{fig:limitedsample}.
-These results are similar to those presented for the full sample, suggesting that physicians become more productive in terms of patient count in the first two years after implementation.
+\begin{figure}[ht]
+    \centering
+    \caption{Effect of EHR Exposure on Claim Count}
+    \includegraphics[scale=.65]{Objects/claim_plot.pdf}
+    \label{fig:claim}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
+\end{figure}
 
@@
 
-\section{Physician EHR Avoidance}
-
-\subsection{Hiring Data Assistants}\label{sec:dataass}
-
-One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals, such as the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
-Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
-I will refer to any employee in these categories as a data assistant.
-In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
-
-First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
-Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
-The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
-From 2005 to 2013, 15-60 more employees were registered each year.
-The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
-If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
-The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
-
-Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
-I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
-I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
-I find that .1\% of physician-year observations are associated with data assistants.
-This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
-However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
-Details of this analysis are given in Appendix \ref{app:DA}.
-
-\subsection{Physician Influence of EHR Decision in Hospitals}
+\section{Physician Influence of EHR Decision in Hospitals}
 
@@
 This contribution is relevant to EHR vendors, hospitals, and policymakers who are involved in further implementation of complex technology.
-
-\section{Tables and Figures}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Patient Count by Group}
-    \includegraphics[scale=.35]{Objects/patient_group.pdf}
-    \label{fig:patientgroup}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Claim Count, Limited Sample}
-    \includegraphics[scale=.65]{Objects/claim_plot.pdf}
-    \label{fig:claim}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \caption{Effect of EHR Exposure on Patient Count, Limited Sample}
-    \includegraphics[scale=.45]{Objects/limitedsample_plot.pdf}
-    \label{fig:limitedsample}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\caption{Frequency of Data Assistant Enumeration by Year}
-\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
-\label{fig:dataassistant_histogram}
-\end{figure}
 

### 2022-06-16 Scripts/analysis.R
add pretrend sensativity analysis from Roth

@@ -43,7 +43,8 @@ models <- lapply(varlist, function(x) {
     est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
     control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
     clustervars = "DocNPI",          # Cluster Variables          
-    anticipation=0                   # can set a number of years to account for anticipation effects
+    anticipation=0,
+    base_period = "universal" # can set a number of years to account for anticipation effects
   )
   
   young <- att_gt(yname = x,                # LHS Variable
@@ -56,7 +57,8 @@ models <- lapply(varlist, function(x) {
          est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
          control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
          clustervars = "DocNPI",          # Cluster Variables          
-         anticipation=0                   # can set a number of years to account for anticipation effects
+         anticipation=0,
+         base_period = "universal" # can set a number of years to account for anticipation effects
   )
   
   old <- att_gt(yname = x,                # LHS Variable
@@ -69,7 +71,8 @@ models <- lapply(varlist, function(x) {
          est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
          control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
          clustervars = "DocNPI",          # Cluster Variables          
-         anticipation=0                   # can set a number of years to account for anticipation effects
+         anticipation=0,
+         base_period = "universal" # can set a number of years to account for anticipation effects
   )
   
   list(all=all,young=young,old=old)

### 2022-06-16 Scripts/sensitivity_analysis.R

@@ -5,6 +5,8 @@ library(ggpubr)
 library(fixest)
 library(did2s)
 library(readxl)
+library(HonestDiD)
+library(mvtnorm)
 
 
 
@@ -1290,7 +1292,317 @@ ggsave("Objects/twfe_plot.pdf", twfe_finalplot, height=9.5, width=8, units = "in
 
 
 
+## Roth Prettrends Stuff ---------------------------------- ####
 
+#(Need results from analysis loaded)
+
+honest_did <- function(es, ...) {
+  UseMethod("honest_did", es)
+}
+
+honest_did.AGGTEobj <- function(es,
+                                e=0,
+                                type=c("smoothness", "relative_magnitude"),
+                                method=NULL,
+                                bound="deviation from parallel trends",
+                                Mvec=NULL,
+                                Mbarvec=NULL,
+                                monotonicityDirection=NULL,
+                                biasDirection=NULL,
+                                alpha=0.05,
+                                parallel=FALSE,
+                                gridPoints=10^3,
+                                grid.ub=NA,
+                                grid.lb=NA,
+                                ...) {
+  
+  
+  type <- type[1]
+  
+  # make sure that user is passing in an event study
+  if (es$type != "dynamic") {
+    stop("need to pass in an event study")
+  }
+  
+  # check if used universal base period and warn otherwise
+  if (es$DIDparams$base_period != "universal") {
+    warning("it is recommended to use a universal base period for honest_did")
+  }
+  
+  # recover influence function for event study estimates
+  es_inf_func <- es$inf.function$dynamic.inf.func.e
+  
+  # recover variance-covariance matrix
+  n <- nrow(es_inf_func)
+  V <- t(es_inf_func) %*% es_inf_func / (n*n) 
+  
+  
+  nperiods <- nrow(V)
+  npre <- sum(1*(es$egt < 0))
+  npost <- nperiods - npre
+  
+  baseVec1 <- basisVector(index=(e+1),size=npost)
+  
+  orig_ci <- constructOriginalCS(betahat = es$att.egt,
+                                 sigma = V, numPrePeriods = npre,
+                                 numPostPeriods = npost,
[…]

### 2022-06-16 Paper/Paper.tex
Updates from Overleaf

@@
     \centering
+    \captionsetup{width=.6\linewidth}
     \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
@@
     \centering
+    \captionsetup{width=.4\linewidth}
     \caption{Screenshot of EHR System}
@@
 \centering
+\captionsetup{width=.45\linewidth}
     \caption{Treatment Variables Over Time}
@@
     \centering
+    \captionsetup{width=.7\linewidth}
     \caption{Hospital EHR Adoption Over Time}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Retirement}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Patient Count}
@@
     \centering
+    \captionsetup{width=.6\linewidth}
     \caption{Effect of EHR Exposure on Patient Count by Group}
@@
     \centering
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Claim Count}
@@
 
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.57\linewidth}
+    \caption{Comparison of ATT, Low-Integration Treatment Variable}
+    \includegraphics[scale=.57]{Objects/LI_graph.pdf}
+    \label{fig:LI}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects are shown for each outcome variable, along with 95\% confidence intervals.
+Low-integration hospitals are defined by hospital management.}}
+\end{figure}
+
 \subsection{Anticipation}\label{app:anticipation}
@@
 The results are virtually identical, just noisier, compared to the estimates from the main specification.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.57\linewidth}
+    \caption{ATT With Anticipation}
+    \includegraphics[scale=.57]{Objects/anticipation_graph.pdf}
+    \label{fig:anticipation}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects are shown for each outcome variable, along with 95\% confidence intervals.}}
+\end{figure}
 
@@
 
+\begin{figure}
+    \centering
+    \captionsetup{width=.8\linewidth}
+    \caption{Results: Two Way Fixed Effects}
+    \includegraphics[scale=.8]{Objects/twfe_plot.pdf}
+    \label{fig:twfe}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Event study plots from two way fixed effects estimation are shown for each outcome variable.}}
+\end{figure}
+
 One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}).
@@
 Generally, the results have the same sign with a slightly larger magnitude, driven by the inclusion of never-treated units.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.57\linewidth}
+    \caption{Other Event Study Estimators}
+    \includegraphics[scale=.57]{Objects/estimators_graph.pdf}
+    \label{fig:esimators}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Results from various methods of estimation are shown, along with 95\% confidence intervals.}}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.57\linewidth}
+    \caption{Limited Sample Years}
+    \includegraphics[scale=.57]{Objects/years_graph.pdf}
+    \label{fig:years}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and a limited sample of years.}}
+\end{figure}
+
 \subsection{Physicians not Exposed to Data Assistants}\label{app:DA}
@@
 
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.57\linewidth}
+    \caption{Physicians without Data Assistants}
+    \includegraphics[scale=.57]{Objects/DA_graph.pdf}
+    \label{fig:DA}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and limited sample of physicians who do not work with data assistants.}}
+\end{figure}
+
 \section{Heterogeneity Analysis}
@@
 
-\clearpage
-
-\section{Tables and Figures}
-
 \begin{figure}[ht]
     \centering
-    \caption{Comparison of ATT, Low-Integration Treatment Variable}
-    \includegraphics[scale=.57]{Objects/LI_graph.pdf}
-    \label{fig:LI}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \caption{ATT With Anticipation}
-    \includegraphics[scale=.57]{Objects/anticipation_graph.pdf}
-    \label{fig:anticipation}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \caption{Results: Two Way Fixed Effects}
-    \includegraphics[scale=.8]{Objects/twfe_plot.pdf}
-    \label{fig:twfe}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \caption{Other Event Study Estimators}
-    \includegraphics[scale=.57]{Objects/estimators_graph.pdf}
-    \label{fig:esimators}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \caption{Limited Sample Years}
-    \includegraphics[scale=.5]{Objects/years_graph.pdf}
-    \label{fig:years}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \caption{Physicians without Data Assistants}
-    \includegraphics[scale=.57]{Objects/DA_graph.pdf}
-    \label{fig:DA}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
+    \captionsetup{width=.65\linewidth}
     \caption{Heterogeneity Analysis}
-    \includegraphics[scale=.57]{Objects/heterogeneity _graph.pdf}
+    \includegraphics[scale=.65]{Objects/heterogeneity _graph.pdf}
     \label{fig:het}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and various limited samples of physicians.}}
 \end{figure}

### 2022-08-04 Paper/Paper.tex
Updates from Overleaf

@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{setspace}
-%\singlespacing
-\doublespacing
+\onehalfspacing
 \addtolength{\skip\footins}{2pc plus 5pt}
@@
  \geometry{,
- left=20mm,
- top=20mm,
- right=20mm,
- bottom=20mm
+ left=25.4mm,
+ top=25.4mm,
+ right=25.4mm,
+ bottom=25.4mm
  }
@@
 }
-
+\usepackage{times}
 \begin{document}
@@
 health care, electronic health records.
-Physician labor market decisions such as retirement, choice of work setting, and productivity potentially have downstream effects for quality of care and access to care for patients.
-I treat EHR implementation in hospitals as exogenous to physicians employed by the hospital and estimate average group time treatment effects on these labor market outcomes.
+I treat EHR implementation in hospitals as exogenous to physicians employed by the hospital and estimate average group time treatment effects on various labor market outcomes.
 I find that physicians of retirement age are more likely to retire, younger physicians are more likely to switch to office settings, and physicians who do not switch work settings see more patients as a results of individual EHR exposure.
-Further, I find that data assistants do not play a meaningful role in physician response to EHR implementation.
 \end{abstract}
@@
 
-How technological innovation affects labor markets is often controversial, in part because of the vast differences in technology and tasks present even within industries.
-Moreover, shocks to an existing labor market could have downstream effects on customers in that market.
-A prime instance of this is in health care, where technological innovation may affect patient care indirectly if it induces physician burnout, which has been shown to affect patient satisfaction (\cite{shanafelt2002burnout}), or if it brings about changes in the composition of aggregate physician labor markets in terms of total number of physicians, number of physicians working in rural settings, or number of physicians working in hospitals, which all effect patient access to care.
+How technological innovation affects labor markets is often debated, in part because of the vast differences in technology and tasks present even within industries.
 The most widespread and health care-altering technological innovation in recent years took place with the implementation of electronic health records.
-Because the technology was incentivized by policy, it was implemented rapidly and on a large scale, making it an opportune setting to examine the relationship between labor markets and technology, and the potential downstream effects that took place.
-In this paper, I investigate physician labor market changes as a result of a major technology shift in the U.S.
-health care system, the implementation of electronic health records.
+Because the technology was incentivized by policy, it was implemented rapidly and on a large scale, making it an opportune setting to examine the relationship between labor markets and technology.
+In this paper, I investigate physician labor market changes as a result of the implementation of electronic health records.
 
@@
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Using an event study research design, I estimate the effect of hospital EHR implementation on individual physician labor market outcomes.
+Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market outcomes.
 
@@
 In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
-I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and list of primary zip codes, (3) number of patients seen, and (4) billing activity.
+I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and location of the physician office, (3) number of patients seen, and (4) billing activity.
 
@@
 There is a robust set of empirical papers examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased costs) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that the median patient experiences no change due to EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased cost) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
 More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
@@
 
-I expand this literature, first by considering a different mechanism by which EHRs could indirectly affect patients through various physician responses, which, to my knowledge, has not yet been examined.
-This is the first empirical study that connects EHR implementation to physician labor market decisions, and focuses on the difference in response across age groups.
+I expand this literature, first by considering a different mechanism by which EHRs could affect physician labor markets (and, indirectly, patients), which, to my knowledge, has not yet been examined.
+This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups.
 Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
@@
 Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects, and further provides insight to short vs.
-long term effects taking place.
+long term effects.
 
@@
 Finally, I find a modest, positive effect on patient count (an increase of 14 patients relative to a mean of 330) completely driven by younger physicians, and no effect on billing activity for either age group.
-Thus, for younger physicians, EHRs allow the claim count to patient ratio to decrease, which, in the best case scenario, points to a decrease in wasteful care being billed for.
+Thus, for younger physicians, EHRs are leading to a decrease in billing activity per patient driven by an increase in patients.
 
@@
 Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
-Second, I investigate whether the decision for hospitals to implement EHRs is exogenous to individual physicians.
+Second, I assume the decision for hospitals to implement EHRs is exogenous to individual physicians.
 That is, a physician does not choose their own EHR exposure.
@@
 For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a lower probability that EHR implementation is endogenous.
-The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint decision making among physicians and hospitals regarding EHR adoptions.
+The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.
 
@@
 The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines.
-The guidelines include having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy.
+The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy.
 The U.S.
@@
 
-Figure \ref{fig:EPIC} shows a two computer screenshots of what physicians might see in their EHR system.
+Figure \ref{fig:EPIC} shows two computer screenshots of what physicians might see in their EHR system.
 On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}).
@@
 
-\section{Physician Incentives}
+\section{EHR and Physician Labor Markets}
 
@@
 For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment.
-An important, and heavily debated, aspect of the implementation of EHRs is whether or not they allow users to be more productive.
+An important aspect of the implementation of EHRs is whether or not they allow users to be more productive.
 A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
@@
 Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
-I first limit the sample of physicians based on specialty type.
-My analysis relies on EHR implementation in hospitals.
-Thus, I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
+I first limit the sample of physicians based on specialty type, since my analysis relies on EHR implementation in hospitals.
+I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
 However, I exclude physicians who list themselves as a specialist in all but one year.
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years and I drop physicians with insufficient claim counts to show indications of behavior.
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years, and I drop physicians with insufficient claim counts to show indications of behavior.
 
@@
 Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen, and overall billing activity.
-I describe these outcome variables and their construction in detail below in Section \ref{sec:outcome}.
+I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.
 
@@
 
-CMS Shared Patient data records annual information on the number of patients who bill two entities under Medicare from 2009 to 2015.
+CMS Shared Patient data records annual information on count of patients and frequency of billing associated with the same patient across providers within a given time period (30, 60, 90, or 180 days).
+These data are available from 2009 to 2015.
 For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common.
 The number of shared patients are collected in 30, 90, or 180 day intervals.
-I focus on the number of shared patient billing two entities in the same day, within a 30 day interval.
-I employ this data to detect physician who work closely with hospitals and which of those hospitals have implemented an EHR, denoting each physician's exposure to EHRs.
-This measure of exposure defines various treatment variables for my analysis.
+I focus on the 30 day interval, and I employ this data to detect physicians who work closely with hospitals.
 I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
@@
 where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$.
-Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future \footnote{Alternatively, I could also define retirement using future number of patients.
+Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future.
+\footnote{Alternatively, I could also define retirement using future number of patients.
 I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \ref{app:years}, I limit the sample years to 2009-2015 to ensure sufficient future year of zero claims.}.
+In Appendix \ref{app:years}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}
 
@@
 While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from EHRs.
-
 Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and total billing activity.
@@
 Only 3\% of physicians (approximately 770) retire over the course of the panel.
-30\% of the physician-year observations have value one for working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting.
+30\% of the physician-year observations indicate working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting.
 By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure.
@@
 In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time.
-In 2009, 32\% of the sample of physicians were already exposed to an EHR.
+In 2009, 28\% of the sample of physicians were already exposed to an EHR.
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
@@
 
-\section{Empirical Strategy}
+\section{Empirical Strategy}\label{sec:empstrat}
 
@@
 Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot.
-When the dis-aggregated estimates are of value, I present those as well.
+In some cases, I present the de-aggregated estimates to show that results are not driven by a particular year of treatment.
 Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each table.
@@
 They may add features or switch vendors, but do not reverse EHR use.
-This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a 2\% random sample of hospitals over time\footnote{Note that out of the 81 hospitals included in the random sample, 4 of them show a reversal of treatment status.
+This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a 2\% random sample of hospitals over time.\footnote{Note that out of the 81 hospitals included in the random sample, 4 of them show a reversal of treatment status.
 Upon further inspection, each of these hospitals show such a reversal due to a discrepancy in the coding of whether the hospital uses an EHR partially or fully.
 Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not.
-Hospitals with this discretion arre coded as implementing an EHR in the first year that they answered "fully" to the survey question.}.
-Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
+Hospitals with this discretion arre coded as implementing an EHR in the first year that they answered "fully" to the survey question.} Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
 Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
@@
 
-\begin{figure}[ht]
+\begin{figure}[ht!]
     \centering
@@
 In this case, the effects seen could be from having the data assistant instead of EHR use.
-Thus, I investigate hteir existence further.
+Thus, I investigate their existence further.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \centering
@@
 In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
-The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after implementation.
-While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .02.
-Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure.
-Considering the magnitude of the decision to retire, it would be alarming to see a numerically large effect on this outcome.
-A 10\% increase in the likelihood of such a decision is a meaningful jump.
-While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that a positive effect on the likelihood of retirement did occur as a result of EHR implementation.
+These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
+The top panel suggests that for any age physician, being exposed to an EHR leads to a .003 ppt increase in the likelihood of retiring in the first year after exposure, and a .0045 ppt increase in the second year after exposure.
+While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .03.
+Relative to this average, the effect represents a 10\% (or 15\% in the second year) increase in the likelihood of retirement due to EHR exposure.
+While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates yields confidence in a positive effect truly occurring.
+Further, even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
+Therefore, in Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
+These results are largely similar to those presented here.
 
@@
     \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.65]{Objects/retire_plot.pdf}
+    \includegraphics[scale=.6]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
 
-Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement and typical retirement age.
-These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above.
-A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
+Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
+These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above in the year after exposure, where younger physicians are driving the result for the second year after exposure.
+A physician who is at least 60 years old is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
 In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure.
-Physicians in the young category have a larger effect on the likelihood of retiring in the second year after exposure relative to the first year.
-While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.
+Physicians in the young category are also 25\% more likely to retire, just in the second year after exposure.
+One reason this could occur is if younger physicians are changing careers due to EHR exposure, where they no longer see patients in a clinical setting, then it may take more time to prepare for such a switch.
+Whereas formal retirement can be decided and carried out more immediately.
+Another possibility is that these physicians are shifting away from Medicare patients.
 
@@
  I consider the decision to switch work settings using multiple outcomes.
-First, I directly investigate the extent to which physicians work in office based setting and whether this changes due to EHR exposure.
-I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has seen any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
-This analysis does not include physicians who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
+First, I investigate the extent to which hospitalists work in office based setting and whether this changes due to EHR exposure.
+I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
+This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
 
 I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
-The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean.
-This is positive, but somewhat inconsequential effect, both economically and compared to the effect of EHRs on retirement discussed above.
+The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .014 percentage points, a 4.6\% increase relative to the mean.
+This result indicates that hospitalists are more likely to start working in an office in the year they are exposed to an EHR, though the effect is not as large as the one found for physicians retiring.
 Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
@@
 This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more stuck to their particular work environment.
+While there is not strong evidence for the existence of a pre-trend, I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}.
+This analysis yields the same result.
 
@@
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.65]{Objects/officeind_plot.pdf}
+    \includegraphics[scale=.6]{Objects/officeind_plot.pdf}
     \label{fig:officefirst}
@@
 
-If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings.
-I show the estimates for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
-This result has varying implications depending on how total number of patients is affected.
+It may be that hospitalists who are already working in office settings shift more of their time or patients towards the office and away from hospitals if EHRs impose a meaningful cost to working in a hospital.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.
+I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
+This result has varying implications depending on whether total number of patients (the denominator of this variable) is affected.
 If total number of patients remains constant, then the absolute number of patients seen in offices is also unchanged.
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Fraction of Total Patients Seen in Office}
-    \includegraphics[scale=.65]{Objects/officefrac_plot.pdf}
+    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
+    \includegraphics[scale=.6]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
 The previous results only capture switching behavior which involves office-based settings.
-A more general way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors.
+Another way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors.
 Again, I exclude physicians who ever retire.
-The outcome variable is equal to 1 if the list of primary zip codes (up to twelve of them) changes.
+The outcome variable is equal to 1 if their list of primary zip codes changes.
 This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes.
@@
 The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}.
-While there is a statistically positive effect in the first 2 years of exposure, there is an upward sloping pre-trend that is probably driving this.
-Thus, I make no claim that EHR exposure is leading physicians to switch zip codes.
+For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is also evidence of a pre-trend that may be driving this.
+To make a more conclusive argument about the true effect under violations of the parallel trends assumption, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}.
+These results reveal that under linear violations of the parallel trends assumption, there is still a positive effect of EHR exposure on changing zip codes for younger physicians and no effect for retirement age physicians.
+A .043 ppt increase in the likelihood of changing zip codes for younger physicians indicates a 36\% increase relative to the mean, a significant increase in probability.
 
@@
     \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.65]{Objects/zip_plot.pdf}
+    \includegraphics[scale=.6]{Objects/zip_plot.pdf}
     \label{fig:zip}
@@
 If this acquisition is correlated with incentives to adopt an EHR, then the effect I find is biased upward.
-Since I do not find strong evidence for switching behavior, I leave this question to future research.
 
@@
 Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
-These physicians do not retire or see patients in another hospital.
-Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the period.
+These hospitalists do not retire or see patients in another hospital.
+Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.
 
 The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
-For physicians of any age, patient count increases by 11.5 patients in the year of EHR exposure, 24 patients in the year after exposure, 15.3 patients in the second year after exposure, and then no effect after the second year.
-These results translate to a .5-1\% increase in the number of patients seen relative to the mean.
+For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then no effect after the second year.
+These results translate to a 5-10\% increase in the number of patients seen relative to the mean.
 As with other outcomes, this result is totally due to changes by the younger group of physicians.
-While there is a statistically larger effect for younger physicians relative to senior physicians (who likely take longer to adapt to technology changes), the observed effect is not largely meaningful.
+Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends}.
 
@@
     \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.65]{Objects/patient_plot.pdf}
+    \includegraphics[scale=.6]{Objects/patient_plot.pdf}
     \label{fig:patient}
@@
 
-Total patient count is directly related to the fraction of patients seen in an office setting.
+This outcome, total patient count, is directly related to the fraction of patients seen in an office setting.
 For younger physicians, since the total number of patients increases in the first couple of years after exposure, the absolute number of patients seen in offices increases as well.
@@
 
+Next, I consider whether EHR exposure affects claim count.
 EHRs may affect physician incentives in such a way that billing activity changes differently from patient count.
-If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the amount of patients has changed or not.
-On average for the entire sample, claim counts are roughly 5.5 times the number of patients seen in a year.
-Shown in Figure \ref{fig:claim}, young physicians increase claim count by 37 in the year of exposure and 96 in the year after exposure, 3-4 times the increase in patient count.
-Thus, billing activity increases proportionally less than average.
-EHRs do not seem to be facilitating more billable activity.
+If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the number of patients has changed or not.
+On average for the entire sample, claim counts are roughly 4 times the number of patients seen in a year.
+Shown in Figure \ref{fig:claim}, there is not strong evidence for an effect of EHR exposure on claim count for physicians of retirement age.
+The point estimates indicate a positive effect in the first two years of exposure for younger physicians, but this is not strong evidence.
+Thus, for hospitalists $>= 60$, it seems EHRs are not making significant changes to practice behavior.
+However, for those $< 60$, the observed claim count increase is two times the observed patient count increase.
+Since the average claim count per patient is 4, these results imply that the number of claims per patient is actually decreasing.
+That is, physicians are billing for less items as they increase the number of patients.
 
@@
     \caption{Effect of EHR Exposure on Claim Count}
-    \includegraphics[scale=.65]{Objects/claim_plot.pdf}
+    \includegraphics[scale=.6]{Objects/claim_plot.pdf}
     \label{fig:claim}
@@
 Using an event study framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR in a hospital, where those of retirement age have an even larger increase relative to younger physicians.
-Further, I find limited evidence that physicians are switching work settings, but that this may be driven by hospital incentives instead of physician incentives.
+Further, I find evidence that younger physicians are switching work settings, but that this may be driven by hospital incentives instead of physician incentives.
 Finally, physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.
@@
 The implications of these results are threefold.
-First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care, especially if these physicians are in rural settings.
+First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care.
 In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
@@
 Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users.
-While the technology may be a burden, physicians are able to see more patients because of their use, and are not more incentivized to over-bill.
+While the technology may be a burden, physicians are able to see more patients because of their use, and do not seem to over-bill because of them.
 
@@
 \section{Sensitivity Analysis}
+
+\subsection{Analyzing Pre-Trends Assumptions}\label{sec:pretrends}
+
+Recent work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference in difference designs.
+In my main specification, I assume that, conditional on covariates, average outcomes for those treated in group g would have followed a parallel trend as those in groups treated in later periods.
+I assess this assumption by testing whether the coefficients showing the effect of EHR exposure on an outcome prior to actual treatment are jointly zero.
+All main specification graphs show p-values from this test.
+In some outcomes, I fail to reject the null that there does not exist a pre-trend.
+Yet, in some cases, I reject this null.
+This section utilizes the tools established by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) to both demonstrate robustness of the results found in the main specification, and to improve our understanding in the cases where a pre-trend exists.
+
+\subsubsection{Retirement}
+
+First, I examine the results with outcome variable retirement.
+While visually there does not appear to be a strong pre-trend, for the full sample of physicians and physicians $< 60$, the p-value indicates a pre-trend may exist.
+Since the positive effect seems to be a possibility in event time 1 and event time 2, I present a plot of robust confidence intervals under various possible assumptions of parallel trends violations for these periods in Figure \ref{fig:pre_retire}.
+In particular, the original confidence interval is shown in yellow, and in gray are confidence intervals robust to varying functions of parallel trend violations.
+When allowing for linear trend violations (M=0), the effect of EHR exposure on retirement is still positive.
+As I allow for more non-linearity (M$>$0) in these variations, this effect becomes zero.
+This is true for both event time 1 and event time 2.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/retire_pretrends_plot.pdf}
+    \label{fig:pre_retire}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
+
+\subsubsection{Work in Office Likelihood}
+
+Next, I examine pre-trends for the outcome variable indicating whether the physician does any work in  an office.
+The p-values do not indicate pre-trends for the full sample or for physicians $< 60$, but I reject no pre-trends for physicians of retirement age.
+As with retirement as the outcome, I show robust confidence intervals in Figure \ref{fig:pre_work}.
+However, the main results show up in event time 0 and 1, so I show confidence intervals for these periods.
+Further, I show these for the full sample of physicians, but also for older physicians separately since the pre-trend seems to differ slightly.
+When allowing for linear pre-trend violations, I find very similar results to the main specification.
+For physicians $>= 60$, there is a higher chance that the effect of EHR exposure on the probability of working in an office is positive.
+This would be theoretically intuitive since older physicians may struggle with the technology more than younger physicians and choose to switch settings.
+However, I do not make strong conclusions given the null results found in most sensitivity analyses.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/work_pretrends_plot.pdf}
+    \label{fig:pre_work}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
+
+\subsubsection{Fraction of Patients in Office}
+
+The next outcome considered is the fraction of patients seen in an office setting.
+All p-values suggest there are no pre-trends.
+I show robust confidence intervals for the full sample of physicians in the first and second year of exposure in Figure \ref{fig:pre_frac}.
+These results hold no matter the assumption on the violation of pre-trends.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/frac_pretrends_plot.pdf}
+    \label{fig:pre_frac}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
+
+\subsubsection{Change Zip Codes Likelihood}
+
+I now consider pre-trends for the effect of EHR exposure on whether the physician changes zip codes.
+In all main specification results, there is indication of a clear pre-trend and thus it is not clear whether there is a positive effect due to EHR exposure or due to this trend occurring prior to exposure.
+Thus, I present confidence intervals that are robust to different specifications of parallel trend violations for each age group at event time 0 and event time 1 in Figure \ref{fig:pre_zip}.
+At event time 0, it seems younger physicians are more likely to change zip codes after EHR exposure under a linear pre-trend violation and slightly non-linear pre-trend violations, although the magnitude of the effect is likely smaller than the main specification shows.
+This is also true for younger physicians in event time 1.
+For physicians of retirement age, it seems the entire effect was driven by a pre-trend and thus there is no effect of EHR exposure on this group's likelihood of changing zip codes.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/zip_pretrends_plot.pdf}
+    \label{fig:pre_zip}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
+
+\subsubsection{Patient Count}
+
+Now I examine the sensitivity of the main specification results where the outcome considered is the number of patients seen.
+None of the p-values indicate a violation of parallel trends, but I present a plot of confidence intervals robust to specified variations in parallel trends as a robustness check in Figure \ref{fig:pre_patient}.
+These plots indicate that, even under nonlinear deviations in parallel trends, there is a positive effect of EHR exposure on patient count.
+Under a linear variation in parallel trends, the effect is positive, yet the magnitude is nearly half of the magnitude found in the main specification.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/patient_pretrends_plot.pdf}
+    \label{fig:pre_patient}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
+
+\subsubsection{Claim Count}
+
+Finally, I assess the results found in the main specification for the outcome variable claim count.
+The main specification indicates a downward trend in the years leading up to exposure and a positive effect in the year of exposure and the year after exposure for younger physicians, and no effect for older physicians.
+In Figure \ref{fig:pre_claim}, I show confidence intervals robust to violations in parallel trends for these two age groups.
+For physicians $>= 60$, this plot shows a null effect just as in the main specification.
+For physicians $< 60$, there still seems to be a positive effect on both event years, although under exactly linear variations in parallel trends, this effect is smaller or zero.
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.5\linewidth}
+    \caption{}
+    \includegraphics[scale=.5]{Objects/claim_pretrends_plot.pdf}
+    \label{fig:pre_claim}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:}}}
+\end{figure}
 

### 2022-08-04 Presentations/YESconference.tex
Updates from Overleaf

--- 2022-01-10 Presentations/ThirdYearPaperpres.tex
@@
 
-\usetheme[progressbar=frametitle]{metropolis}
+\usetheme[progressbar=frametitle, sectionpage=none]{metropolis}
 \usepackage{appendixnumberbeamer}
@@
 \usepackage{booktabs}
-\usepackage{xcolor}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{mathtools}
@@
 
-\subtitle{Hanna Glenn\\ \small Advisor: Dr. Ian McCarthy}
+\subtitle{Hanna Glenn\\ \small Emory University}
  \date{\today}
@@
 
-\begin{frame}{The Broad Question}
+\begin{frame}{Big Picture}
 \begin{itemize}
@@
     \item Investigated in many settings
+    \vspace{3mm}
+    \item The answer depends heavily on the setting
     \vspace{3mm}
@@
             \tiny (Willis and Hudomiet 2021, Labour)
+
+            \footnotesize
+
+            \item Computerization in white collar jobs leads to decrease in employment for the same jobs that implemented the technology \\ \vspace{1mm}
+            \tiny (Dillender & Forsythe 2022, NBER)
         \end{itemize}
@@
             \color{blue}
-                \item Healthcare spending is extremely high
+                \item While doctors are white collar, their labor market incentives differ from many other jobs
                 \item Population of physicians - possible shortages / access to care issues
@@
 \begin{frame}{Data}
-    \underline{Main: CMS Shared Patient Data (2009-2015)}
+    \underline{CMS Shared Patient Data (2009-2015)}
     \begin{itemize}
@@
 
-    \underline{EHR Use:}
+    \underline{Medicare Data on Provider Practice and Specialty (2009-2017)}
     \begin{itemize}
-        \item AHA Survey
+        \item Physician-level number of unique patients, primary locations of practice, and total claims 
     \end{itemize}
@@
 
-    \underline{Other:}
+    \underline{EHR Use:}
     \begin{itemize}
-        \item Physician Compare
+        \item AHA Survey Question
     \end{itemize}
@@
 
-    Aggregated to physician level
+    \underline{Physician Compare}: Age of physician
+
+    \vspace{3mm}
+
+    $\rightarrow$ Final data is at the physician level from 2009-2017, where only physicians who are exposed by 2015 are included 
 \end{frame}
@@
 \begin{frame}{Variables}
+
 Labor Market Outcomes:
 \begin{itemize}
+    \item Retirement
+    \begin{itemize}
+        \item Physicians stop seeing patients $\rightarrow$ drop out of data
+    \end{itemize}
+
+    \item Work Setting:
+    \begin{itemize}
+        \item Percentage of patients seen in an office setting
+        \item Indicator for seeing any patients in an office setting
+        \item Indicator for changing zip codes
+    \end{itemize}
+
     \item Productivity:
     \begin{itemize}
-    \item Patients visited in hospitals
-\end{itemize}
-\end{itemize}
-\begin{itemize}
-\item \color{gray} Retirement
-\end{itemize}
-\begin{itemize}
-\item \color{gray} Work Setting:
-\begin{itemize}
-    \item \color{gray} Maintaining billable activity only outside of hospitals
-    \item \color{gray} Level of billable activity outside of hospitals
+        \item Number of patients
+        \item Ratio of claims per patient
     \end{itemize}
@@
 \begin{itemize}
-    \item Indicator for any exposure to EHR
-    \item \color{gray} Fraction of hospitals exposed to EHR
-\end{itemize}
-
-Other:
-\vspace{-2mm}
-\begin{itemize}
-    \item Years of experience
-    \item Hospital characteristics
+    \item Indicator for any exposure to EHR at related hospitals
 \end{itemize}
@@
     \centering
-    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
-\end{frame}
-
-\begin{frame}{Data}
-\centering
-    \includegraphics[scale=.5]{Objects/cont_treatment_graph.pdf}
-
+    \includegraphics[scale=.4]{Objects/sum_stats_year.pdf}
 \end{frame}
@@
 \begin{frame}{Event Study}
-\begin{equation*}
-    y_{it}=\alpha_i+\delta_t+q_{it}'\lambda+\sum_{k=-4}^4 \beta_kz_{i,t-k} + \varepsilon_{it}
-\end{equation*}
 
-\vspace{4mm}
+\begin{itemize}
+    \item Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects
+    \vspace{3mm}
+    $$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
+    \vspace{3mm}
+    \item Aggregate these over groups to get a familiar event study plot
+\end{itemize}
 
-\begin{align*}
-    y_{it} &: \text{Outcomes} \hspace{30mm}\\
-    \alpha_i, \delta_t &: \text{Fixed Effects}\hspace{30mm}\\
-    q_{it} &: \text{Physician/hospital characteristics}\hspace{30mm}\\
-    z_{i,t-k} &: \text{Exposure to EHR} \hspace{30mm}
-\end{align*}
+\end{frame}
+
+\begin{frame}{Assumptions}
+\begin{itemize}
+    \item No reversal of treatment
+    \item No anticipation of treatment (can be relaxed)
+    \item Parallel trends
+    \item Institutional assumption: physicians are constrained to use to technology implemented in their hospital
+\end{itemize}
 
@@
 
-\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
+\begin{frame}{Retirement}
 \begin{figure}[ht]
-        \begin{minipage}[b]{0.47\linewidth}
     \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample.pdf}
-        \end{minipage}
-        \hspace{0.2cm}
-        \begin{minipage}[b]{0.47\linewidth}
+    \includegraphics[scale=.4]{}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Number of Patients Seen in Office}
+\begin{figure}[ht]
     \centering
-            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample.pdf}
-        \end{minipage}
+    \includegraphics[scale=.4]{}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Indicator for working in office}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.4]{}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Change Zip}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.4]{}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Patient Count}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.4]{}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Claims per Patient}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.4]{}
 \end{figure}
@@
 \begin{itemize}
-    \item Despite physician response to EHRs, those who stick around do become more productive
-    \vspace{3mm}
-
-    \item Complementary relationship between physicians and EHRs
-    \vspace{3mm}
-
-    \item Older physicians see (slightly) greater productivity increases 
-    \vspace{3mm}
+    \item Evidence that physicians are changing their behavior because of this new technology, specifically through retirement or job switching
+    \item Little evidence of work setting changes
+    \item For physicians who endure through the implementation of the technology, claims per patient goes down.
 

### 2022-08-04 Table Code/overall stats.tex

@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 179883 & 0.03 & 0.18 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 179883 & 0.10 & 0.24 & 0 & 1\\
-\hspace{1em}Work in an Office & 179883 & 0.30 & 0.46 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 161558 & 0.12 & 0.33 & 0 & 1\\
-\hspace{1em}Number of Patients & 179883 & 329.71 & 274.35 & 0 & 7613\\
-\hspace{1em}Claim Count & 179883 & 1382.14 & 1789.34 & 0 & 64508\\
+\hspace{1em}Ever Retire & 179,883 & 0.03 & 0.18 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 179,883 & 0.10 & 0.24 & 0 & 1\\
+\hspace{1em}Work in an Office & 179,883 & 0.30 & 0.46 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 161,558 & 0.12 & 0.33 & 0 & 1\\
+\hspace{1em}Number of Patients & 179,883 & 330 & 274 & 0 & 7,613\\
+\hspace{1em}Claim Count & 179,883 & 1,382 & 1,789 & 0 & 64,508\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 179883 & 0.75 & 0.43 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 179883 & 0.63 & 0.44 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 179,883 & 0.75 & 0.43 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 179,883 & 0.63 & 0.44 & 0 & 1\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 179883 & 44.87 & 8.43 & 25 & 86\\
-\hspace{1em}Female & 179883 & 0.34 & 0.47 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 179883 & 1.49 & 0.82 & 1 & 12\\
-\hspace{1em}Number of Systems Worked With & 179883 & 1.31 & 0.60 & 1 & 8\\
+\hspace{1em}Age & 179,883 & 44.9 & 8.4 & 25 & 86\\
+\hspace{1em}Female & 179,883 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 179,883 & 1.49 & 0.82 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 179,883 & 1.31 & 0.60 & 1 & 8\\
 \bottomrule

### 2022-08-16 Paper/Paper.tex

@@
 \begin{abstract}
-    I investigate the relationship between technological innovation and labor market decisions by examining changes in physician behavior as a result of major technology shift in U.S.
-health care, electronic health records.
-I treat EHR implementation in hospitals as exogenous to physicians employed by the hospital and estimate average group time treatment effects on various labor market outcomes.
-I find that physicians of retirement age are more likely to retire, younger physicians are more likely to switch to office settings, and physicians who do not switch work settings see more patients as a results of individual EHR exposure.
+    Technological innovation has affected workers and labor markets for decades, yet the effects of modern technologies in some settings are still not understood.
+I add to the understanding of technology implementation's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
+health care, electronic health records (EHRs).
+I treat EHR implementation in hospitals as an exogenous treatment to hospitalists and estimate average group time treatment effects on various labor market outcomes.
+I find that physicians are 10\% more likely to retire, and 4\% more likely to work in an office if they choose not to retire.
+Finally, for hospitalists who remain exposed to the technology, number of patients seen increases.
+These results point to adverse effects that can occur when technology is poorly implemented in technical occupations.
 \end{abstract}
@@
 
-How technological innovation affects labor markets is often debated, in part because of the vast differences in technology and tasks present even within industries.
-The most widespread and health care-altering technological innovation in recent years took place with the implementation of electronic health records.
-Because the technology was incentivized by policy, it was implemented rapidly and on a large scale, making it an opportune setting to examine the relationship between labor markets and technology.
-In this paper, I investigate physician labor market changes as a result of the implementation of electronic health records.
-
-Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggestions for care.
+How technological innovation affects labor markets is heavily studied, in part because of the vast differences in technology and tasks present even within industries.
+Yet, there is still much to be learned about modern technology and its interaction with workers.
+A recent technology roll-out that altered the U.S.
+health care system dramatically was the electronic health record (EHR).
+The implementation and use of EHRs were incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
+Since many physicians working in hospitals were exposed to this technology beyond their control, this is an opportune setting to investigate the relationship between technology implementation and labor market decisions for highly skilled workers.
+
+Specifically, EHRs are computerized medical records stored in advanced systems that may include advanced capabilities such as decision making assistance.
 EHRs have become increasingly relevant in the U.S.
-since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”.
+According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
-Researchers and government officials expected the movement towards digitization to affect efficiency in health care substantially.
-A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), revealing the nationwide movement towards digitization.
-
-Physicians, as primary users of this technology, play an important role in whether potential benefits of EHRs are realized.
-With the rapid implementation of a complex technology, the practical tasks that take place in a physician's daily life changed just as rapidly.
-In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
+The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy.
+For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
+The desired movement towards digitization was realized; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
+
+Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition.
+The daily tasks a physician faces change drastically when a new system is put in place.
+Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
 Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in certain settings, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor market altogether or shifting towards lower cost settings.
+The frustration of using a new technology raises the cost of working in settings that it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market outcomes.
-
-My primary datasets are the CMS Shared Patient data and the Medicare Data on Provider Practice and Specialty (MD-PPAS), from which I construct an individual-level panel from 2009-2017 consisting of physicians who work primarily in a hospital, known as hospitalists.
-The data captures whether such physicians are exposed to EHRs in their associated hospital(s), several labor market outcomes, and other demographic information.
-For all outcomes, the main independent variable of interest is a binary treatment variable capturing exposure to an EHR.
-In the main analysis, a physician is considered to be exposed to an EHR if they are working in any hospital that implements an EHR.
+Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
+
+I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists, which spans from 2009-2017.
+Using CMS Shared Patient Data, I determine which hospitals each hospitalist is connected to.
+Then, I use the AHA Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
+Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist.
+For all outcomes, the independent variable of interest is a binary treatment variable capturing exposure to an EHR.
 I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and location of the physician office, (3) number of patients seen, and (4) billing activity.
@@
 More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
-One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EMRs being adopted in primary-care sites (\cite{Meyerhoefer}).
-
-I expand this literature, first by considering a different mechanism by which EHRs could affect physician labor markets (and, indirectly, patients), which, to my knowledge, has not yet been examined.
+One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).
+
+I expand this literature, first by considering a different mechanism by which EHRs could affect health care through physicians, which, to my knowledge, has not yet been examined.
 This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups.
@@
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this work, I find that EHR implementation led to a 0.006 ppt (s.e.
-0.002) increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
-Physicians less than 60 were 0.002 ppts (s.e.
-0.0009) more likely to retire after exposure, possibly pointing to career switching since retirement is measured as a lack of Medicare patients seen.
+Contrary to this work, I find that EHR implementation led to a 0.006 ppt increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
+Physicians less than 60 were 0.002 ppts more likely to retire after exposure, possibly pointing to career switching since retirement is measured as a lack of Medicare patients seen.
 These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
-For those who did not retire, I find no evidence of switching towards office settings in either age group.
-However, both age groups are more likely to change zip codes (by 0.045 ppts), suggesting across-hospital switching.
-Finally, I find a modest, positive effect on patient count (an increase of 14 patients relative to a mean of 330) completely driven by younger physicians, and no effect on billing activity for either age group.
-Thus, for younger physicians, EHRs are leading to a decrease in billing activity per patient driven by an increase in patients.
-
-A key assumption that underlies the main analysis is that physicians are constrained to using the technology in a hospital who chooses to adopt.
-I consider two scenarios where this assumption would be violated.
-First, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
+For those who did not retire, I find that young physicians were .014 ppts (4.6\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure.
+Finally, I find a modest, positive effect on patient count completely driven by younger physicians, and no effect on billing activity for either age group.
+
+Two key assumptions underlie the main analysis.
+First, I assume that physicians are constrained to use the technology implemented in their work setting.
+To test this assumption, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
 If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence.
@@
 While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making.
-For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a lower probability that EHR implementation is endogenous.
+For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a low likelihood that EHR implementation is endogenous.
 The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.
 
-\section{Background on Electronic Health Records}
+\section{Institutional Details}
+
+\subsection{Background on EHRs}
 
@@
 
-\section{EHR and Physician Labor Markets}
+\subsection{EHR and Physician Labor Markets}
 
@@
 
-\subsection{Retirement}
+\subsubsection{Retirement}
 
@@
 
-\subsection{Work Setting}
+\subsubsection{Work Setting}
 
@@
 
-\subsection{Productivity}
+\subsubsection{Productivity}
 
@@
 A group indicates all physicians treated, or first exposed to an EHR, in the same year.
-To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{sant2020doubly} (\citeyear{sant2020doubly}).
 Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}.
@@
 However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
-I investigate this possibility and limit the data to address it in Section \ref{sec:patientcount}.
+This is why I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
 
@@
 As is usual in a difference-in-differences framework, I assume a version of parallel trends based on not-yet-treated units.
-I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
+I assume that, conditional on years of experience, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
 This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession.
@@
 
-An institutional assumption that I make is that physicians working in EHR-adopted hospitals are constrained to use the technology.
-My outcomes have two key determinants: the choice of physicians to learn and utilize the technology, and the realized effect of electronic health records themselves.
-The point of this paper is to identify the second determinant, and thus I assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
-That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital.
-It is reasonable to believe that physicians are fully utilizing the EHRs in the hospital they work in, since by 2015 electronic health records were widely implemented and considered unavoidable.
-An objection to this assumption may be that physicians stay in a hospital, but the hospital hires an employee to do technology work for the physician.
-In this case, the effects seen could be from having the data assistant instead of EHR use.
-Thus, I investigate their existence further.
-
-\subsection{Hiring Data Assistants}\label{sec:dataass}
-
-One concern with the presented analysis is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals, such as the existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf, which is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
-Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
-I will refer to any employee in these categories as a data assistant.
-In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
-
-First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
-Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
-The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
-From 2005 to 2013, 15-60 more employees were registered each year.
-The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
-If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
-The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
-
-\begin{figure}[t]
-\centering
-\captionsetup{width=.5\linewidth}
-\caption{Frequency of Data Assistant Enumeration by Year}
-\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
-\label{fig:dataassistant_histogram}
-\vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Histtogram uses NPPES data on the enumeration of certain types of NPIs to estimate the number of new data assistants being hired in each year of my sample.}}
-\end{figure}
-
-Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
-I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
-I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
-I find that .1\% of physician-year observations are associated with data assistants.
-This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
-However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
-Details of this analysis are given in Appendix \ref{app:DA}.
+In order for this to be a causal estimate of EHR exposure on various labor market outcomes, there are additional institutional assumptions necessary.
+First, I assume that when a physician works in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
+That is, there are no ways for the physician to continue practicing without using the EHR.
+I investigate this assumption in Section \ref{sec:dataass}.
+Second, I assume that a physician's exposure to EHRs is exogenous.
+That is, the physician does not influence the hospital's decision to implement an EHR.
+I investigate this assumption in Section \ref{sec:endogeneity}.
 
@@
 \section{Physician Influence of EHR Decision in Hospitals}
+
+\subsection{Hiring Data Assistants}\label{sec:dataass}
+
+One concern with the presented analysis, particularly for the productivity outcomes, is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals.
+The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
+Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
+they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+I will refer to any employee in these categories as a data assistant.
+In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
+
+First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
+Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
+The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
+From 2005 to 2013, 15-60 more employees were registered each year.
+The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
+If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
+The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
+
+\begin{figure}[t]
+\centering
+\captionsetup{width=.5\linewidth}
+\caption{Frequency of Data Assistant Enumeration by Year}
+\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+\label{fig:dataassistant_histogram}
+\vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Histtogram uses NPPES data on the enumeration of certain types of NPIs to estimate the number of new data assistants being hired in each year of my sample.}}
+\end{figure}
+
+Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
+I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
+I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
+I find that .1\% of physician-year observations are associated with data assistants.
+This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
+However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
+Details of this analysis are given in Appendix \ref{app:DA}.
+
+\subsection{Endogeneity of EHR Implementation}\label{sec:endogeneity}
 

### 2022-08-16 Presentations/YESconference.tex

@@
 \begin{frame}{Big Picture}
-\begin{itemize}
-    \item Longstanding topic in economics: relationship between technology and employment
-    \vspace{2mm}
+How does technology impact employment?
+
     \begin{itemize}
         \item Displacement?
+        \vspace{2mm}
         \item Productivity?
@@
     \vspace{3mm}
-    \item Investigated in many settings
-    \vspace{3mm}
-    \item The answer depends heavily on the setting
-    \vspace{3mm}
-    \item This paper: specific healthcare setting
+
+Investigated in different ways
+
+    \begin{itemize}
+        \item Computerization on white collar jobs
+        \vspace{2mm}
+        \item Machines on factory workers
     \end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]{What is an Electronic Health Record?}
-``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"
 
@@
 
-What it can do:
-\begin{itemize}
-    \item Paper $\rightarrow$ computerization
-    \item Communication
-    \item Decision making assistance
-\end{itemize}
+    \pause
 
+    \begin{alertblock}{$\rightarrow$ This paper considers this question in the context of healthcare and physician labor markets}\end{alertblock}
+\end{frame}
+
+\begin{frame}[fragile]{What is an Electronic Health Record (EHR)?}
+``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"
+
+\vspace{5mm}
+\centering
+\includegraphics[scale=.35]{Objects/EHR_picture.PNG}
 \end{frame}
@@
 \end{center}
-
 \end{frame}
@@
 \begin{frame}{This Paper}
+$\rightarrow$ Did EHR implementation in hospitals affect physician labor market decisions?
 
-Did EHR implementation in hospitals affect physician labor market decisions?
-\begin{itemize}
-    \item Intensive Margin / Productivity
-    \item Extensive Margin / Retirement
-    \item Hospital $\rightarrow$ Office
-\end{itemize}
-
+\vspace{5mm}
+\centering
+\includegraphics[scale=.45]{Objects/EHR_FlowChart_General.PNG}
 \end{frame}
 
-\section{Contribution}
+\section{Analysis and Data}
 
-\begin{frame}{Literature}
-\small
-\begin{columns}
-\setlength{\tabcolsep}{-5pt}
-    \column{0.4\textwidth}
-        \centering
-        \underline{ Technology $\rightarrow$ Labor }
-        \vspace{-1mm}
-        \begin{itemize}
-            \item Erosion effect vs. wage effect \\ \vspace{1mm}
-            \tiny (Zeira $\&$ Joseph 2011, EJ)
-
-            \footnotesize
-
-            \item Cause retirement? \\ \vspace{1mm}
-            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
-
-            \footnotesize
-
-            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
-            \tiny (Willis and Hudomiet 2021, Labour)
-
-            \footnotesize
-
-            \item Computerization in white collar jobs leads to decrease in employment for the same jobs that implemented the technology \\ \vspace{1mm}
-            \tiny (Dillender & Forsythe 2022, NBER)
-        \end{itemize}
-
-        \column{0.4\textwidth}
-        \centering
-
-        \pause
-
-        \color{blue}
-        \underline{ Contribution }
-
-        \vspace{1mm}
-
-            \begin{itemize}
-            \color{blue}
-                \item While doctors are white collar, their labor market incentives differ from many other jobs
-                \item Population of physicians - possible shortages / access to care issues
-            \end{itemize}
-
-\end{columns}
-\end{frame}
-
-\begin{frame}{Literature}
-\small
-\begin{columns}
-\setlength{\tabcolsep}{-5pt}
-        \column{0.4\textwidth}
-        \centering
-        \underline{ EHR $\rightarrow$ Healthcare }
-        \vspace{-1mm}
-        \begin{itemize}
-            \item Small improvements for severe health conditions \\ \vspace{1mm}
-            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)
-
-            \footnotesize
-
-            \item No decrease in cost \\ \vspace{1mm}
-            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)
-
-            \footnotesize
-
-            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
-            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
-        \end{itemize}
-
-        \pause
-
-        \column{0.4\textwidth}
-        \centering
-        \color{blue}
-        \underline{ Contribution }
-        \vspace{1mm}
-        \begin{itemize}
-        \color{blue}
-            \item No empirical work on EHR/physician effects
-            \item Speaks to the puzzling results
-            \item Time period 
-        \end{itemize}
-
-\end{columns}
-
-\end{frame}
-
-\section{Data}
-
-\begin{frame}{Data}
-    \underline{CMS Shared Patient Data (2009-2015)}
-    \begin{itemize}
-        \item Hospital-physician pairs: bill for same patient under Medicare
-        \item Physician (PCP, hospitalist, internist) works in the hospital
-    \end{itemize}
+\begin{frame}{Difference in Difference Analysis}
+Treatment: Exposure to EHR
 
@@
 
-    \underline{Medicare Data on Provider Practice and Specialty (2009-2017)}
-    \begin{itemize}
-        \item Physician-level number of unique patients, primary locations of practice, and total claims 
-    \end{itemize}
+Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects
+
+$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
 
@@
 
-    \underline{EHR Use:}
 \begin{itemize}
-        \item AHA Survey Question
+\item Not-yet-treated units as control group
+    \item Aggregate these over groups to get a familiar event study plot
 \end{itemize}
-
-    \vspace{3mm}
-
-    \underline{Physician Compare}: Age of physician
-
-    \vspace{3mm}
-
-    $\rightarrow$ Final data is at the physician level from 2009-2017, where only physicians who are exposed by 2015 are included 
 \end{frame}
 
-\begin{frame}{Variables}
+\begin{frame}{Data}
+Panel of hospitalists from 2009-2017 with information on whether the hospital uses an EHR and various labor market outcomes
 
-Labor Market Outcomes:
+\vspace{4mm}
+\pause
 \begin{itemize}
-    \item Retirement
-    \begin{itemize}
-        \item Physicians stop seeing patients $\rightarrow$ drop out of data
+    \item EHR information ends in 2015, so I limit to only those who were exposed by 2015
 \end{itemize}
-
-    \item Work Setting:
-    \begin{itemize}
-        \item Percentage of patients seen in an office setting
-        \item Indicator for seeing any patients in an office setting
-        \item Indicator for changing zip codes
-    \end{itemize}
-
-    \item Productivity:
-    \begin{itemize}
-        \item Number of patients
-        \item Ratio of claims per patient
-    \end{itemize}
-\end{itemize}
-
-EHR Use:
-\vspace{-2mm}
-\begin{itemize}
-    \item Indicator for any exposure to EHR at related hospitals
-\end{itemize}
-
 \end{frame}
@@
 
-\section{Analysis}
+\begin{frame}{Assumptions}
+\begin{enumerate}
+    \item No reversal of treatment
+    \vspace{3mm}
+    \item No anticipation of treatment (can be relaxed)
+    \vspace{3mm}
+    \item Parallel trends
+\end{enumerate}
+\end{frame}
 
-\begin{frame}{Event Study}
+\section{Retirement}
 
+\begin{frame}{Retirement}
+    \centering
+    \includegraphics[scale=.5]{Objects/EHR_FlowChart_Retire.PNG}
+\end{frame}
+
+\begin{frame}{Retirement Variable}
+Indicator variable equal to 1 if the physician drops out of the data 
+\vspace{3mm}
 \begin{itemize}
-    \item Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects
+    \item No longer seeing patients
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Physician Incentives}
+Decision to retire is different than we usually think of
 \vspace{3mm}
-    $$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
+    \begin{itemize}
+        \item Retirement is thought of as a function of pensions, and sometimes depends on shocks such as health or job loss
+        \tiny (Hall and Johnson 1980, ILRR), (Gordon and Blinder 1980, J Public Econ), (Hamermesh 1985, JOLE), $\cdot$
+
         \vspace{3mm}
-    \item Aggregate these over groups to get a familiar event study plot
+
+        \normalsize \item On average, physicians plan to retire at age 60 but don't actually retire until age 69 \tiny(Collier 2017)
     \end{itemize}
 
+\vspace{3mm}
+\pause
+
+\normalsize Unique set of incentives for which a shock to the work setting \textit{may} lead to retirement more-so than other occupations, for which studies yield mixed results  \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
+
+\vspace{3mm}
+\pause
+
+\normalsize Related studies find that well educated workers exited at higher rates due to computerization \tiny (Willis and Hudomiet 2021, Labour), (Dillender & Forsythe 2022, NBER)
 \end{frame}
 
-\begin{frame}{Assumptions}
+\begin{frame}{Leaving Clinical Setting}
     \begin{itemize}
-    \item No reversal of treatment
-    \item No anticipation of treatment (can be relaxed)
-    \item Parallel trends
-    \item Institutional assumption: physicians are constrained to use to technology implemented in their hospital
+        \item Retirement is loosely defined
+        \vspace{5mm}
+        \item Any age physician may decide to leave a clinical setting
     \end{itemize}
-
 \end{frame}
-
-\section{Results}
 
@@
     \centering
-    \includegraphics[scale=.4]{}
+    \includegraphics[scale=.35]{Objects/Presentation_retire_all.pdf}
 \end{figure}
+$\rightarrow$ EHR exposure led to a .003 (.0045) percentage point increase in the likelihood of retirement in the first (second) year after exposure (10\% and 15\% increase relative to the mean)
 \end{frame}
 
-\begin{frame}{Number of Patients Seen in Office}
+\begin{frame}{Retirement: Senior Physicians}
 \begin{figure}[ht]
     \centering
-    \includegraphics[scale=.4]{}
+    \includegraphics[scale=.35]{Objects/Presentation_retire_old.pdf}
 \end{figure}
+$\rightarrow$ 25\% increase relative to the mean in the first year, but 0 in the second\\
+$\rightarrow$ What does this mean for younger physicians?
+\end{frame}
+
+\section{Job Setting}
+
+\begin{frame}{Changing Job Setting}
+    \centering
+    \includegraphics[scale=.5]{Objects/EHR_FlowChart_JobSwitch.PNG}
+\end{frame}
+
+\begin{frame}{Job Switching Variables}
+    \begin{enumerate}
+        \item Fraction of Patients Seen in Office 
+        \vspace{3mm}
+        \item Indicator for working in office or not
+    \end{enumerate}
+\end{frame}
+
+\begin{frame}{Physician Incentives}
+Most physicians will not be induced to retire because of EHRs
+    \begin{itemize}
+        \item Another way to avoid (or be exposed to) new technology is to switch location of practice
+        \begin{itemize}
+            \vspace{3mm}
+            \item Literature shows job switching is closely tied to job satisfaction \tiny (Aklerlof, Rose and Yellen 1988, Brookings), (Chadi and Hetschko 2017, JEMS)
+        \end{itemize}
+        \vspace{6mm}
+        \pause
+        \item The cost associated with switching depends on the strength of the tie with the hospital
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Fraction of Patients Seen in Office}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.35]{Objects/Presentation_fracoffice_all.pdf}
+\end{figure}
+$\rightarrow$ No effect!
 \end{frame}
@@
     \centering
-    \includegraphics[scale=.4]{}
+    \includegraphics[scale=.35]{Objects/Presentation_office_all.pdf}
 \end{figure}
+$\rightarrow$ 4.6\% increase relative to the mean\\
+$\rightarrow$ Driven by young physicians
 \end{frame}
 
-\begin{frame}{Change Zip}
-\begin{figure}[ht]
+\section{Productivity}
+
+\begin{frame}{Productivity}
     \centering
-    \includegraphics[scale=.4]{}
-\end{figure}
+    \includegraphics[scale=.45]{Objects/EHR_FlowChart_Productivity.PNG}
+\end{frame}
+
+\begin{frame}{Productivity Variables}
+    \begin{enumerate}
+        \item Patient Count
+        \vspace{6mm}
+        \item Claims per patient
+    \end{enumerate}
+\end{frame}
+
+\begin{frame}{Physician Incentives}
+Past research on information technology $\rightarrow$ productivity typically indicates increase \tiny (Bartel, Ichniowski and Shaw 2007, QJE)
+\vspace{3mm}
+\begin{itemize}
+    \normalsize \item Sometimes depends on managerial style \tiny(Garicano and Heaton 2010, JLE)
+\end{itemize}
+
+\vspace{6mm}
+
+\normalsize We can see how productivity is affected for physicians who don't change behavior as a result of EHRs
+\vspace{3mm}
+    \begin{itemize}
+        \item Expected effect is ambiguous
+        \vspace{3mm}
+        \item Claim Count per patient is a matter of physician incentives
+    \end{itemize}
 \end{frame}
@@
     \centering
-    \includegraphics[scale=.4]{}
+    \includegraphics[scale=.35]{Objects/Presentation_patients_all.pdf}
+\end{figure}
+$\rightarrow$ 5-10\% increase relative to the mean\\
+$\rightarrow$ Driven by demand? 
+\end{frame}
+
+\begin{frame}{Patient Count Broken Down by Group}
+\centering
+    \includegraphics[scale=.25]{Objects/patient_group.pdf}
+\end{frame}
+
+\begin{frame}{Patient Count: Senior Physicians}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.4]{Objects/Presentation_patients_old.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics[scale=.4]{}
+    \includegraphics[scale=.35]{Objects/Presentation_claimperpatient_all.pdf}
 \end{figure}
+$\rightarrow$ No significant change, but a downward trend is possible
 \end{frame}
+
+\section{Conclusion}
 
@@
     \item Evidence that physicians are changing their behavior because of this new technology, specifically through retirement or job switching
+    \vspace{3mm}
     \item Little evidence of work setting changes
-    \item For physicians who endure through the implementation of the technology, claims per patient goes down.
+    \vspace{3mm}
+    \item For physicians who endure through the implementation of the technology, number of patients goes up but not conclusive evidence that claims per patient goes down
 
@@
 
-\section{Continuing Work}
-
-\begin{frame}{Plans for this paper}
-    \begin{itemize}
-        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
-        \item See if there are any specific years driving these results
-        \item Endogenous treatment? (Bandwidth, state privacy laws as IV)
-    \end{itemize}
+\begin{frame}{Robustness Checks}
+    \begin{enumerate}
+        \item Endogeneity of treatment
+        \item Data assistants
+        \item Limiting Years and Using Never-treated as control
+        \item Anticipation
+    \end{enumerate}
 \end{frame}
@@
 \centering
-    Thank you!
+    Thank you! \\
+    \vspace{5mm}
+    Email: hkagele@emory.edu
 \end{frame}

### 2022-08-16 Presentations/YESconference.tex

@@
 
-\normalsize Related studies find that well educated workers exited at higher rates due to computerization \tiny (Willis and Hudomiet 2021, Labour), (Dillender & Forsythe 2022, NBER)
+\normalsize Related studies find that well educated workers exited at higher rates due to computerization \tiny (Willis and Hudomiet 2021, Labour), (Dillender \& Forsythe 2022, NBER)
 \end{frame}
@@
     \centering
-    \includegraphics[scale=.35]{Objects/Presentation_retire_old.pdf}
+    \includegraphics[scale=.35]{Objects/Presentation_retire_ages.pdf}
 \end{figure}

### 2022-08-17 Scripts/analysis.R
adjust font sizes of graph labels

@@ -27,7 +27,7 @@ Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 ## ANALYSIS ---------------------------------------------- #########
 ## Write results for each variable in a loop and create graphs ----- ##
 
-varlist <- list("retire", "pos_office_prior", "pos_office_noprior", "work_in_office", "change_zip", "npi_unq_benes", "claim_count_total", "claim_per_patient")
+varlist <- list("retire", "pos_office_prior", "pos_office_noprior", "work_in_office", "change_zip", "npi_unq_benes", "claim_count_total", "claim_per_patient", "majority_in_office", "total_office_prior", "total_office_noprior")
 
 
 
@@ -202,18 +202,24 @@ ggsave(file="Objects/claim_plot.pdf",plot=plots[[7]], width=10, height=7, units=
 ggsave(file="Objects/claim_per_patient_plot.pdf",plot=plots[[8]], width=10, height=7, units="in")
 
 # Save the different plots separately for presentations
-ggsave(file="Objects/Presentation_retire_all.pdf",plot=plots[[1]]$all, width=10, height=7, units="in")
-ggsave(file="Objects/Presentation_retire_ages.pdf",plot=plots[[1]]$ages, width=10, height=7, units="in")
+ggsave(file="Objects/Presentation_retire_all.pdf",plot=graphs[[1]]$all, width=10, height=7, units="in")
+ggsave(file="Objects/Presentation_retire_ages.pdf",plot=graphs[[1]]$ages, width=10, height=7, units="in")
 
-ggsave(file="Objects/Presentation_frac_prior_all.pdf",plot=plots[[2]]$all, width=10, height=7, units="in")
+ggsave(file="Objects/Presentation_frac_prior_all.pdf",plot=graphs[[2]]$all, width=10, height=7, units="in")
 
-ggsave(file="Objects/Presentation_frac_noprior_all.pdf",plot=plots[[3]]$all, width=10, height=7, units="in")
-ggsave(file="Objects/Presentation_office_all.pdf",plot=plots[[4]]$all, width=10, height=7, units="in")
+ggsave(file="Objects/Presentation_frac_noprior_all.pdf",plot=graphs[[3]]$all, width=10, height=7, units="in")
 
-ggsave(file="Objects/Presentation_patients_all.pdf",plot=plots[[6]]$all, width=10, height=7, units="in")
[…]

### 2022-08-17 Scripts/data5_MDPPAS.R

@@ -138,6 +138,7 @@ observe <- Physician_Data %>%
   dplyr::filter(sum>0) %>%
   dplyr::select(DocNPI,year,hosp_patient_count,claim_count_total)
 
+
 # I'm thinking to replace all NAs with 0.
 Physician_Data <- Physician_Data %>%
   dplyr::mutate(claim_count_total=ifelse(is.na(claim_count_total),0,claim_count_total)) %>%
@@ -172,12 +173,23 @@ Physician_Data <- Physician_Data %>%
 
 
 observe <- Physician_Data %>%
-  filter(is.na(office_yearprior)) %>%
-  select(DocNPI, year, minyr_EHR, work_in_office, office_yearprior)
+  filter(office_yearprior==1) %>%
+  select(DocNPI, year, minyr_EHR, pos_office, pos_inpat, work_in_office, office_yearprior, npi_unq_benes, total_office)
   
   # The NAs that are left in this variable are fine because they only occur when minyr_EHR is either 0
   # or 2009, and these observations get dropped in the analysis anyway. 
 
+# Create variable for total number of patients in office
+Physician_Data <- Physician_Data %>%
+  mutate(total_office=pos_office*npi_unq_benes)
+
+Physician_Data <- Physician_Data %>%
+  mutate(total_office_prior=ifelse(office_yearprior==1,total_office,NA),
+         total_office_noprior=ifelse(office_yearprior==0,total_office,NA))
+
+# Create variable for majority patients in office
+Physician_Data <- Physician_Data %>%
+  mutate(majority_in_office=ifelse(pos_office>pos_inpat,1,0))
 
 
 # ZIP CODES ####

### 2022-08-17 Scripts/sumstats.R

@@ -154,13 +154,13 @@ sum_stats_year <- melt(sum_stats_year, id.vars = "year", measure.vars = c("Hospi
 sum_stats_year <- sum_stats_year %>%
   dplyr::rename("Variable"="variable")
 
-ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable)) + 
+ggplot(sum_stats_year,aes(x=year,y=value,shape=Variable,color=Variable, linetype=Variable)) + 
   geom_line() + geom_point() + labs(x="\nYear", y=" ") + 
   scale_colour_manual(values=cbbPalette) + ylim(0,1)  + xlim(2009,2015) +
   theme(legend.key.size=unit(.3,'cm'),
         legend.key.height = unit(.4, 'cm'),
         legend.key.width = unit(.3, 'cm'),
-        text=element_text(size=17, family="lm"))
+        text=element_text(size=25))
 
 ggsave("Objects/sum_stats_year.pdf", width=10, height=7, units="in")
 

### 2022-08-17 Presentations/YESconference.tex

@@
 
-\usetheme[progressbar=frametitle, sectionpage=none]{metropolis}
+\usetheme[progressbar=frametitle]{metropolis}
 \usepackage{appendixnumberbeamer}
-\usepackage{pgfpages}
+
 \usepackage{booktabs}
-\usepackage[dvipsnames]{xcolor}
-\usepackage{mathtools}
+\usepackage[scale=2]{ccicons}
+\usepackage{bbm}
 
 \usepackage{pgfplots}
-
 \usepgfplotslibrary{dateplot}
 
-\setbeameroption{slides only}
-%\setbeameroption{show only notes}
-%\setbeameroption{show notes on second screen=right}
 \usepackage{xspace}
@@
 
-\title{Labor Markets and Technological Change: Evidence from Electronic Health Records}
-
-\subtitle{Hanna Glenn\\ \small Emory University}
- \date{\today}
-
-\setbeamertemplate{caption}{\raggedright\insertcaption\par}
-\setbeamertemplate{itemize item}{\scalebox{.6}{$\blacktriangleright$ }}     
-\setbeamertemplate{itemize subitem}{\scalebox{.8}{$\centerdot$ }}
+\title{Labor Markets and Technological Change: }
+\subtitle{Evidence from Electronic Health Records}
+% \date{\today}
+\date{}
+\author{Hanna Glenn}
+\institute{Emory University}
+% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}
 
@@
 
+\setbeamercolor{background canvas}{bg=white}
+
 \maketitle
-
-\setbeamercolor{background canvas}{bg=white}
 
@@
 
-\begin{frame}{Big Picture}
+\begin{frame}[fragile]{Big Picture}
 How does technology impact employment?
-
+                \vspace{2mm}
     \begin{itemize}
@@
     \end{itemize}
-    \vspace{3mm}
-
+                \vspace{4mm}
 Investigated in different ways
-
+                \vspace{2mm}
     \begin{itemize}
-        \item Computerization on white collar jobs
+        \item Computerization and white collar jobs
                 \vspace{2mm}
-        \item Machines on factory workers
+        \item Machinery and factory workers
     \end{itemize}
-
-    \vspace{3mm}
-
+                \vspace{4mm}
                 \pause
-
 \begin{alertblock}{$\rightarrow$ This paper considers this question in the context of healthcare and physician labor markets}\end{alertblock}
@@
 \begin{frame}[fragile]{What is an Electronic Health Record (EHR)?}
-``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"
-
-\vspace{5mm}
-\centering
-\includegraphics[scale=.35]{Objects/EHR_picture.PNG}
+Health information stored digitally in a software
+                \vspace{4mm}
+\begin{center}
+\includegraphics[scale=.4]{Objects/EHR_picture.PNG}
+\end{center}
 \end{frame}
 
-\begin{frame}[fragile]{HIT: Great (Expected) Potential in Healthcare}
-\begin{alertblock}{Cost Saving}
-\begin{itemize}
-    \item Possible cost reduction of hundreds of billions of dollars \\ \scriptsize (Hillestad et al 2005)
-\end{itemize}
-\end{alertblock}
+\begin{frame}{Adoption in Hospitals}
+\centering
+\includegraphics[scale=.4]{graphics/timeline.PNG}
+\end{frame}
 
-\begin{alertblock}{Quality Improvement}
-\begin{itemize}
-    \item Improved efficiency, patient safety improvements, physicians have decision support that could prevent unnecessary complications, etc.
-\end{itemize}
-\end{alertblock}
+\begin{frame}{Physician Response to EHRs}
+\centering
+\includegraphics[scale=.3]{graphics/News Clip3.PNG}
 
@@
 
-Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs
+\includegraphics[scale=.3]{graphics/News Clip2.PNG}
+\end{frame}
+
+\begin{frame}{This Paper}
+Did EHR implementation in hospitals affect physician labor market decisions?
+                \vspace{4mm}
+\begin{center}
+\includegraphics[scale=.4]{Objects/EHR_FlowChart_General.PNG}
+\end{center}
+\end{frame}
+
+\section{Data and Analysis}
+
+\begin{frame}[fragile]{Data}
+Panel of hospitalists from 2009-2017 with information on whether the hospital uses an EHR and various labor market outcomes\\
+                \vspace{3mm}
 \begin{itemize}
-    \item \textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
+    \item EHR information ends in 2015, so I limit to only those who were exposed by 2015
 \end{itemize}
@@
 
-\begin{frame}{Physician Response to EHRs}
-\begin{center}
-    \includegraphics[scale=.3]{graphics/News Clip3.PNG}
-
+\begin{frame}{Difference in Difference Analysis}
+Treatment: Exposure to EHR\\
+                \vspace{4mm}
+Since I have staggered treatment timing, I use Callaway and Sant'Anna (2021) average group time treatment effects:
+$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
                 \vspace{3mm}
-
-    \includegraphics[scale=.3]{graphics/News Clip2.PNG}
-\end{center}
-\end{frame}
-
-\begin{frame}{This Paper}
-$\rightarrow$ Did EHR implementation in hospitals affect physician labor market decisions?
-
-\vspace{5mm}
-\centering
-\includegraphics[scale=.45]{Objects/EHR_FlowChart_General.PNG}
-\end{frame}
-
-\section{Analysis and Data}
-
-\begin{frame}{Difference in Difference Analysis}
-Treatment: Exposure to EHR
-
-\vspace{3mm}
-
-Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects
-
-$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
-
-\vspace{3mm}
-
 \begin{itemize}
-\item Not-yet-treated units as control group
-    \item Aggregate these over groups to get a familiar event study plot
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Data}
-Panel of hospitalists from 2009-2017 with information on whether the hospital uses an EHR and various labor market outcomes
-
-\vspace{4mm}
-\pause
-\begin{itemize}
-    \item EHR information ends in 2015, so I limit to only those who were exposed by 2015
+    \item Not-yet-treated units
+    \item Aggregate to get familiar event study plot
 \end{itemize}
@@
 \centering
-    \includegraphics[scale=.5]{Objects/EHR_FlowChart_Retire.PNG}
+\includegraphics[scale=.4]{Objects/EHR_FlowChart_Retire.PNG}
 \end{frame}
@@
 \begin{frame}{Retirement Variable}
-Indicator variable equal to 1 if the physician drops out of the data 
+Indicator equal to 1 if physician drops out of the data\\
                     \vspace{3mm}
 \begin{itemize}
-    \item No longer seeing patients
+    \item No longer sees patients
 \end{itemize}
@@
 \begin{frame}{Physician Incentives}
-Decision to retire is different than we usually think of
+Decision to retire is different than we usually think
                 \vspace{3mm}
 \begin{itemize}
-        \item Retirement is thought of as a function of pensions, and sometimes depends on shocks such as health or job loss
-        \tiny (Hall and Johnson 1980, ILRR), (Gordon and Blinder 1980, J Public Econ), (Hamermesh 1985, JOLE), $\cdot$
+    \item Retirement is thought of as a function of pensions, and sometimes depends on shocks such as health or job loss\\ \scriptsize{(Hall and Johnson 1980, ILRR), (Gordon and Blinder 1980, J Public Econ), (Hamermesh 1985, JOLE)}
+                \vspace{3mm}
+                \normalsize
+    \item On average, physicians plan to retire at age 60 but don't actually retire until age 69 \scriptsize{(Collier 2017)}
+\end{itemize}
+\end{frame}
 
-        \vspace{3mm}
-
-        \normalsize \item On average, physicians plan to retire at age 60 but don't actually retire until age 69 \tiny(Collier 2017)
-    \end{itemize}
-
-\vspace{3mm}
-\pause
-
-\normalsize Unique set of incentives for which a shock to the work setting \textit{may} lead to retirement more-so than other occupations, for which studies yield mixed results  \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)
-
-\vspace{3mm}
-\pause
-
-\normalsize Related studies find that well educated workers exited at higher rates due to computerization \tiny (Willis and Hudomiet 2021, Labour), (Dillender \& Forsythe 2022, NBER)
+\begin{frame}{Physician Incentives}
+Unique set of incentives for which a shock to the work setting \textit{may} lead to retirement more-so than other occupations, for which studies yield mixed results\\
+\scriptsize{(Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)}\\
+                \vspace{15mm}
+                \normalsize
+Related studies find that well educated workers exited at higher rates due to computerization\\
+\scriptsize (Willis and Hudomiet 2021, Labour), (Dillender and Forsythe 2022, NBER)
 \end{frame}
@@
 \begin{frame}{Leaving Clinical Setting}
+Retirement is defined using whether the physician sees patients
+                \vspace{3mm}
 \begin{itemize}
-        \item Retirement is loosely defined
-        \vspace{5mm}
-        \item Any age physician may decide to leave a clinical setting
+    \item Physicians of any age could choose to leave a clinical setting
 \end{itemize}
 \end{frame}
-
 \begin{frame}{Retirement}
@@
 
-\begin{frame}{Retirement: Senior Physicians}
+\begin{frame}{Retirement: Age Groups}
 \begin{figure}[ht]
@@
 \end{figure}
-$\rightarrow$ 25\% increase relative to the mean in the first year, but 0 in the second\\
-$\rightarrow$ What does this mean for younger physicians?
+$\rightarrow$ Older: 25\% increase relative to the mean in the first year, but 0 in the second\\
+$\rightarrow$ Younger: 16\% increase in second year
 \end{frame}
@@
 
-\begin{frame}{Job Switching Variables}
+\begin{frame}{Job Setting Variables}
 \begin{enumerate}
-        \item Fraction of Patients Seen in Office 
+    \item Fraction of patients seen in office 
+                \vspace{3mm}
+    \begin{itemize}
+        \item Broken up by whether the physician already worked in an office pre-EHR
+    \end{itemize}
                 \vspace{3mm}
@@
                 \vspace{3mm}
-            \item Literature shows job switching is closely tied to job satisfaction \tiny (Aklerlof, Rose and Yellen 1988, Brookings), (Chadi and Hetschko 2017, JEMS)
+        \item Literature shows job switching is closely tied to job satisfaction\\ \scriptsize (Aklerlof, Rose and Yellen 1988, Brookings), (Chadi and Hetschko 2017, JEMS)
     \end{itemize}
@@
 
-\begin{frame}{Fraction of Patients Seen in Office}
+\begin{frame}{Results: Frac of Patients in Office (Prior Work in Office)}
 \begin{figure}[ht]
 \centering
-    \includegraphics[scale=.35]{Objects/Presentation_fracoffice_all.pdf}
+\includegraphics[scale=.35]{Objects/Presentation_frac_prior_all.pdf}
 \end{figure}
-$\rightarrow$ No effect!
+$\rightarrow$ Decrease fraction of patients
+\end{frame}
+
+\begin{frame}{Results: Frac of Patients in Office (No Prior Work in Office)}
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=.35]{Objects/Presentation_frac_noprior_all.pdf}
+\end{figure}
+$\rightarrow$ Increase fraction of patients
 \end{frame}
@@
 \begin{frame}{Physician Incentives}
-Past research on information technology $\rightarrow$ productivity typically indicates increase \tiny (Bartel, Ichniowski and Shaw 2007, QJE)
+Past research on information technology $\rightarrow$ productivity typically increases \scriptsize (Bartel, Ichniowski and Shaw 2007, QJE)
                 \vspace{3mm}
 \begin{itemize}
-    \normalsize \item Sometimes depends on managerial style \tiny(Garicano and Heaton 2010, JLE)
+                \normalsize 
+    \item Sometimes depends on managerial style \scriptsize(Garicano and Heaton 2010, JLE)
 \end{itemize}
-
                 \vspace{6mm}
-
-\normalsize We can see how productivity is affected for physicians who don't change behavior as a result of EHRs
+                \normalsize 
+We can see how productivity is affected for physicians who don't change behavior as a result of EHRs
                 \vspace{3mm}
@@
                 \vspace{3mm}
-        \item Claim Count per patient is a matter of physician incentives
+    \item Claim count per patient is a matter of physician incentives
 \end{itemize}
@@
 
-\begin{frame}{Patient Count: Senior Physicians}
+\begin{frame}{Patient Count Broken Down by Age}
 \begin{figure}[ht]
 \centering
-    \includegraphics[scale=.4]{Objects/Presentation_patients_old.pdf}
+\includegraphics[scale=.4]{Objects/Presentation_patients_ages.pdf}
 \end{figure}
+$\rightarrow$ Don't see the same increase for older physicians
 \end{frame}
@@
 \end{figure}
-$\rightarrow$ No significant change, but a downward trend is possible
+$\rightarrow$ Downward trend
 \end{frame}
@@
     \item For physicians who endure through the implementation of the technology, number of patients goes up but not conclusive evidence that claims per patient goes down
-
 \end{itemize}
@@
     \item Endogeneity of treatment
+            \vspace{3mm}
     \item Data assistants
-        \item Limiting Years and Using Never-treated as control
+            \vspace{3mm}
+    \item Limiting years and using never-treated as control
+            \vspace{3mm}
     \item Anticipation

### 2022-08-17 Presentations/YESconference.tex

@@
 \begin{enumerate}
-    \item Fraction of patients seen in office 
+    \item Indicator for working in office or not
                     \vspace{3mm}
-    \begin{itemize}
-        \item Broken up by whether the physician already worked in an office pre-EHR
-    \end{itemize}
-                \vspace{3mm}
-    \item Indicator for working in office or not
+    \item Indicator for whether the physician changed zip code
 \end{enumerate}
@@
 
-\begin{frame}{Results: Frac of Patients in Office (Prior Work in Office)}
-\begin{figure}[ht]
-\centering
-\includegraphics[scale=.35]{Objects/Presentation_frac_prior_all.pdf}
-\end{figure}
-$\rightarrow$ Decrease fraction of patients
-\end{frame}
-
-\begin{frame}{Results: Frac of Patients in Office (No Prior Work in Office)}
-\begin{figure}[ht]
-\centering
-\includegraphics[scale=.35]{Objects/Presentation_frac_noprior_all.pdf}
-\end{figure}
-$\rightarrow$ Increase fraction of patients
-\end{frame}
-
-\begin{frame}{Indicator for working in office}
+\begin{frame}{Results: Indicator for Working in Office}
 \begin{figure}[ht]
@@
 $\rightarrow$ Driven by young physicians
+\end{frame}
+
+\begin{frame}{Results: Indicator for Changing Zip Codes}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.35]{Objects/Presentation_zip_all.pdf}
+\end{figure}
+$\rightarrow$ Something happening in relative year -3, may be a pre-trend
+\end{frame}
+
+\begin{frame}{Results: Indicator for Changing Zip Codes}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.35]{Objects/Presentation_zip_ages.pdf}
+\end{figure}
+$\rightarrow$ .043 ppt increase in likelihood of retiring for younger physicians (36\%)
+$\rightarrow$ This effect is persistent
 \end{frame}
@@
     \begin{enumerate}
-        \item Patient Count
+        \item Patient count
         \vspace{6mm}
@@
 
-\begin{frame}{Patient Count}
+\begin{frame}{Results: Patient Count}
 \begin{figure}[ht]
@@
 
-\begin{frame}{Claims per Patient}
+\begin{frame}{Results: Claims per Patient}
 \begin{figure}[ht]
@@
 \begin{frame}{Summary}
-\begin{itemize}
-    \item Evidence that physicians are changing their behavior because of this new technology, specifically through retirement or job switching
-                \vspace{3mm}
-    \item Little evidence of work setting changes
-                \vspace{3mm}
-    \item For physicians who endure through the implementation of the technology, number of patients goes up but not conclusive evidence that claims per patient goes down
-\end{itemize}
+\textcolor{red}{Downside}: evidence that physicians are changing their behavior because of this new technology\\
+                \vspace{6mm}
+\textcolor{teal}{Potential upside}: For physicians who endure through the implementation of the technology, number of patients goes up and claims per patient goes down
+
 \end{frame}

### 2022-08-23 Paper/Paper.tex
Updates from Overleaf

@@
 \begin{abstract}
-    Technological innovation has affected workers and labor markets for decades, yet the effects of modern technologies in some settings are still not understood.
+    Technological innovation has affected workers and labor markets for decades, yet the effects of modern technologies in some vocations with unique incentives are still not understood.
 I add to the understanding of technology implementation's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
 health care, electronic health records (EHRs).
-I treat EHR implementation in hospitals as an exogenous treatment to hospitalists and estimate average group time treatment effects on various labor market outcomes.
+I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes.
 I find that physicians are 10\% more likely to retire, and 4\% more likely to work in an office if they choose not to retire.
-Finally, for hospitalists who remain exposed to the technology, number of patients seen increases.
-These results point to adverse effects that can occur when technology is poorly implemented in technical occupations.
+Finally, for physicians who remain exposed to the technology, the number of patients seen increases.
+However, claims per patient decreases.
+These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential efficiency gains for those who do not respond adversely.
 \end{abstract}
@@
 \section{Introduction}
-
 How technological innovation affects labor markets is heavily studied, in part because of the vast differences in technology and tasks present even within industries.
@@
 The implementation and use of EHRs were incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
-Since many physicians working in hospitals were exposed to this technology beyond their control, this is an opportune setting to investigate the relationship between technology implementation and labor market decisions for highly skilled workers.
+Since many physicians working in hospitals were exposed to this technology beyond their control, this is an opportune setting to investigate the relationship between rapid technology implementation and labor market decisions for highly skilled workers.
 
@@
 Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in settings that it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings.
+The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
@@
 I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists, which spans from 2009-2017.
-Using CMS Shared Patient Data, I determine which hospitals each hospitalist is connected to.
+Using CMS Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital.
 Then, I use the AHA Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
@@
 For all outcomes, the independent variable of interest is a binary treatment variable capturing exposure to an EHR.
-I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting, and location of the physician office, (3) number of patients seen, and (4) billing activity.
+I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) patient and billing activity.
 
 This paper contributes to the literature's understanding of how health information technology generally affects health care.
-There is a robust set of empirical papers examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased cost) (surveyed in \cite{Buntin2011TheResults}), more recent empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
+Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased cost) (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
 More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
@@
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this work, I find that EHR implementation led to a 0.006 ppt increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
-Physicians less than 60 were 0.002 ppts more likely to retire after exposure, possibly pointing to career switching since retirement is measured as a lack of Medicare patients seen.
-These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence physicians learning under them.
+Contrary to this result, I find that EHR implementation led to a 0.006 ppt increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
+Physicians less than 60 were 0.002 ppts more likely to retire after exposure, possibly pointing to career switching.
+These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians.
 For those who did not retire, I find that young physicians were .014 ppts (4.6\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure.
-Finally, I find a modest, positive effect on patient count completely driven by younger physicians, and no effect on billing activity for either age group.
+Finally, I find a modest, positive effect on patient count completely driven by younger physicians, and a decreasing trend in claims per patient.
 
@@
 If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence.
-I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast EHR implementation.
+I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast amount of EHR implementation.
 Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
@@
 \subsection{Background on EHRs}
-
 EHRs have been an important feature of health care since the 1980s.
@@
     \includegraphics[scale=.4]{Objects/epic-ehr-screenshot.jpg}
-    \includegraphics[scale=.115]{Objects/EHRimage2.jpg}
+    \includegraphics[scale=.11]{Objects/EHRimage2.jpg}
     \label{fig:EPIC}
@@
 
-CMS Shared Patient data records annual information on count of patients and frequency of billing associated with the same patient across providers within a given time period (30, 60, 90, or 180 days).
+CMS Shared Patient data records annual information on count of Medicare patients and frequency of billing associated with the same patient across providers within a given time period (30, 60, 90, or 180 days).
 These data are available from 2009 to 2015.
@@
 Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting.
-This binary measure of practice setting offers less variation but captures a more discrete changes in which physicians enter or leave the office-based setting entirely.
+This binary measure of practice setting offers less variation but captures a more discrete change in which physicians enter or leave the office-based setting entirely.
 Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next.
-While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from EHRs.
-Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and total billing activity.
+While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from or towards EHRs.
+Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and claims filed per patient.
 
@@
     \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.45]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
     \label{fig:treatmentgraph}
@@
 
-I treat EHR implementation as an exogenous treatment variable, and I seek to estimate average treatment effects of EHR exposure on various physician outcomes.
+For each outcome variable described in Section \ref{sec:outcome}, I treat EHR implementation as an exogenous treatment variable, and I seek to estimate average treatment effects of EHR exposure.
 In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
@@
 Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot.
-In some cases, I present the de-aggregated estimates to show that results are not driven by a particular year of treatment.
+For select outcomes, I present the original de-aggregated estimates to show that results are not driven by a particular year of treatment.
 Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each table.
@@
 That is, once a physician is exposed to an EHR, they cannot be un-exposed.
-At the hospital level, this assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure.
+For physicians that do not switch hospitals, this assumption is supported by both the institutional details of hospital EHR implementation and the data itself.
 An EHR is a costly technology and requires a significant amount of collaboration to implement.
@@
 This is why I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
+I discuss this limitation to the data further in Section \ref{sec:patientcount}.
 
@@
 In most cases, there is no evidence of a pre-trend.
-In cases where a pre-trend could be driving the results, I make no strong claims about the effect of EHR exposure.
-
-In order for this to be a causal estimate of EHR exposure on various labor market outcomes, there are additional institutional assumptions necessary.
+Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends}.
+
+To interpret these estimators as the causal effect of EHR exposure on various labor market outcomes, there are additional institutional assumptions necessary.
 First, I assume that when a physician works in an electronic record utilizing-hospital, they utilize the electronic health record system fully.

### 2022-08-23 Presentations/YESconference.tex
Updates from Overleaf

@@
 \includegraphics[scale=.3]{graphics/News Clip2.PNG}
+\end{frame}
+
+\begin{frame}{Physician Response to EHRs}
+\centering
+\includegraphics[scale=.4]{graphics/EHR tweet.PNG}
 \end{frame}
@@
 \end{figure}
-$\rightarrow$ 4.6\% increase relative to the mean\\
-$\rightarrow$ Driven by young physicians
+$\rightarrow$ 4.6\% increase relative to the mean
+\end{frame}
+
+\begin{frame}{Results: Indicator for Working in Office (Ages)}
+\begin{figure}[ht]
+    \centering
+    \includegraphics[scale=.35]{Objects/Presentation_office_ages.pdf}
+\end{figure}
+$\rightarrow$ Driven by younger physicians
 \end{frame}
@@
 
-\begin{frame}{Results: Indicator for Changing Zip Codes}
+\begin{frame}{Results: Indicator for Changing Zip Codes (Ages)}
 \begin{figure}[ht]
@@
                 \vspace{6mm}
-\textcolor{teal}{Potential upside}: For physicians who endure through the implementation of the technology, number of patients goes up and claims per patient goes down
+\textcolor{teal}{Potential upside}: For physicians who endure through the implementation of the technology, number of patients goes up and claims per patient goes down\\
+                \vspace{6mm}
+\underline{Implications}: The way technology is implemented and used is important given the response of the user
 

### 2022-08-23 Table Code/overall stats.tex
Updates from Overleaf

@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 179,883 & 0.03 & 0.18 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 179,883 & 0.10 & 0.24 & 0 & 1\\
-\hspace{1em}Work in an Office & 179,883 & 0.30 & 0.46 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 161,558 & 0.12 & 0.33 & 0 & 1\\
-\hspace{1em}Number of Patients & 179,883 & 330 & 274 & 0 & 7,613\\
-\hspace{1em}Claim Count & 179,883 & 1,382 & 1,789 & 0 & 64,508\\
+\hspace{1em}Ever Retire & 179,532 & 0.03 & 0.17 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 179,532 & 0.10 & 0.24 & 0 & 1\\
+\hspace{1em}Work in an Office & 179,532 & 0.30 & 0.46 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 161,273 & 0.12 & 0.33 & 0 & 1\\
+\hspace{1em}Number of Patients & 179,532 & 330.22 & 274.33 & 0 & 7,613\\
+\hspace{1em}Claims per Patient & 163,125 & 4.24 & 4.60 & 0 & 679\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 179,883 & 0.75 & 0.43 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 179,883 & 0.63 & 0.44 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 179,532 & 0.75 & 0.43 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 179,532 & 0.63 & 0.44 & 0 & 1\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 179,883 & 44.9 & 8.4 & 25 & 86\\
-\hspace{1em}Female & 179,883 & 0.34 & 0.47 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 179,883 & 1.49 & 0.82 & 1 & 12\\
-\hspace{1em}Number of Systems Worked With & 179,883 & 1.31 & 0.60 & 1 & 8\\
+\hspace{1em}Age & 179,532 & 44.87 & 8.43 & 25 & 86\\
+\hspace{1em}Female & 179,532 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 179,532 & 1.49 & 0.82 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 179,532 & 1.31 & 0.60 & 1 & 8\\
 \bottomrule
 \end{tabular}
+\end{table}
 \label{tab:sumstats}
-\end{table}

### 2022-08-23 Table Code/split stats.tex

--- 2022-04-15 Table Code/split stats.tex
@@
 \centering
-\begin{tabular}[t]{lccc}
+\begin{tabular}[t]{lcc}
 \toprule
-Variable & Age $>$ 60 & Age $<=$ 60 & Any Who Retire\\
+Variable & Age $>$ 60 & Age $<=$ 60\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 0.02 & 0.02\\
-\hspace{1em}Fraction Patients in Office & 0.44 & 0.27 \\
-\hspace{1em}Work in an Office & 0.78 & 0.52 \\
-\hspace{1em}Change Zip Codes & 0.06 & 0.09 \\
-\hspace{1em}Number of Patients & 431.96 & 357.46 \\
-\hspace{1em}Claim Count & 3216.46 & 2019.17\\
+\multicolumn{3}{l}{\textbf{Outcomes}}\\
+\hspace{1em}Ever Retire & 0.04 & 0.03\\
+\hspace{1em}Fraction Patients in Office & 0.15 & 0.09\\
+\hspace{1em}Work in an Office & 0.48 & 0.28\\
+\hspace{1em}Change Zip Codes & 0.10 & 0.12\\
+\hspace{1em}Number of Patients & 342 & 329\\
+\hspace{1em}Claim Count & 1,767 & 1,332\\
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 0.77 & 0.76 & 0.77\\
-\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.63 & 0.64\\
+\multicolumn{3}{l}{\textbf{Treatment}}\\
+\hspace{1em}Exposure to an EHR & 0.76 & 0.75\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.59 & 0.64\\
 \addlinespace[0.3em]
-\multicolumn{4}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 61.07 & 44.06 & 48.29\\
-\hspace{1em}Female & 0.15 & 0.35 & 0.39\\
-\hspace{1em}Number of Hospitals Worked With & 1.67 & 1.53 & 1.35\\
-\hspace{1em}Number of Systems Worked With & 1.40 & 1.33 & 1.19\\
+\multicolumn{3}{l}{\textbf{Characteristics}}\\
+\hspace{1em}Age & 60.31 & 42.75\\
+\hspace{1em}Female & 0.18 & 0.36\\
+\hspace{1em}Number of Hospitals Worked With & 1.63 & 1.48\\
+\hspace{1em}Number of Systems Worked With & 1.38 & 1.30\\
 \bottomrule
 \end{tabular}
+\end{table}
 \label{tab:splitstats}
-\end{table}

### 2022-08-23 Scripts/analysis.R
add analysis of whether EHR impacts number of physicians

@@ -131,7 +131,11 @@ graph_data <- lapply(Models_agg, function(x){
   
   data <- rbind(data_allages, data_young, data_old)
   
-  list(data)
+  p_all <- x[["p_all"]]
+  p_young <- x[["p_young"]]
+  p_old <- x[["p_old"]]
+  
+  list(data, p_all, p_young, p_old)
 })
 
 dodge <- position_dodge(width=0.3) 
@@ -151,10 +155,12 @@ graphs <- lapply(graph_data, function(x){
     theme_light() +
     geom_hline(yintercept=0, linetype="dashed") +
     theme(legend.position = "none", text = element_text(size = 15)) +
-    xlab("\n") + ylab("Point Estimate and 95% CI\n") +
+    xlab("") + ylab("Point Estimate and 95% CI\n") +
     geom_vline(xintercept=0, linetype="dashed", color="red") +
     theme(panel.grid.major.x = element_blank() ,
-          panel.grid.major.y = element_line(size=.05, color="lightgray" ))
+          panel.grid.major.y = element_line(size=.05, color="lightgray" )) +
+    labs(caption=paste0("pre-trends test p-value: ",round(x[[2]],2))) + 
+    theme(plot.caption=element_text(hjust = 0))
   
   ages <- ggplot(ages_data, aes(year, att, color=group)) +  
     geom_vline(xintercept="0", linetype="dashed", colour="red") +
@@ -164,12 +170,15 @@ graphs <- lapply(graph_data, function(x){
     theme_light() +
     geom_hline(yintercept=0, linetype="dashed") +
     theme(text = element_text(size = 15)) +
-    xlab("\n") + ylab("Point Estimate and 95% CI\n") +
+    xlab("") + ylab("Point Estimate and 95% CI\n") +
     geom_vline(xintercept=0, linetype="dashed", color="red") +
     theme(panel.grid.major.x = element_blank() ,
           panel.grid.major.y = element_line(size=.05, color="lightgray" )) +
     paletteer::scale_colour_paletteer_d("ggthemes::excel_Badge") +
-    theme(legend.position="bottom") + labs(color='Age Group')
+    theme(legend.position="bottom") + 
+    labs(color='Age Group', caption=paste0("pre-trends test p-value (young): ",round(x[[3]],2),
+                                           "\npre-trends test p-value (old): ", round(x[[4]],2))) + 
+    theme(plot.caption=element_text(hjust = 0))
   
   list(all=all, ages=ages)
 })

### 2022-08-23 Scripts/data6_dataassistant.R

@@ -37,7 +37,7 @@ npi.taxcodes <- npi.taxcodes %>%
 
 ggplot(npi.taxcodes, aes(x=enum_year)) + 
   geom_histogram(binwidth=1, colour="black", fill="white") + ylab("Count\n") + xlab("\nNPI Enumeration Year") +
-  theme(text=element_text(size=17,family="lm"))
+  theme(text=element_text(size=15))
 ggsave("Objects/dataassistant_histogram.pdf", width=8, height=5 , units = "in")
 
 

### 2022-08-23 Scripts/sensitivity_analysis.R

@@ -1607,4 +1607,193 @@ ggsave(claim_pretrends_plot, file="Objects/claim_pretrends_plot.pdf", height=7, 
 
 
 
+## ANALYSIS ON WHETHER EHR IMPACTS NUMBER OF PHYSICIANS
 
+# Read in data on physician hospital pairs
+data <- read_rds(paste0(created_data_path, "phys_hosp_pairs.rds"))
+
+#Read in AHA-NPI crosswalk data
+AHANPI_cw <- read_rds(paste0(created_data_path,"AHANPI_cw.rds"))
+
+# Merge crosswalk to main data
+data <- data %>%
+  mutate(HospNPI=as.double(HospNPI)) %>%
+  left_join(AHANPI_cw,by=c("HospNPI"="NPI"))
+
+data <- data %>%
+  filter(!is.na(AHAID))
+
+
+# Read in main AHA survey data
+# I need to think about adding more hospital characteristics to this dataset
+AHAmainsurvey <- read_csv(paste0(raw_data_path,"AHA_mainsurvey.csv"))
+
+# I define low integration and high integration as in Madison (2004, HSR)
+AHAmainsurvey <- AHAmainsurvey %>%
+  mutate(ID=as.character(ID)) %>%
+  mutate(low_integration=ifelse(IPAHOS==1 | OPHOHOS==1 | CPHOHOS==1,1,0),
+         high_integration=ifelse(GPWWHOS==1 | MSOHOS==1 | ISMHOS==1 | EQMODHOS==1,1,0)) %>%
+  dplyr::rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
+                joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
+  select(-IPAHOS, -OPHOHOS, -CPHOHOS, -GPWWHOS, -MSOHOS, -ISMHOS, -EQMODHOS) %>%
+  select(-IPAP, -GPWP, -OPHP, -CPHP, -ISMP, -EQMP, -FNDP, -full_year, -EHLTHI, -EHLTHS,
+         -total_physicians, -joint_phys, -phys_owned)
+
+data <- data %>% ungroup() %>%
+  mutate(AHAID=as.character(AHAID)) %>%
+  left_join(AHAmainsurvey,by=c("AHAID"="ID","year"),na_matches="never")
+
+
+
[…]

### 2022-08-23 Scripts/sumstats.R

@@ -48,7 +48,7 @@ sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
                  "Ever Retire"="ever_retire",
                  "Work in an Office"="work_in_office",
                  "Change Zip Codes"="change_zip",
-                 "Claim Count"="claim_count_total"), 
+                 "Claims per Patient"="claim_per_patient"), 
                list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))), na.rm=TRUE) %>%
   mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
   gather(key=var,value=value) %>%
@@ -73,7 +73,7 @@ knitr::kable(sum_stats_fullsample[c(4,8,12,2,10,3,5,7,1,6,9,11),],
   pack_rows(index = c("Outcomes" = 6, "Treatment" = 2, "Characteristics" = 4))
 
 
-# Summary Stats of all variables by old vs. young vs. those who retire ---------------------------------------------------------
+# Summary Stats of all variables by old vs. young  ---------------------------------------------------------
 means_old <- Physician_Data %>% ungroup() %>%
   filter(minyr_EHR>0) %>%
   filter(max_age>59) %>%
@@ -126,11 +126,10 @@ means_retire <- Physician_Data %>% ungroup() %>%
   dplyr::rename(value_retire=value)
 
 means_bind <- means_old %>%
-  left_join(means_young,by="var") %>%
-  left_join(means_retire,by="var")
+  left_join(means_young,by="var") 
 
 knitr::kable(means_bind[c(9,8,10,11,5,12,7,6,4,2,1,3),], "latex",
-             col.names=c("Variable","Age $>$ 60", "Age $<=$ 60", "Any Who Retire"),
+             col.names=c("Variable","Age $>$ 60", "Age $<=$ 60"),
              digits=2,
              caption="Means by Age Sample",
              booktabs=TRUE,
@@ -393,7 +392,7 @@ AHA_sample <- AHAmainsurvey %>%
 
 
 
-panelview(AHA_sample, Y=NULL,D="EHR", index=c("HospNPI","year"), axis.lab = "time", by.timing=TRUE) +
-  theme(text=element_text(size=10,family="lm"))
+panelview(AHA_sample, Y=NULL,D="EHR", index=c("HospNPI","year"), axis.lab="time", by.timing=TRUE) +
[…]

### 2022-08-23 Paper/Paper.tex

@@
     \caption{Hospital EHR Adoption Over Time}
-    \includegraphics[scale=.7]{Objects/hosp_treat.pdf}
+    \includegraphics[scale=.5]{Objects/hosp_treat.pdf}
     \label{fig:hosp_treat}
@@
 
-This outcome, total patient count, is directly related to the fraction of patients seen in an office setting.
-For younger physicians, since the total number of patients increases in the first couple of years after exposure, the absolute number of patients seen in offices increases as well.
-That is, it seems that younger physicians are shifting a positive amount of work towards offices but still able to increase productivity in the time spent in the hospital, thus the null effect on fraction of office patients.
+Since I found in the earlier stages of analysis that physicians may have shifted away from hospitals due to EHR exposure, I also investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
+For example, if a hospital loses a portion of their physicians, they may divide that physician's patients between the remaining physicians.
+In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away my results could be driven by the retiring physicians instead of EHR use.
+In Figure
+
+\begin{figure}[ht]
+    \centering
+    \captionsetup{width=.85\linewidth}
+    \caption{Effect of EHR Exposure on Claim Count}
+    \includegraphics[scale=.6]{Objects/number.pdf}
+    \label{fig:claim}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
+The bottom show these results for different subgroups of physicians by age.
+The p-value listed for each graph corresponds to a Wald test for pre-trends.
+Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
+\end{figure}
 

### 2022-09-02 Paper/Paper.tex

@@
 
-I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists, which spans from 2009-2017.
+I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
 Using CMS Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital.
@@
 
-In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age.
+In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age.
 These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
-The top panel suggests that for any age physician, being exposed to an EHR leads to a .003 ppt increase in the likelihood of retiring in the first year after exposure, and a .0045 ppt increase in the second year after exposure.
-While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .03.
-Relative to this average, the effect represents a 10\% (or 15\% in the second year) increase in the likelihood of retirement due to EHR exposure.
-While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates yields confidence in a positive effect truly occurring.
-Further, even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
+The estimates in the top panel suggest that being exposed to an EHR leads to a .003 ppt increase in the likelihood of retiring in the first year after exposure, and a .0045 ppt increase in the second year after exposure.
+While numerically small, these effects are statistically and economically meaningful relative to the proportion of physicians who retire in the sample, .03.
+That is, the effect represents a 10\% (or 15\% in the second year) increase in the likelihood of retirement relative to the mean.
+Even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
 Therefore, in Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
-These results are largely similar to those presented here.
+The results are largely similar to those presented here.
 
@@
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
-These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above in the year after exposure, where younger physicians are driving the result for the second year after exposure.
+These results are shown in the bottom panel of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above in the year after exposure, while younger physicians are driving the result for the second year after exposure.
 A physician who is at least 60 years old is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
-In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure.
-Physicians in the young category are also 25\% more likely to retire, just in the second year after exposure.
-One reason this could occur is if younger physicians are changing careers due to EHR exposure, where they no longer see patients in a clinical setting, then it may take more time to prepare for such a switch.
-Whereas formal retirement can be decided and carried out more immediately.
-Another possibility is that these physicians are shifting away from Medicare patients.
-
-\subsection{Work Setting}
-
- I consider the decision to switch work settings using multiple outcomes.
-First, I investigate the extent to which hospitalists work in office based setting and whether this changes due to EHR exposure.
-I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office based setting in a given year.
+In comparison, physicians less than 60 are far less likely, if at all, to respond in the first year after EHR exposure.
+Physicians less than 60 are approximately 16\% more likely to retire two years after exposure.
+There are two potential reasons for the different response time.
+First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
+Alternatively, physicians who are not yet retirement age may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
+
+One argument against this result is that physicians are not leaving a clinical environment, they are simply choosing to avoid seeing Medicare patients.
+While EHR use should not be correlated specifically with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
+The Center for Medicare and Medicaid services publishes a list of physicians who have officially opted out of seeing Medicare patients.
+None of the physicians in my sample have selected into this option.
+Further, the differential timing of retirement from older vs.
+younger physicians is reassuring, as any Medicare change should not affect physicians of different ages differently.
+
+\subsection{Place of Work}
+
+There is still a large portion of physicians not induced to retire due to EHR exposure.
+Thus, I consider the decision to switch work settings using multiple outcomes.
+First, I investigate the extent to which hospitalists work in office and whether this changes due to EHR exposure.
+I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year.
+Second, I investigate whether EHR exposure causes a change in the likelihood of switching zip codes.
 This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
@@
 In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away my results could be driven by the retiring physicians instead of EHR use.
-In Figure
+In Figure \ref{fig:numphys}, I present a graph showing the relationship between EHR adoption and the number of physicians working in the hospital.
+While these are not causal estimates of the effect of EHR implementation on the number of physicians working in a hospital, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation.
+This suggests that the increase in patient count is not due to the exit of physicians observed in the earlier analysis.
 
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Claim Count}
-    \includegraphics[scale=.6]{Objects/number.pdf}
-    \label{fig:claim}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
-The bottom show these results for different subgroups of physicians by age.
-The p-value listed for each graph corresponds to a Wald test for pre-trends.
-Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
+    \caption{Effect of EHR Exposure on Number of Physicians in Hospitals}
+    \includegraphics[scale=.4]{Objects/numberofphysicians_plot.pdf}
+    \label{fig:numphys}
 \end{figure}

### 2022-09-02 Paper/Paper.tex

@@
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
+Further, the cost imposed is likely to vary based on the type of worker, in particular, age group.
 Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
@@
 While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire.
-Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life.
+Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life factors.
 Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction.
@@
 Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role.
-There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management.
-While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching.
-This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how EHRs impact job satisfaction of current clinical work.
+There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management.
+While it is unlikely that physicians under some age threshold are formally retiring, physicians of any age could be career switching.
+This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction of current clinical work.
 As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
@@
 
-For most physicians, EHRs will not impose a costly enough burden to induce retirement.
-However, there are other ways physicians might change their behavior under sufficient incentive to avoid the technology.
-For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over the technologies used in the work environment, affecting overall job satisfaction.
-Generally, physicians may have the ability to move hospitals, move from a hospital to an office, or work in multiple facilities and shift the composition of patients towards other facilities.
+For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of being in the labor market.
+However, there are other ways physicians might change their behavior under drastic changes to their current working environment.
+For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over which technologies they use.
+Generally, physicians may have the ability to change hospitals, move from a hospital to an office, or work in multiple facilities and change the amount of work done in different facilities.
 I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}).
 Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
-
-First, I consider those who are in a relatively binding contract.
+While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faces by each type.
+
+First, I consider those who are in a relatively binding contract with the hospital.
 For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital.
@@
 When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.
-Alternatively, there are hospitalists that work in multiple hospitals or work in both hospitals and offices.
+Thus, I classify these physicians as "sticky", since they are not able to change behavior in the short run.
+
+Alternatively, there may be hospitalists that work in multiple hospitals or work in both hospitals and offices throughout the sample period.
 If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.
+As this type of hospitalist is less sticky, if there is any effect of EHR exposure on work setting it would be among this type.
 
@@
 There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation.
-A lack of response could be due to 3 factors: EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
+A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
 For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment.
@@
 A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
-Such a decrease implies a change in patient care: either a larger number of patients have access to an appointment, or the same amount of patients receive more face-to-face time with physicians during appointments.
-Alternatively, because the technology is not hugely user friendly, it may not decrease administrative burden after all.
+
+Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous.
+If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients.
+In this case, I expect productivity to increase.
+However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
+
+Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician.
+Billing activity refers to the number of claims filed per patient seen.
+EHRs may improve efficiency if they remove the need for repeated or excess testing.
+However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient.
+Thus, both productivity and efficiency are empirical questions.
 
@@
 Thus, I consider the decision to switch work settings using multiple outcomes.
-First, I investigate the extent to which hospitalists work in office and whether this changes due to EHR exposure.
+First, I investigate the extent to which hospitalists work in an office and whether this changes due to EHR exposure.
 I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year.
@@
 I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
-The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .014 percentage points, a 4.6\% increase relative to the mean.
+The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the same year as exposure by .014 ppts, a 4.6\% increase relative to the mean.
 This result indicates that hospitalists are more likely to start working in an office in the year they are exposed to an EHR, though the effect is not as large as the one found for physicians retiring.
 Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
-There is no evidence that older physicians are shifting towards offices.
-This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more stuck to their particular work environment.
+For retirement age physicians, the estimates become very noisy and I do not draw any conclusions.
+Younger physicians, however, show evidence of switching behavior.
+The .015 ppt increase in the likelihood of retiring is equivalent to a 5.4\% increase relative to the mean for this age group.
 While there is not strong evidence for the existence of a pre-trend, I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}.
@@
 
-It may be that hospitalists who are already working in office settings shift more of their time or patients towards the office and away from hospitals if EHRs impose a meaningful cost to working in a hospital.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.
-I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
-This result has varying implications depending on whether total number of patients (the denominator of this variable) is affected.
-If total number of patients remains constant, then the absolute number of patients seen in offices is also unchanged.
-However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in the same direction.
-I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.
+The previous results is driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians shift the amount of work they do in an office.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is also tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office, and the results are the same.} I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
+This does not seem to be a margin where physicians are changing their behavior due to EHR use.
 
@@
     \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
-    \includegraphics[scale=.6]{Objects/officefrac_plot.pdf}
+    \includegraphics[scale=.6]{Objects/officefrac__plot.pdf}
     \label{fig:officesecond}
@@
 
-The previous results only capture switching behavior which involves office-based settings.
+The previous results only capture switching behavior which involves office based settings.
 Another way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors.
@@
 The outcome variable is equal to 1 if their list of primary zip codes changes.
-This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes.
-While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation.
-
-The effects of EHR exposure on zip code switching over time are shown in Figure \ref{fig:zip}.
-For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is also evidence of a pre-trend that may be driving this.
-To make a more conclusive argument about the true effect under violations of the parallel trends assumption, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}.
-These results reveal that under linear violations of the parallel trends assumption, there is still a positive effect of EHR exposure on changing zip codes for younger physicians and no effect for retirement age physicians.
-A .043 ppt increase in the likelihood of changing zip codes for younger physicians indicates a 36\% increase relative to the mean, a significant increase in probability.
+This variable is limited in how broadly it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes.
+While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation that can be combined with the previous outcome variables to draw conclusions about general behavior.
+
+The effects of EHR exposure on changing zip codes over time are shown in Figure \ref{fig:zip}.
+For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is evidence that this result is driven by a pre-trend before exposure.
+Therefore, I do not make any causal arguments about this effect.
+However, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}.
+These results reveal that under linear violations of the parallel trends assumption, there is still a positive relationship between EHR exposure and changing zip codes for younger physicians.
+When taken together, these estimates suggest there is slight evidence of physicians who do not retire changing aspects of their work environment, but not on a large scale, indicating stickiness in place of work.
 
@@
 
-A limitation in this analysis is that physicians may shift work behavior due to the organization of the hospital system changing.
-For example, a hospital may acquire a smaller practice and send physicians to work there for a period of time.
-If this acquisition is correlated with incentives to adopt an EHR, then the effect I find is biased upward.
-
-\subsection{Patient Count and Billable Activity}\label{sec:patientcount}
-
-For both patient and claim count, if the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the move, not due to the EHR.
+\subsection{Patient Count and Billing Activity}\label{sec:patientcount}
+
+As a measure of productivity or efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
+If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR.
 Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
@@
 The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
-For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then no effect after the second year.
-These results translate to a 5-10\% increase in the number of patients seen relative to the mean.
-As with other outcomes, this result is totally due to changes by the younger group of physicians.
+For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then the effect disappears.
+These estimates translate to a 5-10\% increase in the number of patients seen relative to the mean.
+Interestingly, when splitting the sample by age group, the effect becomes null for those of retirement age, suggesting that technology use impacts productivity dependent on worker type (in this case, age).
 Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends}.
@@
 
-Patient count is also driven by demand for health care.
+While this paper is interested the supply side factors of patients seen, this variable is also driven by demand for health care.
 A major change in the U.S.
 health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
-If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the results.
-The ACA has potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
+If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count.
+The ACA had potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
 A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
@@
 Since I found in the earlier stages of analysis that physicians may have shifted away from hospitals due to EHR exposure, I also investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
-For example, if a hospital loses a portion of their physicians, they may divide that physician's patients between the remaining physicians.
-In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away my results could be driven by the retiring physicians instead of EHR use.
+For example, if a hospital loses a portion of their physicians, they may divide those patients between the remaining physicians.
+In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away then these estimates could be driven by the retiring physicians instead of EHR use.
 In Figure \ref{fig:numphys}, I present a graph showing the relationship between EHR adoption and the number of physicians working in the hospital.
-While these are not causal estimates of the effect of EHR implementation on the number of physicians working in a hospital, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation.
+While these are not causal estimates, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation.
 This suggests that the increase in patient count is not due to the exit of physicians observed in the earlier analysis.
@@
 
-Next, I consider whether EHR exposure affects claim count.
-EHRs may affect physician incentives in such a way that billing activity changes differently from patient count.
-If, for example, EHRs reduce the administrative burden of filing extra claims, then physicians might increase the things they bill for whether the number of patients has changed or not.
-On average for the entire sample, claim counts are roughly 4 times the number of patients seen in a year.
-Shown in Figure \ref{fig:claim}, there is not strong evidence for an effect of EHR exposure on claim count for physicians of retirement age.
-The point estimates indicate a positive effect in the first two years of exposure for younger physicians, but this is not strong evidence.
-Thus, for hospitalists $>= 60$, it seems EHRs are not making significant changes to practice behavior.
-However, for those $< 60$, the observed claim count increase is two times the observed patient count increase.
-Since the average claim count per patient is 4, these results imply that the number of claims per patient is actually decreasing.
-That is, physicians are billing for less items as they increase the number of patients.
+Next, I consider whether EHR exposure affects claims filed per patient.
+On average, claim counts are roughly 4 times the number of patients seen in a year.
+Shown in Figure \ref{fig:claim}, there is a decreasing trend in claims per patient in the three years after EHR exposure.
+Two years after exposure, claims per patient decreases by .25, roughly 6\% of the average.
+This indicates a potentially advantageous effect of EHRs, that excessive billing is reduced.
+Though noisier for retirement age physicians, this effect is seen in both age groups.
 
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Claim Count}
-    \includegraphics[scale=.6]{Objects/claim_plot.pdf}
+    \caption{Effect of EHR Exposure on Claims per Patient}
+    \includegraphics[scale=.6]{Objects/claim_per_patient_plot.pdf}
     \label{fig:claim}
@@
 
-In the observed results above, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
-Almost all specifications point to the effect of EHR exposure to become null in the 3-4 years after exposure.
+For the results presented, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
+Almost all specifications point to the effect of EHR exposure becmoing null in the 3-4 years after exposure.
 One reason for this may be the composition of the hospitals which adopt at different times.
@@
 
-One concern with the presented analysis, particularly for the productivity outcomes, is that any effect could be caused by an outside factor also correlated with EHR adoption in hospitals.
+The effects found in the above analysis could be biased if there is an external event occurring that is correlated with EHR implementation and exposure.
 The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
@@
 In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
+This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs improve both outcomes.
 
@@
 This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes.
-The interaction between health information technology and labor markets is vastly important as the technology is continually legislated and growing each year.
-
-I utilize data on shared patients, hospital EHR use, and physician claim activity.
-Using an event study framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR in a hospital, where those of retirement age have an even larger increase relative to younger physicians.
-Further, I find evidence that younger physicians are switching work settings, but that this may be driven by hospital incentives instead of physician incentives.
-Finally, physicians who remain in the labor force increase both number of patients seen and billing activity, but billing activity per patient decreases with EHR exposure.
+Understanding how physicians respond to this technology is important as physician labor markets have implications for access to care, quality of care, and costs.
+
+I utilize data on shared patients, hospital EHR use, and physician billing activity.
+Using a difference in difference framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR, where those of retirement age have an even larger increase relative to younger physicians.
+Further, I find slight evidence that physicians who do not retire change where they work because of EHR exposure.
+Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen and a decrease in claims filed per patient.
 
 The implications of these results are threefold.
-First, a surge of experienced physicians left the labor force in a concentrated number of years due to EHR exposure, potentially worsening access to care.
-In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss may have long term negative effects on patient outcomes, but this is a question left to future research.
-Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid burdens externally placed on them.
-This has implications when considering the current policy movement towards regulated care, that physicians may go to extreme measures to avoid such regulation.
+First, a surge of physicians left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening access to care.
+In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research.
+Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens.
+When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation.
 Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users.
@@
 Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital.
-I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.
-The final list of pairs consists of 1.3 million observations.
+I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60.
+Results are similar and can be found in Figure ____.} The final list of pairs consists of 1.3 million observations.
 
@@
 I combine the physician-hospital pairs created in Section \ref{sec:sharedpat} with various data sets for hospital or physician level information.
-First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation year and gender.
+First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation.
 This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation.
@@
 
-I merge the remaining pairs to an additional survey, the AHA IT Supplement, only sent to hospitals who reported using an EHR.
-There are more detailed questions in this survey including the EHR vendor, and the ranking of documentation and decision making features the hospital uses.
-There is a lot of missing data in this survey, which is why I do not use it in my main analysis.
-
 I use the general survey question about each hospital's EHR use to define my key treatment variable.
@@
 Finally, I aggregate to the physician level by only keeping variables that do not vary by hospital.
-The variables in the physician level data are year, physician NPI, graduation year, average number of beds in hospitals worked with, years of experience, number of hospitals, fraction of hospitals with EHR, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, patient count, patient count in EHR hospitals, patient count in non-EHR hospitals, and never works with a new NPI.
+The variables in the physician level data are year, physician NPI, graduation year, average number of beds in hospitals worked with, years of experience, number of hospitals, fraction of hospitals with EHR, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, and never works with a new NPI.
 I complete this data to include years 2016 and 2017, but leave time varying variables missing for those years.
@@
 
-Using physician NPI, I merge the physician treatment data to Medicare Data on Provider Practice and Specialty (MD-PPAS)\footnote{Information on data located at \hyperlink{https://resdac.org/cms-data/files/md-ppas}{https://resdac.org/cms-data/files/md-ppas}}, which psnas 2009-2017.
+Using physician NPI, I merge the physician treatment data to Medicare Data on Provider Practice and Specialty (MD-PPAS)\footnote{Information on data located at \hyperlink{https://resdac.org/cms-data/files/md-ppas}{https://resdac.org/cms-data/files/md-ppas}}, which spans 2009-2017.
 This data contains variables on physician specialty, Medicare claim counts in various zip codes, unique number of patients seen, fraction of patients seen in specific settings, patient demographics, and physician date of birth.
-Once the data is merged, I make a further limitation to drop any physicians with less than 20\% of their total patients in a hospital setting.
-This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
-With this limitation, there are 403,000 observations left in the data.
+Once the data is merged, I make a further limitation to drop any physicians with less than 70\% of their total patients in a hospital setting.\footnote{I consider different thresholds as a sensitivity analysis.
+The results are similar, and can be found in Section ____.} This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
+With this limitation, there are 214,000 observations left in the data.
 
@@
 Finally, the variables for unique patient count and claim count are already defined.
-This is the final data used in the paper, with 403,020 observations.
+This is the final data used in the paper, with approximately 215,000 observations.
 
@@
 
-Recent work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference in difference designs.
-In my main specification, I assume that, conditional on covariates, average outcomes for those treated in group g would have followed a parallel trend as those in groups treated in later periods.
+Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference in difference designs.
+In my main specification I assume that, conditional on covariates, average outcomes for those treated in group g would have followed a parallel trend as those in groups treated in later periods.
 I assess this assumption by testing whether the coefficients showing the effect of EHR exposure on an outcome prior to actual treatment are jointly zero.
 All main specification graphs show p-values from this test.
-In some outcomes, I fail to reject the null that there does not exist a pre-trend.
+In most analyses, I fail to reject the null that there does not exist a pre-trend.
 Yet, in some cases, I reject this null.
-This section utilizes the tools established by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) to both demonstrate robustness of the results found in the main specification, and to improve our understanding in the cases where a pre-trend exists.
+This section utilizes the tools established by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) to demonstrate robustness of the results found in the main specification, and to investigate whether an effect exists in the cases where a pre-trend may be driving results.
 
@@
 
-First, I examine the results with outcome variable retirement.
+First, I consider the estimated effect of EHR exposure on retirement.
 While visually there does not appear to be a strong pre-trend, for the full sample of physicians and physicians $< 60$, the p-value indicates a pre-trend may exist.
 Since the positive effect seems to be a possibility in event time 1 and event time 2, I present a plot of robust confidence intervals under various possible assumptions of parallel trends violations for these periods in Figure \ref{fig:pre_retire}.
-In particular, the original confidence interval is shown in yellow, and in gray are confidence intervals robust to varying functions of parallel trend violations.
+The original confidence interval is shown in yellow, and in gray are confidence intervals robust to varying functions of parallel trend violations.
 When allowing for linear trend violations (M=0), the effect of EHR exposure on retirement is still positive.

### 2022-09-10 Paper/Paper.tex

@@
 %\captionsetup{justification=raggedright,singlelinecheck=false}
+\usepackage{subcaption}
 
@@
 Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not.
-Hospitals with this discretion arre coded as implementing an EHR in the first year that they answered "fully" to the survey question.} Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
+Hospitals with this discretion are coded as implementing an EHR in the first year that they answered "fully" to the survey question.} Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
 Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
@@
 
-For the results presented, if EHR exposure affects physician behavior, the effect occurs in the first 0-2 years after exposure.
-Almost all specifications point to the effect of EHR exposure becmoing null in the 3-4 years after exposure.
+For the results presented, if EHR exposure affects physician behavior, the effect typically occurs in the first 0-2 years after exposure.
+Almost all specifications point to the effect of EHR exposure becoming null in the 3-4 years after exposure.
 One reason for this may be the composition of the hospitals which adopt at different times.
@@
 For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=.5]{Objects/control_histogram.pdf}
+    \caption{Caption}
+    \label{fig:my_label}
+\end{figure}
 
@@
 I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60.
-Results are similar and can be found in Figure ____.} The final list of pairs consists of 1.3 million observations.
+Results are similar and can be found in Figure \ref{sec:chart}.} The final list of pairs consists of 1.3 million observations.
 
@@
 Once the data is merged, I make a further limitation to drop any physicians with less than 70\% of their total patients in a hospital setting.\footnote{I consider different thresholds as a sensitivity analysis.
-The results are similar, and can be found in Section ____.} This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
+The results are similar, and can be found in Section \ref{sec:chart}} This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
 With this limitation, there are 214,000 observations left in the data.
@@
 
-\subsection{Physician Selection into EHRs}\label{sec:endog}
+\subsection{Various Changes to Specification}\label{sec:changes}
+
+For each outcome, there are various adjustments I make to the main specification in order to show robustness.
+Some of these changes address potential endogeneity concerns, while some provide evidence that the results do not depend on thresholds which may be arbitrary.
+I will outline each of these specification adjustments and then present specification charts which account for different combinations of these changes.
+
+\subsubsection{Physician Selection into EHRs}\label{sec:endog}
 
 The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in.
-These results are biased if physicians endogenously select EHR exposure, particularly by participating in the hospital's decision to implement.
+These results are biased if physicians endogenously select whether they are exposed to EHRs, particularly by participating in the hospital's decision to implement.
 The results do not indicate that this is a huge issue on average, since we would expect physicians not to respond on the margin of retirement or switching work setting if they are choosing to be exposed to EHRs.
 Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions.
-However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for a specification with an alternate binary treatment variable: whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
+However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for specifications with an alternate binary treatment variable: whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
 
@@
 
-I present a comparison of the average treatment effect in the main specification and the average treatment effect with low-integration exposure as the binary treatment variable in Figure \ref{fig:LI}.
-For conciseness, I present group time treatment effects to a single ATT value.
-As expected, the coefficient on retirement is nearly identical to the main specification.
-However, the coefficient for the probability of working in office based settings completely switches signs, pointing to questions about acquisition of physician practices.
-In low-integration settings, physicians are less likely to shift to an office after EHR exposure by a small measure.
-However, when including physicians with more power, the do seem to start office based work.
-If high-integration hospitals are acquiring smaller practices as EHRs are implemented, physicians from the hospital may be sent to the newly acquired practice for a period of time during transition.
-This is an important result because it points to physicians shifting according to hospital incentives instead of as a method to avoid EHR implementation.
-Estimates for other coefficients are more noisy, but similar to the main specification.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.57\linewidth}
-    \caption{Comparison of ATT, Low-Integration Treatment Variable}
-    \includegraphics[scale=.57]{Objects/LI_graph.pdf}
-    \label{fig:LI}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects are shown for each outcome variable, along with 95\% confidence intervals.
-Low-integration hospitals are defined by hospital management.}}
-\end{figure}
-
-\subsection{Anticipation}\label{app:anticipation}
+\subsubsection{Anticipation}\label{app:anticipation}
 
 While there is evidence to suggest the time it takes to implement an EHR does not exceed one year, there still may be some anticipation if physicians are expecting implementation in the hospitals they work in.
-In Figure \ref{fig:anticipation}, I present aggregated estimates in the case of one year of anticipation.
-The results are virtually identical, just noisier, compared to the estimates from the main specification.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.57\linewidth}
-    \caption{ATT With Anticipation}
-    \includegraphics[scale=.57]{Objects/anticipation_graph.pdf}
-    \label{fig:anticipation}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects are shown for each outcome variable, along with 95\% confidence intervals.}}
+Thus, the specification charts that I present in Section \ref{sec:chart} include specifications where one year of anticipation occurs.
+
+\subsubsection{Limiting Sample Years}\label{app:years}
+
+In the main sample, I define dependent variables using years 2009 to 2017 and drop any physicians who were not exposed to an EHR by 2015 due to data constraints.
+This leads to estimation based on not yet treated units, because all never treated units are dropped.
+As a robustness check, I limit sample years to 2009-2015 and use never treated units as the control group.
+
+\subsubsection{Hiring Data Assistants}
+
+As discussed in the body of the paper, I address potential concerns that the hiring of new employees may be driving the results for productivity and efficiency.
+Thus, as another specification, I restrict the data to only include physicians who are exposed in hospitals that do not record employing any data assistants.
+
+\subsubsection{Data Thresholds}
+
+There are two main thresholds I define in the data process that were arbitrarily chosen.
+Thus, I present alternate specifications that adjust these thresholds.
+First, when deciding which hospitals a physician is working closely with, I drop any pairs that fall below 30 shared patients per year.
+The goal of this threshold is to eliminate pairs for which the physician does not have close ties to the hospital.
+In alternate specifications, I lower the threshold to 10 patients per year and raise the threshold to 60 patients per year.
+Second, I only want to include hospitalists in the analysis, and thus drop any physicians that fall below 70\% of patients in an inpatient setting.
+In alternate specifications, I change this threshold to 50\% and 90\%.
+While these are still arbitrary, they provide some evidence that changing the threshold does not dramatically change the results.
+
+\subsection{Specification Charts}\label{sec:chart}
+
+I present overall average treatment effects for each outcome variable under different combinations of the specification adjustments discussed in Section \ref{sec:changes}.
+On each chart, the main specification is labeled in green and the average confidence bands across all combinations is shown as an orange band.
+
+\subsubsection{Retirement}
+
+The specification chart for the retirement outcome is presented in Figure \ref{fig:retire_chart}.
+The main specification yields an average treatment effect of .0025 (significant at all levels).
+This coefficient is on the higher end of coefficients from all combinations of specifications.
+However, almost all of the coefficients are positive.
+Some noisy estimates are to be expected when there are multiple limitations placed on the data.
+The average confidence interval for all combinations is (.001,.003), so the main specification lies in this interval.
+These results are strongly in favor of there being a positive effect of EHR exposure on retirement.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Retire}
+    \includegraphics[scale=.6]{Objects/retire_chart.pdf}
+    \label{fig:retire_chart}
+\end{figure}
+
+\subsubsection{Change Work Setting}
+
+The specification charts for outcomes work in office, fraction of patients seen in office, and changing zip codes are presented in Figures \ref{fig:work_chart}, \ref{fig:fracoffice_chart}, and \ref{fig:zip_chart}, respectively.
+First, I discuss the overall average treatment effect of EHR exposure on the likelihood of working in an office.
+While the main specification showed a slight increase in likelihood in the year of exposure, the overall treatment effect is 0.
+Most of the specifications agree with the main specification, although more coefficients fall below 0 than above 0.
+The average confidence interval is approximately (-.016, .001), encompassing the main specification.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Work in Office}
+    \includegraphics[scale=.6]{Objects/office_chart.pdf}
+    \label{fig:work_chart}
+\end{figure}
+
+Second, I discuss the results for whether physicians change the fraction of patients seen in an office setting due to EHR exposure. The main specification
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Fraction of Patients in Office}
+    \includegraphics[scale=.6]{Objects/office_frac_chart.pdf}
+    \label{fig:fracoffice_chart}
+\end{figure}
+
+Finally, I discuss the specification chart regarding a physician's likelihood of changing zip codes due to EHR exposure.
+These results should be taken in conjunction with Section \ref{sec:pretrends}, which analyzes the extent to which this result is driven by a pre-trend.
+The main specification yields an average treatment effect of .04 (significant at all levels), which is in the average confidence interval across specifications.
+One thing to notice in this graph is that there are two changes to specification for which the resulting average treatment effect is grouped towards one end of the distribution or the other.
+Specifically, when the percent of patients seen in an inpatient setting is 90\%, the coefficient is likely to be higher than when it is 50\% or 70\%.
+That is, a more strict definition of hospitalist yields a higher estimate for the likelihood of changing zip codes (although the variance of estimates is not particularly high).
+Further, using the main EHR definition tends to yield lower estimates than the low integration definition.
+That is, endogeneity is biasing this estimate slightly down.
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Change Zip Codes}
+    \includegraphics[scale=.6]{Objects/change_zip_chart.pdf}
+    \label{fig:zip_chart}
+\end{figure}
+
+\subsubsection{Patient Count and Claims per Patient}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Patient Count}
+    \includegraphics[scale=.6]{Objects/patientcount_chart.pdf}
+    \label{fig:pat_chart}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \caption{Outcome: Claims per Patient}
+    \includegraphics[scale=.6]{Objects/claim_per_patient_chart.pdf}
+    \label{fig:cpp_chart}
 \end{figure}
@@
 
-\subsection{Limiting Sample Years}\label{app:years}
-
-I create the retirement outcome variable as a direct function of future claims.
-Those who retire in the early years of data have zero future claims for multiple years.
-However, those who retire after 2016 have only one or two years of zero future claims to be flagged as retired.
-While some of these physicians are likely retiring, it may be that a fraction of them only drop out of the data temporarily due to pregnancy or other factors.
-Thus, I limit the years in the analysis to 2009-2015 to ensure that those years are not driving the observed results.
-Results for this specification are shown in Figure \ref{fig:years}.
-The observed estimators are virtually identical to those in the main specification.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.57\linewidth}
-    \caption{Limited Sample Years}
-    \includegraphics[scale=.57]{Objects/years_graph.pdf}
-    \label{fig:years}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and a limited sample of years.}}
-\end{figure}
-
-\subsection{Physicians not Exposed to Data Assistants}\label{app:DA}
-
-In this section, I present results for the main specification with a data limitation: only including physicians who are exposed in hospitals that do not record employing any data assistants.
-These physicians may be more likely to respond negatively to EHRs and may be driving results in the main specification.
-I present these results in Figure \ref{fig:DA}.
-These results are surprisingly similar to those for exposure in low integration hospitals discussed in Appendix \ref{sec:endog}.
-Many outcomes have similar coefficients for the effect of EHR exposure, with the exception of probability of working in office settings.
-Those who are not exposed to data assistants are less likely to begin working in an office, an opposite sign from the specification in the main analysis.
-The level of integration could be correlated with hiring data assistants, which both may be correlated with the acquisition of smaller physician practices which are considered office based settings.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.57\linewidth}
-    \caption{Physicians without Data Assistants}
-    \includegraphics[scale=.57]{Objects/DA_graph.pdf}
-    \label{fig:DA}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and limited sample of physicians who do not work with data assistants.}}
-\end{figure}
-
 \section{Heterogeneity Analysis}
 
-While the majority of the analysis focuses on senior vs.
-young physicians, there are other physician characteristics that are also interesting to investigate.
-Figure \ref{fig:het} presents estimates for different subsets of physicians.
-These include female physicians,those in rural or urban zip codes, and physicians who see majority white patients relative to the entire sample.
-Most of the estimates are similar across these characteristics, with some particularly noisy due to low sample sizes (such as physicians in rural settings).
-The implications implication of this table is that EHRs did in fact lead to behavioral changes in many different physicians, not just a specific subgroup.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.65\linewidth}
-    \caption{Heterogeneity Analysis}
-    \includegraphics[scale=.65]{Objects/heterogeneity _graph.pdf}
-    \label{fig:het}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Aggregated average treatment effects and 95\% confidence intervals shown for main specification and various limited samples of physicians.}}
+The main analysis focuses on the differential effects of technology for older vs.
+younger physicians.
+However, another important question is whether physicians in rural vs.
+urban settings respond differently to the technology since the implications for behavior changes are high in areas that already suffer from access to care issues.
+Thus, I present average treatment effects for each outcome where the sample is split between physicians in rural and urban areas.
+Further, to investigate the same idea, I present results for physicians who typically work at small hospitals vs.
+those who typically work at large hospitals.
+These average treatment effects are shown in Figure \ref{fig:1}.
+For most outcomes, the results do not differ along these dimensions.
+The only outcome for which the estimate changes sign is the indicator for whether the physician works in an office setting.
+Physicians in an urban area are less likely to work in an office after EHR exposure, while rural physicians are more likely to work in an office after EHR exposure, but the sign switch does not hold when looking at the hospital size parallel.
+This results is interesting and may be driven by differing physician-hospital relationships.
+I leave further investigation to future research.
+
+\begin{figure}[t!]
+\begin{subfigure}{0.48\textwidth}
+\caption{Retire}
+\includegraphics[width=\linewidth]{Objects/retire_heterog.pdf}
+\label{fig:a}
+\end{subfigure}\hspace*{\fill}
+\begin{subfigure}{0.48\textwidth}
+\caption{Fraction of Patients in Office}
+\includegraphics[width=\linewidth]{Objects/frac_office_heterog.pdf}
+\label{fig:b}
+\end{subfigure}
+
+\medskip
+\begin{subfigure}{0.48\textwidth}
+\caption{Indicator for Office }
+\includegraphics[width=\linewidth]{Objects/office_heterog.pdf}
+ \label{fig:c}
+\end{subfigure}\hspace*{\fill}
+\begin{subfigure}{0.48\textwidth}
+\caption{Change Zip Code} 
+\includegraphics[width=\linewidth]{Objects/change_zip_heterog.pdf}
+\label{fig:d}
+\end{subfigure}
+
+\medskip
+\begin{subfigure}{0.48\textwidth}
+\caption{Patient Count} 
+\includegraphics[width=\linewidth]{Objects/patient_heterog.pdf}
+\label{fig:e}
+\end{subfigure}\hspace*{\fill}
+\begin{subfigure}{0.48\textwidth}
+\caption{Claim per Patient}
+\includegraphics[width=\linewidth]{Objects/claim_per_patient_heterog.pdf}
+ \label{fig:f}
+\end{subfigure}
+
+\caption{Heterogeneity Analysis} \label{fig:1}
 \end{figure}

### 2022-09-10 Table Code/overall stats.tex

--- 2022-08-04 Table Code/overall stats.tex
@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 179,883 & 0.03 & 0.18 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 179,883 & 0.10 & 0.24 & 0 & 1\\
-\hspace{1em}Work in an Office & 179,883 & 0.30 & 0.46 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 161,558 & 0.12 & 0.33 & 0 & 1\\
-\hspace{1em}Number of Patients & 179,883 & 330 & 274 & 0 & 7,613\\
-\hspace{1em}Claim Count & 179,883 & 1,382 & 1,789 & 0 & 64,508\\
+\hspace{1em}Ever Retire & 162018 & 0.03 & 0.17 & 0 & 1.00\\
+\hspace{1em}Fraction Patients in Office & 162018 & 0.08 & 0.21 & 0 & 1.00\\
+\hspace{1em}Work in an Office & 162018 & 0.27 & 0.44 & 0 & 1.00\\
+\hspace{1em}Change Zip Codes & 144787 & 0.12 & 0.32 & 0 & 1.00\\
+\hspace{1em}Number of Patients & 162018 & 330.92 & 273.96 & 0 & 7613.00\\
+\hspace{1em}Claims per Patient & 146443 & 4.09 & 4.09 & 0 & 480.68\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 179,883 & 0.75 & 0.43 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 179,883 & 0.63 & 0.44 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 162018 & 0.76 & 0.43 & 0 & 1.00\\
+\hspace{1em}Fraction of Hospitals with EHR & 162018 & 0.64 & 0.44 & 0 & 1.00\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 179,883 & 44.9 & 8.4 & 25 & 86\\
-\hspace{1em}Female & 179,883 & 0.34 & 0.47 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 179,883 & 1.49 & 0.82 & 1 & 12\\
-\hspace{1em}Number of Systems Worked With & 179,883 & 1.31 & 0.60 & 1 & 8\\
+\hspace{1em}Age & 162018 & 44.55 & 8.31 & 25 & 86.00\\
+\hspace{1em}Female & 162018 & 0.34 & 0.47 & 0 & 1.00\\
+\hspace{1em}Number of Hospitals Worked With & 162018 & 1.49 & 0.82 & 1 & 12.00\\
+\hspace{1em}Number of Systems Worked With & 162018 & 1.31 & 0.61 & 1 & 8.00\\
 \bottomrule
 \end{tabular}
+\end{table}
 \label{tab:sumstats}
-\end{table}

### 2022-09-11 Scripts/analysis.R
update pretrends plots

@@ -13,9 +13,6 @@ library(extrafont)
 #                                       Hanna Glenn, Emory University
 #                                       1/31/2022
 
-font_import()
-loadfonts(device="win")
-windowsFonts(A = windowsFont("Times New Roman"))
 
 # This script reads in "Physician_Data.rds" from data5, the final dataset used in my third year paper. 
 # The first portion of the script considers different potential estimators to use as the main specification for the paper.

### 2022-09-11 Scripts/sensitivity_analysis.R

@@ -261,7 +261,7 @@ honest_did <- function(es, ...) {
 }
 
 honest_did.AGGTEobj <- function(es,
-                                e=0,
+                                e=2,
                                 type=c("smoothness", "relative_magnitude"),
                                 method=NULL,
                                 bound="deviation from parallel trends",
@@ -344,225 +344,79 @@ honest_did.AGGTEobj <- function(es,
 
 smooth_year1 <- lapply(Models_agg, function(x){
   all <- honest_did(x[["agg_all"]], type="smoothness")
-  young <- honest_did(x[["agg_young"]], type="smoothness")
-  old <- honest_did(x[["agg_old"]], type="smoothness")
   
-  list(all=all,young=young,old=old)
+  list(all=all)
 })
 
 plot_smooth_year1 <- lapply(smooth_year1, function(x){
   all <- createSensitivityPlot(x$all$robust_ci, x$all$orig_ci)
-  young <- createSensitivityPlot(x$young$robust_ci, x$young$orig_ci)
-  old <- createSensitivityPlot(x$old$robust_ci, x$old$orig_ci)
   
-  list(all,young,old)
+  list(all)
 })
 
 smooth_year2 <- lapply(Models_agg, function(x){
   all <- honest_did(x[["agg_all"]], type="smoothness")
-  young <- honest_did(x[["agg_young"]], type="smoothness")
-  old <- honest_did(x[["agg_old"]], type="smoothness")
   
-  list(all=all,young=young,old=old)
+  list(all=all)
 })
 
 plot_smooth_year2 <- lapply(smooth_year2, function(x){
   all <- createSensitivityPlot(x$all$robust_ci, x$all$orig_ci)
-  young <- createSensitivityPlot(x$young$robust_ci, x$young$orig_ci)
-  old <- createSensitivityPlot(x$old$robust_ci, x$old$orig_ci)
   
-  list(all,young,old)
-})
-
-smooth_year0 <- lapply(Models_agg, function(x){
-  all <- honest_did(x[["agg_all"]], type="smoothness")
-  young <- honest_did(x[["agg_young"]], type="smoothness")
-  old <- honest_did(x[["agg_old"]], type="smoot whness")
-  
-  list(all=all,young=young,old=old)
-})
-
[…]

### 2022-09-11 Table Code/split stats.tex

@@
 
-\caption{Means by Age Sample}
+\caption{Means by Age}
 \centering
@@
 \hspace{1em}Ever Retire & 0.04 & 0.03\\
-\hspace{1em}Fraction Patients in Office & 0.15 & 0.09\\
-\hspace{1em}Work in an Office & 0.48 & 0.28\\
+\hspace{1em}Fraction Patients in Office & 0.12 & 0.07\\
+\hspace{1em}Work in an Office & 0.44 & 0.25\\
 \hspace{1em}Change Zip Codes & 0.10 & 0.12\\
-\hspace{1em}Number of Patients & 342 & 329\\
-\hspace{1em}Claim Count & 1,767 & 1,332\\
+\hspace{1em}Number of Patients & 340.18 & 329.77\\
+\hspace{1em}Claim Count & 1692.44 & 1295.58\\
 \addlinespace[0.3em]
 \multicolumn{3}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 0.76 & 0.75\\
-\hspace{1em}Fraction of Hospitals with EHR & 0.59 & 0.64\\
+\hspace{1em}Exposure to an EHR & 0.78 & 0.76\\
+\hspace{1em}Fraction of Hospitals with EHR & 0.60 & 0.65\\
 \addlinespace[0.3em]
 \multicolumn{3}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 60.31 & 42.75\\
+\hspace{1em}Age & 60.23 & 42.59\\
 \hspace{1em}Female & 0.18 & 0.36\\
-\hspace{1em}Number of Hospitals Worked With & 1.63 & 1.48\\
-\hspace{1em}Number of Systems Worked With & 1.38 & 1.30\\
+\hspace{1em}Number of Hospitals Worked With & 1.63 & 1.47\\
+\hspace{1em}Number of Systems Worked With & 1.39 & 1.30\\
 \bottomrule

### 2022-09-12 Scripts/analysis.R
add data assistant specification to patient count plot

@@ -21,12 +21,15 @@ library(extrafont)
 # Read in the main data created in "data5_MDPPAS.R"
 Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
+Physician_Data <- Physician_Data %>%
+  mutate(npi_unq_benes_noDA=npi_unq_benes) 
+
 
 
 ## ANALYSIS ---------------------------------------------- #########
 ## Write results for each variable in a loop and create graphs ----- ##
 
-varlist <- list("retire", "pos_office", "work_in_office", "change_zip", "npi_unq_benes", "claim_per_patient")
+varlist <- list("retire", "pos_office", "work_in_office", "change_zip", "npi_unq_benes","npi_unq_benes_noDA", "claim_per_patient")
 
 
 
@@ -38,7 +41,7 @@ models <- lapply(varlist, function(x) {
     tname = "year",                  # Time Variable
     # xformla = NULL                 # No covariates
     xformla = ~grad_year,            # Time-invariant controls
-    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="pos_office_prior" | x=="pos_office_noprior" | x=="work_in_office" | x=="change_zip" | x=="pos_opd") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
+    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="pos_office_prior" | x=="pos_office_noprior" | x=="work_in_office" | x=="change_zip" | x=="pos_opd") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else if (x=="npi_unq_benes_noDA") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1 & works_with_DA==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
     est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
     control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
     clustervars = "DocNPI",          # Cluster Variables          
[…]

### 2022-09-12 Scripts/data4_physician_level.R

@@ -179,10 +179,10 @@ data <- data %>%
 VA <- data %>%
   mutate(VA=ifelse(str_detect(Hospital_name,"Veteran"),1,0)) %>%
   filter(VA==1) %>%
-  distinct(ID,VA)
+  distinct(pairID,VA)
 
 data <- data %>%
-  left_join(VA,by="ID") %>%
+  left_join(VA,by="pairID") %>%
   filter(is.na(VA)) %>%
   select(-VA)
 
@@ -492,6 +492,7 @@ data <- data %>%
   mutate(works_with_DA=ifelse(sum>0,1,0)) %>%
   ungroup()
 
+
 # Aggregate the data to the physician level -------------------------------------------------------------
 
 Aggregated_Pairs <- data %>%
@@ -515,6 +516,7 @@ Aggregated_Pairs <- Aggregated_Pairs %>%
   fill(avg_beds,.direction="downup") %>%
   fill(num_hospitals,.direction="downup") %>%
   fill(num_systems,.direction="downup") %>%
+  fill(works_with_DA,.direction="downup") %>%
   ungroup() %>%
   mutate(experience=year-grad_year) %>%
   mutate(anyEHR_exposed=ifelse(minyr_EHR>0 & year>=minyr_EHR,1,0),

### 2022-09-12 Scripts/data5_MDPPAS.R

@@ -293,11 +293,6 @@ Physician_Data <- Physician_Data %>%
 # The key for this variable is to limit the sample appropriately. 
 
 Physician_Data <- Physician_Data %>%
-  mutate(hosp_patient_count_EHRhosp=ifelse(is.na(hosp_patient_count_EHRhosp),0,hosp_patient_count_EHRhosp),
-         hosp_patient_count_noEHRhosp=ifelse(is.na(hosp_patient_count_noEHRhosp),0,hosp_patient_count_noEHRhosp),
-         hosp_patient_count=ifelse(is.na(hosp_patient_count),0,hosp_patient_count))
-
-Physician_Data <- Physician_Data %>%
   mutate(npi_unq_benes=ifelse(is.na(npi_unq_benes),0,npi_unq_benes))
 
 Physician_Data <- Physician_Data %>%

### 2022-09-12 Paper/Paper.tex

@@
     \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
-    \includegraphics[scale=.6]{Objects/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
+    \includegraphics[scale=.6]{graphics/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
     \caption*{Source: HealthIT.gov}
@@
     \caption{Screenshot of EHR System}
-    \includegraphics[scale=.4]{Objects/epic-ehr-screenshot.jpg}
-    \includegraphics[scale=.11]{Objects/EHRimage2.jpg}
+    \includegraphics[scale=.4]{graphics/epic-ehr-screenshot.jpg}
+    \includegraphics[scale=.11]{graphics/EHRimage2.jpg}
     \label{fig:EPIC}
@@
     \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
-    \includegraphics[scale=.6]{Objects/officefrac__plot.pdf}
+    \includegraphics[scale=.6]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
 Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference in difference designs.
-In my main specification I assume that, conditional on covariates, average outcomes for those treated in group g would have followed a parallel trend as those in groups treated in later periods.
+In my main specification I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
 I assess this assumption by testing whether the coefficients showing the effect of EHR exposure on an outcome prior to actual treatment are jointly zero.
@@
 \caption{Fraction of Patients in Office}
-\includegraphics[width=\linewidth]{Objects/frac_office_heterog.pdf}
+\includegraphics[width=\linewidth]{Objects/office_frac_heterog.pdf}
 \label{fig:b}
@@
 \caption{Indicator for Office }
-\includegraphics[width=\linewidth]{Objects/office_heterog.pdf}
+\includegraphics[width=\linewidth]{Objects/office_ind_heterog.pdf}
  \label{fig:c}

### 2022-09-12 Table Code/overall stats.tex

--- 2022-08-04 Table Code/overall stats.tex
@@
 \multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 179,883 & 0.03 & 0.18 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 179,883 & 0.10 & 0.24 & 0 & 1\\
-\hspace{1em}Work in an Office & 179,883 & 0.30 & 0.46 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 161,558 & 0.12 & 0.33 & 0 & 1\\
-\hspace{1em}Number of Patients & 179,883 & 330 & 274 & 0 & 7,613\\
-\hspace{1em}Claim Count & 179,883 & 1,382 & 1,789 & 0 & 64,508\\
+\hspace{1em}Ever Retire & 162,018 & 0.03 & 0.17 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 162,018 & 0.08 & 0.21 & 0 & 1\\
+\hspace{1em}Work in an Office & 162,018 & 0.27 & 0.44 & 0 & 1\\
+\hspace{1em}Change Zip Codes & 144,787 & 0.12 & 0.32 & 0 & 1\\
+\hspace{1em}Number of Patients & 162,018 & 331 & 274 & 0 & 7,613\\
+\hspace{1em}Claims per Patient & 146,443 & 4.09 & 4.1 & 0 & 481\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 179,883 & 0.75 & 0.43 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 179,883 & 0.63 & 0.44 & 0 & 1\\
+\hspace{1em}Exposure to an EHR & 162,018 & 0.76 & 0.43 & 0 & 1\\
+\hspace{1em}Fraction of Hospitals with EHR & 162,018 & 0.64 & 0.44 & 0 & 1\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 179,883 & 44.9 & 8.4 & 25 & 86\\
-\hspace{1em}Female & 179,883 & 0.34 & 0.47 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 179,883 & 1.49 & 0.82 & 1 & 12\\
-\hspace{1em}Number of Systems Worked With & 179,883 & 1.31 & 0.60 & 1 & 8\\
+\hspace{1em}Age & 162,018 & 44.55 & 8.31 & 25 & 86\\
+\hspace{1em}Female & 162,018 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 162,018 & 1.49 & 0.82 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 162,018 & 1.31 & 0.61 & 1 & 8\\
 \bottomrule
 \end{tabular}
+\end{table}
 \label{tab:sumstats}
-\end{table}

### 2022-10-05 Scripts/analysis.R
update format of estimator plots

@@ -26,17 +26,11 @@ Physician_Data <- Physician_Data %>%
   mutate(npi_unq_benes_LI=npi_unq_benes) %>%
   mutate(DocNPI=as.double(DocNPI))
 
-phys <- Physician_Data %>%
-  ungroup() %>%
-  distinct(DocNPI)
-
-
-
 
 ## ANALYSIS ---------------------------------------------- #########
 ## Write results for each variable in a loop and create graphs ----- ##
 
-varlist <- list("retire", "pos_office","pos_office_prior","pos_office_noprior", "work_in_office", "change_zip", "npi_unq_benes", "claim_per_patient")
+varlist <- list("retire", "pos_office", "work_in_office", "npi_unq_benes", "claim_per_patient")
 
 
 

### 2022-10-05 Scripts/sensitivity_analysis.R

@@ -140,32 +140,33 @@ source(paste0(function_path,"spec_chart_function.R"))
 labels <- list("Stacked Reg.", "Two-Stage DiD", "Main (CS)")
 
 schart(est_retire_data, 
-       order="asis", labels,
+       order="asis", labels = labels,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
-       ylab="Coefficient and 95% C.I.\n") 
+       ylab="Coefficient and 95% C.I.\n",
+       highlight=3) 
 schart(est_frac_office_data, 
-       order="asis", labels,
+       order="asis", labels, highlight=3,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
        ylab="Coefficient and 95% C.I.\n") 
 schart(est_indoffice_data, 
-       order="asis", labels,
+       order="asis", labels, highlight=3,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
        ylab="Coefficient and 95% C.I.\n") 
 schart(est_zip_data, 
-       order="asis", labels,
+       order="asis", labels, highlight=3,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
        ylab="Coefficient and 95% C.I.\n") 
 schart(est_patient_data, 
-       order="asis", labels,
+       order="asis", labels, highlight=3,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
        ylab="Coefficient and 95% C.I.\n") 
 schart(est_claim_data, 
-       order="asis", labels,
+       order="asis", labels, highlight=3,
        col.est=c("grey70","#E69F00"),
        col.dot=c("grey70", "grey90", "#E69F00", "#E69F00"),
        ylab="Coefficient and 95% C.I.\n") 
@@ -373,8 +374,14 @@ retire_all_yr1 <- plot_smooth_year1[[1]][[1]] +
   scale_color_manual(labels = c("FLCI", "Original"), values=c("#999999", "#E69F00"),name="") +
   theme(legend.position = "none", text = element_text(size = 12)) 
 
+retire_all_yr2 <- plot_smooth_year2[[1]][[1]] + 
[…]

### 2022-10-05 Paper/Paper.tex
Updates from Overleaf

@@
 \begin{abstract}
-    Technological innovation has affected workers and labor markets for decades, yet the effects of modern technologies in some vocations with unique incentives are still not understood.
+    Technological innovation has affected workers and labor markets for decades.
 I add to the understanding of technology implementation's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
@@
 I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes.
-I find that physicians are 10\% more likely to retire, and 4\% more likely to work in an office if they choose not to retire.
-Finally, for physicians who remain exposed to the technology, the number of patients seen increases.
-However, claims per patient decreases.
-These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential efficiency gains for those who do not respond adversely.
+I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure.
+Further, for physicians who remain exposed to the technology, the number of patients seen increases.
+These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential productivity gains for those who do not avoid the technology through labor market behavioral changes.
 \end{abstract}
@@
 How technological innovation affects labor markets is heavily studied, in part because of the vast differences in technology and tasks present even within industries.
-Yet, there is still much to be learned about modern technology and its interaction with workers.
+Yet, there is still much to be learned about modern technology and its interaction with how workers behave.
 A recent technology roll-out that altered the U.S.
 health care system dramatically was the electronic health record (EHR).
-The implementation and use of EHRs were incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
-Since many physicians working in hospitals were exposed to this technology beyond their control, this is an opportune setting to investigate the relationship between rapid technology implementation and labor market decisions for highly skilled workers.
-
-Specifically, EHRs are computerized medical records stored in advanced systems that may include advanced capabilities such as decision making assistance.
+The use of EHRs were incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
+Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and labor market behaviors.
+
+Specifically, EHRs are computer systems that house digital medical records and have additional capabilities, including decision making assistance.
 EHRs have become increasingly relevant in the U.S.
-since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”.
-According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 For all outcomes, the independent variable of interest is a binary treatment variable capturing exposure to an EHR.
-I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing billable activity in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) patient and billing activity.
+I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.
 
@@
 Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered "meaningful" by the government's standards, and are likely affecting physicians' daily life.
+This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered ``meaningful" by the government's standards, and are likely affecting physicians' daily life.
 Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects, and further provides insight to short vs.
@@
 These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians.
-For those who did not retire, I find that young physicians were .014 ppts (4.6\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure.
-Finally, I find a modest, positive effect on patient count completely driven by younger physicians, and a decreasing trend in claims per patient.
+For those who did not retire, I find that young physicians were .06 ppts (15\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure.
+I find that patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.
 
@@
 Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation.
-As technological innovation made computers more portable, the usability of EHRs increased, creating what is known as the "physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
+As technological innovation made computers more portable, the usability of EHRs increased, creating what is known as the ``physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
 Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
@@
 Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
-Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), suggesting a negative impact.
+Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
 
@@
 There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management.
-While it is unlikely that physicians under some age threshold are formally retiring, physicians of any age could be career switching.
+While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching.
 This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction of current clinical work.
-As I discuss in the Data Section, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
+As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
 Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
@@
 If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients.
-In this case, I expect productivity to increase.
+In this case, I productivity would increase.
 However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
@@
 
-\section{Data}
+\section{Data}\label{sec:data}
 
@@
 Next, I consider whether physicians exhibit behavior that suggests they have shifted setting or location of work.
-I measure this in the data in three ways.
 First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
@@
 This binary measure of practice setting offers less variation but captures a more discrete change in which physicians enter or leave the office-based setting entirely.
-Third, based on the physician's practice zip code, I form an indicator variable equal to one if the physician's list of zip codes changes from one year to the next.
-While no single variable directly measures physician movement across hospitals, these three outcomes collectively provide insight into whether physicians desire to shift away from or towards EHRs.
 Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and claims filed per patient.
@@
 I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}.
-These results yield similar patterns with slightly smaller magnitudes.
 
 Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot.
-For select outcomes, I present the original de-aggregated estimates to show that results are not driven by a particular year of treatment.
-Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each table.
+For select outcomes, I present the original pre-aggregation estimates to show that results are not driven by a particular year of treatment.
+Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each results plot.
 Simultaneous confidence bands, calculated using bootstrapped standard errors, are presented.
@@
 Second, I assume physicians do not anticipate EHR exposure prior to occurrence.
-Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until a future period.
+This may be a concern if physicians learn that the system will be implemented and change their behavior before implementation occurs.
 Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
@@
 I investigate this assumption in Section \ref{sec:dataass}.
+While physicians may create work-arounds to ease the burden of EHR use, use in some capacity is inevitable.
 Second, I assume that a physician's exposure to EHRs is exogenous.
@@
 
-\subsection{Retirement}
-
-In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age.
-These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
-The estimates in the top panel suggest that being exposed to an EHR leads to a .003 ppt increase in the likelihood of retiring in the first year after exposure, and a .0045 ppt increase in the second year after exposure.
+\subsection{Retirement}\label{sec:retire}
+
+I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age in Figure \ref{fig:retirefirst}.\footnote{Alternatively, one could estimate the effect for more age groups and observe the relationship between the overall ATT and age.
+However, the data is not ideal for more than two age groups since the sample sizes decrease dramatically for certain groups.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
+The estimates in the top panel suggest that being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in the first year after exposure, and a .005 ppt increase in the second year after exposure.
 While numerically small, these effects are statistically and economically meaningful relative to the proportion of physicians who retire in the sample, .03.
-That is, the effect represents a 10\% (or 15\% in the second year) increase in the likelihood of retirement relative to the mean.
+That is, the effect represents a 7\% (16\% in the second year) increase in the likelihood of retirement relative to the mean.
 Even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0008), 0.002 (0.0009), 0.005 (0.002), respectively.}}
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0009), 0.002 (0.001), 0.006 (0.003), respectively.}}
 \end{figure}
@@
 These results are shown in the bottom panel of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above in the year after exposure, while younger physicians are driving the result for the second year after exposure.
-A physician who is at least 60 years old is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
-In comparison, physicians less than 60 are far less likely, if at all, to respond in the first year after EHR exposure.
-Physicians less than 60 are approximately 16\% more likely to retire two years after exposure.
+A retirement age physician is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
+In comparison, physicians less than 60 are not more likely to retire in the first year after exposure, but are approximately 16\% more likely to retire two years after exposure.
 There are two potential reasons for the different response time.
 First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
-Alternatively, physicians who are not yet retirement age may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
-
-One argument against this result is that physicians are not leaving a clinical environment, they are simply choosing to avoid seeing Medicare patients.
-While EHR use should not be correlated specifically with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
+Similarly, younger physicians may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
+
+A potential concern is that physicians are not leaving a clinical setting, but instead they stop seeing Medicare patients.
+While EHR use should not be directly correlated with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
 The Center for Medicare and Medicaid services publishes a list of physicians who have officially opted out of seeing Medicare patients.
 None of the physicians in my sample have selected into this option.
-Further, the differential timing of retirement from older vs.
+Further, the differential timing in the effect of older vs.
 younger physicians is reassuring, as any Medicare change should not affect physicians of different ages differently.
@@
 There is still a large portion of physicians not induced to retire due to EHR exposure.
-Thus, I consider the decision to switch work settings using multiple outcomes.
-First, I investigate the extent to which hospitalists work in an office and whether this changes due to EHR exposure.
+Thus, I investigate another dimension by which physicians could change behavior: the extent to which they work in an office and whether this changes due to EHR exposure.
 I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year.
-Second, I investigate whether EHR exposure causes a change in the likelihood of switching zip codes.
 This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
 
-I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
-The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the same year as exposure by .014 ppts, a 4.6\% increase relative to the mean.
-This result indicates that hospitalists are more likely to start working in an office in the year they are exposed to an EHR, though the effect is not as large as the one found for physicians retiring.
-Splitting the sample by age reveals that this result is driven by the younger sample of physicians.
-For retirement age physicians, the estimates become very noisy and I do not draw any conclusions.
-Younger physicians, however, show evidence of switching behavior.
-The .015 ppt increase in the likelihood of retiring is equivalent to a 5.4\% increase relative to the mean for this age group.
-While there is not strong evidence for the existence of a pre-trend, I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}.
-This analysis yields the same result.
+I first discuss the estimated effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
+The top panel shows aggregated group time treatment effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office by at least .04 ppts in every year after being exposed, where the effect increases over time (up to .1 ppts four years after exposure).
+This estimate is equivalent to a 15-38\% increase relative to the average proportion of physicians who work in an office.
+That is, physicians working in a hospital are more likely to start working in an office after being exposed to an EHR, and these effects are just as large as those on retirement.
+However, unlike the estimated retirement effects, the increased likelihood of working in an office is persistent over time, which suggests that EHRs impact work environment in hospitals for several years after implementation.
 
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0004 (0.008), -0.004 (0.008), 0.02 (0.02), respectively.}}
-\end{figure}
-
-The previous results is driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians shift the amount of work they do in an office.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is also tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office, and the results are the same.} I show the estimates and 95\% confidence intervals for this outcome in Figure \ref{fig:officesecond}, where I find no evidence that EHR exposure affects the fraction of patients seen in office settings.
-This does not seem to be a margin where physicians are changing their behavior due to EHR use.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .07 (0.008), .06 (0.008), 0.12 (0.03), respectively.}}
+\end{figure}
+
+To continue investigating how different ages may respond differently to technology, I also present estimates for physicians $<60$ and $>=60$ separately in the bottom panel of Figure \ref{fig:officefirst}.
+Unlike the choice to retire, the age groups seem to both be responding to EHR implementation in all years after exposure.
+The magnitudes for senior physicians continue to increase over time, whereas the estimates for younger physicians level out.
+However, senior physicians are more likely to work in an office than younger physicians on average, so both groups exhibit similar increases relative to the mean.
+I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}, where the findings are similar.
+
+The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in an office and hospital before EHR exposure change the amount of work they do in an office.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I could also estimate this analysis with the outcome of total number of patients seen in an office.
+The results are the same.} As with previous outcomes, I show estimates and 95\% confidence intervals for the whole sample, as well as different age groups in Figure \ref{fig:officesecond}.
+The estimates show an increase in the fraction of patients seen in office settings in each year once the physician is exposed to an EHR.
+In the year of exposure, the fraction of patients increases by .06, and by the fourth year after exposure the fraction of patients increases by .12 due to exposure.
+Since the physicians included in the sample are meant to have close ties with hospitals, the average fraction of patients seen in an office is low, at .08.
+Thus, the observed effect I estimate here is massive as it is equivalent to a 75-150\% increase in the fraction of patients seen in an office relative to the mean.
+Further, the variation in this variable can either be driven by physicians who already worked in an office or those who did not and then began working in an office after exposure.
+When I partition the sample into prior office work, the effects are identical for both groups.
 
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: -0.0003 (0.004), -0.0007 (0.005), -0.006 (0.018), respectively.}}
-\end{figure}
-
-The previous results only capture switching behavior which involves office based settings.
-Another way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors.
-Again, I exclude physicians who ever retire.
-The outcome variable is equal to 1 if their list of primary zip codes changes.
-This variable is limited in how broadly it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes.
-While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation that can be combined with the previous outcome variables to draw conclusions about general behavior.
-
-The effects of EHR exposure on changing zip codes over time are shown in Figure \ref{fig:zip}.
-For the full sample of hospitalists, there is a statistically positive effect observed in each year after EHR exposure, but there is evidence that this result is driven by a pre-trend before exposure.
-Therefore, I do not make any causal arguments about this effect.
-However, I include a plot of confidence intervals that are robust to specified violations of parallel trends for each age group in Appendix \ref{sec:pretrends}.
-These results reveal that under linear violations of the parallel trends assumption, there is still a positive relationship between EHR exposure and changing zip codes for younger physicians.
-When taken together, these estimates suggest there is slight evidence of physicians who do not retire changing aspects of their work environment, but not on a large scale, indicating stickiness in place of work.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .08 (0.003), .07 (0.003), 0.13 (0.01), respectively.}}
+\end{figure}
+
+The increasing trend over time remains the same for physicians of different age groups.
+However, senior physicians have larger estimated magnitudes in each year after exposure.
+Since senior physicians have more patients in an office on average, the percentage change is similar between the two age groups.
+
+\subsection{Patient Count and Billing Activity}\label{sec:patientcount}
+
+The previous sections reveal that EHRs are inducing some physicians to either retire or work more in an office, but do not give insight to how EHRs impact those who don't change their behavior.
+Thus, as a measure of productivity/efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
+If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR.
+Thus, I limit to physicians who work with the same one hospital through the entire sample.
+These hospitalists do not retire or see patients in another hospital, implying that when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.
+
+The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
+For hospitalists of any age, patient count increases by 22.4 patients in the year of EHR exposure, 45.4 patients in the year after exposure, 28.1 patients in the second year after exposure, and approximately 20 in the third and fourth year after exposure.
+These estimates translate to a 6-14\% increase in the number of patients seen relative to the mean.
+Interestingly, when splitting the sample by age group, the effect decreases three and four years after exposure for younger physicians, while the effect for senior age physicians remains persistent over time, suggesting that technology's effect on productivity in the long run depends on worker type (in this case, age or familiarity with technology of the worker).
+Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends} and find similar results.
 
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Likelihood of Changing Zip Codes}
-    \includegraphics[scale=.6]{Objects/zip_plot.pdf}
-    \label{fig:zip}
+    \caption{Effect of EHR Exposure on Patient Count}
+    \includegraphics[scale=.6]{Objects/patient_plot.pdf}
+    \label{fig:patient}
     \vspace{2mm}
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.045 (0.006), 0.046 (0.007), 0.042 (0.018), respectively.}}
-\end{figure}
-
-\subsection{Patient Count and Billing Activity}\label{sec:patientcount}
-
-As a measure of productivity or efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
-If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR.
-Thus, I limit the sample to only physicians who work with the same one hospital through the entire sample.
-These hospitalists do not retire or see patients in another hospital.
-Thus, when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.
-
-The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
-For hospitalists of any age, patient count increases by 17.5 patients in the year of EHR exposure, 34.7 patients in the year after exposure, 16.6 patients in the second year after exposure, and then the effect disappears.
-These estimates translate to a 5-10\% increase in the number of patients seen relative to the mean.
-Interestingly, when splitting the sample by age group, the effect becomes null for those of retirement age, suggesting that technology use impacts productivity dependent on worker type (in this case, age).
-Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends}.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 28 (5.0), 25.6 (5.2), 40.4 (15.0), respectively.}}
+\end{figure}
+
+While I am investigating the supply side factors of this outcome, patient count is also driven by demand for health care.
+A major change in the U.S.
+health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
+If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count.
+The ACA had potentially opposing effects on Medicare patients: they may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
+A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
+Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}.
+Each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups affected by the ACA are not driving the results.
+
+\begin{figure}[ht!]
+    \centering
+    \captionsetup{width=.6\linewidth}
+    \caption{Effect of EHR Exposure on Patient Count by Group}
+    \includegraphics[scale=.47]{Objects/patient_group.pdf}
+    \label{fig:patientgroup}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Plot shows average group time treatment effects for each treatment group.}}
+\end{figure}
+
+One concern with my estimates is that EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
+That is, maybe really productive physicians are the ones who remain in the same hospital after EHR exposure.
+When including those physicians who are never treated in any year of the sample, the proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
+To investigate whether my results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}).
+For these bounds to be valid, a monotonicity assumption must hold on selection.
+That is, treatment can only affect selection in one direction.
+In my case, that means that EHR exposure increases the likelihood of retiring or moving workplace settings and no physicians remain in the sample under treatment who would have retired or moved in the absence of EHR exposure.
+Since I find in earlier analyses that physicians are more likely to move and retire due to EHRs, this assumption seems reasonable.
+To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate (18\%).
+Trimming the bottom (top) provides an upper (lower) bound.
+The overall average treatment effect falls within 33.08 patients and 66.35 patients, compared to the main result of 28 patients.
+This indicated selection causes downward bias in the effect of EHR exposure on patient count.
+
+Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
+Specifically, if a hospital loses a portion of their physicians, they may divide those patients between the remaining physicians.
+In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away then these estimates could be driven by increased workload instead of EHR use.
+Just by observing the pattern of the estimates on patient count over time, this could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time.
+Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from retired physicians were divided between remaining physicians.
+In the year of exposure the entire increase in patient count is due to EHR use since I find no effect on the likelihood of retirement.
+In the first year after exposure, I estimate the increase in patients due to retiring physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above.
+However, the second year after exposure is when I observe that largest effect on retirement, and so I do not rule out that the increase in patient count could be caused by that.
+After two years, I find no strong evidence for an increase in retirement, so there does seem to be some positive effect on patients seen due to EHR use.
+
+Finally, I consider whether EHR use affects claims filed per patient.
+On average, physicians file roughly 4 claims per patient.
+I present estimates of these effect in Figure \ref{fig:claim}, which shows that EHRs tend to increase claims filed per patient over time.
+The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean.
+There is no evidence of a pre-trend driving this result.
+I also heterogeneity of this result by age group, and the bottom panel of Figure \ref{fig:claim} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
 
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.6]{Objects/patient_plot.pdf}
-    \label{fig:patient}
+    \caption{Effect of EHR Exposure on Claims per Patient}
+    \includegraphics[scale=.6]{Objects/claim_per_patient_plot.pdf}
+    \label{fig:claim}
     \vspace{2mm}
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 14.1 (4.6), 15.8 (5.4), -0.29 (13.0), respectively.}}
-\end{figure}
-
-While this paper is interested the supply side factors of patients seen, this variable is also driven by demand for health care.
-A major change in the U.S.
-health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
-If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count.
-The ACA had potentially opposing effects on Medicare patients: Medicare patients may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
-A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
-Further, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}.
-Each group shows the same pattern of a small increase in patient count in the first 0-2 years after implementation.
-
-\begin{figure}[ht!]
-    \centering
-    \captionsetup{width=.6\linewidth}
-    \caption{Effect of EHR Exposure on Patient Count by Group}
-    \includegraphics[scale=.47]{Objects/patient_group.pdf}
-    \label{fig:patientgroup}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Plot shows average group time treatment effects for each treatment group.}}
-\end{figure}
-
-Since I found in the earlier stages of analysis that physicians may have shifted away from hospitals due to EHR exposure, I also investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
-For example, if a hospital loses a portion of their physicians, they may divide those patients between the remaining physicians.
-In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away then these estimates could be driven by the retiring physicians instead of EHR use.
-In Figure \ref{fig:numphys}, I present a graph showing the relationship between EHR adoption and the number of physicians working in the hospital.
-While these are not causal estimates, they do show that the trend seems to be more physicians working in hospitals, and there is no apparent shock in the short run after EHR implementation.
-This suggests that the increase in patient count is not due to the exit of physicians observed in the earlier analysis.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Number of Physicians in Hospitals}
-    \includegraphics[scale=.4]{Objects/numberofphysicians_plot.pdf}
-    \label{fig:numphys}
-\end{figure}
-
-Next, I consider whether EHR exposure affects claims filed per patient.
-On average, claim counts are roughly 4 times the number of patients seen in a year.
-Shown in Figure \ref{fig:claim}, there is a decreasing trend in claims per patient in the three years after EHR exposure.
-Two years after exposure, claims per patient decreases by .25, roughly 6\% of the average.
-This indicates a potentially advantageous effect of EHRs, that excessive billing is reduced.
-Though noisier for retirement age physicians, this effect is seen in both age groups.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Claims per Patient}
-    \includegraphics[scale=.6]{Objects/claim_per_patient_plot.pdf}
-    \label{fig:claim}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
-The bottom show these results for different subgroups of physicians by age.
-The p-value listed for each graph corresponds to a Wald test for pre-trends.
-Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 4.04 (33.89), 25.77 (31.26), -161.06 (133.35), respectively.}}
-\end{figure}
-
-\subsection{Late Adopters as Comparison Group}
-
-For the results presented, if EHR exposure affects physician behavior, the effect typically occurs in the first 0-2 years after exposure.
-Almost all specifications point to the effect of EHR exposure becoming null in the 3-4 years after exposure.
-One reason for this may be the composition of the hospitals which adopt at different times.
-Consider event time $t+4$.
-The physician included in the estimation for this coefficient are those exposed in 2010 compared with those who did not adopt until 2015, implying that the underlying hospitals implemented EHRs in either 2010 or 2015.
-It is likely that the group of hospitals who adopted in 2015 are compositionally different than early adopters, especially if poorer hospitals adopted later.
-For this reason, I have focused discussion of results on the immediate years after exposure when hospitals the comparison groups are more comprehensive.
-
-\begin{figure}
-    \centering
-    \includegraphics[scale=.5]{Objects/control_histogram.pdf}
-    \caption{Caption}
-    \label{fig:my_label}
-\end{figure}
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .33 (.09), .37 (.08), .05 (.44), respectively.}}
+\end{figure}
+
+I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
+The computed bounds on the average treatment effect are -.02 claims per patient to .28 claims per patient.
+This does not rule out the possibility that the true effect of EHR exposure on claims per patient is zero,  and the upper bound is lower than the ATT found in the main specification, indicating that selection is biasing the effect upward.
+Because of this, I make no strong conclusions about how EHR exposure impacts physician billing activity.
 
@@
 Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator.
+they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist, and Registered Record Administrator.
 I will refer to any employee in these categories as a data assistant.
 In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
-This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs improve both outcomes.
+This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs have an effect.
 
@@
 
-Further, I investigate the occurrence of hospitals and data assistants sharing patients in the CMS Shared Patient data.
-I create a list of hospital-years for which the hospital has a positive number of shared patients with a data assistant.
-I merge this to the physician-hospital pairs and create a physician level variable for whether any hospital worked with is associated with a data assistant.
+Further, I investigate the occurrence of physicians and data assistants sharing patients in the CMS Shared Patient data.
+I create a list of physician-years for which the hospital has a positive number of shared patients with a data assistant.
+I merge this to the physician-hospital pairs and create a physician level variable for they are associated with a data assistant.
 I find that .1\% of physician-year observations are associated with data assistants.
 This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
-However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar, albeit more noisy.
-Details of this analysis are given in Appendix \ref{app:DA}.
+However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar.
+Results from specifications where data assistants are excluded are given in Appendix \ref{sec:chart}.
 
@@
 In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes.
-This technology has been controversial and debated as a burdensome load on physicians that may contribute to physician burnout, and therefore patient outcomes.
-Understanding how physicians respond to this technology is important as physician labor markets have implications for access to care, quality of care, and costs.
-
-I utilize data on shared patients, hospital EHR use, and physician billing activity.
-Using a difference in difference framework, I find that physicians of all ages who work in hospitals are more likely to stop seeing Medicare patients altogether when exposed to an EHR, where those of retirement age have an even larger increase relative to younger physicians.
-Further, I find slight evidence that physicians who do not retire change where they work because of EHR exposure.
-Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen and a decrease in claims filed per patient.
+This technology has been controversial and speculated as a cause of physician burnout to physician burnout.
+Understanding how physicians respond to this technology is important, as physician labor markets have implications for access to care, quality of care, and costs.
+
+I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity.
+Using a difference in difference framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
+Further, I find evidence that physicians who do not retire change where they work because of EHR exposure, particularly shifting from hospitals towards offices.
+Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
 
 The implications of these results are threefold.
-First, a surge of physicians left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening access to care.
-In a field where younger physicians learn a lot by watching those more experienced than them, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research.
-Further, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens.
-When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation.
-Finally, these results speak to the ongoing debate of whether EHRs are generally beneficial to its users.
-While the technology may be a burden, physicians are able to see more patients because of their use, and do not seem to over-bill because of them.
+First, many physicians left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care issues.
+Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research.
+Third, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens.
+When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation and incorporate this into the incentives.
+Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient.
+My findings suggest that EHRs make physicians more productive through the number of patients they see.
+However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, so a future research question is whether this increase in productivity is traded off with time spent with patients.
 
@@
 
-I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI-Files.html}}} to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}.
+I begin by connecting every National Provider Identifier (NPI)\footnote{Data can be downloaded from \hyperlink{https://download.cms.gov/nppes/NPI/Files.html}{\text{https://download.cms.gov/nppes/NPI-Files.html}}}, an identifier of each medical entity, to a description of their tax code\footnote{Data can be downloaded from \hyperlink{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}{\text{https://nucc.org/index.php/code-sets-mainmenu-41/provider-taxonomy-mainmenu-40/pdf-mainmenu-53}}}.
 I then categorize NPIs by key words in their tax code description.
-Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital.
+Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as such.
 I create two data sets, one containing only PCPs and one containing only hospitals.
-These data are saved to use in identifying the relevant NPIs in the CMS Shared Patient Data.
+These data are then used in identifying the relevant NPIs in the CMS Shared Patient Data.
 
@@
 I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs.
-Some pairs are duplicates from the hospital being listed first or second in the shared patient data.
+Some pairs are duplicates from the hospital being listed first or second in the shared patient data (being listed first means the patient went to NPI 1 first, but my analysis does not depend on the order of care, only the relationship between hospital and physician as a whole).
 I combine duplicates into one observation, summing the same day count variable.
-Once duplicated are removed, there are 7.1 million observations.
+Once duplicates are removed, there are 7.1 million observations.
 Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital.
-I drop any pairs who do not have at least 30 shared patients in the same day per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60.
+I drop any pairs who do not have at least 30 same-day shared patients per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60.
 Results are similar and can be found in Figure \ref{sec:chart}.} The final list of pairs consists of 1.3 million observations.
@@
 I combine the physician-hospital pairs created in Section \ref{sec:sharedpat} with various data sets for hospital or physician level information.
-First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation.
+First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation date.
 This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation.
@@
 For hospital level variables, I use an AHA-NPI crosswalk to merge the pairs to the Annual Hospital Administration Survey (AHA Survey) from 2009-2015.
-The variables I include from this data are a hospital's number of beds, organization type, system ID, and EHR use.
+From this data I collect information on each hospital's EHR use.
 There are 4,253 unique AHA hospitals in the shared patient data.
@@
 There are only 58 unique hospitals in the data with less than 10 beds, and these are likely very different from the rest of the sample of hospitals.
-Then, I create a physician-level variable which captures the average number of beds in the hospitals they work with in each year.
 I sum a physician's same day count over their hospitals in a given year to create a physician level patient count variable.
-I also create this variable only for hospitals who are using an EHR and only for hospitals who are not using an EHR.
-
-I use the general survey question about each hospital's EHR use to define my key treatment variable.
+
+Specifically, I use the general survey question about each hospital's EHR use to define my key treatment variable.
 The survey asks the extent to which the hospital utilizes an EHR and allows for three answers: not at all, partially utilize an EHR, and fully utilize an EHR.
@@
 The level of vertical integration between hospital and physician is defined as in \citeauthor{dynan1998assessing} (\citeyear{dynan1998assessing}) using the organization type of the hospital.
-Hospitals classified as IPA or PHO are low-integration, and any other type is high0integration.
-I also create a variable that captures the fraction of hospitals a physician works with that use an EHR, which varies on the dimension of EHR use and on number of hospitals worked with.
+Hospitals classified as IPA or PHO are low-integration, and any other type is high integration.
 Finally, the last variable I create before aggregating to the physician level is an indicator for whether the physician works with the same set of hospitals through the entire sample.
@@
 Finally, I aggregate to the physician level by only keeping variables that do not vary by hospital.
-The variables in the physician level data are year, physician NPI, graduation year, average number of beds in hospitals worked with, years of experience, number of hospitals, fraction of hospitals with EHR, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, and never works with a new NPI.
+The variables in the physician level data are year, physician NPI, graduation year, years of experience, number of hospitals, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, and never works with a new NPI.
 I complete this data to include years 2016 and 2017, but leave time varying variables missing for those years.
@@
 The results are similar, and can be found in Section \ref{sec:chart}} This limitation is to continue ensuring that I am focusing on physicians working in hospitals who will be subject to their EHR use.
-With this limitation, there are 214,000 observations left in the data.
+With this limitation, there are approximately 214,000 observations left in the data.
 
@@
 I then define the outcome variable, retire, as one if a physician ever retires and the year is equal to the first year of retirement.
-I utilize a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.
+I also create a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.
 
@@
 The data already contains a variable for the fraction of patients seen in an office setting, which is used in the analysis, but I further create an indicator variable for whether the physician sees any positive amount of patients in an office setting.
-
-I also create an indicator variable for whether the physician switches zip codes from one year to the next.
-The physicians in this data work in up to 12 different zip codes, so I their zip codes into one string variable, a list of all zip codes in a year.
-I make this variable into wide format: zip list 2009, zip list 2010, ...
-, zip list 2017.
-In year $t$, I define a variable change zip that is equal to one if the zip list in $t-1$ is different from the zip list in $t$, zero otherwise.
 Finally, the variables for unique patient count and claim count are already defined.
+I simply divide the number of claims by the number of patients to get the claims per patient variable used in the analysis.
 This is the final data used in the paper, with approximately 215,000 observations.
@@
     \captionsetup{width=.5\linewidth}
-    \caption{}
+    \caption{Retire Pretrends Plot}
     \includegraphics[scale=.5]{Objects/retire_pretrends_plot.pdf}
@@
     \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:}}}
+    \caption*{\footnotesize{\textit{Notes: This plot shows average treatment effects one year after EHR exposure under different trend assumptions. M=0 indicates a linear pre-trend.}}}
 \end{figure}
@@
     \captionsetup{width=.5\linewidth}
-    \caption{}
-    \includegraphics[scale=.5]{Objects/work_pretrends_plot.pdf}
+    \caption{Work in Office Indicator}
+    \includegraphics[scale=.5]{Objects/office_ind_pretrends_plot.pdf}
     \label{fig:pre_work}
     \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:}}}
+    \caption*{\footnotesize{\textit{Notes: Notes: This plot shows average treatment effects one year after EHR exposure under different trend assumptions. M=0 indicates a linear pre-trend.}}}
 \end{figure}
@@
     \captionsetup{width=.5\linewidth}
-    \caption{}
-    \includegraphics[scale=.5]{Objects/frac_pretrends_plot.pdf}
+    \caption{Fraction of Patients Seen in Office}
+    \includegraphics[scale=.5]{Objects/office_frac_pretrends_plot.pdf}
     \label{fig:pre_frac}
     \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:}}}
-\end{figure}
-
-\subsubsection{Change Zip Codes Likelihood}
-
-I now consider pre-trends for the effect of EHR exposure on whether the physician changes zip codes.
-In all main specification results, there is indication of a clear pre-trend and thus it is not clear whether there is a positive effect due to EHR exposure or due to this trend occurring prior to exposure.
-Thus, I present confidence intervals that are robust to different specifications of parallel trend violations for each age group at event time 0 and event time 1 in Figure \ref{fig:pre_zip}.
-At event time 0, it seems younger physicians are more likely to change zip codes after EHR exposure under a linear pre-trend violation and slightly non-linear pre-trend violations, although the magnitude of the effect is likely smaller than the main specification shows.
-This is also true for younger physicians in event time 1.
-For physicians of retirement age, it seems the entire effect was driven by a pre-trend and thus there is no effect of EHR exposure on this group's likelihood of changing zip codes.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.5\linewidth}
-    \caption{}
-    \includegraphics[scale=.5]{Objects/zip_pretrends_plot.pdf}
-    \label{fig:pre_zip}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:}}}
+    \caption*{\footnotesize{\textit{Notes: This plot shows average treatment effects one year after EHR exposure under different trend assumptions. M=0 indicates a linear pre-trend.}}}
 \end{figure}
@@
     \captionsetup{width=.5\linewidth}
-    \caption{}
+    \caption{Patient Count}
     \includegraphics[scale=.5]{Objects/patient_pretrends_plot.pdf}
@@
     \captionsetup{width=.5\linewidth}
-    \caption{}
+    \caption{Claims per Patient}
     \includegraphics[scale=.5]{Objects/claim_pretrends_plot.pdf}
@@
 As discussed in the body of the paper, I address potential concerns that the hiring of new employees may be driving the results for productivity and efficiency.
-Thus, as another specification, I restrict the data to only include physicians who are exposed in hospitals that do not record employing any data assistants.
+Thus, as another specification, I restrict the data to only include physicians who have no record of working with data assistants as described in Section \ref{sec:dataass}.
 
@@
     \caption{Outcome: Work in Office}
-    \includegraphics[scale=.6]{Objects/office_chart.pdf}
+    \includegraphics[scale=.6]{Objects/office_ind_chart.pdf}
     \label{fig:work_chart}
@@
 
-Finally, I discuss the specification chart regarding a physician's likelihood of changing zip codes due to EHR exposure.
-These results should be taken in conjunction with Section \ref{sec:pretrends}, which analyzes the extent to which this result is driven by a pre-trend.
-The main specification yields an average treatment effect of .04 (significant at all levels), which is in the average confidence interval across specifications.
-One thing to notice in this graph is that there are two changes to specification for which the resulting average treatment effect is grouped towards one end of the distribution or the other.
-Specifically, when the percent of patients seen in an inpatient setting is 90\%, the coefficient is likely to be higher than when it is 50\% or 70\%.
-That is, a more strict definition of hospitalist yields a higher estimate for the likelihood of changing zip codes (although the variance of estimates is not particularly high).
-Further, using the main EHR definition tends to yield lower estimates than the low integration definition.
-That is, endogeneity is biasing this estimate slightly down.
-
-\begin{figure}[ht]
-    \centering
-    \caption{Outcome: Change Zip Codes}
-    \includegraphics[scale=.6]{Objects/change_zip_chart.pdf}
-    \label{fig:zip_chart}
-\end{figure}
-
 \subsubsection{Patient Count and Claims per Patient}
@@
     \caption{Outcome: Patient Count}
-    \includegraphics[scale=.6]{Objects/patientcount_chart.pdf}
+    \includegraphics[scale=.6]{Objects/patient_chart.pdf}
     \label{fig:pat_chart}
@@
 
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.57\linewidth}
-    \caption{Other Event Study Estimators}
-    \includegraphics[scale=.57]{Objects/estimators_graph.pdf}
-    \label{fig:esimators}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Results from various methods of estimation are shown, along with 95\% confidence intervals.}}
+\begin{figure}[t!]
+\begin{subfigure}{0.48\textwidth}
+\caption{Retire}
+\includegraphics[width=\linewidth]{Objects/retire_est_plot.pdf}
+\label{fig:a}
+\end{subfigure}\hspace*{\fill}
+\begin{subfigure}{0.48\textwidth}
+\caption{Fraction of Patients in Office}
+\includegraphics[width=\linewidth]{Objects/frac_office_est_plot.pdf}
+\label{fig:b}
+\end{subfigure}
+
+\medskip
+\begin{subfigure}{0.48\textwidth}
+\caption{Indicator for Office }
+\includegraphics[width=\linewidth]{Objects/office_ind_est_plot.pdf}
+ \label{fig:c}
+\end{subfigure}\hspace*{\fill}
+\begin{subfigure}{0.48\textwidth}
+\caption{Change Zip Code} 
+\includegraphics[width=\linewidth]{Objects/patient_count_est_plot.pdf}
+\label{fig:d}
+\end{subfigure}
+
+\medskip
+\begin{subfigure}{0.48\textwidth}
+\caption{Patient Count} 
+\includegraphics[width=\linewidth]{Objects/claimperpatient_est_plot.pdf}
+\label{fig:e}
+\end{subfigure}\hspace*{\fill}
+
+\caption{Different Estimators} \label{fig:estimators}
 \end{figure}

### 2022-10-12 Paper/Paper.tex
Updates from Overleaf

@@
 Further, for physicians who remain exposed to the technology, the number of patients seen increases.
-These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential productivity gains for those who do not avoid the technology through labor market behavioral changes.
+These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential productivity gains for those who do not avoid the technology.
 \end{abstract}
@@
 Since I find in earlier analyses that physicians are more likely to move and retire due to EHRs, this assumption seems reasonable.
-To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate (18\%).
-Trimming the bottom (top) provides an upper (lower) bound.
-The overall average treatment effect falls within 33.08 patients and 66.35 patients, compared to the main result of 28 patients.
+To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate.
+Trimming the bottom (top) and taking the differences in expectation between treatment and trimmed control provides a lower (upper) bound.
+This bounding yields an interval estimate of 54.6 patients to 125.4 patients, compared to the main result of 28 patients.
 This indicated selection causes downward bias in the effect of EHR exposure on patient count.
@@
 I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
-The computed bounds on the average treatment effect are -.02 claims per patient to .28 claims per patient.
-This does not rule out the possibility that the true effect of EHR exposure on claims per patient is zero,  and the upper bound is lower than the ATT found in the main specification, indicating that selection is biasing the effect upward.
-Because of this, I make no strong conclusions about how EHR exposure impacts physician billing activity.
+The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33.
+Similarly to patient count, selection biased the result in the main specification downward.
 
@@
 While visually there does not appear to be a strong pre-trend, for the full sample of physicians and physicians $< 60$, the p-value indicates a pre-trend may exist.
-Since the positive effect seems to be a possibility in event time 1 and event time 2, I present a plot of robust confidence intervals under various possible assumptions of parallel trends violations for these periods in Figure \ref{fig:pre_retire}.
+I present a plot of robust confidence intervals under various possible assumptions of parallel trends violations for one year after exposure in Figure \ref{fig:pre_retire}.
 The original confidence interval is shown in yellow, and in gray are confidence intervals robust to varying functions of parallel trend violations.
-When allowing for linear trend violations (M=0), the effect of EHR exposure on retirement is still positive.
-As I allow for more non-linearity (M$>$0) in these variations, this effect becomes zero.
+As I allow for different assumptions on parallel trends, the confidence intervals become larger and contain zero.
+Under linearity of pre-trend, the confidence interval contains zero but is more similar to the effect found in the main result.
 This is true for both event time 1 and event time 2.
@@
 The p-values do not indicate pre-trends for the full sample or for physicians $< 60$, but I reject no pre-trends for physicians of retirement age.
-As with retirement as the outcome, I show robust confidence intervals in Figure \ref{fig:pre_work}.
-However, the main results show up in event time 0 and 1, so I show confidence intervals for these periods.
-Further, I show these for the full sample of physicians, but also for older physicians separately since the pre-trend seems to differ slightly.
-When allowing for linear pre-trend violations, I find very similar results to the main specification.
-For physicians $>= 60$, there is a higher chance that the effect of EHR exposure on the probability of working in an office is positive.
-This would be theoretically intuitive since older physicians may struggle with the technology more than younger physicians and choose to switch settings.
-However, I do not make strong conclusions given the null results found in most sensitivity analyses.
+I show robust confidence intervals for the average treatment effect in the first year after exposure in Figure \ref{fig:pre_work}.
+For various assumptions on the form of pre-trends, the positive effect holds.
 
@@
 All p-values suggest there are no pre-trends.
-I show robust confidence intervals for the full sample of physicians in the first and second year of exposure in Figure \ref{fig:pre_frac}.
-These results hold no matter the assumption on the violation of pre-trends.
+I show robust confidence intervals for the average treatment effect in the first year after exposure in Figure \ref{fig:pre_frac}.
+For various assumptions on the form of pre-trends, the positive effect holds.
 
@@
 None of the p-values indicate a violation of parallel trends, but I present a plot of confidence intervals robust to specified variations in parallel trends as a robustness check in Figure \ref{fig:pre_patient}.
-These plots indicate that, even under nonlinear deviations in parallel trends, there is a positive effect of EHR exposure on patient count.
-Under a linear variation in parallel trends, the effect is positive, yet the magnitude is nearly half of the magnitude found in the main specification.
+This plots indicate that, even under nonlinear deviations in parallel trends, there is a positive effect of EHR exposure on patient count with similar magnitude to that of the main specification.
 
@@
 Finally, I assess the results found in the main specification for the outcome variable claim count.
-The main specification indicates a downward trend in the years leading up to exposure and a positive effect in the year of exposure and the year after exposure for younger physicians, and no effect for older physicians.
-In Figure \ref{fig:pre_claim}, I show confidence intervals robust to violations in parallel trends for these two age groups.
-For physicians $>= 60$, this plot shows a null effect just as in the main specification.
-For physicians $< 60$, there still seems to be a positive effect on both event years, although under exactly linear variations in parallel trends, this effect is smaller or zero.
+The main specification indicates an increase in claims per patient after EHR exposure, but under various specification changes this effect becomes null.
+Under linear assumptions of a pre-trend form, the effect is similar to that of the main specification.
+Once more non-linearity is allowed, the effect becomes null.
 
@@
 The average confidence interval for all combinations is (.001,.003), so the main specification lies in this interval.
-These results are strongly in favor of there being a positive effect of EHR exposure on retirement.
+These results support my main finding that EHR exposure positively affects the likelihood of retirement among physicians.
 
@@
 
-The specification charts for outcomes work in office, fraction of patients seen in office, and changing zip codes are presented in Figures \ref{fig:work_chart}, \ref{fig:fracoffice_chart}, and \ref{fig:zip_chart}, respectively.
+The specification charts for outcomes work in office and fraction of patients seen in office are presented in Figures \ref{fig:work_chart} and \ref{fig:fracoffice_chart}, respectively.
 First, I discuss the overall average treatment effect of EHR exposure on the likelihood of working in an office.
-While the main specification showed a slight increase in likelihood in the year of exposure, the overall treatment effect is 0.
-Most of the specifications agree with the main specification, although more coefficients fall below 0 than above 0.
-The average confidence interval is approximately (-.016, .001), encompassing the main specification.
+The main specification yields finding that EHR exposure leads to a .07 ppt increase in the likelihood of working in an office.
+This is on the high end of all combinations of changes to specification, but all reveal a positive effect.
+On average, the confidence interval falls between .035 and .06.
+This supports the finding that EHR exposure positively affects the likelihood of working in an office.
+
+Specification changes that seem to affect the magnitude of the estimate are whether I limit the sample years and which EHR definition is used as the treatment variable.
+Neither of these change the sign of the estimate, but both limiting sample years and using the low-integration definition of EHR exposure push the magnitude slightly down.
 
@@
 
-Second, I discuss the results for whether physicians change the fraction of patients seen in an office setting due to EHR exposure. The main specification
+Now I discuss the results for whether physicians change the fraction of patients seen in an office setting due to EHR exposure.
+Similarly, the ATT from the main specification is on the high end of other ATTs found, but all estimates suggest a positive effect.
+The average of all the confidence intervals found is (.02, .06), suggesting a positive effect on fraction of patients seen in an office.
+Changes that affect the magnitude of the estimate (but not the sign) are limiting sample years and anticipation.
+Limiting the years in the sample and allowing for one year of anticipation both push the magnitude of the estimate down.
 
@@
 
+In Figure \ref{fig:pat_chart}, I present estimates and confidence intervals for different combinations of specification changes.
+The results from the main specification is that EHR exposure increases patient count by 28.
+While most specifications yield positive estimates, about a third of the specifications have a confidence interval that includes zero.
+It doesn't seem as though any specific change is driving the null result, so I hypothesize that it is due to the reduction in observations due to data limiting.
+The average of all of the confidence intervals is 7 to 23, indicating that a positive effect on patient count is reasonable.
+
 \begin{figure}[ht]
@@
 
+Lastly, I examine a chart of estimates for different specifications attempting to capture the effect of EHR exposure on claims per patient.
+My main finding is that EHR exposure increases claims per patient by around .3.
+However, various robustness checks reveal that this is overestimated.
+That is supported in this specification chart as well, where around half of the estimated confidence intervals include zero.
+Thus, I make no conclusion that EHR exposure increases claims per patient.
+
 \begin{figure}[ht]
@@
 $$\text{outcome}_{i,t}=\sum_{j=-4}^{-2} \text{rel\_year}_{j} + \sum_{j=0}^{4} \text{rel\_year}_{j} + \delta_i + \gamma_t,$$
-where outcome$_{i,t}$ is one of the six outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to expansion, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
+where outcome$_{i,t}$ is one of the five outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to expansion, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
 These results from each outcome are presented in Figure \ref{fig:twfe}. These results show similar patterns to the main specification, with slight variations in magnitude.
@@
 This approach is commonly known as two stage difference in differences and is very clever.
-However, the approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
-Nevertheless, I limit the years of data to 2009-2015 and present estimates of this estimator in Figure \ref{fig:esimators}.
+However, the approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
+Nevertheless, I limit the years of data to 2009-2015 and present estimates of this estimator in Figure \ref{fig:estimators}.
 Generally, the results have the same sign with a slightly larger magnitude, driven by the inclusion of never-treated units.
@@
 \includegraphics[width=\linewidth]{Objects/retire_est_plot.pdf}
-\label{fig:a}
+\label{fig:a1}
 \end{subfigure}\hspace*{\fill}
@@
 \caption{Fraction of Patients in Office}
-\includegraphics[width=\linewidth]{Objects/frac_office_est_plot.pdf}
-\label{fig:b}
+\includegraphics[width=\linewidth]{Objects/office_frac_est_plot.pdf}
+\label{fig:b1}
 \end{subfigure}
@@
 \caption{Indicator for Office }
-\includegraphics[width=\linewidth]{Objects/office_ind_est_plot.pdf}
- \label{fig:c}
+\includegraphics[width=\linewidth]{Objects/office_in_est_plot.pdf}
+ \label{fig:c1}
 \end{subfigure}\hspace*{\fill}
 \begin{subfigure}{0.48\textwidth}
-\caption{Change Zip Code} 
-\includegraphics[width=\linewidth]{Objects/patient_count_est_plot.pdf}
-\label{fig:d}
+\caption{Patient Count} 
+\includegraphics[width=\linewidth]{Objects/patient_est_plot.pdf}
+\label{fig:d1}
 \end{subfigure}
@@
 \begin{subfigure}{0.48\textwidth}
-\caption{Patient Count} 
-\includegraphics[width=\linewidth]{Objects/claimperpatient_est_plot.pdf}
-\label{fig:e}
+\caption{Claims Per Patient} 
+\includegraphics[width=\linewidth]{Objects/claim_est_plot.pdf}
+\label{fig:e1}
 \end{subfigure}\hspace*{\fill}
@@
 
-Another estimation method established first used by \citeauthor{cengiz2019effect} (\citeyear{cengiz2019effect}) is commonly known as stacked regression.
-In this method, one re-frames event study data into groups of sub-experiments that are then stacked on top of each other, defining a group of treated units and "clean" control units.
+Another estimation method first used by \citeauthor{cengiz2019effect} (\citeyear{cengiz2019effect}) is commonly known as stacked regression.
+In this method, one re-frames event study data into groups of sub-experiments that are then stacked on top of each other, defining a group of treated units and ``clean" control units.
 The drawback of this method is choosing an event study window that must be the same for each treatment group, where a large window eliminates many treated units and a small window does not allow for estimation over many years.
-I present the results using this method with an event window of one year in Figure \ref{fig:esimators}.
+I present the results using this method with an event window of one year in Figure \ref{fig:estimators}.
 The results are very similar to the main specification, with a smaller magnitude in some cases.
@@
 These average treatment effects are shown in Figure \ref{fig:1}.
-For most outcomes, the results do not differ along these dimensions.
-The only outcome for which the estimate changes sign is the indicator for whether the physician works in an office setting.
-Physicians in an urban area are less likely to work in an office after EHR exposure, while rural physicians are more likely to work in an office after EHR exposure, but the sign switch does not hold when looking at the hospital size parallel.
-This results is interesting and may be driven by differing physician-hospital relationships.
-I leave further investigation to future research.
+The results do not differ along these dimensions.
+This is fascinating in that technology is creating these adverse effects across completely different practice environments.
 
@@
 \begin{subfigure}{0.48\textwidth}
-\caption{Change Zip Code} 
-\includegraphics[width=\linewidth]{Objects/change_zip_heterog.pdf}
+\caption{Patient Count} 
+\includegraphics[width=\linewidth]{Objects/patient_heterog.pdf}
 \label{fig:d}
@@
 \begin{subfigure}{0.48\textwidth}
-\caption{Patient Count} 
-\includegraphics[width=\linewidth]{Objects/patient_heterog.pdf}
+\caption{Claims Per Patient} 
+\includegraphics[width=\linewidth]{Objects/claim_per_patient_heterog.pdf}
 \label{fig:e}
 \end{subfigure}\hspace*{\fill}
-\begin{subfigure}{0.48\textwidth}
-\caption{Claim per Patient}
-\includegraphics[width=\linewidth]{Objects/claim_per_patient_heterog.pdf}
- \label{fig:f}
-\end{subfigure}
 

### 2022-10-16 Paper/Paper.tex

@@
 Further, for physicians who remain exposed to the technology, the number of patients seen increases.
-These results point to adverse responses when technology is poorly implemented in technical occupations, but also indicates potential productivity gains for those who do not avoid the technology.
+These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicates potential productivity gains for those who embrace the technology.
 \end{abstract}
@@
 \section{Introduction}
-How technological innovation affects labor markets is heavily studied, in part because of the vast differences in technology and tasks present even within industries.
-Yet, there is still much to be learned about modern technology and its interaction with how workers behave.
+For decades, technological innovation has brought about changes to production in the economy, including wages, demand for labor, and other aspect of labor markets.
+Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved with the implementation of new technologies.
+This is particularly important in health care, where new technologies can impact patient outcomes.
 A recent technology roll-out that altered the U.S.
 health care system dramatically was the electronic health record (EHR).
-The use of EHRs were incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
+The use of EHRs was incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
 Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and labor market behaviors.
 
-Specifically, EHRs are computer systems that house digital medical records and have additional capabilities, including decision making assistance.
+EHRs are computer systems that house digital medical records and have additional capabilities, including decision making assistance.
 EHRs have become increasingly relevant in the U.S.
-since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech})\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
 In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care.
-Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} (\cite{hitech}).
+Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
 President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
@@
 Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
-Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards lower cost work settings.
+Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}).
+The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
 Further, the cost imposed is likely to vary based on the type of worker, in particular, age group.
-Using a difference in difference research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
+Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
 
 I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
-Using CMS Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital.
-Then, I use the AHA Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
+Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital.
+Then, I use the American Hospital Association (AHA) Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
 Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist.
@@
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects (improved patient outcomes and decreased cost) (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
 More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
@@
 Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered ``meaningful" by the government's standards, and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework, but I improve on this by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of heterogeneous effects, and further provides insight to short vs.
+This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered ``meaningful" and are likely affecting physicians' daily life.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework.
+I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects, and further provides insight to short vs.
 long term effects.
@@
 Generally, physicians may have the ability to change hospitals, move from a hospital to an office, or work in multiple facilities and change the amount of work done in different facilities.
-I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists (for a detailed discussion of included physicians, see Appendix \ref{app:data}).
-Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
-While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faces by each type.
+I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists.
+\footnote{For a detailed discussion of included physicians, see Appendix \ref{app:data}} Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
+While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.
 
@@
 If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients.
-In this case, I productivity would increase.
+In this case, productivity would increase.
 However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
@@
 
-\subsection{MD-PPAS}
+\subsection{Sample Construction}
 
 The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017.
-Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room.
+Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts, and fraction of patients in different settings such as office, inpatient or emergency room.
 I first limit the sample of physicians based on specialty type, since my analysis relies on EHR implementation in hospitals.
@@
 
-\subsection{Shared Patients}
-
-CMS Shared Patient data records annual information on count of Medicare patients and frequency of billing associated with the same patient across providers within a given time period (30, 60, 90, or 180 days).
+CMS Shared Patient data records annual information on count of Medicare patients and frequency of billing associated with the same patient across providers within a given time period.
 These data are available from 2009 to 2015.
@@
 
-\subsection{AHA Survey Data}
-
-Using hospital NPI, I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
+Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
 I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
+A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", ``Not at all".
+Since partial implementation of an EHR does not guarantee sufficient use, I define a hospital as implementing an EHR the first year they answer ``Fully".
 I then aggregate this data to the physician level.
@@
 That is, the data contains no units which are "never-treated".
-
-\subsection{Physician Compare}
 
@@
 I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \ref{app:years}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}
+In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}
 
@@
 Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015.
-The average physician is 45 years old, works with 1.5 hospitals and 1.3 systems.
+The average physician is 45 years old and works with 1.5 hospitals and 1.3 systems.
 \import{Table Code}{overall stats.tex}
@@
 
-\subsection{Assumptions}
+\subsection{Identification Assumptions}
 
@@
 Therefore, I proceed in the main specification assuming no anticipation.
-However, I explore and present results for a one year anticipation period in Appendix \ref{app:anticipation}.
+However, I explore and present results for a one year anticipation period in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change my findings.
 
@@
 In most cases, there is no evidence of a pre-trend.
-Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends}.
+Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends} by allowing for specified violations in the parallel trends assumption (\cite{rambachan2019honest}).
+I find that most outcomes are very robust to violations in the parallel trends assumption.
 
@@
 I investigate this assumption in Section \ref{sec:dataass}.
-While physicians may create work-arounds to ease the burden of EHR use, use in some capacity is inevitable.
+While physicians may create work-arounds to ease the burden of EHR use, it is inevitable that they will need to interact with the technology in some capacity.
 Second, I assume that a physician's exposure to EHRs is exogenous.
@@
 While EHR use should not be directly correlated with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
-The Center for Medicare and Medicaid services publishes a list of physicians who have officially opted out of seeing Medicare patients.
+CMS publishes a list of physicians who have officially opted out of seeing Medicare patients.
 None of the physicians in my sample have selected into this option.
@@
 The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in an office and hospital before EHR exposure change the amount of work they do in an office.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I could also estimate this analysis with the outcome of total number of patients seen in an office.
-The results are the same.} As with previous outcomes, I show estimates and 95\% confidence intervals for the whole sample, as well as different age groups in Figure \ref{fig:officesecond}.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} As with previous outcomes, I show estimates and 95\% confidence intervals for the whole sample, as well as different age groups in Figure \ref{fig:officesecond}.
 The estimates show an increase in the fraction of patients seen in office settings in each year once the physician is exposed to an EHR.
@@
 
-The previous sections reveal that EHRs are inducing some physicians to either retire or work more in an office, but do not give insight to how EHRs impact those who don't change their behavior.
+The previous sections reveal that EHRs induce some physicians to either retire or work more in an office, but do not give insight to how EHRs impact those who don't change their behavior.
 Thus, as a measure of productivity/efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
@@
 A major change in the U.S.
-health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014 for the majority of states that partook.
+health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014.
 If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count.
@@
 One concern with my estimates is that EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
-That is, maybe really productive physicians are the ones who remain in the same hospital after EHR exposure.
+That is, maybe more productive or tech savvy physicians are the ones who remain in the same hospital after EHR exposure.
 When including those physicians who are never treated in any year of the sample, the proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
@@
 This bounding yields an interval estimate of 54.6 patients to 125.4 patients, compared to the main result of 28 patients.
-This indicated selection causes downward bias in the effect of EHR exposure on patient count.
+The bound therefore suggests selection of physicians pushes the estimate down.
 
@@
 
-Finally, I consider whether EHR use affects claims filed per patient.
+Finally, I consider whether EHR exposure affects claims filed per patient.
 On average, physicians file roughly 4 claims per patient.
 I present estimates of these effect in Figure \ref{fig:claim}, which shows that EHRs tend to increase claims filed per patient over time.
-The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean.
+The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as physicians are exposed to the new software, they bill for more items on average.
 There is no evidence of a pre-trend driving this result.
-I also heterogeneity of this result by age group, and the bottom panel of Figure \ref{fig:claim} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
+I also show heterogeneity of this result by age group, and the bottom panel of Figure \ref{fig:claim} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
+I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
+The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33.
+Similarly to patient count, the effect becomes even larger when accounting for a particular type of physician being left in the sample.
 
@@
 
-I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
-The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33.
-Similarly to patient count, selection biased the result in the main specification downward.
-
 \section{Physician Influence of EHR Decision in Hospitals}
@@
 Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-they have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist, and Registered Record Administrator.
-I will refer to any employee in these categories as a data assistant.
-In this Section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
+They have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist, and Registered Record Administrator.
+I refer to any employee in these categories as a data assistant.
+In this section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
 This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs have an effect.
@@
 This would make the binary treatment variable endogenous, biasing the results of the main specification, as it includes physicians who are selecting into their own exposure.
-After the sample period ends, the government began penalizing hospitals for not using health information technology meaningfully, so implementation eventually occurs across the board.
-However, during the sample, physicians may impact the timing or occurrence of implementation.
-I address this question by limiting the sample of hospitals in a meaningful way, only including those hospitals which are considered to have low vertical integration with physicians.
-That is, physicians are employees and less likely to be decision makers.
+After the sample period ends, the government began penalizing hospitals for not using health information technology meaningfully, so implementation eventually occurs in most hospitals.
+However, during the sample, physicians may impact the timing of implementation.
+I address this question by limiting the sample of hospitals, only including those which are considered to have low vertical integration with physicians.
+That is, physicians are not likely to be in leadership capacities, and thus less likely to be decision makers.
 I then redefine exposure as when a physician is exposed to an EHR in low integration settings.
-I present details of how I define integration and results for this specification in Appendix \ref{sec:endog}.
+I present details of how I define integration and results for this specification in Appendix \ref{sec:changes}.
 In general, group time estimates are similar to those in the main specification, with the exception of the probability of working in an office setting, which may indicate some correlation between EHR adoption in high integration settings and acquisition of physician offices.
@@
 In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes.
-This technology has been controversial and speculated as a cause of physician burnout to physician burnout.
+This technology has been controversial and speculated as a cause of physician.
 Understanding how physicians respond to this technology is important, as physician labor markets have implications for access to care, quality of care, and costs.
@@
 I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity.
-Using a difference in difference framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
+Using a difference-in-differences framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
 Further, I find evidence that physicians who do not retire change where they work because of EHR exposure, particularly shifting from hospitals towards offices.
@@
 However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, so a future research question is whether this increase in productivity is traded off with time spent with patients.
-
-This paper contributes to our understanding of health information technology by providing empirical evidence of the theoretically ambiguous physician response to electronic health records being rapidly implemented in hospitals.
-This contribution is relevant to EHR vendors, hospitals, and policymakers who are involved in further implementation of complex technology.
 
@@
 I create a variable capturing the first year that at least one of a physician's connected hospitals uses an EHR.
-For the endogeneity discussion in Section \ref{sec:endog}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR.
+For the endogeneity discussion in Section \ref{sec:endogeneity}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR.
 The level of vertical integration between hospital and physician is defined as in \citeauthor{dynan1998assessing} (\citeyear{dynan1998assessing}) using the organization type of the hospital.
@@
 Some of these changes address potential endogeneity concerns, while some provide evidence that the results do not depend on thresholds which may be arbitrary.
-I will outline each of these specification adjustments and then present specification charts which account for different combinations of these changes.
-
-\subsubsection{Physician Selection into EHRs}\label{sec:endog}
-
+I will outline each of these adjustments and then present specification charts which account for different combinations of these changes.
+
+First, I assess the assumption that exposure to an EHR is exogenous to the physician.
 The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in.
@@
 
-\subsubsection{Anticipation}\label{app:anticipation}
-
 While there is evidence to suggest the time it takes to implement an EHR does not exceed one year, there still may be some anticipation if physicians are expecting implementation in the hospitals they work in.
 Thus, the specification charts that I present in Section \ref{sec:chart} include specifications where one year of anticipation occurs.
-
-\subsubsection{Limiting Sample Years}\label{app:years}
 
@@
 
-\subsubsection{Hiring Data Assistants}
-
 As discussed in the body of the paper, I address potential concerns that the hiring of new employees may be driving the results for productivity and efficiency.
 Thus, as another specification, I restrict the data to only include physicians who have no record of working with data assistants as described in Section \ref{sec:dataass}.
-
-\subsubsection{Data Thresholds}
 
@@
     \caption{Outcome: Claims per Patient}
-    \includegraphics[scale=.6]{Objects/claim_per_patient_chart.pdf}
+    \includegraphics[scale=.6]{Objects/claim__chart.pdf}
     \label{fig:cpp_chart}

### 2022-10-31 Scripts/sensitivity_analysis.R
update appendix plots

@@ -9,6 +9,7 @@ library(HonestDiD)
 library(mvtnorm)
 
 
+cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
 
 ### --------------------  Sensitivity Analysis --------------------------
@@ -22,12 +23,14 @@ library(mvtnorm)
 # In this portion, I only consider "retire" as the dependent outcome and I compare estimators. 
 
 # Read in the main data created in "data5_MDPPAS.R"
-Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
+Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds")) %>%
+  mutate(DocNPI=as.numeric(DocNPI))
+
 
 
 
 ## DIFFERENT ESTIMATORS ############################################################################################
-varlist <- list("retire", "pos_office", "work_in_office", "change_zip", "npi_unq_benes", "claim_per_patient")
+varlist <- list("retire", "pos_office", "work_in_office", "npi_unq_benes", "claim_per_patient")
 
 # Stacked Regression ###
 for (d in 2010:2015){
@@ -50,6 +53,7 @@ models_stacked <- lapply(varlist, function(x){
         data=stacked_data_1)
 })
 
+
 ATT_stacked <- lapply(models_stacked, function(x){
   c(x$fml[[2]],
     round(x[["coefficients"]][["treated_unit:rel_p1"]],5), 
@@ -58,7 +62,12 @@ ATT_stacked <- lapply(models_stacked, function(x){
 })
 
 stacked_values <- as.data.frame(do.call(rbind, ATT_stacked)) %>%
-  dplyr::rename(Variable=V1, ATT=V2, SE=V3, Specification=V4)
+  dplyr::rename(Variable=V1, ATT=V2, SE=V3, Specification=V4) %>%
+  mutate(Variable=ifelse(Variable=="retire", "Retire",Variable),
+         Variable=ifelse(Variable=="pos_office", "Frac. Patients in Office",Variable),
+         Variable=ifelse(Variable=="work_in_office", "Work in Office",Variable),
+         Variable=ifelse(Variable=="npi_unq_benes", "Number Patients",Variable),
+         Variable=ifelse(Variable=="claim_per_patient", "Claims per Patient",Variable))
 
[…]

### 2022-10-31 Scripts/specification_chart.R

@@ -835,14 +835,16 @@ labels <- list("Perc. Inpatient Threshold: "=c("50% ", "70% ", "90% "),
 mu  <- mean(chart_data_frame$coef) # mean of all coefficients
 sd  <- sd(chart_data_frame$coef)
 
+
+
 schart(chart_data_frame, labels, highlight=117, 
        order="increasing", heights=c(1,.75),
-       band.ref=c(mu+sd, mu-sd),
+       band.ref=c(mu+(1.96*sd), mu-(1.96*sd)),
        col.band.ref="#E69F00",
        col.est=c("grey70","#009E73"),
        col.dot=c("grey70", "grey90", "#009E73", "#009E73"),
        ylab="Coefficient and 95% C.I.")
-legend("bottomright", lwd=2:2, col=c("#E69F00", "#009E73"), c("Average Conf. Interval", "Main Specification"), inset=.02)
+legend("bottomright", cex=.8, lwd=2:2, col=c("#E69F00", "#009E73"), c("Average Conf. Interval", "Main Specification"), inset=.02)
 pdf(file="retire_chart.pdf", )
 
 
@@ -901,12 +903,12 @@ sd  <- sd(chart_data_frame$coef)
 
 schart(chart_data_frame, labels, highlight=117, 
        order="increasing", heights=c(1,.75),
-       band.ref=c(mu+sd, mu-sd),
+       band.ref=c(mu+(1.96*sd), mu-(1.96*sd)),
        col.band.ref="#E69F00",
        col.est=c("grey70","#009E73"),
        col.dot=c("grey70", "grey90", "#009E73", "#009E73"),
        ylab="Coefficient and 95% C.I.")
-legend("bottomright", lwd=2:2, col=c("#E69F00", "#009E73"), c("Average Conf. Interval", "Main Specification"), inset=.02)
+legend("bottomright", cex=.8, lwd=2:2, col=c("#E69F00", "#009E73"), c("Average Conf. Interval", "Main Specification"), inset=.02)
 
 
 
@@ -963,12 +965,12 @@ sd  <- sd(chart_data_frame$coef)
 
 schart(chart_data_frame, labels, highlight=117, 
        order="increasing", heights=c(1,.75),
-       band.ref=c(mu+sd, mu-sd),
+       band.ref=c(mu+(1.96*sd), mu-(1.96*sd)),
        col.band.ref="#E69F00",
        col.est=c("grey70","#009E73"),
[…]

### 2022-10-31 Paper/Paper.tex

@@
     Technological innovation has affected workers and labor markets for decades.
-I add to the understanding of technology implementation's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
+I add to our understanding of technology's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
 health care, electronic health records (EHRs).
@@
 I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure.
-Further, for physicians who remain exposed to the technology, the number of patients seen increases.
-These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicates potential productivity gains for those who embrace the technology.
+Further, physicians exposed to EHRs see more patients than those who are not.
+These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicate potential productivity gains for those who embrace the technology.
 \end{abstract}
@@
 Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition.
-The daily tasks a physician faces change drastically when a new system is put in place.
-Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
+Daily tasks change drastically when a new system is put in place; some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels.
 Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}).
@@
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Further, the cost imposed is likely to vary based on the type of worker, in particular, age group.
-Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
+Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices, and I highlight the differences in effects for older vs.
+younger physicians.
 
 I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
-Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital.
-Then, I use the American Hospital Association (AHA) Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
+Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs.
+Then, I use the American Hospital Association (AHA) Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
 Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist.
-For all outcomes, the independent variable of interest is a binary treatment variable capturing exposure to an EHR.
+The independent variable of interest is a binary treatment variable capturing exposure to an EHR.
 I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.
 
-This paper contributes to the literature's understanding of how health information technology generally affects health care.
+This paper contributes to the literature's understanding of how health information technology affects health care.
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
-More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings.
+Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+More relevant to individual physicians' productivity, a number of case studies consider the effect of EHRs on productivity.
 One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).
 
-I expand this literature, first by considering a different mechanism by which EHRs could affect health care through physicians, which, to my knowledge, has not yet been examined.
+I expand this literature, first by considering physicians as the mechanism by which EHRs do not affect health care, which, to my knowledge, has not yet been examined.
 This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups.
 Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered ``meaningful" and are likely affecting physicians' daily life.
+This provides sufficient variation in the timing of implementation and ensures the EHRs are meaningful and are likely affecting physicians' daily life.
 Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework.
@@
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this result, I find that EHR implementation led to a 0.006 ppt increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group.
-Physicians less than 60 were 0.002 ppts more likely to retire after exposure, possibly pointing to career switching.
-These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians.
-For those who did not retire, I find that young physicians were .06 ppts (15\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure.
+Contrary to this result, I find that EHR implementation led to a 20\% increase relative to the average retirement for this age group.
+Physicians less than 60 were 7\% more likely to retire after exposure, possibly pointing to career switching.
+For those who did not retire, I find that physicians were 30\% more likely to see patients in an office setting the same year as EHR exposure.
 I find that patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.
@@
 To test this assumption, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
-If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence.
+If data assistants are used to ease the burden of EHR use for physicians, the results are at least partially driven by their existence.
 I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast amount of EHR implementation.
-Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
+My main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
 Second, I assume the decision for hospitals to implement EHRs is exogenous to individual physicians.
@@
 For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a low likelihood that EHR implementation is endogenous.
-The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.
+The results of the additional analyses indicate that for most outcomes, my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.
+
+The results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians.
+Physicians are also likely to change their behavior due to adverse work environments.
+Given the unintended consequences found here, policymakers should carefully consider how workers may respond to policies that affect their work environment.
 
@@
 Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation.
-As technological innovation made computers more portable, the usability of EHRs increased, creating what is known as the ``physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
+As innovation made computers more portable, the usability of EHRs increased, creating what is known as the ``physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates.
 Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
 Automation in data entry was non-existent, making it extremely time consuming for the user.
-Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry (\cite{evans2016electronic}).
 
 The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines.
-The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy.
+The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and ensuring privacy.
 The U.S.
-allocated subsidies according to stages of "meaningful" use: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, and stage 3 (2014) extended to data sharing between practices.
+allocated subsidies according to stages: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, stage 3 (2014) extended to data sharing between practices.
 This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
@@
 
-Figure \ref{fig:EPIC} shows two computer screenshots of what physicians might see in their EHR system.
-On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}).
-These are the most time consuming portions of EHR use, according to the same study.
-In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR.
+On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively which are the most time consuming portions of EHR use (\cite{arndt2017tethered}).
+In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR, an example of which is shown in Figure \ref{fig:EPIC}.
 Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use.
@@
 A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages meant for other members of a care team, and searching relevant information (\cite{dymek2021building}).
-Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or even longer to load.
+Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or load.
 Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
-Physicians have been voicing concerns over EHR usability since their beginning, indicating that they are likely responding in their behavior as well.
 
@@
 
-This section serves to outline the potential responses of physicians who work primarily in hospitals to EHRs based on the costs and incentives they face.
+This section serves to outline the potential responses of hospitalists to EHRs based on their incentives.
 The physician behaviors that I consider in this analysis are closely related and should not be taken as independent.
-For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting.
+For example, a physician's choice of work setting depends on their decision to prolong retirement, and billing activity depends on work setting.
 Thus, I discuss the behaviors in stages of decision making.
@@
 Workers in developed countries tend to plan for formal retirement well in advance.
-Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of wealth and planning necessary to formally leave the labor force.
+Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of planning necessary to formally leave the labor force.
 Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}).
@@
 It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
-First, I consider physicians already of retirement age when a new technology is implemented.
-While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire.
-Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life factors.
-Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction.
+First, I consider physicians already of retirement age ($>=60$) when a new technology is implemented.
+While payment schemes, wealth, and retirement benefits certainly impact retirement since, on average, those in this age group have already prepared financially to retire.
+Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility depends on job satisfaction and personal life factors.
+Exogenous EHR implementation affects the decision to retire through job satisfaction.
 If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work.
@@
 Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
-Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
+Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
 
@@
 While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching.
-This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction of current clinical work.
+This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction.
 As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
-Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
+Thus, the term retirement for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
 A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
@@
 
-For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of being in the labor market.
+For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of working in clinical settings.
 However, there are other ways physicians might change their behavior under drastic changes to their current working environment.
-For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over which technologies they use.
-Generally, physicians may have the ability to change hospitals, move from a hospital to an office, or work in multiple facilities and change the amount of work done in different facilities.
-I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists.
-\footnote{For a detailed discussion of included physicians, see Appendix \ref{app:data}} Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
+For those who remain practicing, changing the physical place of work may allow the physician more control over which technologies they use.
+Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
 While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.
@@
 When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.
-Thus, I classify these physicians as "sticky", since they are not able to change behavior in the short run.
+Thus, I classify these physicians as ``sticky", since they are not able to change behavior in the short run.
 
@@
 
-There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation.
+There are still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation.
 A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
-For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment.
-An important aspect of the implementation of EHRs is whether or not they allow users to be more productive.
-A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
 
@@
 However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
-
 Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician.
@@
 However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient.
-Thus, both productivity and efficiency are empirical questions.
 
@@
 
-Using various linked data sets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, outcomes related to physician practice location and billable activity, and other relevant physician, hospital, and practice characteristics.
-I describe the different data sets used to construct the panel below.
-I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
+I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labormarket outcomes, and other relevant physician, hospital, and practice characteristics.
+I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
 
@@
 The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017.
-Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts, and fraction of patients in different settings such as office, inpatient or emergency room.
-I first limit the sample of physicians based on specialty type, since my analysis relies on EHR implementation in hospitals.
+Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings.
 I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
-However, I exclude physicians who list themselves as a specialist in all but one year.
 Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years, and I drop physicians with insufficient claim counts to show indications of behavior.
-
-This data includes information on physicians which I use to construct the dependent variables for my analysis.
-Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen, and overall billing activity.
+I use this data to construct the dependent variables for my analysis.
+Generally, these include the decision to retire (leave clinical work), switch to a different work setting, number of patients seen, and overall billing activity.
 I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.
 
-CMS Shared Patient data records annual information on count of Medicare patients and frequency of billing associated with the same patient across providers within a given time period.
+CMS Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period.
 These data are available from 2009 to 2015.
@@
 I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS.
-I am specifically interested in primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital.
 The types of physicians are already limited to primary care, but there still may be office-based primary care physicians in the shared patient data.
 Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data.
-I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year, which gives a threshold of work relationship without excluding physicians who have no claims for a number of years.
-
-Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
+I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year.
+
+Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the Shared Patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
 I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
-A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", ``Not at all".
+A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", or ``Not at all".
 Since partial implementation of an EHR does not guarantee sufficient use, I define a hospital as implementing an EHR the first year they answer ``Fully".
 I then aggregate this data to the physician level.
-Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017.
-That is, the data contains no units which are "never-treated".
+Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015, as I do not observe whether they become treated in 2016 or 2017.
+That is, the data contains no units which are ``never-treated".
 
 Finally, I include a physician's graduation year from Physician Compare.
-I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.
+I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
+The details of such thresholds are explained in Appendix \ref{app:data}.
+To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations.
+The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.
 
@@
 
-The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I loosely call retirement. This variable is constructed as follows: 
+The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as retirement. This variable is constructed as follows: 
 $$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
@@
 Next, I consider whether physicians exhibit behavior that suggests they have shifted setting or location of work.
-First, I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
-Second, I construct an indicator variable equal to one if a physician sees a positive fraction of patients in an office-based setting.
-This binary measure of practice setting offers less variation but captures a more discrete change in which physicians enter or leave the office-based setting entirely.
-Finally, I consider outcome variables which explore inputs to overall efficiency in health care, the number of patients seen and claims filed per patient.
+I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
+Then, I construct an indicator variable equal to one if a physician sees any number of patients in an office-based setting, which offers less variation but captures the extensive margin of working in an office.
+Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.
 
@@
 
-I show summary statistics for the entire sample in Table \ref{tab:sumstats}.
+I show summary statistics of relevant variables for the entire sample in Table \ref{tab:sumstats}.
 Only 3\% of physicians (approximately 770) retire over the course of the panel.
@@
 
-I also include a table of means comparing physicians younger than 60, at least 60, and physicians who ever retire (who can fall in either age category) in Table \ref{tab:splitstats}.
+I also include a table of means comparing physicians below and above retirement age in Table \ref{tab:splitstats}.
 Senior physicians see more patients and work in more hospitals on average than younger physicians.
-These older physicians also see a larger fraction of patients in an office setting and are more likely to work in an office in general.
-Finally, physicians in the higher age bracket are less likely to switch zip codes.
-The age groups are exposed to EHRs in similar ways.
+They also see a larger fraction of patients in an office setting and are more likely to work in an office in general.
+The age groups are exposed to EHRs similarly.
 
@@
 That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
-Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by then.
-The black line represents the average fraction of a physician's hospitals which are utilizing EHRs over time.
-This shows that, along with becoming first exposed over the sample period, physicians are also seeing EHRs implemented in more of their own hospitals over time.
+Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by the end of the sample period.
+The darker line shows the underlying variation in this variable, hospitals implementing EHRs over time.
 
@@
 
-For each outcome variable described in Section \ref{sec:outcome}, I treat EHR implementation as an exogenous treatment variable, and I seek to estimate average treatment effects of EHR exposure.
+I treat EHR implementation as an exogenous treatment variable, and I estimate average treatment effects of EHR exposure on the outcomes discussed.
 In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
@@
 
-Once I estimate the $ATT(g,t)$ for each group and year, I aggregate the estimates over groups to a more familiar event study plot.
+I aggregate the group time estimates over groups to a more familiar event study plot.
 For select outcomes, I present the original pre-aggregation estimates to show that results are not driven by a particular year of treatment.
 Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each results plot.
-Simultaneous confidence bands, calculated using bootstrapped standard errors, are presented.
+Simultaneous confidence bands, calculated using bootstrapped standard errors, are also presented.
 
@@
 There are several assumptions necessary to identify the parameters of interest, $ATT(g,t)$.
-First, I assume that treatment is not reversed once it occurs.
+First, I assume that treatment is not reversed.
 That is, once a physician is exposed to an EHR, they cannot be un-exposed.
@@
 Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not.
-Hospitals with this discretion are coded as implementing an EHR in the first year that they answered "fully" to the survey question.} Further, since we think of treatment at the physician level as exposure to the technology, this assumption is satisfied when thought of as whether the physician has ever used an EHR.
-Once they are exposed to the EHR for the first time, that experience cannot be forgotten.
-However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
-This is why I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
+Hospitals with this discretion are coded as implementing an EHR in the first year that they answered "fully" to the survey question.} When considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
+Thus, I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
 I discuss this limitation to the data further in Section \ref{sec:patientcount}.
@@
 Second, I assume physicians do not anticipate EHR exposure prior to occurrence.
-This may be a concern if physicians learn that the system will be implemented and change their behavior before implementation occurs.
+This may be a concern if physicians learn that the system will be implemented and change their behavior beforhand.
 Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
 Therefore, I proceed in the main specification assuming no anticipation.
-However, I explore and present results for a one year anticipation period in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change my findings.
+However, I explore and present results for a one year anticipation period in the specification charts in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change my findings.
 
@@
 In my analysis, I present p-values for a Wald test of pre-trends.
-In most cases, there is no evidence of a pre-trend.
 Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends} by allowing for specified violations in the parallel trends assumption (\cite{rambachan2019honest}).
@@
 However, the data is not ideal for more than two age groups since the sample sizes decrease dramatically for certain groups.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
-The estimates in the top panel suggest that being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in the first year after exposure, and a .005 ppt increase in the second year after exposure.
+Being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in the first year after exposure, and a .005 ppt increase in the second year after exposure.
 While numerically small, these effects are statistically and economically meaningful relative to the proportion of physicians who retire in the sample, .03.
 That is, the effect represents a 7\% (16\% in the second year) increase in the likelihood of retirement relative to the mean.
+
 Even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
-Therefore, in Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
-The results are largely similar to those presented here.
+In Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
+Due to the rarity of retirement in the sample, the estimate becomes null with any specified variation in the parallel trend assumption.
+Since the pre-trend does not look strong in the first place, this is not too concerning.
+I present estimates for the overall $ATT$ under various specifications of the data in Appendix \ref{sec:chart}.
+The estimate found here falls in the upper range of estimates.
+All estimates are positive, yet some have confidence intervals that include zero.
+Closer examination into these estimates reveals that stricter thresholds on the data yield many of the null results.
+Even with additional noise in the estimates, the average of the upper and lower confidence intervals still indicates a positive effect of EHR exposure of retirement.
+Thus, the main finding is robust to various specifications.
 
@@
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
-These results are shown in the bottom panel of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the positive estimate found above in the year after exposure, while younger physicians are driving the result for the second year after exposure.
+Interestingly, senior physicians are driving the positive estimate in the year after exposure, while younger physicians are driving the result for the second year after exposure.
 A retirement age physician is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
 In comparison, physicians less than 60 are not more likely to retire in the first year after exposure, but are approximately 16\% more likely to retire two years after exposure.
-There are two potential reasons for the different response time.
+There are two potential reasons for the different response time that relate to formal retirement vs.
+career switching.
 First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
@@
 
-A potential concern is that physicians are not leaving a clinical setting, but instead they stop seeing Medicare patients.
+A potential concern is that physicians are not leaving a clinical setting, but only see patients with private insurance.
 While EHR use should not be directly correlated with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
-CMS publishes a list of physicians who have officially opted out of seeing Medicare patients.
-None of the physicians in my sample have selected into this option.
-Further, the differential timing in the effect of older vs.
-younger physicians is reassuring, as any Medicare change should not affect physicians of different ages differently.
+CMS publishes a list of physicians who have officially opted out of seeing Medicare patients and none of the physicians in my sample have selected into this option.
+Additionally, hospitalists do not have complete control over their composition of patients.
 
@@
 
-There is still a large portion of physicians not induced to retire due to EHR exposure.
-Thus, I investigate another dimension by which physicians could change behavior: the extent to which they work in an office and whether this changes due to EHR exposure.
-I use variation in two different variables: (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, and (2) the fraction of patients a physician sees in an office in a given year.
+I investigate another dimension by which physicians could change behavior: the extent to which they work in an office and whether this changes due to EHR exposure.
+The outcomes I consider are (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, shown in Figure \ref{fig:officefirst}, and (2) the fraction of patients a physician sees in an office in a given year, shown in Figure \ref{fig:officesecond}.
 This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
 
-I first discuss the estimated effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}.
-The top panel shows aggregated group time treatment effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office by at least .04 ppts in every year after being exposed, where the effect increases over time (up to .1 ppts four years after exposure).
+EHR exposure increases the likelihood of working in an office by at least .04 ppts in every year after being exposed, where the effect increases over time (up to .1 ppts four years after exposure).
 This estimate is equivalent to a 15-38\% increase relative to the average proportion of physicians who work in an office.
-That is, physicians working in a hospital are more likely to start working in an office after being exposed to an EHR, and these effects are just as large as those on retirement.
+That is, physicians working only in a hospital are more likely to start working in an office after being exposed to an EHR, and these effects are just as large as those on retirement.
 However, unlike the estimated retirement effects, the increased likelihood of working in an office is persistent over time, which suggests that EHRs impact work environment in hospitals for several years after implementation.
+
+I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}, where the findings are similar.
+Further, I compare this finding with various specifications to data thresholds and variable definitions with an estimate chart in Appendix \ref{sec:chart}.
+I find consistently positive and significant coefficients ranging from .02-.08 percentage points.
+Thus, the main finding is on the high end of the range of estimates.
+While all estimates are positive, some variations that drive the estimate down slightly are limiting years to 2009-2015, and using the low integration definition of EHR exposure.
+I discuss this further in Appendix \ref{sec:changes}, and confirm that my finding is robust to specification changes.
 
@@
 To continue investigating how different ages may respond differently to technology, I also present estimates for physicians $<60$ and $>=60$ separately in the bottom panel of Figure \ref{fig:officefirst}.
-Unlike the choice to retire, the age groups seem to both be responding to EHR implementation in all years after exposure.
+Unlike the choice to retire, the age groups are both responding to EHR implementation in all years after exposure.
 The magnitudes for senior physicians continue to increase over time, whereas the estimates for younger physicians level out.
 However, senior physicians are more likely to work in an office than younger physicians on average, so both groups exhibit similar increases relative to the mean.
-I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}, where the findings are similar.
-
-The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in an office and hospital before EHR exposure change the amount of work they do in an office.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} As with previous outcomes, I show estimates and 95\% confidence intervals for the whole sample, as well as different age groups in Figure \ref{fig:officesecond}.
-The estimates show an increase in the fraction of patients seen in office settings in each year once the physician is exposed to an EHR.
+
+The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in both settings before EHR exposure change the amount of work they do in an office.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} The estimates show an increase in the fraction of patients seen in office settings after the physician is exposed to an EHR.
 In the year of exposure, the fraction of patients increases by .06, and by the fourth year after exposure the fraction of patients increases by .12 due to exposure.
@@
 
+I also investigate pre-trends for this outcome in Appendix \ref{sec:pretrends}, where I find that under various assumptions of the form of violation in parallel trends, the estimate remains positive and significant of similar magnitude.
+Further, under different combinations of data specifications I find consistently positive and significant results, as shown in Appendix \ref{sec:changes}.
+However, the range of magnitudes is large: .01 to .1 percentage points.
+Thus, the main finding is large relative to these other estimates.
+Both accounting for anticipation and limiting sample years push the magnitude of the estimate down.
+Taking the average of these confidence intervals shows that the positive effect of EHR exposure on the fraction of patients seen in the office is robust to many combinations of data thresholds.
+
 The increasing trend over time remains the same for physicians of different age groups.
@@
 
-The previous sections reveal that EHRs induce some physicians to either retire or work more in an office, but do not give insight to how EHRs impact those who don't change their behavior.
-Thus, as a measure of productivity/efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
+As a measure of productivity/efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
 If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR.
@@
 These estimates translate to a 6-14\% increase in the number of patients seen relative to the mean.
-Interestingly, when splitting the sample by age group, the effect decreases three and four years after exposure for younger physicians, while the effect for senior age physicians remains persistent over time, suggesting that technology's effect on productivity in the long run depends on worker type (in this case, age or familiarity with technology of the worker).
-Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends} and find similar results.
+Interestingly, when splitting the sample by age group, the effect decreases three and four years after exposure for younger physicians, while the effect for senior age physicians remains persistent over time, suggesting that technology's effect on productivity in the long run depends on worker type (in this case, age).
+
+Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends} and find that this increase in patient count is robust to violations in the parallel trends assumption.
+I also present estimates under different data specifications in Appendix \ref{sec:chart}.
+While the average of the upper and lower confidence interval values still suggest a positive effect, there are many specifications where the confidence interval contains zero, suggesting a null result.
+The main change where we see this difference is in the definition of EHR exposure.
+Using the low-integration definition to account for endogeneity yields no effect.
+This indicates that on the margin of productivity, physicians who are more likely to be involved in hospital decisions do not experience the productivity gains I find in the main analysis.
 
@@
 
-One concern with my estimates is that EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
-That is, maybe more productive or tech savvy physicians are the ones who remain in the same hospital after EHR exposure.
-When including those physicians who are never treated in any year of the sample, the proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
+It may be that that EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
+That is, if productive or tech savvy physicians are the ones who remain in the same hospital after EHR exposure, the estimates are biased.The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
 To investigate whether my results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}).
@@
 In my case, that means that EHR exposure increases the likelihood of retiring or moving workplace settings and no physicians remain in the sample under treatment who would have retired or moved in the absence of EHR exposure.
-Since I find in earlier analyses that physicians are more likely to move and retire due to EHRs, this assumption seems reasonable.
+Since I find in earlier analyses that physicians are more likely to move and retire due to EHRs, this assumption is reasonable.
 To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate.
@@
 Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
-Specifically, if a hospital loses a portion of their physicians, they may divide those patients between the remaining physicians.
-In the long run, the hospital should hire new physicians to replace those that left, but if this does not happen right away then these estimates could be driven by increased workload instead of EHR use.
-Just by observing the pattern of the estimates on patient count over time, this could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time.
+Specifically, a hospital that loses a portion of their physicians, and divides those patients between the remaining physicians.
+In the long run, the hospital should hire new physicians to replace those that left, but in the short run estimates could be driven by increased workload instead of EHR use.
+This could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time.
 Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from retired physicians were divided between remaining physicians.
@@
 In the first year after exposure, I estimate the increase in patients due to retiring physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above.
-However, the second year after exposure is when I observe that largest effect on retirement, and so I do not rule out that the increase in patient count could be caused by that.
-After two years, I find no strong evidence for an increase in retirement, so there does seem to be some positive effect on patients seen due to EHR use.
+However, the second year after exposure is when I observe that largest effect on retirement, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in physicians.
 
@@
 On average, physicians file roughly 4 claims per patient.
-I present estimates of these effect in Figure \ref{fig:claim}, which shows that EHRs tend to increase claims filed per patient over time.
+I present estimates in Figure \ref{fig:claim}, which shows that EHRs tend to increase claims filed per patient over time.
 The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as physicians are exposed to the new software, they bill for more items on average.
-There is no evidence of a pre-trend driving this result.
 I also show heterogeneity of this result by age group, and the bottom panel of Figure \ref{fig:claim} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
+
+There is no evidence of a pre-trend driving this result, and I also show robustness to violations in parallel trends in Appendix \ref{sec:pretrends}.
 I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
 The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33.
-Similarly to patient count, the effect becomes even larger when accounting for a particular type of physician being left in the sample.
+The effect becomes even larger when accounting for a particular type of physician being left in the sample.
+However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, around half of the estimates are null.
+Anticipation and limiting sample years cause the effect to disappear or even become negative.
+Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs since the effects are not necessarily robust to many specifications.
 
@@
 
-The effects found in the above analysis could be biased if there is an external event occurring that is correlated with EHR implementation and exposure.
+The effects found in the above analysis may be biased if there is an external event occurring that is correlated with EHR implementation and exposure.
 The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
@@
 Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
-The total number of these registered employees is 875, and the first data assistants ever registered were in 2005.
-From 2005 to 2013, 15-60 more employees were registered each year.
 The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
@@
 Further, I investigate the occurrence of physicians and data assistants sharing patients in the CMS Shared Patient data.
+If data assistants record patient information during appointments, they will be included on the claim to Medicare.
 I create a list of physician-years for which the hospital has a positive number of shared patients with a data assistant.
@@
 
-Another way physicians may avoid EHR exposure is to exhibit decision making power to either delay or encourage EHR implementation in hospitals.
+Another way certain physicians may avoid EHR exposure is to exhibit decision making power to either delay or encourage EHR implementation in hospitals.
 This would make the binary treatment variable endogenous, biasing the results of the main specification, as it includes physicians who are selecting into their own exposure.
-After the sample period ends, the government began penalizing hospitals for not using health information technology meaningfully, so implementation eventually occurs in most hospitals.
-However, during the sample, physicians may impact the timing of implementation.
 I address this question by limiting the sample of hospitals, only including those which are considered to have low vertical integration with physicians.
@@
 In general, group time estimates are similar to those in the main specification, with the exception of the probability of working in an office setting, which may indicate some correlation between EHR adoption in high integration settings and acquisition of physician offices.
+That mechanism is beyond the scope of this project and left to future research.
 
@@
 
-In this study, I analyze the effect of implementing a now-common technology in hospitals, electronic health records, on various physician labor market outcomes.
-This technology has been controversial and speculated as a cause of physician.
+A technology expected to revolutionize the health care system was the advancement of health information technology in the form of complex electronic health record (EHR) systems.
+While they have the potential to improve efficiency, cut costs, and improve patient outcomes, these have not come to fruition.
+While the effect of EHRs on these outcomes has been directly studied, there is a gap in understanding how physicians, the primary user of this technology, responded to their rapid implementation of EHRs.
+In this study, I analyze the effect of implementing electronic health records on various physician labor market outcomes, since EHRs have been controversial and speculated as a cause of physician burnout.
 Understanding how physicians respond to this technology is important, as physician labor markets have implications for access to care, quality of care, and costs.
 
-I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity.
+I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity to analyze this relationship.
 Using a difference-in-differences framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
@@
     \caption{Outcome: Retire}
-    \includegraphics[scale=.6]{Objects/retire_chart.pdf}
+    \includegraphics[scale=.7]{Objects/retire_chart.pdf}
     \label{fig:retire_chart}
@@
     \caption{Outcome: Work in Office}
-    \includegraphics[scale=.6]{Objects/office_ind_chart.pdf}
+    \includegraphics[scale=.7]{Objects/office_ind_chart.pdf}
     \label{fig:work_chart}
@@
     \caption{Outcome: Fraction of Patients in Office}
-    \includegraphics[scale=.6]{Objects/office_frac_chart.pdf}
+    \includegraphics[scale=.7]{Objects/office_frac_chart.pdf}
     \label{fig:fracoffice_chart}
@@
     \caption{Outcome: Patient Count}
-    \includegraphics[scale=.6]{Objects/patient_chart.pdf}
+    \includegraphics[scale=.7]{Objects/patient_chart.pdf}
     \label{fig:pat_chart}
@@
     \caption{Outcome: Claims per Patient}
-    \includegraphics[scale=.6]{Objects/claim__chart.pdf}
+    \includegraphics[scale=.7]{Objects/claim_chart.pdf}
     \label{fig:cpp_chart}
@@
 
-\begin{figure}[t!]
-\begin{subfigure}{0.48\textwidth}
-\caption{Retire}
-\includegraphics[width=\linewidth]{Objects/retire_est_plot.pdf}
-\label{fig:a1}
-\end{subfigure}\hspace*{\fill}
-\begin{subfigure}{0.48\textwidth}
-\caption{Fraction of Patients in Office}
-\includegraphics[width=\linewidth]{Objects/office_frac_est_plot.pdf}
-\label{fig:b1}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{0.48\textwidth}
-\caption{Indicator for Office }
-\includegraphics[width=\linewidth]{Objects/office_in_est_plot.pdf}
- \label{fig:c1}
-\end{subfigure}\hspace*{\fill}
-\begin{subfigure}{0.48\textwidth}
-\caption{Patient Count} 
-\includegraphics[width=\linewidth]{Objects/patient_est_plot.pdf}
-\label{fig:d1}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{0.48\textwidth}
-\caption{Claims Per Patient} 
-\includegraphics[width=\linewidth]{Objects/claim_est_plot.pdf}
-\label{fig:e1}
-\end{subfigure}\hspace*{\fill}
-
-\caption{Different Estimators} \label{fig:estimators}
+\begin{figure}
+    \centering
+    \includegraphics{Objects/estimators_plot.pdf}
+    \caption{Alternative Estimators}
+    \label{fig:estimators}
 \end{figure}

### 2022-11-07 Paper/Paper.tex
Updates from Overleaf

@@
 
-I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
+I use data on a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
 Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs.
@@
 The independent variable of interest is a binary treatment variable capturing exposure to an EHR.
-I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.
-
-This paper contributes to the literature's understanding of how health information technology affects health care.
+I estimate group-time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.
+
+This paper contributes to our understanding of how health information technology affects health care.
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
 More relevant to individual physicians' productivity, a number of case studies consider the effect of EHRs on productivity.
@@
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this result, I find that EHR implementation led to a 20\% increase relative to the average retirement for this age group.
+Contrary to this result, I find that EHR implementation led to a 20\% increase in retirement relative to the average retirement for this age group.
 Physicians less than 60 were 7\% more likely to retire after exposure, possibly pointing to career switching.
 For those who did not retire, I find that physicians were 30\% more likely to see patients in an office setting the same year as EHR exposure.
-I find that patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.
+I find that the patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.
 
@@
 
-The results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians.
-Physicians are also likely to change their behavior due to adverse work environments.
-Given the unintended consequences found here, policymakers should carefully consider how workers may respond to policies that affect their work environment.
+The results suggest that there was a surge of experienced physicians who left the labor force or hospitals and were no longer available to influence early career physicians.
+Thus, EHR implementation may have intensified access to care issues for patients.
+The unintended consequences highlight the need for careful consideration of policies that affect work environment.
 
@@
 Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry.
-Automation in data entry was non-existent, making it extremely time consuming for the user.
+Automation in data entry was non-existent, making it extremely time consuming for the user (\cite{evans2016electronic}).
 
@@
     \caption{Hospitals Receiving Meaningful Use Stage 1 Subsidy}
-    \includegraphics[scale=.6]{graphics/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
+    \includegraphics[scale=.5]{graphics/QS-Hospitals-Receiving-Payments-for-MU-and-Adoption.png}
     \caption*{Source: HealthIT.gov}
@@
 
-On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively which are the most time consuming portions of EHR use (\cite{arndt2017tethered}).
-In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR, an example of which is shown in Figure \ref{fig:EPIC}.
+On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming portions of EHR use (\cite{arndt2017tethered}).
+In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR.
 Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use.
@@
 Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
-
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.4\linewidth}
-    \caption{Screenshot of EHR System}
-    \includegraphics[scale=.4]{graphics/epic-ehr-screenshot.jpg}
-    \includegraphics[scale=.11]{graphics/EHRimage2.jpg}
-    \label{fig:EPIC}
-\end{figure}
 
@@
 Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
+Additionally, if EHRs cause physicians to substitute time away from patients and they receive less reimbursement, leisure becomes relatively cheaper, leading to retirement.
 Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
@@
 I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years, and I drop physicians with insufficient claim counts to show indications of behavior.
+Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years (I investigate this threshold in Appendix \ref{sec:changes}), and I drop physicians with insufficient claim counts to show indications of behavior.
 I use this data to construct the dependent variables for my analysis.
@@
 I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
-Then, I construct an indicator variable equal to one if a physician sees any number of patients in an office-based setting, which offers less variation but captures the extensive margin of working in an office.
+Then, I construct an indicator variable equal to one if a physician sees any positive number of patients in an office-based setting, which offers less variation but captures the extensive margin of working in an office.
 Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.
@@
     \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.4]{Objects/sum_stats_year.pdf}
     \label{fig:treatmentgraph}
@@
 That is, once a physician is exposed to an EHR, they cannot be un-exposed.
-For physicians that do not switch hospitals, this assumption is supported by both the institutional details of hospital EHR implementation and the data itself.
+For physicians that do not switch hospitals, this assumption is supported by the institutional details of hospital EHR implementation.
 An EHR is a costly technology and requires a significant amount of collaboration to implement.
@@
 They may add features or switch vendors, but do not reverse EHR use.
-This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a 2\% random sample of hospitals over time.\footnote{Note that out of the 81 hospitals included in the random sample, 4 of them show a reversal of treatment status.
-Upon further inspection, each of these hospitals show such a reversal due to a discrepancy in the coding of whether the hospital uses an EHR partially or fully.
-Those who use the EHR partially are coded as not fully implementing an EHR since it is not clear whether physicians in the sample would be exposed yet or not.
-Hospitals with this discretion are coded as implementing an EHR in the first year that they answered "fully" to the survey question.} When considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
+However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
 Thus, I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
 I discuss this limitation to the data further in Section \ref{sec:patientcount}.
-
-\begin{figure}[ht!]
-    \centering
-    \captionsetup{width=.7\linewidth}
-    \caption{Hospital EHR Adoption Over Time}
-    \includegraphics[scale=.5]{Objects/hosp_treat.pdf}
-    \label{fig:hosp_treat}
-\end{figure}
 
@@
 However, the data is not ideal for more than two age groups since the sample sizes decrease dramatically for certain groups.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
-Being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in the first year after exposure, and a .005 ppt increase in the second year after exposure.
+Being exposed to an EHR leads to a .2 ppt increase in the likelihood of retiring in the first year after exposure, and a .5 ppt increase in the second year after exposure.
 While numerically small, these effects are statistically and economically meaningful relative to the proportion of physicians who retire in the sample, .03.
@@
 
+\begin{figure}[H]
+\centering
+   \begin{subfigure}[b]{0.3\textwidth}
+   \includegraphics[width=0.8\textwidth]{retire_plot_ages}
+   \caption{}
+   \label{fig:Ng1} 
+\end{subfigure}
+
+\\
+
+\begin{subfigure}[b]{0.3\textwidth}
+   \includegraphics[width=0.8\textwidth]{Nvariousg2}
+   \caption{}
+   \label{fig:Ng2}
+\end{subfigure}
+\caption{(a) Numerical solutions for the small-time system with a constant-curvature body shape showing the scaled leading-order veritcal reaction force $N_0$ versus the scaled body mass $M$ for various values of $g$.
+Again, $I=M$ for definiteness and $A=0.7$.
+(b) As for (a) but over a wider range of values of $M,I$.}
+\end{figure}
+
 \begin{figure}[ht!]
@@
     \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.6]{Objects/retire_plot.pdf}
+    \includegraphics[scale=.5]{Objects/retire_plot.pdf}
     \label{fig:retirefirst}
@@
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
-Interestingly, senior physicians are driving the positive estimate in the year after exposure, while younger physicians are driving the result for the second year after exposure.
-A retirement age physician is .01 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
+Interestingly, senior physicians are driving the positive estimate in the year after exposure, and younger physicians only retire in the second year after exposure.
+A retirement age physician is 1 ppt more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
 In comparison, physicians less than 60 are not more likely to retire in the first year after exposure, but are approximately 16\% more likely to retire two years after exposure.
@@
 
-EHR exposure increases the likelihood of working in an office by at least .04 ppts in every year after being exposed, where the effect increases over time (up to .1 ppts four years after exposure).
+EHR exposure increases the likelihood of working in an office by at least 4 ppts in every year after being exposed, where the effect increases over time (up to 10 ppts four years after exposure).
 This estimate is equivalent to a 15-38\% increase relative to the average proportion of physicians who work in an office.
@@
 Further, I compare this finding with various specifications to data thresholds and variable definitions with an estimate chart in Appendix \ref{sec:chart}.
-I find consistently positive and significant coefficients ranging from .02-.08 percentage points.
+I find consistently positive and significant coefficients ranging from 2-8 percentage points.
 Thus, the main finding is on the high end of the range of estimates.
@@
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.6]{Objects/officeind_plot.pdf}
+    \includegraphics[scale=.5]{Objects/officeind_plot.pdf}
     \label{fig:officefirst}
@@
     \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
-    \includegraphics[scale=.6]{Objects/officefrac_plot.pdf}
+    \includegraphics[scale=.5]{Objects/officefrac_plot.pdf}
     \label{fig:officesecond}
@@
 Further, under different combinations of data specifications I find consistently positive and significant results, as shown in Appendix \ref{sec:changes}.
-However, the range of magnitudes is large: .01 to .1 percentage points.
+However, the range of magnitudes is large: 1 to 10 percentage points.
 Thus, the main finding is large relative to these other estimates.
@@
     \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.6]{Objects/patient_plot.pdf}
+    \includegraphics[scale=.5]{Objects/patient_plot.pdf}
     \label{fig:patient}
@@
     \caption{Effect of EHR Exposure on Patient Count by Group}
-    \includegraphics[scale=.47]{Objects/patient_group.pdf}
+    \includegraphics[scale=.4]{Objects/patient_group.pdf}
     \label{fig:patientgroup}
@@
     \caption{Effect of EHR Exposure on Claims per Patient}
-    \includegraphics[scale=.6]{Objects/claim_per_patient_plot.pdf}
+    \includegraphics[scale=.5]{Objects/claim_per_patient_plot.pdf}
     \label{fig:claim}
@@
 \caption{Frequency of Data Assistant Enumeration by Year}
-\includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
+\includegraphics[scale=.4]{Objects/dataassistant_histogram.pdf}
 \label{fig:dataassistant_histogram}
@@
     \caption{Outcome: Retire}
-    \includegraphics[scale=.7]{Objects/retire_chart.pdf}
+    \includegraphics[scale=.6]{Objects/retire_chart.pdf}
     \label{fig:retire_chart}
@@
     \caption{Outcome: Work in Office}
-    \includegraphics[scale=.7]{Objects/office_ind_chart.pdf}
+    \includegraphics[scale=.6]{Objects/office_ind_chart.pdf}
     \label{fig:work_chart}
@@
     \caption{Outcome: Fraction of Patients in Office}
-    \includegraphics[scale=.7]{Objects/office_frac_chart.pdf}
+    \includegraphics[scale=.6]{Objects/office_frac_chart.pdf}
     \label{fig:fracoffice_chart}
@@
     \caption{Outcome: Patient Count}
-    \includegraphics[scale=.7]{Objects/patient_chart.pdf}
+    \includegraphics[scale=.6]{Objects/patient_chart.pdf}
     \label{fig:pat_chart}
@@
     \caption{Outcome: Claims per Patient}
-    \includegraphics[scale=.7]{Objects/claim_chart.pdf}
+    \includegraphics[scale=.6]{Objects/claim_chart.pdf}
     \label{fig:cpp_chart}
@@
     \caption{Results: Two Way Fixed Effects}
-    \includegraphics[scale=.8]{Objects/twfe_plot.pdf}
+    \includegraphics[scale=.6]{Objects/twfe_plot.pdf}
     \label{fig:twfe}
@@
     \centering
-    \includegraphics{Objects/estimators_plot.pdf}
     \caption{Alternative Estimators}
+    \includegraphics[scale=.6]{Objects/estimators_plot.pdf}
     \label{fig:estimators}
@@
 those who typically work at large hospitals.
-These average treatment effects are shown in Figure \ref{fig:1}.
+These average treatment effects are shown in Figure \ref{fig:heter}.
 The results do not differ along these dimensions.
@@
 
-\begin{figure}[t!]
-\begin{subfigure}{0.48\textwidth}
-\caption{Retire}
-\includegraphics[width=\linewidth]{Objects/retire_heterog.pdf}
-\label{fig:a}
-\end{subfigure}\hspace*{\fill}
-\begin{subfigure}{0.48\textwidth}
-\caption{Fraction of Patients in Office}
-\includegraphics[width=\linewidth]{Objects/office_frac_heterog.pdf}
-\label{fig:b}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{0.48\textwidth}
-\caption{Indicator for Office }
-\includegraphics[width=\linewidth]{Objects/office_ind_heterog.pdf}
- \label{fig:c}
-\end{subfigure}\hspace*{\fill}
-\begin{subfigure}{0.48\textwidth}
-\caption{Patient Count} 
-\includegraphics[width=\linewidth]{Objects/patient_heterog.pdf}
-\label{fig:d}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{0.48\textwidth}
-\caption{Claims Per Patient} 
-\includegraphics[width=\linewidth]{Objects/claim_per_patient_heterog.pdf}
-\label{fig:e}
-\end{subfigure}\hspace*{\fill}
-
-\caption{Heterogeneity Analysis} \label{fig:1}
+\begin{figure}
+    \centering
+    \caption{Heterogeneity Analysis}
+    \includegraphics[scale=.6]{Objects/heterog_plot.pdf}
+    \label{fig:heter}
 \end{figure}

### 2022-11-07 Paper/Paper.tex
Updates from Overleaf

@@
 
-\begin{figure}[H]
-\centering
-   \begin{subfigure}[b]{0.3\textwidth}
-   \includegraphics[width=0.8\textwidth]{retire_plot_ages}
+\begin{figure}[ht!]
+\centering
+    \captionsetup{width=.85\linewidth}
+    \caption{Effect of EHR Exposure on Retirement}
+   \begin{subfigure}[]{.8\textwidth}
+   \includegraphics[scale=.5]{Objects/retire_plot_all.pdf}
    \caption{}
@@
 
-\\
-
-\begin{subfigure}[b]{0.3\textwidth}
-   \includegraphics[width=0.8\textwidth]{Nvariousg2}
+\begin{subfigure}[]{.8\textwidth}
+   \includegraphics[scale=.5]{Objects/retire_plot_ages.pdf}
    \caption{}
@@
 \end{subfigure}
-\caption{(a) Numerical solutions for the small-time system with a constant-curvature body shape showing the scaled leading-order veritcal reaction force $N_0$ versus the scaled body mass $M$ for various values of $g$.
-Again, $I=M$ for definiteness and $A=0.7$.
-(b) As for (a) but over a wider range of values of $M,I$.}
-\end{figure}
-
-\begin{figure}[ht!]
-    \centering
-    \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Retirement}
-    \includegraphics[scale=.5]{Objects/retire_plot.pdf}
-    \label{fig:retirefirst}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot.
-The bottom show these results for different subgroups of physicians by age.
+\vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Panel (a) shows average group time treatment effects aggregated over groups to an event study plot.
+Panel (b) shows these results for different subgroups of physicians by age.
 The p-value listed for each graph corresponds to a Wald test for pre-trends.
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0009), 0.002 (0.001), 0.006 (0.003), respectively.}}
+ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0009), 0.002 (0.001), 0.006 (0.003), respectively.}}
 \end{figure}

### 2023-04-20 Scripts/analysis.R
add HIMSS data and adjust graphs, summary stats

@@ -19,11 +19,9 @@ library(extrafont)
 # In this portion, I only consider "retire" as the dependent outcome and I compare estimators. 
 
 # Read in the main data created in "data5_MDPPAS.R"
-Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
+Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data2.rds"))
 
 Physician_Data <- Physician_Data %>%
-  mutate(npi_unq_benes_noDA=npi_unq_benes) %>%
-  mutate(npi_unq_benes_LI=npi_unq_benes) %>%
   mutate(DocNPI=as.double(DocNPI))
 
 
@@ -32,17 +30,15 @@ Physician_Data <- Physician_Data %>%
 
 varlist <- list("retire", "pos_office", "work_in_office", "npi_unq_benes", "claim_per_patient")
 
-
-
 # Get results for ATTGT 
 models <- lapply(varlist, function(x) {
   all <- att_gt(yname = x,                # LHS Variable
-    gname = if (x=="npi_unq_benes_LI") "minyr_EHR_int" else "minyr_EHR",             # First year a unit is treated. (set to 0 if never treated)
+    gname = "minyr_EHR",             # First year a unit is treated. (set to 0 if never treated)
     idname = "DocNPI",               # ID
     tname = "year",                  # Time Variable
     # xformla = NULL                 # No covariates
     xformla = ~grad_year,            # Time-invariant controls
-    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="pos_office_prior" | x=="pos_office_noprior" | x=="work_in_office" | x=="change_zip" | x=="pos_opd") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else if (x=="npi_unq_benes_noDA") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1 & ever_Phys_DA==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
+    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="work_in_office") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
[…]

### 2023-04-20 Scripts/data4_physician_level.R

@@ -29,7 +29,7 @@ data <- data %>%
 PhysCompare <- read_csv(paste0(raw_data_path,"DAC_NationalDownloadableFile.csv"))
 
 PhysCompare <- PhysCompare %>%
-  dplyr::rename(gender=gndr,grad_year=Grd_yr) %>%
+  dplyr::rename(gender=gndr,grad_year=`Grd_yr`) %>%
   select(NPI,grad_year) 
 
 # Merge, then drop any physicians that graduated after 2005
@@ -53,7 +53,7 @@ data <- data %>%
 num_hosp <- data %>% ungroup() %>%
   filter(!is.na(AHAID)) %>%
   distinct(HospNPI)
-  # There are 4241 unique hospitals in the data that are not missing AHAID (good)
+  # There are 4267 unique hospitals in the data that are not missing AHAID (good)
 
 # Only keep pairs with AHA hospitals since that is where EHR information comes from
 data <- data %>%
@@ -64,11 +64,8 @@ data <- data %>%
 # I need to think about adding more hospital characteristics to this dataset
 AHAmainsurvey <- read_csv(paste0(raw_data_path,"AHA_mainsurvey.csv"))
 
-# I define low integration and high integration as in Madison (2004, HSR)
 AHAmainsurvey <- AHAmainsurvey %>%
   mutate(ID=as.character(ID)) %>%
-  mutate(low_integration=ifelse(IPAHOS==1 | OPHOHOS==1 | CPHOHOS==1,1,0),
-         high_integration=ifelse(GPWWHOS==1 | MSOHOS==1 | ISMHOS==1 | EQMODHOS==1,1,0)) %>%
   dplyr::rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
          joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
   select(-IPAHOS, -OPHOHOS, -CPHOHOS, -GPWWHOS, -MSOHOS, -ISMHOS, -EQMODHOS) %>%
@@ -79,25 +76,88 @@ data <- data %>% ungroup() %>%
   mutate(AHAID=as.character(AHAID)) %>%
   left_join(AHAmainsurvey,by=c("AHAID"="ID","year"),na_matches="never")
 
-
+# Read in HIMSS data for additional EHR information
+for (t in 2009:2017){
[…]

### 2023-04-20 Scripts/data5_MDPPAS.R

@@ -17,7 +17,7 @@ library(lubridate)
 for (i in 2009:2017) {
   year <- read_csv(paste0(raw_data_path,"/MDPPAS/PhysicianData_",i,".csv"))
   year <- year %>%
-    dplyr::select(-name_middle, -name_first, -spec_broad, -spec_prim_1, -spec_prim_2,
+    dplyr::select(-name_middle, -name_first, -spec_broad, -spec_prim_2,
                   -phy_zip7, -claim_count7, -phy_zip8, -claim_count8, -phy_zip9, -claim_count9,
                   -phy_zip10, -claim_count10, -phy_zip11, -claim_count11, -phy_zip12, 
                   -claim_count12, -spec_prim_2_name, -tin1_legal_name, -tin2_legal_name,
@@ -29,7 +29,7 @@ MDPPAS <- rbind(MDPPAS2009, MDPPAS2010, MDPPAS2011, MDPPAS2012, MDPPAS2013, MDPP
 
 
 # Read in Aggregated Pairs data from "data4_physician_level.R"
-Aggregated_Pairs <- read_rds(paste0(created_data_path,"Aggregated_Pairs.rds"))
+Aggregated_Pairs <- read_rds(paste0(created_data_path,"Aggregated_Pairs2.rds"))
 
 
 # Merge MDPPAS to main dataset and fill time invariant variables
@@ -58,8 +58,8 @@ Physician_Data <- Physician_Data %>%
 Physician_Data <- Physician_Data %>%
   dplyr::group_by(DocNPI) %>%
   dplyr::mutate(max=max(pos_inpat,na.rm=T)) %>%
-  dplyr::filter(max>.7)
-  # 214k obs
+  dplyr::filter(max>.7) 
+  # 240k obs
 
 
 ## Read in Medicare Opt-Out Data
@@ -192,101 +192,6 @@ Physician_Data <- Physician_Data %>%
   mutate(majority_in_office=ifelse(pos_office>pos_inpat,1,0))
 
 
-# ZIP CODES ####
-Physician_Data <- Physician_Data %>% dplyr::ungroup() %>%
-  dplyr::mutate(multiple_zip=ifelse(is.na(phy_zip2),0,1),
-                claim_count1=ifelse(is.na(claim_count1),0,claim_count1)) 
-
-for (i in 1:9){
-zipyear <- Physician_Data %>%
-  dplyr::filter(year==2008+i) %>%
[…]

### 2023-04-20 Scripts/sumstats.R

@@ -22,38 +22,63 @@ cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"
 # The data used is created in "Final_Pairs.R"
 
 # Read in Final_Pairs_Variables.rds
-Physician_Data <- read_rds(paste0(created_data_path,"Physician_data.rds"))
+Physician_Data <- read_rds(paste0(created_data_path,"Physician_data2.rds"))
 
 
 
 # General Summary Stats Tables:  ----------------------------------------------------------------
 
 # Physician Level
-sum_stats_fullsample <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
+sum_stats_exit <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0) %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
-                 "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
-                 "Exposure to an EHR"="anyEHR_exposed",
-                 "Fraction Patients in Office"="pos_office",
-                 "Ever Retire"="ever_retire",
-                 "Work in an Office"="work_in_office",
-                 "Change Zip Codes"="change_zip",
-                 "Claims per Patient"="claim_per_patient"), 
-               list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))), na.rm=TRUE) %>%
+                 "Year of EHR Exposure"="minyr_EHR",
+                 "Ever Exit Clinical Work"="ever_retire"), 
+               list(m=mean,sd=sd,min=min,max=max,n=~sum(!is.na(.))/9), na.rm=TRUE) %>%
   mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
   gather(key=var,value=value) %>%
   extract(col="var",into=c("variable", "statistic"), regex=("(.*)_(.*)$")) %>%
   spread(key=statistic, value=value) %>%
   relocate(variable,n,m,sd,min,max) 
 
+sum_stats_office <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0 & ever_retire==0) %>%
+  summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
[…]

### 2023-04-20 Paper/PaperEdit.tex
Updates from Overleaf

--- 2022-11-07 Paper/Paper.tex
@@
 \begin{abstract}
-    Technological innovation has affected workers and labor markets for decades.
-I add to our understanding of technology's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
+    Technological innovation has affected labor markets drastically over time.
+I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S.
 health care, electronic health records (EHRs).
-I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes.
-I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure.
-Further, physicians exposed to EHRs see more patients than those who are not.
-These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicate potential productivity gains for those who embrace the technology.
+I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital, and estimate average group time treatment effects on various labor market outcomes.
+I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
+Further, physicians exposed to EHRs see more patients due to the technology.
+These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
 \end{abstract}
@@
 \section{Introduction}
-For decades, technological innovation has brought about changes to production in the economy, including wages, demand for labor, and other aspect of labor markets.
-Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved with the implementation of new technologies.
+For decades, technological innovation has brought about changes to production in the economy through adjustments in wages, demand for labor, and other aspects of labor markets.
+Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved in the implementation of new technologies.
 This is particularly important in health care, where new technologies can impact patient outcomes.
-A recent technology roll-out that altered the U.S.
-health care system dramatically was the electronic health record (EHR).
-The use of EHRs was incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
-Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and labor market behaviors.
-
-EHRs are computer systems that house digital medical records and have additional capabilities, including decision making assistance.
-EHRs have become increasingly relevant in the U.S.
-since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech})\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+A recent technology roll-out that altered the US health care system dramatically was the electronic health record (EHR).
+Medical entities nationwide implemented EHRs rapidly from 2008-2015, with the percentage of hospitals with basic EHR capability increasing from 9 percent to 84 percent.
+Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and individual labor market behaviors.
+In this paper, I investigate the effect of EHR adoption in hospitals on several behaviors of physicians working in hospitals, or hospitalists.
+
+EHRs are computer systems that house digital medical records and sometimes have additional functionalities.
+EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech})\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-The desired movement towards digitization was realized; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
 
@@
 Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}).
-The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings.
+When looking solely at hospitalists, 63\% of survey respondents found EHRs time consuming and inefficient (\cite{czernik2022hospitalist}).
+The frustration of a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings.
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices, and I highlight the differences in effects for older vs.
-younger physicians.
-
-I use data on a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017.
-Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs.
-Then, I use the American Hospital Association (AHA) Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use.
-Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist.
-The independent variable of interest is a binary treatment variable capturing exposure to an EHR.
-I estimate group-time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.
-
-This paper contributes to our understanding of how health information technology affects health care.
+For a hospitalist, the hospital's decision to adopt an EHR is plausibly exogenous.
+Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical environments, measured based on no future claims, (2) substituting hospital work for office work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.
+
+While hospitalists are ideal for to answer this question econometrically, they are also important in their own right.
+The Bureau of Labor Statistics estimates approximately 44,000 physicians employed as hospitalists in the US, and studies estimate that 75\% of hospitals employ hospitalists (\cite{bls}, \cite{healthecareers_2022}).
+Further, one third of Medicare admissions are seen by a hospitalist (\cite{messler2015history}).
+I construct a panel of hospitalists spanning from 2009-2017.
+Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify the hospital affiliation of each hospitalist.
+Then, I use the American Hospital Association (AHA) Survey to determine EHR adoption status, and thus connect hospitalists to EHR use.
+Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for information on the patients seen by each hospitalist.
+The independent variable of interest is binary, capturing exposure to an EHR.\footnote{While there are ample nuances to EHRs, I abstract away from these as I focus on the inital shock of a new technology.} I estimate group-time treatment effects of EHR exposure on the physician decisions defined above.
+
+This paper contributes to our understanding of health information technology and its affects in health care.
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
-More relevant to individual physicians' productivity, a number of case studies consider the effect of EHRs on productivity.
+Despite a large number of case studies that find generally positive effects, i.e., improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+A number of case studies consider the effect of EHRs on productivity.
 One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).
-
-I expand this literature, first by considering physicians as the mechanism by which EHRs do not affect health care, which, to my knowledge, has not yet been examined.
-This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups.
-Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of implementation and ensures the EHRs are meaningful and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework.
-I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects, and further provides insight to short vs.
-long term effects.
+There is still much to be learned about EHRs effects on health care, specifically related to labor supply (\cite{NBERw29218}).
+
+I expand this literature first by considering individual physician labor market decisions as an outcome, which, to my knowledge, has not yet been examined.
+Further, I consider the primary time period in which EHRs were rapidly implemented.
+This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting physicians' daily life.
+Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework.
+I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects.
 
 Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}).
-Contrary to this result, I find that EHR implementation led to a 20\% increase in retirement relative to the average retirement for this age group.
-Physicians less than 60 were 7\% more likely to retire after exposure, possibly pointing to career switching.
-For those who did not retire, I find that physicians were 30\% more likely to see patients in an office setting the same year as EHR exposure.
-I find that the patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.
+Contrary to this result, I find that EHR implementation led to a 20\% increase in the likelihood of exiting clinical settings for hospitalists of age 60 plus.
+Physicians less than 60 were 7\% more likely to exit, possibly pointing to career switching.
+Limiting the sample to those who remained in clinical settings, I find that hospitalists were 30\% more likely to start seeing patients in an office setting at the time of exposure.
+Finally, I find that EHRs lead to a patient count increase of at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.
 
 Two key assumptions underlie the main analysis.
-First, I assume that physicians are constrained to use the technology implemented in their work setting.
-To test this assumption, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants.
-If data assistants are used to ease the burden of EHR use for physicians, the results are at least partially driven by their existence.
-I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast amount of EHR implementation.
-My main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results.
-Second, I assume the decision for hospitals to implement EHRs is exogenous to individual physicians.
-That is, a physician does not choose their own EHR exposure.
-While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making.
-For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a low likelihood that EHR implementation is endogenous.
-The results of the additional analyses indicate that for most outcomes, my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.
-
-The results suggest that there was a surge of experienced physicians who left the labor force or hospitals and were no longer available to influence early career physicians.
+First, I assume that hospitalists are constrained to use the technology implemented in their work setting.
+Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals.
+That is, a hospitalist employed by the hospital does not choose their own EHR exposure.
+My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed.
 Thus, EHR implementation may have intensified access to care issues for patients.
-The unintended consequences highlight the need for careful consideration of policies that affect work environment.
+This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.
 
@@
 The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and ensuring privacy.
-The U.S.
-allocated subsidies according to stages: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, stage 3 (2014) extended to data sharing between practices.
-This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
-Figure \ref{fig:meanuse} shows a geographical comparison of U.S.
-hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
+The US allocated subsidies according to stages: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, stage 3 (2014) extended to data sharing between practices.
+Vast EHR adoption coincided with the rollout of these subsidies, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
+However, the subsidies did not necessarily cause all adoption.
+\citeauthor{dranove2015investment} (\citeyear{dranove2015investment}) estimates that adoption would have been ten percentage points lower in the absence of the policy.
+Figure \ref{fig:meanuse} shows a geographical comparison of US hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs.
 2013, revealing the nationwide and rapid expansion of technology.
@@
 
+Research on how physicians interact with EHRs is primarily focused on outpatient settings.
 On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming portions of EHR use (\cite{arndt2017tethered}).
@@
 The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}).
-A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages meant for other members of a care team, and searching relevant information (\cite{dymek2021building}).
-Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or load.
+Another common issue physicians face is the lack of usability of EHR systems, that functions take minutes to either locate or load.
 Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
-
-\subsection{EHR and Physician Labor Markets}
+While most research is specific to physicians in outpatient settings, the limited research available which investigates hospitalists and EHRs seems to be consistent with these findings (\cite{tipping2010did}).
+
+\subsection{Hospitalists and Labor Market Decisions}
 
 This section serves to outline the potential responses of hospitalists to EHRs based on their incentives.
-The physician behaviors that I consider in this analysis are closely related and should not be taken as independent.
-For example, a physician's choice of work setting depends on their decision to prolong retirement, and billing activity depends on work setting.
+The behaviors that I consider in this analysis are closely related and should not be taken as independent.
+For example, hospitalists' choice of work setting depends on their decision to remain in clinical settings, and billing activity depends on work setting.
 Thus, I discuss the behaviors in stages of decision making.
 
-\subsubsection{Retirement}
-
+\subsubsection{The Hospitalist Profession}
+
+For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting.
+However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist.
+Both formal hospitalists and internists are medical professionals who primarily focus on adult patient care, but they differ in their work settings.
+Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting.
+While this distinction is potentially important for some of the decisions outlined below, I do not have the ability to distinguish between the two explicitly in the data.
+I will discuss this limitation in more detail when relevant.
+
+\subsubsection{Exiting Clinical Work}
+
+There are two main ways for a physician to stop seeing patients altogether: retirement from the labor force in general, or changing careers to non-clinical work.
 Workers in developed countries tend to plan for formal retirement well in advance.
@@
 
-It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
-First, I consider physicians already of retirement age ($>=60$) when a new technology is implemented.
-While payment schemes, wealth, and retirement benefits certainly impact retirement since, on average, those in this age group have already prepared financially to retire.
-Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility depends on job satisfaction and personal life factors.
+Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
+As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences.
+Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors.
 Exogenous EHR implementation affects the decision to retire through job satisfaction.
@@
 Additionally, if EHRs cause physicians to substitute time away from patients and they receive less reimbursement, leisure becomes relatively cheaper, leading to retirement.
-Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).
+Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.
 
@@
 This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction.
+In a 2016 survey, 13\% of physicians stated that they were considering careers outside of clinical work because of burnout (\cite{physicians2016physicians}).
 As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
-Thus, the term retirement for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
-A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.
+Thus, the term exit for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
+A physician is said to exit in a given year only if it is the first year in which the physician has no future Medicare claims.
 
@@
 For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of working in clinical settings.
-However, there are other ways physicians might change their behavior under drastic changes to their current working environment.
-For those who remain practicing, changing the physical place of work may allow the physician more control over which technologies they use.
-Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both.
+However, there are other ways hospitalists might change their behavior under drastic changes to their current working environment.
+For those who remain practicing, changing the physical place of work may allow for more control over which technologies they use.
+Some hospitalists work under contract full time for one medical practice where they have a consistent shift schedule (formal hospitalists), while others contract themselves out to multiple hospitals, offices, or both (internists).
 While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.
@@
 This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings.
-A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set.
+A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making it very costly to change work setting or excluding it from the choice set altogether.
 Since these hospitalists are established in their hospital, they work a significant amount of time there.
 When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.
-Thus, I classify these physicians as ``sticky", since they are not able to change behavior in the short run.
 
@@
 If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.
-As this type of hospitalist is less sticky, if there is any effect of EHR exposure on work setting it would be among this type.
+Theoretically, these types of hospitalists are much more likely to change their behavior as a result of a change in one work environment.
 
@@
 
-There are still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation.
+There are still agents who are not induced to change their behavior on the basis of exit or work setting due to EHR implementation.
 A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
-For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+For hospitalists who make no physical changes to work environment after EHR exposure for any of these reasons, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
 
@@
 However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
-Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician.
+Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the hospitalist.
 Billing activity refers to the number of claims filed per patient seen.
@@
 
-I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labormarket outcomes, and other relevant physician, hospital, and practice characteristics.
+I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician, hospital, and practice characteristics.
 I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
@@
 The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017.
-Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings.
-I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year.
-Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years (I investigate this threshold in Appendix \ref{sec:changes}), and I drop physicians with insufficient claim counts to show indications of behavior.
+Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office, emergency room, etc.).
+I only include physicians whose primary specialty as hospitalist\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting.
+I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year.
+I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}).
+I drop hospitalists with low claim counts that do not accurately represent changes in behavior.
 I use this data to construct the dependent variables for my analysis.
-Generally, these include the decision to retire (leave clinical work), switch to a different work setting, number of patients seen, and overall billing activity.
+Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and overall billing activity.
 I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.
 
-CMS Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period.
+Center for Medicare and Medicaid (CMS) Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period.
 These data are available from 2009 to 2015.
@@
 
-Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the Shared Patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics.
-I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with.
-A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", or ``Not at all".
+Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the Shared Patient data to the American Hospital Association (AHA) survey and the Healthcare Information and Management Systems Society (HIMSS) survey, which both contain information on hospital-level EHR use and other characteristics.
+I record the first year a hospitalist is exposed to an EHR based on adoption in the hospitals they share patients with.
+In the AHA survey, a hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", or ``Not at all".
 Since partial implementation of an EHR does not guarantee sufficient use, I define a hospital as implementing an EHR the first year they answer ``Fully".
-I then aggregate this data to the physician level.
+In the HIMSS data, hospitals report the specific vendor they employ.
+From this, I define the first year a hospital reports using a vendor.
+I combine these measures to define the first year a hospital reports using an EHR in either survey.
+While there is an AHA survey supplement which captures more granular details about information technology, only a small subset of hospitals choose to answer this survey.
+In this paper, I focus on the initial shock of any new technology and abstract away from some of the more granular characteristics of the technology.
+
+I aggregate the data to the physician level.
 Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015, as I do not observe whether they become treated in 2016 or 2017.
 That is, the data contains no units which are ``never-treated".
-
 Finally, I include a physician's graduation year from Physician Compare.
@@
 
-\subsection{Outcome Variables}\label{sec:outcome}
-
-The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as retirement. This variable is constructed as follows: 
-$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
-where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$.
-Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future.
-\footnote{Alternatively, I could also define retirement using future number of patients.
+The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as exit.
+Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year.
+I do not define retirement as cumulative (a treatment that turns on and remains on for all following year), but a one-time decision.\footnote{Alternatively, I could also define retirement using future number of patients.
 I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}
-
-Next, I consider whether physicians exhibit behavior that suggests they have shifted setting or location of work.
-I consider the fraction of patients a physician sees in an office-based setting, which is directly available in the data.
-Then, I construct an indicator variable equal to one if a physician sees any positive number of patients in an office-based setting, which offers less variation but captures the extensive margin of working in an office.
+In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.} Next, I consider whether physicians exhibit behavior that suggests they have shifted work setting.
+I consider the fraction of patients a physician sees in an office-based/outpatient-based setting..
+Then, I construct an indicator variable equal to one if a physician sees any positive number of patients in such a setting, which offers less variation but captures the extensive margin of working in an office.
 Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.
@@
 
-I show summary statistics of relevant variables for the entire sample in Table \ref{tab:sumstats}.
-Only 3\% of physicians (approximately 770) retire over the course of the panel.
-30\% of the physician-year observations indicate working in an office in some capacity, with about a tenth of the number of patients being seen in an office setting.
-By construction, all physicians are exposed to an EHR at some point in the sample, so the variation in treatment variables comes from the timing of exposure.
-Physicians are typically exposed about one fourth of the way through the sample, but there is a positive number of physicians first exposed in each year from 2009-2015.
-The average physician is 45 years old and works with 1.5 hospitals and 1.3 systems.
+For each progression of decision making, I limit to physicians who did not change their behavior previously.
+That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who retire or change their work setting.
+I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstats}.
+
 \import{Table Code}{overall stats.tex}
 
-I also include a table of means comparing physicians below and above retirement age in Table \ref{tab:splitstats}.
-Senior physicians see more patients and work in more hospitals on average than younger physicians.
-They also see a larger fraction of patients in an office setting and are more likely to work in an office in general.
-The age groups are exposed to EHRs similarly.
-
-\import{Table Code}{split stats.tex}
+\normalsize
+I follow a maximum of 19,263 hospitalists over time, depending on the sample considered.
+Out of all hospitalists in the sample, 3\% (approximately 578) ever exit.
+A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards the beginning of sample.
+For hospitalists in the office setting sample, 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office.
+For those who never change their work setting, the average hospitalist sees 325 patients and files 4 claims per patient.
+Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.
 
@@
 That is, there are no ways for the physician to continue practicing without using the EHR.
-I investigate this assumption in Section \ref{sec:dataass}.
 While physicians may create work-arounds to ease the burden of EHR use, it is inevitable that they will need to interact with the technology in some capacity.
@@
 That is, the physician does not influence the hospital's decision to implement an EHR.
-I investigate this assumption in Section \ref{sec:endogeneity}.
 
@@
 \centering
-    \captionsetup{width=.85\linewidth}
+    \captionsetup{width=.7\linewidth}
     \caption{Effect of EHR Exposure on Retirement}
    \begin{subfigure}[]{.8\textwidth}
-   \includegraphics[scale=.5]{Objects/retire_plot_all.pdf}
-   \caption{}
-   \label{fig:Ng1} 
+   \caption{all physicians}
+   \includegraphics[scale=.45]{Objects/retire_plot_all.pdf}
+   \label{fig:retirefirst} 
 \end{subfigure}
 
+\vspace{3mm}
+
 \begin{subfigure}[]{.8\textwidth}
-   \includegraphics[scale=.5]{Objects/retire_plot_ages.pdf}
-   \caption{}
-   \label{fig:Ng2}
+\caption{sub-samples of physicians based on age}
+   \includegraphics[scale=.45]{Objects/retire_plot_ages.pdf}
+   \label{fig:retiresecond}
 \end{subfigure}
@@
 While EHR use should not be directly correlated with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
-CMS publishes a list of physicians who have officially opted out of seeing Medicare patients and none of the physicians in my sample have selected into this option.
+The CMS publishes a list of physicians who have officially opted out of seeing Medicare patients and none of the physicians in my sample have selected into this option.
 Additionally, hospitalists do not have complete control over their composition of patients.
@@
 I investigate another dimension by which physicians could change behavior: the extent to which they work in an office and whether this changes due to EHR exposure.
-The outcomes I consider are (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, shown in Figure \ref{fig:officefirst}, and (2) the fraction of patients a physician sees in an office in a given year, shown in Figure \ref{fig:officesecond}.
+The outcomes I consider are (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, shown in Figure \ref{fig:officefirsta}, and (2) the fraction of patients a physician sees in an office in a given year, shown in Figure \ref{fig:officeseconda}.
 This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
@@
 
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
+\begin{figure}[ht!]
+\centering
+    \captionsetup{width=.7\linewidth}
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
-    \includegraphics[scale=.5]{Objects/officeind_plot.pdf}
-    \label{fig:officefirst}
+   \begin{subfigure}[]{.8\textwidth}
+   \caption{all physicians}
+   \includegraphics[scale=.45]{Objects/officeind_plot_all.pdf}
+   \label{fig:officefirsta} 
+\end{subfigure}
+
+\vspace{3mm}
+
+\begin{subfigure}[]{.8\textwidth}
+\caption{sub-samples of physicians based on age}
+   \includegraphics[scale=.45]{Objects/officeind_plot_ages.pdf}
+   \label{fig:officefirstb}
+\end{subfigure}
 \vspace{2mm}
@@
 Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing.
-Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .07 (0.008), .06 (0.008), 0.12 (0.03), respectively.}}
-\end{figure}
-
-To continue investigating how different ages may respond differently to technology, I also present estimates for physicians $<60$ and $>=60$ separately in the bottom panel of Figure \ref{fig:officefirst}.
+Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .07 (0.008), .06 (0.008), 0.12 (0.03), respectively}}
+\end{figure}
+
+To continue investigating how different ages may respond differently to technology, I also present estimates for physicians less than 60 and greater than or equal to 60 separately in Figure \ref{fig:officefirstb}.
 Unlike the choice to retire, the age groups are both responding to EHR implementation in all years after exposure.
@@
 The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in both settings before EHR exposure change the amount of work they do in an office.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} The estimates show an increase in the fraction of patients seen in office settings after the physician is exposed to an EHR.
-In the year of exposure, the fraction of patients increases by .06, and by the fourth year after exposure the fraction of patients increases by .12 due to exposure.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} The estimates, displayed in Figure \ref{fig:officeseconda}, show an increase in the fraction of patients seen in office settings after the physician is exposed to an EHR.
+In the year of exposure, the fraction of patients increases by .06, and by the fourth year after exposure the fraction of patients increases by .12.
 Since the physicians included in the sample are meant to have close ties with hospitals, the average fraction of patients seen in an office is low, at .08.
-Thus, the observed effect I estimate here is massive as it is equivalent to a 75-150\% increase in the fraction of patients seen in an office relative to the mean.
+Thus, the observed effect is massive, as it is equivalent to a 75-150\% increase in the fraction of patients seen in an office relative to the mean.
 Further, the variation in this variable can either be driven by physicians who already worked in an office or those who did not and then began working in an office after exposure.
@@
 
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
+\begin{figure}[ht!]
+\centering
+    \captionsetup{width=.7\linewidth}
     \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
-    \includegraphics[scale=.5]{Objects/officefrac_plot.pdf}
-    \label{fig:officesecond}
+   \begin{subfigure}[]{.8\textwidth}
+   \caption{all physicians}
+   \includegraphics[scale=.45]{Objects/officefrac_plot_all.pdf}
+   \label{fig:officeseconda} 
+\end{subfigure}
+
+\vspace{3mm}
+
+\begin{subfigure}[]{.8\textwidth}
+\caption{sub-samples of physicians based on age}
+   \includegraphics[scale=.45]{Objects/officefrac_plot_ages.pdf}
+   \label{fig:officesecondb}
+\end{subfigure}
 \vspace{2mm}
@@
 
-The effect of EHR exposure on patient count is shown in Figure \ref{fig:patient}.
+The effect of EHR exposure on patient count is shown in Figure \ref{fig:patienta}.
 For hospitalists of any age, patient count increases by 22.4 patients in the year of EHR exposure, 45.4 patients in the year after exposure, 28.1 patients in the second year after exposure, and approximately 20 in the third and fourth year after exposure.
@@
 
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Patient Count}
-    \includegraphics[scale=.5]{Objects/patient_plot.pdf}
-    \label{fig:patient}
+\begin{figure}[ht!]
+\centering
+    \captionsetup{width=.7\linewidth}
+    \caption{Effect of EHR Exposure on Number of Patients Seen}
+   \begin{subfigure}[]{.8\textwidth}
+   \caption{all physicians}
+   \includegraphics[scale=.45]{Objects/patient_plot_all.pdf}
+   \label{fig:patienta} 
+\end{subfigure}
+
+\vspace{3mm}
+
+\begin{subfigure}[]{.8\textwidth}
+\caption{sub-samples of physicians based on age}
+   \includegraphics[scale=.45]{Objects/patient_plot_ages.pdf}
+   \label{fig:patientb}
+\end{subfigure}
 \vspace{2mm}
@@
 
-It may be that that EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
+EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
 That is, if productive or tech savvy physicians are the ones who remain in the same hospital after EHR exposure, the estimates are biased.The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
@@
 On average, physicians file roughly 4 claims per patient.
-I present estimates in Figure \ref{fig:claim}, which shows that EHRs tend to increase claims filed per patient over time.
+I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time.
 The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as physicians are exposed to the new software, they bill for more items on average.
-I also show heterogeneity of this result by age group, and the bottom panel of Figure \ref{fig:claim} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
+I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \captionsetup{width=.85\linewidth}
+\begin{figure}[ht!]
+\centering
+    \captionsetup{width=.7\linewidth}
     \caption{Effect of EHR Exposure on Claims per Patient}
-    \includegraphics[scale=.5]{Objects/claim_per_patient_plot.pdf}
-    \label{fig:claim}
+   \begin{subfigure}[]{.8\textwidth}
+   \caption{all physicians}
+   \includegraphics[scale=.45]{Objects/claim_per_patient_plot_all.pdf}
+   \label{fig:claima} 
+\end{subfigure}
+
+\vspace{3mm}
+
+\begin{subfigure}[]{.8\textwidth}
+\caption{sub-samples of physicians based on age}
+   \includegraphics[scale=.45]{Objects/claim_per_patient_plot_ages.pdf}
+   \label{fig:claimb}
+\end{subfigure}
 \vspace{2mm}
@@
 
-\section{Physician Influence of EHR Decision in Hospitals}
-
-\subsection{Hiring Data Assistants}\label{sec:dataass}
-
-The effects found in the above analysis may be biased if there is an external event occurring that is correlated with EHR implementation and exposure.
-The existence or hiring of employees with the purpose of utilizing electronic health records on a physician's behalf is a way for physicians to continue working in a hospital with an EHR but not experience exposure to the technology.
-Employees of this type are traditionally referred to as scribes, and are assigned a medical tax ID when working in hospitals.
-They have the following official titles in tax data: Coding Specialist (Hospital Based), Health Information Technologist, and Registered Record Administrator.
-I refer to any employee in these categories as a data assistant.
-In this section, I investigate whether data assistants were hired in accordance with EHR implementation and whether the presence of data assistants affects physician response to EHRs.
-This is particularly relevant for the productivity and efficiency outcomes where I find that EHRs have an effect.
-
-First, using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals.
-Figure \ref{fig:dataassistant_histogram} presents a frequency plot of the year of activation for every tax code that falls in the categories listed above.
-The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occurred in 2014.
-If hiring data assistants was directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians first became exposed to EHRs.
-The results of my main specification are not sensitive to leaving out later years, which is a good indication that the results are not driven by the enumeration of data assistants.
-
-\begin{figure}[t]
-\centering
-\captionsetup{width=.5\linewidth}
-\caption{Frequency of Data Assistant Enumeration by Year}
-\includegraphics[scale=.4]{Objects/dataassistant_histogram.pdf}
-\label{fig:dataassistant_histogram}
-\vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Histtogram uses NPPES data on the enumeration of certain types of NPIs to estimate the number of new data assistants being hired in each year of my sample.}}
-\end{figure}
-
-Further, I investigate the occurrence of physicians and data assistants sharing patients in the CMS Shared Patient data.
-If data assistants record patient information during appointments, they will be included on the claim to Medicare.
-I create a list of physician-years for which the hospital has a positive number of shared patients with a data assistant.
-I merge this to the physician-hospital pairs and create a physician level variable for they are associated with a data assistant.
-I find that .1\% of physician-year observations are associated with data assistants.
-This sample of physicians is too small to analyze whether data assistants are more likely with EHR exposure, yielding largely noisy estimates.
-However, I limit the sample to only physicians not associated with data assistants and find that the magnitude of estimates are similar.
-Results from specifications where data assistants are excluded are given in Appendix \ref{sec:chart}.
-
-\subsection{Endogeneity of EHR Implementation}\label{sec:endogeneity}
-
-Another way certain physicians may avoid EHR exposure is to exhibit decision making power to either delay or encourage EHR implementation in hospitals.
-This would make the binary treatment variable endogenous, biasing the results of the main specification, as it includes physicians who are selecting into their own exposure.
-I address this question by limiting the sample of hospitals, only including those which are considered to have low vertical integration with physicians.
-That is, physicians are not likely to be in leadership capacities, and thus less likely to be decision makers.
-I then redefine exposure as when a physician is exposed to an EHR in low integration settings.
-I present details of how I define integration and results for this specification in Appendix \ref{sec:changes}.
-In general, group time estimates are similar to those in the main specification, with the exception of the probability of working in an office setting, which may indicate some correlation between EHR adoption in high integration settings and acquisition of physician offices.
-That mechanism is beyond the scope of this project and left to future research.
-
 \section{Conclusion}
@@
 A technology expected to revolutionize the health care system was the advancement of health information technology in the form of complex electronic health record (EHR) systems.
-While they have the potential to improve efficiency, cut costs, and improve patient outcomes, these have not come to fruition.
+While they have the potential to improve efficiency, cut costs, and increase quality of care, these have not come to fruition.
 While the effect of EHRs on these outcomes has been directly studied, there is a gap in understanding how physicians, the primary user of this technology, responded to their rapid implementation of EHRs.
@@
 Using a difference-in-differences framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
-Further, I find evidence that physicians who do not retire change where they work because of EHR exposure, particularly shifting from hospitals towards offices.
+Further, I find evidence that physicians who do not retire choose to change where they work because of EHR exposure, particularly shifting from hospitals towards offices.
 Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
@@
 \clearpage
-
-\renewcommand*{\bibfont}{\footnotesize}
+\onehalfspacing
 
@@
 I create a variable capturing the first year that at least one of a physician's connected hospitals uses an EHR.
-For the endogeneity discussion in Section \ref{sec:endogeneity}, I create a similar variable but only count a physician as exposed if a hospital categorized as "low-integration" adopts an EHR.
-The level of vertical integration between hospital and physician is defined as in \citeauthor{dynan1998assessing} (\citeyear{dynan1998assessing}) using the organization type of the hospital.
-Hospitals classified as IPA or PHO are low-integration, and any other type is high integration.
 Finally, the last variable I create before aggregating to the physician level is an indicator for whether the physician works with the same set of hospitals through the entire sample.
@@
 
-As discussed in the body of the paper, I address potential concerns that the hiring of new employees may be driving the results for productivity and efficiency.
-Thus, as another specification, I restrict the data to only include physicians who have no record of working with data assistants as described in Section \ref{sec:dataass}.
-
 There are two main thresholds I define in the data process that were arbitrarily chosen.
@@
 $$\text{outcome}_{i,t}=\sum_{j=-4}^{-2} \text{rel\_year}_{j} + \sum_{j=0}^{4} \text{rel\_year}_{j} + \delta_i + \gamma_t,$$
-where outcome$_{i,t}$ is one of the five outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to expansion, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
+where outcome$_{i,t}$ is one of the five outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to treatment, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
 These results from each outcome are presented in Figure \ref{fig:twfe}. These results show similar patterns to the main specification, with slight variations in magnitude.
@@
 However, the approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
-Nevertheless, I limit the years of data to 2009-2015 and present estimates of this estimator in Figure \ref{fig:estimators}.
+Nevertheless, I limit the years of data to 2009-2015 and present estimates as a percentage of the mean of each variable in Figure \ref{fig:estimators}.
 Generally, the results have the same sign with a slightly larger magnitude, driven by the inclusion of never-treated units.
@@
 Further, to investigate the same idea, I present results for physicians who typically work at small hospitals vs.
-those who typically work at large hospitals.
-These average treatment effects are shown in Figure \ref{fig:heter}.
+large hospitals.
+These average treatment effects as a percentage of the mean of each variable are shown in Figure \ref{fig:heter}.
 The results do not differ along these dimensions.

### 2023-04-20 Table Code/overall stats.tex
Updates from Overleaf

@@
-\begin{table}[ht]
-
+\begin{table}[h]
+\small
 \caption{Summary Statistics}
@@
 \toprule
-Variable & N & Mean & Std. Dev. & Min & Max\\
+Variable & No. Hospitalists & Mean & Std. Dev. & Min & Max\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Outcomes}}\\
-\hspace{1em}Ever Retire & 162,018 & 0.03 & 0.17 & 0 & 1\\
-\hspace{1em}Fraction Patients in Office & 162,018 & 0.08 & 0.21 & 0 & 1\\
-\hspace{1em}Work in an Office & 162,018 & 0.27 & 0.44 & 0 & 1\\
-\hspace{1em}Change Zip Codes & 144,787 & 0.12 & 0.32 & 0 & 1\\
-\hspace{1em}Number of Patients & 162,018 & 331 & 274 & 0 & 7,613\\
-\hspace{1em}Claims per Patient & 146,443 & 4.09 & 4.1 & 0 & 481\\
+\multicolumn{6}{l}{\textbf{Exit Sample}}\\
 \addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Treatment}}\\
-\hspace{1em}Exposure to an EHR & 162,018 & 0.76 & 0.43 & 0 & 1\\
-\hspace{1em}Fraction of Hospitals with EHR & 162,018 & 0.64 & 0.44 & 0 & 1\\
+\hspace{1em}Ever Exit Clinical Work & 19,263 & 0.03 & 0.18 & 0 & 1\\
 \addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Characteristics}}\\
-\hspace{1em}Age & 162,018 & 44.55 & 8.31 & 25 & 86\\
-\hspace{1em}Female & 162,018 & 0.34 & 0.47 & 0 & 1\\
-\hspace{1em}Number of Hospitals Worked With & 162,018 & 1.49 & 0.82 & 1 & 12\\
-\hspace{1em}Number of Systems Worked With & 162,018 & 1.31 & 0.61 & 1 & 8\\
+\hspace{1em}Year of EHR Exposure & 19,263 & 2011 & 1.89 & 2009 & 2015\\
+\addlinespace[0.3em]
+\hspace{1em}Age & 19,263 & 44.58 & 8.32 & 25 & 86\\
+\hspace{1em}Female & 19,263 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 19,263 & 1.48 & 0.81 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 19,263 & 1.30 & 0.60 & 1 & 8\\
+\addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Office Setting Sample}}\\
+\addlinespace[0.3em]
+\hspace{1em}Work in an Office & 18,647 & 0.28 & 0.45 & 0 & 1\\
+\hspace{1em}Fraction Patients in Office & 18,647 & 0.09 & 0.22 & 0 & 1\\
+\addlinespace[0.3em]
+\hspace{1em}Year of EHR Exposure & 18,647 & 2011 & 1.89 & 2009 & 2015\\
+\addlinespace[0.3em]
+\hspace{1em}Age & 18,647 & 44.54 & 8.28 & 25 & 86\\
+\hspace{1em}Female & 18,647 & 0.34 & 0.47 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 18,647 & 1.49 & 0.82 & 1 & 12\\
+\hspace{1em}Number of Systems Worked With & 18,647 & 1.30 & 0.60 & 1 & 8\\
+\addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Productivity Sample}}\\
+\addlinespace[0.3em]
+\hspace{1em}Number of Patients & 12,732 & 325 & 277 & 0 & 7,613\\
+\hspace{1em}Claims per Patient & 11,599 & 4.18 & 4.35 & 0 & 481\\
+\addlinespace[0.3em]
+\hspace{1em}Year of EHR Exposure & 12,732 & 2011 & 1.96 & 2009 & 2015\\
+\addlinespace[0.3em]
+\hspace{1em}Age & 12,732 & 44.95 & 8.44 & 25 & 84\\
+\hspace{1em}Female & 12,732 & 0.35 & 0.48 & 0 & 1\\
+\hspace{1em}Number of Hospitals Worked With & 12,732 & 1.34 & 0.70 & 1 & 10\\
+\hspace{1em}Number of Systems Worked With & 12,732 & 1.20 & 0.50 & 1 & 6\\
 \bottomrule
 \end{tabular}
+\label{tab:sumstats}
 \end{table}
-\label{tab:sumstats}

### 2023-05-30 Scripts/sumstats.R
update plot ratios

@@ -73,8 +73,6 @@ sum_stats_prod <- Physician_Data %>% ungroup() %>% filter(minyr_EHR>0 & ever_ret
 sum_stats <- rbind(sum_stats_exit, sum_stats_office, sum_stats_prod)
 
 
-
-
 knitr::kable(sum_stats[c(2,6,1,3,4,5,12,9,13,7,8,10,11,18,15,20,14,16,17,19),],
              format="latex",
              table.envir="table",
@@ -96,13 +94,12 @@ means_old <- Physician_Data %>% ungroup() %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
-                 "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
-                 "Exposure to an EHR"="anyEHR_exposed",
+                 "Number of Patients"="npi_unq_benes",
+                 "Year of EHR Exposure"="minyr_EHR",
                  "Fraction Patients in Office"="pos_office",
                  "Ever Retire"="ever_retire",
                  "Work in an Office"="work_in_office",
-                 "Change Zip Codes"="change_zip",
-                 "Claim Count"="claim_count_total"), list(mean), na.rm=TRUE) %>%
+                 "Claims Per Patient"="claim_per_patient"), list(mean), na.rm=TRUE) %>%
   mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.)) %>%
   gather(key=var,value=value) %>%
   dplyr::rename(value_old=value)
@@ -113,39 +110,21 @@ means_young <- Physician_Data %>% ungroup() %>%
   summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
                  "Female"="female", "Number of Systems Worked With"="num_systems",
                  "Age"="age",
-                 "Number of Patients"="npi_unq_benes","Fraction of Hospitals with EHR"="frac_EHR",
-                 "Exposure to an EHR"="anyEHR_exposed",
+                 "Number of Patients"="npi_unq_benes",
+                 "Year of EHR Exposure"="minyr_EHR",
                  "Fraction Patients in Office"="pos_office",
                  "Ever Retire"="ever_retire",
                  "Work in an Office"="work_in_office",
[…]

### 2023-05-30 Paper/PaperEdit.tex
Updates from Overleaf

@@
 \begin{abstract}
-    Technological innovation has affected labor markets drastically over time.
+    Technology adoption in health care is of great importance to policy makers as some technologies are expected to radically improve the U.S.
+health care system.
+However, some new technologies can be burdensome for physicians and lead to frustration.
 I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S.
-health care, electronic health records (EHRs).
-I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital, and estimate average group time treatment effects on various labor market outcomes.
-I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
-Further, physicians exposed to EHRs see more patients due to the technology.
+hospitals, electronic health records (EHRs).
+I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes.
+I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
+Further, those exposed to EHRs see more patients due to the technology.
 These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
@@
 \section{Introduction}
-For decades, technological innovation has brought about changes to production in the economy through adjustments in wages, demand for labor, and other aspects of labor markets.
+For decades technological innovation has brought about changes to production in the economy through adjustments in wages, demand for labor, and other aspects of labor markets.
 Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved in the implementation of new technologies.
@@
 Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and individual labor market behaviors.
-In this paper, I investigate the effect of EHR adoption in hospitals on several behaviors of physicians working in hospitals, or hospitalists.
+In this paper, I investigate the effect of EHR adoption in hospitals on several behaviors of physicians working in hospitals.
 
 EHRs are computer systems that house digital medical records and sometimes have additional functionalities.
-EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech})\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
 In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care.
-Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}.
-President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
+Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}).
 The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy.
@@
 
-While hospitalists are ideal for to answer this question econometrically, they are also important in their own right.
+While hospitalists create an ideal setting to answer this question econometrically, they are also important in their own right.
 The Bureau of Labor Statistics estimates approximately 44,000 physicians employed as hospitalists in the US, and studies estimate that 75\% of hospitals employ hospitalists (\cite{bls}, \cite{healthecareers_2022}).
@@
 Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify the hospital affiliation of each hospitalist.
-Then, I use the American Hospital Association (AHA) Survey to determine EHR adoption status, and thus connect hospitalists to EHR use.
+Then, I use the American Hospital Association (AHA) Survey to determine EHR adoption status, and thus connect hospitalists to EHR exposure.
 Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for information on the patients seen by each hospitalist.
@@
 
-This paper contributes to our understanding of health information technology and its affects in health care.
+This paper contributes to our understanding of the effects of health information technology.
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
@@
 This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.
+\textcolor{red}{reword the implications and check the numbers are correct}
 
@@
 
-\subsubsection{The Hospitalist Profession}
-
 For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting.
 However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist.
-Both formal hospitalists and internists are medical professionals who primarily focus on adult patient care, but they differ in their work settings.
+Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings.
 Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting.
-While this distinction is potentially important for some of the decisions outlined below, I do not have the ability to distinguish between the two explicitly in the data.
+While this distinction is potentially important for some of the decisions outlined below, the data do not explicitely distinguish between the two.
 I will discuss this limitation in more detail when relevant.
@@
 Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
-Additionally, if EHRs cause physicians to substitute time away from patients and they receive less reimbursement, leisure becomes relatively cheaper, leading to retirement.
 Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.
@@
 As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
-Thus, the term exit for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
-A physician is said to exit in a given year only if it is the first year in which the physician has no future Medicare claims.
+For the purpose of this paper, the term exit indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
 
@@
 
-For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of working in clinical settings.
+Most physicians are likely not on the margin in their decision to exit clinical work, so EHRs will not impose a large enough cost (or benefit) to impose exit.
 However, there are other ways hospitalists might change their behavior under drastic changes to their current working environment.
 For those who remain practicing, changing the physical place of work may allow for more control over which technologies they use.
-Some hospitalists work under contract full time for one medical practice where they have a consistent shift schedule (formal hospitalists), while others contract themselves out to multiple hospitals, offices, or both (internists).
-While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.
-
-First, I consider those who are in a relatively binding contract with the hospital.
-For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital.
-This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings.
-A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making it very costly to change work setting or excluding it from the choice set altogether.
-Since these hospitalists are established in their hospital, they work a significant amount of time there.
-When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.
-
-Alternatively, there may be hospitalists that work in multiple hospitals or work in both hospitals and offices throughout the sample period.
-If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction.
-Theoretically, these types of hospitalists are much more likely to change their behavior as a result of a change in one work environment.
+Specifically, if EHRs change job satisfaction of working in the hospital, a hospitalist may decide to shift more of their working hours towards or away from the hospital.
+Not all physicians working as hospitalists have a binding contract with their associated hospital.
+In fact, a 2013 survey estimated that 36\% of hospitalists also worked outside of the hospital in primary care, long-term care, or hospice settings (\cite{Today}).
+Many physicians trained as internists choose to contract their work out to both hospitals and outpatient settings.
+The fraction of internists who work as hospitalists increased from 25\% to 40\% from 2008 to 2018 (\cite{gray2022evolving}).
+
+The sample of physicians I consider include both hospitalists with formal contracts and internists who contract a majority of their time with the hospital, but could also work in other settings as well.
+If a hospital implements a costly EHR, a hospitalist may choose to shift more of their time towards outpatient settings or initiate a contract with outpatient settings to spend less time in the setting where job satisfaction has decreased.
 
@@
 
-There are still agents who are not induced to change their behavior on the basis of exit or work setting due to EHR implementation.
-A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment.
-For hospitalists who make no physical changes to work environment after EHR exposure for any of these reasons, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
-
-Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous.
-If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients.
-In this case, productivity would increase.
-However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.
+For those not induced to change their behavior by exiting or changing work settings, it is natural to assess whether EHRs impacted productivity.
+A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+This question is empirically ambiguous since EHRs could add daily tasks that make hospitalists less productive, increase efficiency, or some combination of the two where short term productivity is negatively impacted but long term productivity is positively impacted once physicians adjust and learn the new technology.
+I measure productivity in terms of the amount of patients a hospitalist is able to see.
+
 Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the hospitalist.
 Billing activity refers to the number of claims filed per patient seen.
-EHRs may improve efficiency if they remove the need for repeated or excess testing.
-However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient.
+There are many reported reasons that physicians tend to over-treat patients, including defensive medicine, patient request, difficulty accessing patient records, and pressure from their institution (\cite{lyu2017overtreatment}).
+EHRs have the potential to change some of these incentives.
+As more hospitals adopt EHRs, patient information should be more easily accessible and reduce the amount of repeated tests a physician needs to run, leading to a decrease in claims per patient.
+However, institutions have power to customize EHRs and train their hospitalists to increase fee-for-service revenue by increasing claims per patient.
 
@@
 
-I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician, hospital, and practice characteristics.
+I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician and hospital characteristics.
 I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
@@
 The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017.
-Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office, emergency room, etc.).
-I only include physicians whose primary specialty as hospitalist\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting.
+Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.).
+I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting.
 I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year.
 I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}).
-I drop hospitalists with low claim counts that do not accurately represent changes in behavior.
 I use this data to construct the dependent variables for my analysis.
-Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and overall billing activity.
-I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.
+Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and billing activity.
 
@@
 I aggregate the data to the physician level.
-Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015, as I do not observe whether they become treated in 2016 or 2017.
-That is, the data contains no units which are ``never-treated".
-Finally, I include a physician's graduation year from Physician Compare.
+I drop physicians who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated.
+The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated".
+Finally, I include a physician's graduation year from Physician Compare (provided by CMS).
 I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
@@
 The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as exit.
-Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year.
-I do not define retirement as cumulative (a treatment that turns on and remains on for all following year), but a one-time decision.\footnote{Alternatively, I could also define retirement using future number of patients.
-I use claim count for a more conservative measure of retirement, but using patient count yields identical results.
-In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.} Next, I consider whether physicians exhibit behavior that suggests they have shifted work setting.
-I consider the fraction of patients a physician sees in an office-based/outpatient-based setting..
-Then, I construct an indicator variable equal to one if a physician sees any positive number of patients in such a setting, which offers less variation but captures the extensive margin of working in an office.
+Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year and zero in every other year.
+This captures a one time decision to stop seeing patients, not a cumulative decision in all following years.
+The next variables I construct capture whether physicians change work setting.
+I consider the fraction of patients a physician sees in an office-based/outpatient setting, and I construct an indicator variable equal to one if a physician sees any positive number of patients in such a setting.
 Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.
@@
 For each progression of decision making, I limit to physicians who did not change their behavior previously.
-That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who retire or change their work setting.
+That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who exit or change their work setting.
 I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstats}.
@@
 I follow a maximum of 19,263 hospitalists over time, depending on the sample considered.
-Out of all hospitalists in the sample, 3\% (approximately 578) ever exit.
-A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards the beginning of sample.
-For hospitalists in the office setting sample, 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office.
-For those who never change their work setting, the average hospitalist sees 325 patients and files 4 claims per patient.
+Out of all hospitalists in the sample, 3\% (approximately 578) stop seeing patients at some point.
+A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards early years.
+For hospitalists in the office setting sample (those who do not exit), 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office.
+There are 12,732 hospitalists who never exit or change their work setting at any point in the sample.
+These hospitalists see 325 patients and file 4 claims per patient on average.
 Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.
 
-In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time.
-In 2009, 28\% of the sample of physicians were already exposed to an EHR.
-That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period.
+In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time and the underlying hospital adoption that drives this treatment variable.
+Hospitalists are exposed the first time one of their hospital's adopts an EHR.
+Thus, I show the percent of hospitals adopting as well as the percent of hospitalists exposed to the technology.
+In 2009, 41\% of all hospitalists were already exposed to an EHR.
+That is, half of the sample had no affiliation with EHRs at the beginning of the sample period.
 Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by the end of the sample period.
-The darker line shows the underlying variation in this variable, hospitals implementing EHRs over time.
+In 2009, 47\% of hospitals have adopted an EHR, and this percentage grows steadily over time, driving the increase in exposed hospitalists.
 
@@
     \caption{Treatment Variables Over Time}
-    \includegraphics[scale=.4]{Objects/sum_stats_year.pdf}
+    \includegraphics[scale=.6]{Objects/sum_stats_year.pdf}
     \label{fig:treatmentgraph}
@@
 
-I treat EHR implementation as an exogenous treatment variable, and I estimate average treatment effects of EHR exposure on the outcomes discussed.
-In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
+I estimate average treatment effects of EHR exposure on the outcomes discussed.
+In this setting, treatment effects may be heterogeneous across hospitalists who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt.
 Therefore, to avoid negative weighting issues this causes in a classical two way fixed effects event study specification, I instead estimate average treatment effects for a specific group $g$ at time $t$:
@@
 where $G_g=1$ for those in group $g$.
-A group indicates all physicians treated, or first exposed to an EHR, in the same year.
-To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{sant2020doubly} (\citeyear{sant2020doubly}).
+A group indicates all hospitalists treated, or first exposed to an EHR, in the same year.
+I employ the doubly robust estimator established in \citeauthor{sant2020doubly} (\citeyear{sant2020doubly}).
 Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}.
@@
 
-I aggregate the group time estimates over groups to a more familiar event study plot.
-For select outcomes, I present the original pre-aggregation estimates to show that results are not driven by a particular year of treatment.
+I aggregate the group time estimates over groups to a more familiar event study plot with simultaneous confidence bands calculated using bootstrapped standard errors.
 Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each results plot.
-Simultaneous confidence bands, calculated using bootstrapped standard errors, are also presented.
 
@@
 First, I assume that treatment is not reversed.
-That is, once a physician is exposed to an EHR, they cannot be un-exposed.
-For physicians that do not switch hospitals, this assumption is supported by the institutional details of hospital EHR implementation.
-An EHR is a costly technology and requires a significant amount of collaboration to implement.
-A hospital does not have incentive to un-implement an EHR once it is implemented.
+That is, once a hospitalist is exposed to an EHR, they cannot be un-exposed.
+For hospitalists that do not switch hospitals, this assumption is supported by the institutional details of hospital EHR implementation.
+An EHR is a costly technology and requires a significant amount of collaboration to implement, so a hospital does not have incentive to un-implement an EHR.
 They may add features or switch vendors, but do not reverse EHR use.
-However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data.
-Thus, I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period.
-I discuss this limitation to the data further in Section \ref{sec:patientcount}.
-
-Second, I assume physicians do not anticipate EHR exposure prior to occurrence.
-This may be a concern if physicians learn that the system will be implemented and change their behavior beforhand.
+Any hospitalists included in the productivity sample are those which did not switch work settings, satisfying this assumption.
+
+Second, I assume hospitalists do not anticipate EHR exposure prior to occurrence.
+If they learn that the system will be implemented and change their behavior beforehand, the estimates may be biased.
 Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}).
 Therefore, I proceed in the main specification assuming no anticipation.
-However, I explore and present results for a one year anticipation period in the specification charts in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change my findings.
+However, I explore and present results for a one year anticipation period in the specification charts in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change the findings.
 
@@
 I assume that, conditional on years of experience, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
-This assumption would be violated under external conditions in certain years that may be correlated with labor markets, such as a major recession.
-The time period I study, 2009-2017, is reassuringly stable.
-In my analysis, I present p-values for a Wald test of pre-trends.
+While this assumption cannot be tested directly, I present p-values for a Wald test of pre-trends.
 Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends} by allowing for specified violations in the parallel trends assumption (\cite{rambachan2019honest}).
@@
 To interpret these estimators as the causal effect of EHR exposure on various labor market outcomes, there are additional institutional assumptions necessary.
-First, I assume that when a physician works in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
-That is, there are no ways for the physician to continue practicing without using the EHR.
-While physicians may create work-arounds to ease the burden of EHR use, it is inevitable that they will need to interact with the technology in some capacity.
+First, I assume that when a hospitalist works in an electronic record utilizing-hospital, they utilize the electronic health record system fully.
+That is, the hospitalist does not continue practicing without using the EHR.
+While some hospitalists may create work-arounds to ease the burden of EHR use, it is inevitable that they will need to interact with the technology in some capacity.
 Second, I assume that a physician's exposure to EHRs is exogenous.
@@
 
-\subsection{Retirement}\label{sec:retire}
-
-I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age in Figure \ref{fig:retirefirst}.\footnote{Alternatively, one could estimate the effect for more age groups and observe the relationship between the overall ATT and age.
-However, the data is not ideal for more than two age groups since the sample sizes decrease dramatically for certain groups.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
+\subsection{No Longer Seeing Patients}\label{sec:retire}
+
+I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of exiting clinical work for the full sample of hospitalists, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age in Figure \ref{fig:retirefirst}.\footnote{Alternatively, one could estimate the effect for more age groups and observe the relationship between the overall ATT and age.
+However, the data does not support further sub-setting as sample sizes become small.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
 Being exposed to an EHR leads to a .2 ppt increase in the likelihood of retiring in the first year after exposure, and a .5 ppt increase in the second year after exposure.
@@
    \caption{all physicians}
-   \includegraphics[scale=.45]{Objects/retire_plot_all.pdf}
+   \includegraphics[scale=.55]{Objects/retire_plot_all.pdf}
    \label{fig:retirefirst} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.45]{Objects/retire_plot_ages.pdf}
+   \includegraphics[scale=.55]{Objects/retire_plot_ages.pdf}
    \label{fig:retiresecond}

### 2023-05-30 Paper/Paper.tex

@@
 \begin{abstract}
-    Technological innovation has affected workers and labor markets for decades.
-I add to our understanding of technology's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S.
+    Technological innovation has affected labor markets drastically over time.
+I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S.
 health care, electronic health records (EHRs).
-I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes.
-I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure.
-Further, physicians exposed to EHRs see more patients than those who are not.
-These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicate potential productivity gains for those who embrace the technology.
+I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes.
+I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
+Further, those exposed to EHRs start to see more patients after exposure to the technology.
+These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
 \end{abstract}
@@
 \section{Introduction}
-For decades, technological innovation has brought about changes to production in the economy, including wages, demand for labor, and other aspect of labor markets.
+For decades technological innovation has brought about changes to production in the economy, including wages, demand for labor, and other aspect of labor markets.
 Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved with the implementation of new technologies.
@@
 A recent technology roll-out that altered the U.S.
-health care system dramatically was the electronic health record (EHR).
+health care system dramatically was the electronic health record (EHR), which shifted medical record storage from physical to computerized.
 The use of EHRs was incentivized by the government, leading to a rapid implementation by hospitals and practices nationwide.
@@
 
-EHRs are computer systems that house digital medical records and have additional capabilities, including decision making assistance.
+EHRs are computer systems that house digital medical records and sometimes have additional capabilities such as decision making assistance.
 EHRs have become increasingly relevant in the U.S.
@@
 For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).
-The desired movement towards digitization was realized; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
+The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).
 
@@
 However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
-Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices, and I highlight the differences in effects for older vs.
-younger physicians.
+Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.
 
@@
 One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).
+There is still much to be learned about EHRs impacts in health care (\cite{NBERw29218}).
 

### 2023-05-31 Scripts/analysis.R
specification charts

@@ -159,7 +159,7 @@ graphs <- lapply(graph_data, function(x){
     theme(panel.grid.major.x = element_blank() ,
           panel.grid.major.y = element_line(size=.05, color="lightgray" )) +
     labs(caption=paste0("pre-trends test p-value: ",round(x[[2]],2))) + 
-    theme(plot.caption=element_text(hjust = 0))
+    theme(plot.caption=element_text(hjust = 0, size=15))
   
   ages <- ggplot(ages_data, aes(year, att, color=group)) +  
     geom_vline(xintercept="0", linetype="dashed", colour="red") +
@@ -177,22 +177,22 @@ graphs <- lapply(graph_data, function(x){
     theme(legend.position="bottom") + 
     labs(color='Age Group', caption=paste0("pre-trends test p-value (young): ",round(x[[3]],2),
                                            "\npre-trends test p-value (old): ", round(x[[4]],2))) + 
-    theme(plot.caption=element_text(hjust = 0))
+    theme(plot.caption=element_text(hjust = 0, size=15))
   
   list(all=all, ages=ages)
 })
 
 # Save plots -------------------------
-ggsave(file="Objects/retire_plot_all.pdf",plot=graphs[[1]][["all"]], width=10, height=3.5, units="in")
-ggsave(file="Objects/retire_plot_ages.pdf",plot=graphs[[1]][["ages"]], width=10, height=4, units="in")
-ggsave(file="Objects/officefrac_plot_all.pdf",plot=graphs[[2]][["all"]], width=10, height=3.5, units="in")
-ggsave(file="Objects/officefrac_plot_ages.pdf",plot=graphs[[2]][["ages"]], width=10, height=4, units="in")
-ggsave(file="Objects/officeind_plot_all.pdf",plot=graphs[[3]][["all"]], width=10, height=3.5, units="in")
-ggsave(file="Objects/officeind_plot_ages.pdf",plot=graphs[[3]][["ages"]], width=10, height=4, units="in")
-ggsave(file="Objects/patient_plot_all.pdf",plot=graphs[[4]][["all"]], width=10, height=3.5, units="in")
-ggsave(file="Objects/patient_plot_ages.pdf",plot=graphs[[4]][["ages"]], width=10, height=4, units="in")
[…]

### 2023-05-31 Scripts/specification_chart.R

@@ -176,29 +176,6 @@ Data_List <- lapply(Data_List, function(x){
     select(-firstyear_0, -firstyear_1, -firstyear_2, -lastyear_0, -lastyear_1, -lastyear_2)
 })
 
-# Read in hospitals with data assistants
-Hosp_with_DA <- readRDS(paste0(created_data_path,"/Hosp_with_DA.rds"))
-
-Hosp_with_DA <- Hosp_with_DA %>%
-  mutate(DA=1) %>%
-  mutate(HospNPI=as.numeric(HospNPI))
-
-Phys_with_DA <- readRDS(paste0(created_data_path,"/Phys_with_DA.rds")) %>%
-  dplyr::mutate(Phys_DA=1)
-
-Data_List <- lapply(Data_List, function(x){
-    data <- x %>%
-    dplyr::mutate(DocNPI=as.character(DocNPI)) %>%
-    left_join(Phys_with_DA, by=c("year", "DocNPI"))
-  
-  data <- data %>%
-    dplyr::mutate(Phys_DA=ifelse(is.na(Phys_DA),0,Phys_DA)) %>%
-    group_by(DocNPI) %>%
-    dplyr::mutate(sum=sum(Phys_DA)) %>%
-    ungroup() %>%
-    dplyr::mutate(ever_Phys_DA=ifelse(sum>0,1,0))
-})
-
 # Get rid of physicians who only work on VA hospitals
 
 Data_List <- lapply(Data_List, function(x){
@@ -236,11 +213,74 @@ Data_List <- lapply(Data_List, function(x){
   return(data)
 })
 
+# Read in HIMSS data for additional EHR information
+for (t in 2009:2017){
+  year <- readxl::read_xlsx(paste0(raw_data_path,"/HIMSS/HostedSoftware",t,".xlsx")) %>%
+    mutate(year=t) %>%
+    select(HAEntityId, SurveyId, VendorName, year)
+  
+  assign(paste0("HostedSoftware",t),year)
+}
+
+HostedSoftware <- rbind(HostedSoftware2009, HostedSoftware2010, HostedSoftware2011, HostedSoftware2012, HostedSoftware2013, HostedSoftware2014, HostedSoftware2015, HostedSoftware2016, HostedSoftware2017)
+rm(HostedSoftware2009, HostedSoftware2010, HostedSoftware2011, HostedSoftware2012, HostedSoftware2013, HostedSoftware2014, HostedSoftware2015, HostedSoftware2016, HostedSoftware2017)
+
[…]

### 2023-05-31 Paper/PaperEdit.tex
Updates from Overleaf

@@
 
-I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician and hospital characteristics.
+I construct a balanced hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician and hospital characteristics.
 I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
@@
 However, the data does not support further sub-setting as sample sizes become small.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}.
-Being exposed to an EHR leads to a .2 ppt increase in the likelihood of retiring in the first year after exposure, and a .5 ppt increase in the second year after exposure.
-While numerically small, these effects are statistically and economically meaningful relative to the proportion of physicians who retire in the sample, .03.
-That is, the effect represents a 7\% (16\% in the second year) increase in the likelihood of retirement relative to the mean.
-
-Even though there is no visual indication of a violation of the parallel trends assumption in the pre-periods, the p-value of a Wald test for pre-trends indicates some evidence for such a violation.
-In Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
-Due to the rarity of retirement in the sample, the estimate becomes null with any specified variation in the parallel trend assumption.
-Since the pre-trend does not look strong in the first place, this is not too concerning.
-I present estimates for the overall $ATT$ under various specifications of the data in Appendix \ref{sec:chart}.
-The estimate found here falls in the upper range of estimates.
-All estimates are positive, yet some have confidence intervals that include zero.
-Closer examination into these estimates reveals that stricter thresholds on the data yield many of the null results.
-Even with additional noise in the estimates, the average of the upper and lower confidence intervals still indicates a positive effect of EHR exposure of retirement.
-Thus, the main finding is robust to various specifications.
+Being exposed to an EHR leads to a .2 ppt increase in the likelihood of exiting in the first year after exposure, and a .4 ppt increase in the second year after exposure.
+While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03.
+That is, the effect represents a 7\% (13\% in the second year) increase in the likelihood of exiting relative to the mean.
 
@@
 \centering
-    \captionsetup{width=.7\linewidth}
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Retirement}
@@
 
-\vspace{3mm}
+\vspace{1.5mm}
 
@@
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
-Interestingly, senior physicians are driving the positive estimate in the year after exposure, and younger physicians only retire in the second year after exposure.
-A retirement age physician is 1 ppt more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean.
-In comparison, physicians less than 60 are not more likely to retire in the first year after exposure, but are approximately 16\% more likely to retire two years after exposure.
+Interestingly, hospitalists >=60 are driving the positive estimate in the year after exposure, and younger hospitalists only exit in the second year after exposure.
+A retirement age physician is 1 ppt more likely to exit after being exposed to an EHR, a 25\% increase relative to the mean.
+In comparison, hospitalists less than 60 are not more likely to exit in the first year after exposure, but are approximately 16\% more likely to exit two years after exposure.
 There are two potential reasons for the different response time that relate to formal retirement vs.
@@
 First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
-Similarly, younger physicians may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
-
-A potential concern is that physicians are not leaving a clinical setting, but only see patients with private insurance.
-While EHR use should not be directly correlated with Medicare patients, there may be an unobserved event leading physicians to stop seeing Medicare patients that aligns with EHR implementation.
-The CMS publishes a list of physicians who have officially opted out of seeing Medicare patients and none of the physicians in my sample have selected into this option.
-Additionally, hospitalists do not have complete control over their composition of patients.
-
-\subsection{Place of Work}
-
-I investigate another dimension by which physicians could change behavior: the extent to which they work in an office and whether this changes due to EHR exposure.
-The outcomes I consider are (1) an indicator variable equal to 1 if the physician has any positive number of patients in an office in a given year, shown in Figure \ref{fig:officefirsta}, and (2) the fraction of patients a physician sees in an office in a given year, shown in Figure \ref{fig:officeseconda}.
-This analysis does not include individuals who retire at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
+Similarly, younger hospitalists may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
+
+There is no visual indication of a violation of the parallel trends assumption in the pre-periods.
+However, the p-value of a Wald test for pre-trends indicates some evidence for such a violation for younger hospitalists.
+In Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends.
+Due to the rarity of exiting for this sub-sample, the estimate becomes null with any specified variation in the parallel trend assumption.
+I still discuss the estimates for this age group, but focus on the results for the older hospitalists.
+I also present overall treatment effects under different data specifications in Appendix \ref{sec:chart}, and find that this result is robust to various specifications.
+
+A potential concern is that hospitalists are not leaving a clinical setting, but choosing to only see patients with private insurance.
+First, EHR use should not be directly correlated with Medicare patients.
+Further, as a hospitalist, these physicians do not have complete control over their patient composition and are not paid differently based on the types of patients they see.
+As a final check, the CMS publishes a list of physicians who have officially opted out of seeing Medicare patients and none of the hospitalists in the sample have selected into this option.
+
+\subsection{Work Setting}
+
+I investigate another dimension by which hospitalists could change behavior: the extent to which they work in an office (outpatient setting) and whether this changes due to EHR exposure.
+The outcomes I consider are (1) an indicator variable equal to 1 if the hospitalist has any positive number of patients in an office in a given year, shown in Figure \ref{fig:officefirsta}, and (2) the fraction of patients a hospitalist sees in an office in a given year, shown in Figure \ref{fig:officeseconda}.
+This analysis does not include individuals who exit at any point in the sample, as a zero could indicate dropping out of the data instead of working solely in a hospital.
 
 EHR exposure increases the likelihood of working in an office by at least 4 ppts in every year after being exposed, where the effect increases over time (up to 10 ppts four years after exposure).
-This estimate is equivalent to a 15-38\% increase relative to the average proportion of physicians who work in an office.
-That is, physicians working only in a hospital are more likely to start working in an office after being exposed to an EHR, and these effects are just as large as those on retirement.
-However, unlike the estimated retirement effects, the increased likelihood of working in an office is persistent over time, which suggests that EHRs impact work environment in hospitals for several years after implementation.
-
-I consider the robustness of these results to various violations in the parallel trends assumption in Appendix \ref{sec:pretrends}, where the findings are similar.
-Further, I compare this finding with various specifications to data thresholds and variable definitions with an estimate chart in Appendix \ref{sec:chart}.
-I find consistently positive and significant coefficients ranging from 2-8 percentage points.
-Thus, the main finding is on the high end of the range of estimates.
-While all estimates are positive, some variations that drive the estimate down slightly are limiting years to 2009-2015, and using the low integration definition of EHR exposure.
-I discuss this further in Appendix \ref{sec:changes}, and confirm that my finding is robust to specification changes.
+This estimate is equivalent to a 15-38\% increase relative to the average proportion of hospitalists who work in an office.
+That is, those working only in a hospital are more likely to start working in an office after being exposed to an EHR, and these effects are just as large as those on exiting.
+However, unlike the estimated exiting effects, the increased likelihood of working in an office is persistent over time, which suggests that EHRs impact work environment in hospitals for several years after implementation.
 
@@
 \centering
-    \captionsetup{width=.7\linewidth}
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Likelihood of Working in Office}
@@
    \caption{all physicians}
-   \includegraphics[scale=.45]{Objects/officeind_plot_all.pdf}
+   \includegraphics[scale=.55]{Objects/officeind_plot_all.pdf}
    \label{fig:officefirsta} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.45]{Objects/officeind_plot_ages.pdf}
+   \includegraphics[scale=.55]{Objects/officeind_plot_ages.pdf}
    \label{fig:officefirstb}
@@
 
-To continue investigating how different ages may respond differently to technology, I also present estimates for physicians less than 60 and greater than or equal to 60 separately in Figure \ref{fig:officefirstb}.
-Unlike the choice to retire, the age groups are both responding to EHR implementation in all years after exposure.
-The magnitudes for senior physicians continue to increase over time, whereas the estimates for younger physicians level out.
+To continue investigating how different ages may respond differently to technology, I also present estimates for hospitalists less than 60 and greater than or equal to 60 separately in Figure \ref{fig:officefirstb}.
+Unlike the choice to exit, the age groups are both responding to EHR implementation in all years after exposure.
+The magnitudes for senior hospitalists continue to increase over time, whereas the estimates for younger physicians level out.
 However, senior physicians are more likely to work in an office than younger physicians on average, so both groups exhibit similar increases relative to the mean.
 
-The previous results are driven by physicians who are not already working in an office prior to EHR exposure, but it could also be the case that physicians who already work in both settings before EHR exposure change the amount of work they do in an office.
-Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} The estimates, displayed in Figure \ref{fig:officeseconda}, show an increase in the fraction of patients seen in office settings after the physician is exposed to an EHR.
-In the year of exposure, the fraction of patients increases by .06, and by the fourth year after exposure the fraction of patients increases by .12.
-Since the physicians included in the sample are meant to have close ties with hospitals, the average fraction of patients seen in an office is low, at .08.
+The previous results are driven by hospitalists who are not working in an office prior to EHR exposure, but it could also be the case that hospitalists who already work in both settings before exposure change the amount of work they do in an office.
+Thus, I estimate the effect of EHR exposure on the fraction of patients seen in office based settings.\footnote{Since this variable is tied to the total number of patients seen, I also estimate this analysis with the outcome of total number of patients seen in an office and the results are identical.} The estimates, displayed in Figure \ref{fig:officeseconda}, show an increase in the fraction of patients seen in office settings after the hospitalist is exposed to an EHR.
+In the year of exposure, the fraction of patients increases by .04, and by the fourth year after exposure the fraction of patients increases by .12.
+Since the hospitalists included in the sample are meant to have close ties with hospitals, the average fraction of patients seen in an office is low, at .09.
 Thus, the observed effect is massive, as it is equivalent to a 75-150\% increase in the fraction of patients seen in an office relative to the mean.
@@
 \centering
-    \captionsetup{width=.7\linewidth}
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
@@
    \caption{all physicians}
-   \includegraphics[scale=.45]{Objects/officefrac_plot_all.pdf}
+   \includegraphics[scale=.55]{Objects/officefrac_plot_all.pdf}
    \label{fig:officeseconda} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.45]{Objects/officefrac_plot_ages.pdf}
+   \includegraphics[scale=.55]{Objects/officefrac_plot_ages.pdf}
    \label{fig:officesecondb}
@@
 
-I also investigate pre-trends for this outcome in Appendix \ref{sec:pretrends}, where I find that under various assumptions of the form of violation in parallel trends, the estimate remains positive and significant of similar magnitude.
+I also investigate pre-trends for these outcomes in Appendix \ref{sec:pretrends}, where I find that under various assumptions of the form of violation in parallel trends, the estimate remains positive and significant of similar magnitude.
 Further, under different combinations of data specifications I find consistently positive and significant results, as shown in Appendix \ref{sec:changes}.
-However, the range of magnitudes is large: 1 to 10 percentage points.
-Thus, the main finding is large relative to these other estimates.
-Both accounting for anticipation and limiting sample years push the magnitude of the estimate down.
-Taking the average of these confidence intervals shows that the positive effect of EHR exposure on the fraction of patients seen in the office is robust to many combinations of data thresholds.
-
-The increasing trend over time remains the same for physicians of different age groups.
-However, senior physicians have larger estimated magnitudes in each year after exposure.
-Since senior physicians have more patients in an office on average, the percentage change is similar between the two age groups.
 
@@
 As a measure of productivity/efficiency, I now estimate the effect of EHR exposure on patient count and billing activity per patient.
-If the sample includes physicians who shift place of work, this may lead to biased estimates since a physician could switch hospitals due to an EHR and then change practice behavior due to the change in environment, not due to the EHR.
-Thus, I limit to physicians who work with the same one hospital through the entire sample.
-These hospitalists do not retire or see patients in another hospital, implying that when an EHR is implemented, the physician remains utilizing the EHR for the remainder of the sample.
-
+I limit to hospitalists who work with the same one hospital through the entire sample.
+These hospitalists do not exit or see patients in another hospital, so that when an EHR is implemented the they remain utilizing the EHR for the remainder of the sample.
 The effect of EHR exposure on patient count is shown in Figure \ref{fig:patienta}.
-For hospitalists of any age, patient count increases by 22.4 patients in the year of EHR exposure, 45.4 patients in the year after exposure, 28.1 patients in the second year after exposure, and approximately 20 in the third and fourth year after exposure.
-These estimates translate to a 6-14\% increase in the number of patients seen relative to the mean.
-Interestingly, when splitting the sample by age group, the effect decreases three and four years after exposure for younger physicians, while the effect for senior age physicians remains persistent over time, suggesting that technology's effect on productivity in the long run depends on worker type (in this case, age).
-
-Again, I present a sensitivity analysis considering the parallel trends assumption in Appendix \ref{sec:pretrends} and find that this increase in patient count is robust to violations in the parallel trends assumption.
-I also present estimates under different data specifications in Appendix \ref{sec:chart}.
-While the average of the upper and lower confidence interval values still suggest a positive effect, there are many specifications where the confidence interval contains zero, suggesting a null result.
-The main change where we see this difference is in the definition of EHR exposure.
-Using the low-integration definition to account for endogeneity yields no effect.
-This indicates that on the margin of productivity, physicians who are more likely to be involved in hospital decisions do not experience the productivity gains I find in the main analysis.
+For hospitalists of any age, patient count increases by 22 patients in the year of EHR exposure, 55 patients in the year after exposure, 40 patients in the second year after exposure, 30 and 20 patients in the third and fourth year after exposure.
+These estimates translate to a 6-17\% increase in the number of patients seen relative to the mean.
+Interestingly, when splitting the sample by age group, the effect decreases three and four years after exposure for younger hospitalists, while the effect for senior age hospitalists remains persistent over time, suggesting that technology's effect on productivity in the long run depends on worker type (in this case, age).
 
@@
 \centering
-    \captionsetup{width=.7\linewidth}
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Number of Patients Seen}
@@
    \caption{all physicians}
-   \includegraphics[scale=.45]{Objects/patient_plot_all.pdf}
+   \includegraphics[scale=.55]{Objects/patient_plot_all.pdf}
    \label{fig:patienta} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.45]{Objects/patient_plot_ages.pdf}
+   \includegraphics[scale=.55]{Objects/patient_plot_ages.pdf}
    \label{fig:patientb}
@@
 The ACA had potentially opposing effects on Medicare patients: they may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
-A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change is not driving my results.
-Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup}.
+A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change should not drive my results.
+Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup} of the appendix.
 Each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups affected by the ACA are not driving the results.
 
-\begin{figure}[ht!]
-    \centering
-    \captionsetup{width=.6\linewidth}
-    \caption{Effect of EHR Exposure on Patient Count by Group}
-    \includegraphics[scale=.4]{Objects/patient_group.pdf}
-    \label{fig:patientgroup}
-    \vspace{2mm}
-    \caption*{\footnotesize{\textit{Notes:} Plot shows average group time treatment effects for each treatment group.}}
-\end{figure}
-
-EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem.
-That is, if productive or tech savvy physicians are the ones who remain in the same hospital after EHR exposure, the estimates are biased.The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
+EHR exposure affects which hospitalists are included in this analysis (since I drop those who exit or move), indicating a non-random attrition problem.
+That is, if productive or tech savvy hospitalists are the ones who remain in the same hospital after EHR exposure, the estimates are biased.
+The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
 To investigate whether my results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}).
@@
 That is, treatment can only affect selection in one direction.
-In my case, that means that EHR exposure increases the likelihood of retiring or moving workplace settings and no physicians remain in the sample under treatment who would have retired or moved in the absence of EHR exposure.
-Since I find in earlier analyses that physicians are more likely to move and retire due to EHRs, this assumption is reasonable.
+In my case, that means that EHR exposure increases the likelihood of exiting or moving workplace settings and no hospitalists remain in the sample under treatment who would have exited or moved in the absence of EHR exposure.
+Since I find in earlier analyses that hospitalists are more likely to move and exit due to EHRs, this assumption is reasonable.
 To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate.
@@
 
-Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in working physicians.
-Specifically, a hospital that loses a portion of their physicians, and divides those patients between the remaining physicians.
-In the long run, the hospital should hire new physicians to replace those that left, but in the short run estimates could be driven by increased workload instead of EHR use.
+Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in the number of working hospitalists.
+Specifically, a hospital that loses a portion of their physicians, and divides those patients among those remaining.
+In the long run, the hospital should hire new hospitalists to replace those that left, but in the short run estimates could be driven by increased workload instead of EHR use.
 This could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time.
-Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from retired physicians were divided between remaining physicians.
-In the year of exposure the entire increase in patient count is due to EHR use since I find no effect on the likelihood of retirement.
-In the first year after exposure, I estimate the increase in patients due to retiring physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above.
-However, the second year after exposure is when I observe that largest effect on retirement, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in physicians.
+Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from hospitalists who exited were divided between remaining hospitalists.
+In the year of exposure the entire increase in patient count is due to EHR use since I find no effect on the likelihood of exit.
+In the first year after exposure, I estimate the increase in patients due to exiting physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above.
+However, the second year after exposure is when I observe that largest effect on exiting, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in hospitalists.
 
 Finally, I consider whether EHR exposure affects claims filed per patient.
-On average, physicians file roughly 4 claims per patient.
+On average, hospitalists file roughly 4 claims per patient.
 I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time.
-The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as physicians are exposed to the new software, they bill for more items on average.
-I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger physicians are exhibiting more of an increase in claim count than older physicians.
+The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as hospitalists are exposed to the new software, they bill for more items on average.
+I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists.
 
@@
 Anticipation and limiting sample years cause the effect to disappear or even become negative.
-Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs since the effects are not necessarily robust to many specifications.
+Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs.
 
@@
 \centering
-    \captionsetup{width=.7\linewidth}
+    \captionsetup{width=.85\linewidth}
     \caption{Effect of EHR Exposure on Claims per Patient}
@@
    \caption{all physicians}
-   \includegraphics[scale=.45]{Objects/claim_per_patient_plot_all.pdf}
+   \includegraphics[scale=.55]{Objects/claim_per_patient_plot_all.pdf}
    \label{fig:claima} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.45]{Objects/claim_per_patient_plot_ages.pdf}
+   \includegraphics[scale=.55]{Objects/claim_per_patient_plot_ages.pdf}
    \label{fig:claimb}
@@
 A technology expected to revolutionize the health care system was the advancement of health information technology in the form of complex electronic health record (EHR) systems.
-While they have the potential to improve efficiency, cut costs, and increase quality of care, these have not come to fruition.
+While they have the potential to improve efficiency, cut costs, and increase quality of care, these results have not come to fruition as expected.
 While the effect of EHRs on these outcomes has been directly studied, there is a gap in understanding how physicians, the primary user of this technology, responded to their rapid implementation of EHRs.
-In this study, I analyze the effect of implementing electronic health records on various physician labor market outcomes, since EHRs have been controversial and speculated as a cause of physician burnout.
-Understanding how physicians respond to this technology is important, as physician labor markets have implications for access to care, quality of care, and costs.
+In this study, I analyze the effect of implementing electronic health records on various labor market choices of hospitalists, since EHRs have been controversial and speculated as a cause of burnout.
+Understanding how hospitalists respond to this technology is important, as physician labor markets have implications for access to care, quality of care, and costs.
 
 I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity to analyze this relationship.
-Using a difference-in-differences framework, I find that physicians are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
-Further, I find evidence that physicians who do not retire choose to change where they work because of EHR exposure, particularly shifting from hospitals towards offices.
-Finally, physicians who utilize EHRs exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
-
-The implications of these results are threefold.
-First, many physicians left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care issues.
+Using a difference-in-differences framework, I find that hospitalists are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
+Further, I find evidence that hospitalists who do not leave clinical settings choose to change where they work because of EHR exposure, particularly shifting from hospitals towards offices/outpatient settings.
+Finally, hospitalists who utilize EHRs exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
+
+There are various implications of these findings.
+First, many hospitalists left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care.
 Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research.
-Third, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens.
-When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation and incorporate this into the incentives.
+Third, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens in the workplace.
+When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation.
 Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient.
 My findings suggest that EHRs make physicians more productive through the number of patients they see.
-However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, so a future research question is whether this increase in productivity is traded off with time spent with patients.
+However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, as there may be a trade-off between comprehensive patient care and productivity.
+This question is left to future research.
 
@@
 I then categorize NPIs by key words in their tax code description.
-Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as such.
+Any description containing ``Internal Medicine", ``Hospitalist", ``Family Medicine", or ``General Practice" is classified as a primary care physician (PCP), and any description containing ``hospital" is classified as a hospital.
 I create two data sets, one containing only PCPs and one containing only hospitals.
@@
 
-For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}}.
-I use data that captures shared patient activity in 30 day intervals.
+For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals.\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}} I use data that captures shared patient activity in 30 day intervals.
 I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals.
@@
 For hospital level variables, I use an AHA-NPI crosswalk to merge the pairs to the Annual Hospital Administration Survey (AHA Survey) from 2009-2015.
-From this data I collect information on each hospital's EHR use.
+Further, I use a hospital's Medicare ID number to merge this data to the Healthcare Information and Management Systems Society (HIMSS) data.
+From these data sets I collect information on each hospital's EHR use.
 There are 4,253 unique AHA hospitals in the shared patient data.
-I drop any hospitals that aren't in the AHA Survey due to lack of information on EHR use.
-After limiting the hospitals, there are 780,000 observations of pair-years left.
+After limiting to these hospitals, there are 780,000 observations of pair-years left.
+I define EHR use as either answering in the AHA Survey that the hospital uses an EHR fully, or reporting which vendor is used in the HIMSS Survey.
 Next, I investigate the hospitals with missing information for EHR use.
@@
 
-Specifically, I use the general survey question about each hospital's EHR use to define my key treatment variable.
-The survey asks the extent to which the hospital utilizes an EHR and allows for three answers: not at all, partially utilize an EHR, and fully utilize an EHR.
-I create a binary variable equal to one if the answer is fully utilize, zero otherwise.
 The data will be aggregated up to the physician level due to physician level outcome variables, so I first transform the pair-level EHR variable into a treatment variable at the physician level.
@@
 Finally, I aggregate to the physician level by only keeping variables that do not vary by hospital.
-The variables in the physician level data are year, physician NPI, graduation year, years of experience, number of hospitals, minimum year exposed to EHR, minimum year exposed to EHR in low integration hospital, number of systems, and never works with a new NPI.
+The variables in the physician level data are year, physician NPI, graduation year, years of experience, number of hospitals, minimum year exposed to EHR, , number of systems, and never works with a new NPI.
 I complete this data to include years 2016 and 2017, but leave time varying variables missing for those years.
@@
 Next, I create the dependent variables used in the analysis.
-The first outcome is whether a physician chooses to retire or not.
+The first outcome is whether a physician chooses to exit work in a clinical setting.
 First, I sum a physicians claim counts across zip codes into one variable for total claim count in a given year.
@@
 Then I create a variable for the first year that a physician has a future claim count of zero.
-However, this counts retirement one year too early.
-For example, if a physician has no claims from 2014 to 2017, the minimum year variable is set to 2013, but the first year of retirement is 2014.
+However, this counts exit one year too early.
+For example, if a physician has no claims from 2014 to 2017, the minimum year variable is set to 2013, but the first year of exit is 2014.
 Thus, I add one year to the minimum year variable.
-I then define the outcome variable, retire, as one if a physician ever retires and the year is equal to the first year of retirement.
+I then define the outcome variable, exiting, as one if a hospitalist ever exits and the year is equal to the first year of exit.
 I also create a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.
@@
 The first decision to make is how to handle years where a physician has missing data for claims.
-A year of missing data can mean the physician had no claims that year, or the observation was removed due to insufficient claims (less than 11).
+A year of missing data can mean the hospitalist had no claims that year, or the observation was removed due to insufficient claims (less than 11).
 A conservative assumption is to assume a year of missing data means zero claims, thus, I set all missing claim counts to zero.
-The data already contains a variable for the fraction of patients seen in an office setting, which is used in the analysis, but I further create an indicator variable for whether the physician sees any positive amount of patients in an office setting.
+The data already contains a variable for the fraction of patients seen in an office setting, which is used in the analysis, but I further create an indicator variable for whether the hospitalist sees any positive amount of patients in an office setting.
 Finally, the variables for unique patient count and claim count are already defined.
@@
 
-Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference in difference designs.
+Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference-in-differences.
 In my main specification I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
@@
 
-\subsubsection{Retirement}
-
-First, I consider the estimated effect of EHR exposure on retirement.
+\subsubsection{No Longer Seeing Patients}
+
+First, I consider the estimated effect of EHR exposure on leaving clinical settings, or exiting.
 While visually there does not appear to be a strong pre-trend, for the full sample of physicians and physicians $< 60$, the p-value indicates a pre-trend may exist.
@@
     \label{fig:cpp_chart}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    \captionsetup{width=.6\linewidth}
+    \caption{Effect of EHR Exposure on Patient Count by Group}
+    \includegraphics[scale=.4]{Objects/patient_group.pdf}
+    \label{fig:patientgroup}
+    \vspace{2mm}
+    \caption*{\footnotesize{\textit{Notes:} Plot shows average group time treatment effects for each treatment group.}}
 \end{figure}

### 2023-05-31 Scripts/specification_chart.R
update specification charts

@@ -450,7 +450,7 @@ Aggregated_Pairs_list <- lapply(Data_List, function(x){
   data <- x %>%
     distinct(year, DocNPI,grad_year, avg_beds, experience, 
              num_hospitals,
-             frac_EHR, minyr_EHR, minyr_EHR_int,
+             frac_EHR, minyr_EHR,
              num_systems, 
              newnpi, never_newnpi)
   
@@ -464,15 +464,13 @@ Aggregated_Pairs_list <- lapply(Aggregated_Pairs_list, function(x){
   data <- data %>%
     group_by(DocNPI) %>%
     fill(minyr_EHR,.direction="downup") %>%
-    fill(minyr_EHR_int,.direction="downup") %>%
     fill(grad_year,.direction="downup") %>%
     fill(avg_beds,.direction="downup") %>%
     fill(num_hospitals,.direction="downup") %>%
     fill(num_systems,.direction="downup") %>%
     ungroup() %>%
     mutate(experience=year-grad_year) %>%
-    mutate(anyEHR_exposed=ifelse(minyr_EHR>0 & year>=minyr_EHR,1,0),
-           anyEHR_LI_exposed=ifelse(minyr_EHR_int>0 & year>=minyr_EHR_int,1,0))
+    mutate(anyEHR_exposed=ifelse(minyr_EHR>0 & year>=minyr_EHR,1,0))
 
   
   return(data)
@@ -761,16 +759,16 @@ Physician_Data_list <- lapply(Physician_Data_list, function(x){
   
   # Create age bins to do analysis separately 
   Physician_Data <- Physician_Data %>% ungroup() %>%
-    select(DocNPI, year, grad_year, minyr_EHR, minyr_EHR_int, never_newnpi,
+    select(DocNPI, year, grad_year, minyr_EHR, never_newnpi,
            pos_office, npi_unq_benes, max_70, max_90, max_50, claim_count_total, retire,
-           ever_retire, work_in_office, change_zip, num_hospitals, max_age)
+           ever_retire, work_in_office, num_hospitals, max_age)
   
   
   return(Physician_Data)
 })
 
 data_coms <- crossing(inpat_70=c(TRUE, FALSE), inpat_90=c(TRUE, FALSE),
-                      inpat_50=c(TRUE, FALSE), main_EHR=c(TRUE, FALSE),
[…]

### 2023-05-31 Paper/PaperEdit.tex
Updates from Overleaf

@@
 
-Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) reveals the need to carefully assess the parallel trends assumptions commonly used in difference-in-differences.
+Work by \citeauthor{rambachan2019honest} (\citeyear{rambachan2019honest}) points to the need to carefully assess the parallel trends assumptions commonly used in difference-in-differences.
 In my main specification I assume that, conditional on covariates, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods.
@@
 
-First, I assess the assumption that exposure to an EHR is exogenous to the physician.
-The main specification of the paper defines a binary treatment variable for whether a physician is exposed to an EHR in any hospital they work in.
-These results are biased if physicians endogenously select whether they are exposed to EHRs, particularly by participating in the hospital's decision to implement.
-The results do not indicate that this is a huge issue on average, since we would expect physicians not to respond on the margin of retirement or switching work setting if they are choosing to be exposed to EHRs.
-Especially in larger hospitals, this is unlikely to occur as an individual physician will not have much input to high level decisions.
-However, as an attempt to provide more evidence of labor market changes as a result of EHR exposure, I provide results for specifications with an alternate binary treatment variable: whether a physician has been exposed to an EHR in a hospital considered to have low vertical integration with physicians.
-
-I define a hospital's level of integration based on organizational structure, where independent practice associations (IPAs) and physician-hospital organizations (PHOs) are classified as low vertical integration and other organization types are classified as high vertical integration (\cite{dynan1998assessing}).
-
 While there is evidence to suggest the time it takes to implement an EHR does not exceed one year, there still may be some anticipation if physicians are expecting implementation in the hospitals they work in.
@@
 
-\subsubsection{Retirement}
+\subsubsection{Leaving Clinical Setting}
 
-The specification chart for the retirement outcome is presented in Figure \ref{fig:retire_chart}.
-The main specification yields an average treatment effect of .0025 (significant at all levels).
+The specification chart for the exiting outcome is presented in Figure \ref{fig:retire_chart}.
+The main specification yields a statistically significant average treatment effect of .0025.
 This coefficient is on the higher end of coefficients from all combinations of specifications.
@@
 Some noisy estimates are to be expected when there are multiple limitations placed on the data.
-The average confidence interval for all combinations is (.001,.003), so the main specification lies in this interval.
-These results support my main finding that EHR exposure positively affects the likelihood of retirement among physicians.
+These results support the main finding that EHR exposure positively affects the likelihood of leaving a clinical setting among hospitalists.
 
 \begin{figure}[ht]
-    \centering
-    \caption{Outcome: Retire}
-    \includegraphics[scale=.6]{Objects/retire_chart.pdf}
+    \caption{Outcome: Exit Clinical Setting}
+    \includegraphics[scale=.7]{Objects/retire_chart.pdf}
     \label{fig:retire_chart}
@@
 This is on the high end of all combinations of changes to specification, but all reveal a positive effect.
-On average, the confidence interval falls between .035 and .06.
-This supports the finding that EHR exposure positively affects the likelihood of working in an office.
-
-Specification changes that seem to affect the magnitude of the estimate are whether I limit the sample years and which EHR definition is used as the treatment variable.
-Neither of these change the sign of the estimate, but both limiting sample years and using the low-integration definition of EHR exposure push the magnitude slightly down.
 
 \begin{figure}[ht]
-    \centering
     \caption{Outcome: Work in Office}
-    \includegraphics[scale=.6]{Objects/office_ind_chart.pdf}
+    \includegraphics[scale=.7]{Objects/office_ind_chart.pdf}
     \label{fig:work_chart}
@@
 
-Now I discuss the results for whether physicians change the fraction of patients seen in an office setting due to EHR exposure.
+Now I discuss the results for whether hospitalists change the fraction of patients seen in an office setting due to EHR exposure.
 Similarly, the ATT from the main specification is on the high end of other ATTs found, but all estimates suggest a positive effect.
@@
 \begin{figure}[ht]
-    \centering
     \caption{Outcome: Fraction of Patients in Office}
-    \includegraphics[scale=.6]{Objects/office_frac_chart.pdf}
+    \includegraphics[scale=.7]{Objects/office_frac_chart.pdf}
     \label{fig:fracoffice_chart}
@@
 In Figure \ref{fig:pat_chart}, I present estimates and confidence intervals for different combinations of specification changes.
-The results from the main specification is that EHR exposure increases patient count by 28.
+The result from the main specification is that EHR exposure increases patient count by 28.
 While most specifications yield positive estimates, about a third of the specifications have a confidence interval that includes zero.
@@
 \begin{figure}[ht]
-    \centering
     \caption{Outcome: Patient Count}
-    \includegraphics[scale=.6]{Objects/patient_chart.pdf}
+    \includegraphics[scale=.7]{Objects/patient_chart.pdf}
     \label{fig:pat_chart}
@@
 \begin{figure}[ht]
-    \centering
     \caption{Outcome: Claims per Patient}
-    \includegraphics[scale=.6]{Objects/claim_chart.pdf}
+    \includegraphics[scale=.7]{Objects/claim_chart.pdf}
     \label{fig:cpp_chart}

### 2024-03-20 Scripts/analysis.R
update figures for JHE R&R

@@ -19,7 +19,7 @@ library(extrafont)
 # In this portion, I only consider "retire" as the dependent outcome and I compare estimators. 
 
 # Read in the main data created in "data5_MDPPAS.R"
-Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data2.rds"))
+Physician_Data <- readRDS(paste0(created_data_path,"Physician_Data.rds"))
 
 Physician_Data <- Physician_Data %>%
   mutate(DocNPI=as.double(DocNPI))
@@ -28,7 +28,7 @@ Physician_Data <- Physician_Data %>%
 ## ANALYSIS ---------------------------------------------- #########
 ## Write results for each variable in a loop and create graphs ----- ##
 
-varlist <- list("retire", "pos_office", "work_in_office", "npi_unq_benes", "claim_per_patient")
+varlist <- list("retire", "pos_office", "work_in_office", "npi_unq_benes", "claim_per_patient", "only_office")
 
 # Get results for ATTGT 
 models <- lapply(varlist, function(x) {
@@ -38,7 +38,7 @@ models <- lapply(varlist, function(x) {
     tname = "year",                  # Time Variable
     # xformla = NULL                 # No covariates
     xformla = ~grad_year,            # Time-invariant controls
-    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="work_in_office") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
+    data= if (x=="retire") dplyr::filter(Physician_Data,minyr_EHR>0) else (if (x=="pos_office" | x=="work_in_office" | x=="only_office") dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0) else dplyr::filter(Physician_Data, minyr_EHR>0 & ever_retire==0 & never_newnpi==1)),
     est_method = "dr",               # dr is for doubly robust. can also use "ipw" (inverse probability weighting) or "reg" (regression)
     control_group = "notyettreated", # Set the control group to notyettreated or nevertreated
     clustervars = "DocNPI",          # Cluster Variables          
@@ -244,6 +244,117 @@ ggdid(models[[5]][["all"]]) + scale_color_manual(labels = c("Pre", "Post"), valu
[…]

### 2024-03-20 Scripts/data4_physician_level.R

@@ -62,19 +62,61 @@ data <- data %>%
 
 # Read in main AHA survey data
 # I need to think about adding more hospital characteristics to this dataset
-AHAmainsurvey <- read_csv(paste0(raw_data_path,"AHA_mainsurvey.csv"))
+AHAmainsurvey <- read_csv(paste0(raw_data_path,"allAHAvariables_20092017.csv"))
 
 AHAmainsurvey <- AHAmainsurvey %>%
+  select(ID, YEAR, MCRNUM, MNAME, MADMIN, SYSID, SUBS, EHLTH, 
+         BDTOT, FTMDTF, FTRES, PTMDTF, PTRES, JNTPH, PHYGP,
+         IPAHOS, OPHOSHOS, CPHOHOS, ISMHOS, IPAP, OPHP, CPHP, ISMP) %>%
+  filter(!is.na(MCRNUM)) %>%
   mutate(ID=as.character(ID)) %>%
-  dplyr::rename(Hospital_name=MNAME, year=YEAR, full_year=FYR, 
-         joint_phys=JNTPH, total_physicians=FTMT, phys_owned=PHYGP, beds=BDTOT ) %>%
-  select(-IPAHOS, -OPHOHOS, -CPHOHOS, -GPWWHOS, -MSOHOS, -ISMHOS, -EQMODHOS) %>%
-  select(-IPAP, -GPWP, -OPHP, -CPHP, -ISMP, -EQMP, -FNDP, -full_year, -EHLTHI, -EHLTHS,
-         -total_physicians, -joint_phys, -phys_owned)
+  filter(YEAR<=2015)
+
+# drop hospitals that change system ID
+AHAmainsurvey <- AHAmainsurvey %>%
+  group_by(MCRNUM) %>%
+  mutate(minyr = min(YEAR),
+         maxyr = max(YEAR)) %>% 
+  mutate(lag_sys = dplyr::lag(SYSID)) %>%
+  mutate(change_sys = ifelse(is.na(SYSID) & !is.na(lag_sys),1,NA)) %>%
+  mutate(change_sys = ifelse(!is.na(SYSID) & is.na(lag_sys),1,change_sys)) %>%
+  mutate(change_sys = ifelse(!is.na(SYSID) & !is.na(lag_sys) & lag_sys!=SYSID, 1, change_sys)) %>%
+  mutate(change_sys = ifelse((YEAR==minyr),NA,change_sys)) %>%
[…]

### 2024-03-20 Scripts/data5_MDPPAS.R

@@ -27,9 +27,53 @@ for (i in 2009:2017) {
 
 MDPPAS <- rbind(MDPPAS2009, MDPPAS2010, MDPPAS2011, MDPPAS2012, MDPPAS2013, MDPPAS2014, MDPPAS2015, MDPPAS2016, MDPPAS2017)
 
+# Read in Physician Utilization Files for 2013-2015
+for (i in 13:15){
+  year <- read_csv(paste0(raw_data_path,"/MUP_PHY/MUP_PHY_R19_P04_V10_D",i,"_Prov_Svc_zip",".csv")) %>%
+    mutate(year=2000+i)
+  assign(paste0("MUP_PHY",i),year)
+}
+
+MUP_PHY <- rbind(MUP_PHY13, MUP_PHY14, MUP_PHY15)
+rm(MUP_PHY13, MUP_PHY14, MUP_PHY15, year, MDPPAS2009, MDPPAS2010, MDPPAS2011, MDPPAS2012, MDPPAS2013, MDPPAS2014, MDPPAS2015, MDPPAS2016, MDPPAS2017)
+
+# summarize the public use data to the physician level for specific claims I need
+MUP_PHY <- MUP_PHY %>%
+  group_by(Rndrng_NPI, year) %>%
+  mutate(total_claims = sum(Tot_Srvcs, na.rm=T)) %>%
+  ungroup()
+
+MUP_PHY <- MUP_PHY %>%
+  filter(str_detect(HCPCS_Cd, "^992"))
+
+# group HCPCS codes that are the same (with varied time components)
+MUP_PHY <- MUP_PHY %>%
+  mutate(group = ifelse(str_detect(HCPCS_Desc, "Critical care delivery critically ill or injured patient"), "crit_care_delivery", NA),
+         group = ifelse(str_detect(HCPCS_Desc, "Established patient office or other outpatient"), "est_patient_office", group),
+         group = ifelse(str_detect(HCPCS_Desc, "Hospital discharge day management"), "hosp_disch_day", group),
+         group = ifelse(str_detect(HCPCS_Desc, "Hospital observation"), "hosp_obs_care", group),
+         group = ifelse(str_detect(HCPCS_Desc, "Initial hospital inpatient care"), "initial_hosp_inpat", group),
+         group = ifelse(str_detect(HCPCS_Desc, "New patient office or other outpatient visit"), "new_patient_office", group),
[…]

### 2024-03-20 Scripts/heterog_chart.R

@@ -8,6 +8,7 @@ library(lubridate)
 library(did)
 library(ggplot2)
 library(readxl)
+library(dplyr)
 
 
 
@@ -46,24 +47,55 @@ Physician_Data <- Physician_Data %>%
 
 # Define different data sets
 Rural_Data <- Physician_Data %>%
-  filter(ever_rural==1)
+  dplyr::filter(ever_rural==1)
 Urban_Data <- Physician_Data %>%
-  filter(ever_urban==1)
+  dplyr::filter(ever_urban==1)
 Smallhosp_Data <- Physician_Data %>%
-  filter(small==1)
+  dplyr::filter(small==1)
 Largehosp_Data <- Physician_Data %>%
-  filter(large==1)
+  dplyr::filter(large==1)
+onehosp_data <- Physician_Data %>%
+  dplyr::filter(num_systems==1)
+manyhosp_data <- Physician_Data %>%
+  dplyr::filter(num_systems>1)
 
-data_coms <- crossing(rural=c(TRUE,FALSE), urban=c(TRUE,FALSE), small=c(TRUE,FALSE), large=c(TRUE,FALSE)) %>%
-  filter(rural==TRUE & urban==FALSE & small==FALSE & large==FALSE |
-           rural==FALSE & urban==TRUE & small==FALSE & large==FALSE |
-           rural==FALSE & urban==FALSE & small==TRUE & large==FALSE |
-           rural==FALSE & urban==FALSE & small==FALSE & large==TRUE |
-           rural==FALSE & urban==FALSE & small==FALSE & large==FALSE)
+data_coms <- tidyr::crossing(rural=c(TRUE,FALSE), urban=c(TRUE,FALSE), small=c(TRUE,FALSE), large=c(TRUE,FALSE), one=c(TRUE, FALSE), many=c(TRUE,FALSE)) %>%
+  filter(rural==TRUE & urban==FALSE & small==FALSE & large==FALSE & one == FALSE & many == FALSE |
+           rural==FALSE & urban==TRUE & small==FALSE & large==FALSE & one == FALSE & many == FALSE |
+           rural==FALSE & urban==FALSE & small==TRUE & large==FALSE & one == FALSE & many == FALSE |
+           rural==FALSE & urban==FALSE & small==FALSE & large==TRUE & one == FALSE & many == FALSE |
+           rural==FALSE & urban==FALSE & small==FALSE & large==FALSE & one == FALSE & many == FALSE |
[…]

### 2024-03-20 Scripts/hosp_analysis.R

@@ -1,0 +1,183 @@
+library(readr)
+library(dplyr)
+library(did)
+dodge <- position_dodge(width=0.3) 
+
+
+# read in hospital level data
+hosp_data <- readRDS(paste0(created_data_path, "hosp_data.rds"))
+
+# define the first year a hospital implements an EHR
+hosp_data <- hosp_data %>%
+  mutate(EHR=ifelse(EHLTH==2 | !is.na(EHR_HIMSS1),1,ifelse(EHLTH==0 | EHLTH==1,0,NA))) 
+
+minyr_EHR <- hosp_data %>%
+  filter(EHR==1) %>%
+  group_by(HospNPI) %>%
+  mutate(minyr_EHR = min(year)) %>%
+  ungroup() %>%
+  distinct(HospNPI, minyr_EHR)
+
+hosp_data <- hosp_data %>%
+  left_join(minyr_EHR, by="HospNPI")
+
+hosp_data <- hosp_data %>%
+  mutate(minyr_EHR = ifelse(is.na(minyr_EHR),0,minyr_EHR))
+
+# read in HCRIS data
+final_hcris_data <- read_csv(paste0(raw_data_path, "/final_hcris_data.csv")) %>%
+  select(-fy_start, -fy_end, -date_processed, -date_created, -source)
+
+# merge to main hospital data
+hosp_data <- hosp_data %>%
+  left_join(final_hcris_data, by=c("MCRNUM"="provider_number", "year"))
+
+# create summary stats by year
+means <- hosp_data %>%
+  group_by(year) %>%
+  summarise_all(mean, na.rm=T)
+
+# create change in behavior variables
+hosp_data <- hosp_data %>%
+  separate(MADMIN, sep = ",", into = c("admin_name", "other")) %>%
+  select(-other) 
+
+hosp_data <- hosp_data %>%
+  mutate(count=1) %>%
+  group_by(MCRNUM) %>%
+  mutate(sum = sum(count)) %>%
+  filter(sum>=7) %>%
+  mutate(lag_name = dplyr::lag(admin_name)) %>%
+  mutate(lag_sys = dplyr::lag(SYSID)) %>%
+  ungroup() %>%
+  mutate(change_admin = ifelse(admin_name==lag_name,0,1),
+         change_sys = ifelse((is.na(SYSID) & !is.na(lag_sys)) | (!is.na(SYSID) & is.na(lag_sys)),1,0),
+         change_sys = ifelse(year==2009,0,change_sys)) %>%
[…]

### 2024-03-20 Scripts/sensitivity_analysis.R

@@ -7,6 +7,9 @@ library(did2s)
 library(readxl)
 library(HonestDiD)
 library(mvtnorm)
+library(eventstudyr)
+library(dplyr)
+library(modelsummary)
 
 
 cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@ -46,12 +49,13 @@ for (d in 2010:2015){
 
 stacked_data_1 <- rbind(subgroup_2010, subgroup_2011,subgroup_2012,subgroup_2013,subgroup_2014, subgroup_2015)
 
-
-models_stacked <- lapply(varlist, function(x){
-  feols(xpd(..lhs ~ treated_unit:rel_m1 + treated_unit:rel_0 + treated_unit:rel_p1 + experience | DocNPI:minyr_EHR, ..lhs = x),
+models_stacked_IV <- lapply(varlist, function(x){
+  feols(xpd(..lhs ~ 1 | DocNPI + year | anyEHR_exposed ~ weightedavg_exposure, ..lhs = x),
         cluster="DocNPI",
         data=stacked_data_1)
 })
+
+etable(models_stacked_IV[[1]], stage=1, fitstat='ivf', tex=TRUE)
 
 
 ATT_stacked <- lapply(models_stacked, function(x){
@@ -622,5 +626,21 @@ observe <- leebounds %>%
   filter(pctile>22) %>%
   filter(treat==0)
 summary(observe$claim_per_patient)
+
+
+
+# EVENT STUDY WITH LEAST WIGGLY CONFOUNDER
+wiggly <- lapply(varlist, function(x){
+  es <- EventStudy(estimator = "OLS",
+                   data = dplyr::filter(Physician_Data, minyr_EHR>0 & minyr_EHR!=2009),
+                   outcomevar = x,
+                   policyvar = "anyEHR_exposed",
+                   idvar = "DocNPI",
+                   timevar = "year",
+                   pre = 0, post = 2)
+  plot <- EventStudyPlot(estimates = es)
+  return(plot=plot)
+})
+
 
 

### 2024-03-20 Scripts/sumstats.R

@@ -6,6 +6,9 @@ library(ggplot2)
 library(readr)
 library(showtext)
 library(panelView)
+library(maps)
+library(mapdata)
+library(cdlTools)
 
 
 options(knitr.kable.NA=" ")
@@ -22,7 +25,7 @@ cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"
 # The data used is created in "Final_Pairs.R"
 
 # Read in Final_Pairs_Variables.rds
-Physician_Data <- read_rds(paste0(created_data_path,"Physician_data2.rds"))
+Physician_Data <- read_rds(paste0(created_data_path,"Physician_Data.rds"))
 
 
 
@@ -158,6 +161,119 @@ ggplot(sum_stats_year,aes(x=year,y=value,color=Variable, linetype=Variable)) +
         text=element_text(size=20)) 
 
 ggsave("Objects/sum_stats_year.pdf", width=10, height=7, units="in")
+
+
+# summary stats for levels of integration:  ----------------------------------------------------------------
+
+sum_stats_ipa <- Physician_Data %>% ungroup() %>% filter(phys_ever_IPA==1 & minyr_EHR>0) %>%
+  summarise_at(c("Number of Hospitals Worked With"="num_hospitals",
+                 "Female"="female", "Number of Systems Worked With"="num_systems",
+                 "Age"="age",
+                 "Year of EHR Exposure"="minyr_EHR",
+                 "Ever Exit Clinical Work"="ever_retire",
+                 "Fraction Patients in Office"="pos_office",
+                 "Work in an Office"="work_in_office",
+                 "Year of EHR Exposure"="minyr_EHR",
+                 "Claims per Patient"="claim_per_patient",
+                 "Number of Patients"="npi_unq_benes"), 
+               list(m=mean,sd=sd), na.rm=TRUE) %>%
+  mutate_if(is.numeric, ~ifelse(abs(.)==Inf,NA,.))  %>%
+  gather(key=var,value=value) %>%
+  extract(col="var",into=c("variable", "statistic"), regex=("(.*)_(.*)$")) %>%
+  spread(key=statistic, value=value) %>%
+  relocate(variable,m,sd) 
+
[…]

### 2024-03-20 Paper/coverletter.tex

\documentclass{article}
% Define WSU official colors
\usepackage[dvinames]{xcolor}
\definecolor{EmoryBlue}{RGB}{0,43,127} % Emory PMS 280
\definecolor{EmoryBlack}{RGB}{0,0,0}

% Measurements
\usepackage[top=2in,left=1in,bottom=0.5in,right=0.625in]{geometry}
\usepackage{graphicx}
\usepackage[colorlinks=false,
            pdfborder={0 0 0},
            ]{hyperref}
\usepackage[absolute]{textpos}
\usepackage{ifthen}
\usepackage{soul}

% --- A nice serif font (palatino), but not the prescribed nonfree ITC stone
\linespread{1.05}

% Remove paragraph indentation
\parindent0pt
\setlength{\parskip}{0.8\baselineskip}
\raggedright
\pagestyle{empty}
% Ensure consistency in the footer
\urlstyle{sf}

\providecommand\Emoryfrom{\textbf{Emory University}}
\providecommand\Emoryfrombuilding{Rich Memorial Building}
\providecommand\Emoryfromstreet{1602 Fishburne Drive, Room 306}
\providecommand\Emoryfromcity{Atlanta, GA 30322}
\providecommand\Emoryfromtel{404.727.6364 (tel)}
\providecommand\Emoryfromfax{404.727.4639 (fax)}
\providecommand\Emoryfromemail{\url{ian.mccarthy@emory.edu}}
\providecommand\Emorytoname{}
\providecommand\Emorytoaddress{}
\providecommand\Emoryre{}
\providecommand\Emorydate{\today}
\providecommand\Emoryopening{Dear Editors,}
\providecommand\Emoryfromname{Ian McCarthy}
\providecommand\Emoryfromtitle{Associate Professor}
\providecommand\Emoryclosing{Sincerely}

% Update this and the next line to the correct path
\providecommand\Emorysignaturefile{McCarthySignature.jpg}
\providecommand\Emorylogofile{EmoryLetterhead}
\providecommand\Emoryenclosure{}

\usepackage{fancyhdr}
\pagestyle{fancy}

\renewcommand{\footrulewidth}{0pt}
\fancyfoot{}
\fancyfoot[L]{%
    {\footnotesize\color{black}\sffamily
    \hspace{-.15in}
    \begin{tabular}{ll}
      \Emoryfrom & \Emoryfromtel \\[-0.1\baselineskip]
      \Emoryfrombuilding & \Emoryfromfax \\[-0.1\baselineskip]
      \Emoryfromstreet & \\[-0.1\baselineskip]
      \Emoryfromcity   &
    \end{tabular}
   }\color{black}}

\fancyhead{}
\fancyhead[L]{%
    \begin{textblock*}{2in}[0.3066,0.39](1.5in,1in)
        \includegraphics[width=2in]{EmoryLetterhead.pdf}
    \end{textblock*}
    \begin{textblock*}{4in}[0,0](3in,.7in)
      \begin{tabular}{|p{4in}}
        \vspace{.1cm}
        Ian M. McCarthy \\
        Associate Professor, Department of Economics \\
        Research Associate, National Bureau of Economic Research \\
        \vspace{0cm}
      \end{tabular}
    \end{textblock*}
    }
\renewcommand{\headrulewidth}{0pt}

\AtBeginDocument{
    % Text lines should be less than 6in long
    \newgeometry{top=1.8in,left=1in,bottom=1.2in,right=1in}

    \Emoryre
    \bigskip

    \Emorydate
    \bigskip

    \Emorytoname\ifthenelse{\equal{\Emorytoname}{}}{}{\\}
    \Emorytoaddress
    \bigskip

    \Emoryopening\par
    }

\AtEndDocument{
    \par\vspace{2ex}
    \Emoryclosing,

    \ifthenelse{\equal{\Emorysignaturefile}{}}{\bigskip\bigskip}{
      \includegraphics[width=2in]{\Emorysignaturefile}\\[-0.2\baselineskip]
    }

    \Emoryfromname \\
%    \Emoryfromtitle\ifthenelse{\equal{\Emoryfromtitle}{}}{}{\\}
    \Emoryenclosure
    }

\begin{document}
Enclosed please find our manuscript, ``Multimarket Contact in Health Insurance: Evidence from Medicare Advantage,'' to be considered for publication in the \textit{Journal of Industrial Economics}. This paper contributes to the literature on cross-market consolidations, which is of increasing interest to regulatory authorities and policy makers. We consider multimarket contact (MMC) as one potential mechanism by which cross-market consolidations can influence firm behaviors even in markets not directly affected by the merger.

We argue that Medicare Advantage is an excellent market to study these issues, not only because of the size and increasing reliance on Medicare Advantage to provide Medicare benefits, but also because of the transparency of potential deviations from collusive behavior and because of the availability of standardized quality measures made publicly available by CMS. Since our analysis is based on publicly available data, we are happy to provide all data and underlying code files should our paper be accepted. Thank you for your consideration, and we look forward to receiving your comments and suggestions.
\end{document}

### 2024-03-20 Presentations/ASHE2023.tex

\documentclass[notes,11pt, aspectratio=169]{beamer}

\usepackage{pgfpages}
% These slides also contain speaker notes. You can print just the slides,
% just the notes, or both, depending on the setting below. Comment out the want
% you want.
\setbeameroption{hide notes} % Only slide
%\setbeameroption{show only notes} % Only notes
%\setbeameroption{show notes on second screen=right} % Both

\usepackage{helvet}
\usepackage[default]{lato}
\usepackage{array}

\usepackage{tikz}
\usepackage{verbatim}
\setbeamertemplate{note page}{\pagecolor{yellow!5}\insertnote}
\usetikzlibrary{positioning}
\usetikzlibrary{snakes}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{matrix,shapes,arrows,fit,tikzmark}
\usepackage{amsmath}
\usepackage{mathpazo}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{multimedia}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bbm}
\newcolumntype{d}[0]{D{.}{.}{5}}
\usepackage{subfigure}

\usepackage{changepage}
\usepackage{appendixnumberbeamer}
\newcommand{\beginbackup}{
   \newcounter{framenumbervorappendix}
   \setcounter{framenumbervorappendix}{\value{framenumber}}
   \setbeamertemplate{footline}
   {
     \leavevmode%
     \hline
     box{%
       \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right]{footlinecolor}%
%         \insertframenumber  \hspace*{2ex} 
       \end{beamercolorbox}}%
     \vskip0pt%
   }
 }
\newcommand{\backupend}{
   \addtocounter{framenumbervorappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumbervorappendix}} 
}

\usepackage{graphicx}
\usepackage[space]{grffile}
\usepackage{booktabs}

% These are my colors -- there are many like them, but these ones are mine.
\definecolor{sage}{RGB}{102,153,102}
\definecolor{yellow}{RGB}{255,173,1}
\definecolor{purple}{RGB}{153,102,153}

\hypersetup{
  colorlinks=false,
  linkbordercolor = {white},
  linkcolor = {sage}
}

%% I use a beige off white for my background
\definecolor{MyBackground}{RGB}{255,253,218}

%% Uncomment this if you want to change the background color to something else
%\setbeamercolor{background canvas}{bg=MyBackground}

%% Change the bg color to adjust your transition slide background color!
\newenvironment{transitionframe}{
  \setbeamercolor{background canvas}{bg=white}
  \begin{frame}}{
    \end{frame}
}

\setbeamercolor{frametitle}{fg=sage}
\setbeamercolor{title}{fg=sage}
\setbeamertemplate{footline}[frame number]
\setbeamertemplate{navigation symbols}{} 
\setbeamertemplate{itemize items}{$\rightarrow$}
\setbeamercolor{itemize item}{fg=sage}
\setbeamercolor{itemize subitem}{fg=sage}
\setbeamercolor{enumerate item}{fg=sage}
\setbeamercolor{enumerate subitem}{fg=sage}
\setbeamercolor{button}{bg=MyBackground,fg=purple,}

% If you like road maps, rather than having clutter at the top, have a roadmap show up at the end of each section 
% (and after your introduction)
% Uncomment this is if you want the roadmap!
\AtBeginSection[]
{
   \begin{frame}
       \frametitle{Roadmap of Talk}
       \tableofcontents[currentsection]
   \end{frame}
}
\setbeamercolor{section in toc}{fg=sage}
\setbeamercolor{subsection in toc}{fg=sage}
\setbeamersize{text margin left=1em,text margin right=1em}

\newenvironment{wideitemize}{\itemize\addtolength{\itemsep}{10pt}}{\enditemize}

\usepackage{environ}
\NewEnviron{videoframe}[1]{
  \begin{frame}
    \vspace{-8pt}
    \begin{columns}[onlytextwidth, T] % align columns
      \begin{column}{.58\textwidth}
        \begin{minipage}[t][\textheight][t]
          {\dimexpr\textwidth}
          \vspace{8pt}
          \hspace{4pt} {\Large \sc \textcolor{blue}{#1}}
          \vspace{8pt}

          \BODY
        \end{minipage}
      \end{column}%
      \hfill%
      \begin{column}{.42\textwidth}
        \colorbox{green!20}{\begin{minipage}[t][1.2\textheight][t]
            {\dimexpr\textwidth}
            Face goes here
          \end{minipage}}
      \end{column}%
    \end{columns}
  \end{frame}
}

\title[]{\textcolor{sage}{Labor Markets and Technological Change}}
\subtitle[]{Evidence from Electronic Health Records}
\author[]{Hanna Glenn\\PhD Candidate\\Emory University\\hanna.glenn@emory.edu}

\date{\today}

\begin{document}

%%% TIKZ STUFF
\tikzstyle{bigcirc} = [circle, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=sage!30]
\tikzstyle{medcirc} = [circle, minimum width=2.5cm, minimum height=1cm,text centered, draw=black, fill=sage!30]
\tikzstyle{smallcirc} = [circle, minimum width=2cm, minimum height=1cm,text centered, draw=black, fill=sage!30]
\tikzstyle{rect} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=purple!30]
\tikzstyle{text} = [rectangle, minimum width=2cm, minimum height=1cm, text centered, draw=black, fill=green!100]
\tikzstyle{invisible} = [rectangle, minimum width=.001cm, minimum height=.001cm, text centered, draw=black!0, fill=red!0]
\tikzstyle{arrow} = [thick,->,>=stealth]
%%%% END TIKZ STUFF

% Title Slide
\begin{frame}
\maketitle
\end{frame}

\begin{frame}{Technology in Health Care}
    \centering
    \includegraphics[scale=.4]{Objects/tech_graphic.jpg}

    \tiny{Source: \hyperlink{https://academyhealth.org/blog/2015-01/how-coverage-and-technology-interact}{https://academyhealth.org/blog/2015-01/how-coverage-and-technology-interact}}

\end{frame}

\begin{frame}{Electronic Health Records}
    \centering
    \includegraphics[scale=.4]{Objects/EHR_adoption_graph.PNG}

    \tiny{Source (data points): \hyperlink{https://www.healthit.gov/data/quickstats/non-federal-acute-care-hospital-electronic-health-record-adoption}{https://www.healthit.gov/data/quickstats/non-federal-acute-care-hospital-electronic-health-record-adoption}}
\end{frame}

\begin{frame}[fragile]{What is an Electronic Health Record (EHR)?}
Health information stored digitally in a software
                \vspace{2mm}
\begin{center}
\includegraphics[scale=.25]{Objects/cernerscreenshot.png}

\tiny{Source: \hyperlink{https://softwareconnect.com/ehr/epic/}{https://softwareconnect.com/ehr/epic/}}
\vspace{-4mm}
\end{center}
\end{frame}

\begin{frame}{What we know about EHRs already}
Quality of Care:
\begin{itemize}
    \item Haven't seen the improvements expected for median patients
    \item \scriptsize Buntin et al. (2011), Agha (2014), Freedman et al. (2015), McCullough et al. (2016), Meyerhoefer et al. (2016)
\end{itemize}

\vspace{6mm}

Cost and Efficiency:
\begin{itemize}
    \item Maybe some gains, but not in all cases
    \item \scriptsize Agha (2014), Dranove et al. (2014), Hitt \& Tambe (2016)
\end{itemize}

\vspace{6mm}

Productivity:
\begin{itemize}
    \item Depends on the setting, positive effects are small
    \item \scriptsize McCullough et al. (2013), Hitt \& Tambe (2016), Meyerhoefer et al. (2016)
\end{itemize}

\vspace{3mm}
\pause
\textcolor{purple}{this paper: \underline{Did EHR adoption in hospitals affect labor market behaviors of physicians in hospitals?}}
\end{frame}

\begin{frame}{Physician Response to EHRs}
\centering
\includegraphics[scale=.45]{Objects/EHR tweet.PNG}
\end{frame}

\begin{frame}{Physician Response to EHRs}
\centering
\includegraphics[scale=.5]{Objects/News Clip3.PNG}
\end{frame}

\begin{frame}{This Paper}
Did EHR implementation in hospitals affect labor market decisions of hospitalists?

\vspace{5mm}
\centering
\begin{tikzpicture}[node distance=2cm]
\node (inv1) [invisible] {};
\node (bigcirc) [bigcirc, left of=inv1, xshift=.5cm, yshift=.13cm] {Hospitalists: $\mathbbm{H}$};
\node (rect) [rect, below of=bigcirc, yshift=-.5cm] {EHR in Hospital};

\node (inv2) [invisible, right of=inv1, xshift=2.5cm] {};
\node (medcirc) [medcirc, left of=inv2, xshift=.7cm, yshift=.13cm] {$\mathbbm{H}\backslash\mathbbm{E}$};

\node (inv3) [invisible, right of=inv2, xshift=2.5cm] {};
\node (smallcirc) [smallcirc, left of=inv3, xshift=.4cm, yshift=.13cm]{$\mathbbm{H}\backslash\mathbbm{(E\cup\mathbbm{O}})$};

\node  (retire) [text, above of=bigcirc, xshift=2.5cm, yshift=.3cm, align=center] {Exit Clinical Work:  $\mathbbm{E}$\\(Ext. Margin)};
\node  (office) [text, above of=medcirc, xshift=2.5cm, yshift=.3cm, align=center] {Switch Job Setting: $\mathbbm{O}$};
\node  (prod) [text, below of=smallcirc, align=center] {Productivity\\(Int. Margin)};

\draw [arrow] (rect) -- (bigcirc);
\draw [arrow] (bigcirc) -- (medcirc);
\draw [arrow] (medcirc) -- (smallcirc);
\draw [arrow] (inv1) -- (retire);
\draw [arrow] (inv2) -- (office);
\draw [arrow] (smallcirc) -- (prod);
\end{tikzpicture}
\end{frame}

\begin{frame}[fragile]{Data}
Panel of hospitalists from 2009-2017 with information on whether their hospital uses an EHR and various labor market outcomes\\
                \vspace{3mm}
\begin{wideitemize}
    \item MDPPAS data: physician specialty, aggregation of Medicare patients seen
    \item CMS Shared Patient Data: connect hospitals and physicians by the number of patients they share
    \item AHA Survey \& HIMSS: Hospital level EHR information
\end{wideitemize}
\end{frame}

\begin{frame}{Identification Strategy}
    Difference in differences:
    \begin{itemize}
        \item Use variation in timing of hospitals adopting EHRs, compare hospitalists before and after exposure to the EHR
    \end{itemize}

                \vspace{5mm}

    \underline{Key assumption}: Hospital implementing EHR is exogenous shock to hospitalist behavior. That is, hospitalists are not selecting into exposure to EHR
\end{frame}

\begin{frame}{Estimation}
Treatment: Exposure to EHR\\
                \vspace{4mm}
Staggered treatment timing, Callaway and Sant'Anna (2021) average group time treatment effects:
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$

\vspace{-1mm}

\begin{wideitemize}
    \item Not-yet-treated units
    \item Aggregate to get familiar event study plot
\end{wideitemize}

\vspace{3mm}

\pause Assumptions:
\begin{enumerate}
    \item No reversal of treatment
    \item No anticipation of treatment (can be relaxed)
    \item Parallel trends
\end{enumerate} 
\end{frame}

\begin{frame}{EHRs and Exit}
Decision to no longer see patients is multidimensional: retirement? switching type of job?
                \vspace{2mm}
\begin{itemize}
    \item Retirement is thought of as a function of age and wealth, and can depend on shocks such as health or job loss\\ \scriptsize{Hall and Johnson (1980), Gordon and Blinder (1980), Hamermesh (1985)}
                \vspace{3mm}
                \normalsize

    \item On average, physicians plan to retire at age 60 but don't actually retire until age 69 \\ \scriptsize{Collier (2017)}
                \vspace{3mm}
                \normalsize

    \item 13\% of physicians are considering a career change\\ \scriptsize{Physicians Foundation (2016)}
                \normalsize

\end{itemize}

\vspace{5mm}
\textcolor{purple}{I measure this as no longer seeing Medicare patients\ (mean: 3\%)}
\end{frame}

\begin{frame}{Effect of EHR Exposure on Exit}
\label{Effect of EHR Exposure on Exit}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.5]{Objects/retire_plot_all.pdf}
\end{figure}
$\rightarrow$ 7\% and 15\% increase relative to the mean -- 100 additional hospitalists leaving

\hyperlink{Robustness: Retirement}{\beamerbutton{retire}}
\end{frame}

\begin{frame}{Effect of EHR Exposure on Exit: Age Groups}
\label{Effect of EHR Exposure on Retirement: Age Groups}
\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{Objects/retire_plot_ages.pdf}
\end{figure}
\end{frame}

\begin{frame}{EHRs and Job Setting}
Most physicians will not be induced to exit because of EHRs
\begin{itemize}
    \item Another way to avoid (or be exposed to) new technology is to switch location of practice
                \vspace{3mm}
    \item Literature shows job switching is closely tied to job satisfaction\\ \scriptsize Aklerlof, Rose and Yellen (1988), Chadi and Hetschko (2017)
                \normalsize
    \end{itemize}

    \vspace{5mm}
    \textcolor{purple}{Measured as whether fraction of patients seen in office setting is $>$ 0 (mean: 28\%)}
\end{frame}

\begin{frame}{Effect of EHR Exposure on Likelihood of Working in Office}
\label{Effect of EHR Exposure on Likelihood of Working in Office}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.5]{Objects/officeind_plot_all.pdf}
\end{figure}
$\rightarrow$ 15-38\% increase relative to the mean -- roughly 1-2,000 physicians who start seeing patients in an office\\
$\rightarrow$ Similar for both age groups

\hyperlink{Robustness: Indicator for Working in Office}{\beamerbutton{office ind.}}
\end{frame}

\begin{frame}{EHRs and Productivity}
Past research on information technology: productivity typically increases \\ \scriptsize (Brynjolfsson and Hitt 2003; Bartel, Ichniowski and Shaw 2007; Bloom, Sadun, and Reenen 2012b)
                \vspace{6mm}
                \normalsize

We can see how productivity is affected for hospitalists who stay in the same hospital after exposure to EHR
                \vspace{3mm}
\begin{itemize}
    \item Expected effect is ambiguous
\end{itemize}

\vspace{5mm}
\textcolor{purple}{Measured as number of Medicare patients seen (mean: 325)}
\end{frame}

\begin{frame}{Effect of EHR Exposure on Patient Count}
\label{Effect of EHR Exposure on Patient Count}
\label{Results: Patient Count}
\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{Objects/patient_plot_all.pdf}
\end{figure}
$\rightarrow$ 6-14\% increase relative to the mean\\ \hyperlink{patient count broken down by group}{\beamerbutton{groups}}

\hyperlink{Robustness: Patient Count}{\beamerbutton{patient count}}
\end{frame}

\begin{frame}{Summary and Implications}

\underline{Summary:} Hospitalists change labor market behavior as a result of technology change in work environment

                \vspace{5mm}

\underline{Implications}: 
\begin{itemize}
\item Unintended consequences of a policy leading to rapid technology adoption
\item Access to care
\item potentially has downstream affects for patient outcomes
\end{itemize}
\end{frame}

\begin{frame}{}
    \centering
    Thank you! \\
    hanna.glenn@emory.edu
\end{frame}

\begin{frame}[noframenumbering, label=ch1robustness]{Robustness}
\label{Robustness}
\begin{enumerate}
    \item Limiting years and using never-treated as control
            \vspace{3mm}
    \item Anticipation
            \vspace{3mm}
    \item Parallel trends
            \vspace{3mm}
    \item Other estimators
    \vspace{5mm}

\hyperlink{Lee (2009) Bounds}{\beamerbutton{Lee (2009) bounds}}
\hyperlink{TWFE}{\beamerbutton{TWFE}}
\hyperlink{Other Estimators for Staggered Treatment}{\beamerbutton{Estimators}}
\end{enumerate}
\end{frame}

% JUMP SLIDES! %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% first chapter jump slides %

\begin{frame}[noframenumbering]{Patient Count Broken Down by Group}
\label{patient count broken down by group}
\centering
    \includegraphics[scale=.25]{Objects/patient_group.pdf}
\hyperlink{Results: Patient Count}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Lee (2009) Bounds}
\label{Lee (2009) Bounds}
    EHR exposure affects whether physicians are included in the analysis (since I drop those who retire or move), indicating a nonrandom attrition problem for productivity analysis.
    \begin{itemize}
        \item 18\% dropped in control group
        \item 36\% dropped in treatment group 
    \end{itemize}
    Monotonicity $\rightarrow$ EHR exposure increases the likelihood of retiring or moving workplace settings and no physicians remain in the sample under treatment who would have retired or moved in the absence of EHR exposure

    \begin{itemize}
        \item Patient count $\in$ (55,125)
        \hyperlink{Robustness}{\beamerbutton{back}}
    \end{itemize}
\end{frame}

\begin{frame}[noframenumbering]{Robustness: Exit}
\label{Robustness: Retirement}
\centering
    \includegraphics[scale=.5]{Objects/retire_chart.pdf}
    \hyperlink{Effect of EHR Exposure on Exit}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Robustness: Indicator for Working in Office}
\label{Robustness: Indicator for Working in Office}
\centering
    \includegraphics[scale=.45]{Objects/office_ind_chart.pdf}
    \hyperlink{Effect of EHR Exposure on Likelihood of Working in Office}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Robustness: Frac. Patients in Office}
\label{Robustness: Frac. Patients in Office}
\centering
    \includegraphics[scale=.45]{Objects/office_frac_chart.pdf}
    \hyperlink{Effect of EHR Exposure on Fraction of Patients in Office}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Robustness: Patient Count}
\label{Robustness: Patient Count}
\centering
    \includegraphics[scale=.45]{Objects/patient_chart.pdf}
    \hyperlink{Effect of EHR Exposure on Patient Count}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Robustness: Claims per Patient}
\label{Robustness: Claims per Patient}
\centering
    \includegraphics[scale=.45]{Objects/claim_chart.pdf}
    \hyperlink{Effect of EHR Exposure on Claims per Patient}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{TWFE}
\label{TWFE}
\centering
    \includegraphics[scale=.3]{Objects/twfe_plot.pdf}
    \hyperlink{Robustness}{\beamerbutton{back}}
\end{frame}

\begin{frame}[noframenumbering]{Other Estimators for Staggered Treatment}
\label{Other Estimators for Staggered Treatment}
\centering
    \includegraphics[scale=.45]{Objects/estimators_plot.pdf}
    \hyperlink{Robustness}{\beamerbutton{back}}
\end{frame}

\end{document}

### 2024-03-25 Paper/PaperEdit.tex

@@
 
-\usepackage[backend=biber,style=authoryear,
-sorting=ynt,citestyle=authoryear]{biblatex}
+\usepackage[backend=biber,style=authoryear,citestyle=authoryear, natbib=true]{biblatex}
 \addbibresource{papercitations.bib}
@@
 \author{Hanna Glenn}
-%\date{\today}
 
@@
 \usepackage{times}
+
 \begin{document}
 
+\linespread{1.2}\title{\vspace{-0.5in} Labor Markets and Technological Change \\ \Large Evidence from Electronic Health Records}
+
+	\date{\today}
+
+	\author{\vspace{10mm}Hanna Glenn\footnote{Department of Economics, Emory University, 1602 Fishburne Drive, Atlanta, GA 30322, hanna.glenn@emory.edu.} }
+
 	\maketitle
-
-\begin{abstract}
-    Technology adoption in health care is of great importance to policy makers as some technologies are expected to radically improve the U.S.
-health care system.
-However, some new technologies can be burdensome for physicians and lead to frustration.
-I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S.
-hospitals, electronic health records (EHRs).
-I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes.
-I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
-Further, those exposed to EHRs see more patients due to the technology.
-These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
+	%\setlength{\droptitle}{-10pt}
+
+	\vspace{-0.2in}
+
+	\singlespacing\maketitle
+
+\begin{abstract}\small
+    Technology adoption in health care has the potential to improve patient outcomes and, in some cases, decrease costs.
+However, new technologies can also be burdensome for physicians to keep up with.
+I add to our understanding of how new technology can affect those who use it by investigating changes in physician behavior as a result of a major technology shift in US hospitals, electronic health records (EHRs).
+I treat EHR implementation in hospitals as an exogenous shock to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes.
+I find that hospitalists are more likely to retire, and more likely to shift work to an outpatient setting due to EHRs.
+However, those who continue using EHRs in the hospital see more patients due to the technology.
+These findings inform the full costs and benefits of technology adoption, relevant to both technology adopters and policymakers who incentivize technology adoption.
 \end{abstract}
 
-\vspace{1.5cm}
+\vspace{2in}
+
+	\noindent Keywords:
+
+	\noindent JEL Codes:
+
+	\onehalfspacing
+
+	\newpage
 
@@
 
-EHRs are computer systems that house digital medical records and sometimes have additional functionalities.
-EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+EHRs are computer systems that house digital medical records, with the capability to house more complex functionalities.
+EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{Meaningful use (MU) standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 The frustration of a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings.
-However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question.
+However, the extent to which this frustration imposes a meaningful cost is unknown, making physician response an empirical question.
 For a hospitalist, the hospital's decision to adopt an EHR is plausibly exogenous.
-Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical environments, measured based on no future claims, (2) substituting hospital work for office work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.
+Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical work altogether, measured based on no future claims, (2) substituting hospital work for office/outpatient work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.
 
@@
 Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for information on the patients seen by each hospitalist.
-The independent variable of interest is binary, capturing exposure to an EHR.\footnote{While there are ample nuances to EHRs, I abstract away from these as I focus on the inital shock of a new technology.} I estimate group-time treatment effects of EHR exposure on the physician decisions defined above.
+The independent variable of interest is binary, capturing exposure to an EHR.\footnote{While there are ample nuances to EHRs, I abstract away from these as I focus on the initial shock of a new technology.} Due to staggered treatment timing, I estimate group-time treatment effects of EHR exposure on the physician decisions defined above (\cite{callaway2021difference}).
 
@@
 There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use.
-Despite a large number of case studies that find generally positive effects, i.e., improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
+Despite a large number of case studies that find generally positive effects, i.e., improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), quantitative studies have found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}).
 A number of case studies consider the effect of EHRs on productivity.
@@
 I expand this literature first by considering individual physician labor market decisions as an outcome, which, to my knowledge, has not yet been examined.
-Further, I consider the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting physicians' daily life.
-Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework.
-I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects.
+Further, I consider the primary time period in which EHRs were rapidly implemented, in part due to the HITECH Act (\cite{dranove2015investment}).
+This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting hospitalists' daily life.
+Similar to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework.
+I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation of staggered difference-in-differences.
 
@@
 Limiting the sample to those who remained in clinical settings, I find that hospitalists were 30\% more likely to start seeing patients in an office setting at the time of exposure.
-Finally, I find that EHRs lead to a patient count increase of at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.
+I find that this is not necessarily driven by hospital allocation of hospitalists to outpatient facilities.
+Finally, I examine productivity of hospitalists who do not exit or change work settings, and I find that EHRs lead to a patient count increase of at least 20 patients per year.
+This increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.
 
@@
 First, I assume that hospitalists are constrained to use the technology implemented in their work setting.
-Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals.
-That is, a hospitalist employed by the hospital does not choose their own EHR exposure.
-My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed.
-Thus, EHR implementation may have intensified access to care issues for patients.
-This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.
-\textcolor{red}{reword the implications and check the numbers are correct}
+Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists, and uncorrelated with other major changes to the hospitalist work environment.
+I mitigate concerns of the latter by analyzing whether other observable hospital changes are correlated with EHR adoption, and find no evidence that this could be biasing my results.
+Additionally, my main findings are not sensitive to a plethora of robustness checks.
+The results suggest that a surge of hospitalists left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings.
+Thus, EHR implementation may have intensified access to care issues for hospitalized patients.
+This unintended consequence of incentivizing rapid technology adoption highlights the need for careful consideration of technologies and policies that might negatively affect work environment.
 
@@
 The US allocated subsidies according to stages: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, stage 3 (2014) extended to data sharing between practices.
+Any hospital that didn't have an EHR by 2015 was subject to penalties through lower Medicare reimbursements.
 Vast EHR adoption coincided with the rollout of these subsidies, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}).
@@
 Research on how physicians interact with EHRs is primarily focused on outpatient settings.
-On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming portions of EHR use (\cite{arndt2017tethered}).
+There are key differences between outpatient and inpatient EHRs which I will outline below.
+While most research is specific to physicians in outpatient settings, the limited research available which investigates hospitalists and EHRs seems to be consistent with these findings (\cite{tipping2010did}).
+Outpatient EHRs are designed to be much more simplistic than hospital EHRs that communicate across departments.
+An important aspect of modern US health care is the rapid integration of hospitals with outpatient facilities, but it is not clear whether acquired outpatient facility EHRs are interoperable with inpatient EHRs, even if the two have the same EHR vendor.
+The main purpose of an outpatient EHR is to track patient data, but it is unlikely this tracking covers the entire health system in the early years of the HITECH Act before interoperability incentives began.
+
+On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming aspects of EHR use (\cite{arndt2017tethered}).
 In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR.
@@
 The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}).
-Another common issue physicians face is the lack of usability of EHR systems, that functions take minutes to either locate or load.
+Another common issue physicians face is the lack of usability of EHR systems, with functions taking minutes to locate or load.
 Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
-While most research is specific to physicians in outpatient settings, the limited research available which investigates hospitalists and EHRs seems to be consistent with these findings (\cite{tipping2010did}).
+These things taken together imply that there could be a meaningful additional cost imposed on physicians using a new technology.
 
@@
 
-This section serves to outline the potential responses of hospitalists to EHRs based on their incentives.
+This section outlines the potential responses of hospitalists to EHRs based on their incentives and institutions.
 The behaviors that I consider in this analysis are closely related and should not be taken as independent.
@@
 
-For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting.
-However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist.
+For ease of notation, I define a hospitalist as any general practice physician who sees at least 70\% of their patients within a hospital setting.
+However, there are two types of physicians who can be categorized as hospitalists based on this definition: a formal hospitalist who is solely employed by the hospital, or an internist who contracts their services to the hospital.
 Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings.
 Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting.
-While this distinction is potentially important for some of the decisions outlined below, the data do not explicitely distinguish between the two.
-I will discuss this limitation in more detail when relevant.
+While this distinction is potentially important for some of the decisions outlined below, the data do not explicitly distinguish between the two for each individual hospitalist.
+However, I observe whether each hospital integrates with certain types of physicians and include this as supplementary analysis.
 
@@
 Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of planning necessary to formally leave the labor force.
-Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}).
+Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}; \cite{stock1990pension}).
 However, physicians typically make this decision differently than workers in other industries.
@@
 Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently.
-As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences.
+As a worker considers the retirement decision, they weigh their expected earnings with expected costs, along with some preferences.
 Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors.
@@
 Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction.
-Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.
+Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}; \cite{loria_2020}), but this has not been addressed empirically.
 
@@
 As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch.
-For the purpose of this paper, the term exit indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.
-
-\subsubsection{Work Setting}
-
-Most physicians are likely not on the margin in their decision to exit clinical work, so EHRs will not impose a large enough cost (or benefit) to impose exit.
+For the purpose of this paper, the term exit indicates no longer seeing patients.
+An interesting area of future work would be to investigate further the decisions of physicians who decide to switch careers.
+
+\subsubsection{Work Setting}\label{sec:worksetting}
+
+Most physicians are likely not on the margin in their decision to exit clinical work, so EHRs will not impose a large enough cost (or benefit) to affect exit probability.
 However, there are other ways hospitalists might change their behavior under drastic changes to their current working environment.
-For those who remain practicing, changing the physical place of work may allow for more control over which technologies they use.
+For those who remain practicing, different work settings could be in the choice set.
 Specifically, if EHRs change job satisfaction of working in the hospital, a hospitalist may decide to shift more of their working hours towards or away from the hospital.
-Not all physicians working as hospitalists have a binding contract with their associated hospital.
-In fact, a 2013 survey estimated that 36\% of hospitalists also worked outside of the hospital in primary care, long-term care, or hospice settings (\cite{Today}).
+This is relevant for hospitalists who contract out to hospitals, as not all physicians working as hospitalists have a binding contract with their associated hospital.
+A 2013 survey estimated that 36\% of hospitalists also worked outside of the hospital in primary care, long-term care, or hospice settings (\cite{Today}).
 Many physicians trained as internists choose to contract their work out to both hospitals and outpatient settings.
@@
 
+For hospitalists who do have a binding contract with a hospital, there could still be opportunities to shift work to outpatient settings owned by the hospital.
+Particularly, if EHRs change hospital incentives of where to allocate physicians, EHR adoption could affect work setting in a way that is outside of the hospitalists' control.
 The sample of physicians I consider include both hospitalists with formal contracts and internists who contract a majority of their time with the hospital, but could also work in other settings as well.
-If a hospital implements a costly EHR, a hospitalist may choose to shift more of their time towards outpatient settings or initiate a contract with outpatient settings to spend less time in the setting where job satisfaction has decreased.
+I investigate whether these groups differentially change work setting due to EHR through a heterogeneity analysis of hospitalists working in hospitals with different levels of integration with physicians.
 
@@
 
-For those not induced to change their behavior by exiting or changing work settings, it is natural to assess whether EHRs impacted productivity.
-A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.
+For those who do not change their behavior by exiting or changing work settings, it is natural to assess whether EHRs impacted productivity.
+A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burdens.
 This question is empirically ambiguous since EHRs could add daily tasks that make hospitalists less productive, increase efficiency, or some combination of the two where short term productivity is negatively impacted but long term productivity is positively impacted once physicians adjust and learn the new technology.
-I measure productivity in terms of the amount of patients a hospitalist is able to see.
 
@@
 There are many reported reasons that physicians tend to over-treat patients, including defensive medicine, patient request, difficulty accessing patient records, and pressure from their institution (\cite{lyu2017overtreatment}).
-EHRs have the potential to change some of these incentives.
-As more hospitals adopt EHRs, patient information should be more easily accessible and reduce the amount of repeated tests a physician needs to run, leading to a decrease in claims per patient.
-However, institutions have power to customize EHRs and train their hospitalists to increase fee-for-service revenue by increasing claims per patient.
+EHRs may change some of these incentives.
+As more hospitals adopt EHRs, patient information could be more easily accessible and reduce the amount of repeated tests a physician needs to run, leading to a decrease in claims per patient.
+However, institutions could have power to customize EHRs and train their hospitalists to increase fee-for-service revenue by increasing claims per patient.
 
@@
 
-I construct a balanced hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician and hospital characteristics.
-I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.
+I construct a balanced hospitalist-level panel data set spanning from 2009-2017 which measures physician exposure to EHRs over time, several labor market outcomes, and other characteristics.
+I describe the different data sets used to construct the panel below, and I include a more detailed outline of data and variable construction in Appendix \ref{app:data}.
 
@@
 Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.).
-I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting.
+I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 70\% of patients in inpatient setting.
 I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year.
-I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}).
+I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate robustness to this threshold in Appendix \ref{sec:changes}).
 I use this data to construct the dependent variables for my analysis.
-Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and billing activity.
-
-Center for Medicare and Medicaid (CMS) Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period.
+
+The Center for Medicare and Medicaid (CMS) Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period.
 These data are available from 2009 to 2015.
@@
 
-I aggregate the data to the physician level.
-I drop physicians who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated.
+From the AHA survey, I also gather information on the level of integration a hospital has with physicians.
+There are four categories of integration considered: (ranked from least to most integrated): independent practice association (IPA), open physician-hospital organization (OPHO), closed physician-hospital organization (CPHO), and integrated salary model (ISM).
+An IPA has loose contracts with physicians, but the physicians remain independent, an OPHO adds administration tasks to the physician, a CPHO contracts with specific physicians, and a FIO hires physicians as salaried employees (\cite{short2017integration}).
+Further, I limit to hospitailists not connected with a hospital involved in mergers and acquisitions over the time period.
+
+I aggregate the data to the hospitalist level.
+I drop those who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated.
 The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated".
 Finally, I include a physician's graduation year from Physician Compare (provided by CMS).
-I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
+I limit to hospitalists who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
 The details of such thresholds are explained in Appendix \ref{app:data}.
-To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations.
+To ensure that my results  are not sensitive to various choices of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations.
 The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.
@@
 The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as exit.
-Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year and zero in every other year.
-This captures a one time decision to stop seeing patients, not a cumulative decision in all following years.
+Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year.
+This captures a one-time decision to stop seeing patients, not a cumulative decision in all following years.
 The next variables I construct capture whether physicians change work setting.
@@
 That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who exit or change their work setting.
-I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstats}.
-
-\import{Table Code}{overall stats.tex}
+I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstatsnew}.
+
+\import{Table Code}{overall stats new.tex}
 
 \normalsize
-I follow a maximum of 19,263 hospitalists over time, depending on the sample considered.
-Out of all hospitalists in the sample, 3\% (approximately 578) stop seeing patients at some point.
+I follow a maximum of 15,465 hospitalists over time, depending on the sample considered.
+Out of all hospitalists in the sample, 3\% stop seeing patients at some point.
 A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards early years.
-For hospitalists in the office setting sample (those who do not exit), 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office.
-There are 12,732 hospitalists who never exit or change their work setting at any point in the sample.
-These hospitalists see 325 patients and file 4 claims per patient on average.
-Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.
+For hospitalists in the office setting sample (those who do not exit), 27\% of physicians have a positive number of patients seen in an office, and the average physician sees 8\% of their total patient volume in an office.
+Hospitalists who do not exit or switch work setting see 331 patients and file 4 claims per patient on average.
+Consistent across all samples, hospitalists tend to work approximately 1.3 hospitals and 1.2 systems.
 
 In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time and the underlying hospital adoption that drives this treatment variable.
-Hospitalists are exposed the first time one of their hospital's adopts an EHR.
+Hospitalists are exposed the first time one of their hospitals adopts an EHR.
 Thus, I show the percent of hospitals adopting as well as the percent of hospitalists exposed to the technology.
@@
 That is, the physician does not influence the hospital's decision to implement an EHR.
+Finally, I assume that EHR adoption is not correlated with other hospital changes that could also affect physician work environment.
+To mitigate concerns along these lines, I present a hospital level analysis in which observable hospital behaviors are not associated with EHR adoption in Section \ref{sec:hospanalysis}.
 
@@
 Being exposed to an EHR leads to a .2 ppt increase in the likelihood of exiting in the first year after exposure, and a .4 ppt increase in the second year after exposure.
+Coefficients remain similar in the years following, though noisier due to smaller sample not-yet-treated groups.
 While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03.
@@
     \captionsetup{width=.85\linewidth}
-    \caption{Effect of EHR Exposure on Retirement}
+    \caption{Effect of EHR Exposure on No Longer Seeing Patients}
    \begin{subfigure}[]{.8\textwidth}
    \caption{all physicians}
-   \includegraphics[scale=.55]{Objects/retire_plot_all.pdf}
+   \includegraphics[scale=.5]{Objects/retire_plot_all.pdf}
    \label{fig:retirefirst} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.55]{Objects/retire_plot_ages.pdf}
+   \includegraphics[scale=.5]{Objects/retire_plot_ages.pdf}
    \label{fig:retiresecond}
@@
 Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$).
-Interestingly, hospitalists >=60 are driving the positive estimate in the year after exposure, and younger hospitalists only exit in the second year after exposure.
-A retirement age physician is 1 ppt more likely to exit after being exposed to an EHR, a 25\% increase relative to the mean.
+Interestingly, hospitalists $>=60$ are driving the positive estimate in the year after exposure, and younger hospitalists only exit in the second year after exposure.
+A retirement age hospitalist is 1 ppt more likely to exit after being exposed to an EHR, a 25\% increase relative to the mean.
 In comparison, hospitalists less than 60 are not more likely to exit in the first year after exposure, but are approximately 16\% more likely to exit two years after exposure.
@@
 career switching.
-First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
+First, if younger hospitalists are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year.
 Similarly, younger hospitalists may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.
@@
 I still discuss the estimates for this age group, but focus on the results for the older hospitalists.
-I also present overall treatment effects under different data specifications in Appendix \ref{sec:chart}, and find that this result is robust to various specifications.
+I also present overall treatment effects under different data specifications in Appendix \ref{sec:chart}.
+Under various data and specification changes, the coefficients remain non-negative.
+Estimates become noisier under strict limitations of the data, but given the rarity of exiting, this positive effect still seems to be robust.
 
@@
    \caption{all physicians}
-   \includegraphics[scale=.55]{Objects/officeind_plot_all.pdf}
+   \includegraphics[scale=.5]{Objects/officeind_plot_all.pdf}
    \label{fig:officefirsta} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.55]{Objects/officeind_plot_ages.pdf}
+   \includegraphics[scale=.5]{Objects/officeind_plot_ages.pdf}
    \label{fig:officefirstb}
@@
 To continue investigating how different ages may respond differently to technology, I also present estimates for hospitalists less than 60 and greater than or equal to 60 separately in Figure \ref{fig:officefirstb}.
-Unlike the choice to exit, the age groups are both responding to EHR implementation in all years after exposure.
+Unlike the choice to exit, the age groups respond to EHR exposure similarly.
 The magnitudes for senior hospitalists continue to increase over time, whereas the estimates for younger physicians level out.
@@
    \caption{all physicians}
-   \includegraphics[scale=.55]{Objects/officefrac_plot_all.pdf}
+   \includegraphics[scale=.5]{Objects/officefrac_plot_all.pdf}
    \label{fig:officeseconda} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.55]{Objects/officefrac_plot_ages.pdf}
+   \includegraphics[scale=.5]{Objects/officefrac_plot_ages.pdf}
    \label{fig:officesecondb}
@@
 Further, under different combinations of data specifications I find consistently positive and significant results, as shown in Appendix \ref{sec:changes}.
+
+\subsubsection{Mechanisms}
+
+As discussed in Section \ref{sec:worksetting}, hospitalists could be switching to work settings either because of their own preferences to avoid technology, or because EHRs lead hospitals to re-allocate workers to hospital-owned outpatient settings.
+While I do not observe individual physician contract type with each hospital, I do observe hospital level information about the nature of aggregate physician relationships.
+To investigate the potential mechanism of hospital allocation, I present a heterogeneity analysis for all outcomes broken down by level of integration the physician is exposed to in their connected hospitals.
+From least integrated (contracts with multiple facilities) to most integrated (full salaried hospitalists), results are broken down by Independent Practice Association (IPA), Open Physician-Hospital Organization (OPHO), Closed Physician-Hospital Organization (CPHO), and Integrated Salary Model (ISM).
+In practice, hospitals can only re-allocate hospitalists to outpatient facilities if the hospitalist is heavily integrated.
+However, estimates do not differ between the types of integration relationships, indicating that the results are not driven entirely by such re-allocation.
+
+\begin{figure}
+    \centering
+    \caption{Effect of EHR Exposure on Outcomes by Integration Level}
+    \includegraphics[width=\textwidth]{Objects/heterog_plot_int.pdf}
+    \label{fig:het_int}
+\end{figure}
+
+Additionally, I investigate whether this finding is potentially due EHRs changing the type of billing performed by hospitalists.
+For example, maybe they are doing the same services in the same place but billing the differently.
+I do not find this to be the case.
+This analysis can be found in Appendix
 
@@
    \caption{all physicians}
-   \includegraphics[scale=.55]{Objects/patient_plot_all.pdf}
+   \includegraphics[scale=.5]{Objects/patient_plot_all.pdf}
    \label{fig:patienta} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.55]{Objects/patient_plot_ages.pdf}
+   \includegraphics[scale=.5]{Objects/patient_plot_ages.pdf}
    \label{fig:patientb}
@@
 I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists.
-
-There is no evidence of a pre-trend driving this result, and I also show robustness to violations in parallel trends in Appendix \ref{sec:pretrends}.
-I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample.
-The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33.
-The effect becomes even larger when accounting for a particular type of physician being left in the sample.
-However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, around half of the estimates are null.
+However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero.
 Anticipation and limiting sample years cause the effect to disappear or even become negative.
-Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs.
+Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs and I do not make any strong conclusions.
 
@@
    \caption{all physicians}
-   \includegraphics[scale=.55]{Objects/claim_per_patient_plot_all.pdf}
+   \includegraphics[scale=.5]{Objects/claim_per_patient_plot_all.pdf}
    \label{fig:claima} 
@@
 \caption{sub-samples of physicians based on age}
-   \includegraphics[scale=.55]{Objects/claim_per_patient_plot_ages.pdf}
+   \includegraphics[scale=.5]{Objects/claim_per_patient_plot_ages.pdf}
    \label{fig:claimb}
@@
 
+\section{Hospital Analysis}\label{sec:hospanalysis}
+
 \section{Conclusion}
@@
 I utilize data on shared patients between medical entities, hospital EHR use, and physician billing activity to analyze this relationship.
-Using a difference-in-differences framework, I find that hospitalists are more likely to leave clinical settings due to EHR exposure, where the magnitude for senior physicians is larger than that for younger physicians.
+Using a difference-in-differences framework, I find that hospitalists are more likely to exit clinical settings altogether due to EHR exposure, where the magnitude for senior hospitalists is larger than that for younger hospitalists.
 Further, I find evidence that hospitalists who do not leave clinical settings choose to change where they work because of EHR exposure, particularly shifting from hospitals towards offices/outpatient settings.
-Finally, hospitalists who utilize EHRs exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
+Finally, hospitalists who utilize EHRs for the entire sample period exhibit an increase in the number of patients seen that cannot fully be explained by redistributing patients or selection of physicians.
 
@@
 First, many hospitalists left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care.
-Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior physicians may have long term negative effects on patient outcomes, but this is a question left to future research.
-Third, physicians are exhibiting behavior that suggests they incur switching costs to avoid external burdens in the workplace.
+Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior hospitalists may have long term negative effects on patient outcomes, but this is a question left to future research.
+Third, hospitalists are exhibiting behavior that suggests they incur switching costs to avoid external burdens in the workplace.
 When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation.
 Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient.
-My findings suggest that EHRs make physicians more productive through the number of patients they see.
+My findings suggest that EHRs make hospitalists more productive through the number of patients they see.
 However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, as there may be a trade-off between comprehensive patient care and productivity.
@@
 First, I consider the estimated effect of EHR exposure on leaving clinical settings, or exiting.
-While visually there does not appear to be a strong pre-trend, for the full sample of physicians and physicians $< 60$, the p-value indicates a pre-trend may exist.
+While visually there does not appear to be a strong pre-trend, for hospitalists $< 60$, the p-value indicates a pre-trend may exist.
 I present a plot of robust confidence intervals under various possible assumptions of parallel trends violations for one year after exposure in Figure \ref{fig:pre_retire}.
@@
 This is true for both event time 1 and event time 2.
+Given the rarity of exiting in the sample, it is not surprising that this result is sensitive to further assumptions.
+However, I acknowledge the sensitivity of the result for this age group and focus mainly on hospitalists of retirement age.
 
@@
     \captionsetup{width=.5\linewidth}
-    \caption{Retire Pretrends Plot}
+    \caption{Exit Pretrends Plot}
     \includegraphics[scale=.5]{Objects/retire_pretrends_plot.pdf}
@@
 
-In the main sample, I define dependent variables using years 2009 to 2017 and drop any physicians who were not exposed to an EHR by 2015 due to data constraints.
+In the main sample, I define dependent variables using years 2009 to 2017 and drop any hospitalists who were not exposed to an EHR by 2015 due to data constraints.
 This leads to estimation based on not yet treated units, because all never treated units are dropped.
@@
 First, when deciding which hospitals a physician is working closely with, I drop any pairs that fall below 30 shared patients per year.
-The goal of this threshold is to eliminate pairs for which the physician does not have close ties to the hospital.
+The goal of this threshold is to eliminate pairs for which the hospitalist does not have close ties to the hospital.
 In alternate specifications, I lower the threshold to 10 patients per year and raise the threshold to 60 patients per year.
@@
 I present overall average treatment effects for each outcome variable under different combinations of the specification adjustments discussed in Section \ref{sec:changes}.
-On each chart, the main specification is labeled in green and the average confidence bands across all combinations is shown as an orange band.
+On each chart, the main specification is labeled in green and the average confidence bands across all combinations is shown in orange.
 
@@
 
-The specification chart for the exiting outcome is presented in Figure \ref{fig:retire_chart}.
-The main specification yields a statistically significant average treatment effect of .0025.
+The effect of EHR exposure on leaving clinical work under different data specifications is presented in Figure \ref{fig:retire_chart}.
+The main specification yields a marginally significant, positive average treatment effect of .0025.
 This coefficient is on the higher end of coefficients from all combinations of specifications.
-However, almost all of the coefficients are positive.
-Some noisy estimates are to be expected when there are multiple limitations placed on the data.
-These results support the main finding that EHR exposure positively affects the likelihood of leaving a clinical setting among hospitalists.
+However, every data choice leads to a non-negative treatment effect.
+While it is possible that the true effect is zero, a likely reason for such noisy estimates is the rarity of exiting in the sample combined with more restrictive data choices.
+The average confidence band supports a positive effect of EHR exposure on leaving clinical settings.
 
@@
 
-The specification charts for outcomes work in office and fraction of patients seen in office are presented in Figures \ref{fig:work_chart} and \ref{fig:fracoffice_chart}, respectively.
+The specification charts for outcomes regarding switching to outpatient settings are presented in Figures \ref{fig:work_chart} and \ref{fig:fracoffice_chart}, respectively.
 First, I discuss the overall average treatment effect of EHR exposure on the likelihood of working in an office.
-The main specification yields finding that EHR exposure leads to a .07 ppt increase in the likelihood of working in an office.
-This is on the high end of all combinations of changes to specification, but all reveal a positive effect.
+The main specification yields finding that EHR exposure leads to a .08 ppt increase in the likelihood of working in an office.
+This is on the high end of all combinations of changes to specification, but all reveal a positive and significant effect ranging from .03 to .09.
 
@@
 Similarly, the ATT from the main specification is on the high end of other ATTs found, but all estimates suggest a positive effect.
-The average of all the confidence intervals found is (.02, .06), suggesting a positive effect on fraction of patients seen in an office.
+The average of all the confidence intervals found is (.04, .055), suggesting a positive effect on fraction of patients seen in an office.
 Changes that affect the magnitude of the estimate (but not the sign) are limiting sample years and anticipation.
-Limiting the years in the sample and allowing for one year of anticipation both push the magnitude of the estimate down.
+Limiting the years in the sample and allowing for one year of anticipation both push the magnitude of the estimate down, but do not change the sign or significance.
 
@@
 In Figure \ref{fig:pat_chart}, I present estimates and confidence intervals for different combinations of specification changes.
-The result from the main specification is that EHR exposure increases patient count by 28.
-While most specifications yield positive estimates, about a third of the specifications have a confidence interval that includes zero.
-It doesn't seem as though any specific change is driving the null result, so I hypothesize that it is due to the reduction in observations due to data limiting.
-The average of all of the confidence intervals is 7 to 23, indicating that a positive effect on patient count is reasonable.
+The result from the main specification is that EHR exposure increases patient count by 35.
+Under any data or specification adjustments, this estimate is nearly identical.
 
@@
 My main finding is that EHR exposure increases claims per patient by around .3.
-However, various robustness checks reveal that this is overestimated.
-That is supported in this specification chart as well, where around half of the estimated confidence intervals include zero.
-Thus, I make no conclusion that EHR exposure increases claims per patient.
+However, various robustness checks reveal that this is overestimated, and a third of the various specifications yield negative coefficients.
+Specifically, anticipation and limiting sample years seem to flip the sign.
+Since the main finding is not consistent across various specifications, I make no conclusion that EHR exposure affects claims per patient.
 
@@
 
+As referenced in the main paper, the effect of EHR exposure on patient count could be biased under any demand-driven events happening at the same time as EHR adoption in hospitals.
+The only nation-wide demand altering event that I know of during the time of this sample is the Affordable Care Act, which took effect mainly in 2014.
+If this is driving my positive finding, this would show up in a de-aggregated treatment effects by year of exposure.
+In Figure \ref{fig:patientgroup}, I show that the effects persist in all years of exposure, not just in 2014.
+
 \begin{figure}[ht!]
@@
 
+\subsection{Billing Activity as Cause for Work Setting Shift}
+
+I merge the CMS Public Utilization File on Providers by Service (PUF), which provides data beginning in 2013.
+For the purpose of this exercise, I therefore consider only years 2013-2017.
+Using Provider NPI, I merge in the total number of claims and the number of claims that are specific to primary care services in hospitals and offices.
+These are listed on the y-axis of Figure \ref{fig:claims_estimates}.
+Then, using the same difference-in-differences methodology as the main results, I estimate the effect of EHR exposure on physician fraction of claims in each category.
+
+\begin{figure}
+    \centering
+    \caption{Effect of EHR Exposure on Type of Claims}
+    \includegraphics[width=\textwidth]{Objects/claim_type_results.pdf}
+    \label{fig:claims_estimates}
+\end{figure}
+
+The estimates and 95\% CI are shown in Figure \ref{fig:claims_estimates}.
+After EHR exposure, the entire substitution from hospital to outpatient services is driven by a decrease in subsequent care in the hospital and an increase in established patient visits in an office.
+Subsequent visit in a hospital refers to evaluation/management of a patient during their hospital admission, while an established patient visit in an office is when a patient who has been seen by the physician in the last three years returns with a new or existing condition.
+While both services can be performed by physicians who specialize in primary care, they are not substitutes.
+In other words, the hospitalists seem to be performing different tasks after EHR exposure, not simply billing the same task in a different location.
+
 \subsection{Alternative Estimators}\label{app:estimators}
@@
 where outcome$_{i,t}$ is one of the five outcomes discussed previously, rel\_year$_j$ is an indicator equal to 1 if year $t$ is $j$ years relative to treatment, $\delta_i$ are physician fixed effects, and $\gamma_t$ are year fixed effects.
-These results from each outcome are presented in Figure \ref{fig:twfe}. These results show similar patterns to the main specification, with slight variations in magnitude.
+These results from each outcome are presented in Figure \ref{fig:twfe}.
+These results show similar patterns to the main specification, with slight variations in magnitude and more indications of a pre-trend.
 
@@
 One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}).
-This approach is commonly known as two stage difference in differences and is very clever.
-However, the approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
+This approach is commonly known as two stage difference-in-differences.
+The approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
 Nevertheless, I limit the years of data to 2009-2015 and present estimates as a percentage of the mean of each variable in Figure \ref{fig:estimators}.
-Generally, the results have the same sign with a slightly larger magnitude, driven by the inclusion of never-treated units.
+Generally, the results have the same sign with a slightly larger magnitude.
 
@@
 These average treatment effects as a percentage of the mean of each variable are shown in Figure \ref{fig:heter}.
-The results do not differ along these dimensions.
-This is fascinating in that technology is creating these adverse effects across completely different practice environments.
+The results do not differ along these dimensions, implying that a new technology affects hospitalists across many different settings.
 

### 2024-03-25 Paper/PaperEdit.tex

@@
 Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists, and uncorrelated with other major changes to the hospitalist work environment.
-I mitigate concerns of the latter by analyzing whether other observable hospital changes are correlated with EHR adoption, and find no evidence that this could be biasing my results.
-Additionally, my main findings are not sensitive to a plethora of robustness checks.
+I mitigate concerns of the latter by analyzing whether other observable hospital changes are correlated with EHR adoption, and find no evidence that this could be biasing the results.
+Additionally, the main findings are not sensitive to a plethora of robustness checks.
 The results suggest that a surge of hospitalists left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings.
@@
 I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate robustness to this threshold in Appendix \ref{sec:changes}).
-I use this data to construct the dependent variables for my analysis.
+I use this data to construct the dependent variables in my analysis.
 
@@
 The details of such thresholds are explained in Appendix \ref{app:data}.
-To ensure that my results  are not sensitive to various choices of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations.
+To ensure that the results  are not sensitive to various choices of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations.
 The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.
@@
 
-As discussed in Section \ref{sec:worksetting}, hospitalists could be switching to work settings either because of their own preferences to avoid technology, or because EHRs lead hospitals to re-allocate workers to hospital-owned outpatient settings.
-While I do not observe individual physician contract type with each hospital, I do observe hospital level information about the nature of aggregate physician relationships.
+As discussed in Section \ref{sec:worksetting}, hospitalists could be switching work settings either because of their own preferences to avoid technology, or because EHRs, through improved substitution between work settings, cause hospitals to re-allocate workers to hospital-owned outpatient settings.
+Due to lack of granular EHR data, it is not clear whether acquired outpatient facility EHRs are interoperable with inpatient EHRs, even if the two have the same EHR vendor.
+Outpatient EHRs are designed to be much more simplistic than inpatient EHRs which communicate across departments.
+The main purpose of an outpatient EHR is to track patient data, but it is unlikely this tracking covers the entire health system in the years covered by the data, before interoperability initiatives began.
+Still, it is theoretically possible that EHRs facilitate easier substitution between facilities.
+Therefore, I investigate this mechanism through a heterogeneity analysis of hospitals with different contract types with physicians, which will shut down certain channels of re-allocation of resources.
+
+I do not observe individual physician contract type with each hospital, but I do observe hospital level information about the aggregate nature of their physician contracts.
 To investigate the potential mechanism of hospital allocation, I present a heterogeneity analysis for all outcomes broken down by level of integration the physician is exposed to in their connected hospitals.
 From least integrated (contracts with multiple facilities) to most integrated (full salaried hospitalists), results are broken down by Independent Practice Association (IPA), Open Physician-Hospital Organization (OPHO), Closed Physician-Hospital Organization (CPHO), and Integrated Salary Model (ISM).
-In practice, hospitals can only re-allocate hospitalists to outpatient facilities if the hospitalist is heavily integrated.
-However, estimates do not differ between the types of integration relationships, indicating that the results are not driven entirely by such re-allocation.
+In practice, hospitals drive the re-allocation of work setting only if if the hospitalist is heavily integrated.
+For contracted hospitalists, the choice remains whether to contract with a hospital who wants them to work in an outpatient setting.
+However, estimates do not differ between the types of integration relationships, indicating that the results are not driven entirely by hospital-owned outpatient facilities.
+I present summary statistics by contract type and estimates of this heterogeneity analysis for all outcomes in Appendix
 
@@
     \caption{Effect of EHR Exposure on Outcomes by Integration Level}
-    \includegraphics[width=\textwidth]{Objects/heterog_plot_int.pdf}
+    \includegraphics[width=.7\textwidth]{Objects/heterog_plot_int_office.pdf}
     \label{fig:het_int}
@@
 
-Additionally, I investigate whether this finding is potentially due EHRs changing the type of billing performed by hospitalists.
-For example, maybe they are doing the same services in the same place but billing the differently.
+Another proxy for contract type is whether the hospitalists works with one system or is connected to facilities from different systems.
+I present the results from this heterogeneity analysis in Figure \ref{fig:heter}, and find no difference in effects between the two groups.
+
+Finally, I investigate whether this finding is potentially due EHRs changing the type of billing performed by hospitalists.
+For example, they could be performing the same services in the same location, but coding the billing information differently.
 I do not find this to be the case.
-This analysis can be found in Appendix
+This analysis can be found in Appendix \ref{app:coding}.
 
@@
 A major change in the U.S.
-health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014.
-If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count.
+health care system, the Affordable Care Act (ACA) was passed in 2010, and many changes were made that could affect the demand for health care, such as insurance expansions, and hospital quality initiatives.
+In order for additional changes to be biasing my results, they must be taking place in such a way that they either cause EHR adoption in hospitals or are differentially affecting EHR exposed hospitalists vs.
+non-EHR exposed hospitalists.
+I cannot completely rule this out, but it seems unlikely that any one change is causing a large and immediate effect for every treated group year.
+If the results are driven primarily by the group of hospitals who implemented EHRs in 2012 (when pay-for-performance initiatives began) or 2014 (when Medicaid expansion went into effect), one might be concerned that the legislation is driving the increased patient count.
 The ACA had potentially opposing effects on Medicare patients: they may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population.
-A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change should not drive my results.
+A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that this particular change should not drive the results.
 Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup} of the appendix.
-Each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups affected by the ACA are not driving the results.
+Despite different EHR exposure years, each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups most affected by the ACA are not driving the main findings.
 
@@
 The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group.
-To investigate whether my results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}).
+To investigate whether these results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}).
 For these bounds to be valid, a monotonicity assumption must hold on selection.
 That is, treatment can only affect selection in one direction.
-In my case, that means that EHR exposure increases the likelihood of exiting or moving workplace settings and no hospitalists remain in the sample under treatment who would have exited or moved in the absence of EHR exposure.
+In this case, that means that EHR exposure increases the likelihood of exiting or moving workplace settings and no hospitalists remain in the sample under treatment who would have exited or moved in the absence of EHR exposure.
 Since I find in earlier analyses that hospitalists are more likely to move and exit due to EHRs, this assumption is reasonable.
@@
 This bounding yields an interval estimate of 54.6 patients to 125.4 patients, compared to the main result of 28 patients.
-The bound therefore suggests selection of physicians pushes the estimate down.
+The bound therefore suggests selection causes a downward bias in the estimated effect, but the implications of the findings remain.
 
@@
 However, the second year after exposure is when I observe that largest effect on exiting, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in hospitalists.
+Further, I find that EHR adoption is not correlated with the aggregate number of physicians working in the hospital, shown in Table \ref{tab:hosp_assoc}, suggesting that hospitals replace those who exit relatively quickly.
 
@@
 I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists.
-However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero.
+However, when I analyze this relationship under different data specifications, shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero.
 Anticipation and limiting sample years cause the effect to disappear or even become negative.
-Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs and I do not make any strong conclusions.
+Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs, and I do not make any strong conclusions.
 
@@
 
+One concern is that EHR adoption could be occurring simultaneously with a host of other hospital initiatives to improve productivity or quality, which may also affect physician work environment.
+If true, EHR adoption is not exogenous and the results found above are biased.
+Here, I discuss the implications of this type of endogeneity for each outcome, and present evidence that hospital behaviors are not changing in a way that is correlated with EHR adoption.
+
+Anecdotally, there is no other change made in hospitals during this time period that caused the frustration and burn out that EHRs did.
+Thus, on average, it is unlikely that other hospital level changes are imposing a burden on hospitalist work environment that could explain physician exit or work setting changes.
+If work environment changes are occurring, it is more likely that they offset some of the cost imposed by EHRs, and that my estimates for exit and work setting are a lower bound of the true effect of EHRs.
+
+Therefore, this endogeneity is more of a concern when considering productivity of hospitalists.
+If other work environment factors correlated with EHR adoption improve productivity, then the increase I find in productivity could actually be due to other factors.
+I investigate this endogeneity by constructing a hospital level data set from 2009 to 2015 with various hospital-level outcomes that EHR adoption may be correlated with.
+While not all hospital behaviors are observable, it is very likely that if the hospital is making a large portfolio of changes along with EHR, we will see a correlation between EHR adoption and observed hospital behaviors.
+I categorize the behaviors into two categories: non-EHR investments and organizational changes, and level of hospital integration with physicians.
+Non-EHR investments and organization changes include building costs (ex.
+did the hospital renovate or expand?), equipment costs (ex.
+did the hospital invest in other machines?), investment in subsidiaries, changes in the main hospital administrator, and number of beds.
+Relationships with physicians include the number of physicians employed, and indicators for the types of vertical integration with physicians: independent practice organization (IPA), open physician-hospital organization (OPHO), closed physician-hospital organization (CPHO), and fully integrated organization (FIO), listed from the lowest level of integration to the highest.
+An IPA has loose contracts with physicians, but the physicians remain independent, an OPHO adds administration tasks to the physician, a CPHO contracts with specific physicians, and a FIO hires physicians as salaried employees (\cite{short2017integration}).
+The expense variables are pulled from the Healthcare Cost Reporting Information System (HCRIS) database, and the other hospital characteristics come from the AHA data.
+
+\import{Table Code}{hosp_analysis.tex}
+
+I estimate the effect of EHR adoption on each of these outcomes using the same \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) staggered difference-in-differences methodology.
+I present coefficients and standard errors from this analysis in Table \ref{tab:hosp_assoc}.
+Again, I do not present these to suggest that EHR may have a causal impact on other behaviors, but to observe whether the behaviors seem to be happening in conjunction with EHR adoption.
+Based on the confidence intervals, none of the behaviors considered seem to be occurring simultaneously with EHR adoption on average.
+While I cannot rule out all changes that could be occurring at the hospital level, the lack of movement in other important behaviors mitigates the concern.
+
 \section{Conclusion}
@@
 Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient.
-My findings suggest that EHRs make hospitalists more productive through the number of patients they see.
+These findings suggest that EHRs make hospitalists more productive through the number of patients they see.
 However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, as there may be a trade-off between comprehensive patient care and productivity.
@@
 I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs.
-Some pairs are duplicates from the hospital being listed first or second in the shared patient data (being listed first means the patient went to NPI 1 first, but my analysis does not depend on the order of care, only the relationship between hospital and physician as a whole).
+Some pairs are duplicates from the hospital being listed first or second in the shared patient data (being listed first means the patient went to NPI 1 first, but the analysis does not depend on the order of care, only the relationship between hospital and physician as a whole).
 I combine duplicates into one observation, summing the same day count variable.
@@
 This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation.
-I drop any physicians who graduated medical school after 2004, since graduating after that means leaving residency or graduating during the span of my main data, 2009-2017, and potentially exhibiting labor market changes that seem associated with EHRs but are not.
+I drop any physicians who graduated medical school after 2004, since graduating after that means leaving residency or graduating during the span of the main data, 2009-2017, and potentially exhibiting labor market changes that seem associated with EHRs but are not.
 
@@
 I also create a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.
+Since exit is relatively rare, I present a distribution showing where they occur geographically.
+The distribution is fairly uniform.
+
+\begin{figure}[ht!]
+    \centering
+    \caption{Distribution of Exiters}
+    \includegraphics[width=\textwidth]{Objects/retire_dist_map.pdf}
+    \label{fig:retire_dist}
+\end{figure}
 
@@
 The only nation-wide demand altering event that I know of during the time of this sample is the Affordable Care Act, which took effect mainly in 2014.
-If this is driving my positive finding, this would show up in a de-aggregated treatment effects by year of exposure.
+If this is driving the positive finding, this would show up in a de-aggregated treatment effects by year of exposure.
 In Figure \ref{fig:patientgroup}, I show that the effects persist in all years of exposure, not just in 2014.
@@
 
-\subsection{Billing Activity as Cause for Work Setting Shift}
+\subsection{Coding Activity as Cause for Work Setting Shift}\label{app:coding}
 
@@
 There is a robust recent literature considering the issues with estimating heterogeneous treatment effects with a two way fixed effects specification, the classic approach in staggered treatment setups.
-To avoid these problems, in my main specification I use the average group time effect estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
+To avoid these problems, in the main specification I use the average group time effect estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}).
 In this section, I discuss other potential estimators and why I ultimately included average group time treatment effects as the main specification.
@@
 This approach is commonly known as two stage difference-in-differences.
-The approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
+The approach requires a group of never-treated units as the comparison group, a drawback of this study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample.
 Nevertheless, I limit the years of data to 2009-2015 and present estimates as a percentage of the mean of each variable in Figure \ref{fig:estimators}.
@@
 
+\import{Table Code}{int_sumstats.tex}
+
+\begin{figure}
+    \centering
+    \caption{Heterogeneity Analysis, Hospital Contract Types}
+    \includegraphics[scale=.6]{Objects/heterog_plot_int.pdf}
+\end{figure}
+
 \end{document}

### 2024-09-07 Paper/PaperEdit.tex

--- 2023-05-31 Paper/PaperEdit.tex
@@
 \begin{abstract}
-    Technology adoption in health care is of great importance to policy makers as some technologies are expected to radically improve the U.S.
-health care system.
-However, some new technologies can be burdensome for physicians and lead to frustration.
-I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S.
+    Technology adoption in health care has the potential to improve patient outcomes and, in some cases, decrease costs.
+However, new technologies can be burdensome for physicians to keep up with and lead to frustration.
+I add to our understanding of how new technology can affect those who use it by investigating changes in physician behavior as a result of a major technology shift in U.S.
 hospitals, electronic health records (EHRs).
 I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes.
-I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs.
-Further, those exposed to EHRs see more patients due to the technology.
-These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
+I find that hospitalists are more likely to retire, and more likely to shift work to an outpatient setting due to EHRs.
+However, those who continue using EHRs in the hospital see more patients due to the technology.
 \end{abstract}
@@
 EHRs are computer systems that house digital medical records and sometimes have additional functionalities.
-EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time.
+EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{Meaningful use (MU) standards proceeded in three stages over time.
 In Stage 1 (2010), MU focused on data capturing and sharing.
@@
 For a hospitalist, the hospital's decision to adopt an EHR is plausibly exogenous.
-Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical environments, measured based on no future claims, (2) substituting hospital work for office work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.
+Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical work altogether, measured based on no future claims, (2) substituting hospital work for office/outpatient work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.
 
@@
 Further, I consider the primary time period in which EHRs were rapidly implemented.
-This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting physicians' daily life.
+This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting hospitalists' daily life.
 Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework.
@@
 Limiting the sample to those who remained in clinical settings, I find that hospitalists were 30\% more likely to start seeing patients in an office setting at the time of exposure.
-Finally, I find that EHRs lead to a patient count increase of at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.
+Finally, I examine productivity of hospitalists who do not exit or change work settings, and I find that EHRs lead to a patient count increase of at least 20 patients per year.
+This increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.
 
@@
 First, I assume that hospitalists are constrained to use the technology implemented in their work setting.
-Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals.
+Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists.
 That is, a hospitalist employed by the hospital does not choose their own EHR exposure.
-My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed.
-Thus, EHR implementation may have intensified access to care issues for patients.
-This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.
-\textcolor{red}{reword the implications and check the numbers are correct}
+My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of hospitalists who left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings.
+Thus, EHR implementation may have intensified access to care issues for hospitalized patients.
+This unintended consequence of incentivizing rapid technology adoption highlights the need for careful consideration of technologies and policies that might negatively affect work environment.
 
@@
 The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}).
-Another common issue physicians face is the lack of usability of EHR systems, that functions take minutes to either locate or load.
+Another common issue physicians face is the lack of usability of EHR systems, with functions taking minutes to locate or load.
 Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.
@@
 For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting.
-However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist.
+However, there are two types of physicians who can be categorized as hospitalists based on this definition: a formal hospitalist who is solely employed by the hospital, or an internist who contracts their services to the hospital.
 Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings.
 Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting.
-While this distinction is potentially important for some of the decisions outlined below, the data do not explicitely distinguish between the two.
+While this distinction is potentially important for some of the decisions outlined below, the data do not explicitly distinguish between the two.
 I will discuss this limitation in more detail when relevant.
@@
 I use this data to construct the dependent variables for my analysis.
-Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and billing activity.
 
@@
 
-I aggregate the data to the physician level.
-I drop physicians who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated.
+I aggregate the data to the hospitalist level.
+I drop those who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated.
 The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated".
 Finally, I include a physician's graduation year from Physician Compare (provided by CMS).
-I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
+I limit to hospitalists who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship.
 The details of such thresholds are explained in Appendix \ref{app:data}.
@@
 These hospitalists see 325 patients and file 4 claims per patient on average.
-Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.
+Consistent across all samples, hospitalists tend to work approximately 1.5 hospitals and 1.3 systems.
 
@@
 Being exposed to an EHR leads to a .2 ppt increase in the likelihood of exiting in the first year after exposure, and a .4 ppt increase in the second year after exposure.
-While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03.
+Coefficients remain similar 3 and 4 years after exposure, though estimates become noisier due to smaller sample not-yet-treated groups.13 While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03.
 That is, the effect represents a 7\% (13\% in the second year) increase in the likelihood of exiting relative to the mean.
@@
 \end{figure}
+
+As referenced in the main paper, the effect of EHR exposure on patient count could be biased under any demand-driven events happening at the same time as EHR adoption in hospitals.
+The only nation-wide demand altering event that I know of during the time of this sample is the Affordable Care Act, which took effect mainly in 2014.
+If this is driving my positive finding, this would show up in a de-aggregated treatment effects by year of exposure.
+In Figure \ref{fig:patientgroup}, I show that the effects persist in all years of exposure, not just in 2014.
 
